## Why

The six sea turtles and the North Atlantic right whale all have a published OBIS (MPA Europe)
species distribution model on exactly our `global05` grid
(`s3://obis-maps/sdm/species/taxonid={AphiaID}/model=mpaeu/`). A model surface is not a range: the
loggerhead model calls tens of thousands of Alaska cells suitable, and the v9 right whale surface (the
IUCN range at a flat 100) covers the Gulf. This notebook scores five candidate rules for turning a
model surface into a range, on **one scorecard**, by **polygon** (Program Areas and US EEZ
subregions, not latitude / longitude boxes), using only what OBIS published, so nothing here needs a
refit.

| | candidate | needs a refit |
|---|---|---|
| A | OBIS's own recipe: threshold P10, inside the convex hull of fitted records, keeping only patches within 100 km of a record | no |
| B | Tim White's buffer around occurrences: 150, 250, 400 km | no |
| C | the v7.1 rule: AquaMaps suitability of 0.5 or more inside the range, critical habitat kept | no |
| D | threshold plus uncertainty: value minus one bootstrap spread clears the threshold; not an extrapolation | provisional here |
| E | D plus the regional test with search effort | provisional here |

The notebook takes one parameter, `model`. With `mpaeu` (the default) it scores OBIS's published
models, and D and E are **provisional**: their threshold is the published P10, which is fitted on
training data. With one of our own runs (`og`, `ogm`, a control) the threshold of D and E is the
**held-out** one, from the fold predictions our patched pipeline saves. The rule itself is the same
either way and lives in `msens` (`R/range.R`, `tests/testthat/test-range.R`), which this notebook
only calls.

```{mermaid}
flowchart LR
  obis["published OBIS models<br/>ensemble, bootstrap sd,<br/>masks, MESS, thresholds, points"] --> cell["og_cell<br/>one row per species x cell"]
  v9["v9 inputs on global05<br/>AquaMaps, IUCN / SWOT ranges,<br/>critical habitat"] --> cell
  sg["speciesgrids<br/>OBIS + GBIF records<br/>per species x cell"] --> eff["search effort<br/>target-group records"]
  zones["zone_cell<br/>Program Areas, EEZ subregions"] --> reg["regions"]
  cell --> rules["rules A to E<br/>msens::range_*"]
  eff --> rules
  reg --> rules
  rules --> card["scorecard<br/>species x region x rule"]
```

```{r}
#| label: setup
librarian::shelf(
  arrow, DBI, dplyr, duckdb, FNN, fs, glue, here, jsonlite, knitr, msens, purrr, readr, stringr,
  terra, tibble, tidyr,
  quiet = T)
options(readr.show_col_types = F, knitr.kable.NA = "")
source(here("libs/paths.R"))

stopifnot("msens >= 0.54.0 has the range rule" = packageVersion("msens") >= "0.54.0")

# inputs ----
dir_sdm   <- path_expand("~/_big/sdm")
dir_obis  <- glue("{dir_sdm}/obis")              # scripts/og/fetch_obis_published.sh (S3 layout kept)
dir_v9in  <- glue("{dir_sdm}/v9_inputs")         # v9 input surfaces, fetched below from the public store
dir_card  <- glue("{dir_sdm}/scorecard")         # this notebook's working tables
MODEL     <- params$model
is_pub    <- MODEL == "mpaeu"
lbl       <- if (is_pub) "published" else MODEL   # names this run's tables and csv files
stopifnot(grepl("^[a-z0-9]+$", MODEL))
dir_mod   <- if (is_pub) glue("{dir_obis}/species") else path_expand(Sys.getenv("OG_OUT", glue("{dir_sdm}/og")))
db_card   <- glue("{dir_card}/{lbl}.duckdb")
db_sg     <- path_expand(glue("{dir_big}/speciesgrids_global05.duckdb"))
atlas_url <- "https://s3.us-east-1.amazonaws.com/oceanmetrics.io-public/marine-atlas"
dir_create(c(dir_v9in, dir_card, here("data/og")))

redo <- Sys.getenv("REDO_OG_CELLS", "0") == "1"   # rebuild the cached cell table

# species ----
d_sp <- tribble(
  ~taxon,  ~common,                       ~sci,                     ~grp,
  137205L, "Loggerhead",                  "Caretta caretta",        "turtle",
  137206L, "Green",                       "Chelonia mydas",         "turtle",
  137207L, "Hawksbill",                   "Eretmochelys imbricata", "turtle",
  137208L, "Kemp's ridley",               "Lepidochelys kempii",    "turtle",
  137209L, "Leatherback",                 "Dermochelys coriacea",   "turtle",
  220293L, "Olive ridley",                "Lepidochelys olivacea",  "turtle",
  159023L, "North Atlantic right whale",  "Eubalaena glacialis",    "whale")
ids_turtle <- c(137205L, 137206L, 137207L, 137208L, 137209L, 220293L, 344093L)  # + flatback

# rule parameters (the same for every species) ----
SENS      <- 0.90                # sensitivity the threshold fixes
BUF_KM    <- c(150, 250, 400)    # rule B
PATCH_KM  <- 100                 # rule A: OBIS's max_dispersal
AM_MIN    <- 50                  # rule C: AquaMaps probability >= 0.5
UNIT_AGG  <- 10L                 # regional test unit: 10 x 10 cells = 0.5 degree blocks
ALPHA     <- 0.05
RATIO     <- 0.25                # regional test: dropped only if the record rate is below this share of the rate elsewhere

# a file of one species' model run, in the OBIS layout: {root}/taxonid={id}/model={MODEL}/{sub}/taxonid={id}_model={MODEL}_{tail}
f_mod <- function(id, sub, tail, must = T) {
  d <- glue("{dir_mod}/taxonid={id}/model={MODEL}/{sub}")
  f <- dir_ls(d, regexp = glue("taxonid={id}_model={MODEL}_{tail}$"), fail = F)
  if (must) stopifnot("expected exactly one file" = length(f) == 1)
  if (length(f)) as.character(f[1]) else NA_character_
}
f_ens  <- function(id) glue("{dir_mod}/taxonid={id}/model={MODEL}/predictions/taxonid={id}_model={MODEL}_method=ensemble_scen=current_cog.tif")
f_pub  <- function(id, s) glue("{dir_obis}/species/taxonid={id}/model=mpaeu/", glue(s, id = id))   # always OBIS's own files
have   <- file_exists(map_chr(d_sp$taxon, f_ens))
if (is_pub) stopifnot("run scripts/og/fetch_obis_published.sh first" = all(have))
if (!all(have)) message("no ", MODEL, " run yet for: ", paste(d_sp$common[!have], collapse = ", "))
d_sp   <- d_sp[have, ]
stopifnot("no species has a run of this model" = nrow(d_sp) > 0)
need   <- map_chr(d_sp$taxon, f_ens)

g      <- grid_spec_for("global05")
r_cid  <- rast(cellid_tif)
stopifnot(
  "the OBIS rasters are not on global05" = compareGeom(rast(need[1]), r_cid, stopOnError = F),
  ncell(r_cid) == 7200 * 3600)
# cell_id is the pixel index on this grid (asserted on a sample; the cell-id COG is a lookup image)
local({
  i <- seq(1, ncell(r_cid), length.out = 2e5) |> round()
  v <- r_cid[i][, 1]
  stopifnot(all(v[!is.na(v)] == i[!is.na(v)]))
})
unit_of <- function(cell_id, agg = UNIT_AGG) {   # block id of a global05 cell
  row <- (cell_id - 1L) %/% 7200L; col <- (cell_id - 1L) %% 7200L
  (row %/% agg) * (7200L %/% agg) + (col %/% agg)
}
```

## Inputs

### The models scored here (`r MODEL`)

```{r}
#| label: obis-meta
d_meta <- map_dfr(d_sp$taxon, function(id) {
  j  <- read_json(f_mod(id, ".", "what=log.json"))
  th <- read_parquet(f_mod(id, "metrics", "what=thresholds.parquet")) |>
    filter(model == "ensemble")
  tibble(
    taxon     = id,
    hab_depth = unlist(j$hab_depth),
    n_fit     = unlist(j$model_fit_points),
    good      = paste(unlist(j$model_good), collapse = ", "),
    thr_mtp   = th$mtp * 100,
    thr_p10   = th$p10 * 100,
    thr_sss   = th$max_spec_sens * 100)
})
d_sp <- left_join(d_sp, d_meta, by = "taxon")
d_sp |>
  select(common, taxon, hab_depth, n_fit, good, thr_mtp, thr_p10, thr_sss) |>
  kable(digits = 0, caption = "Models scored. Thresholds are on the ensemble's 0 to 100 scale, all fitted on TRAINING data: minimum training presence, P10 (keeps 90 % of fitted records) and maximum sensitivity + specificity.")
```

### The threshold of rules D and E

```{r}
#| label: threshold
# held-out predictions of each accepted algorithm, saved per fold by the patched pipeline
# (metrics/..._method={m}_what=cvpred.parquet; cvpredtrain = the fold's model on its training rows).
# the ensemble is the per-row median of the accepted algorithms, as on the map
algo_file <- c(maxent = "maxent", rf = "rf_classification_ds", xgboost = "xgboost")
cv_read <- function(id, what) {
  good <- str_split_1(d_sp$good[d_sp$taxon == id], ", ")
  map_dfr(good, ~ read_parquet(f_mod(id, "metrics", glue("method={algo_file[.x]}_what={what}\\.parquet"))) |> mutate(algo = .x))
}
cv_ens <- function(d) d |> summarise(pred = median(pred) * 100, .by = c(fold, idx, presence))
if (is_pub) {
  d_sp$thr <- d_sp$thr_p10
  d_thr    <- NULL
} else {
  d_thr <- map_dfr(d_sp$taxon, function(id) {
    te <- cv_read(id, "cvpred"); tr <- cv_read(id, "cvpredtrain")
    stopifnot(n_distinct(count(te, algo)$n) == 1)   # the algorithms share rows and folds
    h  <- heldout_threshold(cv_ens(te), cv_ens(tr), sens = SENS)
    tibble(
      taxon = id, thr = h$thr, sens = h$pooled$sens, spec = h$pooled$spec,
      # what the training threshold of each fold delivers on the rows that fold never saw
      train_thr = mean(h$folds$thr), train_thr_sens = sum(h$folds$tp) / sum(h$folds$tp + h$folds$fn),
      folds = list(h$folds))
  })
  d_sp <- left_join(d_sp, select(d_thr, taxon, thr), by = "taxon")
  write_csv(
    d_thr |> select(taxon, folds) |> unnest(folds), here(glue("data/og/{lbl}_heldout_folds.csv")))
}
```

```{r}
#| label: threshold-table
#| eval: !expr '!is_pub'
d_thr |>
  left_join(select(d_sp, taxon, common, thr_p10), by = "taxon") |>
  transmute(
    common, `training P10` = thr_p10, `held-out threshold` = thr,
    `sensitivity` = sens, `specificity vs background` = spec,
    `mean fold training threshold` = train_thr, `sensitivity it delivers held-out` = train_thr_sens) |>
  kable(digits = c(0, 0, 1, 2, 2, 1, 2), caption = glue("The held-out threshold (keeps {100 * SENS} % of the presences each fold's model never saw) beside the training P10, ensemble scale 0 to 100. The last two columns fix the threshold on each fold's training rows and count on its test rows: the sensitivity a training threshold really delivers."))
```

`r if (is_pub) "For the published models no fold predictions exist, so rules D and E use the training P10." else "Rules D and E use the held-out threshold; P10 and rule A keep the training P10, as OBIS applies it."`

### v9 input surfaces

AquaMaps, the range polygons and critical habitat as v9 ingested them, already on `global05`, from the
public content-addressed store (`tables/native_asset.parquet` names the COG of each input).

```{r}
#| label: v9-inputs
f_na <- glue("{dir_card}/v9_native_asset.parquet")
if (!file_exists(f_na)) download.file(glue("{atlas_url}/v9/tables/native_asset.parquet"), f_na, quiet = T, mode = "wb")
d_asset <- read_parquet(f_na) |>
  filter(
    ms_merge_key %in% glue("ms_merge|WORMS:{d_sp$taxon}"),
    asset_type == "cog", representation == "model") |>
  mutate(
    taxon = as.integer(str_extract(ms_merge_key, "[0-9]+$")),
    file  = glue("{dir_v9in}/{taxon}/{ds_key}__{content_hash}.tif"))
walk2(d_asset$asset_url, d_asset$file, function(u, f) {
  if (file_exists(f)) return()
  dir_create(path_dir(f)); download.file(u, f, quiet = T, mode = "wb")
})
# a v9 input as (cell_id, val) on global05, whatever window the COG was written on
v9_cells <- function(id, ds) {
  fs <- unique(d_asset$file[d_asset$taxon == id & d_asset$ds_key %in% ds])
  if (!length(fs)) return(tibble(cell_id = numeric(), val = numeric()))
  map_dfr(fs, function(f) {
    r <- rast(f)[[1]]
    stopifnot(all(abs(res(r) - 0.05) < 1e-6))   # float noise in the published geotransform
    i <- which(!is.na(values(r, mat = F)))
    tibble(cell_id = cellFromXY(r_cid, xyFromCell(r, i)), val = values(r, mat = F)[i])
  }) |>
    summarise(val = max(val), .by = cell_id)
}
d_asset |>
  count(taxon, ds_key) |>
  pivot_wider(names_from = ds_key, values_from = n) |>
  left_join(select(d_sp, taxon, common), by = "taxon") |>
  relocate(common) |>
  kable(caption = "v9 inputs per species (number of surfaces). `ch_*` = critical habitat, `rng_*` = range polygons, `am` / `ax` = AquaMaps 1.0 / 2.0.")
```

### Regions, by polygon

Program Areas (the 2026 program: 17 in Alaska, 3 off California, 2 in the Gulf, none in the Atlantic)
grouped by BOEM region, and the four US EEZ subregions. Both come from the committed polygon-to-cell
extractions (`zones/{zone_set}/global05/zone_cell.parquet`), the same tables scoring uses. The
regional test runs on the four EEZ subregions, which tile US waters; the Program Area groups are
reported beside them because that is where scores are published.

```{r}
#| label: regions
d_pa <- sf::st_read(glue("{dir_big}/v2/ply_programareas_2026.gpkg"), quiet = T) |>
  sf::st_drop_geometry() |>
  select(zone_key = programarea_key, region_key)
zc_pa <- read_parquet(glue("{dir_big}/zones/programarea_2026-01/global05/zone_cell.parquet")) |>
  left_join(d_pa, by = "zone_key")
zc_sr <- read_parquet(glue("{dir_big}/zones/subregion_2025-06/global05/zone_cell.parquet"))
stopifnot(!anyNA(zc_pa$region_key), setequal(zc_sr$zone_key, c("AK", "AT", "GA", "PA")))

lbl_rgn <- c(AK = "Alaska", AT = "Atlantic", GA = "Gulf", PA = "Pacific")
d_scregion <- bind_rows(
  zc_pa |> distinct(region = paste("PA", lbl_rgn[region_key]), cell_id),
  zc_sr |> distinct(region = paste("EEZ", lbl_rgn[zone_key]),  cell_id))
d_region <- zc_sr |> distinct(region = paste("EEZ", lbl_rgn[zone_key]), cell_id)   # the rule's regions
d_scregion |> count(region, name = "cells") |>
  kable(format.args = list(big.mark = ","), caption = "Regions on the scorecard. `PA` = Program Areas of a BOEM region, `EEZ` = US EEZ subregion (the Pacific one includes Hawaii and the Pacific islands).")
```

### Records and search effort

*Records* of a species are OBIS's quality-controlled points for it (one per cell, 1950 on; the file
the published model was fitted from). *Search effort* is where somebody recorded the species'
**target group**: a cell with any such record was looked at by someone who would have reported the
species. Two definitions, because the answer depends on it:

- `order`: the species' own order. All sea turtles for a turtle, all cetaceans for the right whale.
- `mega`: every air-breathing marine vertebrate (birds, mammals, turtles). Seabird and marine mammal
  surveys record turtles and whales when they see them.

Effort comes from `speciesgrids` (OBIS + GBIF records per species and `global05` cell).

```{r}
#| label: effort
con_sg <- dbConnect(duckdb(), db_sg, read_only = T)
d_spl  <- read_csv(glue("{dir_obis}/source/model=mpaeu/data/all_splist_20240724.csv"))
ids_cetacean <- d_spl |> filter(class == "Mammalia", order == "Cetartiodactyla") |> pull(taxonID) |> unique()

sg_cells <- function(where) dbGetQuery(con_sg, glue("SELECT DISTINCT cell_id FROM sg_cell WHERE {where}"))$cell_id
eff <- list(
  order = list(
    turtle = sg_cells(glue("aphia IN ({paste(ids_turtle, collapse = ',')})")),
    whale  = sg_cells(glue("aphia IN ({paste(ids_cetacean, collapse = ',')})"))),
  mega  = local({
    x <- sg_cells(glue("class IN ('Aves', 'Mammalia') OR aphia IN ({paste(ids_turtle, collapse = ',')})"))
    list(turtle = x, whale = x)
  }))
dbDisconnect(con_sg, shutdown = T)

d_occ <- map_dfr(d_sp$taxon, function(id) {
  read_parquet(glue("{dir_obis}/source/model=mpaeu/data/species/key={id}.parquet")) |>
    transmute(taxon = id, lon = decimalLongitude, lat = decimalLatitude, data_type)
}) |>
  mutate(cell_id = as.integer(cellFromXY(r_cid, cbind(lon, lat))))
d_fit <- map_dfr(d_sp$taxon, function(id) {
  read_parquet(f_mod(id, ".", "what=fitocc.parquet")) |>
    transmute(taxon = id, lon = decimalLongitude, lat = decimalLatitude)
}) |>
  mutate(cell_id = as.integer(cellFromXY(r_cid, cbind(lon, lat))))

tibble(
  effort = c("order: sea turtles", "order: cetaceans", "mega: birds + mammals + turtles"),
  n_species = c(length(ids_turtle), length(ids_cetacean), NA),
  cells  = c(length(eff$order$turtle), length(eff$order$whale), length(eff$mega$turtle))) |>
  mutate(us_cells = c(
    sum(eff$order$turtle %in% zc_sr$cell_id), sum(eff$order$whale %in% zc_sr$cell_id), sum(eff$mega$turtle %in% zc_sr$cell_id))) |>
  kable(format.args = list(big.mark = ","), caption = "Searched cells by effort definition (global, and inside the US EEZ).")
```

## Cell table

One row per species and cell that any rule could keep: the model's value, its bootstrap spread, the
extrapolation flag, OBIS's masks, the distance to the nearest record, and the v9 inputs.

```{r}
#| label: cells
to_xyz <- function(lon, lat) {
  lo <- lon * pi / 180; la <- lat * pi / 180
  cbind(cos(la) * cos(lo), cos(la) * sin(lo), sin(la))
}
# great-circle km from each query cell to the nearest point (kd-tree on the unit sphere)
near_km <- function(cell_id, lon, lat) {
  if (!length(lon)) return(rep(NA_real_, length(cell_id)))
  xy <- xyFromCell(r_cid, cell_id)
  ch <- get.knnx(to_xyz(lon, lat), to_xyz(xy[, 1], xy[, 2]), k = 1)$nn.dist[, 1]
  2 * 6371.0088 * asin(pmin(ch / 2, 1))
}

build_cells <- function(id) {
  sp  <- d_sp[d_sp$taxon == id, ]
  e   <- rast(f_ens(id))
  mk  <- rast(f_mod(id, "predictions", "(what=)?mask_cog\\.tif"))    # our runs write it without `what=`
  ms  <- rast(f_mod(id, "predictions", "what=mess_cog\\.tif"))[["current"]]
  # the bootstrap layer comes from a separate step; without it the spread is missing (the rule then
  # treats it as 0) and the notebook says so in the cell-table summary
  f_bc <- f_mod(id, "predictions", "method=ensemble_scen=current_what=bootcv_cog\\.tif", must = F)
  bc   <- if (is.na(f_bc)) NULL else rast(f_bc)
  stopifnot(compareGeom(e, mk, stopOnError = F), is.null(bc) || (compareGeom(e, bc, stopOnError = F) && names(bc)[1] == "sd"))

  val  <- values(e[[1]], mat = F)
  # the published MESS layer is the count of predictors outside the fitted range, on a 0.6 degree grid
  mess <- values(resample(ms, e[[1]], method = "near"), mat = F)

  ch   <- v9_cells(id, c("ch_nmfs", "ch_fws"))
  am   <- v9_cells(id, "am")
  rng  <- v9_cells(id, c("rng_iucn", "rng_fws", "rng_turtle_swot_dps"))
  iucn <- v9_cells(id, "rng_iucn")
  occ  <- d_occ$cell_id[d_occ$taxon == id]
  fit  <- d_fit[d_fit$taxon == id, ]

  # rule A, as OBIS codes it (functions/post_proc_functions.R limit_by_closest): mask, threshold P10,
  # label 8-connected patches, keep the patches within PATCH_KM of a fitted record
  patch_keep <- function(band) {
    m  <- values(mk[[band]], mat = F) == 1
    ok <- !is.na(val) & val >= sp$thr_p10 & !is.na(m) & m
    r  <- e[[1]]; values(r) <- ifelse(ok, 1L, NA_integer_)
    p  <- values(patches(r, directions = 8), mat = F)
    i  <- which(ok)
    near <- near_km(i, fit$lon, fit$lat) <= PATCH_KM
    ok & p %in% unique(p[i][near])
  }
  a_hull <- patch_keep("convex_hull")
  a_fitr <- patch_keep("fit_region")

  cells <- sort(unique(c(which(!is.na(val) & val >= 1), ch$cell_id, am$cell_id, rng$cell_id, occ)))
  cells <- cells[!is.na(cells)]
  tibble(
    taxon      = id,
    cell_id    = as.integer(cells),
    unit_id    = as.integer(unit_of(cells)),
    val        = val[cells],
    spread     = if (is.null(bc)) NA_real_ else values(bc[[1]], mat = F)[cells],
    sd_algo    = values(e[[2]], mat = F)[cells],
    extrap     = mess[cells] > 0,
    is_ch      = cells %in% ch$cell_id,
    in_hull    = values(mk[["convex_hull"]], mat = F)[cells] == 1,
    a_hull     = a_hull[cells],
    a_fitr     = a_fitr[cells],
    d_occ_km   = near_km(cells, d_occ$lon[d_occ$taxon == id], d_occ$lat[d_occ$taxon == id]),
    am         = am$val[match(cells, am$cell_id)],
    in_rng     = cells %in% rng$cell_id,
    in_iucn    = cells %in% iucn$cell_id)
}

con <- dbConnect(duckdb(), db_card)
source(here("libs/duckdb_budget.R")); duckdb_tune(con, glue("{dir_card}/tmp"))
done <- if (dbExistsTable(con, "og_cell")) dbGetQuery(con, "SELECT DISTINCT taxon FROM og_cell")$taxon else integer()
if (redo && length(done)) { dbExecute(con, "DROP TABLE og_cell"); done <- integer() }
for (id in setdiff(d_sp$taxon, done)) {
  message(glue("cells: {id} {d_sp$common[d_sp$taxon == id]}"))
  dbWriteTable(con, "og_cell", build_cells(id), append = T)
}

dbWriteTable(con, "og_thr",      select(d_sp, taxon, thr), overwrite = T)       # rules D and E
dbWriteTable(con, "og_p10",      transmute(d_sp, taxon, thr = thr_p10), overwrite = T)   # the reference column and rule A
dbWriteTable(con, "sc_species",  select(d_sp, taxon, common, grp, starts_with("thr_")), overwrite = T)
dbWriteTable(con, "og_region",   d_region,   overwrite = T)
dbWriteTable(con, "sc_region",   d_scregion, overwrite = T)
dbWriteTable(con, "og_fit",      distinct(d_fit, taxon, cell_id), overwrite = T)
dbWriteTable(con, "og_occ_cell", distinct(d_occ, taxon, cell_id) |> filter(!is.na(cell_id)), overwrite = T)
for (k in names(eff)) {
  d <- map_dfr(d_sp$taxon, ~ tibble(taxon = .x, cell_id = eff[[k]][[d_sp$grp[d_sp$taxon == .x]]]))
  dbWriteTable(con, glue("og_effort_cell_{k}"), d, overwrite = T)
}

dbGetQuery(con, "
  SELECT taxon, count(*) AS cells, count(*) FILTER (WHERE val >= 1) AS modeled,
         count(*) FILTER (WHERE is_ch) AS critical_habitat, count(*) FILTER (WHERE am >= 1) AS aquamaps,
         count(*) FILTER (WHERE in_rng) AS in_range, count(*) FILTER (WHERE extrap) AS extrapolated,
         count(*) FILTER (WHERE val >= 1 AND spread IS NULL) AS no_bootstrap_spread
  FROM og_cell GROUP BY taxon ORDER BY taxon") |>
  left_join(select(d_sp, taxon, common), by = "taxon") |> relocate(common) |>
  kable(format.args = list(big.mark = ","), caption = "Cell table, whole globe.")
```

## The right whale in the Gulf, redone by polygon

The plan's Gulf numbers came from a latitude / longitude box that also took in northeast Florida,
where the calving ground and four fitted records lie. The same rules, counted inside the polygons:

```{r}
#| label: whale-gulf
#| eval: !expr '159023L %in% d_sp$taxon'
xy <- function(id) { p <- xyFromCell(r_cid, id); tibble(cell_id = id, lon = p[, 1], lat = p[, 2]) }
w  <- dbGetQuery(con, "SELECT * FROM og_cell WHERE taxon = 159023") |> as_tibble()
w  <- left_join(w, xy(w$cell_id), by = "cell_id")
sw <- d_sp[d_sp$taxon == 159023L, ]
in_rg <- function(rg) w$cell_id %in% d_scregion$cell_id[d_scregion$region == rg]
w_rules <- list(
  "Lowest value at any record (MTP {sw$thr_mtp})"         = w$val >= sw$thr_mtp,
  "Best sensitivity + specificity ({round(sw$thr_sss)})"  = w$val >= sw$thr_sss,
  "P10 ({sw$thr_p10})"                                    = w$val >= sw$thr_p10,
  "P10 inside the hull"                                   = w$val >= sw$thr_p10 & w$in_hull,
  "P10 and spread among algorithms <= 10"                 = w$val >= sw$thr_p10 & w$sd_algo <= 10,
  "P10 and spread among algorithms <= 5"                  = w$val >= sw$thr_p10 & w$sd_algo <= 5,
  "Rule D: threshold {round(sw$thr, 1)}, minus one bootstrap spread" = w$val - coalesce(w$spread, 0) >= sw$thr & !coalesce(w$extrap, F))
d_wg <- imap_dfr(w_rules, function(k, nm) {
  k <- coalesce(k, F)
  tibble(
    rule           = as.character(glue(nm)),
    `old Gulf box` = sum(k & w$lat >= 18 & w$lat <= 31 & w$lon >= -98 & w$lon <= -81),
    `PA Gulf`      = sum(k & in_rg("PA Gulf")),
    `EEZ Gulf`     = sum(k & in_rg("EEZ Gulf")),
    `EEZ Atlantic` = sum(k & in_rg("EEZ Atlantic")),
    `EEZ Alaska`   = sum(k & in_rg("EEZ Alaska")),
    `EEZ Pacific`  = sum(k & in_rg("EEZ Pacific")))
})
kable(d_wg, format.args = list(big.mark = ","), caption = glue("North Atlantic right whale, {lbl} ensemble: cells kept by rule. The first column repeats the plan's box; the rest are polygons."))

d_wrec <- bind_rows(
  fitted = d_fit |> filter(taxon == 159023L),
  `all quality-controlled` = d_occ |> filter(taxon == 159023L),
  .id = "records") |>
  left_join(d_scregion, by = "cell_id", relationship = "many-to-many") |>
  mutate(
    region = coalesce(region, "outside the US EEZ"),
    in_box = lat >= 18 & lat <= 31 & lon >= -98 & lon <= -81)
bind_rows(
  d_wrec |> filter(!str_starts(region, "PA ")) |> count(records, region),
  d_wrec |> distinct(records, lon, lat, in_box) |> filter(in_box) |> count(records) |> mutate(region = "old Gulf box")) |>
  pivot_wider(names_from = records, values_from = n, values_fill = 0) |>
  kable(caption = "Right whale records by region. The old box held Atlantic records; the Gulf polygon does not.")

n_v9 <- v9_cells(159023L, "ms_merge")
tibble(
  surface = "v9 merged surface (IUCN range at a flat value)",
  cells   = nrow(n_v9),
  `EEZ Gulf` = sum(n_v9$cell_id %in% d_scregion$cell_id[d_scregion$region == "EEZ Gulf"]),
  `PA Gulf`  = sum(n_v9$cell_id %in% d_scregion$cell_id[d_scregion$region == "PA Gulf"]),
  `EEZ Atlantic` = sum(n_v9$cell_id %in% d_scregion$cell_id[d_scregion$region == "EEZ Atlantic"]),
  `distinct values` = n_distinct(n_v9$val)) |>
  kable(format.args = list(big.mark = ","), caption = "What v9 scores today for the right whale.")
```

## Turtles in Alaska, redone by polygon

```{r}
#| label: turtle-alaska
#| eval: !expr 'any(d_sp$grp == "turtle")'
dbGetQuery(con, "
  SELECT c.taxon, r.region,
         count(*) FILTER (WHERE c.val >= 1)                       AS modeled,
         count(*) FILTER (WHERE c.val >= s.thr_mtp)               AS mtp,
         count(*) FILTER (WHERE c.val >= p.thr)                   AS p10,
         count(*) FILTER (WHERE c.val >= p.thr AND c.in_hull)     AS p10_in_hull,
         count(*) FILTER (WHERE c.val - COALESCE(c.spread, 0) >= t.thr AND NOT COALESCE(c.extrap, FALSE)) AS rule_d
  FROM og_cell c JOIN og_thr t USING (taxon) JOIN og_p10 p USING (taxon) JOIN sc_region r USING (cell_id)
  JOIN sc_species s USING (taxon)
  WHERE r.region IN ('PA Alaska', 'EEZ Alaska') AND c.taxon <> 159023
  GROUP BY ALL ORDER BY r.region DESC, c.taxon") -> d_ak
d_ak |>
  left_join(select(d_sp, taxon, common), by = "taxon") |> select(region, common, modeled:rule_d) |>
  kable(format.args = list(big.mark = ","), caption = glue("Sea turtles, {lbl} ensembles: Alaska cells kept by threshold, by polygon. `rule_d` uses the threshold of rules D and E, minus one bootstrap spread."))
```

## The regional test

For each species and EEZ subregion: among the units the model calls present (rule D's cells), how
many were searched, how many of those hold a record of the species, and how surprising is that count
given the model's precision everywhere else (`msens::region_effort_test()`). Four variants: two effort
definitions, at the cell (0.05°) and at 0.5° blocks. A block is the more honest trial, because
records cluster and neighbouring cells are not independent searches.

```{r}
#| label: region-test
# the block id in SQL (// is integer division) must be the one unit_of() wrote into og_cell
sql_block <- glue("CAST(((cell_id - 1) // 7200 // {UNIT_AGG}) * {7200 %/% UNIT_AGG} + ((cell_id - 1) % 7200) // {UNIT_AGG} AS INTEGER)")
stopifnot(dbGetQuery(con, glue("SELECT count(*) AS n FROM og_cell WHERE unit_id <> {sql_block}"))$n == 0)
variants <- expand_grid(effort = names(eff), unit = c("cell", "block"))
d_count <- pmap_dfr(variants, function(effort, unit) {
  u <- if (unit == "cell") "cell_id" else sql_block
  dbExecute(con, glue("CREATE OR REPLACE TEMP VIEW v_cell   AS SELECT * REPLACE ({u} AS unit_id) FROM og_cell"))
  dbExecute(con, glue("CREATE OR REPLACE TEMP VIEW v_effort AS SELECT taxon, {u} AS unit_id FROM og_effort_cell_{effort}"))
  dbExecute(con, glue("CREATE OR REPLACE TEMP VIEW v_occ    AS SELECT taxon, {u} AS unit_id FROM og_occ_cell"))
  n <- dbGetQuery(con, range_region_sql(cell = "v_cell", effort = "v_effort", occ = "v_occ"))
  bind_cols(tibble(effort = effort, unit = unit), n)
})
# verdicts at the rule's ratio and, for the sensitivity table, at three others (1 = significance alone)
d_verdict_all <- map_dfr(c(1, 0.5, RATIO, 0.1), function(ratio) {
  bind_cols(
    d_count, ratio = ratio,
    region_effort_test(
      d_count$n_searched, d_count$n_occupied, d_count$n_searched_all, d_count$n_occupied_all,
      n_records = d_count$n_records, alpha = ALPHA, ratio = ratio))
})
d_verdict <- filter(d_verdict_all, ratio == RATIO)
dbWriteTable(con, "og_verdict_all", d_verdict_all, overwrite = T)
dir_create(here("data/og"))
write_csv(d_verdict, here(glue("data/og/{lbl}_region_test.csv")))

show_verdicts <- function(eff_k, unit_k) {
  d_verdict |>
    filter(effort == eff_k, unit == unit_k) |>
    left_join(select(d_sp, taxon, common), by = "taxon") |>
    mutate(
      verdict = if_else(!is.na(label), glue("{verdict} ({label})"), verdict),
      p_value = signif(p_value, 2), p_core = signif(p_core, 2), rate = signif(rate, 2)) |>
    select(common, region, `present cells` = n_cells, searched = n_searched, `with a record` = n_occupied,
           records = n_records, rate, `rate elsewhere` = p_core, `needed` = n_min, p = p_value, verdict) |>
    kable(format.args = list(big.mark = ","), caption = glue("Regional test, effort = {eff_k}, unit = {unit_k}. `needed` is the fewest searched units at which zero records would reject."))
}
show_verdicts("order", "block")
show_verdicts("mega", "block")
```

```{r}
#| label: verdict-compare
d_verdict |>
  left_join(select(d_sp, taxon, common), by = "taxon") |>
  mutate(variant = paste(effort, unit)) |>
  select(common, region, variant, verdict) |>
  pivot_wider(names_from = variant, values_from = verdict) |>
  kable(caption = glue("Verdicts under the four variants of the regional test (ratio {RATIO})."))
```

#### Why the test needs an effect size

As the plan worded it, a region is dropped when the binomial tail alone is below 5 %. With hundreds of
searched blocks that rejects any region where the species is merely scarcer than in its stronghold.
The table shows the verdict at four values of `ratio` (the region's record rate must ALSO be below
`ratio` times the rate elsewhere; `1` is significance alone), for the rows where it matters.

```{r}
#| label: ratio-sensitivity
d_verdict_all |>
  filter(unit == "block") |>
  left_join(select(d_sp, taxon, common), by = "taxon") |>
  mutate(`rate / rate elsewhere` = round(rate / p_core, 2)) |>
  select(effort, common, region, searched = n_searched, `with a record` = n_occupied, `rate / rate elsewhere`, ratio, verdict) |>
  pivot_wider(names_from = ratio, values_from = verdict, names_prefix = "ratio ") |>
  filter(n_distinct(c_across(starts_with("ratio "))) > 1, .by = c(effort, common, region)) |>
  arrange(effort, common, region) |>
  kable(caption = "Regions whose verdict depends on the ratio (0.5 degree blocks). Everything not listed gets the same verdict at all four.")
```

## Scorecard

```{r}
#| label: rules
# rule E under one variant of the test = msens::range_constrain_sql() with that variant's verdicts;
# rule D is the same call with no verdict against any region
rule_e <- function(name, eff_k, unit_k, ratio = RATIO) {
  v <- if (is.na(eff_k)) "SELECT taxon, region, verdict FROM og_verdict_all WHERE FALSE" else
    glue("SELECT taxon, region, verdict FROM og_verdict_all WHERE effort = '{eff_k}' AND unit = '{unit_k}' AND ratio = {ratio}")
  dbExecute(con, glue("CREATE OR REPLACE TEMP VIEW v_verdict AS {v}"))
  dbExecute(con, glue("
    CREATE OR REPLACE TEMP TABLE keep_{name} AS
    SELECT taxon, cell_id, status FROM ({range_constrain_sql(verdict = 'v_verdict')}) WHERE keep"))
}
rule_e("d",      NA,      NA)
rule_e("e",      "order", "block")
rule_e("e_mega", "mega",  "block")
rule_e("e_1",    "order", "block", ratio = 1)   # the test as first worded: significance alone

rule_sql <- c(
  "P10"    = "SELECT c.taxon, c.cell_id FROM og_cell c JOIN og_p10 t USING (taxon) WHERE c.val >= t.thr",
  "A"      = "SELECT taxon, cell_id FROM og_cell WHERE a_hull",
  "A fit region" = "SELECT taxon, cell_id FROM og_cell WHERE a_fitr",
  set_names(glue("SELECT taxon, cell_id FROM og_cell WHERE val >= 1 AND d_occ_km <= {BUF_KM}"), glue("B {BUF_KM} km")),
  "C"      = glue("SELECT taxon, cell_id FROM og_cell WHERE (am >= {AM_MIN} AND in_rng) OR is_ch"),
  "D"      = "SELECT taxon, cell_id FROM keep_d",
  "E"      = "SELECT taxon, cell_id FROM keep_e",
  "E mega" = "SELECT taxon, cell_id FROM keep_e_mega",
  "E no guard" = "SELECT taxon, cell_id FROM keep_e_1")
dbExecute(con, glue(
  "CREATE OR REPLACE TABLE sc_keep AS ",
  paste(glue("SELECT '{names(rule_sql)}' AS rule, taxon, cell_id FROM ({rule_sql})"), collapse = " UNION ALL ")))
rule_lvl <- names(rule_sql)

d_card <- dbGetQuery(con, "
  WITH rg AS (SELECT region, cell_id FROM sc_region UNION ALL SELECT DISTINCT 'Globe', cell_id FROM og_cell),
  base AS (
    SELECT c.taxon, rg.region,
           count(*) FILTER (WHERE c.val >= t.thr) AS n_p10,
           count(*) FILTER (WHERE c.is_ch)        AS n_ch,
           count(f.cell_id)                       AS n_fit
    FROM og_cell c JOIN og_p10 t USING (taxon) JOIN rg USING (cell_id)
    LEFT JOIN og_fit f USING (taxon, cell_id)
    GROUP BY ALL),
  kept AS (
    SELECT k.rule, k.taxon, rg.region,
           count(*)                               AS n_kept,
           count(*) FILTER (WHERE c.is_ch)        AS n_ch_kept,
           count(f.cell_id)                       AS n_fit_kept,
           count(*) FILTER (WHERE c.in_iucn)      AS n_kept_in_iucn
    FROM sc_keep k JOIN og_cell c USING (taxon, cell_id) JOIN rg USING (cell_id)
    LEFT JOIN og_fit f USING (taxon, cell_id)
    GROUP BY ALL)
  SELECT b.taxon, b.region, r.rule, b.n_p10, b.n_ch, b.n_fit,
         COALESCE(k.n_kept, 0) AS n_kept, COALESCE(k.n_ch_kept, 0) AS n_ch_kept,
         COALESCE(k.n_fit_kept, 0) AS n_fit_kept, COALESCE(k.n_kept_in_iucn, 0) AS n_kept_in_iucn
  FROM base b CROSS JOIN (SELECT DISTINCT rule FROM sc_keep) r
  LEFT JOIN kept k USING (taxon, region, rule)") |>
  as_tibble() |>
  left_join(select(d_sp, taxon, common), by = "taxon") |>
  mutate(rule = factor(rule, rule_lvl), common = factor(common, d_sp$common))
write_csv(arrange(d_card, taxon, region, rule), here(glue("data/og/{lbl}_scorecard.csv")))
```

### One line per candidate

```{r}
#| label: card-summary
turtles <- d_sp$common[d_sp$grp == "turtle"]
d_core0 <- d_card |>
  filter(str_starts(region, "EEZ"), n_fit >= 20) |>
  summarise(n_p10 = sum(n_p10), n_kept = sum(n_kept), .by = c(common, rule)) |>
  summarise(worst_core_cost = max(1 - n_kept / n_p10), .by = rule)
d_sum <- d_card |>
  summarise(
    `Alaska turtle cells`         = sum(n_kept[region == "EEZ Alaska" & common %in% turtles]),
    `of which loggerhead`         = sum(n_kept[region == "EEZ Alaska" & common == "Loggerhead"]),
    `right whale Gulf cells`      = sum(n_kept[region == "EEZ Gulf" & common == "North Atlantic right whale"]),
    `right whale Atlantic cells`  = sum(n_kept[region == "EEZ Atlantic" & common == "North Atlantic right whale"]),
    `Pacific cells, 6 turtles`    = sum(n_kept[region == "EEZ Pacific" & common %in% turtles]),
    `fewest records kept`         = sprintf("%.0f%%", 100 * min((n_fit_kept / n_fit)[region == "Globe"])),
    `critical-habitat cells lost` = sum((n_ch - n_ch_kept)[region == "Globe"]),
    .by = rule) |>
  left_join(d_core0, by = "rule") |>
  mutate(`worst core cost` = sprintf("%.0f%%", 100 * worst_core_cost), worst_core_cost = NULL) |>
  arrange(rule)
kable(d_sum, format.args = list(big.mark = ","), caption = "Every candidate on one line. EEZ subregions; `fewest records kept` is the lowest share of fitted records kept for any of the seven species; `worst core cost` the largest share of P10 cells lost in any species' core US regions (negative = more cells than P10).")
```

```{r}
#| label: card-read-published
#| eval: !expr is_pub
#| results: asis
#| echo: false
cat(r"---(
What this shows, on the published models:

- **A (OBIS's recipe, with the hull)** clears Alaska for every turtle and costs the core range almost
  nothing, but it does not know about critical habitat and trims it. As OBIS codes it, with the
  `fit_region` mask instead of the hull, the loggerhead keeps its Alaska footprint.
- **B (buffer around occurrences)** keeps every record by construction and brings back what the
  model had removed: the right whale returns to the Gulf at every distance, because four records lie
  there, and the footprint grows with the distance chosen.
- **C (AquaMaps at 0.5 or more)** is a different surface; it loses most loggerhead and olive ridley
  records and keeps the leatherback in Alaska and the right whale in the Gulf.
- **D (threshold minus one bootstrap spread)** trims margins everywhere but leaves the loggerhead's
  Alaska footprint: uncertainty is not what separates Alaska.
- **E (D plus the regional test)** removes the loggerhead from Alaska under either effort
  definition. The green turtle and olive ridley Alaska cells go only with the wider effort layer
  (`E mega`): with turtle records as the only evidence of search, Alaska is "unsearched" for them.
- **E no guard** is the regional test as first worded (significance alone): it also drops the whole
  Pacific EEZ for the loggerhead and the green turtle. That is why `msens::region_effort_test()`
  asks for a material shortfall as well (previous section).
- The right whale has **no cell above P10 in the Gulf EEZ** in the published model. The plan's 487
  came from a box that took in Atlantic waters. Its Gulf problem in v9 is the flat IUCN range, not the model.
)---")
```

```{r}
#| label: card-vs-published
#| eval: !expr '!is_pub && file.exists(here::here("data/og/published_scorecard.csv"))'
# the same line per candidate for OBIS's published models of the same species, beside this run
d_pub <- read_csv(here("data/og/published_scorecard.csv")) |> filter(taxon %in% d_sp$taxon)
sum_line <- function(d, model) d |>
  summarise(
    `Alaska turtle cells`        = sum(n_kept[region == "EEZ Alaska" & common %in% turtles]),
    `right whale Gulf cells`     = sum(n_kept[region == "EEZ Gulf" & common == "North Atlantic right whale"]),
    `Gulf + Atlantic cells`      = sum(n_kept[region %in% c("EEZ Gulf", "EEZ Atlantic")]),
    `Pacific cells`              = sum(n_kept[region == "EEZ Pacific"]),
    `fewest records kept`        = sprintf("%.0f%%", 100 * min((n_fit_kept / n_fit)[region == "Globe"])),
    .by = rule) |>
  mutate(model = model, .before = 1)
bind_rows(sum_line(d_card, lbl), sum_line(d_pub, "published")) |>
  filter(rule %in% c("P10", "A", "B 250 km", "C", "D", "E", "E mega")) |>
  mutate(rule = factor(rule, rule_lvl)) |> arrange(rule, model) |>
  kable(format.args = list(big.mark = ","), caption = glue("This run ({lbl}) beside OBIS's published models of the same {nrow(d_sp)} species. Rule C is the v7.1 AquaMaps rule and does not depend on the model."))
```

### Cells kept

```{r}
#| label: card-cells
card_wide <- function(rg) {
  d_card |> filter(region == rg, rule != "P10") |>
    select(common, rule, n_kept) |> arrange(rule) |>
    pivot_wider(names_from = rule, values_from = n_kept) |>
    left_join(d_card |> filter(region == rg, rule == "P10") |> select(common, P10 = n_kept), by = "common") |>
    relocate(P10, .after = common) |> arrange(common)
}
for (rg in c("PA Alaska", "EEZ Alaska", "PA Gulf", "EEZ Gulf", "EEZ Atlantic", "PA Pacific", "EEZ Pacific"))
  print(kable(card_wide(rg), format.args = list(big.mark = ","), caption = glue("Cells kept in {rg}. P10 alone is the reference column.")))
```

### Fitted records kept, and the cost to the core range

A rule that removes Alaska by removing the Gulf has not helped. *Records kept* is the share of the
fitted records (whole globe) that fall in a kept cell. *Core cost* is the share of P10 cells lost in
the regions where the species has at least 20 fitted records.

```{r}
#| label: card-cost
d_glob <- d_card |> filter(region == "Globe") |>
  transmute(common, rule, records_kept = n_fit_kept / n_fit)
d_core <- d_card |>
  filter(str_starts(region, "EEZ"), n_fit >= 20) |>
  summarise(core = paste(sort(unique(str_remove(region, "EEZ "))), collapse = " + "), n_p10 = sum(n_p10), n_kept = sum(n_kept), .by = c(common, rule)) |>
  mutate(core_cost = 1 - n_kept / n_p10)
d_glob |> filter(rule != "P10") |>
  mutate(records_kept = sprintf("%.0f%%", 100 * records_kept)) |>
  pivot_wider(names_from = rule, values_from = records_kept) |>
  kable(caption = "Share of fitted records (whole globe) inside the kept cells.")
d_core |> filter(rule != "P10") |>
  mutate(core_cost = sprintf("%.0f%%", 100 * core_cost)) |>
  select(common, core, rule, core_cost) |>
  pivot_wider(names_from = rule, values_from = core_cost) |>
  kable(caption = "Share of P10 cells lost in the species' core US regions (EEZ subregions with 20 or more fitted records). Negative = the rule keeps cells below P10 there.")
```

### Critical habitat

Must be zero lost. Rules C, D and E keep critical habitat by construction; A and B do not know about it.

```{r}
#| label: card-ch
d_card |> filter(region == "Globe", n_ch > 0, rule != "P10") |>
  mutate(lost = n_ch - n_ch_kept) |>
  select(common, `critical-habitat cells` = n_ch, rule, lost) |>
  pivot_wider(names_from = rule, values_from = lost) |>
  kable(format.args = list(big.mark = ","), caption = "Critical-habitat cells LOST by each rule (v9's `ch_nmfs` + `ch_fws` cells).")
```

### Agreement with the IUCN range

```{r}
#| label: card-iucn
d_card |> filter(region == "Globe", rule != "P10") |>
  mutate(x = sprintf("%.0f%%", 100 * n_kept_in_iucn / n_kept)) |>
  select(common, rule, x) |>
  pivot_wider(names_from = rule, values_from = x) |>
  kable(caption = "Share of kept cells (whole globe) that lie inside the IUCN range polygon.")
```

## Maps

One panel per rule: kept cells in the lower 48 / Gulf / Atlantic window and in the Alaska window.
Grey = the model's P10 cells the rule dropped; colour = the model value in kept cells.

```{r}
#| label: maps
#| fig-width: 12
#| fig-height: 9
#| results: asis
wins <- list(
  "Gulf and Atlantic" = ext(-100, -60, 17, 48),
  "Alaska and the Pacific coast" = ext(-180, -115, 30, 75))
map_rules <- c("A", "B 250 km", "C", "D", "E", "E mega")   # six panels per window
for (id in d_sp$taxon) {
  cat(glue("\n\n### {d_sp$common[d_sp$taxon == id]}\n\n"))
  d  <- dbGetQuery(con, glue("
    SELECT c.cell_id, c.val, c.val >= t.thr AS p10, k.rule
    FROM og_cell c JOIN og_p10 t USING (taxon) LEFT JOIN sc_keep k USING (taxon, cell_id)
    WHERE c.taxon = {id}"))
  for (wn in names(wins)) {
    tmpl <- crop(r_cid, wins[[wn]])
    par(mfrow = c(2, 3))
    for (rl in map_rules) {
      bg <- tmpl; values(bg) <- NA
      fg <- bg
      i_bg <- cellFromXY(tmpl, xyFromCell(r_cid, unique(d$cell_id[d$p10 %in% TRUE])))
      bg[i_bg[!is.na(i_bg)]] <- 1
      k    <- d[d$rule %in% rl, ]
      i_fg <- cellFromXY(tmpl, xyFromCell(r_cid, k$cell_id))
      fg[i_fg[!is.na(i_fg)]] <- coalesce(k$val[!is.na(i_fg)], 0)
      plot(bg, col = "grey80", legend = F, main = glue("{rl}: {wn}"), mar = c(2, 2, 2, 4))
      if (any(!is.na(i_fg))) plot(fg, add = T, range = c(0, 100), col = hcl.colors(50, "viridis", rev = T), legend = F)
    }
  }
}
```

```{r}
#| label: close
#| include: false
dbDisconnect(con, shutdown = T)
```
