---
title: "Backfill a past release into the multi-version atlas"
msens:
  target_name: backfill_versions
  workflow_type: release
  dependency: [build_zone_cells]
  output: data/manifests/backfill_versions.json
params:
  ver: v7
editor_options:
  chunk_output_type: console
---

Every release v1–v7 exists only as a monolithic `sdm.duckdb` on the server. To render one in the
current app it needs what v8 has: **app-facing tables as Parquet**, a **`mdl_seq` ↔ `mdl_key`
crosswalk**, **COGs** for its surfaces, and a **`manifest.json`** describing all of it.

This does that for one release, parameterized by `ver`, so the same notebook backfills each in turn.

> **Two halves, deliberately separated.** The tables + crosswalk + manifest are cheap (minutes) and
> are what make a version *discoverable*. The COG build is expensive — ~30,000 rasters from a
> 1.18-billion-row surface — and is gated behind `BACKFILL_COGS=1`. That box also serves the live
> apps on 4 cores, so the heavy half should be run deliberately, not as a side effect.

## What makes this affordable

The content-addressed store means a release only pays for surfaces nobody has published yet.
Measured across v3/v5/v6/v7: 120,974 model-rows collapse to **19,766 unique surfaces (6.12×)**, and
v6↔v7 is **30,061/30,061 identical**. So the first `usa05` backfill is the expensive one and each
subsequent release is mostly hashing.

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(arrow, DBI, dplyr, duckdb, glue, knitr, logger, terra,
                 MarineSensitivity/msens, quiet = TRUE)
source(here::here("libs/paths.R"))

bver     <- params$ver
do_cogs  <- Sys.getenv("BACKFILL_COGS") != ""
test_n   <- suppressWarnings(as.integer(Sys.getenv("BACKFILL_TEST_N", "")))
n_work   <- as.integer(Sys.getenv("BACKFILL_WORKERS", "2"))

grid_id  <- msens::grid_for_ver(bver)
# msens::sdm_db_path() knows where each release's database actually is, per
# platform -- swapping the version segment of the configured `sdm_db` does NOT:
# on the server that resolves to derived/{ver}/, while the release databases
# live under big/{ver}/.
db       <- path.expand(msens::sdm_db_path(bver))
dir_v    <- path.expand(dirname(db))
dir_tbl  <- file.path(dir_v, "tables")
dir_cog  <- file.path(dir_v, "backfill_cogs")
dir.create(dir_tbl, recursive = TRUE, showWarnings = FALSE)

stopifnot("release database not found" = file.exists(db))
log_info("backfilling {bver} (grid {grid_id}) from {db}; cogs={do_cogs}")

con <- dbConnect(duckdb(), db, read_only = TRUE)
val <- if ("val" %in% dbListFields(con, "model_cell")) "val" else "value"   # v8 vs v1-v7
log_info("model_cell value column: {val}")
```

## The `mdl_seq` ↔ `mdl_key` crosswalk

`mdl_seq` is an autoincrement that **renumbers on every rebuild**, so a v6 deep link resolves to a
different species in v7 — which is why URLs had to move to the stable `mdl_key`. The old `model`
table happens to carry everything needed to mint one: `ds_key` plus `taxa`, which *is* the source
species id (`am_0.05` / `ITS-Mam-180528`). So the crosswalk is mechanical, not hand-built.

```{r}
#| label: crosswalk

# MERGED models need the taxon's identity, not the model row's `taxa`: v7 stores
# "ms_merge:430645" there, which would mint ms_merge|ms_merge:430645 and match
# nothing. The merged key is ms_merge|{AUTHORITY}:{taxon_id} -- so a v7 merged
# model and a v8 one for the SAME species share a key, which is the whole point.
model <- dbGetQuery(con, "
  SELECT m.mdl_seq, m.ds_key, m.taxa, t.taxon_authority, t.taxon_id
  FROM model m LEFT JOIN taxon t ON t.mdl_seq = m.mdl_seq")

# v1-v7 spell AquaMaps 'am_0.05'; the mdl_key grammar uses 'am'
norm_ds <- function(x) sub("^am_0\\.05$", "am", x)

is_merged <- model$ds_key == "ms_merge"
xwalk <- model |>
  mutate(ds_key  = norm_ds(ds_key),
         mdl_key = ifelse(
           is_merged & !is.na(taxon_authority),
           msens::mdl_key_merged(ifelse(is.na(taxon_authority), "WORMS", taxon_authority),
                                 taxon_id),
           msens::mdl_key_raw(ds_key, taxa)))

# a merged model with no taxon row cannot be keyed; report rather than mint a
# plausible-looking key that resolves to nothing
n_orphan <- sum(is_merged & is.na(model$taxon_authority))
if (n_orphan) log_warn("{n_orphan} merged model(s) have no taxon row - keyed from `taxa` instead")
stopifnot("mdl_key collision" = !any(duplicated(xwalk$mdl_key)))

log_info("{nrow(xwalk)} models crosswalked")
knitr::kable(head(xwalk, 5))
```

## Content-hash every surface

```{r}
#| label: hashes

t0 <- Sys.time()
h  <- msens::content_hashes(con, "model_cell", by = "mdl_seq", cols = c("cell_id", val))
log_info("hashed {nrow(h)} models in {round(as.numeric(difftime(Sys.time(), t0, units='secs')))}s")

ENC <- "int1u-nd0-noovr"   # encoding is part of the object identity; see content_hash_encoded()
h <- h |>
  mutate(payload_hash = content_hash,
         content_hash = msens::content_hash_encoded(content_hash, ENC)) |>
  left_join(xwalk, by = "mdl_seq") |>
  mutate(cog_url = msens::content_url(grid_id, content_hash))

store <- tryCatch(msens::cog_store_index(grid_id), error = function(e) character())
log_info("{length(store)} object(s) already in cog/{grid_id}/; {sum(!h$content_hash %in% store)} new")

knitr::kable(
  h |> summarise(models = n(), unique_surfaces = n_distinct(content_hash),
                 already_published = sum(content_hash %in% store)))
```

## Build the COGs (gated)

```{r}
#| label: cogs

built <- 0L
if (do_cogs) {
  dir.create(dir_cog, recursive = TRUE, showWarnings = FALSE)
  grid <- msens::grid_spec_for(grid_id)

  todo <- h |> distinct(content_hash, .keep_all = TRUE) |> filter(!content_hash %in% store)
  if (!is.na(test_n) && test_n > 0) { todo <- head(todo, test_n); log_warn("TEST_N={test_n}") }
  log_info("{nrow(todo)} raster(s) to build")

  for (i in seq_len(nrow(todo))) {
    r   <- todo[i, ]
    out <- file.path(dir_cog, basename(msens::content_key(grid_id, r$content_hash)))
    if (file.exists(out)) next
    d <- dbGetQuery(con, glue("SELECT cell_id, {val} AS val FROM model_cell WHERE mdl_seq = {r$mdl_seq}"))
    if (nrow(d)) {
      msens::publish_cog(d$cell_id, d$val, out, grid,
                         datatype = "INT1U", nodata = 0, overview = FALSE)
      built <- built + 1L
    }
    if (i %% 500 == 0) log_info("  {i}/{nrow(todo)}")
  }
  log_info("built {built} COG(s), {round(sum(file.info(list.files(dir_cog, full.names=TRUE))$size, na.rm=TRUE)/1024^2)} MB")
} else {
  log_info("BACKFILL_COGS unset - skipping the expensive half")
}
```

## Export the app-facing tables

```{r}
#| label: tables

want <- c("cell", "taxon", "model", "metric", "cell_metric", "zone", "zone_cell",
          "zone_metric", "zone_taxon", "dataset", "taxon_model", "listing")
have <- intersect(want, dbListTables(con))

sizes <- vapply(have, function(t) {
  d <- dbGetQuery(con, glue("SELECT * FROM {t}"))
  f <- file.path(dir_tbl, paste0(t, ".parquet"))
  arrow::write_parquet(d, f, compression = "zstd")
  nrow(d)
}, numeric(1))

# the crosswalk ships as a table of its own: it is what lets a legacy
# ?mdl_seq= link resolve, and what a merged COG href hangs off
model_asset <- h |>
  transmute(mdl_key, mdl_seq, ds_key, sp_id = taxa, n_cells = n,
            content_hash, cog_url, grid_id = grid_id, ver = bver)
arrow::write_parquet(model_asset, file.path(dir_tbl, "model_asset.parquet"), compression = "zstd")

knitr::kable(data.frame(table = c(have, "model_asset"), rows = c(sizes, nrow(model_asset))),
             caption = glue("{bver} tables exported to {dir_tbl}"))
```

## Manifest

```{r}
#| label: manifest

versions <- readr::read_csv(here::here("data/versions.csv"), show_col_types = FALSE)
v_row    <- versions |> filter(ver == bver)
stopifnot("version missing from data/versions.csv" = nrow(v_row) == 1)

mfst <- msens::manifest_build(
  con, bver,
  status = v_row$status,
  base   = msens::atlas_base_url(),
  # COGs are declared only once they exist; until then the app must not offer a
  # layer it cannot draw
  capabilities = list(score_cogs = FALSE, model_cogs = built > 0 || length(store) > 0),
  extra = list(title = v_row$title, released = as.character(v_row$released),
               models = glue("{msens::atlas_base_url()}/{bver}/tables/model_asset.parquet")))

f_mfst <- file.path(dir_v, "manifest.json")
writeLines(as.character(jsonlite::toJSON(mfst, auto_unbox = TRUE, pretty = TRUE, na = "null")), f_mfst)
msens::validate_manifest(jsonlite::fromJSON(f_mfst), ver = bver)
dbDisconnect(con, shutdown = TRUE)

cat(glue("{bver}: id_field={mfst$id_field} grid={mfst$grid_id} tables={length(mfst$tables)} ",
         "metrics={nrow(mfst$metrics)} zones={nrow(mfst$zones)}\n"))
```

## Publish to S3

```{r}
#| label: publish

# The rstudio CONTAINER has no `aws` (the host does). Adding it to
# server/rstudio/Dockerfile is the reproducible fix, but it only takes effect on
# the next container recreate -- and a recreate resets that container's R
# library, taking the live Shiny apps down until they are reinstalled. So when
# the CLI is absent this stages locally and prints the exact host-side command,
# rather than failing with `system2: Function not implemented`.
# presence of the CLI is NOT the question -- whether it can AUTHENTICATE is.
# The container gained `aws` before it gained credentials, and a credential-less
# CLI fails at the publish step AFTER the expensive build, which is the worst
# place to find out.
has_aws <- nzchar(Sys.which("aws")) &&
  identical(0L, suppressWarnings(system2("aws", c("sts", "get-caller-identity"),
                                         stdout = FALSE, stderr = FALSE)))
do_s3   <- Sys.getenv("BACKFILL_NO_S3") == "" && has_aws
if (!has_aws)
  log_warn(paste0("no `aws` in this container - staged locally. Publish from the HOST:\n",
                  "  aws s3 sync {dir_tbl} {s3_atlas}/{bver}/tables\n",
                  "  aws s3 cp {f_mfst} {s3_atlas}/{bver}/manifest.json ",
                  "--content-type application/json --cache-control no-cache"))
if (do_s3) {
  s3_ver <- glue("{s3_atlas}/{bver}")
  push <- function(src, dst, ct = NULL) {
    a <- c("s3", "sync", shQuote(src), shQuote(dst),
           if (!is.null(ct)) c("--content-type", shQuote(ct)),
           "--only-show-errors", "--no-progress")
    o <- system2("aws", a, stdout = TRUE, stderr = TRUE)
    if (!is.null(attr(o, "status")) && attr(o, "status") != 0)
      stop(glue("aws s3 sync {src} failed: "), paste(o, collapse = "\n"))
  }
  push(dir_tbl, glue("{s3_ver}/tables"))
  # the manifest is a MUTABLE pointer: a cached copy advertises capabilities the
  # release no longer has
  system2("aws", c("s3", "cp", shQuote(f_mfst), shQuote(glue("{s3_ver}/manifest.json")),
                   "--content-type", shQuote("application/json"),
                   "--cache-control", shQuote("no-cache"), "--only-show-errors"))
  if (do_cogs && dir.exists(dir_cog))
    push(dir_cog, glue("{s3_atlas}/cog/{grid_id}"))

  # verify by READING BACK through the public URL, as the app would
  got <- msens::atlas_manifest(bver, refresh = TRUE)
  stopifnot("published manifest does not round-trip" = identical(got$ver, bver))
  log_info("verified {msens::atlas_base_url()}/{bver}/manifest.json")
} else {
  log_info("BACKFILL_NO_S3 set - staged locally only")
}
```

## Target manifest

```{r}
#| label: target-manifest

msens::write_manifest(
  here::here("data/manifests/backfill_versions.json"),
  target       = "backfill_versions",
  content_hash = digest::digest(list(bver, h$content_hash), algo = "xxhash64"),
  stats = list(ver = bver, grid_id = grid_id, n_models = nrow(h),
               n_unique_surfaces = length(unique(h$content_hash)),
               n_cogs_built = built, n_tables = length(have)),
  force = msens::force_target("backfill_versions"))
```
