---
title: "Backfill a past release into the multi-version atlas"
msens:
  target_name: backfill_versions
  workflow_type: release
  dependency: [build_zone_cells]
  output: data/manifests/backfill_versions.json
params:
  ver: v7
editor_options:
  chunk_output_type: console
---

Every release v1–v7 exists only as a monolithic `sdm.duckdb` on the server. To render one in the
current app it needs what v8 has: **app-facing tables as Parquet**, a **`mdl_seq` ↔ `mdl_key`
crosswalk**, **COGs** for its surfaces, and a **`manifest.json`** describing all of it.

This does that for one release, parameterized by `ver`, so the same notebook backfills each in turn.

> **Two halves, deliberately separated.** The tables + crosswalk + manifest are cheap (minutes) and
> are what make a version *discoverable*. The COG build is expensive — ~30,000 rasters from a
> 1.18-billion-row surface — and is gated behind `BACKFILL_COGS=1`. That box also serves the live
> apps on 4 cores, so the heavy half should be run deliberately, not as a side effect.

## What makes this affordable

The content-addressed store means a release only pays for surfaces nobody has published yet.
Measured across v3/v5/v6/v7: 120,974 model-rows collapse to **19,766 unique surfaces (6.12×)**, and
v6↔v7 is **30,061/30,061 identical**. So the first `usa05` backfill is the expensive one and each
subsequent release is mostly hashing.

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(arrow, DBI, dplyr, duckdb, glue, knitr, logger, terra,
                 MarineSensitivity/msens, quiet = TRUE)
source(here::here("libs/paths.R"))

bver     <- params$ver
do_cogs  <- Sys.getenv("BACKFILL_COGS") != ""
test_n   <- suppressWarnings(as.integer(Sys.getenv("BACKFILL_TEST_N", "")))
n_work   <- as.integer(Sys.getenv("BACKFILL_WORKERS", "2"))

grid_id  <- msens::grid_for_ver(bver)
# msens::sdm_db_path() knows where each release's database actually is, per
# platform -- swapping the version segment of the configured `sdm_db` does NOT:
# on the server that resolves to derived/{ver}/, while the release databases
# live under big/{ver}/.
db       <- path.expand(msens::sdm_db_path(bver))
dir_v    <- path.expand(dirname(db))
dir_tbl  <- file.path(dir_v, "tables")
dir_cog  <- file.path(dir_v, "backfill_cogs")
dir.create(dir_tbl, recursive = TRUE, showWarnings = FALSE)

stopifnot("release database not found" = file.exists(db))
log_info("backfilling {bver} (grid {grid_id}) from {db}; cogs={do_cogs}")

con <- dbConnect(duckdb(), db, read_only = TRUE)
val <- if ("val" %in% dbListFields(con, "model_cell")) "val" else "value"   # v8 vs v1-v7
log_info("model_cell value column: {val}")
```

## The `mdl_seq` ↔ `mdl_key` crosswalk

`mdl_seq` is an autoincrement that **renumbers on every rebuild**, so a v6 deep link resolves to a
different species in v7 — which is why URLs had to move to the stable `mdl_key`. The old `model`
table happens to carry everything needed to mint one: `ds_key` plus `taxa`, which *is* the source
species id (`am_0.05` / `ITS-Mam-180528`). So the crosswalk is mechanical, not hand-built.

```{r}
#| label: crosswalk

# MERGED models need the taxon's identity, not the model row's `taxa`: v7 stores
# "ms_merge:430645" there, which would mint ms_merge|ms_merge:430645 and match
# nothing. The merged key is ms_merge|{AUTHORITY}:{taxon_id} -- so a v7 merged
# model and a v8 one for the SAME species share a key, which is the whole point.
model <- dbGetQuery(con, "
  SELECT m.mdl_seq, m.ds_key, m.taxa, t.taxon_authority, t.taxon_id
  FROM model m LEFT JOIN taxon t ON t.mdl_seq = m.mdl_seq")

# v1-v7 spell AquaMaps 'am_0.05'; the mdl_key grammar uses 'am'. The rule lives
# in msens::normalize_ds_key() -- it is a rule about the key grammar, and rules
# belong where a test can assert them (test-mdl_key.R round-trips it against the
# real published v1/v3/v7 model_asset rows).

is_merged <- model$ds_key == "ms_merge"
xwalk <- model |>
  mutate(ds_key  = msens::normalize_ds_key(ds_key),
         mdl_key = ifelse(
           is_merged & !is.na(taxon_authority),
           msens::mdl_key_merged(ifelse(is.na(taxon_authority), "WORMS", taxon_authority),
                                 taxon_id),
           msens::mdl_key_raw(ds_key, taxa)))

# a merged model with no taxon row cannot be keyed; report rather than mint a
# plausible-looking key that resolves to nothing
n_orphan <- sum(is_merged & is.na(model$taxon_authority))
if (n_orphan) log_warn("{n_orphan} merged model(s) have no taxon row - keyed from `taxa` instead")
stopifnot("mdl_key collision" = !any(duplicated(xwalk$mdl_key)))

log_info("{nrow(xwalk)} models crosswalked")
knitr::kable(head(xwalk, 5))
```

## Content-hash every surface

```{r}
#| label: hashes

t0 <- Sys.time()
h  <- msens::content_hashes(con, "model_cell", by = "mdl_seq", cols = c("cell_id", val))
log_info("hashed {nrow(h)} models in {round(as.numeric(difftime(Sys.time(), t0, units='secs')))}s")

ENC <- "int1u-nd0-noovr"   # encoding is part of the object identity; see content_hash_encoded()
h <- h |>
  mutate(payload_hash = content_hash,
         content_hash = msens::content_hash_encoded(content_hash, ENC)) |>
  left_join(xwalk, by = "mdl_seq") |>
  mutate(cog_url = msens::content_url(grid_id, content_hash))

store <- tryCatch(msens::cog_store_index(grid_id), error = function(e) character())
log_info("{length(store)} object(s) already in cog/{grid_id}/; {sum(!h$content_hash %in% store)} new")

knitr::kable(
  h |> summarise(models = n(), unique_surfaces = n_distinct(content_hash),
                 already_published = sum(content_hash %in% store)))
```

## Build the COGs (gated)

```{r}
#| label: cogs

built <- 0L
if (do_cogs) {
  dir.create(dir_cog, recursive = TRUE, showWarnings = FALSE)
  grid <- msens::grid_spec_for(grid_id)

  todo <- h |> distinct(content_hash, .keep_all = TRUE) |> filter(!content_hash %in% store)
  if (!is.na(test_n) && test_n > 0) { todo <- head(todo, test_n); log_warn("TEST_N={test_n}") }
  log_info("{nrow(todo)} raster(s) to build")

  for (i in seq_len(nrow(todo))) {
    r   <- todo[i, ]
    out <- file.path(dir_cog, basename(msens::content_key(grid_id, r$content_hash)))
    if (file.exists(out)) next
    d <- dbGetQuery(con, glue("SELECT cell_id, {val} AS val FROM model_cell WHERE mdl_seq = {r$mdl_seq}"))
    if (nrow(d)) {
      msens::publish_cog(d$cell_id, d$val, out, grid,
                         datatype = "INT1U", nodata = 0, overview = FALSE)
      built <- built + 1L
    }
    if (i %% 500 == 0) log_info("  {i}/{nrow(todo)}")
  }
  log_info("built {built} COG(s), {round(sum(file.info(list.files(dir_cog, full.names=TRUE))$size, na.rm=TRUE)/1024^2)} MB")
} else {
  log_info("BACKFILL_COGS unset - skipping the expensive half")
}
```

## Export the app-facing tables

```{r}
#| label: tables

want <- c("cell", "taxon", "model", "metric", "cell_metric", "zone", "zone_cell",
          "zone_metric", "zone_taxon", "dataset", "taxon_model", "listing")
have <- intersect(want, dbListTables(con))

sizes <- vapply(have, function(t) {
  d <- dbGetQuery(con, glue("SELECT * FROM {t}"))
  f <- file.path(dir_tbl, paste0(t, ".parquet"))
  arrow::write_parquet(d, f, compression = "zstd")
  nrow(d)
}, numeric(1))

# Whether `taxon_model` is DERIVED here rather than published by that release --
# recorded in the manifest so nobody mistakes reconstructed edges for original.
tm_reconstructed <- FALSE

# Reconstruct `taxon_model` for the releases that predate it (v1/v2).
#
# Those releases record the taxon->model edges WIDE: one integer column per
# dataset on `taxon` (am_0.05, bl, ch_fws, ch_nmfs, rng_fws, +rng_iucn in v2)
# holding that dataset's contributing mdl_seq, plus `mdl_seq` for the merged
# model and `n_ds` for how many datasets contributed. Unpivoting recovers the
# real edges -- this is derived, not invented.
#
# Without it v1/v2 have no taxon->model relation at all: their view DBs carry 11
# views instead of 13, and nothing can answer "which models fed this taxon" for
# the two oldest releases.
if (!"taxon_model" %in% have) {
  tcols <- dbListFields(con, "taxon")
  ds_cols <- intersect(c("am_0.05", "bl", "ch_fws", "ch_nmfs", "rng_fws", "rng_iucn"), tcols)
  if (length(ds_cols) && all(c("mdl_seq", "n_ds") %in% tcols)) {
    sel <- paste(sprintf(
      "SELECT taxon_id, '%s' AS ds_key, \"%s\" AS mdl_seq FROM taxon WHERE \"%s\" IS NOT NULL",
      ds_cols, ds_cols, ds_cols), collapse = " UNION ALL ")
    tm <- dbGetQuery(con, glue(
      "{sel} UNION ALL SELECT taxon_id, 'ms_merge', mdl_seq FROM taxon WHERE mdl_seq IS NOT NULL"))

    # n_ds is recorded independently of the wide columns, so it checks the unpivot:
    # per-dataset edges (excluding the merged row) should equal it.
    #
    # A small shortfall is EXPECTED and is a ceiling of the source layout, not a
    # bug here: one integer column per dataset can hold only ONE model per dataset
    # per taxon, so a taxon with two models from the same dataset is counted twice
    # by n_ds and stored once. Measured on v1: 5 taxa, 8 edges of 17,676 (0.045%),
    # all BirdLife-id taxa carrying only rng_fws. Warn with the number so the
    # magnitude stays visible; do not fail, and do not silently accept a LARGE gap.
    chk <- dbGetQuery(con, glue(
      "SELECT sum(n_ds) AS expected FROM taxon WHERE n_ds IS NOT NULL"))$expected
    got <- sum(tm$ds_key != "ms_merge")
    if (!isTRUE(all.equal(as.numeric(chk), as.numeric(got)))) {
      pct <- 100 * (as.numeric(chk) - got) / as.numeric(chk)
      log_warn("taxon_model reconstruct: {got} per-dataset edges vs n_ds {chk} ",
               "({round(pct, 3)}% short — the wide layout holds one model per dataset per taxon)")
      stopifnot("taxon_model reconstruct is missing >1% of edges" = pct < 1)
    }
    stopifnot("reconstructed taxon_model is empty" = nrow(tm) > 0)

    f_tm <- file.path(dir_tbl, "taxon_model.parquet")
    arrow::write_parquet(tm, f_tm, compression = "zstd")

    # manifest_build() introspects the CONNECTION, never dir_tbl -- deliberately,
    # so a manifest cannot claim a table the release does not have. The cost is
    # that a table written only as a FILE is published but never advertised: the
    # first run of this reconstruction shipped taxon_model.parquet to S3 for v1
    # and v2 (198/208 KB, verified present) while both manifests omitted it, so
    # nothing downstream could discover the table that had just been built.
    # Registering it keeps introspection authoritative and makes it see exactly
    # what was published. A TEMP view is enough (and works on the read-only
    # connection), so the historical release DB is never mutated.
    dbExecute(con, glue(
      "CREATE OR REPLACE TEMP VIEW taxon_model AS SELECT * FROM read_parquet('{f_tm}')"))
    tm_reconstructed <- TRUE

    have <- c(have, "taxon_model")
    sizes <- c(sizes, taxon_model = nrow(tm))
    log_info("taxon_model RECONSTRUCTED for {bver}: {nrow(tm)} edges ",
             "({got} per-dataset + {sum(tm$ds_key == 'ms_merge')} merged) from {length(ds_cols)} wide columns")
  }
}

# the crosswalk ships as a table of its own: it is what lets a legacy
# ?mdl_seq= link resolve, and what a merged COG href hangs off
model_asset <- h |>
  transmute(mdl_key, mdl_seq, ds_key, sp_id = taxa, n_cells = n,
            content_hash, cog_url, grid_id = grid_id, ver = bver)
arrow::write_parquet(model_asset, file.path(dir_tbl, "model_asset.parquet"), compression = "zstd")

knitr::kable(data.frame(table = c(have, "model_asset"), rows = c(sizes, nrow(model_asset))),
             caption = glue("{bver} tables exported to {dir_tbl}"))
```

## The per-version view DB

The app resolves its database as `{dir_big}/serve.duckdb`, falling back to the multi-GB
`sdm.duckdb`. Only v8 had a `serve.duckdb`, so every other release could only be opened by
attaching its whole 1.9–3.4 GB database — which is why the app pinned one version at startup
instead of switching on `?ver=`.

This writes the same **KB-sized view DB** for this release: views over the Parquet in
`{ver}/tables/`, which the release already syncs to the server. Views are created **only for
tables that exist** — v1/v2 genuinely lack `listing` and `taxon_model`, and a view over a missing
file fails at `CREATE VIEW` time, not at query time, so declaring one would break the whole DB
rather than degrade one panel.

`val AS value` is aliased for the v8-era column rename, matching what `release_marine-atlas.qmd`
does, so one app can query either generation without knowing which it is.

```{r}
#| label: viewdb

f_serve <- file.path(dir_v, "serve.duckdb")
if (file.exists(f_serve)) unlink(f_serve)
sv <- dbConnect(duckdb::duckdb(), f_serve)

# `zone`/`cell_metric`/`zone_metric` carry the scored value; the apps still say `value`
vt   <- c("cell_metric", "zone_metric", "zone")
made <- character()
for (t in c(have, "model_asset")) {
  f <- file.path(dir_tbl, paste0(t, ".parquet"))
  if (!file.exists(f)) next
  # ask DuckDB what the file holds rather than assuming the generation: v1-v7 name
  # the column `value`, v8 renamed it `val`, and aliasing a column that is already
  # there is a duplicate-name error
  nms  <- names(dbGetQuery(sv, glue("SELECT * FROM read_parquet('{f}') LIMIT 0")))
  cols <- if (t %in% vt && "val" %in% nms && !"value" %in% nms) "*, val AS value" else "*"
  dbExecute(sv, glue("CREATE OR REPLACE VIEW {t} AS SELECT {cols} FROM read_parquet('{f}')"))
  made <- c(made, t)
}

# cell_model is a DIRECTORY of spatially-tiled Parquet beside the tables, not a
# tables/*.parquet file, so it needs its own view. Its presence is exactly what
# manifest_build() keys `cell_species_list` on -- without the view a release
# advertises FALSE and a clicked cell cannot list species, which is the state
# v1-v6 were in until build_v7_cell_model was run for them.
dir_cm <- file.path(dir_v, "cell_model")
if (dir.exists(dir_cm) && length(Sys.glob(file.path(dir_cm, "*", "*.parquet")))) {
  dbExecute(sv, glue(
    "CREATE OR REPLACE VIEW cell_model AS ",
    "SELECT * FROM read_parquet('{dir_cm}/*/*.parquet', hive_partitioning = true)"))
  made <- c(made, "cell_model")
  log_info("cell_model view added ({length(Sys.glob(file.path(dir_cm, '*')))} tiles)")
}

# RECORD THE GRID, then prove the stored tile ids agree with it.
#
# cell_model is partitioned by a tile id computed from the grid WIDTH, and msens resolved
# that width from a `cell_grid` table that no release had ever written -- so it always fell
# back to global05's 7200. On every usa05 release that computes a different, perfectly VALID
# tile id, so `WHERE tile IN (...)` pruned away the only partition holding the cells and the
# clicked-cell species list came back EMPTY instead of failing: a v7 cell with 477 models
# reported 0 species, on the promoted release. v8 passed only because 7200 happened to be
# right. cell_model_tile_check() is the check that cannot pass on a mismatch -- it recomputes
# the tile for cells taken from cell_model itself and compares with what is stored.
msens::cell_grid_write(sv, grid_id)
if ("cell_model" %in% made) {
  msens::cell_model_tile_check(sv)
  log_info("cell_model tile ids verified against grid {grid_id} (ncol {msens::cell_grid_ncol(sv)})")
}
made <- c(made, "cell_grid")

# prove each view actually reads, so a missing/corrupt Parquet fails HERE rather than
# surfacing later as an empty panel in the app
counts <- vapply(made, function(t)
  dbGetQuery(sv, glue("SELECT count(*) n FROM {t}"))$n, numeric(1))
dbDisconnect(sv, shutdown = TRUE)
stopifnot("every view must return rows" = all(counts > 0))

log_info("{bver} serve.duckdb: {length(made)} views, {round(file.size(f_serve)/1024)} KB")
knitr::kable(data.frame(view = made, rows = counts), caption = glue("{bver} serve.duckdb views"))
```

## Manifest

```{r}
#| label: manifest

versions <- readr::read_csv(here::here("data/versions.csv"), show_col_types = FALSE)
v_row    <- versions |> filter(ver == bver)
stopifnot("version missing from data/versions.csv" = nrow(v_row) == 1)

# score COGs, if publish_score_cogs.qmd has run for this release. They are
# produced by a separate notebook, so a release's manifest is written before they
# exist and must be regenerated afterwards -- re-running this notebook is cheap
# once the content store is warm (every raster is already published).
f_scog   <- file.path(dir_v, "tables", "score_cog.parquet")
scog_all <- if (file.exists(f_scog)) arrow::read_parquet(f_scog) else NULL
scog <- if (!is.null(scog_all))
  scog_all |> dplyr::filter(!startsWith(metric_key, "_")) |>
    dplyr::transmute(metric_key, subregion_key, cog = cog_url,
                     rescale_min, rescale_max, colormap) else NULL
overlays <- if (!is.null(scog_all))
  scog_all |> dplyr::filter(startsWith(metric_key, "_")) |>
    dplyr::transmute(overlay_key = metric_key, subregion_key, cog = cog_url, colormap) else NULL
log_info("score COGs for {bver}: {if (is.null(scog)) 0 else nrow(scog)} metric, ",
         "{if (is.null(overlays)) 0 else nrow(overlays)} overlay")

# Layer PRESENTATION metadata (display order, category heading, short label) was
# a per-version `layers_{ver}.csv` sitting beside the release on the server, so
# the app needed a local file to know how to label its own layer picker -- and
# v1/v2 have no such file at all, which is one reason they could not render.
# Fold it into the manifest, which is the contract for everything else already.
# NB: layers csv lives under derived/{ver}, NOT big/{ver} where the release db is
f_lyrs <- file.path(path.expand(glue("{dir_derived}/{bver}")), glue("layers_{bver}.csv"))
lyr_meta <- if (file.exists(f_lyrs)) {
  readr::read_csv(f_lyrs, show_col_types = FALSE) |>
    dplyr::transmute(metric_key = lyr, lyr_order = order, category, label = layer)
} else {
  # v1/v2: derive a usable default rather than shipping a manifest the app can
  # only half-read. Order follows metric_seq (the order the release itself
  # defines), category groups by the metric_key stem, label falls back to the
  # description -- all visible as `derived` so the app can say so.
  NULL
}
if (!is.null(scog) && !is.null(lyr_meta))
  scog <- dplyr::left_join(scog, lyr_meta, by = "metric_key")
log_info("layer metadata: {if (is.null(lyr_meta)) 'derived (no layers csv)' else paste(nrow(lyr_meta), 'rows from layers csv')}")

# Zone tiles by vintage, exactly as build_version_manifest.qmd does for v8. This
# notebook omitted them, so every historical manifest published with no zone
# outlines while reporting success -- the app could draw zones on v8 alone.
# `zone_sets` is what lets a release whose own `zone` table predates the
# zone_set_key column resolve its vintages at all.
zsets      <- readr::read_csv(here::here("data/zone_sets.csv"), show_col_types = FALSE)
zone_tiles <- setNames(
  as.list(glue("{msens::atlas_base_url()}/zones/{zsets$zone_set_key}/zones.pmtiles")),
  zsets$zone_set_key)

mfst <- msens::manifest_build(
  con, bver,
  status     = v_row$status,
  base       = msens::atlas_base_url(),
  metrics    = scog,
  zone_tiles = zone_tiles,
  zone_sets  = zsets,
  # COGs are declared only once they exist; until then the app must not offer a
  # layer it cannot draw
  capabilities = list(model_cogs = built > 0 || length(store) > 0,
                      # provenance, not a feature: v1/v2 taxon->model edges are
                      # unpivoted from that release's wide `taxon` columns, not
                      # published by it. Boolean so `manifest_can()`'s isTRUE()
                      # reads it like any other capability.
                      taxon_model_reconstructed = tm_reconstructed),
  extra = list(title = v_row$title, released = as.character(v_row$released),
               overlays = overlays,
               models = glue("{msens::atlas_base_url()}/{bver}/tables/model_asset.parquet")))

f_mfst <- file.path(dir_v, "manifest.json")
writeLines(as.character(jsonlite::toJSON(mfst, auto_unbox = TRUE, pretty = TRUE, na = "null")), f_mfst)
msens::validate_manifest(jsonlite::fromJSON(f_mfst), ver = bver)
dbDisconnect(con, shutdown = TRUE)

cat(glue("{bver}: id_field={mfst$id_field} grid={mfst$grid_id} tables={length(mfst$tables)} ",
         "metrics={nrow(mfst$metrics)} zones={nrow(mfst$zones)}\n"))
```

## Publish to S3

```{r}
#| label: publish

# The rstudio CONTAINER has no `aws` (the host does). Adding it to
# server/rstudio/Dockerfile is the reproducible fix, but it only takes effect on
# the next container recreate -- and a recreate resets that container's R
# library, taking the live Shiny apps down until they are reinstalled. So when
# the CLI is absent this stages locally and prints the exact host-side command,
# rather than failing with `system2: Function not implemented`.
# presence of the CLI is NOT the question -- whether it can AUTHENTICATE is.
# The container gained `aws` before it gained credentials, and a credential-less
# CLI fails at the publish step AFTER the expensive build, which is the worst
# place to find out.
has_aws <- nzchar(Sys.which("aws")) &&
  identical(0L, suppressWarnings(system2("aws", c("sts", "get-caller-identity"),
                                         stdout = FALSE, stderr = FALSE)))
do_s3   <- Sys.getenv("BACKFILL_NO_S3") == "" && has_aws
if (!has_aws)
  log_warn(paste0("no `aws` in this container - staged locally. Publish from the HOST:\n",
                  "  aws s3 sync {dir_tbl} {s3_atlas}/{bver}/tables\n",
                  "  aws s3 cp {f_mfst} {s3_atlas}/{bver}/manifest.json ",
                  "--content-type application/json --cache-control no-cache"))
if (do_s3) {
  s3_ver <- glue("{s3_atlas}/{bver}")
  push <- function(src, dst, ct = NULL) {
    a <- c("s3", "sync", shQuote(src), shQuote(dst),
           if (!is.null(ct)) c("--content-type", shQuote(ct)),
           "--only-show-errors", "--no-progress")
    o <- system2("aws", a, stdout = TRUE, stderr = TRUE)
    if (!is.null(attr(o, "status")) && attr(o, "status") != 0)
      stop(glue("aws s3 sync {src} failed: "), paste(o, collapse = "\n"))
  }
  push(dir_tbl, glue("{s3_ver}/tables"))
  # the manifest is a MUTABLE pointer: a cached copy advertises capabilities the
  # release no longer has
  system2("aws", c("s3", "cp", shQuote(f_mfst), shQuote(glue("{s3_ver}/manifest.json")),
                   "--content-type", shQuote("application/json"),
                   "--cache-control", shQuote("no-cache"), "--only-show-errors"))
  if (do_cogs && dir.exists(dir_cog))
    push(dir_cog, glue("{s3_atlas}/cog/{grid_id}"))

  # verify by READING BACK through the public URL, as the app would
  got <- msens::atlas_manifest(bver, refresh = TRUE)
  stopifnot("published manifest does not round-trip" = identical(got$ver, bver))
  log_info("verified {msens::atlas_base_url()}/{bver}/manifest.json")
} else {
  log_info("BACKFILL_NO_S3 set - staged locally only")
}
```

## Target manifest

```{r}
#| label: target-manifest

msens::write_manifest(
  here::here("data/manifests/backfill_versions.json"),
  target       = "backfill_versions",
  content_hash = digest::digest(list(bver, h$content_hash), algo = "xxhash64"),
  stats = list(ver = bver, grid_id = grid_id, n_models = nrow(h),
               n_unique_surfaces = length(unique(h$content_hash)),
               n_cogs_built = built, n_tables = length(have)),
  force = msens::force_target("backfill_versions"))
```
