---
title: "Build taxon common names — NMFS > FWS > BirdLife > WoRMS (English)"
msens:
  target_name: build_common_names
  workflow_type: merge
  dependency: [merge_taxon]
  output: data/manifests/build_common_names.json
editor_options:
  chunk_output_type: console
---

Resolve one English **common name** per taxon by a preference order — US federal names win
(they're the authoritative/managed names), then BirdLife for birds, then WoRMS:

1. **NOAA NMFS** species directory (`common_name`) — highest preference.
2. **USFWS** species explorer (`Common Name`).
3. **BirdLife** checklist (`common`, from `model_bl.csv`) — authoritative for birds (WoRMS
   covers birds poorly).
4. **WoRMS** vernacular (`vernacularname.txt`, `language == "ENG"`, `isPreferredName` first) —
   for the rest. WoRMS can list several English candidates, so we take the *preferred* one.

Taxa still uncovered (not in NMFS/FWS/BirdLife and with no English WoRMS vernacular) are logged
as **Wikipedia candidates** (`COMMON_WIKI=1` fetches + caches best-fit names — a gated follow-up,
since it's a large network job). Writes `common_name` to `merge.duckdb` + `sdm.duckdb` `taxon`.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "One English common name per taxon by preference: NMFS > FWS > BirdLife > WoRMS (Wikipedia gated)"
flowchart LR
  src["NMFS · FWS · BirdLife · WoRMS vern<br/>(+ gated Wikipedia)"] --> res["coalesce by preference<br/>per clean_sci_name / sisid / aphia"]
  res --> tx[("common_name → taxon<br/>merge.duckdb + sdm.duckdb")]
  tx --> mf["hash_query(taxon)<br/>→ content-addressed manifest"]
```

## Setup + sources

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, janitor, jsonlite, logger, readr,
                 stringr, tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))
options(readr.show_col_types = F)

merge_db  <- glue("{dir_big_v}/marine-atlas/merge.duckdb")
model_bl  <- glue("{dir_big_v}/marine-atlas/dist/model_bl.csv")
nmfs_csv  <- here("data/nmfs_species-directory.csv")
fws_csv   <- glue("{dir_raw}/fws.gov/species/FWS_Species_Data_Explorer.csv")
vern_txt  <- glue("{dir_raw}/marinespecies.org/checklistbank.org_dataset-2011_v2026-07-02/vernacularname.txt")
manifest  <- here("data/manifests/build_common_names.json"); dir_create(path_dir(manifest))
stopifnot(all(file_exists(c(merge_db, nmfs_csv, fws_csv, vern_txt))))

first_common <- function(d) d |> filter(!is.na(common), common != "") |>
  group_by(sci) |> summarise(common = first(common), .groups = "drop")

nmfs <- read_csv(nmfs_csv) |> transmute(sci = clean_sci_name(scientific_name), common = common_name) |> first_common()
fws  <- read_csv(fws_csv)  |> clean_names() |>
  transmute(sci = clean_sci_name(scientific_name), common = common_name) |> first_common()

# WoRMS English vernacular, preferred first, one per AphiaID
vern <- read_tsv(vern_txt) |>
  filter(language == "ENG") |>
  mutate(aphia = str_remove(taxonID, "urn:lsid:marinespecies.org:taxname:"),
         pref  = as.integer(isPreferredName %in% c("1", "true", "preferred", "TRUE"))) |>
  arrange(desc(pref)) |>
  group_by(aphia) |> summarise(common = first(vernacularName), .groups = "drop")

# BirdLife common by sisid (taxon_id for botw)
bl <- read_csv(model_bl) |> transmute(sisid = as.character(sisid), common) |>
  filter(!is.na(common)) |> distinct(sisid, .keep_all = TRUE)
```

## Resolve per taxon

```{r}
#| label: resolve

con <- dbConnect(duckdb(merge_db))
tx  <- dbGetQuery(con, "SELECT ms_merge_key, taxon_authority, taxon_id, scientific_name FROM taxon") |>
  mutate(sci = clean_sci_name(scientific_name))

tx <- tx |>
  left_join(nmfs |> rename(cn_nmfs = common), by = "sci") |>
  left_join(fws  |> rename(cn_fws  = common), by = "sci") |>
  left_join(bl   |> rename(cn_bl   = common), by = c("taxon_id" = "sisid")) |>
  left_join(vern |> rename(cn_worms = common), by = c("taxon_id" = "aphia")) |>
  mutate(
    common_name = coalesce(cn_nmfs, cn_fws, cn_bl, cn_worms),
    common_src  = case_when(!is.na(cn_nmfs) ~ "NMFS", !is.na(cn_fws) ~ "FWS",
                            !is.na(cn_bl) ~ "BirdLife", !is.na(cn_worms) ~ "WoRMS", TRUE ~ NA))

src_tbl <- tx |> count(common_src)
knitr::kable(src_tbl)
log_info("common names resolved for {sum(!is.na(tx$common_name))} of {nrow(tx)} taxa")
```

## Wikipedia fallback (gated: `COMMON_WIKI=1`)

```{r}
#| label: wikipedia

wiki_csv <- here("data/common_names_wikipedia.csv")
need <- tx |> filter(is.na(common_name)) |> distinct(scientific_name)
log_info("{nrow(need)} taxa lack a common name (Wikipedia candidates)")
if (nzchar(Sys.getenv("COMMON_WIKI")) && nrow(need)) {
  librarian::shelf(httr2, quiet = T)
  cache <- if (file_exists(wiki_csv)) read_csv(wiki_csv) else tibble(scientific_name = character(), common = character())
  todo  <- setdiff(need$scientific_name, cache$scientific_name)
  wiki_common <- function(sci) tryCatch({
    r <- httr2::request("https://en.wikipedia.org/api/rest_v1/page/summary/") |>
      httr2::req_url_path_append(utils::URLencode(sci, reserved = TRUE)) |>
      httr2::req_user_agent("marinesensitivity.org common-name resolver") |>
      httr2::req_error(is_error = \(x) FALSE) |> httr2::req_perform() |> httr2::resp_body_json()
    ti <- r$title
    if (is.null(ti) || grepl("Not found", r$type %||% "")) NA_character_ else ti
  }, error = \(e) NA_character_)
  fetched <- tibble(scientific_name = todo, common = vapply(todo, wiki_common, character(1)))
  cache   <- bind_rows(cache, fetched) |> distinct(scientific_name, .keep_all = TRUE)
  write_csv(cache, wiki_csv)
  tx <- tx |> left_join(cache |> rename(cn_wiki = common), by = "scientific_name") |>
    mutate(common_name = coalesce(common_name, cn_wiki),
           common_src  = ifelse(is.na(common_src) & !is.na(cn_wiki), "Wikipedia", common_src))
  log_info("Wikipedia filled {sum(!is.na(tx$cn_wiki))} of {length(todo)} fetched")
} else if (file_exists(wiki_csv)) {
  tx <- tx |> left_join(read_csv(wiki_csv) |> rename(cn_wiki = common), by = "scientific_name") |>
    mutate(common_name = coalesce(common_name, cn_wiki),
           common_src  = ifelse(is.na(common_src) & !is.na(cn_wiki), "Wikipedia", common_src))
}
```

## Write `common_name` → taxon (merge + sdm)

```{r}
#| label: write

duckdb_register(con, "tmp_cn", tx |> select(ms_merge_key, common_name))
dbExecute(con, "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS common_name VARCHAR")
dbExecute(con, "UPDATE taxon SET common_name = tmp_cn.common_name FROM tmp_cn WHERE taxon.ms_merge_key = tmp_cn.ms_merge_key")
dbDisconnect(con, shutdown = TRUE)

con2 <- dbConnect(duckdb(sdm_db))
if ("taxon" %in% dbListTables(con2)) {
  duckdb_register(con2, "tmp_cn", tx |> select(ms_merge_key, common_name))
  dbExecute(con2, "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS common_name VARCHAR")
  dbExecute(con2, "UPDATE taxon SET common_name = tmp_cn.common_name FROM tmp_cn WHERE taxon.ms_merge_key = tmp_cn.ms_merge_key")
  n_sdm <- dbGetQuery(con2, "SELECT count(common_name) n FROM taxon")$n
} else n_sdm <- 0
dbDisconnect(con2, shutdown = TRUE)
log_info("wrote common_name to merge.duckdb + sdm.duckdb ({n_sdm} in sdm taxon)")
```

## Outputs + manifest

```{r}
#| label: manifest

# content-addressed: fingerprint of the taxon table (now carrying common_name);
# reopen read-only since the write chunks already disconnected
con <- dbConnect(duckdb(merge_db, read_only = TRUE))
h   <- msens::hash_query(con, "taxon")
dbDisconnect(con, shutdown = TRUE)

msens::report_table(src_tbl, caption = "common-name source counts")

msens::write_manifest(
  manifest, target = "build_common_names", content_hash = h,
  stats = list(ver = ver, n_taxa = nrow(tx), n_common = sum(!is.na(tx$common_name)),
               by_source = as.list(setNames(src_tbl$n, src_tbl$common_src)),
               n_uncovered = sum(is.na(tx$common_name))),
  force = msens::force_target("build_common_names"))
```
