---
title: "Build cell_model — the cell-oriented twin of a release's model_cell"
params:
  ver: v7
msens:
  target_name: build_v7_cell_model
  workflow_type: release
  dependency: []
  output: data/manifests/build_v7_cell_model.json
editor_options:
  chunk_output_type: console
---

v7 is the **default live** generation (`/scores`, `/species`, and the API's `ver=latest`), and its
drawn-area report path scans `model_cell` — partitioned by nothing, keyed by `mdl_seq` — for every
per-cell question. This builds v7 the same cell-oriented surface v8 already has, so
`msens::species_for_cells()` prunes to a handful of spatial partitions instead.

Two things make this **not** a rerun of the v8 recipe:

1. **A different grid.** v8 tiles a global 7200 × 3600 grid; v7's `cell_id`s come from the regional
   0–360 bio-oracle raster, **3103 × 2006**. The tile key must use 3103, and readers must *discover*
   that rather than assume it — hence `msens::cell_grid_ncol()` and the `cell_grid` sidecar table
   written below. Using 7200 here would not error; it would simply scatter a compact polygon across
   many tiles and forfeit the pruning.
2. **Half the rows are unreachable.** `model_cell` holds 1,184,895,607 rows across 30,061 models,
   but `species_for_cells()` joins `taxon ON taxon.mdl_seq = model_cell.mdl_seq` — and **13,678
   models / 613,315,679 rows (52%) are not referenced by any taxon**. They are the raw per-dataset
   inputs the species app browses individually (10,997 `am_0.05`, 1,518 `rng_iucn`, …), reachable
   via their own COGs. Including them would double the artifact for no reader. `V7_CELLMODEL_ALL=1`
   overrides.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "v7 model_cell (by mdl_seq) → cell_model (by 2.5° spatial tile), taxon-linked only"
flowchart LR
  mc["v7 model_cell<br/>1.18B rows / 30,061 models"] --> f{"referenced<br/>by a taxon?"}
  f -- "no: 613.3M / 13,678" --> x["skipped<br/>(raw inputs, own COGs)"]
  f -- "yes: 571.6M / 16,383" --> t["tile = f(cell_id, ncol=3103)"]
  t --> p["cell_model/tile=*/​*.parquet"]
  p --> v["v7 sdm.duckdb:<br/>cell_model VIEW + cell_grid"]
```

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, rlang, quiet = T)
source(here("libs/paths.R"))

# Parameterized by release. v1-v6 need this surface exactly as v7 did -- without
# it their manifests declare `cell_species_list: false` and a clicked cell cannot
# list species. They share v7's `usa05` grid, so the tile arithmetic is identical;
# the width still comes from the REGISTRY rather than a literal, because using the
# wrong ncol does not error, it just scatters a compact polygon across many tiles
# and silently forfeits the pruning this artifact exists for.
cver      <- params$ver
V7_NCOL   <- as.integer(msens::grid_spec_for(msens::grid_for_ver(cver))$nc)
v7_db     <- msens::sdm_db_path(cver)
dir_v7    <- path_dir(v7_db)
cell_mc   <- glue("{dir_v7}/cell_model")   # local build target
srv_v7    <- glue("/share/data/big/{cver}")
do_all    <- nzchar(Sys.getenv("V7_CELLMODEL_ALL"))     # include non-taxon models
do_build  <- !dir_exists(cell_mc) || nzchar(Sys.getenv("V7_CELLMODEL_REDO"))
do_deploy <- nzchar(Sys.getenv("DEPLOY_V7_CELLMODEL"))

stopifnot("sdm.duckdb not found for this release" = file_exists(v7_db))
log_info("{cver} db: {v7_db} ({round(file_size(v7_db)/1e9, 2)} GB), grid ncol {V7_NCOL} -> {cell_mc}")

# ONE connection for the whole notebook, closed in the last chunk. `on.exit()` is
# the wrong idiom here: knitr fires it at the end of the chunk that registered it,
# so a connection opened in one chunk is already dead in the next.
con <- dbConnect(duckdb(v7_db, read_only = TRUE))
dbExecute(con, "PRAGMA memory_limit='12GB'"); dbExecute(con, "PRAGMA threads=6")
dbExecute(con, glue("PRAGMA temp_directory='{dir_v7}/duckdb_tmp'"))
```

## Build the partitioned surface

Partitioned, never globally sorted: an `ORDER BY` over ~572M rows previously spilled ~500 GB on the
v8 build. `partitioned_write_max_open_files` is raised because every tile is written concurrently.

```{r}
#| label: build

if (!do_build) {
  log_info("{cell_mc} exists — skipping build (V7_CELLMODEL_REDO=1 to force)")
} else {
  if (dir_exists(cell_mc)) dir_delete(cell_mc)
  dbExecute(con, "SET partitioned_write_max_open_files=2048")

  # the tile key comes from msens so writer and reader cannot drift (test-cell-model.R
  # asserts they agree on this exact grid)
  tile_sql <- msens::cell_model_tile_sql("mc.cell_id", ncol = V7_NCOL)
  join_sql <- if (do_all) "" else
    "JOIN (SELECT DISTINCT mdl_seq FROM taxon WHERE mdl_seq IS NOT NULL) t USING (mdl_seq)"

  t0 <- Sys.time()
  dbExecute(con, glue(
    "COPY (SELECT {tile_sql} AS tile, mc.mdl_seq, mc.cell_id, mc.value AS val ",
    "      FROM model_cell mc {join_sql}) ",
    "TO '{cell_mc}' (FORMAT PARQUET, PARTITION_BY tile, OVERWRITE_OR_IGNORE)"))
  log_info("built in {round(as.numeric(difftime(Sys.time(), t0, units='mins')), 1)} min")
}
```

## Validate

A wrong tile id is still a *valid* tile id, so a writer/reader mismatch returns **fewer rows, not an
error**. These checks would catch that: the same cells must resolve identically whether read by
`tile` (pruned) or by `cell_id` alone (unpruned).

```{r}
#| label: validate

mc_glob <- glue("{cell_mc}/*/*.parquet")

smry <- dbGetQuery(con, glue(
  "SELECT count(*) n_rows, count(DISTINCT mdl_seq) n_models,
          count(DISTINCT tile) n_tiles, count(DISTINCT cell_id) n_cells
   FROM read_parquet('{mc_glob}', hive_partitioning = true)"))
log_info("cell_model: {format(smry$n_rows, big.mark=',')} rows, {smry$n_models} models, ",
         "{smry$n_tiles} tiles, {format(smry$n_cells, big.mark=',')} cells, ",
         "{round(as.numeric(sum(file_size(dir_ls(cell_mc, recurse = TRUE, type = 'file'))))/1e9, 2)} GB")

# pruned vs unpruned must agree for a sample of cells
set.seed(42)
cid   <- dbGetQuery(con, glue(
  "SELECT DISTINCT cell_id FROM read_parquet('{mc_glob}', hive_partitioning = true)
   USING SAMPLE 200 ROWS"))$cell_id
tiles <- msens::cell_model_tiles(cid, ncol = V7_NCOL)
ids   <- paste(cid, collapse = ",")
n_pruned <- dbGetQuery(con, glue(
  "SELECT count(*) n FROM read_parquet('{mc_glob}', hive_partitioning = true)
   WHERE tile IN ({paste(tiles, collapse=',')}) AND cell_id IN ({ids})"))$n
n_plain  <- dbGetQuery(con, glue(
  "SELECT count(*) n FROM read_parquet('{mc_glob}', hive_partitioning = true)
   WHERE cell_id IN ({ids})"))$n
log_info("tile pruning: {n_pruned} == {n_plain} over {length(cid)} cells, {length(tiles)} tiles")
stopifnot("tile pruning drops rows — writer/reader grid mismatch" = n_pruned == n_plain)

# and against the source table, so the taxon filter did not lose anything it should keep
n_src <- dbGetQuery(con, glue(
  "SELECT count(*) n FROM model_cell mc
   {if (do_all) '' else 'JOIN (SELECT DISTINCT mdl_seq FROM taxon WHERE mdl_seq IS NOT NULL) t USING (mdl_seq)'}"))$n
stopifnot("row count differs from source" = smry$n_rows == n_src)
log_info("row count matches source: {format(n_src, big.mark=',')}")

msens::report_table(tibble::tibble(
  rows = smry$n_rows, models = smry$n_models, tiles = smry$n_tiles,
  cells = smry$n_cells, grid_ncol = V7_NCOL,
  scope = if (do_all) "all models" else "taxon-linked only"))
```

## Deploy to msens1 (`DEPLOY_V7_CELLMODEL=1`)

rsync the parquet, then register it in the server's v7 `sdm.duckdb` — a `cell_model` VIEW (so
`species_for_cells()` picks it up via `use_cm`) plus the `cell_grid` sidecar so
`msens::cell_grid_ncol()` returns 3103 rather than the v8 default.

The DDL runs against a **copy** which is then swapped in atomically: every live v7 Shiny session
holds a read lock, so an in-place write fails with `Conflicting lock is held`. `mv` is atomic within
the filesystem, so in-flight sessions finish on the old inode and `restart.txt` rolls them over.

```{r}
#| label: deploy

if (!do_deploy) {
  log_info("DEPLOY_V7_CELLMODEL unset — built locally only")
} else {
  # HARD GUARD, ordering matters. msens < 0.13.0 has no cell_grid_ncol(), so
  # species_for_cells() would tile v7 cells with the 7200 default, prune away the
  # very partitions holding them, and return FEWER SPECIES WITH NO ERROR. Ship
  # msens first (DEPLOY_APPS=1 / DEPLOY_API=1), then register the view.
  # Rendering ON the server (scripts/srv_render.sh) means the build target ALREADY
  # is the server path: nothing to rsync, and `ssh msens` from inside the container
  # fails outright -- which is how the first v6 run died after a fully successful
  # build (571,579,928 rows, pruning verified) with only "Error in system2()".
  on_server <- identical(fs::path_real(dir_v7), fs::path_real(srv_v7))
  v <- if (on_server) as.character(packageVersion("msens")) else {
    o <- system2("ssh", c("msens", shQuote(glue(
      "docker exec -u shiny rstudio Rscript -e ",
      shQuote("cat(as.character(packageVersion('msens')))")))), stdout = TRUE, stderr = TRUE)
    utils::tail(grep("^[0-9]+\\.[0-9]+", o, value = TRUE), 1)
  }
  if (!length(v) || package_version(v) < "0.13.0")
    stop(glue("server msens is {v %||% '?'}; need >= 0.13.0 for cell_grid_ncol() — ",
              "run DEPLOY_APPS=1 (and DEPLOY_API=1) on release_marine-atlas.qmd first, ",
              "else v7 reports silently lose species"))
  log_info("server msens {v} >= 0.13.0 — safe to register the v7 cell_model view")

  # check every exit status explicitly: system2() reports failure only through its
  # return value here, so an unchecked mkdir/rsync fails the render with NO usable
  # message (the first attempt died exactly this way — 962 bytes of log, no error)
  ok <- function(rc, what)
    if (!identical(as.integer(rc), 0L)) stop(glue("{what} failed, exit {rc}"))

  if (on_server) {
    log_info("already on the server ({dir_v7}) — no rsync needed")
  } else {
    ok(system2("ssh", c("msens", shQuote(glue("mkdir -p {srv_v7}/cell_model")))), "mkdir")
    ok(system2("rsync", c("-a", "--delete", glue("{cell_mc}/"),
                          glue("msens:{srv_v7}/cell_model/"))), "rsync cell_model")
    log_info("synced cell_model -> msens:{srv_v7}/cell_model")
  }

  register <- glue(
    "con <- DBI::dbConnect(duckdb::duckdb('{srv_v7}/sdm.duckdb.new')); ",
    "DBI::dbExecute(con, \"CREATE OR REPLACE VIEW cell_model AS SELECT cm.tile, cm.mdl_seq, ",
    "cm.cell_id, cm.val, cm.val AS value FROM read_parquet('{srv_v7}/cell_model/*/*.parquet', ",
    "hive_partitioning = true) cm\"); ",
    "DBI::dbExecute(con, 'CREATE OR REPLACE TABLE cell_grid AS SELECT {V7_NCOL} AS ncol'); ",
    "cat('V7CM_OK', DBI::dbGetQuery(con, ",
    "'SELECT count(*) n FROM cell_model WHERE tile = (SELECT min(tile) FROM cell_model)')$n, '\\n'); ",
    "DBI::dbDisconnect(con, shutdown = TRUE)")
  # Post-cutover (2026-08-12) there is ONE app, at apps_v8, serving every release
  # from ?ver=. `mapgl`/`mapsp` are retired to /share/shiny_apps_retired, so
  # touching their restart.txt would reload nothing.
  #
  # The view also has to land in serve.duckdb, not only sdm.duckdb: the app
  # resolves `{dir_big}/serve.duckdb` first and every release now has one, so a
  # view registered solely in sdm.duckdb would never be read. backfill_versions'
  # viewdb chunk adds it from the parquet on the next `--stage manifest`, which is
  # also what flips `cell_species_list` to true in the manifest.
  live_apps <- c("species", "scores")
  cmd <- glue(
    "set -e; ",
    "cp -f {srv_v7}/sdm.duckdb {srv_v7}/sdm.duckdb.new; ",
    "docker exec rstudio Rscript -e {shQuote(register)}; ",
    "mv -f {srv_v7}/sdm.duckdb.new {srv_v7}/sdm.duckdb; ",
    "{paste(sprintf('touch /share/github/MarineSensitivity/apps_v8/%s/restart.txt', live_apps), collapse='; ')}")
  out <- if (on_server) system2("bash", c("-c", shQuote(cmd)), stdout = TRUE, stderr = TRUE)
         else            system2("ssh", c("msens", shQuote(cmd)), stdout = TRUE, stderr = TRUE)
  if (!any(grepl("V7CM_OK", out)))
    stop(glue("register failed (exit {attr(out, 'status') %||% 0}):\n"),
         paste(utils::tail(out, 25), collapse = "\n"))
  log_info("{cver} cell_model registered + apps reloaded: {grep('V7CM_OK', out, value = TRUE)}")
}
```

## Manifest

```{r}
#| label: manifest

msens::write_manifest(
  here(glue("data/manifests/build_v7_cell_model{if (cver == 'v7') '' else paste0('_', cver)}.json")),
  target       = glue("build_v7_cell_model{if (cver == 'v7') '' else paste0('_', cver)}"),
  content_hash = msens::hash_query(
    con, glue("SELECT tile, mdl_seq, cell_id, val FROM read_parquet('{mc_glob}', hive_partitioning = true)")),
  stats = list(
    rows = smry$n_rows, models = smry$n_models, tiles = smry$n_tiles,
    cells = smry$n_cells, grid_ncol = V7_NCOL,
    scope = if (do_all) "all" else "taxon_linked"),
  force = msens::force_target("build_v7_cell_model"))   # v7 keeps the original target name

dbDisconnect(con, shutdown = TRUE)
```
