---
title: "Build the zone-set registry (spatial units as vintages, not per-version copies)"
msens:
  target_name: build_zone_sets
  workflow_type: schema
  dependency: [build_cell_grid]
  output: data/manifests/build_zone_sets.json
editor_options:
  chunk_output_type: console
---

Every release so far baked its own copy of the spatial units — `ply_programareas_2026_v3.gpkg`
through `…_v8.gpkg`, each with its own `zone_cell` re-extracted by `score_zones.qmd`. That makes
the question users actually ask — *how did **this** Program Area's score change from v3 to v8?* —
unanswerable, because nothing guarantees two releases meant the same polygon.

This notebook identifies a spatial unit by its **geometry**, not by the release that used it:

```
zone_set_key = {zone_type}_{YYYY-MM}      e.g. programarea_2026-03
```

The vintage is the year-month the geometry **first appeared on disk**, so the label is a checkable
fact rather than an editorial guess, and keys sort chronologically.

> **The per-version files are mostly the same layer copied forward.** Measured across every
> published gpkg, `programarea` is 8 files but only **2** distinct geometries (v2 alone; v3–v8 are
> byte-identical), and `ecoregion` is 10 files but **2**. So `zone_cell` — which depends only on
> (geometry × grid) — is computed **once** per `(zone_set_key, grid_id)` and reused by every release
> on that grid, instead of being re-extracted per version.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "Fingerprint every zone layer, collapse identical ones into vintages, emit the registry"
flowchart LR
  gp["ply_*.gpkg<br/>(top level + every version dir)"] --> gh["msens::zone_geom_hash()<br/>order-invariant WKB fingerprint"]
  gh --> gr["group by (zone_type, geom_hash)<br/>vintage = earliest mtime YYYY-MM"]
  gr --> vz["msens::validate_zone_sets()<br/>one key ↔ one geometry"]
  vz --> csv["data/zone_sets.csv"]
```

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(dplyr, glue, knitr, readr, MarineSensitivity/msens, quiet = TRUE)
source(here::here("libs/paths.R"))

# Every zone layer anywhere under derived/ — the top level and every version dir.
# `ply_ecoregion_programarea_*` is a CROSSWALK (ecoregion x programarea slivers),
# not a spatial unit in its own right, and `ply_boem-usa` is the study-area
# outline; neither is a zone set.
#
# Globbed lazily, at the point of USE, because the canonical subregions are
# DERIVED below and must be picked up by the scan. Globbing here at setup ran
# before that file existed, so the first run silently omitted it from the
# registry while still writing the layer.
zone_gpkgs <- function() {
  g <- Sys.glob(c(path.expand(glue("{dir_derived}/ply_*.gpkg")),
                  path.expand(glue("{dir_derived}/*/ply_*.gpkg"))))
  grep("ecoregion_programarea|boem-usa", g, value = TRUE, invert = TRUE)
}

message(glue("derived tree: {dir_derived}"))
```

::: {.callout-warning}
This scan is only as complete as the machine it runs on. The **server** holds every version's
`derived/` tree; a laptop holds only the current version's. Render this where the whole tree is,
or the registry will silently omit vintages.
:::

## Derive the canonical subregions

Subregions were never a published layer: each release synthesized them in the database as unions of
**its own** Planning/Program Areas, so they drifted (v1 `AK/AKL48/L48/USA`, v4–v6 `AK/GA/PA/USA`,
v7 adds `FULL`) and v2/v3 carry *two different sets under one `fld`*. Worse, a set derived from the
2026 Program Areas **cannot cover the Atlantic at all** — the 2026 National OCS Program has no
Atlantic areas — so it can never be the frame for comparing releases across all US waters.

The ecoregion layer already solves this. It is **byte-identical across v1–v8**, it spans every US
water, and it carries the rollup as a column: `region_key` / `region_name`. So the canonical
subregions are a dissolve of that column — no hand-written lookup to curate, and nothing to guess:

| region | ecoregions |
|---|---|
| `AK` Alaska | High Arctic, Chukchi & Beaufort Seas, East Bering Sea, **Gulf of Alaska** |
| `AT` Atlantic | NE Continental Shelf, SE Continental Shelf, Puerto Rico & USVI |
| `GA` Gulf of America | **Eastern**, **Western & Central** Gulf of America |
| `PA` Pacific | California Current, Washington/Oregon, Pacific Island Territories |

> Note `GOA` is the Gulf of **Alaska** while `EGOA`/`WCGOA` are the Gulf of **America** — the codes
> invite exactly the wrong inference, which is why this reads `region_key` from the layer rather
> than mapping ecoregion codes by hand.

The vintage is **inherited from the source ecoregions** (not the build date), since the geometry is
a pure function of them — so rebuilding never mints a spurious new vintage.

```{r}
#| label: subregions
#| message: false

librarian::shelf(sf, quiet = TRUE)

eco_src <- path.expand(glue("{dir_derived}/v1/ply_ecoregions_2025.gpkg"))
eco_vin <- format(file.info(eco_src)$mtime, "%Y-%m")     # the ecoregion vintage
f_sub   <- path.expand(glue("{dir_derived}/ply_subregions_usa_{eco_vin}.gpkg"))

eco <- sf::st_read(eco_src, quiet = TRUE)
stopifnot("ecoregion layer lacks region_key" = all(c("region_key", "region_name") %in% names(eco)))

sub <- eco |>
  group_by(subregion_key = region_key, subregion_name = region_name) |>
  summarise(n_ecoregions = n(), .groups = "drop") |>
  sf::st_make_valid()

sf::st_write(sub, f_sub, delete_dsn = TRUE, quiet = TRUE)
knitr::kable(sf::st_drop_geometry(sub), caption = glue("Canonical subregions -> {basename(f_sub)}"))
```

`USA` (all waters) and `FULL` (the whole map extent) stay **pseudo-subregions**: they are not
polygons but the absence of a filter, so materializing them as geometry would double-count.

## Fingerprint every layer

```{r}
#| label: fingerprint

zone_type_of <- function(f)
  dplyr::case_when(
    grepl("programarea", f) ~ "programarea",
    grepl("planarea",    f) ~ "planarea",
    grepl("ecoregion",   f) ~ "ecoregion",
    grepl("subregion",   f) ~ "subregion",
    .default = NA_character_)

gpkgs <- zone_gpkgs()          # re-globbed AFTER the derived subregions exist
message(glue("scanning {length(gpkgs)} zone layer(s)"))

scan <- lapply(gpkgs, function(f) {
  # zone_type FIRST: a program-area gpkg carries region_key, planarea_key AND
  # programarea_key, so "the first *_key column" picks region_key (3 values for
  # 20 features) -- which both mislabels the zones and, being a non-total
  # ordering, leaves ties whose order depends on the file's feature order
  zt <- zone_type_of(basename(f))
  h  <- msens::zone_geom_hash(f, zone_type = if (is.na(zt)) NULL else zt)
  tibble::tibble(
    source    = sub(paste0(path.expand(dir_derived), "/"), "", f),
    zone_type = zone_type_of(basename(f)),
    n_zones   = h$n,
    key_col   = h$key_col,
    geom_hash = h$geom_hash,
    mtime     = file.info(f)$mtime)
}) |> bind_rows() |> filter(!is.na(zone_type))

# which release used it: the version directory it sits in ("" = top level, i.e.
# a staging copy no release scored)
scan <- scan |> mutate(ver = ifelse(grepl("/", source), sub("/.*", "", source), NA_character_))

knitr::kable(scan |> select(source, zone_type, n_zones, ver, geom_hash))
```

## Collapse identical layers into vintages

```{r}
#| label: vintages

vint <- scan |>
  group_by(zone_type, geom_hash) |>
  summarise(n_zones = first(n_zones),
            # First appearance on disk -- EXCEPT the derived canonical subregions,
            # whose vintage is inherited from the ecoregions they dissolve (it is
            # encoded in their filename). Using their build mtime would mint a
            # new vintage on every rebuild for geometry that never changed.
            vintage = {
              inherited <- sub(".*subregions_usa_([0-9]{4}-[0-9]{2})\\.gpkg$", "\\1",
                               grep("subregions_usa_", source, value = TRUE)[1])
              if (!is.na(inherited) && grepl("^[0-9]{4}-[0-9]{2}$", inherited))
                inherited else format(min(mtime), "%Y-%m")
            },
            n_files = n(),
            versions = paste(sort(na.omit(ver)), collapse = " "),
            source   = source[which.min(mtime)],
            .groups  = "drop") |>
  mutate(zone_set_key = msens::zone_set_key(zone_type, vintage)) |>
  arrange(zone_type, vintage)

# Two DISTINCT geometries of one type first appearing in the same month would
# collide on the key. Stop rather than silently merge two different maps.
dup <- vint$zone_set_key[duplicated(vint$zone_set_key)]
if (length(dup))
  stop(glue("zone_set_key collision ({paste(dup, collapse=', ')}): two distinct geometries ",
            "first appeared in the same month. Disambiguate the vintage by hand (e.g. -a/-b) ",
            "before publishing."))

msens::validate_zone_sets(vint)

knitr::kable(vint |> select(zone_set_key, n_zones, n_files, versions, geom_hash))
```

```{r}
#| label: summary

vint |>
  count(zone_type, name = "distinct_vintages") |>
  left_join(scan |> count(zone_type, name = "files"), by = "zone_type") |>
  relocate(files, .after = zone_type) |>
  knitr::kable(caption = "How many spatial units really exist, versus how many files")
```

## Which releases actually *use* each vintage

`versions` above records which version directory a **gpkg file** was found in — provenance, not
usage. Those are different: v7 draws Planning Areas and Ecoregions but ships no gpkg of its own for
them, so it went unlisted while plainly using them. Manifests resolve a release's zone tiles by
looking itself up in this column, so under-reporting usage published seven manifests with no zone
outlines at all.

So read usage where it is actually recorded — each release's own `zone` table — and match to a
vintage by `(zone_type, n_zones)`. A count that matches nothing stays unmatched rather than being
attached to the nearest vintage: v7's 5-zone subregion set is its own `AK/GA/PA/USA/FULL` synthesis,
not the canonical 4-zone layer, and quietly equating them would draw the wrong outlines.

```{r}
#| label: used-by

used <- lapply(msens::atlas_versions()$ver, function(v) {
  f <- msens::sdm_db_path(v)
  if (!file.exists(f)) return(NULL)
  con <- DBI::dbConnect(duckdb::duckdb(), f, read_only = TRUE)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  if (!"zone" %in% DBI::dbListTables(con)) return(NULL)
  DBI::dbGetQuery(con, "SELECT fld, count(*) n FROM zone GROUP BY 1") |>
    transmute(ver = v, zone_type = sub("_key$", "", fld), n_zones = n)
}) |> bind_rows()

used <- used |>
  left_join(vint |> select(zone_type, n_zones, zone_set_key), by = c("zone_type", "n_zones"))

knitr::kable(used |> arrange(zone_type, ver),
             caption = "Zone usage read from each release's `zone` table")

unmatched <- used |> filter(is.na(zone_set_key))
if (nrow(unmatched))
  message(glue("unmatched zone usage (no vintage with that count): ",
               "{paste(unmatched$ver, unmatched$zone_type, unmatched$n_zones, collapse='; ')}"))

# fold usage into `versions`, so a manifest resolving itself by version finds
# every vintage it draws -- not only the ones whose file happened to live in its
# own directory
vint <- vint |>
  left_join(
    used |> filter(!is.na(zone_set_key)) |>
      group_by(zone_set_key) |>
      summarise(used_by = paste(sort(unique(ver)), collapse = " "), .groups = "drop"),
    by = "zone_set_key") |>
  mutate(versions = vapply(seq_len(n()), function(i) {
    both <- unique(c(strsplit(versions[i], "\\s+")[[1]],
                     strsplit(ifelse(is.na(used_by[i]), "", used_by[i]), "\\s+")[[1]]))
    both <- both[nzchar(both)]
    paste(both[order(as.integer(sub("^v([0-9]+).*", "\\1", both)), both)], collapse = " ")
  }, character(1))) |>
  select(-used_by)

knitr::kable(vint |> select(zone_set_key, n_zones, versions))
```

## Write the registry

```{r}
#| label: write

f_csv <- here::here("data/zone_sets.csv")

# generated-but-committed, like a lockfile: deterministic given the files on
# disk, and reviewable in a diff when a new vintage appears
# The canonical vintage of a zone type is the frame applied uniformly to every
# release, rather than one matched per version. Subregions are the case that
# needs it: the two 4-zone vintages are indistinguishable by count, and v7's own
# 5-zone set (which adds FULL) matches neither, so without a declared canonical
# every release resolved to NA and drew no subregion outline at all.
out <- vint |>
  mutate(canonical = zone_type == "subregion" & grepl("subregions_usa_", source)) |>
  select(zone_set_key, zone_type, vintage, n_zones, geom_hash, versions, canonical, source)

stopifnot("exactly one canonical vintage per canonical zone_type" =
  all(tapply(out$canonical, out$zone_type, sum) <= 1))
readr::write_csv(out, f_csv)

msens::validate_zone_sets(readr::read_csv(f_csv, show_col_types = FALSE))
message(glue("wrote {f_csv} ({nrow(out)} zone sets)"))
```

## Publish each vintage as PMTiles

Zone outlines are drawn by both apps, and until now came from
`ply_programareas_2026.pmtiles` / `ply_ecoregions_2025.pmtiles` on the file host —
**unversioned**, and inherited artifacts of an archived v7 notebook that nothing
re-runs. So the tiles a v3 map draws were whatever the last hand-run produced.

Built here per vintage, keyed like everything else, and published to S3 beside the
data:

```
zones/{zone_set_key}/zones.pmtiles
```

Vector tiles are versioned by their OWN vintage and shared across MST releases —
six releases share one program-area geometry, so they share one tile set.

```{r}
#| label: pmtiles

do_pmt <- Sys.getenv("ZONE_PMTILES") != "" && nzchar(Sys.which("tippecanoe"))
if (Sys.getenv("ZONE_PMTILES") != "" && !nzchar(Sys.which("tippecanoe")))
  message("ZONE_PMTILES set but tippecanoe is not installed - skipping")

pmt <- tibble::tibble(zone_set_key = character(), pmtiles = character())
if (do_pmt) {
  dir_pmt <- path.expand(glue("{dir_derived}/zones"))
  for (i in seq_len(nrow(vint))) {
    z   <- vint[i, ]
    src <- path.expand(glue("{dir_derived}/{z$source}"))
    out <- glue("{dir_pmt}/{z$zone_set_key}/zones.pmtiles")
    dir.create(dirname(out), recursive = TRUE, showWarnings = FALSE)

    ply <- sf::st_read(src, quiet = TRUE)
    kc  <- msens::zone_key_col(z$zone_type, names(ply))
    # carry the KEY plus any name column, so the app can label without a join
    keep <- intersect(c(kc, sub("_key$", "_name", kc)), names(ply))
    msens::publish_pmtiles(ply[, keep], out, layer = z$zone_type,
                           keep_attrs = keep, quiet = TRUE)
    pmt <- dplyr::bind_rows(pmt, tibble::tibble(
      zone_set_key = z$zone_set_key, pmtiles = as.character(out)))
    message(glue("  {z$zone_set_key}: {round(file.info(out)$size/1024^2, 1)} MB"))
  }

  if (Sys.getenv("ZONE_PMTILES_NO_S3") == "") {
    out <- system2("aws", c("s3", "sync", shQuote(dir_pmt),
                            shQuote(glue("{s3_atlas}/zones")),
                            "--exclude", shQuote("*"), "--include", shQuote("*.pmtiles"),
                            "--only-show-errors", "--no-progress"),
                   stdout = TRUE, stderr = TRUE)
    if (!is.null(attr(out, "status")) && attr(out, "status") != 0)
      stop("aws s3 sync failed: ", paste(out, collapse = "\n"))
    message(glue("published {nrow(pmt)} zone tile set(s) to {s3_atlas}/zones"))
  }
  knitr::kable(pmt |> dplyr::mutate(mb = round(file.info(pmtiles)$size/1024^2, 1)) |>
                 dplyr::select(zone_set_key, mb))
} else {
  message("ZONE_PMTILES unset - skipping tile build")
}
```

## What each release used, and what replaces it

The per-release subregions this supersedes — note they are not comparable to each other, and v2/v3
are ambiguous even within one database:

| release | subregion keys |
|---|---|
| v1 | `AK`, `AKL48`, `L48`, `USA` |
| v2, v3 | *two different sets under one `fld`* — `AK/AKL48/L48/USA` **and** `AK/GA/PA/USA` |
| v4–v6 | `AK`, `GA`, `PA`, `USA` |
| v7 | `AK`, `FULL`, `GA`, `PA`, `USA` |
| v8 | `AK`, `GA`, `PA`, `USA` |

Every one of them is missing the **Atlantic**, because all were unions of Program Areas and the 2026
National OCS Program has none there. The canonical set derived above does not have that hole, and
being a function of the v1–v8-identical ecoregions, it applies unchanged to every release.

## Target manifest

```{r}
#| label: target-manifest

msens::write_manifest(
  here::here("data/manifests/build_zone_sets.json"),
  target       = "build_zone_sets",
  content_hash = digest::digest(out, algo = "xxhash64"),
  stats = list(
    n_files       = nrow(scan),
    n_zone_sets   = nrow(out),
    programarea_n = sum(vint$zone_type == "programarea"),
    ecoregion_n   = sum(vint$zone_type == "ecoregion"),
    planarea_n    = sum(vint$zone_type == "planarea"),
    subregion_n   = sum(vint$zone_type == "subregion")),
  force = msens::force_target("build_zone_sets"))
```
