---
title: "Build the atlas `app/` bundle — the browser's data contract"
msens:
  target_name: build_app_bundle
  workflow_type: publish
  # after release_marine-atlas (needs the release's own sdm.duckdb / serve.duckdb
  # and, for a legacy generation, its published tables/model_asset.parquet), and
  # BEFORE build_version_manifest (which records the `app{}` block by probing
  # THIS notebook's push with an anonymous HTTPS HEAD, never by copying
  # local presence).
  dependency: [release_marine_atlas]
  output: data/manifests/build_app_bundle.json
params:
  ver: null
editor_options:
  chunk_output_type: console
---

Master plan D5 ("normalize at publish, not in the browser") and atlas-1 Deliverable 4. Both
Shiny apps carry a per-version adapter today — two grids, `mdl_seq` vs `mdl_key`, `value` vs
`val`, `is_ok` vs `is_valid_usa`, a `taxon_model` self-edge on one side only — each with a
production bug recorded behind it. `msens::app_bundle_build()` (msens 0.43.0, `R/app_bundle.R`)
already resolves every one of those quirks, with tests. This notebook is the thin publishing
shell around it: **build → validate → push (gated) → verify (gated)**, one version-independent
`{ver}/app/` bundle per release.

**Nothing here publishes anything.** The push and verification chunks are real, runnable code,
gated behind `APP_BUNDLE_S3=1` (unset by default) exactly like every other side-effecting step in
this repo (`../CLAUDE.md` reproducibility rule 1) — this notebook can be rendered top to bottom,
safely, with no S3 credentials and no network write of any kind.

`APP_BUNDLE_VERS=v7,v9` narrows a run to that set (comma-separated); unset, every version in
`data/versions.csv` is attempted, and a version whose release database is not on this machine is
**skipped, not failed** (the same convention `build_version_manifest.qmd`'s `caps_override`
uses: a capability this run cannot see is `FALSE`, never invented). `APP_BUNDLE_V7_CELLMODEL=1`
additionally uploads v7's missing `serve/cell_model/` (757 MB locally, absent from S3 today even
though v7's manifest advertises `cell_species_list: true` — the atlas-1 "Facts" note) under its
own flag, separate from the ordinary `{ver}/app/` push. Both flags are strict
([app_bundle_flag()]): only exactly `1` turns them on — `0`, `true`, `false`, `no`, anything else
STOPS the render naming the variable, rather than guessing.

**Two things worth saying out loud so nobody "fixes" them.** First, a `restricted` release gets
**exactly the same objects** under `{ver}/app/` as a `public` one — the review gate that matters is
which HOST serves them (atlas-9: the public site never renders a restricted version; the preview
host does, behind Access), not what this notebook writes, so `build_app_bundle.qmd` never branches
on `access`. Second, even with `APP_BUNDLE_S3` and `APP_BUNDLE_V7_CELLMODEL` both unset — the
default, publish-nothing case — a legacy-generation dry run (v1–v7b) still downloads roughly
**165 MB anonymously** from the public bucket (`.attach_s3_legacy()`'s `tables/*.parquet` cache,
measured against v7, 2026-09-21: 13 tables, 164,894,811 bytes, dominated by `cell_metric` at 109
MB): reading nothing is not the same as touching nothing.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "One release database -> app_bundle_build() -> {ver}/app/, validated before it is ever written to S3"
flowchart LR
  subgraph src["this version's release database"]
    sdm[("sdm.duckdb<br/>(v8/v9: has model_asset->native_asset)")]
    srv[("serve.duckdb<br/>(v1-v7b: model_asset lives HERE,<br/>not in sdm.duckdb -- see Setup)")]
  end
  sdm -->|"sdm_db_path(ver)"| con(("read-only<br/>DuckDB connection"))
  srv -.->|"fallback when sdm.duckdb<br/>has no asset registry"| con
  con --> abb["msens::app_bundle_build()"]
  abb --> boot["boot.json"]
  abb --> taxa["taxa.json"]
  abb --> tsh["taxon/{xx}.json x256"]
  abb --> alsh["alias/{xx}.json x256"]
  abb --> txp["taxon.parquet"]
  abb --> ztp["zone_taxon.parquet"]
  abb --> cell["cell/tile={t}/data_0.parquet"]
  csv[("data/taxonomic_hierarchy_worms_…csv<br/>(curated, not a DB table)")] -->|"app_taxonomy_table()<br/>libs/app_bundle.R"| txm["taxonomy.parquet"]
  boot & taxa & tsh & alsh & txp & ztp & cell & txm --> val{{"app_validate()<br/>every object vs its<br/>inst/schema/app_*.schema.json"}}
  val -->|"APP_BUNDLE_S3=1"| grd{{"app_bundle_assert_prefix()<br/>refuses any key outside<br/>{ver}/app/"}}
  grd --> s3[["s3://…/marine-atlas/{ver}/app/"]]
  s3 -->|"APP_BUNDLE_S3=1"| anon{{"anonymous HTTPS HEAD/GET<br/>per object"}}
  anon --> mfst["build_version_manifest.qmd<br/>probes the SAME way -><br/>manifest.json $app{}"]

  classDef gate fill:#3a1e1e,stroke:#e03131,color:#e6edf3;
  classDef db fill:#1e3a5f,stroke:#3bc9db,color:#e6edf3;
  class val,grd,anon gate;
  class sdm,srv,con db;
```

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(
  DBI, dplyr, duckdb, fs, glue, here, jsonlite, knitr, logger, purrr, readr,
  MarineSensitivity/msens, quiet = TRUE)
source(here("libs/duckdb_budget.R"))
source(here("libs/app_bundle.R"))

stopifnot(
  "msens 0.43.0 or newer is required (app_bundle_build(), app_manifest_block())" =
    packageVersion("msens") >= "0.43.0")
message(glue("msens {packageVersion('msens')} @ {find.package('msens')}"))

# --- run scope ------------------------------------------------------------
versions_csv <- readr::read_csv(here("data/versions.csv"), show_col_types = FALSE)
narrow       <- Sys.getenv("APP_BUNDLE_VERS")
vers         <- if (nzchar(narrow)) trimws(strsplit(narrow, ",")[[1]]) else versions_csv$ver

# strict: exactly "1" is on, unset/"" is off, anything else (0/true/false/no/...)
# STOPS the render naming the variable -- see libs/app_bundle.R::app_bundle_flag()
do_s3       <- app_bundle_flag("APP_BUNDLE_S3")          # push to S3 (default OFF)
do_v7_cm    <- app_bundle_flag("APP_BUNDLE_V7_CELLMODEL") # also upload v7's serve/cell_model/
base        <- msens::atlas_base_url()
dir_derived <- if (Sys.info()[["sysname"]] == "Linux") "/share/data/derived" else "~/_big/msens/derived"
dir_out_root <- here("_output", "app_bundle")                 # gitignored (see .gitignore)
fs::dir_create(dir_out_root)

message(glue(
  "versions: {paste(vers, collapse = ', ')}  push_s3={do_s3}  v7_cell_model={do_v7_cm}  ",
  "base={base}"))
```

**Which database.** `msens::sdm_db_path(ver)` returns the release's build database, and prefers
`sdm.duckdb` over `serve.duckdb` whenever the former exists on this machine. That is right for
v8/v9, whose `native_asset` table lives IN `sdm.duckdb` — but wrong for v1–v7b: their local
`sdm.duckdb` has neither `model_asset` nor `native_asset` at all (verified 2026-09-21 against the
real v7 release: `model`, not `model_asset`, holds only dataset/taxon/region metadata, no COG
URL). `model_asset` is written only by `backfill_versions.qmd`, as a view in `serve.duckdb`. Opening
the wrong database is not an error — `app_bundle_build()` runs to completion and every taxon's
`merged` and `inputs[].assets` come back silently empty, which is exactly the failure mode this
contract exists to prevent everywhere else.

Three tiers, each PROBED (a view existing by name is not enough — `serve.duckdb`'s views embed
the LOCAL paths of the machine that built them, so a copy rsynced to a laptop lists a `model_asset`
view that then fails the moment it is queried; discovered running this notebook for real against
v7):

1. `sdm_db_path(ver)` (`sdm.duckdb`, or `serve.duckdb` if that is what `sdm_db_path()` itself
   falls back to) — right for v8/v9.
2. `serve.duckdb` beside it, if (1) has no working asset table — right on the SERVER that built it.
3. An ephemeral, **anonymous, read-only** DuckDB attached directly to this release's *published*
   `tables/*.parquet` on S3 (httpfs, path-style; no credentials, no write) — right anywhere,
   including this laptop, for any `released`/public version. Never used for a `restricted` version
   (nothing to attach yet, or nothing this notebook should read anonymously either).

```{r}
#| label: db-open

# does `tbl` exist AND actually return rows when queried (not just listed)?
.probe_table <- function(con, tbl) {
  if (!tbl %in% DBI::dbListTables(con)) return(FALSE)
  isTRUE(tryCatch(DBI::dbGetQuery(con, glue("SELECT count(*) n FROM {tbl} LIMIT 1"))$n >= 0,
                  error = function(e) FALSE))
}
.has_asset_tbl <- function(con) any(vapply(c("model_asset", "native_asset"), .probe_table, logical(1), con = con))

# tier 3: views over a LOCAL CACHE of the PUBLISHED tables/*.parquet -- downloaded
# once, anonymously, read-only (no S3 write, no credentials). attach_atlas() does
# the remote-view version of this for v8+'s native_asset shape, but has no
# model_asset/taxon_model view -- and a REMOTE view here was measured to be
# impractically slow: a full aggregation over `cell_metric` (unprunable row
# groups) via anonymous httpfs range reads did not finish in over an hour, where
# the equivalent LOCAL query takes seconds. `tables/` for a whole legacy release
# is under 200 MB (atlas-1 plan "Facts": v7 cell 22.7 MB + cell_metric 109 MB +
# the rest under 25 MB combined), so caching it locally costs a one-time download,
# not a redesign.
.attach_s3_legacy <- function(ver, access) {
  if (!identical(access, "public")) return(NULL)
  cache_dir <- here("_output", "app_bundle", "_s3cache", ver, "tables")
  fs::dir_create(cache_dir)
  http <- sprintf("%s/%s", base, ver)
  tbls <- c("cell", "taxon", "model", "model_asset", "metric", "cell_metric",
           "zone", "zone_cell", "zone_metric", "zone_taxon", "dataset", "taxon_model", "listing")
  got <- vapply(tbls, function(t) {
    dest <- file.path(cache_dir, paste0(t, ".parquet"))
    if (file.exists(dest)) return(TRUE)
    isTRUE(tryCatch(
      utils::download.file(sprintf("%s/tables/%s.parquet", http, t), dest,
                           mode = "wb", quiet = TRUE) == 0L,
      error = function(e) FALSE, warning = function(w) FALSE))
  }, logical(1))
  if (!any(got)) return(NULL)
  con <- DBI::dbConnect(duckdb::duckdb())
  ok <- vapply(tbls[got], function(t) {
    isTRUE(tryCatch({
      DBI::dbExecute(con, glue(
        "CREATE OR REPLACE VIEW {t} AS SELECT * FROM read_parquet('{cache_dir}/{t}.parquet')"))
      .probe_table(con, t)
    }, error = function(e) FALSE))
  }, logical(1))
  if (!.has_asset_tbl(con)) { DBI::dbDisconnect(con, shutdown = TRUE); return(NULL) }
  log_warn("{ver}: cached + attached {sum(ok)}/{length(tbls)} tables from {http}/tables/ ({cache_dir})")
  con
}

app_bundle_open <- function(ver) {
  v_row  <- versions_csv |> filter(ver == !!ver)
  access <- if (nrow(v_row)) v_row$access[1] else "restricted"
  db  <- path.expand(msens::sdm_db_path(ver))
  if (file.exists(db)) {
    con <- DBI::dbConnect(duckdb::duckdb(), db, read_only = TRUE)
    if (.has_asset_tbl(con)) return(list(con = con, db = db, ok = TRUE))
    DBI::dbDisconnect(con, shutdown = TRUE)
  }
  alt <- file.path(fs::path_dir(db), "serve.duckdb")
  if (fs::path_file(db) != "serve.duckdb" && file.exists(alt)) {
    con2 <- DBI::dbConnect(duckdb::duckdb(), alt, read_only = TRUE)
    if (.has_asset_tbl(con2)) {
      log_warn("{ver}: sdm.duckdb has no working model_asset/native_asset; using serve.duckdb ({alt})")
      return(list(con = con2, db = alt, ok = TRUE))
    }
    DBI::dbDisconnect(con2, shutdown = TRUE)
  }
  con3 <- .attach_s3_legacy(ver, access)
  if (!is.null(con3)) return(list(con = con3, db = sprintf("s3 (anonymous, tables/) %s/%s", base, ver), ok = TRUE))
  if (!file.exists(db)) return(list(con = NULL, db = db, ok = FALSE))
  log_warn(paste(
    "{ver}: no database anywhere (local sdm.duckdb, local serve.duckdb, S3 tables/) carries",
    "model_asset/native_asset -- merged COG urls and input assets will be EMPTY (not an error)"))
  list(con = DBI::dbConnect(duckdb::duckdb(), db, read_only = TRUE), db = db, ok = TRUE)
}

# the gates below only need cell/cell_metric/zone*/taxon/metric -- always present
# in sdm.duckdb (or its serve.duckdb fallback) for every generation -- so they
# open directly rather than paying app_bundle_open()'s S3-probe escalation,
# which exists only for the asset registry (model_asset/native_asset).
app_bundle_open_scoring <- function(ver)
  DBI::dbConnect(duckdb::duckdb(), path.expand(msens::sdm_db_path(ver)), read_only = TRUE)
```

## Guard: the push may never write outside `{ver}/app/`, and a publish flag is never guessed at

Run once, at render time, so the render itself is the proof — not just a definition nobody
exercised. `app_bundle_assert_prefix()` (whitelist, not blocklist) and `app_bundle_flag()` are both
pure: no S3, no side effect.

```{r}
#| label: guard-selftest
app_bundle_assert_prefix_selftest()
app_bundle_flag_selftest()
app_bundle_gzip_selftest()
```

## Build

**One `msens::app_bundle_build()` call per version** — review round 1 (2026-09-21): the prior
version of this notebook composed the bundle itself (`app_boot(..., tables = list())`,
`manifest_build()` with no `zone_tiles`), which is why `boot$tables` (the OPFS invalidation key)
and `boot$units` (the drawable units) came out EMPTY on both real dry runs. `app_bundle_build()`
is the one function that wires `app_table_descriptor()`/`app_table_manifest()` into `app_boot()`
correctly; calling it directly is what makes `tables` and `units` non-empty, not a bigger
`tryCatch`. `geom_keys` — the published geometry keys `app_units()` needs to exclude
whole-study-area rollups (`USA`/`FULL`) — comes from [app_bundle_geom_keys()], which reads the
zone-set registry's own source GeoPackage rather than guessing a filename.

`app/taxonomy.parquet` still has no msens builder (a curated CSV, not a database table), so it is
written here, from the JUST-WRITTEN `taxon.parquet` (never a second query), via
`app_taxonomy_table()` (`libs/app_bundle.R`).

`app_bundle_build()` does not yet isolate its own internal stages (an msens change landing in
parallel with this round); until that lands, a single broken stage for one version aborts that
version's WHOLE build here — the outer `tryCatch` below keeps that from aborting every OTHER
version in the same run, which is the one thing this notebook can guarantee on its own side.

```{r}
#| label: build

build_one <- function(ver) {
  t0 <- Sys.time()
  opened <- app_bundle_open(ver)
  if (!isTRUE(opened$ok)) {
    log_warn("{ver}: no local release database at {opened$db} -- SKIPPED")
    return(list(ver = ver, ok = FALSE, reason = "no local database"))
  }
  con <- opened$con
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  duckdb_tune(con, file.path(tempdir(), "app_bundle_spill", ver))

  dir_out <- file.path(dir_out_root, ver)
  if (fs::dir_exists(dir_out)) fs::dir_delete(dir_out)   # clean slate every run
  fs::dir_create(dir_out)

  v_row <- versions_csv |> filter(ver == !!ver)
  status <- if (nrow(v_row)) v_row$status[1] else "prerelease"
  access <- if (nrow(v_row)) v_row$access[1] else "restricted"

  f_scog  <- file.path(fs::path_dir(opened$db), "tables", "score_cog.parquet")
  metrics <- if (file.exists(f_scog)) {
    arrow::read_parquet(f_scog) |> filter(!startsWith(metric_key, "_")) |>
      transmute(metric_key, subregion_key, cog = cog_url, rescale_min, rescale_max, colormap)
  } else NULL

  zone_sets <- readr::read_csv(here("data/zone_sets.csv"), show_col_types = FALSE)
  mfst <- msens::manifest_build(con, ver, status = status, access = access, base = base,
                                metrics = metrics, zone_sets = zone_sets)
  geom_keys <- app_bundle_geom_keys(mfst$zones, zone_sets, dir_derived)

  out <- tryCatch(
    msens::app_bundle_build(con, ver, dir_out, manifest = mfst, base = base, geom_keys = geom_keys),
    error = function(e) {
      log_warn("{ver}: app_bundle_build() FAILED: {substr(conditionMessage(e), 1, 300)}")
      NULL
    })

  txm_ok <- FALSE
  if (!is.null(out)) {
    txm_ok <- isTRUE(tryCatch({
      tx <- arrow::read_parquet(file.path(dir_out, "taxon.parquet"))
      d  <- app_taxonomy_table(tx)
      msens::write_atlas_parquet(d, file.path(dir_out, "taxonomy.parquet"))
      TRUE
    }, error = function(e) {
      log_warn("{ver}: taxonomy.parquet FAILED: {conditionMessage(e)}")
      FALSE
    }))
  }

  list(ver = ver, ok = !is.null(out), dir_out = dir_out, con_db = opened$db, manifest = mfst,
       geom_keys = geom_keys, out = out, taxonomy_ok = txm_ok,
       wall_s = as.numeric(Sys.time() - t0, units = "secs"))
}

builds <- purrr::map(vers, build_one) |> setNames(vers)
knitr::kable(
  bind_rows(lapply(builds, function(b) data.frame(
    ver = b$ver, ok = b$ok, n_taxa = if (!is.null(b$out)) b$out$n_taxa else NA,
    n_tables = if (!is.null(b$out)) length(b$out$tables) else NA,
    n_units = if (!is.null(b$out)) length(b$out$boot$units) else NA,
    taxonomy_ok = b$taxonomy_ok, wall_s = round(b$wall_s, 1)))),
  caption = "Per-version build result (this run)")
```

## Validate

`app_bundle_build()`'s builders already validate their own output as they write it
(`app_validate()`, `R/app_bundle.R`); this chunk re-validates the WRITTEN BYTES, round-tripped
through disk exactly as an app would fetch them — the same discipline
`build_version_manifest.qmd` applies to `manifest.json` — and proves the schemas actually reject
a broken shape, not just accept a good one.

```{r}
#| label: validate-schema

revalidate_written <- function(b) {
  if (!isTRUE(b$ok)) return(tibble())
  files <- list(
    boot  = list(path = file.path(b$dir_out, "boot.json"),  schema = "boot"),
    taxa  = list(path = file.path(b$dir_out, "taxa.json"),  schema = "taxa"))
  rows <- purrr::imap(files, function(f, nm) {
    if (!file.exists(f$path)) return(data.frame(ver = b$ver, object = nm, ok = NA, note = "not built"))
    x <- jsonlite::fromJSON(f$path, simplifyVector = FALSE)
    ok <- tryCatch({ msens::app_validate(x, f$schema); TRUE },
                    error = function(e) { message(conditionMessage(e)); FALSE })
    data.frame(ver = b$ver, object = nm, ok = ok, note = "")
  })
  for (kind in c("taxon", "alias")) {
    d <- file.path(b$dir_out, kind)
    if (!dir.exists(d)) next
    fs_ <- list.files(d, full.names = TRUE)
    bad <- 0L
    for (f in fs_) {
      x <- jsonlite::fromJSON(f, simplifyVector = FALSE)
      ok <- tryCatch({ msens::app_validate(x, kind); TRUE }, error = function(e) FALSE)
      if (!ok) bad <- bad + 1L
    }
    rows[[kind]] <- data.frame(ver = b$ver, object = sprintf("%s/*.json (n=%d)", kind, length(fs_)),
                               ok = bad == 0L, note = if (bad) sprintf("%d shard(s) FAILED", bad) else "")
  }
  bind_rows(rows)
}
val_tbl <- bind_rows(purrr::map(builds, revalidate_written))
knitr::kable(val_tbl, caption = "Schema re-validation of the WRITTEN bytes, per version")
```

### `boot.json` actually carries its OPFS keys and drawable units

Review round 1's root finding: composing the bundle by hand (rather than calling
`app_bundle_build()`) left `boot$tables` and `boot$units` EMPTY on both real dry runs. These four
gates are what a regression back to hand-composing would fail.

```{r}
#| label: validate-boot-contract

boot_contract_gate <- function(b) {
  if (!isTRUE(b$ok) || is.null(b$out)) return(tibble())
  boot   <- b$out$boot
  tables <- b$out$tables      # named list of app_table_descriptor() results
  digest_ok <- vapply(tables, function(d) {
    got <- as.character(msens::parquet_digest(file.path(b$dir_out, d$path))$digest[1])
    identical(got, d$digest)
  }, logical(1))
  data.frame(
    ver = b$ver,
    n_tables            = length(tables),
    tables_ge_3         = length(tables) >= 3,
    all_digests_match   = length(digest_ok) > 0 && all(digest_ok),
    n_digest_mismatches = sum(!digest_ok),
    n_units             = length(boot$units),
    units_ge_1          = length(boot$units) >= 1,
    # boot.json does not (yet) carry its own id_field -- msens issue, tracked
    # in the atlas-1 review; check what CAN be checked today: the manifest's
    # id_field is present and one of the two known values.
    manifest_id_field    = b$manifest$id_field %||% NA_character_,
    id_field_present_ok  = !is.null(b$manifest$id_field) &&
      b$manifest$id_field %in% c("mdl_key", "mdl_seq"))
}
`%||%` <- function(a, b) if (is.null(a)) b else a
boot_contract_tbl <- bind_rows(purrr::map(builds, boot_contract_gate))
knitr::kable(boot_contract_tbl,
             caption = "boot.json contract: tables (OPFS digests), units (drawable), id_field")
stopifnot(
  "boot$tables has fewer than 3 entries"          = nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$tables_ge_3),
  "a table descriptor's digest disagrees with the written file's" =
    nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$all_digests_match),
  "boot$units has fewer than 1 entry"             = nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$units_ge_1),
  "manifest id_field missing or unrecognized"     = nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$id_field_present_ok))
```

```{r}
#| label: validate-malformed-fixture
#| eval: !expr any(purrr::map_lgl(builds, "ok"))

# a deliberately malformed fixture must fail -- proves the two checks above are
# not vacuously green (msens's own test-app_bundle.R exercises the same five
# malformations against the builder output; this repeats it against ONE real
# written boot.json so the notebook, not just the package, is proven to enforce it)
good <- jsonlite::fromJSON(
  file.path(builds[[which(purrr::map_lgl(builds, "ok"))[1]]]$dir_out, "boot.json"),
  simplifyVector = FALSE)
bad_cases <- list(
  "required key removed"    = { x <- good; x$grid <- NULL; x },
  "wrong type (nc as string)" = { x <- good; x$grid$nc <- "7200"; x },
  "undeclared key"          = { x <- good; x$surprise <- 1; x },
  "out-of-range enum"       = { x <- good; x$release$access <- "semi-public"; x })
mal <- purrr::imap(bad_cases, function(x, nm) {
  ok <- tryCatch({ msens::app_validate(x, "boot"); TRUE }, error = function(e) FALSE)
  data.frame(case = nm, rejected = !ok)
}) |> bind_rows()
knitr::kable(mal, caption = "A deliberately malformed boot.json must be REJECTED (rejected = TRUE)")
stopifnot("schema validation let a malformed boot.json through" = all(mal$rejected))
```

### Exactly one file per tile directory

Review round 1 found 290 of v9's 422 `cell/tile=*/` directories holding SEVERAL part files
(`data_0.parquet`, `data_1.parquet`, ...) — an msens `COPY ... PARTITION_BY` bug under
parallelism, fixed in `../msens` in parallel with this round. The browser only ever fetches the
one fixed filename `tile={t}/data_0.parquet` (never a glob — anonymous LIST is denied), so a tile
with more than one file is missing data from every part after the first, silently. This gate is
what would have caught it; it is expected to be RED until the msens fix lands and this notebook
re-renders against it.

```{r}
#| label: validate-one-file-per-tile

one_file_per_tile_gate <- function(b) {
  if (!isTRUE(b$ok) || is.null(b$out$tiles)) return(tibble())
  d <- file.path(b$dir_out, "cell")
  tiledirs <- list.dirs(d, full.names = TRUE, recursive = FALSE)
  n_files <- vapply(tiledirs, function(td) length(list.files(td, pattern = "\\.parquet$")), integer(1))
  data.frame(ver = b$ver, n_tile_dirs = length(tiledirs),
             n_with_exactly_1_file = sum(n_files == 1L),
             n_with_multiple_files = sum(n_files > 1L),
             n_with_zero_files     = sum(n_files == 0L),
             worst_example = if (any(n_files > 1L))
               basename(tiledirs[which(n_files > 1L)[1]]) else NA_character_,
             worst_example_n_files = if (any(n_files > 1L)) max(n_files) else NA_integer_)
}
one_file_tbl <- bind_rows(purrr::map(builds, one_file_per_tile_gate))
knitr::kable(one_file_tbl, caption = "cell/tile=*/ directories: exactly one file expected per tile")
if (nrow(one_file_tbl) && any(one_file_tbl$n_with_multiple_files > 0))
  message("RED (expected until the msens partition-writer fix lands): ",
          paste(sprintf("%s has %d/%d tile dirs with >1 file (e.g. tile=%s has %d)",
                        one_file_tbl$ver, one_file_tbl$n_with_multiple_files, one_file_tbl$n_tile_dirs,
                        one_file_tbl$worst_example, one_file_tbl$worst_example_n_files),
                collapse = "; "))
```

### Tile-width gate (real nc, then the wrong one)

```{r}
#| label: validate-tile-width

tile_width_gate <- function(b) {
  if (!isTRUE(b$ok) || is.null(b$out$tiles)) return(tibble())
  ncol_real <- b$out$tiles$ncol
  ncol_wrong <- if (ncol_real == 7200L) 3103L else 7200L
  ok_real  <- tryCatch({ msens::app_cell_tile_check(file.path(b$dir_out, "cell"), ncol_real); TRUE },
                       error = function(e) FALSE)
  bad_wrong <- tryCatch({ msens::app_cell_tile_check(file.path(b$dir_out, "cell"), ncol_wrong); FALSE },
                        error = function(e) TRUE)
  data.frame(ver = b$ver, ncol_real = ncol_real, real_passes = ok_real,
             ncol_wrong = ncol_wrong, wrong_fails_as_expected = bad_wrong)
}
tw_tbl <- bind_rows(purrr::map(builds, tile_width_gate))
knitr::kable(tw_tbl, caption = "Tile-width gate: PASS at the release's own nc, FAIL at the wrong one")
stopifnot(
  "tile-width gate did not pass at the real nc"    = all(tw_tbl$real_passes),
  "tile-width gate did not fail at the wrong nc"   = all(tw_tbl$wrong_fails_as_expected))
```

### Wide-tile digest gate (per-metric multiset digest vs `cell_metric`, read the way the browser does)

`msens::app_cell_tile_digests()` globs every `*.parquet` under `cell/`; the browser fetches only
`tile={t}/data_0.parquet`. Read the SAME single file per tile here — a tile with extra part files
(see above) then correctly shows a digest MISMATCH, because the rows in `data_1.parquet` etc. are
invisible to this query exactly as they are to the app.

```{r}
#| label: validate-digest

digest_gate <- function(b, ver) {
  if (!isTRUE(b$ok) || is.null(b$out$tiles)) return(tibble())
  con <- app_bundle_open_scoring(ver)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  duckdb_tune(con, file.path(tempdir(), "app_bundle_spill_digest", ver))
  keys <- b$out$tiles$metric_keys
  vc   <- msens::sdm_val_col(con, "cell_metric")
  glob <- file.path(b$dir_out, "cell", "tile=*", "data_0.parquet")   # browser-equivalent, not **/*.parquet
  bind_rows(lapply(keys, function(k) {
    kq <- DBI::dbQuoteIdentifier(con, k)
    a <- msens::hash_query(con, glue::glue(
      "SELECT cell_id, {kq} AS val FROM read_parquet('{glob}', hive_partitioning = true)
        WHERE {kq} IS NOT NULL"), cols = c("cell_id", "val"))
    bb <- msens::hash_query(con, glue::glue(
      "SELECT cm.cell_id, cm.{vc} AS val FROM cell_metric cm JOIN metric m USING (metric_seq)
        WHERE m.metric_key = {DBI::dbQuoteString(con, k)} AND cm.{vc} IS NOT NULL"),
      cols = c("cell_id", "val"))
    data.frame(ver = ver, metric_key = k, tiles = a, cell_metric = bb, ok = identical(a, bb))
  }))
}
dg_tbl <- purrr::imap(builds, digest_gate) |> bind_rows()
knitr::kable(dg_tbl |> select(ver, metric_key, ok),
             caption = "cell/tile=*/data_0.parquet digest == cell_metric digest, per metric (browser-equivalent read)")
if (nrow(dg_tbl) && any(!dg_tbl$ok))
  message("RED (expected while any tile has >1 part file): ",
          paste(sprintf("%s/%s", dg_tbl$ver[!dg_tbl$ok], dg_tbl$metric_key[!dg_tbl$ok]), collapse = ", "))
```

### `boot.json` zone metrics == `zone_metric` (joined on `zone_type` AND `metric_key`)

```{r}
#| label: validate-zone-metrics

zone_metric_gate <- function(b, ver) {
  if (!isTRUE(b$ok) || is.null(b$out$boot)) return(tibble())
  con <- app_bundle_open_scoring(ver)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  duckdb_tune(con, file.path(tempdir(), "app_bundle_spill_zm", ver))
  vm <- msens::sdm_val_col(con, "zone_metric"); vz <- msens::sdm_val_col(con, "zone")
  pub <- DBI::dbGetQuery(con, glue::glue("
    SELECT z.fld, z.{vz} AS zkey, m.metric_key, zm.{vm} AS val
      FROM zone z JOIN zone_metric zm USING (zone_seq) JOIN metric m USING (metric_seq)"))
  pub$zone_type <- sub("_key$", "", pub$fld)
  boot <- b$out$boot
  boot_rows <- do.call(rbind, lapply(names(boot$zones), function(ztype) {
    zl <- boot$zones[[ztype]]
    do.call(rbind, lapply(zl, function(zr) {
      if (!length(zr$metrics)) return(NULL)
      data.frame(zone_type = ztype, zkey = zr$key, metric_key = names(zr$metrics),
                val = as.numeric(unlist(zr$metrics)), stringsAsFactors = FALSE)
    }))
  }))
  j <- merge(pub, boot_rows, by = c("zone_type", "zkey", "metric_key"), suffixes = c("_pub", "_boot"))
  data.frame(ver = ver, n_pub = nrow(pub), n_boot = nrow(boot_rows), n_joined = nrow(j),
             max_abs_delta = if (nrow(j)) max(abs(j$val_pub - j$val_boot)) else NA_real_)
}
zm_tbl <- purrr::imap(builds, zone_metric_gate) |> bind_rows()
knitr::kable(zm_tbl, caption = "boot.json zone metrics vs published zone_metric (join on zone_type + zkey + metric_key)")
stopifnot("boot.json zone metrics disagree with zone_metric" =
  nrow(zm_tbl) == 0 || all(zm_tbl$n_pub == zm_tbl$n_boot & zm_tbl$max_abs_delta < 1e-6, na.rm = TRUE))
```

### Layer list == manifest's drawable (cell-level) metrics; shard keys == `taxa.json`; every alias resolves

Reads the WRITTEN `taxa.json` / `taxon/*.json` / `alias/*.json` from disk — `app_bundle_build()`
does not return the full shard contents in memory, only counts, and the point of this gate is the
published bytes anyway, not what the builder happened to hold before writing them.

```{r}
#| label: validate-sets

set_gates <- function(b, ver) {
  if (!isTRUE(b$ok) || is.null(b$out$boot)) return(tibble())
  con <- app_bundle_open_scoring(ver)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  cell_keys  <- sort(unique(msens:::.app_cell_metric_keys(con)))
  layer_keys <- sort(vapply(b$out$boot$layers, function(l) l$metric_key, ""))
  layers_ok  <- identical(cell_keys, layer_keys)

  taxa_path <- file.path(b$dir_out, "taxa.json")
  taxa_keys <- if (file.exists(taxa_path))
    unique(unname(unlist(jsonlite::fromJSON(taxa_path, simplifyVector = FALSE)$key))) else character()

  shards_ok <- NA; n_shard_keys <- NA_integer_
  taxon_dir <- file.path(b$dir_out, "taxon")
  if (dir.exists(taxon_dir) && length(taxa_keys)) {
    shard_keys <- unique(unname(unlist(lapply(list.files(taxon_dir, full.names = TRUE), function(f)
      names(jsonlite::fromJSON(f, simplifyVector = FALSE)$taxa)))))
    shards_ok <- setequal(shard_keys, taxa_keys)
    n_shard_keys <- length(shard_keys)
  }

  alias_ok <- NA; n_alias_bad <- NA_integer_
  alias_dir <- file.path(b$dir_out, "alias")
  if (dir.exists(alias_dir) && length(taxa_keys)) {
    targets <- unname(unlist(lapply(list.files(alias_dir, full.names = TRUE), function(f) {
      al <- jsonlite::fromJSON(f, simplifyVector = FALSE)$alias
      vapply(al, function(x) x[[1]], "")
    })))
    bad <- setdiff(unique(targets), taxa_keys)
    alias_ok <- length(bad) == 0
    n_alias_bad <- length(bad)
  }
  data.frame(ver = ver, layers_eq_cell_metrics = layers_ok,
             n_cell_keys = length(cell_keys), n_layer_keys = length(layer_keys),
             shard_keys_eq_taxa = shards_ok, n_shard_keys = n_shard_keys, n_taxa_keys = length(taxa_keys),
             every_alias_resolves = alias_ok, n_alias_unresolved = n_alias_bad)
}
sets_tbl <- purrr::imap(builds, set_gates) |> bind_rows()
knitr::kable(sets_tbl, caption = "Set-equality gates: layers, taxon shard keys, alias targets")
```

## Budgets (bytes vs the subplan's per-object budgets — FAILS the render, not just flags)

Review round 1: the notebook used to carry its own private thresholds (9 MB for
`zone_taxon.parquet`, 1.5 MB for `taxon.parquet`) that the two real measurements could never
exceed, and only ever printed a table rather than stopping anything. [app_bundle_budgets()]
(`libs/app_bundle.R`) is now the ONE place those numbers live — the subplan's own, with
`zone_taxon.parquet` corrected to 8 MB — and this chunk `stopifnot()`s on the result: an object
over budget fails the render, exactly like every other gate here. Sizes are measured the same
way a real push would produce them: gzip via [app_bundle_gzip_file()] (byte-safe, never
`readLines()`/`writeLines()`) for JSON, raw bytes for Parquet; the cell-tile row reads ONLY
`tile=*/data_0.parquet` (browser-equivalent, see the digest gate above).

```{r}
#| label: budgets

budgets_spec <- app_bundle_budgets()

.budget_row <- function(object, sizes, kind, budget_bytes) {
  if (!length(sizes)) return(NULL)
  data.frame(object = object, kind = kind, n = length(sizes),
             bytes_p50 = as.integer(round(median(sizes))),
             bytes_p90 = as.integer(round(quantile(sizes, 0.9))),
             bytes_max = as.integer(max(sizes)),
             budget_bytes = as.integer(budget_bytes),
             n_over_budget = sum(sizes > budget_bytes))
}
.sizes_for <- function(dir_out, object, kind) {
  gz_of <- function(p) { tmp <- tempfile(fileext = ".gz"); app_bundle_gzip_file(p, tmp)
                         sz <- file.size(tmp); unlink(tmp); sz }
  if (object %in% c("boot.json", "taxa.json")) {
    p <- file.path(dir_out, object)
    if (!file.exists(p)) return(numeric())
    if (kind == "gzip") gz_of(p) else file.size(p)
  } else if (object %in% c("taxon/*.json", "alias/*.json")) {
    d <- file.path(dir_out, sub("/\\*\\.json$", "", object))
    if (!dir.exists(d)) return(numeric())
    fs_ <- list.files(d, full.names = TRUE)
    if (kind == "gzip") vapply(fs_, gz_of, numeric(1)) else file.size(fs_)
  } else if (object == "cell/tile=*/data_0.parquet") {
    fs_ <- Sys.glob(file.path(dir_out, "cell", "tile=*", "data_0.parquet"))  # browser-equivalent
    file.size(fs_)
  } else {  # taxon.parquet, taxonomy.parquet, zone_taxon.parquet
    p <- file.path(dir_out, object)
    if (!file.exists(p)) numeric() else file.size(p)
  }
}

budgets_one <- function(b) {
  if (!isTRUE(b$ok)) return(tibble())
  rows <- lapply(seq_len(nrow(budgets_spec)), function(i) {
    spec <- budgets_spec[i, ]
    .budget_row(spec$object, .sizes_for(b$dir_out, spec$object, spec$kind), spec$kind, spec$budget_bytes)
  })
  out <- bind_rows(Filter(Negate(is.null), rows))
  if (nrow(out)) out$ver <- b$ver
  out
}
budgets_tbl <- bind_rows(purrr::map(builds, budgets_one))
knitr::kable(
  budgets_tbl |> select(ver, object, kind, n, bytes_p50, bytes_p90, bytes_max, budget_bytes, n_over_budget),
  caption = "Object budgets vs the subplan (bytes; n_over_budget must be 0)")
stopifnot(
  "an object exceeds its subplan budget" = nrow(budgets_tbl) == 0 || all(budgets_tbl$n_over_budget == 0))
```

## Per-version summary (objects, bytes, taxa, tiles)

```{r}
#| label: summary-per-version

summary_one <- function(b) {
  if (!isTRUE(b$ok))
    return(data.frame(ver = b$ver, ok = FALSE, reason = b$reason %||% NA_character_, n_objects = NA,
                      bytes_raw = NA, bytes_gzip_est = NA, n_taxa = NA, n_tiles = NA, wall_s = NA))
  d <- b$dir_out
  files <- list.files(d, recursive = TRUE, full.names = TRUE)
  is_json <- grepl("\\.json$", files)
  gz1 <- function(p) { tmp <- tempfile(fileext = ".gz"); app_bundle_gzip_file(p, tmp)
                       sz <- file.size(tmp); unlink(tmp); sz }
  raw <- sum(file.size(files))
  gz  <- sum(vapply(files[is_json], gz1, numeric(1))) + sum(file.size(files[!is_json]))
  data.frame(ver = b$ver, ok = TRUE, reason = "",
             n_objects = length(files), bytes_raw = raw, bytes_gzip_est = gz,
             n_taxa = if (!is.null(b$out)) b$out$n_taxa else NA,
             n_tiles = if (!is.null(b$out$tiles)) b$out$tiles$tiles else NA,
             wall_s = round(b$wall_s, 1))
}
summary_tbl <- bind_rows(purrr::map(builds, summary_one))
knitr::kable(summary_tbl, caption = "Per-version summary: objects, raw/gzip-estimated bytes, taxa, tiles, wall clock")
```

## Push (`APP_BUNDLE_S3=1` only — never run by this notebook's author)

Lists the target prefix before and after, refuses (via [app_bundle_assert_prefix()]) any key
outside `{ver}/app/` — and `v7/serve/cell_model/`, only when `APP_BUNDLE_V7_CELLMODEL=1` — gzips
every JSON object before upload (`Content-Encoding: gzip`, S3 does not compress on the fly),
leaves Parquet as-is (already zstd-compressed by [msens::write_atlas_parquet()] /
[msens::copy_atlas_parquet()]).

```{r}
#| label: push
#| eval: !expr do_s3

s3_root <- "s3://oceanmetrics.io-public/marine-atlas"

list_prefix <- function(prefix) {
  out <- system2("aws", c("s3", "ls", shQuote(glue("{s3_root}/{prefix}")), "--recursive"),
                stdout = TRUE, stderr = TRUE)
  if (!is.null(attr(out, "status")) && attr(out, "status") != 0) character(0) else out
}

push_one <- function(b) {
  if (!isTRUE(b$ok)) return(invisible(NULL))
  ver <- b$ver
  before <- list_prefix(glue("{ver}/app/"))
  log_info("{ver}/app/ BEFORE push: {length(before)} object(s)")

  # a controlled directory walk of dir_out -- every file that exists under it,
  # nothing else -- then EVERY key derived from it still goes through the
  # whitelist guard before anything touches S3; the walk being "controlled"
  # is not itself the protection, the guard call two lines down is
  files <- list.files(b$dir_out, recursive = TRUE, full.names = TRUE)
  rels  <- fs::path_rel(files, b$dir_out)
  keys  <- glue("{ver}/app/{rels}")
  app_bundle_assert_prefix(keys, ver)                       # refuse before touching S3

  for (i in seq_along(files)) {
    is_json <- grepl("\\.json$", files[i])
    src <- files[i]
    if (is_json) {
      gz <- tempfile(fileext = ".gz")
      app_bundle_gzip_file(files[i], gz)                    # byte-safe, never readLines()
      app_bundle_assert_gzip_roundtrip(gz, files[i])         # proves upload bytes == validated bytes
      src <- gz
    }
    args <- c("s3", "cp", shQuote(src), shQuote(glue("{s3_root}/{keys[i]}")),
              if (is_json) c("--content-type", shQuote("application/json"),
                             "--content-encoding", "gzip"))
    out <- system2("aws", args, stdout = TRUE, stderr = TRUE)
    if (!is.null(attr(out, "status")) && attr(out, "status") != 0)
      stop(glue("aws s3 cp failed for {keys[i]}: {paste(out, collapse = '\n')}"))
    if (is_json) unlink(src)
  }

  if (isTRUE(do_v7_cm) && ver == "v7") {
    cm_local <- path.expand("~/_big/msens/derived/v7/cell_model")
    if (dir.exists(cm_local)) {
      cm_files <- list.files(cm_local, recursive = TRUE, full.names = TRUE)
      cm_keys  <- glue("v7/serve/cell_model/{fs::path_rel(cm_files, cm_local)}")
      app_bundle_assert_prefix(cm_keys, "v7", allow_v7_cell_model = TRUE)
      system2("aws", c("s3", "sync", shQuote(cm_local), shQuote(glue("{s3_root}/v7/serve/cell_model")),
                       "--only-show-errors"), stdout = TRUE, stderr = TRUE)
    } else log_warn("APP_BUNDLE_V7_CELLMODEL=1 but {cm_local} does not exist locally")
  }

  after <- list_prefix(glue("{ver}/app/"))
  log_info("{ver}/app/ AFTER push: {length(after)} object(s) ({length(after) - length(before)} new)")
  invisible(list(before = before, after = after))
}
push_out <- purrr::map(builds, push_one)
```

## Anonymous verification (`APP_BUNDLE_S3=1` only)

HEAD (falling back to a ranged GET) per published object, anonymously — no credentials, exactly
as the browser will fetch it: 200, `content-encoding: gzip` on every JSON object, path-style URL,
size within budget.

```{r}
#| label: verify-anonymous
#| eval: !expr do_s3

verify_one <- function(b) {
  if (!isTRUE(b$ok)) return(tibble())
  ver <- b$ver
  files <- list.files(b$dir_out, recursive = TRUE, full.names = TRUE)
  rels  <- fs::path_rel(files, b$dir_out)
  urls  <- glue("{base}/{ver}/app/{rels}")
  purrr::map_dfr(seq_along(urls), function(i) {
    r <- tryCatch(
      httr2::req_perform(httr2::req_error(httr2::req_method(httr2::request(urls[i]), "HEAD"),
                                          is_error = function(resp) FALSE)),
      error = function(e) NULL)
    is_json <- grepl("\\.json$", files[i])
    data.frame(
      ver = ver, key = rels[i],
      status = if (is.null(r)) NA_integer_ else httr2::resp_status(r),
      path_style = grepl("^https://s3\\.[a-z0-9-]+\\.amazonaws\\.com/", urls[i]),
      gzip_ok = if (is.null(r) || !is_json) NA else
        identical(httr2::resp_header(r, "content-encoding"), "gzip"),
      bytes = if (is.null(r)) NA_integer_ else
        suppressWarnings(as.integer(httr2::resp_header(r, "content-length"))))
  })
}
verify_tbl <- bind_rows(purrr::map(builds, verify_one))
knitr::kable(
  verify_tbl |> summarise(n = n(), n_200 = sum(status == 200, na.rm = TRUE),
                          n_gzip_ok = sum(gzip_ok, na.rm = TRUE), .by = ver),
  caption = "Anonymous verification summary per version")
stopifnot("not every pushed object answered 200 anonymously" =
  all(verify_tbl$status == 200, na.rm = TRUE) && !anyNA(verify_tbl$status))
```

## Target manifest

```{r}
#| label: target-manifest

msens::write_manifest(
  here("data/manifests/build_app_bundle.json"),
  target       = "build_app_bundle",
  content_hash = digest::digest(
    list(summary_tbl, val_tbl, boot_contract_tbl, one_file_tbl, tw_tbl, dg_tbl, zm_tbl, sets_tbl,
        budgets_tbl),
    algo = "xxhash64"),
  stats = list(
    versions_attempted = vers,
    versions_built      = summary_tbl$ver[summary_tbl$ok],
    push_s3             = do_s3,
    v7_cell_model       = do_v7_cm,
    n_objects_total     = sum(summary_tbl$n_objects, na.rm = TRUE),
    n_tile_dirs_with_multiple_files = sum(one_file_tbl$n_with_multiple_files, na.rm = TRUE)),
  force = msens::force_target("build_app_bundle"))
```
