---
title: "Build global 0.05° cell grid (v8 sampling unit)"
subtitle: "Global ocean 0.05° raster cells in [-180,180] with env covariates + US study-area / Program-Area membership"
msens:
  target_name: build_cell_grid
  workflow_type: grid
  dependency: []
  output: data/manifests/build_cell_grid.json
format: html
editor_options:
  chunk_output_type: console
---

The v8 spatial sampling unit is the **global 0.05° raster cell** in **[-180,180]**
(rolled back from H3 hex). The grid is anchored to the shared 0.05°, 3600×7200,
EPSG:4326 topology of **Bio-Oracle v3 AND the AquaX SDM TIFs** — so `cell_id` is
the raster cell index (`1:ncell`), and any SDM raster on this grid (Bio-Oracle,
AquaX, resampled AquaMaps) maps to `cell_id` by position with **no projection**.

This notebook builds the canonical `cell` table (ocean cells only) — `cell_id`,
the 12 Bio-Oracle env covariates, `area_km2`, and boolean `in_usa` / `in_pra`
membership — plus the global cell-id COG that titiler paints.

```{r setup}
librarian::shelf(DBI, dplyr, duckdb, glue, here, jsonlite, sf, terra, quiet = TRUE)
source(here("libs/paths.R"))   # ver, sdm_db, bio_oracle_tif, cellid_tif, ply_usa_gpkg, ply_pra_gpkg
stopifnot(file.exists(bio_oracle_tif), file.exists(ply_usa_gpkg), file.exists(ply_pra_gpkg))

# optional bbox restriction for prototype runs: CELL_BBOX="xmin,xmax,ymin,ymax"
cell_bbox <- Sys.getenv("CELL_BBOX", "")
cell_db   <- Sys.getenv("CELL_DB", sdm_db)
dir.create(dirname(cell_db), recursive = TRUE, showWarnings = FALSE)
dir.create(here("data/manifests"), recursive = TRUE, showWarnings = FALSE)
```

## Read Bio-Oracle, assign global cell_id, membership + area

```{r build_grid}
r <- rast(bio_oracle_tif)   # global [-180,180], 3600 x 7200, 12 env layers
r$cell_id  <- 1:ncell(r)    # global raster-position id (shared with AquaX/AquaMaps grids)
r$area_km2 <- cellSize(r$depth_mean, unit = "km")

# membership: rasterize the study-area + Program-Area polygons onto the same grid
r$in_usa <- rasterize(vect(st_read(ply_usa_gpkg, quiet = TRUE)), r$depth_mean, field = 1, background = 0)
r$in_pra <- rasterize(vect(st_read(ply_pra_gpkg, quiet = TRUE)), r$depth_mean, field = 1, background = 0)

if (nzchar(cell_bbox)) {
  bb <- as.numeric(strsplit(cell_bbox, ",")[[1]])
  r  <- crop(r, ext(bb[1], bb[2], bb[3], bb[4]))
  message(glue("restricted to bbox [{cell_bbox}]"))
}
```

## Write cell table (ocean cells) + global cell-id COG

```{r write_cell}
# ocean cells only (depth_mean non-NA); cell_id stays the global raster index
d_cell <- as.data.frame(r, xy = TRUE, na.rm = FALSE) |>
  filter(!is.na(depth_mean)) |>
  transmute(
    cell_id = as.integer(cell_id),
    lon = x, lat = y,
    across(c(depth_mean, depth_min, depth_max, oxy_b_mean, oxy_mean, prim_prod_mean,
             ice_con_ann, salinity_b_mean, salinity_mean, sbt_an_mean, sst_an_mean, fao_area_m)),
    area_km2,
    in_usa = in_usa == 1,
    in_pra = in_pra == 1)

con <- dbConnect(duckdb(cell_db))
dbWriteTable(con, "cell", d_cell, overwrite = TRUE)
dbExecute(con, "ALTER TABLE cell ADD PRIMARY KEY (cell_id)")

# global cell-id COG in [-180,180] for titiler (uint32; no overviews to keep ids intact)
writeRaster(r$cell_id, cellid_tif, filetype = "COG", datatype = "INT4U", overwrite = TRUE,
            gdal = c("COMPRESS=LZW", "OVERVIEWS=NONE"))
```

## Summary + manifest

```{r summary}
smry <- dbGetQuery(con, "
  SELECT count(*) AS n_cell,
         count(*) FILTER (WHERE in_usa) AS n_usa,
         count(*) FILTER (WHERE in_pra) AS n_pra,
         round(min(area_km2), 3) AS area_min_km2,
         round(max(area_km2), 3) AS area_max_km2,
         round(min(lon),1) AS lon_min, round(max(lon),1) AS lon_max,
         count(*) FILTER (WHERE depth_mean IS NULL) AS n_null_depth
  FROM cell")
knitr::kable(smry)

manifest <- c(as.list(smry), list(
  version = ver, bbox = cell_bbox, db = cell_db,
  cellid_tif = cellid_tif, source = basename(bio_oracle_tif)))
write_json(manifest, here("data/manifests/build_cell_grid.json"), auto_unbox = TRUE, pretty = TRUE)

dbDisconnect(con, shutdown = TRUE)
message(glue("cell grid: {smry$n_cell} cells ({smry$n_usa} in US, {smry$n_pra} in PRAs) -> {cell_db}"))
```
