---
title: "Build the version registry (`latest.txt`, `versions.json`, `manifest.json`)"
msens:
  target_name: build_version_manifest
  workflow_type: release
  dependency: [score_zone_metrics]
  output: data/manifests/build_version_manifest.json
editor_options:
  chunk_output_type: console
---

Every MST release so far has shipped by **forking the apps repo** — freeze `apps` at a commit,
clone it to `apps_v{n}`, symlink `/share/shiny_apps/{scores,species}_v{n}` at it. Nine such forks
exist, and every app improvement is stranded in the newest one: v6's app will never get v8's
deep-link handling or a bug fix, because it is a different checkout of a different commit.

The cause is that the version is a **code** fact — `ver <- "v8"` hardcoded in each app, propagating
into data paths, table names, a per-version titiler service and the symlink name. This notebook
publishes the files that make it a **data** fact instead, so one app can render any release:

| file | what it settles |
|---|---|
| `marine-atlas/latest.txt` | which version an app shows when the URL says nothing |
| `marine-atlas/versions.json` | which versions exist, whether each is showable, and to whom (`access`) |
| `marine-atlas/{ver}/manifest.json` | everything needed to *render* that version |

The manifest is the contract: publishing v9 means publishing a manifest, **not** editing an app.

> **Promotion is gated and separate from publishing.** Writing data does not make it live —
> `latest.txt` is only rewritten under `PROMOTE_LATEST=1`, after the validation chunk passes. So
> `latest.txt` today names **v7** (the default-live release), while v8 remains a `prerelease`
> reachable only as `?ver=v8`. This mirrors the CalCOFI release registry, where promotion is gated
> on a query-test pass rather than done by the build.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "Curated registry + release introspection → the three published registry files"
flowchart LR
  csv["data/versions.csv<br/>(curated: status, access, date, title)"] --> vj["versions.json"]
  db[("{ver}/sdm.duckdb")] --> mb["msens::manifest_build()<br/>introspects tables, metrics,<br/>zones, public model id"]
  mb --> mj["{ver}/manifest.json"]
  vj --> s3[["s3://…/marine-atlas/"]]
  mj --> s3
  vj -.->|"PROMOTE_LATEST=1<br/>after validation"| lt["latest.txt"]
  lt --> s3
```

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(
  arrow, DBI, dplyr, duckdb, glue, jsonlite, knitr, readr,
  MarineSensitivity/msens, quiet = TRUE)
source(here::here("libs/paths.R"))

# env flags (default OFF, per the repo's reproducibility rule: expensive or
# side-effecting steps are opt-in, and every step lives in this notebook)
do_s3    <- Sys.getenv("MANIFEST_NO_S3")   == ""   # push to S3 unless told not to
promote  <- Sys.getenv("PROMOTE_LATEST")   != ""   # rewrite latest.txt (gated below)

s3_root  <- s3_atlas                                # s3://oceanmetrics.io-public/marine-atlas
http_root <- msens::atlas_base_url()
dir_reg  <- glue("{dir_big_v}/registry-files")      # local staging, mirrors the S3 layout
dir.create(glue("{dir_reg}/{ver}"), recursive = TRUE, showWarnings = FALSE)

message(glue("ver={ver}  s3={s3_root}  push_s3={do_s3}  promote_latest={promote}"))
```

## The curated version registry

Release *status* and *dates* are editorial facts — they cannot be introspected from a database, and
they must not be guessed. They live in a committed CSV so the registry is reviewable in a diff, and
so the apps, the versioned docs and the storage index all read one source rather than three.

Deliberately a **registry, not a changelog**: `ver`, `status`, `access`, `released`, `title` and
nothing more. What each release *changed* is prose, it already lives in the docs timeline, and a
second copy here would drift from it — which is the very problem the versioned-docs work exists to
fix.

`status` and `access` are **separate axes**: `status` says where a release is in its life
(`prerelease` / `released` / `retired`); `access` says *who may be shown it* (`public` /
`restricted`). A pre-release under review by SDM data providers and BOEM/NOAA colleagues is
`restricted` — the apps and docs serve it only through the signed-in
`preview.marinesensitivity.org` host (`msens::atlas_allow_access()`), while retired releases stay
`public` because citations point at them. The column is written explicitly here; a reader that
meets a `versions.json` predating it derives `access` fail-closed (`prerelease → restricted`, see
`msens::atlas_access_default()`), so the column being present with `public` is what keeps a
public pre-release public.

```{r}
#| label: versions

versions <- readr::read_csv(here::here("data/versions.csv"), show_col_types = FALSE)

# a registry that predates the column gets the same fail-closed default the readers apply
if (!"access" %in% names(versions)) versions$access <- msens::atlas_access_default(versions$status)

stopifnot(
  "versions.csv has duplicate ver"        = !any(duplicated(versions$ver)),
  "unknown status in versions.csv"        = all(versions$status %in% c("released", "prerelease", "retired")),
  "unknown access in versions.csv"        = all(versions$access %in% c("public", "restricted")),
  "current ver missing from versions.csv" = ver %in% versions$ver,
  "no released version to promote"        = any(versions$status == "released"))

# The PROMOTED release -- what every app and the docs root show when the URL says
# nothing -- can never sit behind a login. Asserted here on the registry, so a
# `latest.txt` naming a restricted version cannot be written in the first place.
promoted <- versions |> filter(status == "released") |> arrange(desc(released)) |> slice_head(n = 1)
stopifnot("the newest released version (what latest.txt names) must be public" =
            promoted$access == "public")

# newest first, so a consumer can take the head without sorting
versions <- versions |> arrange(desc(released))
knitr::kable(versions |> select(ver, status, access, released, title))
```

## The manifest for this version

`msens::manifest_build()` *introspects* the release rather than being told about it — which model
id is public (`mdl_key` from v8, `mdl_seq` before it), which tables exist, which metrics are
cell-level, which spatial units were scored. That is what keeps a manifest from drifting from the
data it describes, and it is why capabilities derive from **presence** and default to FALSE: a
release without `cell_model` must not advertise a per-cell species list.

```{r}
#| label: manifest

con <- dbConnect(duckdb(), path.expand(sdm_db), read_only = TRUE)

# Some surfaces are NOT tables in the build database, and are not on S3 either.
# v8's `cell_model` (the per-cell species list) is a partitioned Parquet
# DIRECTORY that lives only on the server — deliberately, since its per-cell and
# per-polygon queries touch many partitions and fail outright over HTTPS. So:
#   - deriving from `con` alone advertises cell_species_list = FALSE and switches
#     off a panel that works today;
#   - checking S3 also says FALSE, because it is intentionally not published there.
# The authoritative test is the SERVING surface on the server. Check it there
# (read-only) and print the answer, so a manifest built from a laptop describes
# the release rather than the laptop.
srv_has <- function(path) {
  if (dir.exists(path.expand(glue("{dir_big_v}/{path}")))) return(TRUE)   # running ON the server
  out <- suppressWarnings(system2(
    "ssh", c("-o", "ConnectTimeout=15", "msens",
             shQuote(glue("test -e /share/data/big/{ver}/{path} && echo yes || echo no"))),
    stdout = TRUE, stderr = FALSE))
  isTRUE(any(grepl("^yes$", trimws(out))))
}

caps_override <- list(cell_species_list = srv_has("cell_model"))
message(glue("serving surface check: cell_model present = {caps_override$cell_species_list}"))

# score COGs, when published: their presence is what flips capabilities$score_cogs,
# so the app only offers a raster layer once the raster exists
f_scog   <- path.expand(glue("{dir_big_v}/tables/score_cog.parquet"))
scog_all <- if (file.exists(f_scog)) arrow::read_parquet(f_scog) else NULL

# `_`-prefixed keys are OVERLAYS, not metrics: the app's "cells outside Program
# Areas" mask is a raster it draws but not a quantity it scores. They must be
# separated, because manifest_build() left-joins metrics onto the release's
# `metric` table -- an overlay has no row there and was silently dropped.
scog <- if (!is.null(scog_all))
  scog_all |> filter(!startsWith(metric_key, "_")) |>
    transmute(metric_key, subregion_key, cog = cog_url,
              rescale_min, rescale_max, colormap) else NULL
overlays <- if (!is.null(scog_all))
  scog_all |> filter(startsWith(metric_key, "_")) |>
    transmute(overlay_key = metric_key, subregion_key, cog = cog_url, colormap) else NULL
message(glue("score COGs: {if (is.null(scog)) 0 else nrow(scog)} metric, ",
             "{if (is.null(overlays)) 0 else nrow(overlays)} overlay"))

# layer presentation metadata (order/category/short label) moves into the
# manifest so the app stops needing a per-version layers_{ver}.csv on disk --
# see backfill_versions.qmd for the same join on the historical releases
f_lyrs <- path.expand(glue("{dir_derived}/{ver}/layers_{ver}.csv"))
if (!is.null(scog) && file.exists(f_lyrs)) {
  lyr_meta <- readr::read_csv(f_lyrs, show_col_types = FALSE) |>
    dplyr::transmute(metric_key = lyr, lyr_order = order, category, label = layer)
  scog <- dplyr::left_join(scog, lyr_meta, by = "metric_key")
  message(glue("layer metadata: {nrow(lyr_meta)} rows"))
}

# zone PMTiles by vintage, so the app resolves outlines from the manifest rather
# than a hardcoded unversioned filename on the file host
zone_sets <- readr::read_csv(here::here("data/zone_sets.csv"), show_col_types = FALSE)
zone_tiles <- setNames(
  as.list(glue("{http_root}/zones/{zone_sets$zone_set_key}/zones.pmtiles")),
  zone_sets$zone_set_key)
message(glue("zone tile sets: {length(zone_tiles)}"))

v_row <- versions |> filter(ver == !!ver)
mfst  <- msens::manifest_build(
  con, ver,
  status       = v_row$status,
  access       = v_row$access,
  base         = http_root,
  metrics      = scog,
  zone_tiles   = zone_tiles,
  capabilities = caps_override,
  extra  = list(
    title    = v_row$title,
    released = as.character(v_row$released),
    overlays = overlays,
    # score COGs and the vintage-labelled zone-set registry arrive in later
    # phases; until then the manifest declares what is true today, and
    # `capabilities$score_cogs` stays FALSE so no app promises a layer it
    # cannot draw.
    grid     = as.list(msens::grid_spec_for(msens::grid_for_ver(ver))[
                 c("nc", "nr", "xmin", "ymax", "resx", "resy", "lon360")])))

dbDisconnect(con, shutdown = TRUE)

cat(glue(
  "ver={mfst$ver}  status={mfst$status}  access={mfst$access}  grid={mfst$grid_id}  id_field={mfst$id_field}\n",
  "tables={length(mfst$tables)}  metrics={nrow(mfst$metrics)}  zone_sets={nrow(mfst$zones)}\n"))
knitr::kable(mfst$zones)
```

```{r}
#| label: capabilities

knitr::kable(
  data.frame(capability = names(mfst$capabilities),
             supported  = unlist(mfst$capabilities)),
  row.names = FALSE)
```

## Validate, then write

The same `validate_manifest()` the apps call on read is applied here on write, so a manifest that
would break an app can never be published in the first place.

```{r}
#| label: write

msens::validate_manifest(mfst, ver = ver)

# jsonlite writes NULL as [] and NA as null; pretty-print for reviewable diffs
wr <- function(obj, path) {
  writeLines(as.character(jsonlite::toJSON(obj, auto_unbox = TRUE, pretty = TRUE, na = "null")), path)
  path
}

f_versions <- wr(list(versions = versions), glue("{dir_reg}/versions.json"))
f_manifest <- wr(mfst,                      glue("{dir_reg}/{ver}/manifest.json"))

# round-trip: parse what we just wrote, exactly as an app would
msens::validate_manifest(jsonlite::fromJSON(f_manifest), ver = ver)
cat(glue("wrote {f_versions}\n     {f_manifest}\n"))
```

## Promote `latest.txt` (gated)

`latest.txt` is the one file that changes what users see by default, so it is written **only** when
`PROMOTE_LATEST=1` and only naming a `released` version. A pre-release is never promoted implicitly
— it stays reachable at `?ver={ver}` until someone decides otherwise.

```{r}
#| label: promote

latest <- versions |> filter(status == "released") |> slice_head(n = 1) |> pull(ver)

if (promote) {
  f_latest <- glue("{dir_reg}/latest.txt")
  writeLines(latest, f_latest)
  cat(glue("PROMOTED latest.txt -> {latest}\n"))
} else {
  cat(glue("latest.txt would be '{latest}' (set PROMOTE_LATEST=1 to write it)\n"))
}
```

## Publish to S3

```{r}
#| label: s3

if (do_s3) {
  # system2() runs the command through a shell, so unquoted glob patterns are
  # expanded by the shell against the CWD before aws ever sees them (an
  # unquoted `--exclude *` becomes the entire repo listing). shQuote them, and
  # expand `~` ourselves since the CLI does not.
  args <- c("s3", "sync", shQuote(path.expand(dir_reg)), shQuote(s3_root),
            "--exclude", shQuote("*"),
            "--include", shQuote("*.json"), "--include", shQuote("latest.txt"),
            "--content-type", shQuote("application/json"),
            # these three are MUTABLE pointers; S3 has no default cache-control but
            # CDNs and browsers will happily hold a stale registry, which shows the
            # wrong version list long after a promotion
            "--cache-control", shQuote("no-cache"),
            "--only-show-errors", "--no-progress")
  out <- system2("aws", args, stdout = TRUE, stderr = TRUE)
  if (!is.null(attr(out, "status")) && attr(out, "status") != 0)
    stop("aws s3 sync failed: ", paste(out, collapse = "\n"))

  # verify by READING BACK through the public URL, not by trusting the exit
  # status — a zero exit has already proved not to mean the objects are there
  # (CalCOFI's index publisher learned this the expensive way)
  got <- msens::atlas_manifest(ver, base = http_root, refresh = TRUE)
  stopifnot("published manifest does not round-trip" = identical(got$ver, ver))
  # ...and the registry itself, read exactly as an app reads it: the `access` an
  # app will enforce must be the `access` this notebook just wrote (a stale
  # versions.json here would let the readers' fail-closed default hide a public
  # pre-release, or an edited CSV fail to restrict one)
  got_v <- msens::atlas_versions(base = http_root, refresh = TRUE)
  stopifnot("published versions.json access does not match versions.csv" =
              identical(got_v$access[match(versions$ver, got_v$ver)], versions$access))
  cat(glue("verified {http_root}/{ver}/manifest.json and versions.json (access column)\n"))
} else {
  cat("MANIFEST_NO_S3 set - staged locally only\n")
}
```

## Target manifest

```{r}
#| label: target-manifest

msens::write_manifest(
  here::here("data/manifests/build_version_manifest.json"),
  target        = "build_version_manifest",
  content_hash  = digest::digest(list(mfst, versions), algo = "xxhash64"),
  stats = list(
    ver          = ver,
    status       = mfst$status,
    access       = mfst$access,
    grid_id      = mfst$grid_id,
    id_field     = mfst$id_field,
    n_versions   = nrow(versions),
    n_metrics    = nrow(mfst$metrics),
    n_zone_sets  = nrow(mfst$zones),
    latest       = latest,
    promoted     = promote),
  force = msens::force_target("build_version_manifest"))
```
