---
title: "Build H3 hex grid (v8 sampling unit)"
subtitle: "Global res-7 ocean hexagons with env covariates + US study-area / Program-Area membership"
msens:
  target_name: build_hex_grid
  workflow_type: grid
  dependency: []
  output: data/manifests/build_hex_grid.json
format: html
editor_options:
  chunk_output_type: console
---

The v8 spatial sampling unit is the **H3 resolution-7 hexagon** (~5.16 km²),
replacing the 0.05° raster cell. This notebook builds the canonical `hex` table:
every ocean hexagon globally, each carrying the 12 Bio-Oracle environmental
covariates (inherited from the 0.05° pixel that contains the hex center), its
`area_km2`, and boolean `in_usa` / `in_pra` membership flags.

H3 res-7 ids are 64-bit and exceed R's exact-double range, so all indexing,
area, and membership run inside DuckDB via the `h3` (+ `spatial`) extension — the
same extension the OBIS `h3t` tile service uses, so hex ids align exactly for the
OBIS overlay. See `msens::hex_grid_from_raster()` / `msens::hex_add_membership()`.

```{r setup}
librarian::shelf(DBI, dplyr, duckdb, glue, here, jsonlite, terra, quiet = TRUE)
devtools::load_all(here("../msens"), quiet = TRUE)  # hex + grid helpers
source(here("libs/paths.R"))                        # ver, sdm_db, bio_oracle_tif

# study-area (to-shore US EEZ) + Program-Area polygons for membership
ply_usa <- glue("{dir_derived}/v1/ply_boem-usa.gpkg")                    # shared
ply_pra <- glue("{dir_derived}/{ver_prev}/ply_programareas_2026_{ver_prev}.gpkg")
stopifnot(file.exists(bio_oracle_tif), file.exists(ply_usa), file.exists(ply_pra))

# optional bbox restriction for prototype runs: HEX_BBOX="xmin,xmax,ymin,ymax"
# (unset => full global build). DB path can be overridden with HEX_DB.
hex_bbox <- Sys.getenv("HEX_BBOX", "")
hex_db   <- Sys.getenv("HEX_DB", sdm_db)
dir.create(dirname(hex_db), recursive = TRUE, showWarnings = FALSE)
dir.create(here("data/manifests"), recursive = TRUE, showWarnings = FALSE)
```

## Read Bio-Oracle + build hex grid

```{r build_grid}
r <- rast(bio_oracle_tif)
if (nzchar(hex_bbox)) {
  bb <- as.numeric(strsplit(hex_bbox, ",")[[1]])
  r  <- crop(r, ext(bb[1], bb[2], bb[3], bb[4]))
  message(glue("restricted to bbox [{hex_bbox}] ({ncell(r)} pixels)"))
}

# bigint="integer64" so BIGINT hex_id round-trips to R exactly (not a lossy double)
con <- dbConnect(duckdb(hex_db), bigint = "integer64")

# 1) enumerate the ocean hex tiling (geometry only: hex_id + area_km2)
hex_ocean(r, con, res = hex_res, mask_layer = "depth_mean", out_tbl = "hex")

# 2) interpolate the 12 Bio-Oracle env layers onto the hex CENTROIDS by inverse-
#    distance weighting from the (ocean) pixel centroids. This is real
#    interpolation — NOT inheriting the value of the pixel that contains the hex
#    centre. Using ocean-only source points also means coastal hexes draw from the
#    nearest ocean pixels rather than going NA. (See msens R/hex.R note.)
pix <- as.data.frame(r, xy = TRUE, na.rm = FALSE)
pix <- pix[!is.na(pix$depth_mean), ]
names(pix)[match(c("x", "y"), names(pix))] <- c("lon", "lat")
hex_interp_idw(con, pix, "hex", val_cols = names(r), k = 8, power = 2)
```

## Membership: US study area + Program Areas

```{r membership}
# polyfill each polygon at res-7 to enumerate in-polygon hexes (scale-independent)
hex_add_membership(con, "hex", ply_usa, "in_usa", res = hex_res)
hex_add_membership(con, "hex", ply_pra, "in_pra", res = hex_res)
```

## Partition columns + spatial clustering

Add coarse H3 parent ids (`h3_p0` res-0, `h3_p2` res-2) as partition keys and
rewrite the table **ordered by `hex_id`** so row groups (and the marine-atlas
Parquet release) get tight, spatially-local hex ranges — enabling partition
pruning and row-group skipping for bbox/tile queries.

```{r partition_order}
dbExecute(con, glue("
  CREATE TABLE hex_ordered AS
  SELECT hex_id,
         CAST(h3_cell_to_parent(hex_id, 0) AS BIGINT) AS h3_p0,
         CAST(h3_cell_to_parent(hex_id, 2) AS BIGINT) AS h3_p2,
         * EXCLUDE (hex_id)
  FROM hex
  ORDER BY hex_id"))
dbExecute(con, "DROP TABLE hex")
dbExecute(con, "ALTER TABLE hex_ordered RENAME TO hex")
dbExecute(con, "ALTER TABLE hex ADD PRIMARY KEY (hex_id)")
```

## Summary + manifest

```{r summary}
smry <- dbGetQuery(con, "
  SELECT count(*) AS n_hex,
         count(*) FILTER (WHERE in_usa) AS n_usa,
         count(*) FILTER (WHERE in_pra) AS n_pra,
         count(DISTINCT h3_p0) AS n_part_p0,
         count(DISTINCT h3_p2) AS n_part_p2,
         round(min(area_km2), 3) AS area_min_km2,
         round(max(area_km2), 3) AS area_max_km2,
         count(*) FILTER (WHERE depth_mean IS NULL) AS n_null_depth
  FROM hex")
knitr::kable(smry)

manifest <- c(as.list(smry), list(
  version = ver, hex_res = hex_res, bbox = hex_bbox,
  db = hex_db, source = basename(bio_oracle_tif)))
write_json(manifest, here("data/manifests/build_hex_grid.json"),
           auto_unbox = TRUE, pretty = TRUE)

dbDisconnect(con, shutdown = TRUE)
message(glue("hex grid: {smry$n_hex} hexes ({smry$n_usa} in US, {smry$n_pra} in PRAs) -> {hex_db}"))
```
