---
title: "Build the zone-set registry (spatial units as vintages, not per-version copies)"
msens:
  target_name: build_zone_sets
  workflow_type: schema
  dependency: [build_cell_grid]
  output: data/manifests/build_zone_sets.json
editor_options:
  chunk_output_type: console
---

Every release so far baked its own copy of the spatial units — `ply_programareas_2026_v3.gpkg`
through `…_v8.gpkg`, each with its own `zone_cell` re-extracted by `score_zones.qmd`. That makes
the question users actually ask — *how did **this** Program Area's score change from v3 to v8?* —
unanswerable, because nothing guarantees two releases meant the same polygon.

This notebook identifies a spatial unit by its **geometry**, not by the release that used it:

```
zone_set_key = {zone_type}_{YYYY-MM}      e.g. programarea_2026-03
```

The vintage is the year-month the geometry **first appeared on disk**, so the label is a checkable
fact rather than an editorial guess, and keys sort chronologically.

> **The per-version files are mostly the same layer copied forward.** Measured across every
> published gpkg, `programarea` is 8 files but only **2** distinct geometries (v2 alone; v3–v8 are
> byte-identical), and `ecoregion` is 10 files but **2**. So `zone_cell` — which depends only on
> (geometry × grid) — is computed **once** per `(zone_set_key, grid_id)` and reused by every release
> on that grid, instead of being re-extracted per version.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "Fingerprint every zone layer, collapse identical ones into vintages, emit the registry"
flowchart LR
  gp["ply_*.gpkg<br/>(top level + every version dir)"] --> gh["msens::zone_geom_hash()<br/>order-invariant WKB fingerprint"]
  gh --> gr["group by (zone_type, geom_hash)<br/>vintage = earliest mtime YYYY-MM"]
  gr --> vz["msens::validate_zone_sets()<br/>one key ↔ one geometry"]
  vz --> csv["data/zone_sets.csv"]
```

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(dplyr, glue, knitr, readr, MarineSensitivity/msens, quiet = TRUE)
source(here::here("libs/paths.R"))

# Every zone layer anywhere under derived/ — the top level and every version dir.
# `ply_ecoregion_programarea_*` is a CROSSWALK (ecoregion x programarea slivers),
# not a spatial unit in its own right, and `ply_boem-usa` is the study-area
# outline; neither is a zone set.
gpkgs <- Sys.glob(c(path.expand(glue("{dir_derived}/ply_*.gpkg")),
                    path.expand(glue("{dir_derived}/*/ply_*.gpkg"))))
gpkgs <- grep("ecoregion_programarea|boem-usa", gpkgs, value = TRUE, invert = TRUE)

message(glue("scanning {length(gpkgs)} zone layer(s) under {dir_derived}"))
```

::: {.callout-warning}
This scan is only as complete as the machine it runs on. The **server** holds every version's
`derived/` tree; a laptop holds only the current version's. Render this where the whole tree is,
or the registry will silently omit vintages.
:::

## Fingerprint every layer

```{r}
#| label: fingerprint

zone_type_of <- function(f)
  dplyr::case_when(
    grepl("programarea", f) ~ "programarea",
    grepl("planarea",    f) ~ "planarea",
    grepl("ecoregion",   f) ~ "ecoregion",
    grepl("subregion",   f) ~ "subregion",
    .default = NA_character_)

scan <- lapply(gpkgs, function(f) {
  h <- msens::zone_geom_hash(f)
  tibble::tibble(
    source    = sub(paste0(path.expand(dir_derived), "/"), "", f),
    zone_type = zone_type_of(basename(f)),
    n_zones   = h$n,
    key_col   = h$key_col,
    geom_hash = h$geom_hash,
    mtime     = file.info(f)$mtime)
}) |> bind_rows() |> filter(!is.na(zone_type))

# which release used it: the version directory it sits in ("" = top level, i.e.
# a staging copy no release scored)
scan <- scan |> mutate(ver = ifelse(grepl("/", source), sub("/.*", "", source), NA_character_))

knitr::kable(scan |> select(source, zone_type, n_zones, ver, geom_hash))
```

## Collapse identical layers into vintages

```{r}
#| label: vintages

vint <- scan |>
  group_by(zone_type, geom_hash) |>
  summarise(n_zones = first(n_zones),
            vintage = format(min(mtime), "%Y-%m"),   # first appearance on disk
            n_files = n(),
            versions = paste(sort(na.omit(ver)), collapse = " "),
            source   = source[which.min(mtime)],
            .groups  = "drop") |>
  mutate(zone_set_key = msens::zone_set_key(zone_type, vintage)) |>
  arrange(zone_type, vintage)

# Two DISTINCT geometries of one type first appearing in the same month would
# collide on the key. Stop rather than silently merge two different maps.
dup <- vint$zone_set_key[duplicated(vint$zone_set_key)]
if (length(dup))
  stop(glue("zone_set_key collision ({paste(dup, collapse=', ')}): two distinct geometries ",
            "first appeared in the same month. Disambiguate the vintage by hand (e.g. -a/-b) ",
            "before publishing."))

msens::validate_zone_sets(vint)

knitr::kable(vint |> select(zone_set_key, n_zones, n_files, versions, geom_hash))
```

```{r}
#| label: summary

vint |>
  count(zone_type, name = "distinct_vintages") |>
  left_join(scan |> count(zone_type, name = "files"), by = "zone_type") |>
  relocate(files, .after = zone_type) |>
  knitr::kable(caption = "How many spatial units really exist, versus how many files")
```

## Write the registry

```{r}
#| label: write

f_csv <- here::here("data/zone_sets.csv")

# generated-but-committed, like a lockfile: deterministic given the files on
# disk, and reviewable in a diff when a new vintage appears
out <- vint |> select(zone_set_key, zone_type, vintage, n_zones, geom_hash, versions, source)
readr::write_csv(out, f_csv)

msens::validate_zone_sets(readr::read_csv(f_csv, show_col_types = FALSE))
message(glue("wrote {f_csv} ({nrow(out)} zone sets)"))
```

## Not yet covered: subregions

Subregions are **not** a published layer for most releases — they are synthesized in the database as
unions of that version's Planning/Program Areas, and they drifted accordingly:

| release | subregion keys |
|---|---|
| v1 | `AK`, `AKL48`, `L48`, `USA` |
| v2, v3 | *two different sets under one `fld`* — `AK/AKL48/L48/USA` **and** `AK/GA/PA/USA` |
| v4–v6 | `AK`, `GA`, `PA`, `USA` |
| v7 | `AK`, `FULL`, `GA`, `PA`, `USA` |
| v8 | 4 keys |

Because they were derived from each release's own Program Areas, a subregion is not comparable
across versions — and v2/v3 are ambiguous even within one database. The next step is a single
**canonical, version-independent** `subregion_usa` set spanning all US waters, registered here like
any other vintage, so every release shares one subregion list and the score-COG matrix
(metrics × subregions) has the same shape everywhere.

## Target manifest

```{r}
#| label: target-manifest

msens::write_manifest(
  here::here("data/manifests/build_zone_sets.json"),
  target       = "build_zone_sets",
  content_hash = digest::digest(out, algo = "xxhash64"),
  stats = list(
    n_files       = nrow(scan),
    n_zone_sets   = nrow(out),
    by_type       = paste(sprintf("%s=%d", vint$zone_type, 1), collapse = ""),
    programarea_n = sum(vint$zone_type == "programarea"),
    ecoregion_n   = sum(vint$zone_type == "ecoregion"),
    planarea_n    = sum(vint$zone_type == "planarea")),
  force = msens::force_target("build_zone_sets"))
```
