---
title: "Build the atlas `app/` bundle — the browser's data contract"
msens:
  target_name: build_app_bundle
  workflow_type: publish
  # after release_marine-atlas (needs the release's own sdm.duckdb / serve.duckdb
  # and, for a legacy generation, its published tables/model_asset.parquet), and
  # BEFORE build_version_manifest (which records the `app{}` block by probing
  # THIS notebook's push with an anonymous HTTPS HEAD, never by copying
  # local presence).
  dependency: [release_marine_atlas]
  output: data/manifests/build_app_bundle.json
params:
  ver: null
editor_options:
  chunk_output_type: console
---

Master plan D5 ("normalize at publish, not in the browser") and atlas-1 Deliverable 4. Both
Shiny apps carry a per-version adapter today — two grids, `mdl_seq` vs `mdl_key`, `value` vs
`val`, `is_ok` vs `is_valid_usa`, a `taxon_model` self-edge on one side only — each with a
production bug recorded behind it. `msens::app_bundle_build()` (msens 0.43.0, `R/app_bundle.R`)
already resolves every one of those quirks, with tests. This notebook is the thin publishing
shell around it: **build → validate → push (gated) → verify (gated)**, one version-independent
`{ver}/app/` bundle per release.

**Nothing here publishes anything.** The push and verification chunks are real, runnable code,
gated behind `APP_BUNDLE_S3=1` (unset by default) exactly like every other side-effecting step in
this repo (`../CLAUDE.md` reproducibility rule 1) — this notebook can be rendered top to bottom,
safely, with no S3 credentials and no network write of any kind.

**Rendering.** `quarto render build_app_bundle.qmd` — with or without an explicit `--output` —
always writes its HTML BESIDE THE SOURCE (`./build_app_bundle.html`), never into `_output/`
despite this project's `_quarto.yml` `output-dir: _output` (confirmed empirically 2026-09-21: a
project-wide render like `scripts/run_version.sh` respects it, a single ad hoc `quarto render` of
THIS file does not — a quirk of quarto's single-file vs. project render path, not this notebook).
The documented, ONE-TRUE command therefore always ends in a move — never render this file any
other way, and never leave the moved file's name as the version-less default:

```sh
APP_BUNDLE_VERS=v7 quarto render build_app_bundle.qmd && \
  mv -f build_app_bundle.html _output/build_app_bundle_v7.html && \
  rm -rf build_app_bundle_files
```

`scripts/render_app_bundle.sh <ver> [ver ...]` wraps exactly this (one version at a time,
sequentially, in the background — see its header for why: a foreground `quarto render` here runs
90–190s per legacy release and longer per v8/v9, and this notebook must never be waited on in a
chat turn) and is the only invocation that should be used from here on.

**A cache-invalidation rule from the atlas-2 review, stated here because a publish turn must know
it BEFORE it decides what to re-run**: re-publishing ANY object the app reads OUTSIDE `app/` —
`{ver}/tables/model.parquet`, `{ver}/serve/cell_model/**` — requires rebuilding and re-publishing
`app/` too, because the browser caches those objects keyed on `boot.built_at` (OPFS), not on the
object's own URL or ETag. A `tables/model.parquet` re-push with no matching `app/` re-push (a new
`built_at`) leaves every already-cached client reading stale `model.parquet` bytes under an
unchanged cache key — it will never refetch until something else bumps `built_at` for an unrelated
reason. There is no partial-republish path for these two: either both `app/` and the object move
together, in the same publish turn, or neither does.

`APP_BUNDLE_VERS=v7,v9` narrows a run to that set (comma-separated); unset, every version in
`data/versions.csv` is attempted (v1, v2, v3, v4, v4b, v5, v6, v7, v7b, v8, v9 — **not** v3b, which
is not a registered release), and a version whose release database is not on this machine is
**skipped, not failed** (the same convention `build_version_manifest.qmd`'s `caps_override`
uses: a capability this run cannot see is `FALSE`, never invented). Every entry is checked
against BOTH the version-label shape (`^v[0-9]+[a-z]?$`) AND membership in `data/versions.csv` —
round 3 (2026-09-21): shape alone let a shape-valid-but-unregistered label (e.g. `v3b`) silently
build a bundle for a release that does not exist.

`APP_BUNDLE_CELLMODEL_VERS=v7,v7b` (round 3: generalized from the single, v7-only
`APP_BUNDLE_V7_CELLMODEL` flag) additionally uploads the named version(s)' missing
`serve/cell_model/` — today v7 (757 MB locally, 428 tiles) and v7b (753 MB locally, 428 tiles),
both absent from S3 despite their manifest advertising `cell_species_list: true` (the atlas-1
"Facts" note). Default empty (nothing planned). Every entry is validated the same way as
`APP_BUNDLE_VERS` (shape + registry) and must also be one of THIS run's `vers`; each version's
plan is then checked — local tile-width (nc=3103) and one-file-per-tile gates, the exact key
shape `^{ver}/serve/cell_model/tile=[0-9]+/data_0\.parquet$`, and an anonymous HEAD refusing any
version whose tiles already exist on S3 — **before** any upload, in [validate-cell-model-plan]
below, which runs even when `APP_BUNDLE_S3` is unset (a pure dry run: it prints the exact key list
and byte total per version and writes nothing). v8 and v9 are never at risk even if
mis-listed: neither has a local `{ver}/cell_model/` export (their asset registry lives in the
database, not a filesystem directory), so the plan refuses them with "no local source directory"
before a single key is constructed.

Both `APP_BUNDLE_S3` and `APP_BUNDLE_V7_CELLMODEL`'s replacement are strict
([app_bundle_flag()]/allow-list validation): `APP_BUNDLE_S3` only exactly `1` turns it on — `0`,
`true`, `false`, `no`, anything else STOPS the render naming the variable, rather than guessing.

**Two things worth saying out loud so nobody "fixes" them.** First, a `restricted` release gets
**exactly the same objects** under `{ver}/app/` as a `public` one — the review gate that matters is
which HOST serves them (atlas-9: the public site never renders a restricted version; the preview
host does, behind Access), not what this notebook writes, so `build_app_bundle.qmd` never branches
on `access`. Second, even with `APP_BUNDLE_S3` and `APP_BUNDLE_V7_CELLMODEL` both unset — the
default, publish-nothing case — a legacy-generation dry run (v1–v7b) still downloads roughly
**165 MB anonymously** from the public bucket (`.attach_s3_legacy()`'s `tables/*.parquet` cache,
measured against v7, 2026-09-21: 13 tables, 164,894,811 bytes, dominated by `cell_metric` at 109
MB): reading nothing is not the same as touching nothing.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "One release database -> app_bundle_build() -> {ver}/app/, validated before it is ever written to S3"
flowchart LR
  subgraph src["this version's release database"]
    sdm[("sdm.duckdb<br/>(v8/v9: has model_asset->native_asset)")]
    srv[("serve.duckdb<br/>(v1-v7b: model_asset lives HERE,<br/>not in sdm.duckdb -- see Setup)")]
  end
  sdm -->|"sdm_db_path(ver)"| con(("read-only<br/>DuckDB connection"))
  srv -.->|"fallback when sdm.duckdb<br/>has no asset registry"| con
  con --> abb["msens::app_bundle_build()"]
  abb --> boot["boot.json"]
  abb --> taxa["taxa.json"]
  abb --> tsh["taxon/{xx}.json x256"]
  abb --> alsh["alias/{xx}.json x256"]
  abb --> txp["taxon.parquet"]
  abb --> ztp["zone_taxon.parquet"]
  abb --> cell["cell/tile={t}/data_0.parquet"]
  csv[("data/taxonomic_hierarchy_worms_…csv<br/>(curated, not a DB table)")] -->|"app_taxonomy()<br/>msens, taxonomy_csv="| txm["taxonomy.parquet"]
  abb --> mdl["model.parquet (v8+ only)"]
  boot & taxa & tsh & alsh & txp & ztp & cell & txm & mdl --> val{{"app_validate()<br/>every object vs its<br/>inst/schema/app_*.schema.json"}}
  val -->|"APP_BUNDLE_S3=1"| grd{{"app_bundle_assert_prefix()<br/>refuses any key outside<br/>{ver}/app/"}}
  grd --> s3[["s3://…/marine-atlas/{ver}/app/"]]
  s3 -->|"APP_BUNDLE_S3=1"| anon{{"anonymous HTTPS HEAD/GET<br/>per object"}}
  anon --> mfst["build_version_manifest.qmd<br/>probes the SAME way -><br/>manifest.json $app{}"]

  classDef gate fill:#3a1e1e,stroke:#e03131,color:#e6edf3;
  classDef db fill:#1e3a5f,stroke:#3bc9db,color:#e6edf3;
  class val,grd,anon gate;
  class sdm,srv,con db;
```

## Setup

```{r}
#| label: setup
#| message: false

librarian::shelf(
  DBI, dplyr, duckdb, fs, glue, here, jsonlite, knitr, logger, purrr, readr,
  MarineSensitivity/msens, quiet = TRUE)
source(here("libs/duckdb_budget.R"))
source(here("libs/app_bundle.R"))

stopifnot(
  "msens 0.43.0 or newer is required (app_bundle_build(), app_manifest_block())" =
    packageVersion("msens") >= "0.43.0",
  # round 4: 71aa231, then ba4b6d0, both landed WITHOUT a Version: bump (still
  # 0.43.0), so the version-string check above cannot tell either apart from
  # an older 0.43.0 build (e.g. 7c6eb3a) -- check the formal arguments this
  # notebook actually depends on existing, not just a string.
  "installed msens::app_bundle_build() has no `taxonomy_csv` argument -- reinstall from atlas-contract @ 71aa231 or later" =
    "taxonomy_csv" %in% names(formals(msens::app_bundle_build)),
  "installed msens::app_bundle_build() has no `zone_sets` argument -- reinstall from atlas-contract @ ba4b6d0 or later (the v2 zone-table fix)" =
    "zone_sets" %in% names(formals(msens::app_bundle_build)))
message(glue("msens {packageVersion('msens')} @ {find.package('msens')}"))

# --- run scope ------------------------------------------------------------
versions_csv <- readr::read_csv(here("data/versions.csv"), show_col_types = FALSE)
narrow       <- Sys.getenv("APP_BUNDLE_VERS")
vers         <- if (nzchar(narrow)) trimws(strsplit(narrow, ",")[[1]]) else versions_csv$ver
# review round 2: validate every entry BEFORE anything is built from it --
# same shape app_bundle_assert_prefix() requires of its own `ver`, and the
# same reason (a `ver` that isn't this shape could smuggle a path segment
# into every key this run ever constructs)
bad_vers <- vers[!vapply(vers, app_bundle_valid_ver, logical(1))]
if (length(bad_vers))
  stop(sprintf("APP_BUNDLE_VERS contains invalid version label(s): %s (must match ^v[0-9]+[a-z]?$)",
               paste(bad_vers, collapse = ", ")), call. = FALSE)
# round 3 item 2: shape alone is not enough -- v3b is shape-valid but is NOT a
# registered release (skipped per the owner's explicit instruction), and any
# other shape-valid-but-unregistered label would otherwise silently build a
# bundle for a release that doesn't exist in data/versions.csv
unreg_vers <- setdiff(vers, versions_csv$ver)
if (length(unreg_vers))
  stop(sprintf(
    "APP_BUNDLE_VERS contains version label(s) not registered in data/versions.csv: %s (registered: %s)",
    paste(unreg_vers, collapse = ", "), paste(versions_csv$ver, collapse = ", ")), call. = FALSE)

# strict: exactly "1" is on, unset/"" is off, anything else (0/true/false/no/...)
# STOPS the render naming the variable -- see libs/app_bundle.R::app_bundle_flag()
do_s3       <- app_bundle_flag("APP_BUNDLE_S3")          # push to S3 (default OFF)
base        <- msens::atlas_base_url()
dir_derived <- if (Sys.info()[["sysname"]] == "Linux") "/share/data/derived" else "~/_big/msens/derived"
dir_out_root <- here("_output", "app_bundle")                 # gitignored (see .gitignore)
fs::dir_create(dir_out_root)

# round 3 item 3: generalized from the single v7-only APP_BUNDLE_V7_CELLMODEL
# flag to an explicit allow-list of version labels -- validated exactly like
# APP_BUNDLE_VERS above (shape + registry), PLUS must be inside this run's own
# `vers` (uploading cell_model for a version this run did not even build makes
# no sense). Default empty: nothing is planned unless asked for.
narrow_cm   <- Sys.getenv("APP_BUNDLE_CELLMODEL_VERS")
cm_vers     <- if (nzchar(narrow_cm)) trimws(strsplit(narrow_cm, ",")[[1]]) else character(0)
bad_cm_vers <- cm_vers[!vapply(cm_vers, app_bundle_valid_ver, logical(1))]
if (length(bad_cm_vers))
  stop(sprintf(
    "APP_BUNDLE_CELLMODEL_VERS contains invalid version label(s): %s (must match ^v[0-9]+[a-z]?$)",
    paste(bad_cm_vers, collapse = ", ")), call. = FALSE)
unreg_cm_vers <- setdiff(cm_vers, versions_csv$ver)
if (length(unreg_cm_vers))
  stop(sprintf(
    "APP_BUNDLE_CELLMODEL_VERS contains version label(s) not registered in data/versions.csv: %s",
    paste(unreg_cm_vers, collapse = ", ")), call. = FALSE)
notinrun_cm_vers <- setdiff(cm_vers, vers)
if (length(notinrun_cm_vers))
  stop(sprintf(
    "APP_BUNDLE_CELLMODEL_VERS names version(s) outside this run's APP_BUNDLE_VERS scope (%s): %s",
    paste(vers, collapse = ", "), paste(notinrun_cm_vers, collapse = ", ")), call. = FALSE)

message(glue(
  "versions: {paste(vers, collapse = ', ')}  push_s3={do_s3}  ",
  "cell_model_vers={if (length(cm_vers)) paste(cm_vers, collapse = ',') else '(none)'}  base={base}"))
```

**Which database.** `msens::sdm_db_path(ver)` returns the release's build database, and prefers
`sdm.duckdb` over `serve.duckdb` whenever the former exists on this machine. That is right for
v8/v9, whose `native_asset` table lives IN `sdm.duckdb` — but wrong for v1–v7b: their local
`sdm.duckdb` has neither `model_asset` nor `native_asset` at all (verified 2026-09-21 against the
real v7 release: `model`, not `model_asset`, holds only dataset/taxon/region metadata, no COG
URL). `model_asset` is written only by `backfill_versions.qmd`, as a view in `serve.duckdb`. Opening
the wrong database is not an error — `app_bundle_build()` runs to completion and every taxon's
`merged` and `inputs[].assets` come back silently empty, which is exactly the failure mode this
contract exists to prevent everywhere else.

Three tiers, each PROBED (a view existing by name is not enough — `serve.duckdb`'s views embed
the LOCAL paths of the machine that built them, so a copy rsynced to a laptop lists a `model_asset`
view that then fails the moment it is queried; discovered running this notebook for real against
v7):

1. `sdm_db_path(ver)` (`sdm.duckdb`, or `serve.duckdb` if that is what `sdm_db_path()` itself
   falls back to) — right for v8/v9.
2. `serve.duckdb` beside it, if (1) has no working asset table — right on the SERVER that built it.
3. An ephemeral, **anonymous, read-only** DuckDB attached directly to this release's *published*
   `tables/*.parquet` on S3 (httpfs, path-style; no credentials, no write) — right anywhere,
   including this laptop, for any `released`/public version. Never used for a `restricted` version
   (nothing to attach yet, or nothing this notebook should read anonymously either).

```{r}
#| label: db-open

# does `tbl` exist AND actually return rows when queried (not just listed)?
.probe_table <- function(con, tbl) {
  if (!tbl %in% DBI::dbListTables(con)) return(FALSE)
  isTRUE(tryCatch(DBI::dbGetQuery(con, glue("SELECT count(*) n FROM {tbl} LIMIT 1"))$n >= 0,
                  error = function(e) FALSE))
}
.has_asset_tbl <- function(con) any(vapply(c("model_asset", "native_asset"), .probe_table, logical(1), con = con))

# tier 3: views over a LOCAL CACHE of the PUBLISHED tables/*.parquet -- downloaded
# once, anonymously, read-only (no S3 write, no credentials). attach_atlas() does
# the remote-view version of this for v8+'s native_asset shape, but has no
# model_asset/taxon_model view -- and a REMOTE view here was measured to be
# impractically slow: a full aggregation over `cell_metric` (unprunable row
# groups) via anonymous httpfs range reads did not finish in over an hour, where
# the equivalent LOCAL query takes seconds. `tables/` for a whole legacy release
# is under 200 MB (atlas-1 plan "Facts": v7 cell 22.7 MB + cell_metric 109 MB +
# the rest under 25 MB combined), so caching it locally costs a one-time download,
# not a redesign.
.attach_s3_legacy <- function(ver, access) {
  if (!identical(access, "public")) return(NULL)
  cache_dir <- here("_output", "app_bundle", "_s3cache", ver, "tables")
  fs::dir_create(cache_dir)
  http <- sprintf("%s/%s", base, ver)
  tbls <- c("cell", "taxon", "model", "model_asset", "metric", "cell_metric",
           "zone", "zone_cell", "zone_metric", "zone_taxon", "dataset", "taxon_model", "listing")
  got <- vapply(tbls, function(t) {
    dest <- file.path(cache_dir, paste0(t, ".parquet"))
    if (file.exists(dest)) return(TRUE)
    isTRUE(tryCatch(
      utils::download.file(sprintf("%s/tables/%s.parquet", http, t), dest,
                           mode = "wb", quiet = TRUE) == 0L,
      error = function(e) FALSE, warning = function(w) FALSE))
  }, logical(1))
  if (!any(got)) return(NULL)
  con <- DBI::dbConnect(duckdb::duckdb())
  ok <- vapply(tbls[got], function(t) {
    isTRUE(tryCatch({
      DBI::dbExecute(con, glue(
        "CREATE OR REPLACE VIEW {t} AS SELECT * FROM read_parquet('{cache_dir}/{t}.parquet')"))
      .probe_table(con, t)
    }, error = function(e) FALSE))
  }, logical(1))
  if (!.has_asset_tbl(con)) { DBI::dbDisconnect(con, shutdown = TRUE); return(NULL) }
  log_warn("{ver}: cached + attached {sum(ok)}/{length(tbls)} tables from {http}/tables/ ({cache_dir})")
  con
}

app_bundle_open <- function(ver) {
  v_row  <- versions_csv |> filter(ver == !!ver)
  access <- if (nrow(v_row)) v_row$access[1] else "restricted"
  db  <- path.expand(msens::sdm_db_path(ver))
  if (file.exists(db)) {
    con <- DBI::dbConnect(duckdb::duckdb(), db, read_only = TRUE)
    if (.has_asset_tbl(con)) return(list(con = con, db = db, ok = TRUE))
    DBI::dbDisconnect(con, shutdown = TRUE)
  }
  alt <- file.path(fs::path_dir(db), "serve.duckdb")
  if (fs::path_file(db) != "serve.duckdb" && file.exists(alt)) {
    con2 <- DBI::dbConnect(duckdb::duckdb(), alt, read_only = TRUE)
    if (.has_asset_tbl(con2)) {
      log_warn("{ver}: sdm.duckdb has no working model_asset/native_asset; using serve.duckdb ({alt})")
      return(list(con = con2, db = alt, ok = TRUE))
    }
    DBI::dbDisconnect(con2, shutdown = TRUE)
  }
  con3 <- .attach_s3_legacy(ver, access)
  if (!is.null(con3)) return(list(con = con3, db = sprintf("s3 (anonymous, tables/) %s/%s", base, ver), ok = TRUE))
  if (!file.exists(db)) return(list(con = NULL, db = db, ok = FALSE))
  log_warn(paste(
    "{ver}: no database anywhere (local sdm.duckdb, local serve.duckdb, S3 tables/) carries",
    "model_asset/native_asset -- merged COG urls and input assets will be EMPTY (not an error)"))
  list(con = DBI::dbConnect(duckdb::duckdb(), db, read_only = TRUE), db = db, ok = TRUE)
}

# the gates below only need cell/cell_metric/zone*/taxon/metric -- always present
# in sdm.duckdb (or its serve.duckdb fallback) for every generation -- so they
# open directly rather than paying app_bundle_open()'s S3-probe escalation,
# which exists only for the asset registry (model_asset/native_asset).
app_bundle_open_scoring <- function(ver)
  DBI::dbConnect(duckdb::duckdb(), path.expand(msens::sdm_db_path(ver)), read_only = TRUE)
```

## Guard: the push may never write outside `{ver}/app/`, and a publish flag is never guessed at

Run once, at render time, so the render itself is the proof — not just a definition nobody
exercised. `app_bundle_assert_prefix()`, `app_bundle_flag()`, `app_bundle_assert_cell_model_keys()`
and `app_bundle_geom_keys()`'s own selftest are pure/offline (no S3, no side effect);
`app_bundle_head_check()`'s selftest is the one exception — it makes two real anonymous HTTPS HEAD
requests (one known-200, one known-non-200) to prove the URL/status pairing is never misaligned by
`xargs -P`'s out-of-order completion, which is exactly the hazard this function exists to close.

```{r}
#| label: guard-selftest
app_bundle_assert_prefix_selftest()
app_bundle_flag_selftest()
app_bundle_gzip_selftest()
app_bundle_geom_keys_selftest()
app_bundle_cell_model_selftest()
app_bundle_head_check_selftest()
app_bundle_head_check_retry_selftest()
app_bundle_zone_taxon_check_selftest()
app_bundle_d17_check_selftest()
```

## Build

**One `msens::app_bundle_build()` call per version** — review round 1 (2026-09-21): the prior
version of this notebook composed the bundle itself (`app_boot(..., tables = list())`,
`manifest_build()` with no `zone_tiles`), which is why `boot$tables` (the OPFS invalidation key)
and `boot$units` (the drawable units) came out EMPTY on both real dry runs. `app_bundle_build()`
is the one function that wires `app_table_descriptor()`/`app_table_manifest()` into `app_boot()`
correctly; calling it directly is what makes `tables` and `units` non-empty, not a bigger
`tryCatch`. `geom_keys` — the published geometry keys `app_units()` needs to exclude
whole-study-area rollups (`USA`/`FULL`) — comes from [app_bundle_geom_keys()], which reads the
zone-set registry's own source GeoPackage rather than guessing a filename.

`app/taxonomy.parquet` and `app/model.parquet` (round 4, msens 0.43.0 @ `71aa231`): both moved INTO
`app_bundle_build()` itself — `taxonomy_csv=` feeds `app_taxonomy()` (still a curated CSV join, not
a database table, but now one of the builder's own stages, with a `boot$tables` digest like every
other object) and `app_model()` writes the v8+ `mdl_id -> mdl_key` mapping unconditionally,
returning `NULL` (not a failure) for v1-v7b, which have no such mapping to give. This notebook no
longer composes either writer itself, and `app_bundle_build()` now asserts
`names(boot$tables)` equals the Parquet objects actually on disk (`msens::app_tables_match()`,
called inside the `boot.json` stage) — the exact drift `app/taxonomy.parquet` and `model.parquet`
used to be invisible to (OPFS could never invalidate an object with no digest).

`app_bundle_build()` now isolates its own internal stages (msens 0.43.0 @ `7c6eb3a`, landed
during review round 1): `strict = TRUE` (its default) still stops the WHOLE build at the first
failing stage — right for an actual publish, wrong for this notebook's diagnostic pass — so this
notebook always passes `strict = FALSE`, matching the function's own documented "notebook's
diagnostic pass" use case: every stage that CAN complete does, `out$failed` names whichever
didn't, and the outer `tryCatch` below is only a safety net for anything upstream of `stage()`
(e.g. `manifest_build()` itself).

```{r}
#| label: build

build_one <- function(ver) {
  t0 <- Sys.time()
  opened <- app_bundle_open(ver)
  if (!isTRUE(opened$ok)) {
    log_warn("{ver}: no local release database at {opened$db} -- SKIPPED")
    return(list(ver = ver, ok = FALSE, reason = "no local database"))
  }
  con <- opened$con
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  duckdb_tune(con, file.path(tempdir(), "app_bundle_spill", ver))

  dir_out <- file.path(dir_out_root, ver)
  if (fs::dir_exists(dir_out)) fs::dir_delete(dir_out)   # clean slate every run
  fs::dir_create(dir_out)

  v_row <- versions_csv |> filter(ver == !!ver)
  status <- if (nrow(v_row)) v_row$status[1] else "prerelease"
  access <- if (nrow(v_row)) v_row$access[1] else "restricted"

  f_scog  <- file.path(fs::path_dir(opened$db), "tables", "score_cog.parquet")
  metrics <- if (file.exists(f_scog)) {
    arrow::read_parquet(f_scog) |> filter(!startsWith(metric_key, "_")) |>
      transmute(metric_key, subregion_key, cog = cog_url, rescale_min, rescale_max, colormap)
  } else NULL

  zone_sets <- readr::read_csv(here("data/zone_sets.csv"), show_col_types = FALSE)
  # round-2 fix: manifest_build()'s zone_tiles was never passed, so every
  # zone's pmtiles stayed NA and app_units() -- which requires a non-NA
  # pmtiles -- returned ZERO units regardless of geom_keys. Same construction
  # build_version_manifest.qmd uses.
  zone_tiles <- setNames(
    as.list(glue("{base}/zones/{zone_sets$zone_set_key}/zones.pmtiles")),
    zone_sets$zone_set_key)
  mfst <- msens::manifest_build(con, ver, status = status, access = access, base = base,
                                metrics = metrics, zone_tiles = zone_tiles, zone_sets = zone_sets)
  # round 10 (2026-09-22): app_bundle_build() (and this notebook's own
  # geom_keys, below) must be handed the PUBLISHED manifest.json, not the
  # one just rebuilt above. manifest_build()'s one-row-per-zone_set_key
  # collapse breaks a genuine tie (v2's subregion_key: two zone tables,
  # n = 4 each) by DuckDB's own row order -- non-deterministic between an
  # installed-package build and a source-tree run of the IDENTICAL commit.
  # `mfst` (above) is kept only for the app{} manifest-patch proposal
  # (PROPOSAL.md) and the id_field cross-check below, neither of which
  # touches the ambiguous zone-table resolution. A missing/unreadable
  # published manifest is a hard stop: a release with no manifest cannot
  # have a bundle at all.
  pub_manifest <- app_bundle_fetch_published_manifest(
    ver, base, here("_output", "app_bundle", "_s3cache", ver))
  stopifnot(
    "could not fetch/parse the published {ver}/manifest.json (round-10 hard-fail)" =
      !is.null(pub_manifest) && !is.null(pub_manifest$zones) && nrow(pub_manifest$zones) > 0)

  # master plan D17 (Ben's decision, 2026-09-22): drawable units are Program
  # Areas ONLY (v2-v9), Planning Areas on v1 -- no subregion or ecoregion
  # unit on ANY release. This supersedes D16 entirely (its per-zone-type
  # derivation, the subregion cut/omit logic, the v8 AT warning): there is
  # exactly ONE unit type per release now, so ONLY that one geometry is ever
  # resolved or handed to app_bundle_build() -- not "every zone type the
  # manifest names," which is what let ecoregion/subregion geometry checks
  # (and their false mismatches) into the build at all.
  d17_type <- if (ver == "v1") "planarea" else "programarea"
  d17_fld  <- paste0(d17_type, "_key")
  zones_for_unit <- pub_manifest$zones[pub_manifest$zones$fld == d17_fld, , drop = FALSE]
  stopifnot(
    "D17: this release's manifest has no row for its own drawable unit's field" =
      nrow(zones_for_unit) == 1)
  # round 3: a missing/unreadable GeoPackage is now a HARD FAILURE inside
  # app_bundle_geom_keys() itself (it stops, naming the path).
  geom_keys <- app_bundle_geom_keys(zones_for_unit, zone_sets, dir_derived)
  stopifnot(
    "D17: exactly the release's ONE drawable unit's geometry must be passed -- no more, no fewer" =
      length(geom_keys) == 1 && identical(names(geom_keys), d17_type))

  # round 4: msens 0.43.0 @ 71aa231 moved app/taxonomy.parquet INTO
  # app_bundle_build() itself (as app_taxonomy(), fed by taxonomy_csv) and
  # added app_model() for the v8+ mdl_id -> mdl_key mapping -- both now get
  # boot$tables entries (with digests) the same way every other object does,
  # and app_bundle_build() already asserts names(boot$tables) == the Parquet
  # objects on disk (msens::app_tables_match(), called inside the boot.json
  # stage) -- so this notebook's own taxonomy writer (app_taxonomy_table(),
  # removed from libs/app_bundle.R) and its separate ok-flag are both gone;
  # a taxonomy-stage failure is caught the same way every other stage is,
  # via out$failed below.
  # round 4 (msens 0.43.0 @ ba4b6d0, round 5-7 fix): a zone_fld with more than
  # one source zone table (v2's subregion_key: ply_subregions_2025 AND
  # ply_subregions_2026) is resolved ONCE, inside app_bundle_build(), via
  # app_zone_tbl(con, manifest, geom_keys, zone_sets) -- manifest$zones[].tbl
  # is the deciding evidence (already present: manifest_build() has always
  # queried `tbl` alongside `fld`), geom_keys is the agreement check, and
  # zone_sets is a cross-check recorded in `why`. zone_sets MUST be passed
  # here too now, not only to manifest_build() above, or app_zone_tbl() loses
  # its cross-check silently (still resolves, just with one fewer thing
  # checked). app_zones_unique() (msens, a builder-level hard stop) then
  # asserts (zone_fld, zone_value, key) unique in zone_taxon and each key
  # once per boot$zones unit -- the exact 13,077-row v2 defect item (c)'s own
  # independent gate caught on the PRE-fix build.
  out <- tryCatch(
    msens::app_bundle_build(con, ver, dir_out, manifest = pub_manifest, base = base,
                            geom_keys = geom_keys, zone_sets = zone_sets, strict = FALSE,
                            taxonomy_csv = here("data/taxonomic_hierarchy_worms_2025-10-30.csv")),
    error = function(e) {
      log_warn("{ver}: app_bundle_build() FAILED (upstream of stage isolation): {substr(conditionMessage(e), 1, 300)}")
      NULL
    })
  if (!is.null(out) && length(out$failed))
    for (nm in names(out$failed))
      log_warn("{ver}: stage '{nm}' FAILED: {substr(out$failed[[nm]], 1, 300)}")

  # every drawable unit's keys must be a subset of >= 2 of its REAL geometry
  # (the assertion the original bug -- 5 subregion keys where only 4 are
  # drawable -- would have failed had it existed)
  geom_check <- if (!is.null(out) && !is.null(out$boot))
    app_bundle_assert_units_match_geometry(out$boot$units, geom_keys) else tibble()

  list(ver = ver, ok = !is.null(out), dir_out = dir_out, con_db = opened$db, manifest = mfst,
       pub_manifest = pub_manifest, d17_type = d17_type,
       geom_keys = geom_keys, geom_check = geom_check, out = out,
       has_taxonomy = !is.null(out) && "taxonomy" %in% names(out$tables),
       has_model    = !is.null(out) && "model"    %in% names(out$tables),
       n_failed_stages = if (is.null(out)) NA_integer_ else length(out$failed),
       wall_s = as.numeric(Sys.time() - t0, units = "secs"))
}

builds <- purrr::map(vers, build_one) |> setNames(vers)
build_tbl <- bind_rows(lapply(builds, function(b) data.frame(
  ver = b$ver, ok = b$ok, n_taxa = if (!is.null(b$out)) b$out$n_taxa else NA,
  n_tables = if (!is.null(b$out)) length(b$out$tables) else NA,
  n_units = if (!is.null(b$out)) length(b$out$boot$units) else NA,
  n_failed_stages = b$n_failed_stages, failed_stage_names = if (!is.null(b$out) && length(b$out$failed))
    paste(names(b$out$failed), collapse = ", ") else "",
  has_taxonomy = b$has_taxonomy, has_model = b$has_model, wall_s = round(b$wall_s, 1))))
knitr::kable(build_tbl, caption = "Per-version build result (this run)")
stopifnot(
  "app_bundle_build() aborted upstream of stage isolation for at least one version" =
    all(build_tbl$ok),
  "at least one app_bundle_build() STAGE failed" = all(build_tbl$n_failed_stages == 0, na.rm = TRUE),
  # every release's taxa are a slice of the SAME curated CSV -- taxonomy.parquet
  # must exist for every release, unlike model.parquet (has_model), which is
  # TRUE only for v8+ (app_model() correctly returns NULL for v1-v7b, no
  # mdl_id/mdl_key mapping to give -- not a failure, a capability)
  "taxonomy.parquet is missing for at least one version" = all(build_tbl$has_taxonomy))
```

### Every drawable unit's keys are a subset of >= 2 keys in its REAL geometry

Round 3's fix: `geom_keys` used to resolve empty (message + continue) when a zone GeoPackage was
missing, so `app_units()`'s own `>= 2 keys in the geometry` check was silently skipped for every
type — v9's `boot.json` listed **5** subregion keys where only **`AK`, `AT`, `GA`, `PA`** (4) are
actually drawable. `app_bundle_geom_keys()` now hard-fails on a missing/unreadable GeoPackage
(seeded fault verified 2026-09-21: renaming `v1/ply_ecoregions_2025.gpkg` away made this render
STOP with `app_bundle_geom_keys(): missing GeoPackage for zone_set_key 'ecoregion_2025-06'
(ecoregion): .../v1/ply_ecoregions_2025.gpkg does not exist`; file restored before this render).

```{r}
#| label: validate-geom-keys

geom_check_tbl <- bind_rows(lapply(builds, function(b) {
  if (!nrow(b$geom_check)) return(tibble())
  cbind(ver = b$ver, b$geom_check)
}))
knitr::kable(geom_check_tbl,
             caption = "Drawable units vs their real GeoPackage keys (keys_ge_2 and all_keys_in_geometry must be TRUE)")

# the literal keys per unit, for the subregion case specifically named in the review
unit_keys_tbl <- bind_rows(lapply(builds, function(b) {
  if (is.null(b$out) || is.null(b$out$boot)) return(tibble())
  bind_rows(lapply(b$out$boot$units, function(u)
    data.frame(ver = b$ver, zone_type = u$zone_type, n_keys = length(u$keys),
              keys = paste(sort(unlist(u$keys)), collapse = ","))))
}))
knitr::kable(unit_keys_tbl, caption = "Every drawable unit's actual keys, per version")
stopifnot(
  "a drawable unit failed the geometry-consistency check" =
    nrow(geom_check_tbl) == 0 || (all(geom_check_tbl$keys_ge_2) && all(geom_check_tbl$all_keys_in_geometry)))

# master plan D17 (Ben's decision, 2026-09-22, supersedes D16 in full):
# drawable units are Program Areas ONLY (v2-v9), Planning Areas on v1 -- no
# subregion or ecoregion unit on any release. Exactly ONE unit per release,
# with an EXACT key count (v1 planarea=36; v2-v9 programarea=20) --
# app_bundle_d17_check() asserts it, never derives or hand-cuts it. Every
# release still gets a ROW here regardless of pass/fail, per "never silent."
d17_gate_tbl <- bind_rows(lapply(builds, function(b) {
  if (!isTRUE(b$ok) || is.null(b$out$boot)) return(tibble())
  n_keys_expected <- length(b$geom_keys[[b$d17_type]])
  chk <- app_bundle_d17_check(b$d17_type, n_keys_expected, b$out$boot$units)
  data.frame(ver = b$ver, expected_zone_type = b$d17_type, n_keys_expected = n_keys_expected,
             n_units_actual = chk$n_units_actual, n_keys_actual = chk$n_keys_actual,
             matches_expectation = chk$matches, reason = chk$reason)
}))
knitr::kable(d17_gate_tbl, caption = paste(
  "D17: exactly ONE drawable unit per release (programarea v2-v9, planarea v1),",
  "with the exact published key count -- matches_expectation must be TRUE"))
stopifnot(
  "D17: a release does not have exactly one drawable unit with the exact expected key count" =
    nrow(d17_gate_tbl) == 0 || all(d17_gate_tbl$matches_expectation))
```

## Validate

`app_bundle_build()`'s builders already validate their own output as they write it
(`app_validate()`, `R/app_bundle.R`); this chunk re-validates the WRITTEN BYTES, round-tripped
through disk exactly as an app would fetch them — the same discipline
`build_version_manifest.qmd` applies to `manifest.json` — and proves the schemas actually reject
a broken shape, not just accept a good one.

```{r}
#| label: validate-schema

revalidate_written <- function(b) {
  if (!isTRUE(b$ok)) return(tibble())
  files <- list(
    boot  = list(path = file.path(b$dir_out, "boot.json"),  schema = "boot"),
    taxa  = list(path = file.path(b$dir_out, "taxa.json"),  schema = "taxa"))
  rows <- purrr::imap(files, function(f, nm) {
    if (!file.exists(f$path)) return(data.frame(ver = b$ver, object = nm, ok = NA, note = "not built"))
    x <- jsonlite::fromJSON(f$path, simplifyVector = FALSE)
    ok <- tryCatch({ msens::app_validate(x, f$schema); TRUE },
                    error = function(e) { message(conditionMessage(e)); FALSE })
    data.frame(ver = b$ver, object = nm, ok = ok, note = "")
  })
  for (kind in c("taxon", "alias")) {
    d <- file.path(b$dir_out, kind)
    if (!dir.exists(d)) next
    fs_ <- list.files(d, full.names = TRUE)
    bad <- 0L
    for (f in fs_) {
      x <- jsonlite::fromJSON(f, simplifyVector = FALSE)
      ok <- tryCatch({ msens::app_validate(x, kind); TRUE }, error = function(e) FALSE)
      if (!ok) bad <- bad + 1L
    }
    rows[[kind]] <- data.frame(ver = b$ver, object = sprintf("%s/*.json (n=%d)", kind, length(fs_)),
                               ok = bad == 0L, note = if (bad) sprintf("%d shard(s) FAILED", bad) else "")
  }
  bind_rows(rows)
}
val_tbl <- bind_rows(purrr::map(builds, revalidate_written))
knitr::kable(val_tbl, caption = "Schema re-validation of the WRITTEN bytes, per version")
```

### `boot.json` actually carries its OPFS keys and drawable units

Review round 1's root finding: composing the bundle by hand (rather than calling
`app_bundle_build()`) left `boot$tables` and `boot$units` EMPTY on both real dry runs. `boot.json`
now carries its own `id_field` (msens `7c6eb3a`), so this gate cross-checks it against the
manifest's directly, rather than only checking the manifest's in isolation (round 1's stopgap,
removed).

```{r}
#| label: validate-boot-contract

boot_contract_gate <- function(b) {
  if (!isTRUE(b$ok) || is.null(b$out)) return(tibble())
  boot   <- b$out$boot
  tables <- b$out$tables      # named list of app_table_descriptor() results
  digest_ok <- vapply(tables, function(d) {
    got <- as.character(msens::parquet_digest(file.path(b$dir_out, d$path))$digest[1])
    identical(got, d$digest)
  }, logical(1))
  data.frame(
    ver = b$ver,
    n_tables            = length(tables),
    tables_ge_3         = length(tables) >= 3,
    all_digests_match   = length(digest_ok) > 0 && all(digest_ok),
    n_digest_mismatches = sum(!digest_ok),
    n_units             = length(boot$units),
    units_ge_1          = length(boot$units) >= 1,
    boot_id_field       = boot$id_field %||% NA_character_,
    manifest_id_field   = b$manifest$id_field %||% NA_character_,
    id_field_matches    = !is.null(boot$id_field) && !is.null(b$manifest$id_field) &&
      identical(boot$id_field, b$manifest$id_field) &&
      boot$id_field %in% c("mdl_key", "mdl_seq"))
}
`%||%` <- function(a, b) if (is.null(a)) b else a
boot_contract_tbl <- bind_rows(purrr::map(builds, boot_contract_gate))
knitr::kable(boot_contract_tbl,
             caption = "boot.json contract: tables (OPFS digests), units (drawable), id_field")
stopifnot(
  "boot$tables has fewer than 3 entries"          = nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$tables_ge_3),
  "a table descriptor's digest disagrees with the written file's" =
    nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$all_digests_match),
  "boot$units has fewer than 1 entry"             = nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$units_ge_1),
  "boot$id_field does not equal the manifest's"   = nrow(boot_contract_tbl) == 0 || all(boot_contract_tbl$id_field_matches))
```

```{r}
#| label: validate-malformed-fixture
#| eval: !expr any(purrr::map_lgl(builds, "ok"))

# a deliberately malformed fixture must fail -- proves the two checks above are
# not vacuously green (msens's own test-app_bundle.R exercises the same five
# malformations against the builder output; this repeats it against ONE real
# written boot.json so the notebook, not just the package, is proven to enforce it)
good <- jsonlite::fromJSON(
  file.path(builds[[which(purrr::map_lgl(builds, "ok"))[1]]]$dir_out, "boot.json"),
  simplifyVector = FALSE)
bad_cases <- list(
  "required key removed"    = { x <- good; x$grid <- NULL; x },
  "wrong type (nc as string)" = { x <- good; x$grid$nc <- "7200"; x },
  "undeclared key"          = { x <- good; x$surprise <- 1; x },
  "out-of-range enum"       = { x <- good; x$release$access <- "semi-public"; x })
mal <- purrr::imap(bad_cases, function(x, nm) {
  ok <- tryCatch({ msens::app_validate(x, "boot"); TRUE }, error = function(e) FALSE)
  data.frame(case = nm, rejected = !ok)
}) |> bind_rows()
knitr::kable(mal, caption = "A deliberately malformed boot.json must be REJECTED (rejected = TRUE)")
stopifnot("schema validation let a malformed boot.json through" = all(mal$rejected))
```

### Exactly one file per tile directory

Review round 1 found 290 of v9's 422 `cell/tile=*/` directories holding SEVERAL part files
(`data_0.parquet`, `data_1.parquet`, ...) — an msens `COPY ... PARTITION_BY` bug under
parallelism. The browser only ever fetches the one fixed filename `tile={t}/data_0.parquet`
(never a glob — anonymous LIST is denied), so a tile with more than one file was missing data
from every part after the first, silently. **Fixed in msens 0.43.0 @ `7c6eb3a`** (review round 2):
this gate now FAILS the render on any tile directory with anything other than exactly one file.

```{r}
#| label: validate-one-file-per-tile

one_file_per_tile_gate <- function(b) {
  if (!isTRUE(b$ok) || is.null(b$out$tiles)) {
    app_bundle_gate_note("one_file_per_tile", b$ver, FALSE, if (!isTRUE(b$ok))
      "build did not complete" else "release has no cell table (no tiles were built)")
    return(tibble())
  }
  app_bundle_gate_note("one_file_per_tile", b$ver, TRUE)
  d <- file.path(b$dir_out, "cell")
  tiledirs <- list.dirs(d, full.names = TRUE, recursive = FALSE)
  n_files <- vapply(tiledirs, function(td) length(list.files(td, pattern = "\\.parquet$")), integer(1))
  data.frame(ver = b$ver, n_tile_dirs = length(tiledirs),
             n_with_exactly_1_file = sum(n_files == 1L),
             n_with_multiple_files = sum(n_files > 1L),
             n_with_zero_files     = sum(n_files == 0L),
             worst_example = if (any(n_files > 1L))
               basename(tiledirs[which(n_files > 1L)[1]]) else NA_character_,
             worst_example_n_files = if (any(n_files > 1L)) max(n_files) else NA_integer_)
}
one_file_tbl <- bind_rows(purrr::map(builds, one_file_per_tile_gate))
knitr::kable(one_file_tbl, caption = "cell/tile=*/ directories: exactly one file per tile")
stopifnot(
  "a tile directory has zero or multiple .parquet files" =
    nrow(one_file_tbl) == 0 ||
    all(one_file_tbl$n_with_multiple_files == 0 & one_file_tbl$n_with_zero_files == 0))
```

### Tile-width gate (real nc, then the wrong one)

```{r}
#| label: validate-tile-width

tile_width_gate <- function(b) {
  if (!isTRUE(b$ok) || is.null(b$out$tiles)) {
    app_bundle_gate_note("tile_width", b$ver, FALSE, if (!isTRUE(b$ok))
      "build did not complete" else "release has no cell table (no tiles were built)")
    return(tibble())
  }
  app_bundle_gate_note("tile_width", b$ver, TRUE)
  ncol_real <- b$out$tiles$ncol
  ncol_wrong <- if (ncol_real == 7200L) 3103L else 7200L
  ok_real  <- tryCatch({ msens::app_cell_tile_check(file.path(b$dir_out, "cell"), ncol_real); TRUE },
                       error = function(e) FALSE)
  bad_wrong <- tryCatch({ msens::app_cell_tile_check(file.path(b$dir_out, "cell"), ncol_wrong); FALSE },
                        error = function(e) TRUE)
  data.frame(ver = b$ver, ncol_real = ncol_real, real_passes = ok_real,
             ncol_wrong = ncol_wrong, wrong_fails_as_expected = bad_wrong)
}
tw_tbl <- bind_rows(purrr::map(builds, tile_width_gate))
knitr::kable(tw_tbl, caption = "Tile-width gate: PASS at the release's own nc, FAIL at the wrong one")
stopifnot(
  "tile-width gate did not pass at the real nc"    = all(tw_tbl$real_passes),
  "tile-width gate did not fail at the wrong nc"   = all(tw_tbl$wrong_fails_as_expected))
```

### Wide-tile digest gate (per-metric multiset digest vs `cell_metric`, read the way the browser does)

`msens::app_cell_tile_digests()` globs every `*.parquet` under `cell/`; the browser fetches only
`tile={t}/data_0.parquet`. Read the SAME single file per tile here — a tile with extra part files
(see above) then correctly shows a digest MISMATCH, because the rows in `data_1.parquet` etc. are
invisible to this query exactly as they are to the app.

```{r}
#| label: validate-digest

digest_gate <- function(b, ver) {
  if (!isTRUE(b$ok) || is.null(b$out$tiles)) {
    app_bundle_gate_note("digest", ver, FALSE, if (!isTRUE(b$ok))
      "build did not complete" else
      "release has no cell table (no tiles were built, so no per-metric digest to check)")
    return(tibble())
  }
  app_bundle_gate_note("digest", ver, TRUE)
  con <- app_bundle_open_scoring(ver)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  duckdb_tune(con, file.path(tempdir(), "app_bundle_spill_digest", ver))
  keys <- b$out$tiles$metric_keys
  vc   <- msens::sdm_val_col(con, "cell_metric")
  glob <- file.path(b$dir_out, "cell", "tile=*", "data_0.parquet")   # browser-equivalent, not **/*.parquet
  bind_rows(lapply(keys, function(k) {
    kq <- DBI::dbQuoteIdentifier(con, k)
    a <- msens::hash_query(con, glue::glue(
      "SELECT cell_id, {kq} AS val FROM read_parquet('{glob}', hive_partitioning = true)
        WHERE {kq} IS NOT NULL"), cols = c("cell_id", "val"))
    # round 4 (found rendering v1 for real, 2026-09-21): v1's cell_metric has 703
    # rows whose cell_id is NOT in `cell` at all (an orphan, pre-existing in the
    # legacy database, nothing to do with this notebook) -- app_cell_tiles()
    # correctly never writes them (its own query JOINs against `cell`), so the
    # ground-truth query here must scope to the SAME cells or every legacy
    # release with an orphan cell_metric row shows a false digest mismatch. A
    # cell_metric row for a cell_id the bundle could never contain is out of
    # scope for "does the published bundle disagree with the database" -- it
    # was never a candidate to publish in the first place.
    bb <- msens::hash_query(con, glue::glue(
      "SELECT cm.cell_id, cm.{vc} AS val FROM cell_metric cm JOIN metric m USING (metric_seq)
        JOIN cell c USING (cell_id)
        WHERE m.metric_key = {DBI::dbQuoteString(con, k)} AND cm.{vc} IS NOT NULL"),
      cols = c("cell_id", "val"))
    data.frame(ver = ver, metric_key = k, tiles = a, cell_metric = bb, ok = identical(a, bb))
  }))
}
dg_tbl <- purrr::imap(builds, digest_gate) |> bind_rows()
knitr::kable(dg_tbl |> select(ver, metric_key, ok),
             caption = "cell/tile=*/data_0.parquet digest == cell_metric digest, per metric (browser-equivalent read)")
stopifnot(
  "a browser-equivalent wide-tile digest disagreed with cell_metric" =
    nrow(dg_tbl) == 0 || all(dg_tbl$ok))
```

### `boot.json` zone metrics == `zone_metric` (joined on `zone_type` AND `metric_key`)

```{r}
#| label: validate-zone-metrics

zone_metric_gate <- function(b, ver) {
  if (!isTRUE(b$ok) || is.null(b$out$boot)) {
    app_bundle_gate_note("zone_metric", ver, FALSE, if (!isTRUE(b$ok))
      "build did not complete" else "release has no boot.json (build did not reach that stage)")
    return(tibble())
  }
  con <- app_bundle_open_scoring(ver)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  # round 3 item 2: a release with no zone_metric table would otherwise hit
  # the JOIN below and ERROR (unknown table), not gracefully return zero rows
  # -- this is exactly the "no capability -> SKIP, don't fail" case, caught
  # BEFORE the query rather than relying on the query to fail safely
  if (!"zone_metric" %in% DBI::dbListTables(con)) {
    app_bundle_gate_note("zone_metric", ver, FALSE, "release has no zone_metric table")
    return(tibble())
  }
  app_bundle_gate_note("zone_metric", ver, TRUE)
  duckdb_tune(con, file.path(tempdir(), "app_bundle_spill_zm", ver))
  vm <- msens::sdm_val_col(con, "zone_metric"); vz <- msens::sdm_val_col(con, "zone")
  # round 4 (found rendering v7b for real, 2026-09-21; correction from review):
  # `zone_metric` carries "*_coverage" rows (e.g.
  # extrisk_bird_ecoregion_rescaled_coverage) beside the real score metrics --
  # my FIRST fix here excluded them from the ground truth, on the theory
  # msens never publishes them. Wrong: `msens::app_boot()` (R/app_bundle.R
  # ~407-423) DOES publish every one, split into two objects per zone --
  # `metrics` (keys NOT ending "_coverage") and `coverage` (keys ending
  # "_coverage", published with the suffix STRIPPED, e.g.
  # `coverage$extrisk_bird_ecoregion_rescaled`). Proved against the real v7b
  # bundle: programarea alone has 458 `metrics` + 146 `coverage` values;
  # every unit together is 983 + 146 = 1129 = n_pub exactly. The bug was this
  # GATE never reading `zr$coverage` at all, not the ground truth -- fixed by
  # reading both and mapping a coverage entry's key back to `{key}_coverage`
  # to rejoin it against `pub`.
  pub <- DBI::dbGetQuery(con, glue::glue("
    SELECT z.fld, z.{vz} AS zkey, m.metric_key, zm.{vm} AS val
      FROM zone z JOIN zone_metric zm USING (zone_seq) JOIN metric m USING (metric_seq)"))
  pub$zone_type <- sub("_key$", "", pub$fld)
  boot <- b$out$boot
  boot_rows <- do.call(rbind, lapply(names(boot$zones), function(ztype) {
    zl <- boot$zones[[ztype]]
    do.call(rbind, lapply(zl, function(zr) {
      m_rows <- if (length(zr$metrics)) data.frame(
        zone_type = ztype, zkey = zr$key, metric_key = names(zr$metrics),
        val = as.numeric(unlist(zr$metrics)), stringsAsFactors = FALSE) else NULL
      c_rows <- if (length(zr$coverage)) data.frame(
        zone_type = ztype, zkey = zr$key, metric_key = paste0(names(zr$coverage), "_coverage"),
        val = as.numeric(unlist(zr$coverage)), stringsAsFactors = FALSE) else NULL
      rbind(m_rows, c_rows)
    }))
  }))
  j <- merge(pub, boot_rows, by = c("zone_type", "zkey", "metric_key"), suffixes = c("_pub", "_boot"))
  data.frame(ver = ver, n_pub = nrow(pub), n_boot = nrow(boot_rows), n_joined = nrow(j),
             max_abs_delta = if (nrow(j)) max(abs(j$val_pub - j$val_boot)) else NA_real_)
}
zm_tbl <- purrr::imap(builds, zone_metric_gate) |> bind_rows()
knitr::kable(zm_tbl, caption = "boot.json zone metrics vs published zone_metric (join on zone_type + zkey + metric_key)")
stopifnot("boot.json zone metrics disagree with zone_metric" =
  nrow(zm_tbl) == 0 || all(zm_tbl$n_pub == zm_tbl$n_boot & zm_tbl$max_abs_delta < 1e-6, na.rm = TRUE))
```

### Layer list == manifest's drawable (cell-level) metrics; shard keys == `taxa.json`; every alias resolves

Reads the WRITTEN `taxa.json` / `taxon/*.json` / `alias/*.json` from disk — `app_bundle_build()`
does not return the full shard contents in memory, only counts, and the point of this gate is the
published bytes anyway, not what the builder happened to hold before writing them.

```{r}
#| label: validate-sets

set_gates <- function(b, ver) {
  if (!isTRUE(b$ok) || is.null(b$out$boot)) {
    app_bundle_gate_note("sets", ver, FALSE, if (!isTRUE(b$ok))
      "build did not complete" else "release has no boot.json (build did not reach that stage)")
    return(tibble())
  }
  app_bundle_gate_note("sets", ver, TRUE)
  con <- app_bundle_open_scoring(ver)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  cell_keys  <- sort(unique(msens:::.app_cell_metric_keys(con)))
  layer_keys <- sort(vapply(b$out$boot$layers, function(l) l$metric_key, ""))
  layers_ok  <- identical(cell_keys, layer_keys)

  taxa_path <- file.path(b$dir_out, "taxa.json")
  taxa_keys <- if (file.exists(taxa_path))
    unique(unname(unlist(jsonlite::fromJSON(taxa_path, simplifyVector = FALSE)$key))) else character()

  shards_ok <- NA; n_shard_keys <- NA_integer_
  taxon_dir <- file.path(b$dir_out, "taxon")
  if (dir.exists(taxon_dir) && length(taxa_keys)) {
    shard_keys <- unique(unname(unlist(lapply(list.files(taxon_dir, full.names = TRUE), function(f)
      names(jsonlite::fromJSON(f, simplifyVector = FALSE)$taxa)))))
    shards_ok <- setequal(shard_keys, taxa_keys)
    n_shard_keys <- length(shard_keys)
  }

  alias_ok <- NA; n_alias_bad <- NA_integer_
  alias_dir <- file.path(b$dir_out, "alias")
  if (dir.exists(alias_dir) && length(taxa_keys)) {
    targets <- unname(unlist(lapply(list.files(alias_dir, full.names = TRUE), function(f) {
      al <- jsonlite::fromJSON(f, simplifyVector = FALSE)$alias
      vapply(al, function(x) x[[1]], "")
    })))
    bad <- setdiff(unique(targets), taxa_keys)
    alias_ok <- length(bad) == 0
    n_alias_bad <- length(bad)
  }
  data.frame(ver = ver, layers_eq_cell_metrics = layers_ok,
             n_cell_keys = length(cell_keys), n_layer_keys = length(layer_keys),
             shard_keys_eq_taxa = shards_ok, n_shard_keys = n_shard_keys, n_taxa_keys = length(taxa_keys),
             every_alias_resolves = alias_ok, n_alias_unresolved = n_alias_bad)
}
sets_tbl <- purrr::imap(builds, set_gates) |> bind_rows()
knitr::kable(sets_tbl, caption = "Set-equality gates: layers, taxon shard keys, alias targets")
```

### Manifest identity — the manifest `app_bundle_build()` actually used, never a rebuilt one

Round 10 (2026-09-22): `msens::manifest_build()`'s one-row-per-`zone_set_key` collapse can lose a
genuine tie between two zone tables for the same field to DuckDB's own row order — found on v2's
`subregion_key` (`ply_subregions_2025` vs `ply_subregions_2026`, `n = 4` each): an installed
package build and a source-tree run of the SAME commit picked different tables. The fix moved to
[app_bundle_fetch_published_manifest()] — the fixed, already-published `{ver}/manifest.json`, not a
fresh rebuild — but the fix needs its own gate, or a future regression (something reintroducing a
rebuilt manifest into the `app_bundle_build()` call) would be silent again. This gate checks the
PUBLISHED manifest against the release DATABASE's own `zone_taxon.zone_tbl` column (where the
release has one: v2, v3, ..., v7b) and against `boot$units[].zone_tbl` — no rebuilt manifest
anywhere in the comparison.

```{r}
#| label: validate-manifest-identity

manifest_identity_gate <- function(b) {
  none <- data.frame(ver = character(), zone_fld = character(), manifest_tbl = character(),
                     n_manifest_rows = integer(), boot_zone_tbl = character(),
                     content_matches_chosen_tbl = logical(), ok = logical())
  if (!isTRUE(b$ok) || is.null(b$pub_manifest) || is.null(b$out$boot)) return(none)
  mz <- b$pub_manifest$zones
  if (is.null(mz) || !nrow(mz)) return(none)
  p <- file.path(b$dir_out, "zone_taxon.parquet")
  if (!file.exists(p)) return(none)
  zt <- arrow::read_parquet(p)
  dup_flds <- unique(mz$fld[duplicated(mz$fld)])   # requirement 1: exactly one row per fld

  con <- app_bundle_open_scoring(b$ver)
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE), add = TRUE)
  has_zt_tbl_col <- "zone_taxon" %in% DBI::dbListTables(con) &&
    "zone_tbl" %in% DBI::dbListFields(con, "zone_taxon")

  units_by_type <- setNames(b$out$boot$units, vapply(b$out$boot$units, function(u) u$zone_type, ""))
  bind_rows(lapply(seq_len(nrow(mz)), function(i) {
    fld  <- mz$fld[i]; tbl <- mz$tbl[i]; type <- sub("_key$", "", fld)
    u        <- units_by_type[[type]]
    boot_tbl <- if (!is.null(u) && !is.null(u$zone_tbl)) u$zone_tbl else NA_character_
    # requirement 3: the WRITTEN zone_taxon.parquet must hold the CHOSEN
    # table's own rows -- checked as content (its zone_value SET), not just a
    # name, because the raw db zone_taxon table legitimately carries EVERY
    # candidate table's rows for a genuine tie (v2 subregion_key: both
    # ply_subregions_2025 and _2026 are really there) -- a name-only check
    # would pass no matter which table's rows actually got written.
    content_ok <- TRUE
    if (has_zt_tbl_col) {
      db_keys_for_tbl <- sort(unique(DBI::dbGetQuery(con, glue::glue(
        "SELECT DISTINCT zone_value FROM zone_taxon
          WHERE zone_fld = {DBI::dbQuoteString(con, fld)} AND zone_tbl = {DBI::dbQuoteString(con, tbl)}"))$zone_value))
      written_keys <- sort(unique(as.character(zt$zone_value[zt$zone_fld == fld])))
      content_ok <- length(db_keys_for_tbl) > 0 && identical(written_keys, db_keys_for_tbl)
    }
    ok <- !(fld %in% dup_flds) && (is.null(u) || identical(boot_tbl, tbl)) && content_ok
    data.frame(ver = b$ver, zone_fld = fld, manifest_tbl = tbl, n_manifest_rows = sum(mz$fld == fld),
               boot_zone_tbl = boot_tbl, content_matches_chosen_tbl = content_ok, ok = ok)
  }))
}
manifest_identity_tbl <- bind_rows(lapply(builds, manifest_identity_gate))
knitr::kable(manifest_identity_tbl, caption = paste(
  "Manifest identity: the PUBLISHED manifest's zones[].tbl (one row per fld) must equal",
  "boot$units[].zone_tbl for every published unit, and the WRITTEN zone_taxon.parquet's content",
  "must match that table's own rows in the release db -- ok must be TRUE"))
stopifnot(
  "the published manifest, boot$units[].zone_tbl, or the written zone_taxon.parquet disagree on which zone table won" =
    nrow(manifest_identity_tbl) == 0 || all(manifest_identity_tbl$ok))
```

### `zone_taxon.parquet` integrity — independent of msens, by design

msens owns `app_zone_taxon()`'s internals (which of a zone type's source tables to keep, how a
duplicate resolves). This gate does not call into any of that: it reads the WRITTEN
`zone_taxon.parquet` and `boot.json` bytes and checks them against each other and against the
same GeoPackage geometry `app_bundle_geom_keys()` already resolved for the geometry-consistency
gate above — three independent sources that must agree, with no msens function anywhere in the
comparison. Two checks:

1. `(zone_fld, zone_value, mdl_key)` is unique in `zone_taxon.parquet` — the same taxon, in the
   same zone, cannot appear twice with two different `area_km2`.
2. For every drawable unit, `boot$units[].keys` (what the app offers), the real GeoPackage's keys
   (`app_bundle_geom_keys()`), and the distinct `zone_value`s actually published in
   `zone_taxon.parquet` for that field must all agree — after dropping any whole-study-area
   rollup (`USA`/`FULL`) from the `zone_taxon.parquet` side, since a rollup is scored, has no
   polygon, and is never in the GeoPackage or in `boot$units[].keys` by round 3's own rule (found
   running v8/v9 for real, 2026-09-22: both list `USA` as a `subregion_key` value beside
   `AK`/`AT`/`GA`/`PA`, which is correct, not a defect).

**Proved RED on today's v2 bundle** (built on msens 0.43.0 @ `71aa231`, before its zone-table fix
landed): v2's source has two zone tables for `subregion_key`
(`ply_subregions_2025`/`ply_subregions_2026`); `app_zone_taxon()` merged rows from both instead of
choosing one, so 13,077 `(zone_fld, zone_value, mdl_key)` triples are duplicated with two different
`area_km2` values each (e.g. `subregion_key`/`AK`/mdl_key `1` appears twice) — verified directly
against `_output/app_bundle/v2/zone_taxon.parquet`: `n_dup_groups = 13077`, `n_dup_rows = 13077`.

```{r}
#| label: validate-zone-taxon-integrity

zone_taxon_integrity_gate <- function(b) {
  none_dup  <- data.frame(ver = character(), n_dup_groups = integer(), n_dup_rows = integer(),
                          zone_taxon_unique = logical())
  none_unit <- data.frame(ver = character(), zone_type = character(),
                          n_boot_keys = integer(), n_geom_keys = integer(), n_zt_keys = integer(),
                          missing_from_zt = character(), extra_in_zt = character(),
                          warn_missing = logical(), stop_extra = logical())
  if (!isTRUE(b$ok) || is.null(b$out$boot)) return(list(dup = none_dup, unit = none_unit))
  p <- file.path(b$dir_out, "zone_taxon.parquet")
  if (!file.exists(p)) return(list(dup = none_dup, unit = none_unit))
  zt <- arrow::read_parquet(p)

  dupg <- zt |> dplyr::count(zone_fld, zone_value, mdl_key) |> dplyr::filter(n > 1)
  dup_row <- data.frame(ver = b$ver, n_dup_groups = nrow(dupg),
                        n_dup_rows = if (nrow(dupg)) sum(dupg$n - 1) else 0L,
                        zone_taxon_unique = nrow(dupg) == 0)

  # round 9 (msens 0.43.0 @ 9e66946, "the geometry check asks about units"):
  # a scored, published unit key can legitimately have ZERO zone_taxon rows
  # (v8 subregion AT: scored, 52,674 cells, but no species-per-zone rows --
  # boot$zones[unit][].n_taxa == 0, a gap in the release's OWN table, not an
  # error msens invents around). That direction -- a boot key missing from
  # zone_taxon -- is a WARNING here, never a stop. The OTHER direction -- a
  # zone_taxon key that ISN'T a boot key at all -- is still a real integrity
  # violation (data for a key that was never even a valid unit) and stays a
  # hard stop. A key with no polygon (USA/FULL rollups) is dropped from the
  # zone_taxon side before either comparison, same as before.
  units <- b$out$boot$units
  unit_rows <- bind_rows(lapply(units, function(u) {
    fld       <- paste0(u$zone_type, "_key")
    boot_keys <- sort(unique(as.character(unlist(u$keys))))
    geom_k    <- sort(unique(as.character(b$geom_keys[[u$zone_type]])))
    zt_keys   <- sort(intersect(unique(as.character(zt$zone_value[zt$zone_fld == fld])), geom_k))
    missing <- setdiff(boot_keys, zt_keys)   # scored+published, but 0 zone_taxon rows -- WARN
    extra   <- setdiff(zt_keys, boot_keys)   # zone_taxon data for a non-unit key -- STOP
    data.frame(ver = b$ver, zone_type = u$zone_type,
              n_boot_keys = length(boot_keys), n_geom_keys = length(geom_k), n_zt_keys = length(zt_keys),
              missing_from_zt = paste(missing, collapse = ","),
              extra_in_zt     = paste(extra, collapse = ","),
              warn_missing = length(missing) > 0, stop_extra = length(extra) > 0)
  }))
  list(dup = dup_row, unit = if (is.null(unit_rows)) none_unit else unit_rows)
}
zti <- lapply(builds, zone_taxon_integrity_gate)
zti_dup_tbl  <- bind_rows(lapply(zti, `[[`, "dup"))
zti_unit_tbl <- bind_rows(lapply(zti, `[[`, "unit"))
knitr::kable(zti_dup_tbl, caption = paste(
  "zone_taxon.parquet uniqueness: (zone_fld, zone_value, mdl_key) must never repeat",
  "-- zone_taxon_unique=FALSE is a real defect (a fan-out from a stale zone-table join)"))
knitr::kable(zti_unit_tbl, caption = paste(
  "Per unit: boot$units[].keys vs the distinct (in-geometry) zone_value published in",
  "zone_taxon.parquet -- warn_missing (scored+published, 0 zone_taxon rows) is a WARNING;",
  "stop_extra (zone_taxon data for a non-unit key) is a hard stop"))
if (any(zti_unit_tbl$warn_missing))
  for (i in which(zti_unit_tbl$warn_missing))
    log_warn(paste0("{zti_unit_tbl$ver[i]} {zti_unit_tbl$zone_type[i]} {zti_unit_tbl$missing_from_zt[i]}: ",
                    "scored but 0 zone_taxon rows"))
stopifnot(
  "zone_taxon.parquet has a duplicated (zone_fld, zone_value, mdl_key) triple" =
    nrow(zti_dup_tbl) == 0 || all(zti_dup_tbl$zone_taxon_unique),
  "zone_taxon.parquet has rows for a key that is not a published unit key" =
    nrow(zti_unit_tbl) == 0 || !any(zti_unit_tbl$stop_extra))
```

### Anonymous COG spot-check (species assets) — SKIPS, with a reason, when a release has no merged COG

Formalizes what was previously an ad-hoc post-render `curl` script into a real notebook gate.
Samples ALL `sp_cat == "turtle"` taxa (a handful, always checked in full) plus up to 300 other
taxa, from the WRITTEN `taxon/*.json` shards ([app_bundle_cog_sample()]) — the exact bytes an app
would fetch, not a query against the database — then HEADs each URL anonymously, in parallel
([app_bundle_head_check()]). `xargs -P`'s workers finish in an unpredictable order, so each result
line carries both its URL and its status code and is re-associated by CONTENT, never by position —
a naive positional read would silently pair the wrong status with the wrong URL under load. A
release whose written bundle has no taxon with a merged COG (no asset registry — `model_asset` /
`native_asset` — was reachable at build time; see [app_bundle_open()]'s 3-tier probe above) SKIPS
this gate with that reason rather than failing it: there is nothing to check, which is not the same
as everything being broken.

```{r}
#| label: validate-cog-sample

cog_gate <- function(b) {
  none <- function(reason) data.frame(
    ver = b$ver, ran = FALSE, reason = reason, n_sampled = NA_integer_,
    n_checked = NA_integer_, n_200 = NA_integer_, n_other = NA_integer_,
    n_no_answer = NA_integer_, worst_status = NA_integer_, worst_url = NA_character_)
  if (!isTRUE(b$ok)) {
    app_bundle_gate_note("cog_species_spotcheck", b$ver, FALSE, "build did not complete")
    return(none("build did not complete"))
  }
  samp <- app_bundle_cog_sample(b$dir_out)
  if (!nrow(samp)) {
    reason <- "no taxon in this release's written bundle carries a merged COG (no asset registry at build time)"
    app_bundle_gate_note("cog_species_spotcheck", b$ver, FALSE, reason)
    return(none(reason))
  }
  app_bundle_gate_note("cog_species_spotcheck", b$ver, TRUE)
  # round 4 (found for real 2026-09-21: this gate reported "an object is
  # missing" while four concurrent Playwright suites saturated the laptop's
  # own network, load ~40 -- every request timed out (curl's `000`), and the
  # SAME urls answered 200 seconds later with the load gone).
  # app_bundle_head_check_retry() retries only a `000`/no-answer result, at
  # LOWER parallelism, with backoff; app_bundle_head_classify() then keeps a
  # REAL non-200 (the actual defect: the object answered and it is wrong)
  # separate from "no answer after every retry" (nothing was verified for
  # that URL -- NOT the same claim, and still not a PASS: see `ok` below).
  chk <- app_bundle_head_check_retry(samp$url)
  cl  <- app_bundle_head_classify(chk)
  data.frame(ver = b$ver, ran = TRUE, reason = "", n_sampled = nrow(samp), n_checked = nrow(chk),
             n_200 = cl$n_200, n_other = cl$n_other, n_no_answer = cl$n_no_answer,
             worst_status = cl$worst_status, worst_url = cl$worst_url)
}
cog_tbl <- bind_rows(lapply(builds, cog_gate))
knitr::kable(cog_tbl, caption = paste(
  "Anonymous COG spot-check (all turtle taxa + up to 300 random others);",
  "ran=FALSE with a reason is a SKIP, not a failure. n_200/n_other/n_no_answer is the",
  "status histogram AFTER retries -- n_other is a real non-200 (the defect);",
  "n_no_answer means nothing was verified for that URL (network/local load, not a missing object)."))
stopifnot(
  "an anonymous COG spot-check found a real non-200 response for a version that DOES have merged COGs" =
    nrow(cog_tbl) == 0 || all(!cog_tbl$ran | cog_tbl$n_other == 0),
  "an anonymous COG spot-check could not verify every URL even after retries (no-answer, not a PASS)" =
    nrow(cog_tbl) == 0 || all(!cog_tbl$ran | cog_tbl$n_no_answer == 0))
```

### Gate coverage — which capability-conditioned gates ran per version, and why one was skipped

Every gate above that can legitimately have nothing to check ([one_file_per_tile_gate],
[tile_width_gate], [digest_gate], `zone_metric_gate`, `set_gates`, `cog_gate`) calls
[app_bundle_gate_note()] right beside the condition that decides whether it runs, so this table can
never drift from what the gates actually did.

```{r}
#| label: validate-gate-coverage

gate_coverage_tbl <- app_bundle_gate_coverage()
knitr::kable(gate_coverage_tbl,
             caption = "Capability-conditioned gates: ran (TRUE/FALSE) per version, and the reason when FALSE")
stopifnot(
  "every attempted version must have exactly one coverage row per capability-conditioned gate" =
    nrow(gate_coverage_tbl) == 0 ||
    (setequal(unique(gate_coverage_tbl$ver), vers) &&
     length(unique(table(gate_coverage_tbl$ver))) == 1L))
```

## Budgets (bytes vs the subplan's per-object budgets — FAILS the render, not just flags)

Review round 1: the notebook used to carry its own private thresholds (9 MB for
`zone_taxon.parquet`, 1.5 MB for `taxon.parquet`) that the two real measurements could never
exceed, and only ever printed a table rather than stopping anything. [app_bundle_budgets()]
(`libs/app_bundle.R`) is now the ONE place those numbers live — the subplan's own, with
`zone_taxon.parquet` corrected to 8 MB — and this chunk `stopifnot()`s on the result: an object
over budget fails the render, exactly like every other gate here. Sizes are measured the same
way a real push would produce them: gzip via [app_bundle_gzip_file()] (byte-safe, never
`readLines()`/`writeLines()`) for JSON, raw bytes for Parquet; the cell-tile row reads ONLY
`tile=*/data_0.parquet` (browser-equivalent, see the digest gate above).

```{r}
#| label: budgets

budgets_spec <- app_bundle_budgets()

.budget_row <- function(object, sizes, kind, budget_bytes) {
  if (!length(sizes)) return(NULL)
  data.frame(object = object, kind = kind, n = length(sizes),
             bytes_p50 = as.integer(round(median(sizes))),
             bytes_p90 = as.integer(round(quantile(sizes, 0.9))),
             bytes_max = as.integer(max(sizes)),
             budget_bytes = as.integer(budget_bytes),
             n_over_budget = sum(sizes > budget_bytes))
}
.sizes_for <- function(dir_out, object, kind) {
  gz_of <- function(p) { tmp <- tempfile(fileext = ".gz"); app_bundle_gzip_file(p, tmp)
                         sz <- file.size(tmp); unlink(tmp); sz }
  if (object %in% c("boot.json", "taxa.json")) {
    p <- file.path(dir_out, object)
    if (!file.exists(p)) return(numeric())
    if (kind == "gzip") gz_of(p) else file.size(p)
  } else if (object %in% c("taxon/*.json", "alias/*.json")) {
    d <- file.path(dir_out, sub("/\\*\\.json$", "", object))
    if (!dir.exists(d)) return(numeric())
    fs_ <- list.files(d, full.names = TRUE)
    if (kind == "gzip") vapply(fs_, gz_of, numeric(1)) else file.size(fs_)
  } else if (object == "cell/tile=*/data_0.parquet") {
    fs_ <- Sys.glob(file.path(dir_out, "cell", "tile=*", "data_0.parquet"))  # browser-equivalent
    file.size(fs_)
  } else {  # taxon.parquet, taxonomy.parquet, zone_taxon.parquet
    p <- file.path(dir_out, object)
    if (!file.exists(p)) numeric() else file.size(p)
  }
}

budgets_one <- function(b) {
  if (!isTRUE(b$ok)) return(tibble())
  rows <- lapply(seq_len(nrow(budgets_spec)), function(i) {
    spec <- budgets_spec[i, ]
    .budget_row(spec$object, .sizes_for(b$dir_out, spec$object, spec$kind), spec$kind, spec$budget_bytes)
  })
  out <- bind_rows(Filter(Negate(is.null), rows))
  if (nrow(out)) out$ver <- b$ver
  out
}
budgets_tbl <- bind_rows(purrr::map(builds, budgets_one))
knitr::kable(
  budgets_tbl |> select(ver, object, kind, n, bytes_p50, bytes_p90, bytes_max, budget_bytes, n_over_budget),
  caption = "Object budgets vs the subplan (bytes; n_over_budget must be 0)")
stopifnot(
  "an object exceeds its subplan budget" = nrow(budgets_tbl) == 0 || all(budgets_tbl$n_over_budget == 0))
```

## Per-version summary (objects, bytes, taxa, tiles)

```{r}
#| label: summary-per-version

summary_one <- function(b) {
  if (!isTRUE(b$ok))
    return(data.frame(ver = b$ver, ok = FALSE, reason = b$reason %||% NA_character_, n_objects = NA,
                      bytes_raw = NA, bytes_gzip_est = NA, n_taxa = NA, n_tiles = NA, wall_s = NA))
  d <- b$dir_out
  files <- list.files(d, recursive = TRUE, full.names = TRUE)
  is_json <- grepl("\\.json$", files)
  gz1 <- function(p) { tmp <- tempfile(fileext = ".gz"); app_bundle_gzip_file(p, tmp)
                       sz <- file.size(tmp); unlink(tmp); sz }
  raw <- sum(file.size(files))
  gz  <- sum(vapply(files[is_json], gz1, numeric(1))) + sum(file.size(files[!is_json]))
  data.frame(ver = b$ver, ok = TRUE, reason = "",
             n_objects = length(files), bytes_raw = raw, bytes_gzip_est = gz,
             n_taxa = if (!is.null(b$out)) b$out$n_taxa else NA,
             n_tiles = if (!is.null(b$out$tiles)) b$out$tiles$tiles else NA,
             wall_s = round(b$wall_s, 1))
}
summary_tbl <- bind_rows(purrr::map(builds, summary_one))
knitr::kable(summary_tbl, caption = "Per-version summary: objects, raw/gzip-estimated bytes, taxa, tiles, wall clock")
```

## `serve/cell_model/` upload plan (allow-listed versions only) — local gates, dry-run, never a write

Round 3 item 3: generalized from a single v7-only flag to an explicit ALLOW-LIST parameter
(`APP_BUNDLE_CELLMODEL_VERS`, default empty). This chunk always runs, even when `APP_BUNDLE_S3` is
unset — it is a pure dry run: a local directory walk plus one anonymous HEAD per allow-listed
version, nothing is ever written here. Per version: the local tile-width gate (`nc=3103`) and
one-file-per-tile gate run on the LOCAL source tiles [app_bundle_cell_model_plan()], every planned
key is checked against the EXACT shape `^{ver}/serve/cell_model/tile=[0-9]+/data_0\.parquet$`
[app_bundle_assert_cell_model_keys()], and the version is refused outright if its cell_model is
ALREADY on S3 [app_bundle_cell_model_already_on_s3()] — upload is all-or-nothing, never a partial
overwrite. v8/v9 are refused before any of that even runs: neither has a local `{ver}/cell_model/`
export (verified 2026-09-21: `~/_big/msens/derived/v9/cell_model` does not exist on this machine),
so [app_bundle_cell_model_dir()] returns `NA` and the plan reports "no local source directory".

```{r}
#| label: validate-cell-model-plan

cm_plan_one <- function(ver) {
  plan <- app_bundle_cell_model_plan(ver, dir_derived)
  if (!length(plan$keys))
    return(data.frame(ver = ver, n_tiles = 0L, n_keys = 0L, bytes = 0, keys_ok = NA,
                      one_file_per_tile_ok = NA, tile_width_ok = NA, already_on_s3 = NA,
                      clear_to_upload = FALSE,
                      reason = sprintf("no local source directory (%s)",
                                       file.path(dir_derived, ver, "cell_model"))))
  keys_ok <- isTRUE(tryCatch({ app_bundle_assert_cell_model_keys(plan$keys, ver); TRUE },
                             error = function(e) { message(conditionMessage(e)); FALSE }))
  tiledirs <- list.dirs(plan$dir, full.names = TRUE, recursive = FALSE)
  n_files  <- vapply(tiledirs, function(td) length(list.files(td, pattern = "\\.parquet$")), integer(1))
  one_ok   <- length(n_files) > 0 && all(n_files == 1L)
  tw_ok    <- isTRUE(tryCatch({ msens::app_cell_tile_check(plan$dir, 3103L); TRUE },
                              error = function(e) { message(conditionMessage(e)); FALSE }))
  already  <- app_bundle_cell_model_already_on_s3(ver, base)
  clear    <- isTRUE(keys_ok) && isTRUE(one_ok) && isTRUE(tw_ok) && !isTRUE(already)
  reason   <- if (!isTRUE(keys_ok)) "a key failed the exact tile-filename shape check" else
              if (!isTRUE(one_ok)) "a local tile directory has zero or multiple .parquet files" else
              if (!isTRUE(tw_ok))  "a local tile failed the width check (nc=3103)" else
              if (isTRUE(already)) "already present on S3 (refusing to overwrite)" else ""
  data.frame(ver = ver, n_tiles = plan$n_tiles, n_keys = length(plan$keys), bytes = plan$bytes,
             keys_ok = keys_ok, one_file_per_tile_ok = one_ok, tile_width_ok = tw_ok,
             already_on_s3 = already, clear_to_upload = clear, reason = reason)
}
cm_plan_tbl <- bind_rows(lapply(cm_vers, cm_plan_one))
if (!length(cm_vers))
  message("APP_BUNDLE_CELLMODEL_VERS is empty -- no serve/cell_model/ upload planned")
# `knitr::kable()` must be its OWN top-level chunk statement to auto-print --
# a plain print() (or one buried inside an `if {}`/`for` block) bypasses
# knitr's knit_print() "asis" hook and pandoc is left with a literal pipe-table
# STRING instead of an HTML <table> (found rendering this exact chunk
# 2026-09-21: the table above showed as unrendered `|ver|n_tiles|...` text).
# A zero-row cm_plan_tbl (the ordinary case: the allow-list is empty) still
# kables cleanly -- header only, no data rows.
knitr::kable(cm_plan_tbl, caption = paste(
  "serve/cell_model/ upload plan: local gates + already-on-S3 refusal,",
  "per allow-listed version (dry run; nothing written here)"))
if (nrow(cm_plan_tbl))
  for (i in seq_len(nrow(cm_plan_tbl)))
    log_info(paste0(
      "cell_model plan {cm_plan_tbl$ver[i]}: {cm_plan_tbl$n_keys[i]} keys, ",
      "{cm_plan_tbl$bytes[i]} bytes, clear_to_upload={cm_plan_tbl$clear_to_upload[i]}",
      if (!cm_plan_tbl$clear_to_upload[i]) " ({cm_plan_tbl$reason[i]})" else ""))
stopifnot(
  "a version in the cell_model allow-list is not clear to upload (see the reason column)" =
    nrow(cm_plan_tbl) == 0 || all(cm_plan_tbl$clear_to_upload))
```

## Push (`APP_BUNDLE_S3=1` only — never run by this notebook's author)

Lists the target prefix before and after, refuses (via [app_bundle_assert_prefix()]) any key
outside `{ver}/app/` — and, for a version in `APP_BUNDLE_CELLMODEL_VERS`, `{ver}/serve/cell_model/`
— gzips every JSON object before upload (`Content-Encoding: gzip`, S3 does not compress on the
fly), leaves Parquet as-is (already zstd-compressed by [msens::write_atlas_parquet()] /
[msens::copy_atlas_parquet()]). The `serve/cell_model/` sync itself reuses [cm_plan_tbl] computed
above — a version is only synced if THIS run's plan already marked it `clear_to_upload`, so the
dry-run gates above are not just printed, they gate the write.

```{r}
#| label: push
#| eval: !expr do_s3

s3_root <- "s3://oceanmetrics.io-public/marine-atlas"

list_prefix <- function(prefix) {
  out <- system2("aws", c("s3", "ls", shQuote(glue("{s3_root}/{prefix}")), "--recursive"),
                stdout = TRUE, stderr = TRUE)
  if (!is.null(attr(out, "status")) && attr(out, "status") != 0) character(0) else out
}

push_one <- function(b) {
  if (!isTRUE(b$ok)) return(invisible(NULL))
  ver <- b$ver
  before <- list_prefix(glue("{ver}/app/"))
  log_info("{ver}/app/ BEFORE push: {length(before)} object(s)")

  # a controlled directory walk of dir_out -- every file that exists under it,
  # nothing else -- then EVERY key derived from it still goes through the
  # whitelist guard before anything touches S3; the walk being "controlled"
  # is not itself the protection, the guard call two lines down is
  files <- list.files(b$dir_out, recursive = TRUE, full.names = TRUE)
  rels  <- fs::path_rel(files, b$dir_out)
  keys  <- glue("{ver}/app/{rels}")
  app_bundle_assert_prefix(keys, ver)                       # refuse before touching S3

  for (i in seq_along(files)) {
    is_json <- grepl("\\.json$", files[i])
    src <- files[i]
    if (is_json) {
      gz <- tempfile(fileext = ".gz")
      app_bundle_gzip_file(files[i], gz)                    # byte-safe, never readLines()
      app_bundle_assert_gzip_roundtrip(gz, files[i])         # proves upload bytes == validated bytes
      src <- gz
    }
    args <- c("s3", "cp", shQuote(src), shQuote(glue("{s3_root}/{keys[i]}")),
              if (is_json) c("--content-type", shQuote("application/json"),
                             "--content-encoding", "gzip"))
    out <- system2("aws", args, stdout = TRUE, stderr = TRUE)
    if (!is.null(attr(out, "status")) && attr(out, "status") != 0)
      stop(glue("aws s3 cp failed for {keys[i]}: {paste(out, collapse = '\n')}"))
    if (is_json) unlink(src)
  }

  if (ver %in% cm_vers) {
    plan_row <- cm_plan_tbl[cm_plan_tbl$ver == ver, , drop = FALSE]
    if (nrow(plan_row) && isTRUE(plan_row$clear_to_upload[1])) {
      plan <- app_bundle_cell_model_plan(ver, dir_derived)
      app_bundle_assert_cell_model_keys(plan$keys, ver)          # exact shape, again, right before the write
      app_bundle_assert_prefix(plan$keys, ver, allow_cell_model = TRUE)  # whitelist, again
      system2("aws", c("s3", "sync", shQuote(plan$dir),
                       shQuote(glue("{s3_root}/{ver}/serve/cell_model")), "--only-show-errors"),
              stdout = TRUE, stderr = TRUE)
      log_info("{ver}: synced serve/cell_model/ -- {length(plan$keys)} keys, {plan$bytes} bytes")
    } else {
      stop(glue(
        "{ver} is in APP_BUNDLE_CELLMODEL_VERS but this run's plan did not mark it clear_to_upload ",
        "-- refusing to sync (see the validate-cell-model-plan chunk's reason column)"))
    }
  }

  after <- list_prefix(glue("{ver}/app/"))
  log_info("{ver}/app/ AFTER push: {length(after)} object(s) ({length(after) - length(before)} new)")
  invisible(list(before = before, after = after))
}
push_out <- purrr::map(builds, push_one)
```

## Anonymous verification (`APP_BUNDLE_S3=1` only)

HEAD (falling back to a ranged GET) per published object, anonymously — no credentials, exactly
as the browser will fetch it: 200, `content-encoding: gzip` on every JSON object, path-style URL,
size within budget.

```{r}
#| label: verify-anonymous
#| eval: !expr do_s3

verify_one <- function(b) {
  if (!isTRUE(b$ok)) return(tibble())
  ver <- b$ver
  files <- list.files(b$dir_out, recursive = TRUE, full.names = TRUE)
  rels  <- fs::path_rel(files, b$dir_out)
  urls  <- glue("{base}/{ver}/app/{rels}")
  purrr::map_dfr(seq_along(urls), function(i) {
    r <- tryCatch(
      httr2::req_perform(httr2::req_error(httr2::req_method(httr2::request(urls[i]), "HEAD"),
                                          is_error = function(resp) FALSE)),
      error = function(e) NULL)
    is_json <- grepl("\\.json$", files[i])
    data.frame(
      ver = ver, key = rels[i],
      status = if (is.null(r)) NA_integer_ else httr2::resp_status(r),
      path_style = grepl("^https://s3\\.[a-z0-9-]+\\.amazonaws\\.com/", urls[i]),
      gzip_ok = if (is.null(r) || !is_json) NA else
        identical(httr2::resp_header(r, "content-encoding"), "gzip"),
      bytes = if (is.null(r)) NA_integer_ else
        suppressWarnings(as.integer(httr2::resp_header(r, "content-length"))))
  })
}
verify_tbl <- bind_rows(purrr::map(builds, verify_one))
knitr::kable(
  verify_tbl |> summarise(n = n(), n_200 = sum(status == 200, na.rm = TRUE),
                          n_gzip_ok = sum(gzip_ok, na.rm = TRUE), .by = ver),
  caption = "Anonymous verification summary per version")
stopifnot("not every pushed object answered 200 anonymously" =
  all(verify_tbl$status == 200, na.rm = TRUE) && !anyNA(verify_tbl$status))
```

## Target manifest

```{r}
#| label: target-manifest

msens::write_manifest(
  here("data/manifests/build_app_bundle.json"),
  target       = "build_app_bundle",
  content_hash = digest::digest(
    list(summary_tbl, val_tbl, boot_contract_tbl, one_file_tbl, tw_tbl, dg_tbl, zm_tbl, sets_tbl,
        cog_tbl, gate_coverage_tbl, budgets_tbl, cm_plan_tbl),
    algo = "xxhash64"),
  stats = list(
    versions_attempted  = vers,
    versions_built      = summary_tbl$ver[summary_tbl$ok],
    push_s3             = do_s3,
    cell_model_vers     = cm_vers,
    cell_model_clear_to_upload = if (nrow(cm_plan_tbl)) cm_plan_tbl$ver[cm_plan_tbl$clear_to_upload] else character(0),
    n_objects_total     = sum(summary_tbl$n_objects, na.rm = TRUE),
    n_tile_dirs_with_multiple_files = sum(one_file_tbl$n_with_multiple_files, na.rm = TRUE),
    n_gates_skipped     = sum(!gate_coverage_tbl$ran)),
  force = msens::force_target("build_app_bundle"))
```
