---
title: "Tabulate Species"
format: html
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
librarian::shelf(
  DBI, dplyr, dbplyr, glue, here, httr2, jsonlite, 
  MarineSensitivity/msens,
  readr, sf, stringr, tictoc, tidyr)
source("libs/db.R")
```

```{r}
# TODO: add req_timeout(60*20) to R fxn so timeout is 20 min, not 10 min
# https://httr2.r-lib.org/reference/req_timeout.html
# erEGOM-paEGM: 921.262 sec / 60 = 15.4 min
get_species_by_feature_localdb <- function(
    schema.table = "raw.mr_eez",
    where        = "mrgid = 8442"){
  
  # schema.table = "public.ply_ecoareas"
  # where        = "ecoarea_key = 'erCAC-paCEC'"

  source(here("../workflows/libs/db.R")) # define: con
  
  schema <- str_split(schema.table, fixed("."))[[1]][1]
  table  <- str_split(schema.table, fixed("."))[[1]][2]

  fld_geom <- dbGetQuery(
    con,
    glue("
      SELECT f_geometry_column
      FROM geometry_columns
      WHERE
        f_table_schema = '{schema}' AND
        f_table_name   = '{table}';")) |>
    pull(f_geometry_column)
  # message(glue("schema.table: '{schema}.{table}'; fld_geom: '{fld_geom}'"))

  q <- glue("
    WITH
      c AS (
        SELECT
        c.cell_id,
        c.area_km2 AS cell_km2,
        CASE
          WHEN ST_CoveredBy(c.geom, a.{fld_geom})
            THEN c.geom
          ELSE
            ST_Multi(ST_Intersection(c.geom, a.{fld_geom}))
        END AS geom
      FROM
        aquamaps.cells c
        INNER JOIN {table} a
          ON ST_Intersects(c.geom, a.{fld_geom})
            AND NOT ST_Touches(c.geom, a.{fld_geom})
      WHERE
        a.{where}
      ORDER BY
        c.cell_id),
    k AS (
      SELECT *,
        ST_AREA(geom::geography) / (1000 * 1000) AS aoi_km2
      FROM c),
    a AS (
      SELECT *,
        aoi_km2 / cell_km2 AS pct_cell
      FROM k),
    s AS (
      SELECT a.*,
        sc.sp_key, sc.probability
      FROM a
        LEFT JOIN aquamaps.spp_cells sc
          ON a.cell_id = sc.cell_id),
    g AS (
      SELECT
        sp_key,
        COUNT(*)                                  AS n_cells,
        AVG(pct_cell)                             AS avg_pct_cell,
        SUM(aoi_km2)                              AS area_km2,
        SUM(probability * aoi_km2) / SUM(aoi_km2) AS avg_suit
      FROM s
      GROUP BY sp_key)
    SELECT
      sp_key, n_cells, avg_pct_cell, area_km2, avg_suit,
      n_cells * avg_pct_cell * avg_suit AS amt,
      phylum, class, \"order\", family, genus, species
    FROM
      g
      LEFT JOIN aquamaps.spp USING (sp_key)")
  # message(q)

  dbGetQuery(con, q)
}
```


## Fetch and Store Species Data

```{r}
insert_ply_species <- function(
    schema       = "public",
    tbl          = "ply_ecoareas",
    fld_key      = "ecoarea_key",
    tbl_ply_spp  = "ply_species",
    redo = F){
  
  # keys, all
  keys <- dbGetQuery(con, glue("SELECT {fld_key} FROM {schema}.{tbl} ORDER BY {fld_key}")) |> 
    pull(fld_key)
  
  # keys, done
  if (dbExistsTable(con, tbl_ply_spp) & !redo) {
    keys_done <- dbGetQuery(con, glue("SELECT DISTINCT ply_key FROM {tbl_ply_spp} WHERE ply_tbl = '{schema}.{tbl}' ORDER BY ply_key")) |> 
      pull(ply_key)
    
    keys <- setdiff(keys, keys_done)
  }
  
  if (length(keys) == 0){
    message(glue("All keys in {tbl}.{fld_key} were already processed and inserted into {tbl_ply_spp} ~ {Sys.time()}"))
    return(NULL)
  }
  
  # Iterate over each region key
  for (i in 1:length(keys)) {  # key = rgn_keys[1] # bad: i = 1; key = "ALA" # good: key = "WGM"
    key = keys[i]
    
    # message with time stamp
    tic()
    message(glue("{sprintf('%02d',i)}/{length(keys)}: {key} ~ {Sys.time()}"))
    
    # Fetch species data from API; wrap in tryCatch to avoid breaking the loop
    d_spp <- tryCatch({
      # msens::get_species_by_feature(
      get_species_by_feature_localdb(
        glue("public.{tbl}"),
        glue("{fld_key} = '{key}'")) }, 
      error = function(e) {
        message(glue(
          "
        Error fetching species data: {e$message}
        ~ {Sys.time()}"))
        source("libs/db.R") # reset database con
        return(NULL)
      })
    
    if (is.null(d_spp))
      next
    
    # add columns identifying polygon for species
    d_spp <- d_spp |>
      mutate(
        ply_tbl = glue("{schema}.{tbl}"),
        ply_fld = fld_key,
        ply_key = key) |>
      relocate(ply_tbl, ply_fld, ply_key)
    
    # clear existing if table exists
    if (dbExistsTable(con, tbl_ply_spp)) {
      dbExecute(con, glue("DELETE FROM {tbl_ply_spp} WHERE ply_tbl = '{schema}.{tbl}' AND ply_fld = '{fld_key}' AND ply_key = '{key}'"))
    }
      
    # append to database
    dbWriteTable(con, tbl_ply_spp, d_spp, append = TRUE, row.names = FALSE)

    toc <- toc(quiet = T)
    message(glue("{sprintf('%02d',i)}/{length(keys)}: {key} ~ {toc$callback_msg}"))
  }
}

# dbRemoveTable(con, "ply_species")

insert_ply_species(
  schema       = "public",
  tbl          = "ply_ecoareas",
  fld_key      = "ecoarea_key",
  tbl_ply_spp  = "ply_species",
  redo         = F)

# 01/17: erEGOM-paEGM ~ 2024-12-21 15:43:57.250182
# Error fetching species data: Failed to perform HTTP request.
# ~ 2024-12-21 15:53:58.042009
# 06/17: erGOA-paGOA ~ 2024-12-21 16:03:20.086948
# Error fetching species data: Failed to perform HTTP request.
# ~ 2024-12-21 16:13:20.706389
# 10/17: erNECS-paNOA ~ 2024-12-21 16:26:41.31597
# Error fetching species data: Failed to perform HTTP request.
# ~ 2024-12-21 16:36:41.934764
# 11/17: erSECS-paFLS ~ 2024-12-21 16:36:42.155532
# Error fetching species data: Failed to perform HTTP request.
# ~ 2024-12-21 16:46:42.77022

insert_ply_species(
  schema       = "public",
  tbl          = "ply_ecorgns",
  fld_key      = "ecorgn_key",
  tbl_ply_spp  = "ply_species",
  redo = F)
```



## OLD db: Fetch and Store Species Data

```{r}
# show region keys
dbGetQuery(con, glue("SELECT rgn_key FROM public.ply_rgns ORDER BY rgn_key")) |> 
  pull(rgn_key)
```

region keys (n = 27):
```
 [1] "ALA" "ALB" "BFT" "BOW" "CEC" "CGM" "CHU" "COK" "EGM"
[10] "FLS" "GEO" "GOA" "HI"  "HOP" "KOD" "MAT" "MDA" "NAL"
[19] "NAV" "NOA" "NOC" "NOR" "SHU" "SOA" "SOC" "WAO" "WGM"
```

```{r}
#| eval: false
insert_ply_species(
  schema       = "public",
  tbl          = "ply_shlfs",
  fld_key      = "shlf_key",
  tbl_ply_spp  = "ply_species",
  redo         = F)
```

```{r}
#| eval: false
insert_ply_species(
  schema       = "public",
  tbl          = "ply_rgns",
  fld_key      = "rgn_key",
  tbl_ply_spp  = "ply_species",
  redo = F)
```


Executing above:
```
01: ALA ~ 2024-12-11 23:32:20.493056
Error fetching species data for ALA: HTTP 500 Internal Server Error.
01: ALA ~ 1.41 sec elapsed
02: ALB ~ 2024-12-11 23:32:21.903263
Error fetching species data for ALB: HTTP 500 Internal Server Error.
02: ALB ~ 1.509 sec elapsed
03: BFT ~ 2024-12-11 23:32:23.413059
03: BFT ~ 35.018 sec elapsed
04: BOW ~ 2024-12-11 23:32:58.431392
Error fetching species data for BOW: HTTP 500 Internal Server Error.
04: BOW ~ 1.391 sec elapsed
05: CEC ~ 2024-12-11 23:32:59.822888
05: CEC ~ 93.324 sec elapsed
06: CGM ~ 2024-12-11 23:34:33.14729
06: CGM ~ 343.167 sec elapsed
07: CHU ~ 2024-12-11 23:40:16.315158
07: CHU ~ 20.115 sec elapsed
08: COK ~ 2024-12-11 23:40:36.431037
08: COK ~ 156.458 sec elapsed                                         
09: EGM ~ 2024-12-11 23:43:12.889954
Error fetching species data for EGM: HTTP 500 Internal Server Error.
09: EGM ~ 27.012 sec elapsed
10: FLS ~ 2024-12-11 23:43:39.902751
10: FLS ~ 607.625 sec elapsed                                 
11: GEO ~ 2024-12-11 23:53:47.529269
11: GEO ~ 199.316 sec elapsed                                 
12: GOA ~ 2024-12-11 23:57:06.845524
Error fetching species data for GOA: Failed to perform HTTP request.
12: GOA ~ 328.706 sec elapsed
13: HI ~ 2024-12-12 00:02:35.552604
Error fetching species data for HI: Failed to perform HTTP request.
...
06: HI ~ 1.417 sec elapsed
07: HOP ~ 2024-12-12 00:46:31.430755
07: HOP ~ 31.362 sec elapsed          
08: KOD ~ 2024-12-12 00:47:02.792726
08: KOD ~ 144.358 sec elapsed         
09: MAT ~ 2024-12-12 00:49:27.151273
09: MAT ~ 30.936 sec elapsed          
10: MDA ~ 2024-12-12 00:49:58.088405
10: MDA ~ 480.974 sec elapsed         
11: NAL ~ 2024-12-12 00:57:59.063676
11: NAL ~ 172.836 sec elapsed         
12: NAV ~ 2024-12-12 01:00:51.900245
12: NAV ~ 3.234 sec elapsed           
13: NOA ~ 2024-12-12 01:00:55.13549
13: NOA ~ 667.073 sec elapsed         
14: NOC ~ 2024-12-12 01:12:02.20922
14: NOC ~ 102.602 sec elapsed         
15: NOR ~ 2024-12-12 01:13:44.812222
15: NOR ~ 93.109 sec elapsed          
16: SHU ~ 2024-12-12 01:15:17.922195
16: SHU ~ 94.259 sec elapsed          
17: SOA ~ 2024-12-12 01:16:52.182344
17: SOA ~ 664.422 sec elapsed         
18: SOC ~ 2024-12-12 01:27:56.605175
18: SOC ~ 467.248 sec elapsed         
19: WAO ~ 2024-12-12 01:35:43.853924
19: WAO ~ 154.774 sec elapsed         
20: WGM ~ 2024-12-12 01:38:18.627972
20: WGM ~ 182.608 sec elapsed
```

Error fetching species data for:
```
ALA
ALB
BOW
EGM
GOA
HI
```

## Assign Taxa Groups

```{r}
where_taxa <- list(
  mammals = "class = 'Mammalia' 
    AND \"order\" IN ('Cetacea', 'Carnivora')
    AND family IN ('Balaenopteridae', 'Delphinidae', 'Phocidae', 'Otariidae')",
  
  benthos = "(phylum = 'Annelida' AND class = 'Polychaeta')
    OR (phylum = 'Mollusca' AND class IN ('Bivalvia', 'Gastropoda'))
    OR (phylum = 'Echinodermata')
    OR (phylum = 'Arthropoda' AND class = 'Malacostraca')",
  
  corals = "phylum = 'Cnidaria'
    AND class IN ('Anthozoa', 'Hydrozoa')
    AND \"order\" IN ('Scleractinia', 'Alcyonacea')",
  
  pelagic_fish = "phylum = 'Chordata'
    AND class = 'Actinopterygii'
    AND \"order\" IN ('Clupeiformes', 'Perciformes', 'Gadiformes')
    AND family IN ('Scombridae', 'Clupeidae', 'Gadidae')",
  
  seabirds = "class = 'Aves'
    AND \"order\" IN ('Charadriiformes', 'Procellariiformes', 'Pelecaniformes')
    AND family IN ('Laridae', 'Procellariidae', 'Sulidae', 'Phalacrocoracidae')",
  
  turtles = "class = 'Reptilia'
    AND \"order\" = 'Testudines'
    AND family IN ('Cheloniidae', 'Dermochelyidae')",
  
  elasmo = "class = 'Chondrichthyes'
    AND (\"order\" IN ('Lamniformes', 'Carcharhiniformes', 'Rajiformes', 'Myliobatiformes')
    OR family IN ('Sphyrnidae', 'Pristidae', 'Mobulidae'))",
  
  other_fish = "phylum = 'Chordata'
    AND class = 'Actinopterygii'
    AND \"order\" NOT IN ('Clupeiformes', 'Perciformes', 'Gadiformes')
    AND family NOT IN ('Scombridae', 'Clupeidae', 'Gadidae')",
  
  other_inverts = "phylum IN ('Porifera', 'Bryozoa', 'Brachiopoda', 'Chaetognatha', 'Nemertea')
    OR (phylum = 'Arthropoda' AND class NOT IN ('Malacostraca'))
    OR (phylum = 'Mollusca' AND class NOT IN ('Bivalvia', 'Gastropoda'))"
)

# add column taxa if not already present
if (!("taxa" %in% dbListFields(con, tbl_ply_spp))) {
  dbExecute(con, glue("ALTER TABLE {tbl_ply_spp} ADD COLUMN taxa TEXT"))
}

# Assign taxa groupings to species
for (taxa in names(where_taxa)) { # taxa = names(where_taxa)[1]
  dbExecute(con, glue("UPDATE {tbl_ply_spp} SET taxa = '{taxa}' WHERE {where_taxa[[taxa]]}"))
}
```


## Visualize Species Data

### Flower Plot

```{r}
librarian::shelf(
  dplyr, ggplot2)

# Function to create the flower plot
plot_flower <- function(
    data,
    fld_category,
    fld_width,
    fld_height,
    score,
    colors = "Set2"){
  
  # Calculate positions
  data <- data  |> 
    arrange({{ fld_category }}) |> 
    mutate(
      # Calculate angles for plotting
      ymax = cumsum({{ fld_width }}),
      ymin = c(0, head(ymax, n=-1)),
      xmax = {{ fld_height }},
      xmin = 0)
  
  browser()
  
  # Create the donut chart
  ggplot(data) +
    geom_rect(aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      ymax = ymax,
      fill = {{ fld_category }}),
      color = "white") +
    coord_polar(theta = "y") +
    # Create donut hole
    xlim(c(-20, max(data |> pull({{ fld_height }})))) +
    # Add center score
    annotate(
      "text", x = -20, y = 0, 
      label = "85", 
      size = 8, 
      fontface = "bold") +
    # Customize appearance
    scale_fill_brewer(palette = colors) +
    theme_void() +
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(20, 20, 20, 20), "pt"))
}

# Sample data
ex_data <- data.frame(
  category     = c("A", "B", "C", "D"),
  width_value  = c(10, 25, 40, 25),    # Controls segment width
  height_value = c(80, 65, 45, 60))    # Controls segment height)
sum(ex_data$width_value)

ex_data |> 
  plot_flower(
    category,
    width_value,
    height_value,
    sum(ex_data$width_value))
```

### Species

```{r}
ply_key <- "BFT"

# get species data from db
schema       = "public"
tbl          = "ply_rgns"
fld_key      = "rgn_key"
tbl_ply_spp  = "ply_species"

d_spp <- dbGetQuery(con, glue(
  "SELECT * FROM {tbl_ply_spp} 
  WHERE
    ply_tbl = '{schema}.{tbl}' AND 
    ply_fld = '{fld_key}'      AND 
    ply_key = '{ply_key}'")) |> 
  replace_na(list(taxa = "other_taxa")) |> 
  # mutate(
  #   taxa = factor(taxa, levels = sort(names(where_taxa))) |> 
  #     forcats::fct_na_value_to_level("other_taxa")) |> 
  tibble()

d_taxa <- d_spp |> 
  group_by(taxa) |>
  summarize(
    n_spp       = n(),
    amt         = sum(amt),
    amt_per_spp = amt / n_spp)

d_taxa |> 
  plot_flower(
    taxa,
    n_spp,
    amt_per_spp,
    "99")
```

$$
amt = n\_cells * avg\_pct * avg\_suit
$$

Amount ($amt$) is the multiplication of the number of cells ($n\_cells$), average percent ($avg\_pct$) of a cell's contents within the selected polygon, and the average Suitability ($avg\_suit$; 0 to 100%) of the species given by AquaMaps.

```{r cleanup, include=FALSE}
# Disconnect from the database
DBI::dbDisconnect(con)
```
