---
title: "Calculate Scores"
subtitle: "Calculate scores from models loaded into sdm db and taxon table"
format:
  html:
    code-fold: true
    code-tools: true
editor_options: 
  chunk_output_type: console
---

## Overview

This notebook calculates scores from models loaded into sdm db and taxon table, including:

- Extinction risk metrics by class (eg `extrisk_bird`, `extrisk_fish`, etc)
- Other metrics TBD

**Source**:

- Copied this notebook from  [ingest_aquamaps_to_sdm_duckdb.qmd](https://github.com/MarineSensitivity/workflows/blob/3880f9edf3a204d8ec9768e7be1a1aa51f70d9f0/ingest_aquamaps_to_sdm_duckdb.qmd) to work up only calculation component from models already loaded

**Process**:

- Before:
  1. `ingest_taxon.qmd`
    - Added `con_spp.botw` (Birds of the World) as another taxonomic authority with [ingest_taxon.qmd](https://github.com/MarineSensitivity/workflows/blob/3880f9edf3a204d8ec9768e7be1a1aa51f70d9f0/ingest_taxon.qmd)
  1.`merge_models.qmd`
    - Built `con_sdm.taxon` table as authoritative unique taxa reference from multiple entries in `species` with [merge_models.qmd](https://github.com/MarineSensitivity/workflows/blob/3880f9edf3a204d8ec9768e7be1a1aa51f70d9f0/merge_models.qmd)
- After:
  1. Summary data/figs/README sections are now integrated into this notebook (merged from `msens-summary_programareas.qmd`)

To render this notebook on Ben's laptop:

```bash
cd ~/Github/MarineSensitivity/workflows
quarto render calc_scores.qmd
```


## TODO

- [x] fix "reptile" in flower plots (addressed by merge_models.qmd reclassification)
- [ ] Add functions here and in `libs/sdm_functions.R` to `msens` R package
- [ ] All R chunks under `## Add Zones` should be re-evaluated with to check existence and creation with the version suffix (`{v_sfx} = "_v3"`), eg `ply_programareas_2026_v3`, `ply_subregions_2026_v3`, etc. For now evaluation is turned off (`eval = F`). - [ ] Delete orphans, like zone_cell without zone, zone_metric without zone & metric,... Reference [DuckDB Raster Data Metric Sequence Insertion - Claude](https://claude.ai/chat/748f00aa-d2f6-46a5-91d9-fb56b27e3ebe)
- [ ] Add relationships and primary keys to sdm db tables (eg taxon_id in taxon, ds_seq in species, etc) for data integrity, and schema viz (ERD) for query understanding
- [ ] Show tabular results of R chunks before / after execution with `DT::datatable()`
- [ ] Migrate from PostGIS & pg_tileserv to [creating PMTiles](https://docs.protomaps.com/pmtiles/create) and map with `mapgl::add_pmtiles_source()`
- [ ] Redo parquet export into dir outside Google Drive (sync issues)
- include more complex models:
  - unique by: ds_key, taxa_id, time_period
  - pkey: mdl_seq
  - [ ] ∆ taxa -> taxa_id (unique worms_id, botw_id)
  - [ ] ∆ time_period -> time_type == `2019` (AquaMaps) -> "climatic annual"; need "climatic monthly" for GoMMAPS, plus annualized version
  - [ ] ∆ time_period == "month:January"? ISO?
  - [ ] + year_published: `2019` (AquaMaps)
  - [ ] ∆ ds_key -> ds_seq; eg "am_0.05" -> 1; allows for later changing name, year version, etc
- [ ] rename table `species` -> `dataset_taxon`
- [ ] `taxon`: all unique taxa cross-listed with identifiers

```{r}
#| label: setup

librarian::shelf(
  DBI,
  dbplyr,
  dplyr,
  duckdb,
  DT,
  flextable,
  fs,
  ftExtra,
  ggiraph,
  ggplot2,
  glue,
  gt,
  here,
  htmlwidgets,
  kableExtra,
  knitr,
  leaflet,
  logger,
  mapview,
  writexl,
  purrr,
  quarto,
  RColorBrewer,
  readr,
  scales,
  sf,
  shiny,
  stringr,
  terra,
  tibble,
  tidyr,
  units,
  webshot2,
  quiet = T
)
options(readr.show_col_types = F)

stopifnot(packageVersion("terra") >= "1.8.60")
# [rotate() issue](https://github.com/rspatial/terra/issues/1876)

source(here("libs/sdm_functions.R"))
source(here("libs/paths.R"))
dir_create(dir_v)
dir_create(dir_big_v)

# copy sdm.duckdb from prior version if missing
sdm_db_prev <- glue("{dir_big}/{ver_prev}/sdm.duckdb")
if (!file_exists(sdm_db) && file_exists(sdm_db_prev)) {
  message(glue("Copying sdm.duckdb from {ver_prev} to {ver}..."))
  file_copy(sdm_db_prev, sdm_db)
  message(glue("Copied sdm.duckdb ({file_size(sdm_db)})"))
}

# inputs (shared, unversioned)
pra_raw_gpkg <- glue("{dir_raw}/boem.gov/boem_program_areas_2025-11.gpkg")
cell_tif <- glue("{dir_derived}/r_bio-oracle_planarea.tif")
stopifnot(all(file_exists(c(cell_tif, pra_raw_gpkg))))

# version-independent input polygons (find from prior versions)
# note: search stable prior versions first; current dir_v files on Google Drive
#   can be evicted during long renders
find_input <- function(fname, search_dirs) {
  for (d in search_dirs) {
    f <- glue("{d}/{fname}")
    if (file_exists(f)) return(f)
  }
  stop(glue(
    "Input not found: {fname} (searched {paste(search_dirs, collapse=', ')})"
  ))
}
prev_dirs <- glue("{dir_derived}/{c(ver_prev, 'v2', 'v1')}")
er_raw_gpkg <- find_input("ply_ecoregions_2025.gpkg", prev_dirs)
pa_gpkg <- find_input("ply_planareas_2025.gpkg", prev_dirs)

# flags, typical
do_zones <- F # no, except if new spatial zones added
do_metrics <- T # YES
do_zone_taxon <- F # slow in past
do_postgis <- F # NO, since migrated to pmtiles
do_downloads <- T # YES
do_summary_data <- T # ecoregion x PRA intersections, label points, score tables
do_summary_figs <- T # maps, flower plots, primprod charts, Word/Excel tables
do_readme <- T # programmatic README.md + README.html
do_map <- F # no, but useful for visual checks during development
do_parquet <- T # no, but produces smaller faster duckdb
do_pmtiles <- F # generate PMTiles archives for mapgl/mapsp apps
do_pra_primprod <- T # NO, unless zones or primary prod ∆; zonal primary productivity for program areas
do_deploy <- T # sync derived data + app code to server

stopifnot(!do_pmtiles || (do_pmtiles && do_downloads)) # pmtiles should be generated from downloads

# spatial outputs (year reflects input data vintage, version suffix model version of derived metric columns)
pra_gpkg <- glue("{dir_v}/ply_programareas_2026_{ver}.gpkg")
sr_gpkg <- glue("{dir_v}/ply_subregions_2026_{ver}.gpkg")

# computed outputs (version suffix only — no obvious input year)
sr_pra_csv <- glue("{dir_v}/subregion_programareas.csv")
metrics_tif <- glue("{dir_v}/r_metrics_{ver}.tif")
metrics_pra_tif <- glue("{dir_v}/r_metrics_programareas_{ver}.tif")
zone_taxon_csv <- glue("{dir_v}/zone_taxon_{ver}.csv")
lyrs_csv <- glue("{dir_v}/layers_{ver}.csv")
# file_exists(c(sr_pra_csv, metrics_tif, metrics_pra_tif, zone_taxon_csv, lyrs_csv))

# summary outputs (merged from msens-summary_programareas.qmd)
er_pra_gpkg <- glue("{dir_v}/ply_ecoregion_programarea_{ver}.gpkg")
er_pra_csv <- glue("{dir_v}/ply_ecoregion_programarea_{ver}.csv")
er_pra_n_csv <- glue("{dir_v}/tbl_er_pra_scores_{ver}.csv")
pra_n_csv <- glue("{dir_v}/tbl_pra_scores_{ver}.csv")
lbls_csv <- glue("{dir_v}/ply_label_placement_pra.csv")
dir_figs <- here(glue("figs/msens-summary_programareas_{ver}"))
tbl_pra <- glue("ply_programareas_2026_{ver}")
tbl_er <- glue("ply_ecoregions_2025")
pp_csv <- glue("{dir_v}/programarea_primprod.csv")
pp_png <- glue("{dir_v}/programarea_primprod.png")
pp_pdf <- glue("{dir_v}/programarea_primprod.pdf")

# redo flags for summary sections
redo_summary_data <- T
redo_summary_figs <- T

# parquet (big files)
dir_pq <- glue("{dir_big_v}/sdm_parquet")

# helper functions ----
get_rast <- function(m_key) {
  d <- dbGetQuery(
    con_sdm,
    glue(
      "
        SELECT
          cm.cell_id,
          cm.value
        FROM cell_metric cm
        WHERE cm.metric_seq = (
          SELECT metric_seq
          FROM metric
          WHERE metric_key = '{m_key}' )"
    )
  )
  stopifnot(sum(duplicated(d$cell_id)) == 0)

  r <- init(r_cell[[1]], NA)
  r[d$cell_id] <- d$value

  r
}

plot_flower <- function(
  data,
  fld_category,
  fld_height,
  fld_width,
  colors,
  tooltip_expr = NULL,
  score = NULL,
  title = NULL,
  interactive = T
) {
  stopifnot(is.numeric(data |> pull({{ fld_height }})))
  stopifnot(is.numeric(data |> pull({{ fld_width }})))

  if (is.null(score)) {
    score <- data |>
      mutate(
        "{{fld_height}}" := as.double({{ fld_height }}),
        "{{fld_width}}" := as.double({{ fld_width }})
      ) |>
      summarize(
        score = weighted.mean({{ fld_height }}, {{ fld_width }}, na.rm = T)
      ) |>
      pull(score)
  }

  d <- data |>
    arrange({{ fld_category }}) |>
    mutate(across(!where(is.character), as.double)) |>
    mutate(
      ymax = cumsum({{ fld_width }}),
      ymin = lag(ymax, default = 0),
      xmax = {{ fld_height }},
      xmin = 0
    )

  sym_category <- ensym(fld_category)
  sym_height <- ensym(fld_height)
  sym_width <- ensym(fld_width)

  if (!is.null(tooltip_expr)) {
    d <- d |>
      mutate(tooltip = glue(tooltip_expr))
  } else {
    d <- d |>
      mutate(tooltip = glue("{!!fld_category}"))
  }

  g <- ggplot(d) +
    geom_rect_interactive(
      aes(
        xmin = xmin,
        xmax = xmax,
        ymin = ymin,
        ymax = ymax,
        fill = {{ fld_category }},
        color = "white",
        data_id = {{ fld_category }},
        tooltip = tooltip
      ),
      color = "white",
      alpha = 0.5
    ) +
    coord_polar(theta = "y") +
    xlim(c(-10, max(data |> pull({{ fld_height }})))) +
    annotate(
      "text",
      x = -10,
      y = 0,
      label = round(score),
      size = 8,
      fontface = "bold"
    ) +
    scale_fill_manual(values = colors) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(20, 20, 20, 20), "pt")
    )

  if (!is.null(title)) {
    g <- g + ggtitle(title)
  }

  if (interactive) {
    g <- girafe(
      ggobj = g,
      options = list(
        opts_sizing(rescale = TRUE, width = 1),
        opts_tooltip(
          css = "background-color:white;color:black;padding:5px;border-radius:3px;"
        )
      )
    )
  }

  g
}

# conditional summary setup ----
if (do_summary_figs) {
  mapbox_tkn_txt <- glue("{dir_private}/mapbox_token_bdbest.txt")
  Sys.setenv(MAPBOX_PUBLIC_TOKEN = readLines(mapbox_tkn_txt))
  librarian::shelf(mapgl, quiet = T)
  dir_create(dir_figs)
}

con_sdm <- dbConnect(duckdb(), dbdir = sdm_db, read_only = F)
# dbDisconnect(con_sdm, shutdown = T); duckdb_shutdown(duckdb()); rm(con_sdm)
if (do_postgis) {
  source(here("libs/db.R"))
} # generates Postgres `con`; fails if ssh-msens not connected

# clean up prior version zone references in DuckDB ----
zone_tbls_db <- tbl(con_sdm, "zone") |> distinct(tbl) |> pull(tbl)

# rename version-independent zones (ecoregions, planning areas) to remove old suffix
for (tbl_base in c("ply_ecoregions_2025", "ply_planareas_2025")) {
  old_tbl <- glue("{tbl_base}_{ver_prev}")
  if (old_tbl %in% zone_tbls_db) {
    dbExecute(
      con_sdm,
      glue("UPDATE zone SET tbl = '{tbl_base}' WHERE tbl = '{old_tbl}'")
    )
    message(glue("Renamed zone tbl: {old_tbl} -> {tbl_base}"))
  }
}

# rename versioned zones (program areas, subregions) from ver_prev to ver
for (old_tbl in zone_tbls_db[str_detect(zone_tbls_db, glue("_{ver_prev}$"))]) {
  new_tbl <- str_replace(old_tbl, glue("_{ver_prev}$"), glue("_{ver}"))
  dbExecute(
    con_sdm,
    glue("UPDATE zone SET tbl = '{new_tbl}' WHERE tbl = '{old_tbl}'")
  )
  # zone_taxon uses zone_tbl column
  dbExecute(
    con_sdm,
    glue(
      "UPDATE zone_taxon SET zone_tbl = '{new_tbl}' WHERE zone_tbl = '{old_tbl}'"
    )
  )
  message(glue("Renamed zone tbl: {old_tbl} -> {new_tbl}"))
}

sp_cats <- tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  distinct(sp_cat) |>
  arrange(sp_cat) |>
  pull(sp_cat)

# for visual checks
r_cell <- rast(cell_tif)
ext(r_cell) <- round(ext(r_cell), 3) # TODO: culprit with warlus wonk?
```


```{r}
#| label: check_scores
#| eval: false

old_csv <- glue("{dir_v}/tbl_pra_scores_v3_old-pre-mmpa-mbta-floor.csv")
out_csv <- glue("{dir_v}/tbl_pra_scores_v3_new-vs-old-pre-mmpa-mbta-floor.csv")
new_csv <- pra_n_csv

d_old <- read_csv(old_csv)
d_new <- read_csv(new_csv)
d_comp <- d_old |>
  full_join(
    d_new,
    by = join_by(`Program Area`),
    suffix = c("_old", "_new")
  ) |>
  mutate(
    Score_diff = Score_new - Score_old
  ) |>
  select(`Program Area`, Score_old, Score_new, Score_diff)
write_csv(d_comp, out_csv)
```


Flags for evaluating R chunks:

- `do_zones`: `r do_zones`
- `do_metrics`: `r do_metrics`
- `do_zone_taxon`: `r do_zone_taxon`
- `do_postgis`: `r do_postgis`
- `do_downloads`: `r do_downloads`
- `do_summary_data`: `r do_summary_data`
- `do_summary_figs`: `r do_summary_figs`
- `do_readme`: `r do_readme`
- `do_map`: `r do_map`
- `do_parquet`: `r do_parquet`
- `do_deploy`: `r do_deploy`
  
## Add Zones
  
### Add `cell` table to DuckDB if missing
  
```{r}
#| label: re-add_con_sdm
#| eval: !expr do_zones

if (!"cell" %in% dbListTables(con_sdm)) {
  if (!exists("r_cell")) {
    r_cell <- rast(cell_tif)
  }

  d_cell <- values(r_cell, dataframe = T, na.rm = T) |>
    tibble() |>
    mutate(
      cell_id = as.integer(cell_id)
    )

  # dbRemoveTable(con_sdm, "cell")
  dbWriteTable(con_sdm, "cell", d_cell, append = F)
  dbExecute(con_sdm, "ALTER TABLE cell ADD PRIMARY KEY (cell_id);")
}
```


### Add Program Areas to SDM zones

Add BOEM Program Areas (11th National Draft Proposed Program, 2025-11) as a new zone type.
Program Areas are derived from Planning Areas with additional GOA subdivisions.

```{r}
#| label: add_programareas
#| eval: !expr do_zones

zone_tbls <- tbl(con_sdm, "zone") |>
  distinct(tbl) |>
  pull(tbl)

if (!tbl_pra %in% zone_tbls) {
  source(here("libs/sdm_functions.R"))
  # TODO: move these into msens R
  # - set_geom_ctr_area()
  # - register_zones()
  # - insert_zone_cell_data()

  # load planning areas for field join
  pa <- read_sf(pa_gpkg) |>
    st_drop_geometry() |>
    select(planarea_name, region_key, planarea_key)

  # load and process program areas with field renaming
  pra <- read_sf(pra_raw_gpkg) |>
    rename(
      region_name = BOEM_OCS_REGION,
      planarea_name = PLANNING_AREA,
      programarea_name = Label,
      programarea_id = OBJECTID
    ) |>
    # left join to get region_key and planarea_key from planning areas
    left_join(pa, by = "planarea_name") |>
    # programarea_key defaults to planarea_key
    mutate(
      programarea_key = planarea_key
    ) |>
    # handle NA values for GOA Program Areas (not in planning areas)
    mutate(
      programarea_key = case_when(
        programarea_name == "GOA Program Area A" ~ "GAA",
        programarea_name == "GOA Program Area B" ~ "GAB",
        TRUE ~ programarea_key
      ),
      planarea_key = case_when(
        programarea_name == "GOA Program Area A" ~ "WGA,CGA,EGA",
        programarea_name == "GOA Program Area B" ~ "CGA,EGA",
        TRUE ~ planarea_key
      ),
      region_key = case_when(
        str_detect(programarea_name, "GOA") ~ "GA",
        TRUE ~ region_key
      )
    ) |>
    select(
      programarea_id,
      region_key,
      region_name,
      planarea_key,
      planarea_name,
      programarea_key,
      programarea_name
    ) |>
    st_transform(st_crs(4326))

  # add centroid and area
  pra <- pra |> set_geom_ctr_area()

  # write derived gpkg
  write_sf(pra, pra_gpkg)
  message(glue("Wrote {nrow(pra)} program areas to {pra_gpkg}"))
  # Wrote 20 program areas to ~/My Drive/projects/msens/data/derived/ply_programareas_2026.gpkg

  # shift lon [-180, 180] to [0, 360]
  pra_sl <- st_shift_longitude(pra)

  # load cell raster template
  r_cell <- rast(cell_tif, lyrs = "cell_id")

  # get percent each feature covers each cell
  r_pra <- rasterize(
    pra_sl,
    r_cell,
    fun = "max",
    touches = T,
    cover = T,
    by = "programarea_key"
  )

  # convert to data frame
  d_pra <- terra::as.data.frame(
    r_pra,
    cells = T,
    na.rm = F,
    wide = F
  ) |>
    filter(!is.na(values)) |>
    tibble()

  message(glue("Program areas cover {length(unique(d_pra$cell))} cells"))

  # register zones and insert zone data into con_sdm
  register_zones(pra, tbl_pra, "programarea_key")
  insert_zone_cell_data(d_pra, tbl_pra, "programarea_key")

  # verify registration
  tbl(con_sdm, "zone") |>
    group_by(tbl) |>
    summarise(n_zones = n()) |>
    collect() |>
    print()
}
```

### Create ply_programareas_2026 in PostGIS database (deprecated)

Deprecated: migrated to PMTiles. Kept for reference.

```{r}
#| label: create_pg_ply_programareas_2026
#| eval: !expr do_postgis

# devtools::install_local("../msens")
librarian::shelf(
  here,
  leaflet,
  leaflet.extras, # mapview,
  marinesensitivity / msens,
  sf,
  tibble,
  quiet = T
)

source(here("libs/db.R")) # define: con

# load from gpkg if not already in memory (eg zones already existed)
if (!exists("pra")) {
  pra <- read_sf(pra_gpkg)
}

pra |>
  st_write(
    dsn = con,
    layer = tbl_pra,
    driver = "PostgreSQL",
    layer_options = "OVERWRITE=true"
  )

# TODO: need to convert to 1st column to integer for mapgl viewing?

# enforce SRID so shows up in tile.marinesensitivity.org
dbExecute(
  con,
  glue(
    "SELECT UpdateGeometrySRID('{tbl_pra}','geom',4326);"
  )
)

# fix any invalid geometries
dbExecute(
  con,
  glue(
    "UPDATE {tbl_pra} SET geom = ST_MakeValid(geom) WHERE NOT ST_IsValid(geom)"
  )
) # 6 rows fixed

# add spatial index for faster queries
dbExecute(
  con,
  glue(
    "CREATE INDEX IF NOT EXISTS {tbl_pra}_geom_idx ON {tbl_pra} USING gist(geom);"
  )
)
```

### Create Subregion-ProgramArea Lookup

Create lookup table for which program areas belong to which subregions.

Used in dropdown of mapgl app.

```{r}
#| label: subregion_programareas_lookup
#| eval: !expr do_zones

# calculate intersections between subregions and program areas using zone_cell
d_sr_pra <- tbl(con_sdm, "zone_cell") |>
  inner_join(
    tbl(con_sdm, "zone") |>
      filter(tbl == !!tbl_sr),
    by = "zone_seq"
  ) |>
  rename(subregion_key = value) |>
  select(cell_id, subregion_key) |>
  inner_join(
    tbl(con_sdm, "zone_cell") |>
      inner_join(
        tbl(con_sdm, "zone") |>
          filter(tbl == !!tbl_pra),
        by = "zone_seq"
      ) |>
      select(cell_id, programarea_key = value),
    by = "cell_id"
  ) |>
  distinct(subregion_key, programarea_key) |>
  collect()


get_sr_bbox <- function(sr_key) {
  pra_sr <- d_sr_pra |>
    filter(subregion_key == !!sr_key) |>
    pull(programarea_key)
  r <- init(r_cell[[1]], NA)
  r_sr <- r_metrics[["programarea_key"]] %in% pra_sr
  trim(r_sr) |> st_bbox() |> as.numeric()
}
# TODO: pre-calc bboxes for all subregions

write_csv(d_sr_pra, sr_pra_csv)
message(glue(
  "Wrote {nrow(d_sr_pra)} subregion-programarea mappings to {sr_pra_csv}"
))
# Wrote 60 subregion-programarea mappings to ~/My Drive/projects/msens/data/derived/subregion_programareas.csv
```

### Add SubRegions

Original:

- `USA`: USA mainland, Alaska, Hawaii and territories
- `AK`: Alaska
- `L48`: USA mainland
- `AKL48`: USA mainland & Alaska

Conceived for later:

- `HI`: Hawaii
- `L48Pac`: Pacific USA mainland
- `L48Atl`: Atlantic USA mainland
- `L48Goa`: Gulf of America USA mainland
- `PAC`: Pacific USA (Hawaii, Pacific mainland and USA Territories)

version for 2026:

- `USA`: USA mainland and Alaska
- `AK`: Alaska
- `L48Pac`: Pacific USA mainland
- `L48Atl`: Atlantic USA mainland
- `L48Goa`: Gulf of America USA mainland

v3:

- `USA`: All USA (Alaska, Gulf of America and Pacific)
- `AK`: Alaska
- `GA`: Gulf of America
- `PA`: Pacific

```{r}
#| label: add_subregions_2026
#| eval: !expr do_zones

zones <- tbl(con_sdm, "zone") |>
  distinct(tbl) |>
  pull(tbl)

# delete existing tbl_sr
if (tbl_sr %in% zones) {
  dbExecute(con_sdm, glue("DELETE FROM zone WHERE tbl = '{tbl_sr}'"))
  # dbExecute(con_sdm, "DELETE FROM zone_cell WHERE zone_seq NOT IN (SELECT zone_seq FROM zone)")
}

d_sr <- tribble(
  ~subregion_key , ~subregion_name   ,
  "AK"           , "Alaska"          ,
  "GA"           , "Gulf of America" ,
  "PA"           , "Pacific"         ,
  "USA"          , "All USA"
) # note: subregion_name not used (yet)
register_zones(d_sr, tbl_sr, "subregion_key")

# load pra from gpkg if not already in memory
if (!exists("pra")) {
  if (file_exists(pra_gpkg)) {
    pra <- read_sf(pra_gpkg)
  } else {
    pra_prev <- glue(
      "{dir_derived}/{ver_prev}/ply_programareas_2026_{ver_prev}.gpkg"
    )
    pra <- read_sf(pra_prev)
  }
}

# temporarily push tmp_pra into duckdb for later joining
tmp_pra <- pra |>
  st_drop_geometry() |>
  select(programarea_key, region_key)
copy_to(con_sdm, tmp_pra, "tmp_pra", temporary = T, overwrite = T)

df_z_sr_notusa <- tbl(con_sdm, "zone") |>
  filter(
    tbl == !!tbl_sr,
    value != "USA"
  ) |>
  select(region_key = value, zone_seq_sr = zone_seq) |>
  left_join(
    tbl(con_sdm, "tmp_pra"),
    by = "region_key"
  ) |>
  left_join(
    tbl(con_sdm, "zone") |>
      filter(tbl == !!tbl_pra) |>
      select(
        programarea_key = value,
        zone_seq
      ),
    by = "programarea_key"
  ) |>
  left_join(
    tbl(con_sdm, "zone_cell"),
    by = "zone_seq"
  ) |>
  group_by(zone_seq_sr, cell_id) |>
  summarize(
    pct_covered = sum(pct_covered, na.rm = T),
    .groups = "drop"
  ) |>
  mutate(
    pct_covered = pmin(pct_covered, 100, na.rm = T)
  ) |>
  select(
    zone_seq = zone_seq_sr,
    cell_id,
    pct_covered
  ) |>
  collect() # 349,139 × 3
# summary(df_z_sr_notusa)

# duplicated(df_z_sr_notusa$cell_id) |> any() # FALSE
# TODO: check why no overlapping cells across subregions?

# zone_seqs_sr <- tbl(con_sdm, "zone") |>
#   filter(
#     tbl == !!tbl_sr) |>
#   pull(zone_seq)
# dbExecute(con_sdm, glue("
#   DELETE FROM zone_cell
#   WHERE zone_seq IN ({paste(zone_seqs_sr, collapse = ',')})"))

# append USA, ie all cells in other subregions
zone_seq_USA <- tbl(con_sdm, "zone") |>
  filter(
    tbl == !!tbl_sr,
    value == "USA"
  ) |>
  pull(zone_seq) # 217

df_z_sr_usa <- df_z_sr_notusa |>
  group_by(cell_id) |>
  summarize(
    pct_covered = sum(pct_covered, na.rm = T),
    .groups = "drop"
  ) |>
  mutate(
    zone_seq = !!zone_seq_USA,
    pct_covered = pmin(pct_covered, 100, na.rm = T)
  ) # 349,139 × 3
# summary(df_z_sr_usa)

df_z_sr <- bind_rows(
  df_z_sr_notusa,
  df_z_sr_usa
)

# TODO: ....
copy_to(
  dest = con_sdm,
  df = df_z_sr,
  name = "zone_cell",
  append = T
)

tbl(con_sdm, "zone") |>
  filter(tbl == !!tbl_sr) |>
  select(zone_seq, subregion_key = value) |>
  left_join(
    tbl(con_sdm, "zone_cell"),
    by = "zone_seq"
  ) |>
  group_by(subregion_key) |>
  summarize(n_cell = n()) |>
  arrange(subregion_key)
#   subregion_key n_cell
#   <chr>          <dbl>
# 1 AK            307,445
# 2 GA             17,274
# 3 PA             24,420
# 4 USA           349,139
```


```{r}
#| label: old_add_subregions_2025
#| eval: false

zone_tbls <- tbl(con_sdm, "zone") |>
  distinct(tbl) |>
  pull(tbl)

if (!"ply_subregions_2025" %in% zone_tbls) {
  source(here("libs/sdm_functions.R"))
  # - set_geom_ctr_area()
  # - register_zones()
  # - insert_zone_cell_data()

  # spatial sources
  er <- read_sf(er_raw_gpkg)
  r_cell <- rast(cell_tif, lyrs = "cell_id")

  # USA: USA mainland, Alaska, Hawaii and territories
  sr_usa <- er |>
    st_union() |>
    st_as_sf() |>
    mutate(
      subregion_key = "USA",
      subregion_name = "USA mainland, Alaska, Hawaii and territories"
    )

  # AK: Alaska
  sr_ak <- er |>
    filter(region_key == "AK") |>
    st_union() |>
    st_as_sf() |>
    mutate(
      subregion_key = "AK",
      subregion_name = "Alaska"
    )

  # L48: USA mainland
  sr_l48 <- er |>
    filter(
      ecoregion_key %in%
        c(
          "CAC",
          "WAOR",
          "SECS",
          "WCGOA",
          "EGOA",
          "NECS"
        )
    ) |>
    st_union() |>
    st_as_sf() |>
    mutate(
      subregion_key = "L48",
      subregion_name = "USA mainland"
    )

  # AKL48: USA mainland & Alaska
  sr_akl48 <- bind_rows(sr_ak, sr_l48) |>
    st_union() |>
    st_as_sf() |>
    mutate(
      subregion_key = "AKL48",
      subregion_name = "USA mainland & Alaska"
    )

  # combine
  sr <- bind_rows(
    sr_ak,
    sr_l48,
    sr_akl48,
    sr_usa
  ) |>
    set_geom_ctr_area()

  # write gpkg
  write_sf(sr, sr_gpkg)

  # shift lon [-180, 180] to [0, 360]
  sr_sl <- st_shift_longitude(sr)

  # get percent each feature covers each cell
  r_sr <- rasterize(
    sr_sl,
    r_cell,
    fun = "max",
    touches = T,
    cover = T,
    by = "subregion_key"
  )

  # convert to data frame
  d_sr <- terra::as.data.frame(
    r_sr,
    cells = T,
    na.rm = F,
    wide = F
  ) |>
    filter(
      !is.na(values)
    ) |> # 839,534 × 3
    tibble()

  # d_sr |>
  #   group_by(layer) |>
  #   summarise(n_cells  = n()) |>
  #   collect() |>
  #   arrange(desc(n_cells))
  #
  #   layer  n_cells
  #   <chr>    <int>
  # 1 USA    630,806
  # 2 AKL48  419,767
  # 3 AK     315,369
  # 4 L48    104,398

  # register zones and insert zone data into con_sdm
  register_zones(sr, "ply_subregions_2025", "subregion_key")
  insert_zone_cell_data(d_sr, "ply_subregions_2025", "subregion_key")

  # tbl(con_sdm, "zone") |>
  #   group_by(tbl) |>
  #   summarise(n_zones = n()) |>
  #   collect()
  #
  #   tbl                 n_zones
  #   <chr>                 <dbl>
  # 1 ply_ecoregions_2025      12
  # 2 ply_planareas_2025       36
  # 3 ply_subregions_2025       4

  # TODO: load ply_subregions_2025 into PostGIS
}
```

## Calculate Metrics

### Calculate Extinction Risk Metrics by Taxonomic Component

```{r}
#| label: delete_extrisk_metrics
#| eval: !expr do_metrics

# Start fresh with all extrisk_* metrics
existing_seqs <- tbl(con_sdm, "metric") |>
  filter(
    str_starts(metric_key, "extrisk_")
  ) |>
  pull(metric_seq)
if (length(existing_seqs) > 0) {
  metric_sql <- paste(existing_seqs, collapse = ", ")
  dbExecute(
    con_sdm,
    glue("DELETE FROM cell_metric WHERE metric_seq IN ({metric_sql})")
  )
  dbExecute(
    con_sdm,
    glue("DELETE FROM zone_metric WHERE metric_seq IN ({metric_sql})")
  )
  dbExecute(
    con_sdm,
    glue("DELETE FROM metric WHERE metric_seq IN ({metric_sql})")
  )
}
```

```{r}
#| label: calc_cell_metric_redlist
#| eval: !expr do_metrics

d_taxon_ok <- tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  collect()

for (sp_cat in sp_cats) {
  # sp_cat = sp_cats[1]

  message(glue("Calculating extinction risk metrics for {sp_cat}..."))

  metric_key <- glue("extrisk_{str_replace(sp_cat, ' ', '_')}")
  description <- glue("Extinction risk for {sp_cat}")
  # TODO: rename metric_key -> metric_key, description

  # Get metric_seq first, then delete/insert
  existing_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT metric_seq FROM metric WHERE metric_key = '{metric_key}'"
    )
  )

  if (nrow(existing_seq) > 0) {
    metric_seq <- existing_seq$metric_seq
    dbExecute(
      con_sdm,
      glue("DELETE FROM cell_metric WHERE metric_seq = {metric_seq}")
    )
    dbExecute(
      con_sdm,
      glue("DELETE FROM zone_metric WHERE metric_seq = {metric_seq}")
    )
    dbExecute(
      con_sdm,
      glue("DELETE FROM metric WHERE metric_seq = {metric_seq}")
    )
  }

  dbExecute(
    con_sdm,
    glue(
      "
    INSERT INTO metric (metric_key, description)
    VALUES ('{metric_key}', '{description}')"
    )
  )

  metric_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT max(metric_seq) AS metric_seq FROM metric WHERE metric_key = '{metric_key}'"
    )
  ) |>
    pull(metric_seq)

  # use er_score from taxon (pre-computed in ingest_nmfs-fws-mmpa-mbta-listings.qmd
  # and merged in merge_models.qmd)
  # for is_er_spatial species (turtles), ER is already baked into cell values
  # so set er_score = 100 to avoid double-counting (100 * value / 100 = value)
  d_er <- tbl(con_sdm, "taxon") |>
    filter(
      is_ok,
      if (!!sp_cat == "all") T else sp_cat == !!sp_cat
    ) |>
    select(mdl_seq, er_score, is_er_spatial) |>
    collect() |>
    mutate(
      er_score = ifelse(is_er_spatial, 100, er_score)
    ) |>
    select(mdl_seq, er_score)

  duckdb_register(con_sdm, "tmp_er", d_er |> select(mdl_seq, er_score))
  dbExecute(
    con_sdm,
    glue(
      "
    INSERT INTO cell_metric (cell_id, metric_seq, value)
    SELECT
      mc.cell_id,
      {metric_seq} AS metric_seq,
      ROUND(SUM(er.er_score * mc.value) / 100.0, 2) AS value
    FROM model_cell mc
    JOIN tmp_er er ON mc.mdl_seq = er.mdl_seq
    GROUP BY mc.cell_id"
    )
  ) # 6,662,075
  duckdb_unregister(con_sdm, "tmp_er")

  # visual check ----
  # d <- tbl(con_sdm, "cell_metric") |>
  #   filter(metric_seq == metric_seq) |>
  #   select(cell_id, value) |>
  #   collect()
  # r <- init(r_cell[[1]], NA)
  # r[d$cell_id] <- d$value
  # r_r <- rotate(r)
  # plet(r_r)
}

tbl(con_sdm, "metric") |>
  filter(str_starts(metric_key, "extrisk_")) |>
  arrange(metric_seq) |>
  select(-date_created) |>
  collect() |>
  print(n = Inf)
```

### Calculate ExtinctionRisk min/max per SpeciesCategory and Ecoregion

```{r}
#| label: calc_ecoregion_minmax
#| eval: !expr do_metrics

for (sp_cat in sp_cats) {
  # sp_cat = sp_cats[1]
  for (fxn in c("min", "max")) {
    # fxn = "min"

    m_key <- glue("extrisk_{sp_cat}_ecoregion_{fxn}")
    m_description <- glue("Extinction risk for {sp_cat} Ecoregional {fxn}")
    message(glue("m_key: {m_key}"))

    # Get metric_seq first, then delete/insert
    existing_seq <- dbGetQuery(
      con_sdm,
      glue(
        "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
      )
    ) |>
      pull(metric_seq)
    if (length(existing_seq) > 0) {
      dbExecute(
        con_sdm,
        glue("DELETE FROM metric      WHERE metric_seq = {existing_seq}")
      )
      dbExecute(
        con_sdm,
        glue("DELETE FROM cell_metric WHERE metric_seq = {existing_seq}")
      )
      dbExecute(
        con_sdm,
        glue("DELETE FROM zone_metric WHERE metric_seq = {existing_seq}")
      )
    }

    dbExecute(
      con_sdm,
      glue(
        "INSERT INTO metric (metric_key, description)
      VALUES ('{m_key}', '{m_description}')"
      )
    )

    # get metric ID
    m_seq <- dbGetQuery(
      con_sdm,
      glue(
        "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
      )
    ) |>
      pull(metric_seq)

    # calculate min/max extinction risk per ecoregion
    dbExecute(
      con_sdm,
      glue(
        "INSERT INTO zone_metric (zone_seq, metric_seq, value)
      SELECT 
        z.zone_seq,
        -- z.tbl, z.fld, z.value AS tbl_fld_val,
        {m_seq} as metric_seq,
        {fxn}(cm.value) as value
      FROM zone z
      LEFT JOIN zone_cell zc ON z.zone_seq = zc.zone_seq
      LEFT JOIN cell_metric cm ON zc.cell_id = cm.cell_id
      WHERE
        z.fld = 'ecoregion_key'
        AND cm.metric_seq IN (
          SELECT metric_seq FROM metric WHERE metric_key = 'extrisk_{sp_cat}')
      -- GROUP BY z.zone_seq, z.tbl, z.fld, z.value
      GROUP BY z.zone_seq "
      )
    )
  }
}

tbl(con_sdm, "metric") |>
  filter(str_detect(metric_key, "^extrisk_.*_ecoregion_(min|max)$")) |>
  arrange(metric_seq) |>
  select(-date_created) |>
  collect() |>
  print(n = Inf)
```

### Calculate rescaled ExtinctionRisk per cell based on EcoregionMinMax per SpeciesCategory

```{r}
#| label: extrisk_spcat_ecoregion_rescaled_cell
#| eval: !expr do_metrics

for (sp_cat in sp_cats) {
  # sp_cat = sp_cats[1]

  m_key <- glue("extrisk_{sp_cat}_ecoregion_rescaled")
  m_description <- glue(
    "Extinction risk for {sp_cat}, rescaled to [0,100] based on Ecoregional min/max values"
  )

  # Get metric_seq first, then delete/insert
  existing_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
    )
  )
  if (nrow(existing_seq) > 0) {
    m_seq <- existing_seq$metric_seq
    dbExecute(
      con_sdm,
      glue("DELETE FROM metric      WHERE metric_seq = {m_seq}")
    )
    dbExecute(
      con_sdm,
      glue("DELETE FROM cell_metric WHERE metric_seq = {m_seq}")
    )
    dbExecute(
      con_sdm,
      glue("DELETE FROM zone_metric WHERE metric_seq = {m_seq}")
    )
  }

  # insert new metric
  dbExecute(
    con_sdm,
    glue(
      "
      INSERT INTO metric (metric_key, description)
      VALUES ('{m_key}', '{m_description}')"
    )
  )
  m_seq <- dbGetQuery(
    con_sdm,
    glue(
      "
      SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
    )
  ) |>
    pull(metric_seq)

  dbExecute(
    con_sdm,
    glue(
      "
      WITH
      -- get ecoregion_minmax from zone_metric
      ecoregion_minmax AS (
        SELECT 
          z.zone_seq,
          MIN(zm.value) as min_value,
          MAX(zm.value) as max_value
        FROM zone z
        JOIN zone_metric zm USING (zone_seq)
        WHERE 
          z.fld = 'ecoregion_key' AND 
          zm.metric_seq IN (
            SELECT metric_seq FROM metric WHERE metric_key IN (
              'extrisk_{sp_cat}_ecoregion_min',
              'extrisk_{sp_cat}_ecoregion_max') )
        GROUP BY z.zone_seq
      ),
      -- get cell_ecoregion from zone_cell with pct_covered
      cell_ecoregion_raw AS (
        SELECT 
          cell_id, 
          z.zone_seq, 
          zone_cell.pct_covered
        FROM zone_cell
        JOIN zone z ON zone_cell.zone_seq = z.zone_seq
        WHERE z.fld = 'ecoregion_key'
      ),
      -- calculate total coverage per cell for normalization
      cell_coverage_total AS (
        SELECT 
          cell_id, 
          SUM(pct_covered) as total_coverage
        FROM cell_ecoregion_raw
        GROUP BY cell_id
      ),
      -- normalize coverage percentages
      -- for cells with only one ecoregion, use 100% regardless of actual coverage
      -- for cells with multiple ecoregions, rescale to ensure percentages sum to 100
      cell_ecoregion AS (
        SELECT 
          ce.cell_id, 
          ce.zone_seq,
          CASE 
            WHEN COUNT(*) OVER (PARTITION BY ce.cell_id) = 1 THEN 100
            ELSE ce.pct_covered * 100.0 / ct.total_coverage
          END as normalized_pct
        FROM cell_ecoregion_raw ce
        JOIN cell_coverage_total ct ON ce.cell_id = ct.cell_id
      )
      -- insert rescaled extinction risk values
      INSERT INTO cell_metric (cell_id, metric_seq, value)
      SELECT 
        cm.cell_id,
        {m_seq} as metric_seq,
        SUM(
          (cm.value     - er.min_value) / 
          (er.max_value - er.min_value) * 
          (ce.normalized_pct / 100.0) ) * 100 AS value  -- scale to 0-100 range
      FROM cell_metric      cm
      JOIN cell_ecoregion   ce ON cm.cell_id  = ce.cell_id
      JOIN ecoregion_minmax er ON ce.zone_seq = er.zone_seq
      WHERE cm.metric_seq IN (
        SELECT metric_seq FROM metric WHERE metric_key = 'extrisk_{sp_cat}')
      AND cm.value IS NOT NULL
      AND er.min_value IS NOT NULL
      AND er.max_value IS NOT NULL
      AND er.min_value < er.max_value  -- avoid division by zero
      GROUP BY cm.cell_id
    "
    )
  )
}

tbl(con_sdm, "metric") |>
  filter(str_detect(metric_key, "^extrisk_.*_ecoregion_rescaled$")) |>
  arrange(metric_seq) |>
  select(-date_created) |>
  collect() |>
  print(n = Inf)
```


### Transfer raw PrimProd into cell_metric

```{r}
#| label: primprod_cell_metric
#| eval: !expr do_metrics

# Create metric for primary productivity
m_key <- "primprod"
m_description <- "Primary productivity: Oregon State Vertically Generalized Production Model (VGPM) from Visible Infrared Imaging Radiometer Suite (VIIRS) satellite data (mg C / m^2 / day) from daily averages available as monthly averaged to annual and averaged to overall for the most recently available full years of data 2014 to 2023"
# TODO: consider log rescaling PrimaryProductivity

# workflow: ingest_productivity.qmd
dir_vgpm <- glue("{dir_data}/raw/oregonstate.edu")
vgpm_tif <- glue(
  "{dir_vgpm}/vgpm.r2022.v.chl.v.sst.2160x4320_2014-2023.avg.sd.tif"
)

r_cell <- rast(cell_tif) |> subset("cell_id")
ext(r_cell) <- round(ext(r_cell), 3)
# r_cell_r <- rotate(r_cell)
# ext(r_cell_r) <- round(ext(r_cell_r), 3)

r_vgpm <- rast(vgpm_tif) |> subset("npp_avg")
ext(r_vgpm) <- round(ext(r_vgpm), 3)
# r_vgpm_rc <- rotate(r_vgpm) |>
#   crop(r_cell) |>
#   mask(r_cell)

r_v_c <- resample(
  r_vgpm,
  r_cell,
  method = "bilinear"
) |>
  mask(r_cell) |>
  crop(r_cell)
# plot(r_v_c)
r_cv <- c(r_cell, r_v_c)

# Check if metric exists and delete if it does
existing_seq <- dbGetQuery(
  con_sdm,
  glue(
    "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
  )
)
if (nrow(existing_seq) > 0) {
  m_seq <- existing_seq$metric_seq
  dbExecute(con_sdm, glue("DELETE FROM metric      WHERE metric_seq = {m_seq}"))
  dbExecute(con_sdm, glue("DELETE FROM cell_metric WHERE metric_seq = {m_seq}"))
  dbExecute(con_sdm, glue("DELETE FROM zone_metric WHERE metric_seq = {m_seq}"))
}

# Create new metric
dbExecute(
  con_sdm,
  glue(
    "
    INSERT INTO metric (metric_key, description)
    VALUES ('{m_key}', '{m_description}')"
  )
)
m_seq <- dbGetQuery(
  con_sdm,
  glue(
    "
    SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
  )
) |>
  pull(metric_seq)

d_cv <- values(r_cv, dataframe = T, na.rm = T) |>
  tibble() |>
  rename(
    value = npp_avg
  ) |>
  mutate(
    cell_id = as.integer(cell_id),
    metric_seq = !!m_seq
  )

dbWriteTable(con_sdm, "cell_metric", d_cv, append = T)

tbl(con_sdm, "metric") |>
  filter(str_detect(metric_key, "^primprod")) |>
  arrange(metric_seq) |>
  select(-date_created) |>
  collect() |>
  print(n = Inf)
#   metric_seq metric_key description
#        <int> <chr>      <chr>
# 1        160 primprod   Primary productivity: Oregon State Vertically Generali…

# visual check ----
# d <- tbl(con_sdm, "cell_metric") |>
#   filter(metric_seq == m_seq) |>
#   select(cell_id, value) |>
#   collect()
# r <- init(r_cell[[1]], NA)
# r[d$cell_id] <- d$value
# r_r <- rotate(r)
# plet(r_r)
```


### Calculate PrimProd min/max by Ecoregion

```{r}
#| label: primprod_ecoregion_minmax
#| eval: !expr do_metrics

pp_metric <- "primprod"
pp_seq <- dbGetQuery(
  con_sdm,
  glue(
    "SELECT metric_seq FROM metric WHERE metric_key = '{pp_metric}'"
  )
) |>
  pull(metric_seq)

for (fxn in c("min", "max")) {
  # fxn = "min"

  # get existing primprod ecoregion min/max metric_seq
  m_key <- glue("primprod_ecoregion_{fxn}")
  m_description <- glue("Primary productivity Ecoregional {fxn}")

  # Check if metric exists and delete if it does
  existing_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
    )
  )
  if (nrow(existing_seq) > 0) {
    m_seq <- existing_seq$metric_seq
    dbExecute(
      con_sdm,
      glue("DELETE FROM metric      WHERE metric_seq = {m_seq}")
    )
    dbExecute(
      con_sdm,
      glue("DELETE FROM cell_metric WHERE metric_seq = {m_seq}")
    )
    dbExecute(
      con_sdm,
      glue("DELETE FROM zone_metric WHERE metric_seq = {m_seq}")
    )
  }

  # insert new metric
  dbExecute(
    con_sdm,
    glue(
      "INSERT INTO metric (metric_key, description) VALUES ('{m_key}', '{m_description}')"
    )
  )

  # get metric_seq
  m_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
    )
  ) |>
    pull(metric_seq)

  # Calculate min/max primary productivity per ecoregion
  dbExecute(
    con_sdm,
    glue(
      "INSERT INTO zone_metric (zone_seq, metric_seq, value)
    SELECT 
      z.zone_seq,
      -- z.tbl, z.fld, z.value AS tbl_fld_val,
      {m_seq} as metric_seq,
      {fxn}(cm.value) as value
    FROM      zone        z
    LEFT JOIN zone_cell   zc ON z.zone_seq = zc.zone_seq
    LEFT JOIN cell_metric cm ON zc.cell_id = cm.cell_id
    WHERE
      z.fld         = 'ecoregion_key' AND
      cm.metric_seq = {pp_seq}
    GROUP BY z.zone_seq"
    )
  )
}
tbl(con_sdm, "metric") |>
  filter(
    metric_key == "primprod_ecoregion_min"
  ) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "metric_seq"
  ) |>
  pull(value) |>
  summary()
# OLD:
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 65.61  130.50  237.31  247.02  358.66  435.62
# NEW:
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 35.15  135.08  267.47  257.26  366.07  454.43
tbl(con_sdm, "metric") |>
  filter(
    metric_key == "primprod_ecoregion_max"
  ) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "metric_seq"
  ) |>
  pull(value) |>
  summary()
```

### Calculate PrimProd avg/sd by Planning Area

```{r}
#| label: primprod_ecoregion_avg_stddev
#| eval: !expr do_metrics

pp_metric <- "primprod"
pp_seq <- dbGetQuery(
  con_sdm,
  glue(
    "SELECT metric_seq FROM metric WHERE metric_key = '{pp_metric}'"
  )
) |>
  pull(metric_seq)

for (fxn in c("avg", "stddev")) {
  # fxn = "avg"

  # get existing primprod avg/stddev metric_seq
  m_key <- glue("primprod_{fxn}")
  m_description <- glue("Primary productivity {fxn}")

  # Check if metric exists and delete if it does
  existing_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
    )
  ) |>
    pull(metric_seq)
  if (length(existing_seq) > 0) {
    dbExecute(
      con_sdm,
      glue(
        "DELETE FROM metric      WHERE metric_seq IN ({paste(existing_seq, collapse = ',')})"
      )
    )
    dbExecute(
      con_sdm,
      glue(
        "DELETE FROM cell_metric WHERE metric_seq IN ({paste(existing_seq, collapse = ',')})"
      )
    )
    dbExecute(
      con_sdm,
      glue(
        "DELETE FROM zone_metric WHERE metric_seq IN ({paste(existing_seq, collapse = ',')})"
      )
    )
  }

  # insert new metric
  dbExecute(
    con_sdm,
    glue(
      "INSERT INTO metric (metric_key, description) VALUES ('{m_key}', '{m_description}')"
    )
  )

  # get metric_seq
  m_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
    )
  ) |>
    pull(metric_seq)

  stopifnot(length(m_seq) == 1)

  # calculate avg/stddev primary productivity per planning area
  # dbGetQuery(con_sdm, glue(
  #   "SELECT
  dbExecute(
    con_sdm,
    glue(
      "INSERT INTO zone_metric (zone_seq, metric_seq, value)
     SELECT
      z.zone_seq,
      {m_seq} as metric_seq,
      {fxn}(cm.value) as value
    FROM      zone        z
    LEFT JOIN zone_cell   zc ON z.zone_seq = zc.zone_seq
    LEFT JOIN cell_metric cm ON zc.cell_id = cm.cell_id
    WHERE
      z.fld         = 'planarea_key' AND
      cm.metric_seq = {pp_seq}
    GROUP BY z.zone_seq"
    )
  )
}

# tbl(con_sdm, "cell_metric") |>
#   filter(
#     metric_seq == !!pp_seq) |>
#   pull(value) |>
#   summary()

tbl(con_sdm, "metric") |>
  filter(
    metric_key == "primprod_avg"
  ) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "metric_seq"
  ) |>
  pull(value) |>
  summary()
```

### Calculate rescaled PrimProd per cell based on EcoregionMinMax

```{r}
#| label: primprod_ecoregion_rescaled
#| eval: !expr do_metrics

pp_metric <- "primprod"
pp_seq <- dbGetQuery(
  con_sdm,
  glue(
    "SELECT metric_seq FROM metric WHERE metric_key = '{pp_metric}'"
  )
) |>
  pull(metric_seq)

# Create new metric for rescaled primary productivity
m_key <- "primprod_ecoregion_rescaled"
m_description <- "Primary productivity rescaled to [0,100] based on Ecoregional min/max values"
# TODO: consider log rescaling PrimaryProductivity

# Check if metric exists and delete if it does
existing_seq <- dbGetQuery(
  con_sdm,
  glue(
    "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
  )
) |>
  pull(metric_seq)
if (length(existing_seq) > 0) {
  dbExecute(
    con_sdm,
    glue("DELETE FROM metric      WHERE metric_seq = {existing_seq}")
  )
  dbExecute(
    con_sdm,
    glue("DELETE FROM cell_metric WHERE metric_seq = {existing_seq}")
  )
  dbExecute(
    con_sdm,
    glue("DELETE FROM zone_metric WHERE metric_seq = {existing_seq}")
  )
}

# Create new metric
dbExecute(
  con_sdm,
  glue(
    "
    INSERT INTO metric (metric_key, description)
    VALUES ('{m_key}', '{m_description}')"
  )
)
m_seq <- dbGetQuery(
  con_sdm,
  glue(
    "
    SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
  )
) |>
  pull(metric_seq)

# Rescale primary productivity by ecoregion
# dbGetQuery(con_sdm, glue("
dbExecute(
  con_sdm,
  glue(
    "
    WITH
    -- get ecoregion_minmax from cell_metric for primary productivity
    ecoregion_minmax AS (
      SELECT 
        z.zone_seq,
        MIN(zm.value) as min_value,
        MAX(zm.value) as max_value
      FROM zone z
      JOIN zone_metric zm USING (zone_seq)
      WHERE 
        z.fld = 'ecoregion_key' AND 
        zm.metric_seq IN (
          SELECT metric_seq FROM metric WHERE metric_key IN (
            'primprod_ecoregion_min',
            'primprod_ecoregion_max') )
      GROUP BY z.zone_seq ),
    -- get cell_ecoregion from zone_cell with pct_covered
    cell_ecoregion_raw AS (
      SELECT 
        cell_id, 
        z.zone_seq, 
        zone_cell.pct_covered
      FROM zone_cell
      JOIN zone z ON zone_cell.zone_seq = z.zone_seq
      WHERE z.fld = 'ecoregion_key'
    ),
    -- calculate total coverage per cell for normalization
    cell_coverage_total AS (
      SELECT 
        cell_id, 
        SUM(pct_covered) as total_coverage
      FROM cell_ecoregion_raw
      GROUP BY cell_id
    ),
    -- normalize coverage percentages
    cell_ecoregion AS (
      SELECT 
        ce.cell_id, 
        ce.zone_seq,
        CASE 
          WHEN COUNT(*) OVER (PARTITION BY ce.cell_id) = 1 THEN 100
          ELSE ce.pct_covered * 100.0 / ct.total_coverage
        END as normalized_pct
      FROM cell_ecoregion_raw ce
      JOIN cell_coverage_total ct ON ce.cell_id = ct.cell_id
    )
    -- insert rescaled primary productivity values
    INSERT INTO cell_metric (cell_id, metric_seq, value)
    SELECT 
      cm.cell_id,
      {m_seq} as metric_seq,
      SUM(
        -- (c.prim_prod_mean - er.min_value) / 
        (cm.value     - er.min_value) / 
        (er.max_value - er.min_value) * 
        (ce.normalized_pct / 100.0) ) * 100 AS value  -- scale to 0-100 range
    -- FROM cell c
    FROM cell_metric cm
    JOIN cell_ecoregion   ce ON cm.cell_id  = ce.cell_id
    JOIN ecoregion_minmax er ON ce.zone_seq = er.zone_seq
    WHERE 
    --c.prim_prod_mean IS NOT NULL AND 
    cm.metric_seq = {pp_seq} AND
    cm.value IS NOT NULL AND 
    ce.zone_seq IS NOT NULL AND
    er.min_value IS NOT NULL AND 
    er.max_value IS NOT NULL AND 
    er.min_value < er.max_value  -- avoid division by zero
    GROUP BY cm.cell_id
"
  )
)

# tbl(con_sdm, "cell_metric") |>
#   filter(
#     metric_seq == !!m_seq)

tbl(con_sdm, "metric") |>
  filter(
    metric_key == "primprod_ecoregion_rescaled"
  ) |>
  left_join(
    tbl(con_sdm, "cell_metric"),
    by = "metric_seq"
  ) |>
  pull(value) |>
  summary()
```

### Calculate ProgramArea metrics contributing to score based on cell metrics

```{r}
#| label: cell_metrics_to_programarea_metrics
#| eval: !expr do_metrics

# get zone_seqs for program areas
z_seqs <- tbl(con_sdm, "zone") |>
  filter(tbl == !!tbl_pra) |>
  pull(zone_seq)

message(glue("Calculating metrics for {length(z_seqs)} program areas..."))
# Calculating metrics for 20 program areas...

metrics <- c(
  glue("extrisk_{sp_cats}"),
  glue("extrisk_{sp_cats}_ecoregion_rescaled"),
  "primprod",
  "primprod_ecoregion_rescaled"
)

for (m_key in metrics) {
  # m_key = metrics[1]

  # get existing metric_seq
  m_seq <- tbl(con_sdm, "metric") |>
    filter(metric_key == !!m_key) |>
    pull(metric_seq)

  if (length(m_seq) != 1) {
    message(glue("  Skipping {m_key} - metric not found"))
    next
  }

  # delete existing zone_metric values for this metric and program areas
  dbExecute(
    con_sdm,
    glue(
      "DELETE FROM zone_metric
    WHERE metric_seq = {m_seq}
      AND zone_seq IN ({paste(z_seqs, collapse = ',')})"
    )
  )

  # stop if no cell_metric found
  n_cm <- tbl(con_sdm, "cell_metric") |>
    filter(metric_seq == !!m_seq) |>
    tally() |>
    pull(n)

  if (n_cm == 0) {
    message(glue("  Skipping {m_key} - no cell_metric values"))
    next
  }

  # calculate average score per program area weighted by cell coverage
  dbExecute(
    con_sdm,
    glue(
      "
    INSERT INTO zone_metric (zone_seq, metric_seq, value)
    SELECT
      zc.zone_seq,
      {m_seq} as metric_seq,
      SUM(cm.value * zc.pct_covered) / SUM(zc.pct_covered) as value
    FROM zone_cell zc
    JOIN cell_metric cm ON zc.cell_id = cm.cell_id
    WHERE
      cm.metric_seq = {m_seq}
      AND zc.zone_seq IN ({paste(z_seqs, collapse = ',')})
      AND cm.value IS NOT NULL
    GROUP BY zc.zone_seq"
    )
  )

  message(glue("  Calculated {m_key}"))
}

# verify
d_ck <- tbl(con_sdm, "metric") |>
  select(metric_seq, metric_key) |>
  filter(metric_key %in% metrics) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "metric_seq"
  ) |>
  left_join(
    tbl(con_sdm, "zone") |>
      filter(tbl == !!tbl_pra) |>
      select(zone_seq, pra_key = value),
    by = "zone_seq"
  ) |>
  filter(!is.na(pra_key)) |>
  collect()

d_ck |>
  filter(str_detect(metric_key, "_ecoregion_rescaled")) |>
  pull(value) |>
  summary()
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 1.195  11.186  21.902  27.317  41.150  91.748
```

### Calculate Ecoregion metrics contributing to score based on cell metrics

```{r}
#| label: cell_metrics_to_ecoregion_metrics
#| eval: !expr do_metrics

# get zone_seqs for ecoregions
z_seqs <- tbl(con_sdm, "zone") |>
  filter(
    tbl == !!tbl_er
  ) |>
  pull(zone_seq)
stopifnot(
  `No ecoregion zones found in zone table — check tbl_er value matches zone$tbl` = length(
    z_seqs
  ) >
    0
)

# tbl(con_sdm, "metric") |> pull(metric_key) |>  sort()

metrics <- c(
  glue("extrisk_{sp_cats}"),
  glue("extrisk_{sp_cats}_ecoregion_rescaled"),
  "primprod",
  "primprod_ecoregion_rescaled"
)

for (m_key in metrics) {
  # m_key = metrics[1]

  # get existing metric_seq
  m_seq <- tbl(con_sdm, "metric") |>
    filter(metric_key == !!m_key) |>
    pull(metric_seq)
  stopifnot(length(m_seq) == 1)

  # delete existing zone_metric values for this metric and ecoregions
  dbExecute(
    con_sdm,
    glue(
      "DELETE FROM zone_metric
    WHERE metric_seq = {m_seq}
      AND zone_seq IN ({paste(z_seqs, collapse = ',')})"
    )
  )

  # stop if no cell_metric found
  n_cm <- tbl(con_sdm, "cell_metric") |>
    filter(
      metric_seq == !!m_seq
    ) |>
    tally() |>
    pull(n)
  stopifnot(n_cm > 0)

  # calculate average score per ecoregion weighted by cell coverage
  dbExecute(
    con_sdm,
    glue(
      "
    INSERT INTO zone_metric (zone_seq, metric_seq, value)
    SELECT 
      zc.zone_seq,
      {m_seq} as metric_seq,
      SUM(cm.value * zc.pct_covered) / SUM(zc.pct_covered) as value
    FROM zone_cell zc
    JOIN cell_metric cm ON zc.cell_id = cm.cell_id
    WHERE 
      cm.metric_seq = {m_seq} 
      AND zc.zone_seq IN ({paste(z_seqs, collapse = ',')})
      AND cm.value IS NOT NULL
    GROUP BY zc.zone_seq"
    )
  )
}

d_ck <- tbl(con_sdm, "metric") |>
  select(metric_seq, metric_key) |>
  filter(
    metric_key %in% metrics
  ) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "metric_seq"
  ) |>
  left_join(
    tbl(con_sdm, "zone") |>
      filter(
        tbl == !!tbl_er
      ) |>
      select(zone_seq, pa_key = value),
    by = "zone_seq"
  ) |>
  filter(!is.na(pa_key)) |>
  collect()

d_ck |>
  filter(str_detect(metric_key, "_ecoregion_rescaled")) |>
  pull(value) |>
  summary()
# OLD:
#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 0.1106 12.9840 24.9402 28.6058 43.1917 83.1341
# NEW:
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 2.365  13.926  21.852  26.010  35.288  80.208
# NEWEST:
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 2.365  13.923  22.475  27.350  38.309  80.039
```

### Calculate cell_metric score: ExtinctionRisk + PrimProd (already EcoregionRescaled), equal weight

```{r}
#| label: cell_metric_score
#| eval: !expr do_metrics

# Create combined metric for extinction risk and primary productivity
m_key <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
m_description <- "Combined score of extinction risk per species category and primary productivity, equally weighted (and each previously rescaled [0,100] based on Ecoregional min/max values)"

# Check if metric exists and delete if it does
m_seq <- dbGetQuery(
  con_sdm,
  glue(
    "SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
  )
) |>
  pull(metric_seq)
if (length(m_seq) > 0) {
  dbExecute(con_sdm, glue("DELETE FROM metric      WHERE metric_seq = {m_seq}"))
  dbExecute(con_sdm, glue("DELETE FROM cell_metric WHERE metric_seq = {m_seq}"))
  dbExecute(con_sdm, glue("DELETE FROM zone_metric WHERE metric_seq = {m_seq}"))
}

# Create new metric
dbExecute(
  con_sdm,
  glue(
    "
  INSERT INTO metric (metric_key, description)
  VALUES ('{m_key}', '{m_description}')"
  )
)
m_seq <- dbGetQuery(
  con_sdm,
  glue(
    "
  SELECT metric_seq FROM metric WHERE metric_key = '{m_key}'"
  )
) |>
  pull(metric_seq)

# Get all metric_seqs for extinction risk metrics and primary productivity
metric_keys <- c(
  sapply(
    sp_cats,
    \(sc) glue("extrisk_{str_replace(sc, ' ', '_')}_ecoregion_rescaled"),
    USE.NAMES = F
  ),
  "primprod_ecoregion_rescaled"
)

# Create SQL to join all metrics with equal weights
dbExecute(
  con_sdm,
  glue(
    "
  WITH metric_values AS (
    SELECT 
      cm.cell_id,
      m.metric_key,
      cm.value
    FROM cell_metric cm
    JOIN metric m ON cm.metric_seq = m.metric_seq
    WHERE m.metric_key IN ({paste(shQuote(metric_keys), collapse=', ')}) )
  INSERT INTO cell_metric (cell_id, metric_seq, value)
  SELECT 
    cell_id,
    {m_seq} as metric_seq,
    ROUND(AVG(value)) AS value
  FROM metric_values
  GROUP BY cell_id"
  )
)

# Check the new metric
d_ck <- tbl(con_sdm, "metric") |>
  select(metric_seq, metric_key) |>
  filter(
    metric_key == !!m_key
  ) |>
  left_join(
    tbl(con_sdm, "cell_metric"),
    by = "metric_seq"
  ) |>
  collect()

d_ck |>
  pull(value) |>
  summary()
# OLD:
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 0.00   10.00   16.00   21.28   31.00   89.00
# NEW:
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#  0.00   12.00   17.00   21.29   26.00   89.00
# NEWEST:
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#  0.0    12.0    18.0    22.1    28.0    89.0
```

### Calculate Score per ProgramArea from ProgramArea metrics

```{r}
#| label: calc_programarea_score_from_programarea_metrics
#| eval: !expr do_metrics

# metric: calculate score per program area
m_key <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
m_seq <- tbl(con_sdm, "metric") |>
  filter(metric_key == !!m_key) |>
  pull(metric_seq)

metric_input_keys <- c(
  glue("extrisk_{sp_cats}_ecoregion_rescaled"),
  "primprod_ecoregion_rescaled"
)
m_input_seqs <- tbl(con_sdm, "metric") |>
  filter(metric_key %in% metric_input_keys) |>
  pull(metric_seq)
stopifnot(length(metric_input_keys) == length(m_input_seqs))

# get zone_seqs for program areas
z_seqs <- tbl(con_sdm, "zone") |>
  filter(
    tbl == !!tbl_pra
  ) |>
  pull(zone_seq)

# DEBUG check
# tbl(con_sdm, "zone_metric") |>
#   filter(
#     metric_seq == !!m_seq,
#     zone_seq %in% z_seqs) |>
#   arrange(desc(value)) |>
#   pull(value) |>
#   range()
# 5.704735 61.050130
#   View()
#   nrow() -> n_existing_zm
# z_seq == 134 # (ALA)

# delete existing zone_metric values for this metric and planning areas
dbExecute(
  con_sdm,
  glue(
    "DELETE FROM zone_metric 
  WHERE metric_seq = {m_seq} 
    AND zone_seq IN ({paste(z_seqs, collapse = ',')})"
  )
)

# calculate average score per program area
dbExecute(
  con_sdm,
  glue(
    "
  INSERT INTO zone_metric (zone_seq, metric_seq, value)
  SELECT 
    zone_seq,
    {m_seq} as metric_seq,
    SUM(value) / COUNT(value) as value
  FROM zone_metric zm
  WHERE 
    metric_seq IN ({paste(m_input_seqs, collapse = ',')}) 
    AND zone_seq IN ({paste(z_seqs, collapse = ',')})
    AND value IS NOT NULL
  GROUP BY zone_seq"
  )
) # 20

# check the new metric
d_ck <- tbl(con_sdm, "metric") |>
  select(metric_seq, metric_key) |>
  filter(
    metric_key == !!m_key
  ) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "metric_seq"
  ) |>
  left_join(
    tbl(con_sdm, "zone") |>
      filter(
        tbl == !!tbl_pra
      ) |>
      select(zone_seq, pa_key = value),
    by = "zone_seq"
  ) |>
  filter(!is.na(pa_key)) |>
  collect()

d_ck |>
  pull(value) |>
  summary()
# PlanAreas:
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#  7.16   17.10   22.86   25.42   31.59   60.07
# NEW ProgramAreas:
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 7.642  17.372  24.131  26.439  32.972  60.068
```

### Metric labels

Add display label columns to metric table as single source of truth for display names used in score tables, Word/Excel output, and flower plots.

```{r}
#| label: metric_labels
#| eval: true

# add display label columns to metric table if missing
for (col in c("metric_title", "metric_abbrev")) {
  cols <- dbGetQuery(
    con_sdm,
    "SELECT column_name FROM information_schema.columns WHERE table_name = 'metric'"
  )
  if (!col %in% cols$column_name) {
    dbExecute(con_sdm, glue("ALTER TABLE metric ADD COLUMN {col} VARCHAR"))
  }
}

# populate labels for score-contributing metrics
metric_labels <- tribble(
  ~metric_key                                                  , ~metric_title        , ~metric_abbrev ,
  "score_extriskspcat_primprod_ecoregionrescaled_equalweights" , "Score"              , "Score"        ,
  "extrisk_bird_ecoregion_rescaled"                            , "Bird"               , "Bird"         ,
  "extrisk_coral_ecoregion_rescaled"                           , "Coral"              , "Coral"        ,
  "extrisk_fish_ecoregion_rescaled"                            , "Fish"               , "Fish"         ,
  "extrisk_invertebrate_ecoregion_rescaled"                    , "Invertebrate"       , "Invertebrate" ,
  "extrisk_mammal_ecoregion_rescaled"                          , "Mammal"             , "Mammal"       ,
  "extrisk_other_ecoregion_rescaled"                           , "Other"              , "Other"        ,
  "extrisk_turtle_ecoregion_rescaled"                          , "Turtle"             , "Turtle"       ,
  "primprod_ecoregion_rescaled"                                , "Primary Production" , "Prim. Prod."
)

duckdb_register(con_sdm, "tmp_labels", metric_labels)
dbExecute(
  con_sdm,
  "
  UPDATE metric
  SET metric_title  = t.metric_title,
      metric_abbrev = t.metric_abbrev
  FROM tmp_labels t
  WHERE metric.metric_key = t.metric_key"
)
duckdb_unregister(con_sdm, "tmp_labels")
```

## Add/update `zone_taxon`

For showing species composition per zone (subregion, program area) in mapgl app.

```{r}
#| label: zone_taxon
#| eval: !expr do_zone_taxon

tbl_taxon <- tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  select(
    sp_cat,
    sp_common = common_name,
    sp_scientific = scientific_name,
    taxon_id,
    taxon_authority,
    rl_code = extrisk_code,
    er_score,
    is_er_spatial,
    is_mmpa,
    is_mbta,
    is_bcc,
    esa_code,
    esa_source,
    mdl_seq
  )

d <- tbl(con_sdm, "zone") |>
  filter(
    tbl %in% c(tbl_sr, tbl_pra)
  ) |>
  select(zone_seq, zone_tbl = tbl, zone_fld = fld, zone_value = value) |> # zone_tbl = tbl,
  inner_join(
    tbl(con_sdm, "zone_cell") |>
      select(zone_seq, cell_id),
    by = join_by(zone_seq)
  ) |>
  inner_join(
    tbl(con_sdm, "model_cell") |>
      select(mdl_seq, cell_id, value) |>
      inner_join(
        tbl_taxon,
        by = join_by(mdl_seq)
      ),
    by = join_by(cell_id)
  ) |>
  inner_join(
    tbl(con_sdm, "cell") |>
      select(cell_id, area_km2),
    by = join_by(cell_id)
  ) |>
  group_by(
    zone_tbl,
    zone_fld,
    zone_value, # zone_seq,
    mdl_seq,
    sp_cat,
    sp_common,
    sp_scientific,
    taxon_id,
    taxon_authority,
    rl_code,
    er_score,
    is_er_spatial,
    is_mmpa,
    is_mbta,
    is_bcc,
    esa_code,
    esa_source
  ) |>
  summarize(
    area_km2 = sum(area_km2, na.rm = T),
    avg_suit = mean(value, na.rm = T) / 100,
    .groups = "drop"
  ) |>
  collect()


d <- d |>
  mutate(
    # for is_er_spatial (turtles), ER is already baked into cell values
    suit_rl = ifelse(is_er_spatial, avg_suit, avg_suit * er_score / 100),
    suit_rl_area = suit_rl * area_km2
  ) |>
  group_by(zone_fld, zone_value, sp_cat) |>
  mutate(
    cat_suit_rl_area = sum(suit_rl_area, na.rm = T)
  ) |>
  ungroup() |>
  mutate(
    pct_cat = suit_rl_area / cat_suit_rl_area
  )

write_csv(d, zone_taxon_csv)
dbWriteTable(con_sdm, glue("zone_taxon"), d, overwrite = T)
```

## Update PostGIS layers with metrics from DuckDB

These layers are used in the mapgl app (via pg_tileserv) and need to be updated with the new metrics calculated in DuckDB. The `update_table_with_clear` function adds new metric columns if they don't exist, clears existing values for those metrics, and then updates with new values from DuckDB.

### Update Zone Tables in PostGIS with Metrics

```{r}
#| label: setup_update_pgtable_with_clear
#| eval: !expr do_postgis

update_table_with_clear <- function(con, df, tbl, key_col = "programarea_key") {
  # df = d_pa_metrics; tbl = "ply_planareas_2025"; key_col = "planarea_key"

  stopifnot(dbExistsTable(con, tbl))
  existing_cols <- dbListFields(con, tbl)
  stopifnot(key_col %in% existing_cols)
  stopifnot(key_col %in% names(df))
  source_cols <- setdiff(names(df), key_col)

  # Add missing columns
  for (col in source_cols) {
    if (!col %in% existing_cols) {
      add_col_query <- glue(
        "ALTER TABLE {tbl} ADD COLUMN {col} DOUBLE PRECISION"
      )
      dbExecute(con, add_col_query)
      message("Added column: ", col)
    }
  }

  # Clear existing values for columns we're updating
  existing_source_cols <- intersect(source_cols, existing_cols)
  if (length(existing_source_cols) > 0) {
    clear_clauses <- paste0(existing_source_cols, " = NULL", collapse = ", ")
    clear_query <- glue("UPDATE {tbl} SET {clear_clauses}")
    dbExecute(con, clear_query)
    message(
      "Cleared existing values for: ",
      paste(existing_source_cols, collapse = ", "),
      "\n"
    )
  }

  # Now update with new values
  dbWriteTable(
    con,
    "temp_metrics_update",
    df,
    temporary = TRUE,
    overwrite = TRUE
  )

  set_clauses <- paste0(source_cols, " = temp.", source_cols, collapse = ", ")
  update_query <- glue(
    "
    UPDATE {tbl} 
    SET {set_clauses}
    FROM temp_metrics_update temp
    WHERE {tbl}.{key_col} = temp.{key_col}
  "
  )

  rows_affected <- dbExecute(con, update_query)
  message("Updated ", rows_affected, " rows")

  dbRemoveTable(con, "temp_metrics_update", force = TRUE)

  return(rows_affected)
}

# update zone.tbl to include version suffix ----
zone_tbls <- tbl(con_sdm, "zone") |> distinct(tbl) |> pull(tbl)
for (zt in zone_tbls) {
  zt_v <- if (!grepl("_v\\d+$", zt)) paste0(zt, v_sfx) else zt
  if (zt != zt_v) {
    # dbExecute(
    #   con_sdm,
    #   glue(
    #     "UPDATE zone SET tbl = '{zt_v}' WHERE tbl = '{zt}'"
    #   )
    # )
    cat(glue("zone.tbl: {zt} -> {zt_v} ~ {Sys.time()}"))
  }
}
```

```{r}
#| label: create_pg_v3_tables
#| eval: !expr do_postgis

# create versioned PostGIS tables from existing tables ----
pg_zone_tbls <- c(
  tbl_er,
  tbl_pra
)
for (pg_tbl in pg_zone_tbls) {
  # pg_tbl already includes v_sfx (e.g. _v3) from tbl_er, tbl_pra
  pg_tbl_base <- str_remove(pg_tbl, v_sfx)
  if (!dbExistsTable(con, pg_tbl)) {
    dbExecute(con, glue("CREATE TABLE {pg_tbl} AS SELECT * FROM {pg_tbl_base}"))
    message(glue("Created {pg_tbl} from {pg_tbl_base}"))
  } else {
    message(glue("{pg_tbl} already exists"))
  }
}
```

### Load Program Area Scores to PostGIS

```{r}
#| label: pg_update_programarea_metrics
#| eval: !expr do_postgis

# delete zone_metric rows without a metric_key
dbExecute(
  con_sdm,
  "
  DELETE FROM zone_metric 
  WHERE metric_seq NOT IN (SELECT metric_seq FROM metric)"
) # 0

# check if table exists in PostGIS
if (!dbExistsTable(con, tbl_pra)) {
  stop("Run create_pg_v3_tables R chunk above")
}

# get program area metrics from DuckDB
d_pra_metrics <- tbl(con_sdm, "zone") |>
  filter(tbl == !!tbl_pra) |>
  select(programarea_key = value, zone_seq) |>
  left_join(
    tbl(con_sdm, "zone_metric") |>
      mutate(value = round(value, 1)),
    by = "zone_seq"
  ) |>
  left_join(
    tbl(con_sdm, "metric") |>
      select(metric_seq, metric_key),
    by = "metric_seq"
  ) |>
  select(programarea_key, metric_key, value) |>
  filter(
    !is.na(programarea_key),
    !is.na(metric_key),
    !is.na(value)
  ) |>
  pivot_wider(
    names_from = metric_key,
    values_from = value
  ) |>
  collect()
message(glue("Loaded {nrow(d_pra_metrics)} program area metrics"))

# For complete refresh (clears existing values first)
update_table_with_clear(
  con,
  d_pra_metrics,
  tbl_pra,
  "programarea_key"
)
```

```{r}
#| label: pg_rm_old_metrics
#| eval: !expr do_postgis

flds_pra_df <- names(d_pra_metrics)
flds_pra_db <- dbListFields(con, tbl_pra)
fdls_pra_rm <- setdiff(flds_pra_db, flds_pra_df) |>
  str_subset("(extrisk_)|(^na$)")

d_rm <-
  tibble(
    tbl = tbl_pra,
    fld = fdls_pra_rm
  )

for (i in seq_len(nrow(d_rm))) {
  # i = 1
  tbl <- d_rm$tbl[i]
  fld <- d_rm$fld[i]

  a <- dbExecute(con, glue('ALTER TABLE {tbl} DROP COLUMN "{fld}"'))
  message("Removed column: ", fld, " from ", tbl)
}
```


### Add index for vector tiles

per [Feature id | pg_tileserv](https://github.com/CrunchyData/pg_tileserv?tab=readme-ov-file#feature-id)

```{r}
#| label: add_index_for_vector_tiles
#| eval: !expr do_postgis

# add integer primary key to postgres programareas table
dbExecute(
  con,
  glue(
    "ALTER TABLE public.{tbl_pra} ALTER COLUMN programarea_id TYPE INTEGER;"
  )
)
dbExecute(
  con,
  glue(
    "ALTER TABLE public.{tbl_er} ADD COLUMN IF NOT EXISTS ecoregion_id SERIAL PRIMARY KEY;"
  )
)
```

## Generate downloads

For sharing with BOEM and external partners.

```{r}
#| label: downloads
#| eval: !expr do_downloads

redo_dl <- T

# ecoregion gpkg destination (input copy, no metrics — just copy to dir_v)
er_gpkg_dst <- glue("{dir_v}/ply_ecoregions_2025.gpkg")

# redo_dl: delete existing files
if (redo_dl) {
  for (f in c(
    pra_gpkg,
    lyrs_csv,
    metrics_tif,
    metrics_pra_tif,
    er_gpkg_dst
  )) {
    if (file_exists(f)) {
      file_delete(f)
    }
  }
}

# * pra_gpkg (with metrics from DuckDB) ----
if (!file.exists(pra_gpkg)) {
  message("Generating Program Areas geopackage with metrics...")

  # get metrics from DuckDB
  d_pra_metrics <- tbl(con_sdm, "zone") |>
    filter(tbl == !!tbl_pra) |>
    select(programarea_key = value, zone_seq) |>
    left_join(
      tbl(con_sdm, "zone_metric") |>
        mutate(value = round(value, 1)),
      by = "zone_seq"
    ) |>
    left_join(
      tbl(con_sdm, "metric") |>
        select(metric_seq, metric_key),
      by = "metric_seq"
    ) |>
    select(programarea_key, metric_key, value) |>
    filter(!is.na(metric_key), !is.na(value)) |>
    pivot_wider(names_from = metric_key, values_from = value) |>
    collect()

  # read base geometry (spatial columns only, no metrics)
  base_cols <- c(
    "programarea_id",
    "region_key",
    "region_name",
    "planarea_key",
    "planarea_name",
    "programarea_key",
    "programarea_name",
    "ctr_lon",
    "ctr_lat",
    "area_km2"
  )
  pra_prev_gpkg <- glue(
    "{dir_derived}/{ver_prev}/ply_programareas_2026_{ver_prev}.gpkg"
  )
  pra_base <- read_sf(pra_prev_gpkg) |>
    select(any_of(base_cols))

  # join metrics to geometry
  pra_out <- pra_base |>
    left_join(d_pra_metrics, by = "programarea_key")

  st_write(pra_out, pra_gpkg, delete_dsn = T, quiet = T)
}

# * lyrs_csv ----
if (!file.exists(lyrs_csv)) {
  message("Generating layers CSV...")

  d_lyrs <- bind_rows(
    tibble(
      order = 1,
      category = "Overall",
      layer = "score",
      lyr = "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
    ),
    tibble(
      order = 2,
      category = "Species, rescaled by Ecoregion",
      layer = glue("{sp_cats}: ext. risk, ecorgn"),
      lyr = glue("extrisk_{sp_cats}_ecoregion_rescaled")
    ),
    tibble(
      order = 3,
      category = "Primary Productivity, rescaled by Ecoregion",
      layer = glue("prim prod, ecorgn"),
      lyr = glue("primprod_ecoregion_rescaled")
    ),
    tibble(
      order = 4,
      category = "Species, raw Extinction Risk",
      layer = glue("{sp_cats}: ext. risk"),
      lyr = glue("extrisk_{sp_cats}")
    ),
    tibble(
      order = 5,
      category = "Primary Productivity, raw Phytoplankton",
      lyr = "primprod",
      layer = "prim prod, 2014-2023 avg (mg C/m^2/day)"
    )
  )

  write_csv(d_lyrs, lyrs_csv)
}

# * metrics_tif ----
if (!file.exists(metrics_tif)) {
  message("Generating Metrics raster...")

  d_lyrs <- read_csv(lyrs_csv)

  # add zones
  lst <- list()
  for (zone_fld in c("ecoregion_key", "programarea_key")) {
    # zone_fld = "ecoregion_key"
    # zone_fld = "programarea_key"

    d <- tbl(con_sdm, "zone") |>
      filter(fld == !!zone_fld) |>
      left_join(
        tbl(con_sdm, "zone_cell"),
        by = "zone_seq"
      ) |>
      group_by(cell_id) |>
      window_order(pct_covered) |>
      summarize(
        value = last(value),
        .groups = "drop"
      ) |>
      collect() |>
      mutate(
        "{zone_fld}" := as.factor(value)
      ) |>
      select(all_of(c("cell_id", zone_fld)))

    lvls <- levels(d[[zone_fld]])
    if (length(lvls) == 0) {
      message(glue("Skipping {zone_fld}: no zone data found"))
      next
    }

    r <- init(r_cell[[1]], NA) # |> as.factor() #sort(unique(d$value)))
    # r$chr <- NA_character_ # ensure chr type
    r[d$cell_id] <- d[[zone_fld]]
    varnames(r) <- zone_fld
    names(r) <- zone_fld
    levels(r) <- tibble(
      id = seq_along(lvls),
      "{zone_fld}" := lvls
    )
    lst[[zone_fld]] <- r
  }
  r_zones <- do.call(c, lst |> unname())

  # add layers
  lst <- list()
  for (i in 1:nrow(d_lyrs)) {
    # i = 2
    lyr <- d_lyrs$lyr[i]
    layer <- d_lyrs$layer[i]

    r <- get_rast(lyr) # plot(r)
    names(r) <- lyr
    varnames(r) <- layer
    lst[[i]] <- r
  }
  r_metrics <- do.call(c, lst)

  # add log of primprod
  r_metrics[["primprod_log"]] <- log(r_metrics[["primprod"]])

  # show layer names
  names(r_metrics) |> sort() |> print()

  r_metrics <- c(r_zones, r_metrics)
  writeRaster(r_metrics, metrics_tif, overwrite = T)
}

# * metrics_pra_tif ----
if (!file.exists(metrics_pra_tif)) {
  message("Generating Metrics raster masked to Program Areas...")
  # mask to study area of Program Areas

  # subregion_keys in sr_pra_csv: USA (all 20), AK (15), PA (3), GA (2)
  # AK + L48 = USA (all program areas)
  sr <- "USA"
  pra_sr <- read_csv(sr_pra_csv) |>
    filter(subregion_key == !!sr) |>
    pull(programarea_key)

  r_metrics <- rast(metrics_tif)
  r_mask <- subset(r_metrics, "programarea_key")

  r_metrics_pra <- r_metrics |>
    mask(r_mask) |>
    trim()

  writeRaster(r_metrics_pra, metrics_pra_tif, overwrite = T)
}

# * er_gpkg_dst (input, no metrics — copy to dir_v for downloads) ----
if (!file.exists(er_gpkg_dst) && er_raw_gpkg != er_gpkg_dst) {
  message("Copying ecoregion gpkg to dir_v...")
  file_copy(er_raw_gpkg, er_gpkg_dst)
}
```

```{r}
#| label: old_pa_metrics_geo
#| eval: false

pa_metrics_geo <- glue(
  "{dir_derived}/OCS_Proposed_Planning_Area_Polygon_Env_Sensitivity_Metrics.geojson"
)

# * pa_metrics_geo ----
if (!file.exists(pa_metrics_geo)) {
  pa_boem_geo <- glue("{dir_data}/raw/boem.gov/OCS_Area_Polygons.geojson")

  pa <- read_sf(pa_gpkg)

  pa_boem <- read_sf(pa_boem_geo)

  (pa_miss <- pa_boem |>
    st_drop_geometry() |>
    anti_join(
      pa |>
        st_drop_geometry() |>
        select(planarea_key, planarea_name),
      by = join_by(PLANNING_AREA == planarea_key)
    ) |>
    pull(PLANNING_AREA))
  # [1] "CG" "EG" "WG"

  pa_boem |>
    filter(PLANNING_AREA %in% pa_miss) |>
    mapview::mapView(zcol = "PLANNING_AREA")

  pa |>
    st_drop_geometry() |>
    filter(
      region_name == "Gulf of America"
    ) |>
    select(planarea_key, planarea_name)
  #   planarea_key planarea_name
  #   <chr>        <chr>
  # 1 CGA          Central Gulf of America
  # 2 EGA          Eastern Gulf of America
  # 3 WGA          Western Gulf of America

  pa_metrics <- pa_boem |>
    left_join(
      pa |>
        st_drop_geometry() |>
        mutate(
          PLANNING_AREA = case_match(
            planarea_key,
            "CGA" ~ "CG",
            "EGA" ~ "EG",
            "WGA" ~ "WG",
            .default = planarea_key
          )
        ),
      by = join_by(PLANNING_AREA)
    )

  write_sf(pa_metrics, pa_metrics_geo)
}
```

## Generate PMTiles

Static PMTiles archives for mapgl/mapsp apps, replacing PostGIS/pg_tileserv tile pipeline.

```{r}
#| label: generate_pmtiles
#| eval: !expr do_pmtiles

# check / install tippecanoe and pmtiles CLI ----
check_brew_bin <- function(bin) {
  if (Sys.which(bin) == "") {
    if (Sys.which("brew") == "") {
      stop(glue("{bin} not found and homebrew not available"))
    }
    message(glue("installing {bin} via homebrew..."))
    system2("brew", c("install", bin))
    stopifnot(Sys.which(bin) != "")
  }
  message(glue("{bin}: {Sys.which(bin)}"))
}
check_brew_bin("tippecanoe") # tippecanoe: /opt/homebrew/bin/tippecanoe
check_brew_bin("pmtiles") # pmtiles:       /opt/homebrew/bin/pmtiles

dir_create(dir_pmtiles)

# helper: sf → pmtiles via tippecanoe ----
sf_to_pmtiles <- function(sf_obj, layer_name, out_dir = dir_pmtiles) {
  tmp_geojson <- tempfile(fileext = ".geojson")
  on.exit(unlink(tmp_geojson), add = TRUE)

  out_pmtiles <- glue("{out_dir}/{layer_name}.pmtiles")

  # wrap polygons crossing the antimeridian so GeoJSON [-180,180] doesn't clip
  #sf_obj <- st_wrap_dateline(sf_obj, options = c("WRAPDATELINE=YES"))

  st_write(sf_obj, tmp_geojson, driver = "GeoJSON") #, delete_dsn = TRUE)

  # tippecanoe: full detail for small polygon layers
  system2(
    "tippecanoe",
    c(
      "-z12",
      "-Z0",
      "--no-feature-limit",
      "--no-tile-size-limit",
      "-l",
      layer_name,
      "-o",
      out_pmtiles,
      "--force",
      tmp_geojson
    )
  )
  stopifnot(file_exists(out_pmtiles))
  message(glue("wrote {out_pmtiles} ({file_size(out_pmtiles)})"))
  out_pmtiles
}

# program areas WITH metrics ----
d_pra_metrics <- tbl(con_sdm, "zone") |>
  filter(tbl == !!tbl_pra) |>
  select(programarea_key = value, zone_seq) |>
  left_join(
    tbl(con_sdm, "zone_metric") |>
      mutate(value = round(value, 1)),
    by = "zone_seq"
  ) |>
  left_join(
    tbl(con_sdm, "metric") |>
      select(metric_seq, metric_key),
    by = "metric_seq"
  ) |>
  select(programarea_key, metric_key, value) |>
  filter(!is.na(programarea_key), !is.na(metric_key), !is.na(value)) |>
  pivot_wider(names_from = metric_key, values_from = value) |>
  collect()

# load geometry from memory or prior version (pra_gpkg may not exist yet)
pra_geom <- if (exists("pra")) {
  pra
} else if (file_exists(pra_gpkg)) {
  read_sf(pra_gpkg)
} else {
  pra_prev <- glue(
    "{dir_derived}/{ver_prev}/ply_programareas_2026_{ver_prev}.gpkg"
  )
  read_sf(pra_prev)
}

# strip any existing metric columns before joining fresh ones
base_cols <- c(
  "programarea_id",
  "region_key",
  "region_name",
  "planarea_key",
  "planarea_name",
  "programarea_key",
  "programarea_name",
  "ctr_lon",
  "ctr_lat",
  "area_km2"
)
pra_sf <- pra_geom |>
  select(any_of(base_cols)) |>
  left_join(d_pra_metrics, by = "programarea_key")

# mapView(pra_sf, zcol = "score_extriskspcat_primprod_ecoregionrescaled_equalweights")

# program areas for pmtiles: base columns only (scores joined at render time)
pra_pm_sf <- pra_geom |> select(any_of(base_cols))
sf_to_pmtiles(pra_pm_sf, tbl_pra_pm)

# plot(pra_sf["score_extriskspcat_primprod_ecoregionrescaled_equalweights"])

# ecoregions WITHOUT metrics (outlines only) ----
er_sf <- read_sf(er_raw_gpkg)
sf_to_pmtiles(er_sf, tbl_er_pm)
```

## Summary data products

Merged from `msens-summary_programareas.qmd`: ecoregion × program area intersections, label placement points, and score tables.

```{r}
#| label: summary_data_redo
#| eval: !expr do_summary_data

# delete existing summary data files if redo (except lbls_csv — manually adjusted)
if (redo_summary_data) {
  for (f in c(er_pra_gpkg, er_pra_csv, er_pra_n_csv, pra_n_csv)) {
    if (file_exists(f)) {
      file_delete(f)
    }
  }
}
```

### Ecoregion × Program Area intersections

```{r}
#| label: read_plys
#| eval: !expr do_summary_data

er <- read_sf(er_raw_gpkg)
pra <- read_sf(pra_gpkg)

if (!file.exists(er_pra_gpkg)) {
  er_pra <- st_intersection(
    er |>
      select(ecoregion_key, ecoregion_name),
    pra |>
      select(region_key, region_name, programarea_key, programarea_name)
  ) |>
    st_make_valid() |>
    mutate(
      area_km2 = st_area(geom) |> set_units(km^2) |> drop_units(),
      ctr_lon = st_coordinates(st_centroid(geom))[, 1],
      ctr_lat = st_coordinates(st_centroid(geom))[, 2]
    ) |>
    filter(area_km2 > 1) |>
    arrange(programarea_key, ecoregion_key) |>
    group_by(programarea_key) |>
    mutate(
      pct_programarea_in_ecoregion = round(
        area_km2 / sum(area_km2) * 100,
        2
      ),
      pct_programarea_in_ecoregion = na_if(
        pct_programarea_in_ecoregion,
        100.0
      )
    ) |>
    ungroup()

  er_pra <- er_pra |>
    select(
      region_key,
      region_name,
      ecoregion_key,
      ecoregion_name,
      programarea_key,
      programarea_name,
      pct_programarea_in_ecoregion,
      area_km2,
      ctr_lon,
      ctr_lat
    ) |>
    arrange(region_key, ecoregion_key, programarea_key) |>
    mutate(
      area_km2 = round(area_km2, 2),
      ctr_lon = round(ctr_lon, 4),
      ctr_lat = round(ctr_lat, 4)
    )

  er_pra |>
    st_drop_geometry() |>
    write_csv(er_pra_csv, na = "")

  write_sf(er_pra, er_pra_gpkg, delete_dsn = T)
}
er_pra <- read_sf(er_pra_gpkg)
```

### Label placement points

```{r}
#| label: ply_lbl_pts
#| eval: !expr do_summary_data

er_pts <- er |>
  st_drop_geometry() |>
  select(
    ecoregion_key,
    ecoregion_name,
    region_name,
    region_key,
    ctr_lon,
    ctr_lat
  ) |>
  st_as_sf(coords = c("ctr_lon", "ctr_lat"), crs = 4326)
pra_pts <- pra |>
  st_drop_geometry() |>
  select(
    programarea_key,
    programarea_name,
    region_name,
    region_key,
    ctr_lon,
    ctr_lat
  ) |>
  st_as_sf(coords = c("ctr_lon", "ctr_lat"), crs = 4326)

# keep file-exists guard — label placement is manually adjusted
if (!file.exists(lbls_csv)) {
  bind_rows(
    er_pts |> st_drop_geometry() |> select(ecoregion_key),
    pra_pts |> st_drop_geometry() |> select(programarea_key)
  ) |>
    arrange(ecoregion_key, programarea_key) |>
    mutate(
      text_anchor = ifelse(!is.na(programarea_key), "center", "top"),
      text_justify = "center",
      text_offset_right = 0,
      text_offset_down = 0
    ) |>
    write_csv(lbls_csv, na = "")
}
```

### Score tables

```{r}
#| label: pra_score_table
#| eval: !expr do_summary_data

d_pra <- read_sf(pra_gpkg) |>
  st_drop_geometry()

# get field-to-label mapping from metric table (single source of truth)
d_metric_labels <- tbl(con_sdm, "metric") |>
  filter(!is.na(metric_title)) |>
  select(metric_key, metric_title, metric_abbrev) |>
  collect()

# fields present in program area data that have labels
flds <- intersect(names(d_pra), d_metric_labels$metric_key)

# named vector: metric_key -> short name for column renaming
flds_rn <- d_metric_labels |>
  filter(metric_key %in% flds) |>
  mutate(
    short = metric_key |>
      str_replace(
        "score_extriskspcat_primprod_ecoregionrescaled_equalweights",
        "score"
      ) |>
      str_replace("_ecoregion_rescaled", "") |>
      str_replace("extrisk_", "")
  ) |>
  pull(short, name = metric_key)

# ecoregion header rows with NA for all metric columns
na_cols <- setNames(
  rep(list(NA_real_), length(flds_rn) + 1),
  c("area_km2", unname(flds_rn))
)
d_er_n <- read_csv(er_pra_csv) |>
  distinct(
    region_key,
    region_name,
    ecoregion_key,
    ecoregion_name
  ) |>
  bind_cols(as_tibble(na_cols))

d_pra_n <- read_csv(er_pra_csv) |>
  select(
    region_name,
    ecoregion_name,
    programarea_key,
    programarea_name,
    pct_programarea_in_ecoregion
  ) |>
  left_join(
    d_pra |>
      select(all_of(c("programarea_key", "area_km2", flds))) |>
      rename_with(~ flds_rn[.x], all_of(flds)),
    by = "programarea_key"
  )

d_er_pra_n <- bind_rows(
  d_er_n |> mutate(row_type = "ecoregion"),
  d_pra_n |> mutate(row_type = "programarea")
) |>
  relocate(
    row_type,
    programarea_key,
    programarea_name,
    pct_programarea_in_ecoregion,
    .after = ecoregion_name
  ) |>
  arrange(region_name, ecoregion_name, row_type, programarea_name) |>
  mutate(
    pct = ifelse(
      !is.na(pct_programarea_in_ecoregion),
      glue(" [{round(pct_programarea_in_ecoregion, 1)}%]", .trim = F),
      ""
    ),
    row_lbl = case_match(
      row_type,
      "ecoregion" ~ glue("{region_name}: {ecoregion_name} ({ecoregion_key})"),
      "programarea" ~ glue("· {programarea_name} ({programarea_key}){pct}")
    )
  ) |>
  relocate(row_lbl)

d_er_pra_n |>
  select(
    row_type,
    region_name,
    ecoregion_name,
    ecoregion_key,
    programarea_name,
    programarea_key,
    pct_programarea_in_ecoregion,
    row_lbl,
    score,
    primprod,
    any_of(sp_cats),
    area_km2
  ) |>
  write_excel_csv(er_pra_n_csv)
```

## Summary figures and tables

Merged from `msens-summary_programareas.qmd`: maps, flower plots, primary productivity charts, and Word/Excel tables.

```{r}
#| label: summary_figs_redo
#| eval: !expr do_summary_figs

dir_create(dir_figs)

# delete figs if redo requested
if (redo_summary_figs) {
  fig_files <- dir_ls(dir_figs)
  if (length(fig_files) > 0) {
    file_delete(fig_files)
  }
}
```

### Program Area boundary maps

```{r}
#| label: map_imgs_pra_ply_l48_ak
#| eval: !expr do_summary_figs

var_pra <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
d_lbls <- read_csv(lbls_csv)

for (is_ak in c(TRUE, FALSE)) {
  if (is_ak) {
    pra_area <- pra |>
      filter(region_key == "AK") |>
      st_shift_longitude()
    er_area <- er |>
      filter(region_key == "AK") |>
      st_shift_longitude()

    er_pts_area <- er_pts |>
      filter(region_key == "AK") |>
      st_shift_longitude() |>
      left_join(
        d_lbls |> select(-programarea_key),
        by = "ecoregion_key"
      )
    pra_pts_area <- pra_pts |>
      filter(region_key == "AK") |>
      st_shift_longitude() |>
      left_join(
        d_lbls |> select(-ecoregion_key),
        by = "programarea_key"
      )

    lgnd_pos <- "top-left"
  } else {
    box_l48 <- pra |>
      filter(region_key != "AK") |>
      st_bbox() |>
      st_as_sfc()
    pra_area <- pra |>
      st_filter(box_l48, .predicate = st_intersects)
    er_area <- er |>
      st_filter(box_l48, .predicate = st_intersects)

    er_pts_area <- er_pts |>
      filter(ecoregion_key %in% er_area$ecoregion_key) |>
      left_join(
        d_lbls |> select(-programarea_key),
        by = "ecoregion_key"
      )
    pra_pts_area <- pra_pts |>
      filter(programarea_key %in% pra_area$programarea_key) |>
      left_join(
        d_lbls |> select(-ecoregion_key),
        by = "programarea_key"
      )

    lgnd_pos <- "bottom-left"
  }

  m <- mapboxgl(
    style = mapbox_style("dark"),
    projection = "globe"
  ) |>
    fit_bounds(bbox = er_area) |>
    msens::add_pmfill(
      url = glue("{pmtiles_base_url}/{tbl_pra_pm}.pmtiles"),
      source_layer = tbl_pra_pm,
      col_key = "programarea_key",
      d = pra_area |> sf::st_drop_geometry(),
      col_value = var_pra,
      filter_keys = pra_area$programarea_key
    )

  legend_meta <- attr(m, "legend_meta")

  m <- m |>
    msens::add_pmline(list(
      list(
        url = glue("{pmtiles_base_url}/{tbl_er_pm}.pmtiles"),
        source_layer = tbl_er_pm,
        line_color = "lightgray",
        line_width = 5,
        filter = c("in", "ecoregion_key", er_area$ecoregion_key)
      )
    )) |>
    msens::add_pmlabel(list(
      list(source = pra_pts_area, text_field = "programarea_key"),
      list(
        source = er_pts_area,
        text_field = "ecoregion_name",
        text_color = "black",
        text_halo_color = "lightgray",
        text_halo_width = 2,
        text_halo_blur = 2,
        text_line_height = 1.5,
        text_size = 20
      )
    )) |>
    mapgl::add_legend(
      "Score",
      values = legend_meta$rng,
      colors = legend_meta$colors,
      position = lgnd_pos
    ) |>
    mapgl::add_scale_control(position = "bottom-right")

  b <- glue("{dir_figs}/map_pra_{ifelse(is_ak, 'ak', 'l48')}{v_sfx}")
  b_html <- glue("{b}.html")
  b_files <- glue("{b}_files")
  b_png <- glue("{b}.png")
  saveWidget(m, b_html, selfcontained = FALSE, background = "transparent")
  webshot(
    b_html,
    b_png,
    vwidth = 1200,
    vheight = 800,
    zoom = 2,
    selector = "#htmlwidget_container",
    delay = 2,
    quiet = TRUE
  )
  file_delete(b_html)
  dir_delete(b_files)
}
```

### Cell score & NPP maps

```{r}
#| label: map_imgs_cell_l48_ak
#| eval: !expr do_summary_figs

vgpm_tif <- glue(
  "{dir_data}/raw/oregonstate.edu/vgpm.r2022.v.chl.v.sst.2160x4320_2014-2023.avg.sd_planareas.tif"
)

r_metrics <- rast(metrics_tif)
r_vgpm <- rast(vgpm_tif)
d_lbls <- read_csv(lbls_csv)

lyr_score <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
lyr_npp <- "npp_avg"

# mask to program areas to exclude Atlantic coast
ply_mask <- vect(pra |> st_shift_longitude())
r_score <- r_metrics[[lyr_score]] |> mask(ply_mask)
r_npp <- r_vgpm |> log() |> mask(ply_mask)

n_cols <- 11
rng_r_score <- range(values(r_score, na.rm = T))
cols_r_score <- rev(RColorBrewer::brewer.pal(n_cols, "Spectral"))
brks_r_score <- seq(rng_r_score[1], rng_r_score[2], length.out = n_cols)

rng_r_npp <- range(values(r_npp, na.rm = T))
cols_r_npp <- pal_viridis()(n_cols)
brks_r_npp <- seq(rng_r_npp[1], rng_r_npp[2], length.out = n_cols)

for (is_vgpm in c(T, F)) {
  # npp: is_vgpm = T # score: is_vgpm = F
  if (is_vgpm) {
    r <- r_npp
    rng_r <- round(rng_r_npp, 2)
    cols_r <- cols_r_npp
    brks_r <- brks_r_npp
    lyr <- lyr_npp
    title <- "Primary Productivity (log(NPP))"
  } else {
    r <- r_score
    rng_r <- rng_r_score
    cols_r <- cols_r_score
    brks_r <- brks_r_score
    lyr <- lyr_score
    title <- "Score"
  }

  for (is_ak in c(T, F)) {
    # is_ak = T
    if (is_ak) {
      rgn_mask <- pra |>
        filter(region_key == "AK") |>
        st_shift_longitude() |>
        vect()

      r_area <- r |>
        mask(rgn_mask) |>
        trim()

      er_area <- er |>
        filter(region_key == "AK") |>
        st_shift_longitude()

      pra_area <- pra |>
        filter(region_key == "AK") |>
        st_shift_longitude()

      er_pts_area <- er_pts |>
        filter(region_key == "AK") |>
        st_shift_longitude() |>
        left_join(
          d_lbls |> select(-programarea_key),
          by = "ecoregion_key"
        )
      pra_pts_area <- pra_pts |>
        filter(region_key == "AK") |>
        st_shift_longitude() |>
        left_join(
          d_lbls |> select(-ecoregion_key),
          by = "programarea_key"
        )

      lgnd_pos <- "top-left"
    } else {
      rgn_mask <- pra |>
        filter(region_key != "AK") |>
        st_shift_longitude() |>
        vect()

      r_area <- r |>
        mask(rgn_mask) |>
        trim()

      box_lower48 <- pra |>
        filter(region_key != "AK") |>
        st_bbox() |>
        st_as_sfc()

      er_area <- er |>
        st_filter(box_lower48, .predicate = st_intersects)
      pra_area <- pra |>
        st_filter(box_lower48, .predicate = st_intersects)

      er_pts_area <- er_pts |>
        filter(ecoregion_key %in% er_area$ecoregion_key) |>
        left_join(
          d_lbls |> select(-programarea_key),
          by = "ecoregion_key"
        )
      pra_pts_area <- pra_pts |>
        filter(programarea_key %in% pra_area$programarea_key) |>
        left_join(
          d_lbls |> select(-ecoregion_key),
          by = "programarea_key"
        )

      lgnd_pos <- "bottom-left"
    }
    scale_pos <- "bottom-right"

    er_filter <- c("in", "ecoregion_key", er_area$ecoregion_key)
    pra_filter <- c("in", "programarea_key", pra_area$programarea_key)

    m <- msens::map_cells(
      r = r_area,
      colors = cols_r,
      bounds = er_area,
      raster_opacity = 0.9,
      legend_title = title,
      legend_position = lgnd_pos,
      legend_values = rng_r,
      pmtiles_outlines = list(
        list(
          url = glue("{pmtiles_base_url}/{tbl_er_pm}.pmtiles"),
          source_layer = tbl_er_pm,
          line_color = "lightgray",
          line_width = 5,
          filter = er_filter
        ),
        list(
          url = glue("{pmtiles_base_url}/{tbl_pra_pm}.pmtiles"),
          source_layer = tbl_pra_pm,
          line_color = "black",
          line_width = 1,
          filter = pra_filter
        )
      ),
      labels = list(
        list(source = pra_pts_area, text_field = "programarea_key"),
        list(
          source = er_pts_area,
          text_field = "ecoregion_name",
          text_color = "black",
          text_halo_color = "lightgray",
          text_halo_width = 2,
          text_halo_blur = 2,
          text_line_height = 1.5,
          text_size = 20
        )
      )
    )

    b <- glue(
      "{dir_figs}/map_cell_{ifelse(is_vgpm, 'npp', 'score')}_{ifelse(is_ak, 'ak', 'l48')}{v_sfx}"
    )
    b_html <- glue("{b}.html")
    b_files <- glue("{b}_files")
    b_png <- glue("{b}.png")
    saveWidget(m, b_html, selfcontained = F, background = "transparent")
    webshot(
      b_html,
      b_png,
      vwidth = 1200,
      vheight = 800,
      zoom = 2,
      selector = "#htmlwidget_container",
      delay = 2,
      quiet = T
    )
    file_delete(b_html)
    dir_delete(b_files)
  }
}
```

### Ecoregion × Program Area table (docx)

```{r}
#| label: flextable_er_pra
#| eval: !expr do_summary_figs

tbl_docx <- glue("{dir_figs}/table_ecoregion_programarea_scores{v_sfx}.docx")

set_flextable_defaults(
  font.family = "Arial",
  font.size = 10,
  border.color = "gray",
  big.mark = ","
)

# build rename vector from metric table: metric_abbrev -> short_name
abbrev_rn <- d_metric_labels |>
  filter(metric_key %in% flds) |>
  mutate(short = flds_rn[metric_key]) |>
  filter(short != "score") |>
  pull(short, name = metric_abbrev)

d_er_pra_n |>
  mutate(
    area_kkm2 = area_km2 / 1000
  ) |>
  mutate(across(where(is.numeric), round)) |>
  mutate(
    row_lbl = ifelse(
      row_type == "ecoregion",
      glue("**{row_lbl}**"),
      row_lbl
    )
  ) |>
  select(
    Area = row_lbl,
    Score = score,
    all_of(abbrev_rn),
    `Area (1,000 km^2^)` = area_kkm2
  ) |>
  flextable() |>
  ftExtra::colformat_md(part = "all") |>
  theme_vanilla() |>
  save_as_docx(path = tbl_docx)

# browseURL(tbl_docx)
```

### Program Area table (docx + xlsx)

```{r}
#| label: flextable_pra
#| eval: !expr do_summary_figs

tbl_docx <- glue("{dir_figs}/table_programarea_scores{v_sfx}.docx")
tbl_xlsx <- glue("{dir_figs}/table_programarea_scores{v_sfx}.xlsx")

set_flextable_defaults(
  font.family = "Arial",
  font.size = 10,
  border.color = "gray",
  big.mark = ","
)

# build rename vector: metric_title -> short_name
title_rn <- d_metric_labels |>
  filter(metric_key %in% flds) |>
  mutate(short = flds_rn[metric_key]) |>
  filter(short != "score") |>
  pull(short, name = metric_title)

d_pra_only <- d_er_pra_n |>
  filter(row_type == "programarea") |>
  arrange(desc(score)) |>
  select(
    -row_type,
    -ecoregion_name,
    -ecoregion_key,
    -pct_programarea_in_ecoregion
  ) |>
  mutate(
    area_kkm2 = area_km2 / 1000
  ) |>
  mutate(across(where(is.numeric), round)) |>
  mutate(
    row_lbl = glue("{programarea_name} ({programarea_key})")
  ) |>
  select(
    row_lbl,
    score,
    all_of(unname(title_rn)),
    area_kkm2
  ) |>
  distinct() |>
  select(
    `Program Area` = row_lbl,
    Score = score,
    all_of(title_rn),
    `Area (1,000 km^2^)` = area_kkm2
  )

write_excel_csv(d_pra_only, pra_n_csv, na = "")

d_pra_only |>
  flextable() |>
  ftExtra::colformat_md(part = "all") |>
  theme_vanilla() |>
  save_as_docx(path = tbl_docx)

# xlsx version (plain column names without superscript)
d_pra_only |>
  rename(`Area (1,000 km2)` = `Area (1,000 km^2^)`) |>
  openxlsx::write.xlsx(tbl_xlsx)

# browseURL(tbl_docx)
```

### Compare scores with previous version

```{r}
#| label: compare_scores_prev
#| eval: !expr do_summary_figs

# compare pra scores between two versions ----
# returns list with long (d_comp), wide diff (d_comp_wide), and
# wide prev/curr/diff (d_comp_pvt) data frames; NULL if files missing
compare_scores <- function(ver_curr, ver_prev) {
  csv_curr <- glue("{dir_derived}/{ver_curr}/tbl_pra_scores_{ver_curr}.csv")
  csv_prev <- glue("{dir_derived}/{ver_prev}/tbl_pra_scores_{ver_prev}.csv")

  if (!file_exists(csv_prev) || !file_exists(csv_curr)) {
    message(glue(
      "Skipping {ver_curr} vs {ver_prev}: ",
      "prev={file_exists(csv_prev)}, curr={file_exists(csv_curr)}"
    ))
    return(NULL)
  }

  d_prev <- read_csv(csv_prev, show_col_types = FALSE) |>
    rename(`Area (1,000 km2)` = `Area (1,000 km^2^)`)
  d_curr <- read_csv(csv_curr, show_col_types = FALSE) |>
    rename(`Area (1,000 km2)` = `Area (1,000 km^2^)`)

  # component columns (exclude Program Area and Area)
  comp_cols <- intersect(
    setdiff(names(d_prev), c("Program Area", "Area (1,000 km2)")),
    setdiff(names(d_curr), c("Program Area", "Area (1,000 km2)"))
  )

  # use version names for value columns (e.g., Score_v4b, Score_v4)
  v_c <- ver_curr
  v_p <- ver_prev

  d_comp <- d_prev |>
    select(`Program Area`, all_of(comp_cols)) |>
    pivot_longer(
      cols = all_of(comp_cols),
      names_to = "Component",
      values_to = v_p
    ) |>
    full_join(
      d_curr |>
        select(`Program Area`, all_of(comp_cols)) |>
        pivot_longer(
          cols = all_of(comp_cols),
          names_to = "Component",
          values_to = v_c
        ),
      by = c("Program Area", "Component")
    ) |>
    mutate(
      diff = .data[[v_c]] - .data[[v_p]]
    ) |>
    arrange(Component, `Program Area`)

  comp_csv <- glue("{dir_v}/tbl_pra_scores_{ver_curr}_vs_{ver_prev}.csv")
  write_csv(d_comp, comp_csv)
  message(glue("Wrote score comparison to {comp_csv}"))

  # wide diff table
  d_comp_wide <- d_comp |>
    select(`Program Area`, Component, diff) |>
    pivot_wider(
      names_from = Component,
      values_from = diff
    ) |>
    arrange(`Program Area`)

  # wide prev/curr/diff pivot (like Excel comparison) ----
  # only rows where Score differs; columns: Score_v4b, Score_v4, Score_diff, ...
  pra_diff <- d_comp |>
    filter(Component == "Score", diff != 0) |>
    pull(`Program Area`)

  score_diff_col <- glue("Score_diff")

  if (length(pra_diff) > 0) {
    d_comp_pvt <- d_comp |>
      filter(`Program Area` %in% pra_diff) |>
      pivot_wider(
        names_from = Component,
        values_from = c(all_of(v_p), all_of(v_c), diff),
        names_glue = "{Component}_{.value}"
      ) |>
      # rename ver columns: Score_v4 / Score_v4b / Score_diff
      rename_with(
        ~ str_replace(.x, glue("_{v_p}$"), glue("_{v_p}")),
        ends_with(glue("_{v_p}"))
      ) |>
      rename_with(
        ~ str_replace(.x, glue("_{v_c}$"), glue("_{v_c}")),
        ends_with(glue("_{v_c}"))
      ) |>
      select(
        `Program Area`,
        starts_with("Score_"),
        !starts_with("Score_")
      ) |>
      arrange(desc(.data[[score_diff_col]]))

    pvt_xlsx <- glue(
      "{dir_v}/tbl_pra_scores_{ver_curr}_vs_{ver_prev}_pivot.xlsx"
    )
    write_xlsx(d_comp_pvt, pvt_xlsx)
    message(glue("Wrote pivot comparison to {pvt_xlsx}"))
  } else {
    d_comp_pvt <- NULL
    message(glue("No score differences between {ver_curr} and {ver_prev}"))
  }

  list(
    d_comp = d_comp,
    d_comp_wide = d_comp_wide,
    d_comp_pvt = d_comp_pvt,
    comp_cols = comp_cols,
    ver_curr = ver_curr,
    ver_prev = ver_prev
  )
}

# compare ver vs ver_prev (v4b vs v4) ----
res <- compare_scores(ver, ver_prev)
if (!is.null(res)) {
  res$d_comp_wide |>
    DT::datatable(
      caption = glue("Score differences ({res$ver_curr} - {res$ver_prev})"),
      options = list(pageLength = 25, dom = "t")
    ) |>
    DT::formatRound(columns = res$comp_cols, digits = 0)
}
```

```{r}
#| label: compare_scores_prev_pivot
#| eval: !expr do_summary_figs

# pivot table: rows where scores differ (ver vs ver_prev) ----
if (!is.null(res) && !is.null(res$d_comp_pvt)) {
  res$d_comp_pvt |>
    DT::datatable(
      caption = glue(
        "Score pivot where different ({res$ver_curr} vs {res$ver_prev})"
      ),
      options = list(pageLength = 25, dom = "t", scrollX = TRUE)
    ) |>
    DT::formatRound(
      columns = setdiff(names(res$d_comp_pvt), "Program Area"),
      digits = 0
    )
}
```

```{r}
#| label: compare_scores_v3
#| eval: !expr do_summary_figs

# compare ver vs v3 ----
res_v3 <- compare_scores(ver, "v3")
if (!is.null(res_v3)) {
  res_v3$d_comp_wide |>
    DT::datatable(
      caption = glue(
        "Score differences ({res_v3$ver_curr} - {res_v3$ver_prev})"
      ),
      options = list(pageLength = 25, dom = "t")
    ) |>
    DT::formatRound(columns = res_v3$comp_cols, digits = 0)
}
```

```{r}
#| label: compare_scores_v3_pivot
#| eval: !expr do_summary_figs

# pivot table: rows where scores differ (ver vs v3) ----
if (!is.null(res_v3) && !is.null(res_v3$d_comp_pvt)) {
  res_v3$d_comp_pvt |>
    DT::datatable(
      caption = glue(
        "Score pivot where different ({res_v3$ver_curr} vs {res_v3$ver_prev})"
      ),
      options = list(pageLength = 25, dom = "t", scrollX = TRUE)
    ) |>
    DT::formatRound(
      columns = setdiff(names(res_v3$d_comp_pvt), "Program Area"),
      digits = 0
    )
}
```

### Multi-version comparison pivot

```{r}
#| label: compare_scores_multi
#| eval: !expr do_summary_figs

# multi-version pivot: Score and Turtle across v3, v4, v4b ----
vers_compare <- c("v3", "v4", ver)
csvs <- glue("{dir_derived}/{vers_compare}/tbl_pra_scores_{vers_compare}.csv")
names(csvs) <- vers_compare

if (all(file_exists(csvs))) {
  # read all versions into one long table
  d_all <- imap_dfr(csvs, \(f, v) {
    read_csv(f, show_col_types = FALSE) |>
      rename(`Area (1,000 km2)` = `Area (1,000 km^2^)`) |>
      select(`Program Area`, Score, Turtle) |>
      mutate(ver = v)
  })

  # pivot wide: Score_v3, Score_v4, Score_v4b, Turtle_v3, ...
  d_wide <- d_all |>
    pivot_wider(
      names_from = ver,
      values_from = c(Score, Turtle),
      names_glue = "{.value}_{ver}"
    )

  # add diff columns
  d_wide <- d_wide |>
    mutate(
      Score_v4_v3 = Score_v4 - Score_v3,
      Score_v4b_v3 = Score_v4b - Score_v3,
      Turtle_v4_v3 = Turtle_v4 - Turtle_v3,
      Turtle_v4b_v3 = Turtle_v4b - Turtle_v3
    ) |>
    select(
      `Program Area`,
      Score_v3,
      Score_v4,
      Score_v4_v3,
      Score_v4b,
      Score_v4b_v3,
      Turtle_v3,
      Turtle_v4,
      Turtle_v4_v3,
      Turtle_v4b,
      Turtle_v4b_v3
    ) |>
    arrange(desc(Score_v4b_v3))

  # write xlsx
  multi_xlsx <- glue("{dir_v}/tbl_pra_scores_v3_vs_v4_v4b.xlsx")
  write_xlsx(d_wide, multi_xlsx)
  message(glue("Wrote multi-version comparison to {multi_xlsx}"))

  # display
  num_cols <- setdiff(names(d_wide), "Program Area")
  d_wide |>
    DT::datatable(
      caption = "Program Area scores: v3 vs v4 vs v4b (Score and Turtle)",
      options = list(pageLength = 25, dom = "t", scrollX = TRUE)
    ) |>
    DT::formatRound(columns = num_cols, digits = 0) |>
    DT::formatStyle(
      columns = num_cols[str_detect(num_cols, "_v[34]")],
      backgroundColor = DT::styleInterval(
        c(-0.5, 0.5),
        c("#f4cccc", "white", "#d9ead3")
      )
    )
}
```

### Program Area flower plots

```{r}
#| label: plot_programarea_flowers
#| eval: !expr do_summary_figs

# pra list from gpkg (previously from PostGIS)
pra_all <- read_sf(pra_gpkg) |>
  st_drop_geometry() |>
  select(programarea_key, programarea_name)

# components
metric_input_keys <- c(
  glue("extrisk_{sp_cats}_ecoregion_rescaled"),
  "primprod_ecoregion_rescaled"
)
m_input_seqs <- tbl(con_sdm, "metric") |>
  filter(metric_key %in% metric_input_keys) |>
  pull(metric_seq)

# build components from metric table
components <- d_metric_labels |>
  filter(metric_key %in% flds, metric_title != "Score") |>
  pull(metric_title)
component_cols <- setNames(
  hue_pal()(length(components)),
  components
)

# d_scores
m_score <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
m_key_score <- tbl(con_sdm, "metric") |>
  filter(metric_key == !!m_score) |>
  pull(metric_seq)

d_scores <- tbl(con_sdm, "zone") |>
  filter(
    tbl == !!tbl_pra,
    fld == "programarea_key",
    value %in% pra_all$programarea_key
  ) |>
  select(programarea_key = value, zone_seq) |>
  inner_join(
    tbl(con_sdm, "zone_metric") |>
      filter(metric_seq == !!m_key_score),
    by = join_by(zone_seq)
  ) |>
  select(programarea_key, score = value) |>
  mutate(score = round(score)) |>
  collect() |>
  arrange(desc(score), programarea_key) |>
  left_join(
    pra_all,
    by = "programarea_key"
  )

d_scores$programarea_lbl <- factor(
  str_wrap(d_scores$programarea_name, width = 20),
  levels = str_wrap(d_scores$programarea_name, width = 20),
  ordered = T
)

# d_fl
d_fl <- tbl(con_sdm, "zone") |>
  filter(
    tbl == !!tbl_pra,
    fld == "programarea_key",
    value %in% pra_all$programarea_key
  ) |>
  select(programarea_key = value, zone_seq) |>
  inner_join(
    tbl(con_sdm, "zone_metric"),
    by = "zone_seq"
  ) |>
  inner_join(
    tbl(con_sdm, "metric") |>
      filter(metric_seq %in% m_input_seqs) |>
      select(metric_seq, metric_key),
    by = "metric_seq"
  ) |>
  select(programarea_key, metric_key, value) |>
  arrange(programarea_key, metric_key) |>
  collect() |>
  mutate(
    component = metric_key |>
      str_replace("extrisk_", "") |>
      str_replace("_ecoregion_rescaled", "") |>
      str_replace("_", " ")
  ) |>
  select(programarea_key, component, value) |>
  arrange(programarea_key, component, value)

# add missing components with value 0
d_0 <- expand_grid(
  programarea_key = d_scores$programarea_key,
  component = sort(unique(d_fl$component)),
  value = 0
) |>
  anti_join(
    d_fl,
    by = join_by(programarea_key, component)
  )

d_fl <- d_fl |>
  bind_rows(d_0) |>
  arrange(programarea_key, component) |>
  mutate(even = 1)

# build component display name lookup from metric table
component_labels <- d_metric_labels |>
  filter(metric_key %in% flds, metric_title != "Score") |>
  mutate(
    component = metric_key |>
      str_replace("extrisk_", "") |>
      str_replace("_ecoregion_rescaled", "") |>
      str_replace("_", " ")
  ) |>
  pull(metric_title, name = component)

d_fl <- d_fl |>
  left_join(
    d_scores |> select(programarea_key, programarea_lbl),
    by = "programarea_key"
  ) |>
  mutate(
    component = component_labels[component]
  ) |>
  arrange(programarea_lbl, component)

# all flowers
d <- d_fl |>
  arrange(programarea_lbl, component) |>
  group_by(programarea_lbl) |>
  mutate(across(!where(is.character), as.double)) |>
  mutate(
    ymax = cumsum(even),
    ymin = lag(ymax, default = 0),
    xmax = value,
    xmin = 0
  )

for (theme_name in c("minimal", "linedraw", "min_line", "dark", "void")) {
  theme_fxn <- switch(
    theme_name,
    "minimal" = theme_minimal,
    "linedraw" = theme_linedraw,
    "min_line" = function(...) {
      theme_minimal(...) %+replace%
        theme(
          panel.grid = element_line(colour = "black"),
          panel.grid.major = element_line(linewidth = rel(0.1)),
          panel.grid.minor = element_line(linewidth = rel(0.05))
        )
    },
    "dark" = theme_dark,
    "void" = theme_void
  )

  g <- d |>
    ggplot() +
    geom_rect_interactive(
      aes(
        xmin = xmin,
        xmax = xmax,
        ymin = ymin,
        ymax = ymax,
        fill = component,
        color = "white",
        data_id = component
      ),
      color = "white",
      alpha = 1
    ) +
    scale_fill_manual(
      values = component_cols,
      name = "Component"
    ) +
    facet_wrap(~programarea_lbl) +
    coord_polar(theta = "y") +
    xlim(c(-20, max(d_scores$score, d$xmax))) +
    annotate(
      "rect",
      xmin = -Inf,
      xmax = 0,
      ymin = -Inf,
      ymax = Inf,
      fill = ifelse(theme_name == "dark", "grey20", "white"),
      color = NA
    ) +
    geom_text(
      aes(x, y, label = score),
      data = d_scores |> mutate(x = -20, y = 0),
      size = 2.5,
      fontface = "bold",
      color = ifelse(theme_name == "dark", "white", "black")
    ) +
    theme_fxn() +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6),
      plot.margin = unit(c(1, 1, 1, 1), "pt"),
      strip.text.x = element_text(size = 6),
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )

  pra_fl_png <- glue(
    "{dir_figs}/programarea_flower-plot_scores_{theme_name}{v_sfx}.png"
  )
  pra_fl_pdf <- glue(
    "{dir_figs}/programarea_flower-plot_scores_{theme_name}{v_sfx}.pdf"
  )
  png(pra_fl_png, width = 1200, height = 800, res = 150)
  print(g)
  dev.off()
  pdf(pra_fl_pdf, width = 7, height = 5)
  print(g)
  dev.off()
}
```

### Primary productivity by Program Area

```{r}
#| label: plot_programarea_primprod
#| eval: !expr do_pra_primprod

# zonal NPP from VGPM rasters, adapted from ingest_productivity.qmd pa_zonal_plot
dir_vgpm_yrs <- glue(
  "{dir_data}/raw/oregonstate.edu/vgpm.r2022.v.chl.v.sst.2160x4320"
)

pra <- read_sf(pra_gpkg)

r_yrs <- dir_ls(dir_vgpm_yrs, glob = "*.tif") |>
  rast()
names(r_yrs) <- glue("npp_{time(r_yrs)}")

d_pra_npp <- r_yrs |>
  zonal(
    pra |>
      select(programarea_key, programarea_name) |>
      vect(),
    fun = "mean",
    na.rm = T,
    as.polygons = T
  ) |>
  st_as_sf() |>
  st_drop_geometry() |>
  tibble() |>
  pivot_longer(
    cols = starts_with("npp_"),
    names_to = "year",
    values_to = "npp"
  ) |>
  filter(!is.na(npp)) |>
  mutate(
    year = str_replace(year, "npp_", "") |> as.integer(),
    npp = set_units(npp, mg / m^2 / day) |>
      set_units(t / km^2 / yr)
  ) |>
  group_by(programarea_key, programarea_name) |>
  summarize(
    npp_avg = mean(npp, na.rm = T),
    npp_sd = sd(npp, na.rm = T),
    .groups = "drop"
  ) |>
  arrange(desc(npp_avg)) |>
  mutate(
    npp_avg = npp_avg |> drop_units(),
    npp_sd = npp_sd
  ) |>
  mutate(
    programarea_name = factor(programarea_name, levels = programarea_name)
  )

write_csv(d_pra_npp, pp_csv)

g <- d_pra_npp |>
  ggplot(aes(x = programarea_name, y = npp_avg)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    fill = "darkgreen"
  ) +
  geom_errorbar(
    aes(ymin = npp_avg - npp_sd, ymax = npp_avg + npp_sd),
    width = .4
  ) +
  labs(
    title = NULL,
    x = NULL,
    y = expression("Areal NPP (metric tons C km"^-2 ~ "yr"^-1 * ")")
  ) +
  scale_y_continuous(expand = c(0, 20)) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 1,
      hjust = 1
    )
  )

png(pp_png, width = 1200, height = 800, res = 150)
print(g)
dev.off()
pdf(pp_pdf, width = 7, height = 5)
print(g)
dev.off()
```

### Copy figures to `dir_v`

```{r}
#| label: copy_figs_to_dir_v
#| eval: !expr do_summary_figs

# data files are already written directly to dir_v;
# copy rendered figures from local figs/ into dir_v
fig_files <- dir_ls(dir_figs, regexp = "\\.(png|pdf|docx|xlsx|csv)$")
for (f in fig_files) {
  file_copy(f, glue("{dir_v}/{path_file(f)}"), overwrite = T)
}

message(glue("Copied {length(fig_files)} figure files to {dir_v}"))
```

## Generate README

Programmatically generate `README.md` and `README.html` for the versioned output directory. Uses `data/derived_products.csv` for file descriptions.

```{r}
#| label: generate_readme
#| eval: !expr do_readme

dp_csv <- here("data/derived_products.csv")
stopifnot(file_exists(dp_csv))

# read and expand {ver} placeholders in output_file and dependencies
d_dp <- read_csv(dp_csv) |>
  rowwise() |>
  mutate(
    output_file = glue(output_file),
    dependencies = glue(dependencies)
  ) |>
  ungroup()

# list actual files in dir_v (non-recursive, exclude README itself)
v_files <- dir_ls(dir_v, recurse = F) |>
  path_file() |>
  str_subset("^README", negate = T) |>
  # exclude ~$ temp files from Word/Excel
  str_subset("^~\\$", negate = T)

# match files to descriptions
d_files <- tibble(file = v_files) |>
  left_join(
    d_dp |> select(file = output_file, description),
    by = "file"
  ) |>
  mutate(
    description = ifelse(is.na(description), "", description)
  ) |>
  arrange(file)

lns_files <- d_files |>
  knitr::kable(format = "pipe") |>
  paste(collapse = '\n')

# build README.md content
readme_lines <- c(
  glue(
    "
    ---
    format:
      html:
        embed-resources: true
    ---
    
    # Derived Products — {ver}
     
    Version: `{v_sfx}` | Generated: {Sys.Date()}
    
    ## File listing

    {lns_files}

    ## Naming conventions

    - `ply_*`: polygon spatial data (GeoPackage)
    - `r_*`: raster data (GeoTIFF)
    - `tbl_*`: tabular summary data (CSV) 
    - `layers_*`: layer metadata for metrics and species
    - `zone_taxon_*`: zone × taxon matrix
    - `*_{ver}`: version {ver} model outputs
    
    ## Large files
    
    Some files exceed GitHub size limits and are stored in the 
    `big-files/{ver}/` directory (not tracked by git). 
    See `sdm_parquet/` for parquet exports.
    
    _Auto-generated by `calc_scores.qmd` on {Sys.time()}_"
  )
)

readme_md <- glue("{dir_v}/README.md")
writeLines(readme_lines, readme_md)
message(glue("Wrote {readme_md}"))

# render to HTML
quarto_render(readme_md |> path_real())
message(glue("Rendered {readme_md} to HTML"))
```


## Map score cells

```{r}
#| label: map_raster
#| eval: !expr do_map

librarian::shelf(mapgl)

# load global raster template
# r_global <- readRDS(glue("{dir_data}/derived/r_global_0.05.rds"))
cell_tif <- glue("{dir_data}/derived/r_bio-oracle_planarea.tif")

r_cell <- rast(cell_tif)

# d_metric <- tbl(con_sdm, "metric") |> collect()
# View(d_metric)
# TODO: add sequences for date_created post-export-parquet

# get species richness metric ID
# m_key <- "extrisk_all_ecoregion_rescaled"
m_key <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
m_seq <- tbl(con_sdm, "metric") |>
  filter(metric_key == !!m_key) |>
  pull(metric_seq)

# query species richness for a region
d_m <- dbGetQuery(
  con_sdm,
  glue(
    "
  SELECT
    cm.cell_id,
    cm.value
  FROM cell_metric cm
  WHERE cm.metric_seq = (
    SELECT metric_seq
    FROM metric
    WHERE metric_key = '{m_key}' )"
  )
) |>
  tibble()
stopifnot(sum(duplicated(d_m$cell_id)) == 0)

# convert to raster
r_m <- create_raster_from_df(d_m, r_cell, value_col = "value")

# display with mapgl
# TODO: make into function in msens R package
mapgl_rast <- function(
  r,
  cols_r = rev(RColorBrewer::brewer.pal(11, "Spectral")),
  layer_name = "SDM"
) {
  # r = r_m; cols_r = cols_r; layer_name = m_key

  rng_r <- minmax(r) |> as.numeric() |> signif(digits = 2)

  mapboxgl(
    style = mapbox_style("dark"),
    zoom = 3.5,
    center = c(-106, 40.1)
  ) |> # ply_pa centroid
    add_image_source(
      id = "r_src",
      data = r,
      colors = cols_r
    ) |>
    add_raster_layer(
      id = "r_lyr",
      source = "r_src",
      raster_resampling = "nearest"
    ) |>
    mapgl::add_legend(
      layer_name,
      values = rng_r,
      colors = cols_r,
      position = "bottom-right"
    ) |>
    add_fullscreen_control(position = "top-left") |>
    add_navigation_control() |>
    add_scale_control()
}

# visualize
mapgl_rast(r_m, layer_name = m_key)
```

## Parquet export/import

This cleans up the DuckDB database by exporting tables to Parquet files (with sort order for faster reads) and then re-importing them to a new DuckDB database. This also allows us to remove indexes from the DuckDB database, which are not needed for the mapgl app and can slow down writes.

### Export to Parquet

```{r}
#| label: export_parquet
#| eval: !expr do_parquet

# TODO: model_cell: PARTITION BY (mdl_seq);") # https://duckdb.org/docs/stable/data/parquet/overview

source(here("libs/sdm_functions.R"))
con_sdm <- dbConnect(duckdb(), dbdir = sdm_db, read_only = F) |>
  load_duckdb_extensions()

# dbListTables(con_sdm)
dbExecute(con_sdm, "SET threads = 7;")
dbExecute(con_sdm, "SET memory_limit = '20GB';")

dir_pq <- normalizePath(dir_pq)
if (dir_exists(dir_pq)) {
  dir_delete(dir_pq)
}
dir_create(dir_pq)

# OLD: add primary keys, indexes
# NEW: skip indexes, sort on export, per [performance](https://duckdb.org/docs/stable/guides/performance/indexing.html)

# dbRemoveTable(con_sdm, "cell_model", force = T)
tbl_keys <- list(
  cell = "cell_id",
  cell_metric = c("cell_id", "metric_seq"),
  # cell_model  = c("cell_id", "mdl_seq"),   # source: model_cell; alternative key sorting order
  dataset = "ds_key",
  listing = "spp_id",
  metric = "metric_seq",
  model = "mdl_seq",
  model_cell = c("mdl_seq", "cell_id"), # *big* one
  species = "sp_seq",
  taxon = "taxon_id",
  taxon_model = c("taxon_id", "ds_key"),
  zone = "zone_seq",
  zone_cell = c("zone_seq", "cell_id"),
  zone_metric = c("zone_seq", "metric_seq"),
  zone_taxon = c("zone_fld", "zone_value", "mdl_seq", "taxon_id")
)
stopifnot(length(setdiff(names(tbl_keys), dbListTables(con_sdm))) == 0)
stopifnot(length(setdiff(dbListTables(con_sdm), names(tbl_keys))) == 0)

# export parquet files with sort order
for (tbl in names(tbl_keys)) {
  # tbl = names(tbl_keys)[1]  # tbl = "cell_model"
  keys <- tbl_keys[[tbl]]
  keys_str <- paste(keys, collapse = ', ')
  pq_file <- glue("{dir_pq}/{tbl}.parquet")

  message(glue(
    "Exporting {tbl} (ORDER BY {keys_str}) to parquet  ~  {Sys.time()}"
  ))

  fro_tbl <- tbl
  if (tbl == "cell_model") {
    fro_tbl <- "model_cell"
  }
  stopifnot(all(keys %in% dbListFields(con_sdm, fro_tbl)))

  if (tbl %in% c("cell_model", "model_cell")) {
    flds <- dbListFields(con_sdm, fro_tbl) |>
      setdiff(keys)
    flds_str <- paste(flds, collapse = '\", \"')
    sql <- glue(
      "COPY (
      SELECT {keys_str}, {flds_str} FROM {fro_tbl} 
      ORDER BY {keys_str}) 
      TO '{pq_file}' 
    (FORMAT parquet, PER_THREAD_OUTPUT, COMPRESSION zstd) "
    )
  } else {
    sql <- glue(
      "COPY (
      SELECT * FROM {fro_tbl} 
      ORDER BY {keys_str}) 
      TO '{pq_file}' 
    (FORMAT parquet, PER_THREAD_OUTPUT, COMPRESSION zstd) "
    )
  }
  message(sql)
  dbExecute(con_sdm, sql)
}
message(glue("Exporting finished  ~  {Sys.time()}")) # ~ 1 min

dbDisconnect(con_sdm, shutdown = T)
rm(con_sdm)
```

### Import from Parquet

```{r}
#| label: import_parquet
#| eval: !expr do_parquet

# dir_pq and sdm_db defined in setup chunk

# close any existing connection and force GC to clean up orphaned drivers
if (exists("con_sdm") && dbIsValid(con_sdm)) {
  dbDisconnect(con_sdm, shutdown = T)
}
if (exists("con_sdm")) {
  rm(con_sdm)
}
gc() # flush orphaned duckdb driver instances from earlier chunks

if (file_exists(sdm_db)) {
  file_delete(sdm_db)
}
# also remove stale WAL file
sdm_wal <- paste0(sdm_db, ".wal")
if (file_exists(sdm_wal)) {
  file_delete(sdm_wal)
}

source(here("libs/sdm_functions.R"))
con_sdm <- dbConnect(duckdb(), dbdir = sdm_db, read_only = F) |>
  load_duckdb_extensions()

# dbExecute(con_sdm, "SET threads = 8;")
# dbExecute(con_sdm, "SET memory_limit = '20GB';")

# import parquet ----
for (pq in dir_ls(dir_pq, glob = "*.parquet")) {
  # pq = dir_ls(dir_pq, glob = "*.parquet")[2]
  tbl <- path_file(pq) |> str_remove("\\.parquet$")
  if (!dbExistsTable(con_sdm, tbl)) {
    message(glue("Importing {tbl} from {basename(pq)}/*.parquet"))
    dbExecute(
      con_sdm,
      glue("CREATE TABLE {tbl} AS SELECT * FROM '{pq}/*.parquet'")
    )
  }
}

# create index on 2nd column of multi-column index from parquet export
# tbl_idx <- list(
#   cell_metric = "metric_seq",
#   model_cell  = "mdl_seq",
#   zone_cell   = "cell_id",
#   zone_metric = "metric_seq")
# for (tbl in names(tbl_idx)){      # tbl = "model_cell"
#   fld <- tbl_idx[[tbl]]
#   idx <- glue("idx_{tbl}_{fld}")  # idx_model_cell_mdl_seq
#   stopifnot(fld %in% dbListFields(con_sdm, tbl))
#
#   sql <- glue("CREATE INDEX IF NOT EXISTS {idx} ON {tbl} ({fld});")
#   message(sql)
#   dbExecute(con_sdm, sql)
# }

# sequences ----
# for auto-incrementing primary keys
tbl_seq <- list(
  metric = "metric_seq",
  model = "mdl_seq",
  species = "sp_seq",
  zone = "zone_seq"
)
stopifnot(all(names(tbl_seq) %in% dbListTables(con_sdm)))

for (tbl in names(tbl_seq)) {
  # tbl = "metric"
  fld <- tbl_seq[[tbl]]
  id <- glue("seq_{tbl}")

  nxt <- dbGetQuery(con_sdm, glue("SELECT MAX({fld}) FROM {tbl}")) |>
    pull(1) +
    1

  dbExecute(con_sdm, glue("DROP SEQUENCE IF EXISTS {id}"))
  sql <- glue("CREATE SEQUENCE {id} START {nxt}")
  message(sql)
  dbExecute(con_sdm, sql)
  sql <- glue(
    "ALTER TABLE {tbl} ALTER COLUMN {fld} SET DEFAULT nextval('{id}')"
  )
  message(sql)
  dbExecute(con_sdm, sql)
}

# date_created ----
# default to current_date
for (tbl in dbListTables(con_sdm)) {
  # tbl = "dataset"
  fld <- dbListFields(con_sdm, tbl) |>
    str_subset("date_created")
  if (length(fld) == 1) {
    sql <- glue(
      "ALTER TABLE {tbl} ALTER COLUMN date_created SET DEFAULT CURRENT_DATE;"
    )
    message(sql)
    a <- dbExecute(con_sdm, sql)
  }
}

# checkpoint WAL to main DB file, then shut down driver cleanly
dbExecute(con_sdm, "CHECKPOINT")
dbDisconnect(con_sdm, shutdown = T)
rm(con_sdm)
gc() # flush duckdb driver finalizer so file handles are fully released
message(glue(
  "sdm.duckdb exists: {file_exists(sdm_db)}  size: {file_size(sdm_db)}"
))
```

## Deploy to server

Sync derived data and app code to the msens server.

```{r}
#| label: deploy
#| eval: !expr do_deploy

deploy_sh <- here("libs/deploy_to_server.sh")
stopifnot(file_exists(deploy_sh))

# duckdb shutdown may need a moment for file handles to release
if (!file_exists(sdm_db)) {
  gc() # flush any orphaned duckdb driver finalizers
  Sys.sleep(2)
}
stopifnot(file_exists(sdm_db))
system2("bash", args = c(deploy_sh, ver), wait = T)
```

## Close Connection

```{r}
#| label: cleanup
if (exists("con_sdm") && dbIsValid(con_sdm)) {
  dbDisconnect(con_sdm, shutdown = T)
}
```
