---
title: "Explore marine mammal vulnerabilities to climate change in U.S. Atlantic (Lettrich et al, 2023)"
editor_options: 
  chunk_output_type: console
---

## Introduction

This report by Lettrich et al. [-@lettrichVulnerabilityClimateChange2023] includes 3 Supplemental tables:

-   S1 Dataset. **Raw attribute and factor scores**. \
    `s001.csv`\
    Expert scores for each exposure factor and sensitivity attribute aggregated by stock.

-   S1 Table. **Stocks scored**.\
    (XLSX) List of all marine mammal stocks and stock groupings in the assessment.\
    `s005.xlsx`\
    S1. Marine mammal stocks scored in the assessment of climate vulnerability. Status based on NOAA Fisheries 2020 Stock Assessment Reports (Hayes et al. 2021). Status: E - Endangered under ESA, D - Depleted under MMPA, S - Strategic under MMPA. GOMx - Gulf of Mexico. WNA - Western North Atlantic. BSE - Bay, sound, and estuary.

-   S2 Table. **Stock exposure, sensitivity, and vulnerability results summary table**.\
    `s006.xlsx`\
    S3. Stock vulnerability, exposure, sensitivity, data quality, and response category scores of 108 stocks of marine mammals from the western North Atlantic, Gulf of Mexico, and Caribbean Sea. Status based on NOAA Fisheries 2020 Stock Assessment Reports (Hayes et al. 2021). Status: E - Endangered under ESA, D - Depleted under MMPA, S - Strategic under MMPA. GOMx - Gulf of Mexico. WNA - Western North Atlantic. BSE - Bay, sound, and estuary.							

## S1 Dataset. Raw attribute and factor scores

```{r}
#| label: setup
#| warning: false

librarian::shelf(
  dplyr, DT, fs, glue, here, knitr, leaflet, mapview,
  MarineSensitivity/msens,
  purrr, readr, readxl, sf, stringr, terra, tibble, tidyr, units,
  quiet = T)

dir_data  <- "/Users/bbest/My Drive/projects/msens/data"
dir_suppl <- glue("{dir_data}/raw/studies/lettrich2023/suppl")
d_csv     <- glue("{dir_suppl}/s001.csv")
t1_xls    <- glue("{dir_suppl}/s005.xlsx")
t2_xls    <- glue("{dir_suppl}/s006.xlsx")

d <- read_csv(d_csv)

n_StockName     <- length(unique(d$StockName))
n_AttributeName <- length(unique(d$AttributeName))

datatable(d) |> 
  formatRound(columns=c('MeanScore', 'MeanDataQuality'), digits=1)
```

- `n_StockName`:  `r n_StockName`
- `n_AttributeName`:  `r n_AttributeName`
- `nrow(d)`:  `r nrow(d)` (`n_StockName` * `n_AttributeName`)

### `StockName`

```{r}
d |> 
  group_by(StockName) |>
  summarize(n = n(), .groups = "drop") |>
  datatable()
```

### `FunctionalGroup`

```{r}
d |> 
  group_by(FunctionalGroup) |>
  summarize(n = n(), .groups = "drop") |>
  datatable()
```

### `Attribute`

```{r}
d |> 
  group_by(Attribute) |>
  summarize(n = n(), .groups = "drop") |>
  datatable()
```

### `AttributeName`

```{r}
d |> 
  group_by(Attribute, AttributeName) |> 
  summarize(n = n(), .groups = "drop") |>
  datatable()
```

## S1 Table. Stocks scored

List of all marine mammal stocks and stock groupings in the assessment.

`s005.xlsx`

S1. Marine mammal stocks scored in the assessment of climate vulnerability. Status based on NOAA Fisheries 2020 Stock Assessment Reports (Hayes et al. 2021). 

Status:

- E: Endangered under ESA
- D: Depleted under MMPA
- S: Strategic under MMPA

Places:

- GOMx: Gulf of Mexico
- WNA: Western North Atlantic
- BSE: Bay, sound, and estuary

```{r}
#| label: t1_xls

t1 <- read_excel(t1_xls, skip = 1)

datatable(t1)
```

## S2 Table. Stock exposure, sensitivity, and vulnerability results summary table

`s006.xlsx`

S3. Stock vulnerability, exposure, sensitivity, data quality, and response category scores of 108 stocks of marine mammals from the western North Atlantic, Gulf of Mexico, and Caribbean Sea. Status based on NOAA Fisheries 2020 Stock Assessment Reports (Hayes et al. 2021). 

Status:

- E: Endangered under ESA
- D: Depleted under MMPA
- S: Strategic under MMPA

Places:

- GOMx: Gulf of Mexico
- WNA: Western North Atlantic
- BSE: Bay, sound, and estuary

```{r}
#| label: t2_xls

t2 <- read_excel(t2_xls, skip = 1)

datatable(t2)
```


## References {.unnumbered}
