---
title: "GoA Marine Mammal Species Review"
subtitle: "Response to SEFSC feedback on 12 species in Eastern Gulf (v2 → v6)"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Executive Summary

NOAA SEFSC flagged 12 marine mammal species that the Marine Sensitivity Tool
(MST) showed as occurring in the Eastern Gulf of America but shouldn't be
there. **As of v6, all 12 species are resolved** — none remain erroneously
in the Gulf:

| Resolution | # | Species | Mechanism |
|---|---|---|---|
| **Removed from GOA (v5)** | 7 | Hooded Seal, Long-Finned Pilot Whale, Gray Seal, Northern Bottlenose Whale, White-Beaked Dolphin, Sowerby's Beaked Whale, Harp Seal | IUCN range mask clipped AquaMaps to expert range (outside Gulf) |
| **Excluded as duplicate (v5)** | 1 | *B. brydei* | Taxonomic synonym of *B. edeni*; `is_ok=FALSE` |
| **Reduced GOA footprint (v5)** | 2 | *B. edeni* (GAA+GAB only), *P. phocoena* (GOA only) | IUCN range mask trimmed to actual Gulf extent |
| **Removed from GOA (v6)** | 1 | Harbor Seal (*Phoca vitulina*) | IUCN range from newer `MAMMALS.zip` download; range is Atlantic/Pacific coasts only |
| **Excluded — range outside EEZ (v6)** | 1 | Guiana Dolphin (*Sotalia guianensis*) | IUCN range produces 0 cells in US EEZ; merged model truncated, `is_ok=FALSE` |

The two species that remained problematic through v5 were resolved in v6 by:

1. Ingesting 3 missing mammal IUCN ranges from a newer `MAMMALS.zip` download
   (the original `MAMMALS_MARINE_ONLY.zip` and
   `MAMMALS_MARINE_AND_TERRESTRIAL.zip` were incomplete).
2. Excluding 371 species (343 fish, 20 invertebrates, 6 corals, 2 mammals)
   whose IUCN ranges fall entirely outside the US EEZ — their AquaMaps
   predictions in US waters were edge-of-range artifacts with near-zero
   suitability.

See [investigate_iucn_masking.qmd](investigate_iucn_masking.qmd) for the
broader impact analysis across all taxa groups.

```{r}
#| label: setup
#| warning: false

librarian::shelf(
  DBI,
  dplyr,
  duckdb,
  DT,
  fs,
  glue,
  here,
  htmltools,
  knitr,
  purrr,
  readr,
  stringr,
  tibble,
  tidyr,
  webshot2,
  quiet = T)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

source(here("libs/paths.R"))

# v6 database (current version, from paths.R)
con_sdm <- dbConnect(duckdb(dbdir = sdm_db, read_only = TRUE))

# v2 database (outside dir_big/derived on laptop)
sdm_db_v2 <- ifelse(
  is_server,
  "/share/data/big/v2/sdm.duckdb",
  "~/_big/msens/v2/sdm.duckdb")
con_sdm_v2 <- dbConnect(duckdb(dbdir = sdm_db_v2, read_only = TRUE))

# output directory for screenshots
dir_figs <- here("figs/explore_goa-marmam")
dir_create(dir_figs)

# goa program area codes
goa_zones <- c("GAA", "GAB", "GOA")

# zone table names per version
tbl_pra_v2 <- "ply_programareas_2026"
# tbl_pra (v6) already defined by paths.R
```

## Background

NOAA SEFSC (Avery Paxton, 2026-01-29) flagged 12 marine mammal species:

> One of my action items was to share the list of other marine mammals that the
> MST shows as occurring in the Eastern Gulf planning area (that's where we
> focused our initial review) that don't currently occur there. There were 12
> more (13 if include walrus) of 44 marine mammals.

**References:**

- [GitHub apps#5 comment](https://github.com/MarineSensitivity/apps/issues/5#issuecomment-4118545339)
  — Harbor seal IUCN range map issue
- [Google Doc notes](https://docs.google.com/document/d/1s6HWNkyethIL5-VjE3tBGu9o_YIu3oFcEutw7ouN1Bs/edit?usp=sharing)
- [SEFSC meeting notes](dev/2026-02-05_v3.md)
- [investigate_iucn_masking.qmd](investigate_iucn_masking.qmd)
  — quantifies impact of 0-cell IUCN masking across all taxa groups

## Resolution Table

```{r}
#| label: resolution_table

d_spp <- tibble::tribble(
  ~scientific_name,              ~common_name,                ~v2_mdl_seq,
  "Balaenoptera brydei",         "Bryde's whale",             295L,
  "Balaenoptera edeni",          "Bryde's Whale",             170L,
  "Cystophora cristata",         "Hooded Seal",               1808L,
  "Globicephala melas",          "Long-Finned Pilot Whale",   1434L,
  "Halichoerus grypus",          "Gray Seal",                 4454L,
  "Hyperoodon ampullatus",       "Northern Bottlenose Whale", 1354L,
  "Lagenorhynchus albirostris",  "White-Beaked Dolphin",      1679L,
  "Mesoplodon bidens",           "Sowerby's Beaked Whale",    1308L,
  "Pagophilus groenlandicus",    "Harp Seal",                 3246L,
  "Phoca vitulina",              "Harbor Seal",               770L,
  "Phocoena phocoena",           "Harbor Porpoise",           593L,
  "Sotalia guianensis",          "Guiana dolphin",            9013L)

# look up v6 taxon info
d_taxon_spp <- tbl(con_sdm, "taxon") |>
  filter(scientific_name %in% !!d_spp$scientific_name) |>
  collect()

d_spp <- d_spp |>
  left_join(
    d_taxon_spp |>
      select(
        scientific_name, taxon_id,
        v6_mdl_seq = mdl_seq,
        is_ok, is_mmpa, er_score,
        extrisk_code, redlist_code, n_ds),
    by = "scientific_name")

# v2 and v6 GOA zone_taxon
d_goa_v2 <- dbGetQuery(con_sdm_v2, glue(
  "SELECT sp_scientific, zone_value, avg_suit, rl_score, suit_rl, area_km2
   FROM zone_taxon
   WHERE zone_tbl = '{tbl_pra_v2}'
     AND zone_value IN ('GAA', 'GAB', 'GOA')
     AND sp_scientific IN ({paste0(\"'\", d_spp$scientific_name, \"'\", collapse = ', ')})
   ORDER BY sp_scientific, zone_value")) |>
  as_tibble()

d_goa_v6 <- dbGetQuery(con_sdm, glue(
  "SELECT sp_scientific, zone_value, avg_suit, er_score, suit_rl, area_km2
   FROM zone_taxon
   WHERE zone_tbl = '{tbl_pra}'
     AND zone_value IN ('GAA', 'GAB', 'GOA')
     AND sp_scientific IN ({paste0(\"'\", d_spp$scientific_name, \"'\", collapse = ', ')})
   ORDER BY sp_scientific, zone_value")) |>
  as_tibble()

# rng_iucn cell counts per species
d_iucn_cells <- tbl(con_sdm, "taxon_model") |>
  filter(
    ds_key   == "rng_iucn",
    taxon_id %in% !!d_spp$taxon_id) |>
  left_join(
    tbl(con_sdm, "model_cell") |>
      group_by(mdl_seq) |>
      summarize(n_iucn_cells = n(), .groups = "drop"),
    by = "mdl_seq") |>
  collect() |>
  mutate(n_iucn_cells = coalesce(n_iucn_cells, 0L)) |>
  select(taxon_id, n_iucn_cells)

# build resolution table
spp_in_goa_v6 <- d_goa_v6 |> distinct(sp_scientific) |> pull()
goa_zones_v2  <- d_goa_v2 |>
  group_by(sp_scientific) |>
  summarize(goa_zones_v2 = paste(sort(zone_value), collapse = ", "), .groups = "drop")
goa_zones_v6  <- d_goa_v6 |>
  group_by(sp_scientific) |>
  summarize(goa_zones_v6 = paste(sort(zone_value), collapse = ", "), .groups = "drop")

d_resolution <- d_spp |>
  left_join(d_iucn_cells, by = "taxon_id") |>
  left_join(goa_zones_v2, by = c("scientific_name" = "sp_scientific")) |>
  left_join(goa_zones_v6, by = c("scientific_name" = "sp_scientific")) |>
  mutate(
    in_goa_v2 = scientific_name %in% d_goa_v2$sp_scientific,
    in_goa_v6 = scientific_name %in% spp_in_goa_v6,
    resolution = case_when(
      scientific_name == "Balaenoptera brydei" ~
        "excluded (v5): taxonomic duplicate of B. edeni",
      scientific_name == "Sotalia guianensis" ~
        "excluded (v6): IUCN range outside US EEZ (0 cells)",
      scientific_name == "Phoca vitulina" ~
        "removed from GOA (v6): IUCN range ingested from MAMMALS.zip; Atlantic/Pacific only",
      scientific_name == "Balaenoptera edeni" ~
        "reduced (v5): IUCN range limits to GAA+GAB (no GOA)",
      scientific_name == "Phocoena phocoena" ~
        "reduced (v5): IUCN range limits to GOA only",
      !is_ok & !in_goa_v6 ~
        "removed from GOA (v5): IUCN range mask outside Gulf",
      in_goa_v2 & !in_goa_v6 ~
        "removed from GOA (v5): IUCN range mask outside Gulf",
      .default = "unchanged")) |>
  select(
    scientific_name, common_name,
    is_ok, n_iucn_cells,
    goa_zones_v2, goa_zones_v6,
    resolution) |>
  arrange(resolution, scientific_name)

datatable(
  d_resolution,
  caption = "resolution status for all 12 SEFSC-flagged species in v6",
  escape  = FALSE,
  filter  = "top",
  options = list(dom = "ft", pageLength = 15, scrollX = TRUE))
```

## Database Interrogation (v6)

```{r}
#| label: db_interrogate

# all models per species (per-dataset and merged)
d_models_spp <- tbl(con_sdm, "taxon_model") |>
  filter(taxon_id %in% !!d_taxon_spp$taxon_id) |>
  collect() |>
  left_join(
    d_taxon_spp |> select(taxon_id, scientific_name, common_name),
    by = "taxon_id")

# cell counts per model (left_join: 0-cell rng_iucn models have no model_cell rows)
d_counts_spp <- tbl(con_sdm, "model_cell") |>
  filter(mdl_seq %in% !!d_models_spp$mdl_seq) |>
  group_by(mdl_seq) |>
  summarize(
    n_cells = n(),
    v_min   = min(value, na.rm = TRUE),
    v_max   = max(value, na.rm = TRUE),
    .groups = "drop") |>
  collect()

d_summary_spp <- d_models_spp |>
  left_join(d_counts_spp, by = "mdl_seq") |>
  mutate(n_cells = coalesce(n_cells, 0L)) |>
  select(scientific_name, common_name, ds_key, mdl_seq, n_cells, v_min, v_max) |>
  arrange(scientific_name, ds_key)

datatable(
  d_summary_spp,
  caption = "v6 models per species by dataset (n_cells=0 means IUCN range outside US EEZ)",
  filter  = "top",
  options = list(dom = "ft", pageLength = 50))
```

### Dataset summary (wide)

```{r}
#| label: ds_wide

d_ds_wide <- d_summary_spp |>
  select(scientific_name, ds_key, n_cells) |>
  pivot_wider(
    names_from  = ds_key,
    values_from = n_cells,
    names_prefix = "n_")

d_taxon_wide <- d_taxon_spp |>
  select(
    scientific_name, common_name, sp_cat,
    er_score, is_mmpa, is_ok,
    extrisk_code, redlist_code, n_ds) |>
  left_join(d_ds_wide, by = "scientific_name")

datatable(
  d_taxon_wide,
  caption = "v6 species summary with cell counts per dataset",
  filter  = "top",
  options = list(dom = "ft", scrollX = TRUE))
```

## GOA Scores: v2 vs v6

```{r}
#| label: goa_comparison

d_goa_v2_r <- d_goa_v2 |>
  rename(er_score_v2 = rl_score, avg_suit_v2 = avg_suit,
         suit_rl_v2  = suit_rl,  area_km2_v2 = area_km2)
d_goa_v6_r <- d_goa_v6 |>
  rename(er_score_v6 = er_score, avg_suit_v6 = avg_suit,
         suit_rl_v6  = suit_rl,  area_km2_v6 = area_km2)

d_goa_cmp <- d_goa_v2_r |>
  full_join(d_goa_v6_r, by = c("sp_scientific", "zone_value")) |>
  left_join(
    d_spp |> select(scientific_name, common_name),
    by = c("sp_scientific" = "scientific_name")) |>
  relocate(common_name, .after = sp_scientific) |>
  arrange(sp_scientific, zone_value)

datatable(
  d_goa_cmp,
  caption = "GOA Program Area scores: v2 vs v6 (NA = absent from zone in that version)",
  filter  = "top",
  options = list(dom = "ft", scrollX = TRUE, pageLength = 30))
```

## Screenshots: v2 vs v6

```{r}
#| label: webshots

for (i in 1:nrow(d_spp)) {
  sp <- d_spp[i, ]
  sp_slug <- str_replace_all(sp$scientific_name, " ", "-")

  # v2 screenshot
  url_v2 <- glue("https://app.marinesensitivity.org/mapsp_v2/?mdl_seq={sp$v2_mdl_seq}")
  png_v2 <- glue("{dir_figs}/{sp_slug}_v2.png")
  if (!file.exists(png_v2)) {
    tryCatch(
      webshot(url_v2, png_v2, vwidth = 1200, vheight = 800, zoom = 2, delay = 15),
      error = function(e) message(glue("webshot failed for {sp$scientific_name} v2: {e$message}")))
  }

  # v6 screenshot (skip if no v6 mdl_seq)
  if (!is.na(sp$v6_mdl_seq)) {
    url_v6 <- glue("https://app.marinesensitivity.org/mapsp/?mdl_seq={sp$v6_mdl_seq}&splash=false")
    png_v6 <- glue("{dir_figs}/{sp_slug}_v6.png")
    if (!file.exists(png_v6)) {
      tryCatch(
        webshot(url_v6, png_v6, vwidth = 1200, vheight = 800, zoom = 2, delay = 15),
        error = function(e) message(glue("webshot failed for {sp$scientific_name} v6: {e$message}")))
    }
  }
}
```

## Per-Species Details

```{r}
#| label: species_details
#| results: asis
#| echo: false

for (i in 1:nrow(d_spp)) {
  sp <- d_spp[i, ]
  sp_slug <- str_replace_all(sp$scientific_name, " ", "-")
  res <- d_resolution |> filter(scientific_name == sp$scientific_name) |> pull(resolution)

  cat(glue("\n\n### {sp$common_name} (*{sp$scientific_name}*) {{#{sp_slug}}}\n\n"))

  if (!is.na(sp$is_ok) && !sp$is_ok) {
    cat(glue(
      "::: {{.callout-warning}}\n",
      "**Excluded in v6** (`is_ok=FALSE`): {res}\n",
      ":::\n\n"))
  } else {
    cat(glue(
      "::: {{.callout-note}}\n",
      "**Resolution**: {res}\n",
      ":::\n\n"))
  }

  cat("::: {.panel-tabset}\n\n")

  # tab: v2 vs v6 screenshots ----
  cat("#### v2 vs v6\n\n")
  cat("::: {.columns}\n")
  cat("::: {.column width='50%'}\n")
  url_v2 <- glue("https://app.marinesensitivity.org/mapsp_v2/?mdl_seq={sp$v2_mdl_seq}")
  cat(glue("**v2** ([app link]({url_v2}))\n\n"))
  png_v2 <- glue("figs/explore_goa-marmam/{sp_slug}_v2.png")
  if (file.exists(here(png_v2))) {
    cat(glue("![]({png_v2})\n\n"))
  } else {
    cat("*Screenshot not yet captured.*\n\n")
  }
  cat(":::\n")
  cat("::: {.column width='50%'}\n")
  if (!is.na(sp$v6_mdl_seq)) {
    url_v6 <- glue("https://app.marinesensitivity.org/mapsp/?mdl_seq={sp$v6_mdl_seq}&splash=false")
    cat(glue("**v6** ([app link]({url_v6}))\n\n"))
    png_v6 <- glue("figs/explore_goa-marmam/{sp_slug}_v6.png")
    if (file.exists(here(png_v6))) {
      cat(glue("![]({png_v6})\n\n"))
    } else {
      cat("*Screenshot not yet captured.*\n\n")
    }
  } else {
    cat("**v6**: no merged model (`is_ok=FALSE`)\n\n")
  }
  cat(":::\n:::\n\n")

  # tab: goa cell counts ----
  cat("#### GOA Comparison\n\n")
  d_sp_goa <- d_goa_cmp |>
    filter(sp_scientific == sp$scientific_name) |>
    select(-sp_scientific, -common_name)
  if (nrow(d_sp_goa) > 0) {
    tbl_html <- knitr::kable(d_sp_goa, digits = 4, format = "html",
                             caption = "v2 vs v6 scores in GOA Program Areas")
    cat(tbl_html)
    cat("\n\n")
  } else {
    cat("*Species not present in any GOA program area in either version.*\n\n")
  }

  # tab: database ----
  cat("#### Datasets (v6)\n\n")
  d_sp_ds <- d_summary_spp |>
    filter(scientific_name == sp$scientific_name) |>
    select(-scientific_name, -common_name)
  if (nrow(d_sp_ds) > 0) {
    tbl_html <- knitr::kable(d_sp_ds, format = "html",
                             caption = "v6 datasets and cell counts")
    cat(tbl_html)
    cat("\n\n")
  } else {
    cat("*No models in v6 database.*\n\n")
  }

  cat(":::\n\n")
}
```

```{r}
#| label: cleanup
#| include: false

dbDisconnect(con_sdm)
dbDisconnect(con_sdm_v2)
```
