---
title: "Explore OBIS species distribution models (MPA Europe) for v11"
author: "Ben Best"
date: "2026-09-15"
editor_options:
  chunk_output_type: console
---

## Why

AquaMaps 2.0 (the AquaX pipeline, ingested as `ax` in v9) cannot be redistributed: the FishBase
consortium keeps the rasters, and BOEM-funded products are expected to be open. OBIS's MPA Europe
contribution ([iobis/mpaeu_sdm](https://github.com/iobis/mpaeu_sdm), docs at
[iobis.github.io/mpaeu_docs](https://iobis.github.io/mpaeu_docs/)) is a fully open alternative: an
ensemble SDM framework (`obissdm`, Maxent + random forest + XGBoost; "ESM" ensembles of small models
for 10–29 records) fitted on **global** OBIS + GBIF occurrences and predicted to the **whole globe on
exactly our `global05` grid** (7200 × 3600, 0.05°, −180..180), published as COGs on
`s3://obis-maps/sdm/species/taxonid={AphiaID}/` with a STAC catalogue.

This notebook answers three questions with numbers, all reproducible from public sources plus the
local OBIS snapshot:

1. What species coverage does the OBIS/MPA Europe SDM set give us compared to AquaMaps 1.0 (`am`) and
   2.0 (`ax`)?
2. What is the minimum number of OBIS records the framework needs, and how many US species clear it
   (so the same framework could fill the gaps)?
3. Does an OBIS model clipped to the IUCN range (our merge rule) look like AquaX? (walrus test)

```{r}
#| label: setup
librarian::shelf(
  arrow, DBI, dplyr, duckdb, fs, glue, here, jsonlite, knitr, readr, terra, tibble,
  quiet = T)
options(readr.show_col_types = F)
source(here("libs/paths.R"))

# inputs ----
# local OBIS snapshot (the obisindicators export, 136 M rows); on the server this is the
# continuously-synced s3://obis-open-data/occurrence/ mirror (newer) — point OBIS_OCC at either
obis_occ    <- Sys.getenv("OBIS_OCC", "~/_big/obis.duckdb")          # duckdb (table occ) or a dir of parquet
dir_obismap <- glue("{dir_raw}/obis-maps")                            # cache of MPA Europe files
dir_create(dir_obismap)
mp_list_url <- "https://raw.githubusercontent.com/iobis/mpaeu_docs/main/files/sdm_list_iss_cons_20250903.csv"
mp_list_csv <- glue("{dir_obismap}/sdm_list_iss_cons_20250903.csv")
if (!file_exists(mp_list_csv)) download.file(mp_list_url, mp_list_csv, quiet = T)
aws_zip_url <- "https://github.com/iobis/mpaeu_sdm/raw/main/aws_files_list.zip"
aws_zip     <- glue("{dir_obismap}/aws_files_list.zip")
if (!file_exists(aws_zip)) download.file(aws_zip_url, aws_zip, quiet = T)
if (!file_exists(glue("{dir_obismap}/species_files.parquet"))) unzip(aws_zip, exdir = dir_obismap)

# outputs ----
obis_grid_db <- glue("{dir_derived}/obis_grid.duckdb")   # (species x cell) counts on global05
gap_csv      <- here("data/obis_sdm_gap_species.csv")    # US species with >=30 cells and no am/ax
```

## The framework's thresholds (from the code)

`mpaeu_sdm/functions/model_species.R` sets `minptslim <- 30` quality-controlled occurrence points for
the standard model and `model_species_esm.R` sets `minptslim <- 10` for the ensemble-of-small-models
path; a model is kept when its cross-validated continuous Boyce index is ≥ 0.3
(`codes/model_fit.R`: `tg_metric <- "cbi"; tg_threshold <- 0.3`). The published species list
(`sdm_list_iss_cons_20250903.csv`) records the outcome per species.

```{r}
#| label: mp-list
mp <- read_csv(mp_list_csv) |>
  rename(aphia = taxonID, mp_sdm = SDM, mp_info = SDMinfo, mp_rec = availableRecords)
mp |>
  count(mp_sdm, mp_info) |>
  left_join(
    mp |> group_by(mp_sdm, mp_info) |>
      summarise(min_rec = min(mp_rec, na.rm = T), median_rec = median(mp_rec, na.rm = T), .groups = "drop"),
    by = c("mp_sdm", "mp_info")) |>
  kable(caption = "MPA Europe species list: SDM outcome by reason, with cleaned-record counts")

sf <- read_parquet(glue("{dir_obismap}/species_files.parquet"))
sf |>
  select(taxonid, available_models) |> tidyr::unnest(available_models) |>
  group_by(taxonid) |> summarise(models = paste(sort(unique(method)), collapse = "+")) |>
  count(models, sort = T) |>
  kable(caption = "Algorithms published per species on s3://obis-maps (esm = ensemble of small models, 10-29 points)")
```

## OBIS occurrences onto the 0.05° cell grid

Species-rank records (`speciesid` = accepted WoRMS AphiaID) are binned to the `global05` cell whose
pixel index is `cell_id` (asserted against the `cell` table), then joined to `in_usa`.

```{r}
#| label: obis-grid
con <- dbConnect(duckdb(), obis_grid_db)
dbExecute(con, glue("attach '{sdm_db}' as sdm (read_only)"))
occ_sql <- if (file_exists(obis_occ) && !is_dir(obis_occ)) {
  dbExecute(con, glue("attach '{obis_occ}' as ob (read_only)")); "ob.occ"
} else {
  glue("read_parquet('{obis_occ}/*.parquet')")
}
if (!"occ_cell" %in% dbListTables(con)) {
  dbExecute(con, glue("
    create table occ_cell as
    select speciesid::int aphia, any_value(species) species, any_value(phylum) phylum, any_value(\"class\") \"class\",
      (floor((90 - decimalLatitude)/0.05)::bigint * 7200 + floor((decimalLongitude+180)/0.05)::bigint + 1) cell_id,
      count(*) records, min(date_year) min_year, max(date_year) max_year
    from {occ_sql}
    where speciesid is not null and decimalLongitude is not null and decimalLatitude is not null
    group by 1, 5"))
}
# the pixel-index formula must reproduce cell_id exactly
chk <- dbGetQuery(con, "select count(*) n, sum((cell_id = floor((90 - lat)/0.05)::bigint * 7200 + floor((lon+180)/0.05)::bigint + 1)::int) n_ok from sdm.cell")
stopifnot(chk$n == chk$n_ok)

dbExecute(con, "create or replace table occ_usa as select o.* from occ_cell o join sdm.cell c using (cell_id) where c.in_usa")
dbExecute(con, glue("
  create or replace table spp as
  with g as (select aphia, any_value(species) species, any_value(phylum) phylum, any_value(\"class\") \"class\",
               sum(records) rec_global, count(*) cells_global from occ_cell group by 1),
       u as (select aphia, sum(records) rec_usa, count(*) cells_usa, min(min_year) min_year_usa, max(max_year) max_year_usa
             from occ_usa group by 1),
       v as (select try_cast(replace(t.taxon_id,'WORMS:','') as int) aphia, t.is_valid_usa, t.sp_cat,
               bool_or(m.ds_key='am') has_am, bool_or(m.ds_key='ax') has_ax, bool_or(m.ds_key like 'rng%') has_rng
             from sdm.taxon t left join sdm.taxon_model tm using (taxon_id) left join sdm.model m using (mdl_key)
             where t.taxon_authority='worms' group by all),
       b as (select distinct scientific_name from sdm.taxon where taxon_authority='botw'),
       mp as (select taxonID aphia, SDM mp_sdm, SDMinfo mp_info, availableRecords mp_rec from read_csv('{mp_list_csv}'))
  select g.*, coalesce(u.rec_usa,0) rec_usa, coalesce(u.cells_usa,0) cells_usa, u.min_year_usa, u.max_year_usa,
    v.aphia is not null in_v9, coalesce(v.is_valid_usa,false) is_valid_usa, v.sp_cat,
    coalesce(v.has_am,false) has_am, coalesce(v.has_ax,false) has_ax, coalesce(v.has_rng,false) has_rng,
    (g.\"class\" = 'Aves' or b.scientific_name is not null) is_bird,
    mp.mp_sdm, mp.mp_info, mp.mp_rec
  from g left join u using (aphia) left join v using (aphia)
    left join b on b.scientific_name = g.species left join mp using (aphia)"))
q <- \(sql) dbGetQuery(con, sql)
q("select count(*) n_spp, sum(rec_global) records, count(*) filter (where rec_usa>0) n_spp_usa, sum(rec_usa) records_usa from spp") |>
  kable(caption = "OBIS snapshot: species-rank records, globally and inside in_usa cells")
```

Birds are keyed by BirdLife (`botw`) ids in the taxon table, not AphiaID, so they are matched by name
and excluded from the gap counts below (BirdLife covers them).

```{r}
#| label: coverage
bins <- "case when cells_usa>=1000 then 'e >=1000' when cells_usa>=100 then 'd 100-999' when cells_usa>=30 then 'c 30-99'
         when cells_usa>=10 then 'b 10-29' else 'a 1-9' end"
q(glue("select {bins} us_cells, count(*) n_spp, sum(has_ax::int) n_ax, sum((has_am and not has_ax)::int) n_am_only,
   sum((not has_am and not has_ax)::int) n_gap, sum((not in_v9)::int) n_not_in_v9,
   sum((mp_sdm='available')::int) n_mpaeu_sdm, sum((not has_am and not has_ax and mp_sdm='available')::int) n_gap_mpaeu_covers
   from spp where rec_usa>0 and not is_bird group by 1 order by 1")) |>
  kable(caption = "Non-bird species with OBIS records in US cells, by distinct 0.05° cells occupied, vs v9 suitability coverage")

q("select case when has_ax then 'ax' when has_am then 'am_only' else 'other' end src, count(*) n_in_obis,
   sum((rec_usa>0)::int) n_any_us, sum((cells_usa>=10)::int) n_ge10, sum((cells_usa>=30)::int) n_ge30, sum((cells_usa>=100)::int) n_ge100
   from spp where in_v9 and not is_bird group by 1 order by 1") |>
  kable(caption = "v9 taxa by suitability source: OBIS evidence in US cells (how many could be re-modelled openly)")

q("select mp_sdm, count(*) n, sum((rec_usa>0)::int) n_any_us, sum((cells_usa>=30)::int) n_ge30_us,
   sum(in_v9::int) n_in_v9, sum((has_am or has_ax)::int) n_am_or_ax, sum(has_ax::int) n_ax
   from spp where mp_sdm is not null and not is_bird group by 1") |>
  kable(caption = "MPA Europe species (non-bird) that also occur in US waters, and whether v9 already models them")

q("select \"class\", count(*) n_ge30, sum(has_ax::int) n_ax, sum((has_am and not has_ax)::int) n_am_only,
   sum((not has_am and not has_ax)::int) n_gap, sum((mp_sdm='available')::int) n_mpaeu
   from spp where cells_usa>=30 and not is_bird and \"class\" in ('Mammalia','Elasmobranchii','Holocephali','Teleostei','Myxini','Petromyzonti')
   group by 1 order by 1") |>
  kable(caption = "Vertebrates with >=30 US cells: coverage by source")

gap <- q("select aphia, species, phylum, \"class\", rec_usa, cells_usa, min_year_usa, max_year_usa, in_v9, has_rng, mp_sdm
          from spp where cells_usa>=30 and not has_am and not has_ax and not is_bird order by cells_usa desc")
write_csv(gap, gap_csv)
gap |> count(phylum, class, sort = T) |> head(20) |>
  kable(caption = glue("Gap species (>=30 US cells, no am/ax, non-bird; n = {nrow(gap)}) by class"))
gap |> head(20) |> kable(caption = "Gap species with the most US cells")
```

## Walrus test: OBIS ensemble vs IUCN range vs AquaX

The OBIS models are predicted to the whole globe and only carry a marine-realm mask
(`native_ecoregions` = Costello realms with records), which for the walrus admits most of the US
Pacific. Our merge rule masks suitability to the range dataset, so the useful comparison is the OBIS
surface inside the IUCN range against AquaX.

```{r}
#| label: walrus
aphia_w <- 137077   # Odobenus rosmarus
u <- glue("https://obis-maps.s3.amazonaws.com/sdm/species/taxonid={aphia_w}/model=mpaeu")
d_w <- glue("{dir_obismap}/taxonid={aphia_w}"); dir_create(d_w)
get <- \(rel, f) { p <- glue("{d_w}/{f}"); if (!file_exists(p)) download.file(glue("{u}/{rel}"), p, quiet = T, mode = "wb"); p }
f_ens  <- get(glue("predictions/taxonid={aphia_w}_model=mpaeu_method=ensemble_scen=current_cog.tif"), "ensemble_current.tif")
f_mask <- get(glue("predictions/taxonid={aphia_w}_model=mpaeu_what=mask_cog.tif"), "mask.tif")
f_thr  <- get(glue("metrics/taxonid={aphia_w}_model=mpaeu_what=thresholds.parquet"), "thresholds.parquet")
f_log  <- get(glue("taxonid={aphia_w}_model=mpaeu_what=log.json"), "log.json")
lg  <- fromJSON(f_log)
thr <- read_parquet(f_thr); p10 <- thr$p10[thr$model == "ensemble"] * 100
r   <- rast(f_ens); m <- rast(f_mask)
stopifnot(ncol(r) == 7200, nrow(r) == 3600, max(abs(as.vector(ext(r)) - c(-180, 180, -90, 90))) < 1e-4)
v <- values(r[[1]])[, 1]; mv <- values(m)

D <- glue("{dir_big_v}/marine-atlas/dist")
iucn <- q(glue("select cell_id from read_parquet('{D}/dataset=rng_iucn/*.parquet', hive_partitioning=false) where mdl_key = 'rng_iucn|15106'"))$cell_id
axd  <- q(glue("select cell_id, val from read_parquet('{D}/dataset=ax/*.parquet', hive_partitioning=false) where mdl_key = 'ax|{aphia_w}'"))
amd  <- q(glue("select cell_id, val from read_parquet('{D}/dataset=am/*.parquet', hive_partitioning=false) where mdl_key = 'am|ITS-Mam-180639'"))
usa  <- q("select cell_id from sdm.cell where in_usa")$cell_id
in_iucn <- usa %in% iucn; in_ax <- usa %in% axd$cell_id; in_am <- usa %in% amd$cell_id
ens <- v[usa]
r_shared <- \(d, sel) cor(ens[sel & !is.na(ens)], setNames(d$val, d$cell_id)[as.character(usa[sel & !is.na(ens)])])
jacc <- \(a, b) sum(a & b) / sum(a | b)
obis_p10 <- !is.na(ens) & ens >= p10
tibble(
  fit_points = unlist(lg$model_fit_points), good_models = paste(unlist(lg$model_good), collapse = "+"), p10 = p10,
  us_cells_iucn = sum(in_iucn), us_cells_am = sum(in_am), us_cells_ax = sum(in_ax),
  am_cells_outside_iucn = sum(in_am & !in_iucn), ax_cells_outside_iucn = sum(in_ax & !in_iucn),
  us_cells_obis_gt0 = sum(ens > 0, na.rm = T), us_cells_obis_ge_p10 = sum(obis_p10),
  obis_ge_p10_in_iucn = sum(obis_p10 & in_iucn),
  obis_ge_p10_in_native_realm = sum(obis_p10 & mv[usa, "native_ecoregions"] == 1, na.rm = T),
  jaccard_obis_p10_vs_am = jacc(obis_p10, in_am), jaccard_obis_p10_vs_ax = jacc(obis_p10, in_ax),
  r_obis_vs_am_shared = r_shared(amd, in_am), r_obis_vs_ax_shared = r_shared(axd, in_ax),
  r_am_vs_ax_shared = cor(setNames(amd$val, amd$cell_id)[as.character(usa[in_am & in_ax])], setNames(axd$val, axd$cell_id)[as.character(usa[in_am & in_ax])])) |>
  t() |> kable(caption = "Walrus: OBIS ensemble (current) in US cells vs IUCN range, AquaMaps 1.0 (am) and AquaMaps 2.0 (ax)")
dbDisconnect(con, shutdown = T)
```

Reading: unmasked, the OBIS walrus surface is nonzero over most US cells; at its own P10 threshold it
shrinks to the Bering/Chukchi and almost entirely inside the IUCN range. Against AquaMaps 1.0 the
question is improvement (the trapezoid envelope over-predicts outside the range); against AquaMaps 2.0
it is agreement on the shared cells. The realm mask is not a substitute for the range mask.
