---
title: "Ingest US federal listings → MMPA / MBTA / ESA taxon attributes"
msens:
  target_name: ingest_listings
  workflow_type: ingest
  dependency: []
  output: data/manifests/ingest_listings.json
editor_options:
  chunk_output_type: console
---

Read the US federal species lists and derive per-species attributes that drive the
taxon-level extinction-risk score (`compute_er_score` floors): **ESA** status
(NMFS + FWS), **`is_mmpa`** (Marine Mammal Protection Act, from the NMFS directory),
**`is_mbta`** (Migratory Bird Treaty Act, from the FWS 10.13 list — essential now
that v8 ingests **all** birds, not v7's 571), and **`is_bcc`** (Birds of
Conservation Concern). Keyed by `clean_sci_name()` so the merge crosswalk joins it
to each taxon; the US-vs-IUCN precedence + `compute_er_score` happen in `merge_taxon`.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "US federal lists → per-taxon MMPA/MBTA/ESA/BCC attributes joined on clean scientific name"
flowchart LR
  nmfs["NMFS directory<br/>(ESA + is_mmpa)"] --> comb["full_join by<br/>clean_sci_name()"]
  fws["FWS Explorer<br/>(ESA + is_bcc)"] --> comb
  mbta["CFR 50 §10.13<br/>(is_mbta)"] --> comb
  comb --> tbl[("listing table<br/>merge.duckdb")]
  tbl --> mf["hash_query(listing)<br/>→ content-addressed manifest"]
```

## Setup + read the three lists

```{r}
#| label: listings

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, janitor, jsonlite, logger,
                 readr, readxl, stringr, tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))
options(readr.show_col_types = F)

nmfs_csv  <- here("data/nmfs_species-directory.csv")
fws_csv   <- glue("{dir_raw}/fws.gov/species/FWS_Species_Data_Explorer.csv")
mbta_xlsx <- glue("{dir_raw}/fws.gov/birds_mbta/CFR50-Part10.13-2023.xlsx")
merge_db  <- glue("{dir_big_v}/marine-atlas/merge.duckdb")
manifest  <- here("data/manifests/ingest_listings.json")
dir_create(path_dir(manifest))
stopifnot(all(file_exists(c(nmfs_csv, fws_csv, mbta_xlsx))))

# NMFS species directory -> ESA status (NMFS) + is_mmpa (parse_noaa_status)
d_nmfs_raw <- read_csv(nmfs_csv)
d_nmfs <- d_nmfs_raw |>
  bind_cols(parse_noaa_status(d_nmfs_raw$protected_status)) |>
  transmute(sci = clean_sci_name(scientific_name), nmfs_esa = esa_status, is_mmpa) |>
  group_by(sci) |>
  summarise(nmfs_esa = if (any(nmfs_esa == "EN")) "EN" else if (any(nmfs_esa == "TN")) "TN" else "LC",
            is_mmpa = any(is_mmpa), .groups = "drop")

# FWS Data Explorer -> ESA status (FWS) + is_bcc
d_fws <- read_csv(fws_csv) |> clean_names() |>
  transmute(
    sci     = clean_sci_name(scientific_name),
    fws_esa = case_when(str_detect(esa_listing_status, "Endangered") ~ "EN",
                        str_detect(esa_listing_status, "Threatened") ~ "TN", TRUE ~ "LC"),
    is_bcc  = is_bcc %in% c(TRUE, "TRUE")) |>
  group_by(sci) |>
  summarise(fws_esa = if (any(fws_esa == "EN")) "EN" else if (any(fws_esa == "TN")) "TN" else "LC",
            is_bcc = any(is_bcc), .groups = "drop")

# MBTA 10.13 list -> is_mbta (Genus + species binomial in cols 4-5)
mbta <- read_excel(mbta_xlsx, sheet = "10.13 List", skip = 3, col_names = FALSE)
mbta_sci <- clean_sci_name(paste(mbta[[4]], mbta[[5]])) |> unique()
```

## Combine → listing table (keyed by clean scientific name)

```{r}
#| label: combine

listing <- full_join(d_nmfs, d_fws, by = "sci") |>
  mutate(
    is_mmpa = coalesce(is_mmpa, FALSE),
    is_bcc  = coalesce(is_bcc,  FALSE),
    is_mbta = sci %in% mbta_sci,
    nmfs_esa = coalesce(nmfs_esa, "LC"),
    fws_esa  = coalesce(fws_esa,  "LC")) |>
  # add MBTA-only birds not otherwise in the NMFS/FWS tables
  bind_rows(tibble(sci = setdiff(mbta_sci, full_join(d_nmfs, d_fws, by = "sci")$sci),
                   nmfs_esa = "LC", fws_esa = "LC",
                   is_mmpa = FALSE, is_bcc = FALSE, is_mbta = TRUE)) |>
  distinct(sci, .keep_all = TRUE)

con <- dbConnect(duckdb(merge_db))
dbWriteTable(con, "listing", listing, overwrite = TRUE)
smry <- dbGetQuery(con, "SELECT count(*) n, sum(is_mmpa::int) mmpa, sum(is_mbta::int) mbta,
  sum(is_bcc::int) bcc, count(*) FILTER(WHERE nmfs_esa<>'LC' OR fws_esa<>'LC') esa_listed FROM listing")
# content fingerprint of the listing table (before disconnect)
h <- msens::hash_query(con, "listing")
dbDisconnect(con, shutdown = TRUE)
msens::report_table(smry, caption = "US federal listing table summary")
```

## Manifest

```{r}
#| label: manifest

# content-addressed manifest: deterministic (no wall-clock) -> downstream targets
# re-run only when the listing table's content actually changes
msens::write_manifest(
  manifest, target = "ingest_listings", content_hash = h,
  stats = list(ver = ver, n = smry$n, n_mmpa = smry$mmpa,
               n_mbta = smry$mbta, n_bcc = smry$bcc),
  force = msens::force_target("ingest_listings"))
```
