---
title: "Ingest WoRMS — (re)build the spp.duckdb `worms` table from the monthly marinespecies.org DwC-A"
msens:
  target_name: ingest_worms
  workflow_type: ingest
  dependency: []
  output: data/manifests/ingest_worms.json
editor_options:
  chunk_output_type: console
---

Rebuild the `worms` taxonomy table in **`spp.duckdb`** from the monthly **World Register
of Marine Species** Darwin Core Archive downloaded from
[marinespecies.org/download](https://www.marinespecies.org/download/). Run this whenever a
newer download is dropped in `{dir_raw}/marinespecies.org/WoRMS_download_<YYYY-MM-DD>/`; it
is **version-tracked** (a `worms_meta` row records the loaded download date) and idempotent
— it rebuilds only when the latest download on disk is newer than what is loaded (or
`REDO_WORMS=1`).

**Why this source (not checklistbank / GBIF):** the marinespecies.org DwC-A is the
authoritative WoRMS export and carries the `isMarine`/`isBrackish`/… habitat flags. It is
updated monthly. **AlgaeBase-sourced taxa (macroalgae/seagrass) are license-excluded** from
this (and every redistributable) WoRMS download, so ~hundreds of algae used by our SDM
datasets are absent here — `merge_taxon.qmd` fills those on demand from the **WoRMS REST
API** (`worrms::wm_record`) and caches them in `data/worms_taxonomy_supplement.csv`.

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "Latest WoRMS DwC-A → rebuilt `worms` taxonomy table with habitat flags, content-addressed manifest"
flowchart LR
  dwc["WoRMS DwC-A<br/>taxon.txt + speciesprofile.txt"] --> pick["pick latest<br/>WoRMS_download_&lt;date&gt;"]
  pick --> bld["CREATE TABLE worms<br/>(strip LSID, join isMarine/…)"]
  bld --> tbl[("worms table<br/>spp.duckdb")]
  tbl --> mf["hash_query(worms)<br/>→ content-addressed manifest"]
```

## Setup + locate the latest download

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, readr, stringr, quiet = T)
source(here("libs/paths.R"))   # spp_db, dir_raw
manifest <- here("data/manifests/ingest_worms.json"); dir_create(path_dir(manifest))
redo     <- nzchar(Sys.getenv("REDO_WORMS"))

# newest WoRMS_download_<date> folder holding the DwC-A core + speciesprofile
dl_root <- glue("{dir_raw}/marinespecies.org")
dls <- dir_ls(dl_root, type = "directory", regexp = "WoRMS_download_\\d{4}-\\d{2}-\\d{2}$")
stopifnot("no WoRMS_download_<date> folder found" = length(dls) > 0)
dl_dir   <- dls[order(path_file(dls))] |> tail(1)                    # latest by name (date)
worms_ver <- str_extract(path_file(dl_dir), "\\d{4}-\\d{2}-\\d{2}")
taxon_txt <- glue("{dl_dir}/taxon.txt")
sprof_txt <- glue("{dl_dir}/speciesprofile.txt")
stopifnot(all(file_exists(c(taxon_txt, sprof_txt))))
log_info("latest WoRMS download: {path_file(dl_dir)} (version {worms_ver})")
```

## Fetch a fresh monthly download (procedure)

The [WoRMS download](https://www.marinespecies.org/download/) is **login-gated and
request-based** (there is no stable "latest" URL): sign in, request a full Darwin Core
Archive, and download the generated zip. It cannot be reliably scripted end-to-end, so the
monthly refresh is: **download → unzip into `WoRMS_download_<YYYY-MM-DD>/` → run this
notebook** (it auto-detects the newest folder).

Credentials live outside git in `{dir_private}/…marinespecies.org_<user>.txt` (password
only; username is `ben@ecoquants.com`). The chunk below is a **best-effort** helper (kept
`eval: false`) — if WoRMS changes to expose a direct authenticated archive URL, wire it here.

```{r}
#| label: download
#| eval: false

# username + password (password file is the plain secret; never commit it)
dir_private <- ifelse(Sys.info()[["sysname"]] == "Linux", "/share/private", "~/My Drive/private")
pw_txt <- Sys.glob(path.expand(glue("{dir_private}/*marinespecies.org_*.txt")))[1]
worms_user <- "ben@ecoquants.com"
worms_pass <- readr::read_file(pw_txt) |> trimws()
# NOTE: marinespecies.org serves the DwC-A via a login + request flow, not a static URL.
# Do the request in the browser, then unzip into
#   {dir_raw}/marinespecies.org/WoRMS_download_<YYYY-MM-DD>/
# and re-run this notebook. (Placeholder for a future authenticated-URL fetch.)
```

## (Re)build the `worms` table

```{r}
#| label: build

con <- dbConnect(duckdb(spp_db))
loaded <- tryCatch(
  dbGetQuery(con, "SELECT worms_ver FROM worms_meta ORDER BY built DESC LIMIT 1")$worms_ver,
  error = function(e) NA_character_)

if (!redo && isTRUE(loaded == worms_ver) && "worms" %in% dbListTables(con)) {
  log_info("worms table already at version {worms_ver}; skipping (REDO_WORMS=1 to force)")
} else {
  strip <- function(col) glue(
    "TRY_CAST(regexp_replace({col}, 'urn:lsid:marinespecies.org:taxname:', '') AS BIGINT)")
  dbExecute(con, glue("
    CREATE OR REPLACE TABLE worms AS
    WITH sp AS (
      SELECT {strip('taxonID')} AS taxonID, isMarine, isFreshwater, isTerrestrial,
             isExtinct, isBrackish
      FROM read_csv('{sprof_txt}', delim='\t', header=true, quote='\"', ignore_errors=true,
                    types={{'taxonID':'VARCHAR'}}))
    SELECT
      {strip('t.taxonID')}             AS taxonID,
      {strip('t.scientificNameID')}    AS scientificNameID,
      {strip('t.acceptedNameUsageID')} AS acceptedNameUsageID,
      {strip('t.parentNameUsageID')}   AS parentNameUsageID,
      t.scientificName, t.acceptedNameUsage, t.parentNameUsage,
      t.kingdom, t.phylum, t.\"class\", t.\"order\", t.family, t.genus, t.subgenus,
      t.specificEpithet, t.infraspecificEpithet, t.taxonRank, t.scientificNameAuthorship,
      t.nomenclaturalCode, t.taxonomicStatus, t.nomenclaturalStatus,
      sp.isMarine, sp.isFreshwater, sp.isTerrestrial, sp.isExtinct, sp.isBrackish
    FROM read_csv('{taxon_txt}', delim='\t', header=true, quote='\"', ignore_errors=true,
                  types={{'taxonID':'VARCHAR','scientificNameID':'VARCHAR',
                          'acceptedNameUsageID':'VARCHAR','parentNameUsageID':'VARCHAR'}}) t
    LEFT JOIN sp ON {strip('t.taxonID')} = sp.taxonID"))

  for (ix in c("taxonID", "scientificNameID", "acceptedNameUsageID", "scientificName"))
    try(dbExecute(con, glue("CREATE INDEX worms_{ix}_idx ON worms ({ix})")), silent = TRUE)

  n <- dbGetQuery(con, "SELECT count(*) n FROM worms")$n
  dbExecute(con, "CREATE TABLE IF NOT EXISTS worms_meta (worms_ver VARCHAR, n_rows BIGINT, source VARCHAR, built TIMESTAMP)")
  dbExecute(con, glue("INSERT INTO worms_meta VALUES ('{worms_ver}', {n}, '{path_file(dl_dir)}', now())"))
  log_info("rebuilt worms: {n} rows from {path_file(dl_dir)}")
}
```

## Verify

```{r}
#| label: verify

smry <- dbGetQuery(con, "SELECT count(*) n_rows,
  count(*) FILTER (WHERE scientificNameID IS NOT NULL) n_aphia,
  count(*) FILTER (WHERE isMarine) n_marine,
  count(DISTINCT kingdom) n_kingdoms FROM worms")
msens::report_table(smry, caption = "worms taxonomy table summary")
knitr::kable(dbGetQuery(con, "SELECT kingdom, count(*) n FROM worms GROUP BY 1 ORDER BY n DESC LIMIT 8"))
# AlgaeBase check: these should be ABSENT here (filled by the WoRMS API in merge_taxon)
alg <- dbGetQuery(con, "SELECT count(*) n FROM worms WHERE scientificName IN ('Caulerpa sertularioides','Ulva intestinalis')")$n
log_info("AlgaeBase spot-check (expect 0, filled via API): {alg}")
# content fingerprint of the worms table (before disconnect)
h <- msens::hash_query(con, "worms")
dbDisconnect(con, shutdown = TRUE)
```

## Manifest

```{r}
#| label: manifest

# content-addressed manifest: deterministic (no wall-clock) -> downstream targets
# re-run only when the worms table's content actually changes
msens::write_manifest(
  manifest, target = "ingest_worms", content_hash = h,
  stats = list(worms_ver = worms_ver, source = path_file(dl_dir),
               n_rows = smry$n_rows, n_marine = smry$n_marine),
  force = msens::force_target("ingest_worms"))
manifest
```
