---
title: "Ingest IUCN Red List expert ranges → global 0.05° cells"
msens:
  target_name: ingest_rng_iucn
  workflow_type: ingest
  dependency: [build_cell_grid]
  output: data/manifests/ingest_rng_iucn.json
  dataset: {ds_key: rng_iucn, response_type: range, source_authority: IUCN, temporal_interval: static, native_format: vector}
editor_options:
  chunk_output_type: console
---

IUCN Red List **expert ranges** for marine taxa onto the **global 0.05° cell grid**
— the non-bird complement to AquaMaps/BOTW. Source: the 24 marine taxonomic-group
downloads under `raw/iucnredlist.org/` (corals, fish, sharks, marine mammals,
turtles, seagrasses, mangroves…).

Per species (`id_no`): union its **extant** polygons (`presence ∈ {1,2,3}`),
capture the **whole range** (no land mask), value = `msens::compute_er_score(
"IUCN:{category}")` (never hard-coded; CR=50/EN=25/VU=5/NT=2/LC,DD=1). `mdl_key =
"rng_iucn|{id_no}"`. Species are filtered to IUCN's own **`marine = true`** flag —
authoritative for non-birds (e.g. it drops freshwater turtles from the TURTLES
group) — so the plain-`MAMMALS` terrestrial species are excluded too.

Parallel (mclapply, per group in memory), resumable; `REDO_INGEST=1` rebuilds all.

## Setup

```{r}
#| label: setup

librarian::shelf(
  arrow, DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, readr, sf, terra,
  tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))   # dir_raw, dir_big_v, cellid_tif, ver
source(here("libs/vars.R"))    # redo_ingest
options(readr.show_col_types = F)
sf_use_s2(FALSE)

ds_key    <- "rng_iucn"
dir_iucn  <- glue("{dir_raw}/iucnredlist.org")
dir_dist  <- glue("{dir_big_v}/marine-atlas/dist/dataset={ds_key}")
manifest  <- here("data/manifests/ingest_rng_iucn.json")
n_workers <- max(1L, parallel::detectCores())
dir_create(c(dir_dist, path_dir(manifest)))
stopifnot("run build_cell_grid.qmd first" = file_exists(cellid_tif), dir_exists(dir_iucn))

# valid IUCN Red List categories for compute_er_score (drop EX/EW/blank)
iucn_ok <- c("CR", "EN", "VU", "NT", "LC", "DD")
if (redo_ingest && dir_exists(dir_dist)) { dir_delete(dir_dist); dir_create(dir_dist) }
```

## Iterate taxonomic groups → species → cells

```{r}
#| label: interpolate

groups <- dir_ls(dir_iucn, type = "directory")
sp_all <- list()

for (grp in groups) {
  shps <- dir_ls(grp, glob = "*.shp", recurse = TRUE)
  if (length(shps) == 0) next
  g <- suppressWarnings(bind_rows(lapply(shps, st_read, quiet = TRUE))) |>
    rlang::set_names(tolower) |>
    filter(presence %in% 1:3, tolower(marine) == "true", category %in% iucn_ok)
  if (nrow(g) == 0) next

  sp <- g |> st_drop_geometry() |>
    group_by(id_no, sci_name, class, category) |> summarise(.groups = "drop") |>
    mutate(er_score = compute_er_score(paste0("IUCN:", category)),
           mdl_key  = mdl_key_raw(ds_key, id_no), group = path_file(grp))
  sp_all[[path_file(grp)]] <- sp
  done <- as.integer(path_ext_remove(path_file(dir_ls(dir_dist, glob = "*.parquet"))))
  todo <- sp |> filter(!id_no %in% done)
  log_info("{path_file(grp)}: {nrow(sp)} marine species, {nrow(todo)} to do")

  one <- function(i) {
    r      <- todo[i, ]
    ply_sp <- g |> filter(id_no == r$id_no) |> summarise(.groups = "drop")
    d <- tryCatch(cells_from_ranges(ply_sp, cellid_tif, value = r$er_score),
                  error = function(e) NULL)
    if (is.null(d) || nrow(d) == 0) return("empty")
    arrow::write_parquet(
      tibble::tibble(mdl_key = r$mdl_key, cell_id = d$cell_id, value = d$value),
      fs::path(dir_dist, r$id_no, ext = "parquet")); "done"
  }
  invisible(parallel::mclapply(seq_len(nrow(todo)),
    function(i) tryCatch(one(i), error = function(e) "error"),
    mc.cores = n_workers, mc.preschedule = FALSE))
}
sp_all <- bind_rows(sp_all)
write_csv(sp_all, glue("{dir_dist}/../model_{ds_key}.csv"))
glue("{nrow(sp_all)} IUCN marine species across {length(unique(sp_all$group))} groups")
```

## Verify + manifest

```{r}
#| label: verify

pq  <- dir_ls(dir_dist, glob = "*.parquet")
con <- dbConnect(duckdb())
smry <- dbGetQuery(con, glue(
  "SELECT count(DISTINCT mdl_key) n_models, count(*) n_cells, min(value) v_min, max(value) v_max
     FROM read_parquet('{dir_dist}/*.parquet')"))
dbDisconnect(con, shutdown = TRUE)
knitr::kable(smry)

write_json(
  list(target = "ingest_rng_iucn", ds_key = ds_key, ver = ver,
       n_species = nrow(sp_all), n_parquet = length(pq),
       dir_dist = dir_dist, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
```
