---
title: "Ingest NMFS/FWS/MMPA/MBTA Listings"
subtitle: "Compute extrisk_code and er_score from US federal species listings"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Overview

Ingest US federal species listings (ESA from NMFS/FWS, MMPA from NMFS, MBTA from FWS), compute
`extrisk_code` and `er_score`, and write a staging `listing` table to
`sdm_2026.duckdb`. This notebook runs **BEFORE** `merge_models.qmd`.

**Scoring rules:**

- US-listed species get max of ESA score, MMPA (20), MBTA (10)
- Any species in NMFS or FWS directory but not EN/TN gets "LC" (not NA)
- Non-US species fall back to IUCN scale in `merge_models.qmd`

```{r}
#| label: setup

devtools::load_all("../msens")
librarian::shelf(
  DBI,
  dplyr,
  DT,
  duckdb,
  glue,
  here,
  htmltools,
  janitor,
  litedown,
  MarineSensitivity / msens,
  readr,
  readxl,
  stringr,
  tibble,
  tidyr,
  quiet = T
)

options(readr.show_col_types = F)

# helper functions ----

# helper: convert *_url columns to HTML links in datatable
datatable_with_links <- function(data, url_rx = "_url$", ...) {
  # find all *_url columns
  url_cols <- names(data)[str_detect(names(data), url_rx)]

  if (length(url_cols) > 0) {
    # get corresponding display columns (remove _url suffix)
    display_cols <- str_remove(url_cols, url_rx)

    # check that display columns exist in data
    display_cols <- display_cols[display_cols %in% names(data)]
    stopifnot(length(display_cols) == length(url_cols))

    # create links for each pair
    for (i in seq_along(url_cols)) {
      url_col <- url_cols[i]
      display_col <- display_cols[i]

      data <- data |>
        mutate(
          !!display_col := ifelse(
            !is.na(.data[[url_col]]) & .data[[url_col]] != "",
            glue(
              '<a href="{.data[[url_col]]}" target="_blank">{.data[[display_col]]}</a>'
            ),
            as.character(.data[[display_col]])
          )
        ) |>
        select(-all_of(url_col))
    }
  }

  # determine which columns have HTML links
  has_links <- any(str_detect(names(data), "_url$")) || length(url_cols) > 0

  datatable(data, escape = !has_links, ...)
}

common_words <- function(names) {
  # find longest common subsequence of words across common names

  words_list <- str_split(names, "\\s+")
  common <- Reduce(intersect, words_list)
  if (length(common) == 0) {
    return(paste(names, collapse = " / "))
  }
  paste(common, collapse = " ")
}

# paths ----

dir_data <- msens::sdm_db_path() |> dirname() |> dirname()

con_spp <- msens::spp_db_con()
con_sdm <- msens::sdm_db_con(read_only = FALSE)

nmfs_src_csv <- here("data/nmfs_species-directory.csv")
nmfs_worms_csv <- here("data/nmfs_species-directory_worms-matched.csv")

fws_spp_csv <- glue(
  "{dir_data}/raw/fws.gov/species/FWS_Species_Data_Explorer.csv"
)
f_bw_csv <- here("data/fws_species_worms-botw.csv")

mbta_xlsx <- glue(
  "{dir_data}/raw/fws.gov/birds_mbta/CFR50-Part10.13-2023.xlsx"
)
mbta_botw_worms_csv <- here("data/mbta_botw_worms.csv")

overrides_csv <- here("data/species_overrides_v3.csv")
overrides_audit_csv <- here("data/species_overrides_v3_.csv")

# values ----

# esa severity ranking for taking the max
esa_levels <- c("LC", "TN", "EN")

# lookup tables for corrections ----
rx_todrop <- ",| and |Multiple species"

spp_sci_to_cmn <- tribble(
  ~scientific_name_dataset , ~common_name ,
  "Coryphaena hippurus"    , "Mahi Mahi"  ,
)

spp_sci_renames <- tribble(
  ~scientific_name_dataset             , ~scientific_name_corrected  ,
  "Sardinops sagax caerulea"           , "Sardinops sagax caeruleus" ,
  "Lampris guttatus, Lampris spp."     , "Lampris"                   ,
  "Crassostrea gigas; Magallana gigas" , "Magallana gigas"           ,
  "Mercluccius productus"              , "Merluccius productus"      ,
  "Delphinus capensis"                 , "Delphinus delphis"         ,
  "Acanthocybium solanderi"            , "Acanthocybium solandri"
)
# "Delphinus capensis" has 2 accepted WoRMS matches: _Delphinus delphis_ and _Cephalorhynchus heavisidii_

d_worms_missing <- tribble(
  ~sci_ds                , ~scientific_name_accepted , ~worms_id , ~taxon_rank , ~is_marine ,
  "Saccharina latissima" , "Saccharina latissima"    ,    234483 , "Species"   , TRUE
) # Wierdly, algaes (from AlgaeBase) are searchable on WoRMS but not downloadable
```

## Ingest NMFS Species Directory

Source: `r nmfs_src_csv` (309 species)

Parse `protected_status` using `msens::parse_noaa_status()`. Any species in
the NMFS directory gets at minimum `"LC"` (not NA).

- Chum Salmon	_Oncorhynchus keta_: LC, TN
- Coho Salmon _Oncorhynchus kisutch_: LC, EN

```{r}
#| label: ingest_nmfs

d_nmfs_raw <- read_csv(nmfs_src_csv)

# parse protected status into esa_status and is_mmpa
d_nmfs_parsed <- msens::parse_noaa_status(d_nmfs_raw$protected_status)

d_nmfs <- d_nmfs_raw |>
  bind_cols(d_nmfs_parsed) |>
  rename(
    scientific_name_dataset = scientific_name,
    scientific_name_dataset_url = species_url
  ) |>
  mutate(
    scientific_name_corrected = recode_values(
      scientific_name_dataset,
      from = spp_sci_renames$scientific_name_dataset,
      to = spp_sci_renames$scientific_name_corrected
    ),
    common_name = recode_values(
      scientific_name_dataset,
      from = spp_sci_to_cmn$scientific_name_dataset,
      to = spp_sci_to_cmn$common_name
    ),
    # sci_ds: scientific_name_dataset (corrected)
    sci_ds = ifelse(
      !is.na(scientific_name_corrected),
      scientific_name_corrected,
      scientific_name_dataset
    ),
  ) |>
  relocate(sci_ds)

d_nmfs |>
  filter(!is.na(scientific_name_corrected)) |>
  select(
    scientific_name_dataset,
    scientific_name_dataset_url,
    scientific_name_corrected,
    common_name
  ) |>
  arrange(scientific_name_dataset) |>
  datatable_with_links(caption = "NMFS species with corrected scientific names")

d_nmfs |>
  filter(duplicated(sci_ds) | duplicated(sci_ds, fromLast = TRUE)) |>
  arrange(esa_status, sci_ds) |>
  select(-scientific_name_corrected) |>
  datatable_with_links(
    caption = HTML(mark(
      "NMFS duplicated scientific names (dataset corrected) `sci_ds`"
    ))
  )


# handle duplicates by taking max ESA status
d_nmfs <- d_nmfs |>
  mutate(
    nmfs_esa = factor(esa_status, levels = esa_levels, ordered = TRUE)
  ) |>
  summarize(
    scientific_name_dataset = first(scientific_name_dataset),
    scientific_name_dataset_url = first(scientific_name_dataset_url),
    category = first(category),
    region = paste(unique(region), collapse = "; "),
    common_name = common_words(common_name),
    nmfs_esa = max(nmfs_esa),
    nmfs_esa_all = paste(unique(nmfs_esa), collapse = "; "),
    is_mmpa = any(is_mmpa),
    .by = sci_ds
  )

d_nmfs_todrop <- d_nmfs |>
  filter(
    is.na(scientific_name_dataset) | str_detect(sci_ds, rx_todrop)
  )

d_nmfs_todrop |>
  select(scientific_name_dataset, common_name, nmfs_esa) |>
  datatable_with_links(
    caption = HTML(mark(glue(
      "NMFS species dropped due to missing or multi-species scientific names (regex: `{rx_todrop}`)."
    )))
  )

d_nmfs <- d_nmfs |> filter(!sci_ds %in% d_nmfs_todrop$sci_ds)

# d_wn_direct: worms to nmfs direct matches to WoRMS via scientific name (Rule A)
d_wn_direct <- tbl(con_spp, "worms") |>
  filter(
    scientificName %in% !!unique(d_nmfs$sci_ds),
    taxonomicStatus == "accepted"
  ) |>
  select(
    sci_ds = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    worms_id = taxonID,
    taxon_rank = taxonRank,
    is_marine = isMarine
  ) |>
  collect()

sci_indirect = setdiff(d_nmfs$sci_ds, d_wn_direct$sci_ds) |>
  unique()

# for indirect, get acceptedNameUsageID  (Rule B)
d_wn_indirect <- tbl(con_spp, "worms") |>
  filter(
    scientificName %in% !!sci_indirect,
  ) |>
  select(
    sci_ds = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    worms_id = acceptedNameUsageID,
  ) |>
  left_join(
    tbl(con_spp, "worms") |>
      select(
        worms_id = taxonID,
        taxon_rank = taxonRank,
        is_marine = isMarine
      ),
    by = join_by(worms_id)
  ) |>
  collect()

d_wn <- bind_rows(d_wn_direct, d_wn_indirect, d_worms_missing) |>
  mutate(
    worms_id_url = glue(
      "https://www.marinespecies.org/aphia.php?p=taxdetails&id={worms_id}"
    )
  )

stopifnot(length(setdiff(d_nmfs$sci_ds, d_wn$sci_ds)) == 0)

d_wn |>
  datatable_with_links(
    caption = HTML(mark(
      "WoRMS matches to NMFS species directory via scientific name (`sci_ds`), including indirect matches via acceptedNameUsage."
    ))
  )

d_wn |>
  filter(sci_ds != scientific_name_accepted) |>
  datatable_with_links(
    caption = HTML(mark(
      "NMFS scientific names not matching WoRMS accepted name. These may be synonyms that can still be matched via `worms_acc`."
    ))
  )

stopifnot(sum(duplicated(d_wn$sci_ds)) == 0)
stopifnot(sum(duplicated(d_wn$worms_id)) == 0)

d_wn |>
  filter(taxon_rank != "Species") |>
  datatable_with_links(
    caption = "WoRMS matches that are not a taxonomic rank of Species"
  )

# d_nw: nmfs joined to worms
d_nw <- d_nmfs |>
  left_join(d_wn, by = join_by(sci_ds))

# write to csv
d_nw |>
  write_csv(nmfs_worms_csv)

d_nw |>
  datatable_with_links(
    caption = "NMFS species directory filtered and matched to WoRMS."
  )

# summary
d_nw |>
  count(nmfs_esa, is_mmpa, name = "n_species") |>
  arrange(desc(n_species)) |>
  datatable(
    caption = HTML(mark(
      "Summary of NMFS species directory by ESA status and MMPA listing"
    ))
  )
```

## Ingest FWS Species Data Explorer

Source: [FWS Species Data Explorer](https://ecos.fws.gov/ecp/report/adhoc-creator?catalogId=species&reportId=species&columns=%2Fspecies@cn,sn,gn;%2Fspecies%2Ftaxonomy@tsn,spcode&sort=%2Fspecies@cn%20asc;%2Fspecies@sn%20asc&distinct=true)

Exclude taxonomic groups that do not contain marine species:
```{r}
#| label: fws_taxonomic_exclusions
#| echo: false
#| results: asis

# filter to marine-relevant taxonomic groups
fws_taxonomic_groups_exclude <- c(
  "Amphibians",
  "Arachnids",
  "Conifers and Cycads",
  "Ferns and Allies",
  "Flowering Plants",
  "Insects",
  "Lichens",
  "Millipedes"
)

cat(paste0("- ", fws_taxonomic_groups_exclude, collapse = "\n"))
```


ESA status:

- "Endangered" -> "EN",
- "Threatened" -> "TN", 
- otherwise -> "LC". 

Match via ITIS TSN crosswalk.

```{r}
#| label: ingest_fws

d_fws_raw <- read_csv(fws_spp_csv) |>
  clean_names() |>
  mutate(
    taxonomic_group_excluded = taxonomic_group %in% fws_taxonomic_groups_exclude
  )

d_fws_raw |>
  filter(is_bcc) |>
  datatable(
    caption = "Number of FWS species that are Birds of Conservation Concern (BCC)"
  )

d_fws_raw |>
  filter(taxonomic_group_excluded) |>
  group_by(taxonomic_group) |>
  summarize(n = n()) %>%
  bind_rows(
    summarize(., taxonomic_group = "TOTAL", n = sum(n))
  ) |>
  datatable(
    caption = "Number of FWS species in excluded non-marine taxonomic groups"
  ) |>
  formatCurrency("n", digits = 0, currency = "")

d_fws <- d_fws_raw |>
  filter(!taxonomic_group_excluded)

d_fws |>
  group_by(taxonomic_group) |>
  summarize(n = n()) %>%
  bind_rows(
    summarize(., taxonomic_group = "TOTAL", n = sum(n))
  ) |>
  datatable(
    caption = "Number of FWS species by taxonomic group after excluding non-marine groups"
  ) |>
  formatCurrency("n", digits = 0, currency = "")


d_fws |>
  datatable_with_links(
    caption = "FWS species after non-marine taxonomic group exclusion"
  )

d_fws |>
  count(status_category) |>
  arrange(desc(n)) |>
  datatable(
    caption = HTML(mark("FWS `status_category` distribution in raw data"))
  ) |>
  formatCurrency("n", digits = 0, currency = "")

d_fws |>
  count(esa_listing_status) |>
  arrange(desc(n)) |>
  datatable(
    caption = HTML(mark("FWS `esa_listing_status` distribution in raw data"))
  ) |>
  formatCurrency("n", digits = 0, currency = "")

d_fws <- d_fws |>
  rename(
    scientific_name_dataset = scientific_name,
    scientific_name_dataset_url = scientific_name_url
  ) |>
  mutate(
    # clean scientific name: strip subspecific qualifiers
    sci_ds = scientific_name_dataset |>
      str_remove_all("\\(=[^)]+\\)") |>
      str_remove_all("\\s+ssp\\.$") |>
      str_remove_all("\\s+sp\\.$") |>
      str_remove_all("\\s+sp\\.\\s+#?\\d+") |>
      str_remove_all("\\s+species$") |>
      str_squish(),
    itis_id = ifelse(
      taxonomic_serial_number < 0,
      NA,
      taxonomic_serial_number
    ) |>
      as.integer(),
    # map ESA status
    fws_esa = case_when(
      str_detect(esa_listing_status, "Endangered") ~ "EN",
      str_detect(esa_listing_status, "Threatened") ~ "TN",
      TRUE ~ "LC"
    ),
    is_bcc = is_bcc %in% c(TRUE, "TRUE")
  ) |>
  relocate(sci_ds)

d_fws |>
  filter(scientific_name_dataset == "Acipenser oxyrinchus oxyrinchus") |>
  datatable_with_links(
    caption = "Example of FWS species with multiple distinct population segments (DPS)"
  )

d_fws |>
  filter(sci_ds != scientific_name_dataset) |>
  select(
    sci_ds,
    scientific_name_dataset,
    scientific_name_dataset_url,
    status_category,
    esa_listing_status
  ) |>
  arrange(scientific_name_dataset, sci_ds) |>
  datatable_with_links(
    caption = "FWS species where scientific name has been cleaned."
  )

d_fws |>
  datatable_with_links(
    caption = "FWS species with cleaned scientific names (`sci_ds`) that differ from original `scientific_name_dataset`."
  )

# ensure no NA sci_ds
stopifnot(!any(is.na(d_fws$sci_ds)))

# handle birds ----
spp_fws_birds <- d_fws |>
  filter(
    taxonomic_group == "Birds"
  ) |>
  pull(sci_ds)

# TODO: add spp.botw.taxonRank
# TODO: add spp.botw.profile_code to spp.botw eg _Rhea pennata_; botw_id: 22678081; commonName: Lesser Rhea; https://birdsoftheworld.org/bow/species/lesrhe2/cur/introduction; so profile_code: lesrhe2

# d_bf: birds of the world & fws matches
d_bf <- tbl(con_spp, "botw") |>
  filter(
    scientificName %in% !!unique(spp_fws_birds)
  ) |>
  select(
    sci_ds = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    botw_id = taxonID,
    redlist_code
  ) |>
  collect()

# d_wf: worms_fws ---
d_wf_direct <- tbl(con_spp, "worms") |>
  filter(
    scientificName %in% !!unique(d_fws$sci_ds),
    taxonomicStatus == "accepted"
  ) |>
  select(
    worms_id = taxonID,
    sci_ds = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    is_marine = isMarine,
    taxon_rank = taxonRank
  ) |>
  collect()

sci_indirect = setdiff(d_fws$sci_ds, d_wf_direct$sci_ds) |>
  unique()
# Lontra longicaudis (incl. platensis)

d_wf_indirect <- tbl(con_spp, "worms") |>
  filter(
    scientificName %in% !!sci_indirect
  ) |>
  select(
    sci_ds = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    worms_id = acceptedNameUsageID,
  ) |>
  left_join(
    tbl(con_spp, "worms") |>
      select(
        worms_id = taxonID,
        taxon_rank = taxonRank,
        is_marine = isMarine
      ),
    by = join_by(worms_id)
  ) |>
  collect()

# nrow(d_wf_direct) # nrow: 2,070
# nrow(d_wf_indirect) # nrow:   276

d_wf <- bind_rows(d_wf_direct, d_wf_indirect) |>
  mutate(
    worms_id_url = glue(
      "https://www.marinespecies.org/aphia.php?p=taxdetails&id={worms_id}"
    )
  ) |>
  filter(!is.na(worms_id))

# intersect(names(d_bf), names(d_wf)) # "sci_ds"    "scientific_name_accepted"
# setdiff(names(d_bf), names(d_wf))   # "botw_id"   "redlist_code"
# setdiff(names(d_wf), names(d_bf))   # "worms_id"  "is_marine"    "taxon_rank"   "worms_id_url"

d_b_wf <- d_bf |>
  inner_join(
    d_wf |>
      select(
        sci_ds,
        worms_id,
        worms_id_url,
        taxon_rank,
        is_marine
      ),
    by = join_by(sci_ds)
  ) |>
  filter(!is.na(worms_id) & !is.na(botw_id))

d_b_notwf <- d_bf |>
  anti_join(
    d_b_wf,
    by = join_by(sci_ds)
  ) |>
  filter(!is.na(botw_id))

d_wf_notb <- d_wf |>
  anti_join(
    d_b_wf,
    by = join_by(sci_ds)
  ) |>
  filter(!is.na(worms_id))

d_bwf <- bind_rows(
  d_b_wf,
  d_b_notwf,
  d_wf_notb
) |>
  arrange(sci_ds, scientific_name_accepted, botw_id, worms_id)

d_bwf |>
  select(-worms_id, -worms_id_url, -taxon_rank, -is_marine) |>
  filter(
    !is.na(botw_id)
  ) |>
  arrange(sci_ds, scientific_name_accepted, botw_id) |>
  datatable_with_links(
    caption = HTML(mark(
      "BOTW matches to FWS species with duplicated BOTW IDs to be collapsed with rest of FWS species data."
    ))
  )

d_bwf |>
  filter(!is.na(worms_id)) |>
  count(is_marine) |>
  datatable(
    caption = HTML(mark(
      "Summary of WoRMS matches to FWS species by whether occurs in the marine environment (`is_marine`). Those explicitly not (ie 'false') get excluded."
    ))
  )

sci_w_notmarine <- d_bwf |>
  filter(is_marine == FALSE) |>
  pull(sci_ds)
d_bwf <- d_bwf |> filter(!sci_ds %in% sci_w_notmarine)

d_bwf |>
  filter(!is.na(worms_id)) |>
  select(-botw_id, -redlist_code) |>
  datatable_with_links(
    caption = HTML(mark(
      "WoRMS matches to FWS species via cleaned dataset scientific name (`sci_ds`) and `is_marine == TRUE` filter."
    ))
  )

stopifnot(sum(duplicated(d_bwf$sci_ds)) == 0)
# sum(duplicated(d_bwf$worms_id)) # 5

d_bwf |>
  select(-botw_id, -redlist_code) |>
  filter(
    !is.na(worms_id),
    duplicated(worms_id) |
      duplicated(worms_id, fromLast = TRUE)
  ) |>
  arrange(worms_id, sci_ds, scientific_name_accepted) |>
  datatable_with_links(
    caption = HTML(mark(
      "WoRMS matches to FWS species with duplicated WoRMS IDs to be collapsed with rest of FWS species data."
    ))
  )

d_fbw <- d_fws |>
  left_join(
    d_bwf,
    by = join_by(sci_ds)
  ) |>
  filter(
    !is.na(worms_id) | !is.na(botw_id),
  )

# nrow(d_fws_raw) # 10,998
# nrow(d_fws)     #  4,470
# nrow(d_bwf)      #   984
# nrow(d_fbw)      # 1,114

d_fbw |>
  filter(
    duplicated(worms_id) | duplicated(worms_id, fromLast = TRUE)
  ) |>
  relocate(worms_id, sci_ds, scientific_name_dataset) |>
  arrange(worms_id, sci_ds, scientific_name_dataset) |>
  datatable_with_links(
    caption = HTML(mark(
      "FWS species with duplicated WoRMS IDs after join (`d_fbw`) to be collapsed by taking max ESA status."
    ))
  )

d_fbw |>
  filter(
    duplicated(botw_id) | duplicated(botw_id, fromLast = TRUE)
  ) |>
  relocate(botw_id, sci_ds, scientific_name_dataset) |>
  arrange(botw_id, sci_ds, scientific_name_dataset) |>
  datatable_with_links(
    caption = HTML(mark(
      "FWS species with duplicated BOTW IDs after join (`d_fbw`) to be collapsed by taking max ESA status."
    ))
  )

# d_fw$esa_listing_status |> table(useNA = "ifany")
# d_fw$fws_esa |> table(useNA = "ifany")

d_f_w <- d_fbw |>
  mutate(
    fws_esa = factor(fws_esa, levels = esa_levels, ordered = TRUE)
  ) |>
  summarize(
    botw_id = first(botw_id),
    redlist_code = first(redlist_code),
    worms_id_url = first(worms_id_url),
    scientific_names_dataset = paste(
      unique(scientific_name_dataset),
      collapse = "; "
    ),
    scientific_names_dataset_url = first(scientific_name_dataset_url),
    scientific_name_accepted = first(scientific_name_accepted),
    taxon_rank = first(taxon_rank),
    common_name = common_words(common_name),
    taxonomic_group = first(taxonomic_group),
    entity_descriptions = paste(unique(entity_description), collapse = "; "),
    lead_agency = paste(unique(lead_agency), collapse = "; "),
    fws_esa = max(fws_esa),
    fws_esa_all = paste(unique(fws_esa), collapse = "; "),
    is_bcc = any(is_bcc),
    .by = worms_id
  )

spp_w_b <- d_f_w |>
  filter(!is.na(botw_id)) |>
  pull(worms_id)

d_f_b <- d_fbw |>
  filter(
    !worms_id %in% spp_w_b
  ) |>
  mutate(
    fws_esa = factor(fws_esa, levels = esa_levels, ordered = TRUE)
  ) |>
  summarize(
    redlist_code = first(redlist_code),
    worms_id     = first(worms_id),
    worms_id_url = first(worms_id_url),
    scientific_names_dataset = paste(
      unique(scientific_name_dataset),
      collapse = "; "
    ),
    scientific_names_dataset_url = first(scientific_name_dataset_url),
    scientific_name_accepted = first(scientific_name_accepted),
    taxon_rank = first(taxon_rank),
    common_name = common_words(common_name),
    taxonomic_group = first(taxonomic_group),
    entity_descriptions = paste(unique(entity_description), collapse = "; "),
    lead_agency = paste(unique(lead_agency), collapse = "; "),
    fws_esa = max(fws_esa),
    fws_esa_all = paste(unique(fws_esa), collapse = "; "),
    is_bcc = any(is_bcc),
    .by = botw_id
  )

d_f_bw <- d_f_w |>
  bind_rows(
    d_f_b
  ) |>
  arrange(scientific_name_accepted)

write_csv(d_f_bw, f_bw_csv)

d_f_bw |>
  datatable_with_links(
    caption = HTML(mark(
      "FWS species after collapsing duplicates by WoRMS ID (`worms_id`) and Birds of the World ID (`botw_id`) and taking max ESA status."
    ))
  )

d_f_bw |>
  count(fws_esa, taxonomic_group, name = "n_species") |>
  arrange(desc(n_species)) |>
  datatable(
    caption = HTML(mark(
      "Summary of FWS species by ESA status and taxonomic group sorted by `n_species` after collapsing duplicates by WoRMS ID (`worms_id`) and Birds of the World ID (`botw_id`)."
    ))
  )

d_f_bw |> 
  filter(taxonomic_group == "Birds") |> 
  count(!is.na(botw_id), !is.na(worms_id)) |> 
  datatable(
    caption = HTML(mark(
      "Summary of FWS bird species by presence of WoRMS ID (`worms_id`) and Birds of the World ID (`botw_id`)."
    ))
  )
```
  
## Ingest MBTA Bird List
  
Source: CFR 50 Part 10.13 Excel file listing MBTA-protected birds.
  
```{r}
#| label: ingest_mbta

d_mbta_raw <- read_excel(mbta_xlsx, sheet = 2) |>
  clean_names()

d_mbta <- d_mbta_raw |>
  rename(
    common_name = english_name,
  ) |> 
  mutate(
    scientific_name = str_squish(scientific_name),
    is_mbta = TRUE
  )

d_wm_direct <- tbl(con_spp, "worms") |>
  filter(
    scientificName %in% !!unique(d_mbta$scientific_name),
    taxonomicStatus == "accepted"
  ) |>
  select(
    scientific_name = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    worms_id = taxonID,
    taxon_rank = taxonRank,
    is_marine = isMarine
  ) |>
  collect()

sci_indirect <- setdiff(d_mbta$scientific_name, d_wm_direct$scientific_name) |>
  unique()

d_wm_indirect <- tbl(con_spp, "worms") |>
  filter(
    scientificName %in% !!sci_indirect
  ) |>
  select(
    scientific_name = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    worms_id = acceptedNameUsageID,
  ) |>
  left_join(
    tbl(con_spp, "worms") |>
      select(
        worms_id = taxonID,
        taxon_rank = taxonRank,
        is_marine = isMarine
      ),
    by = join_by(worms_id)
  ) |>
  collect()

d_wm <- bind_rows(d_wm_direct, d_wm_indirect) |>
  mutate(
    worms_id_url = glue(
      "https://www.marinespecies.org/aphia.php?p=taxdetails&id={worms_id}"
    )
  ) |>
  filter(!is.na(worms_id))

d_bm <- tbl(con_spp, "botw") |>
  filter(
    scientificName %in% !!unique(d_mbta$scientific_name)
  ) |>
  select(
    scientific_name = scientificName,
    scientific_name_accepted = acceptedNameUsage,
    botw_id = taxonID,
    redlist_code
  ) |>
  collect()

d_b_wm <- d_bm |>
  inner_join(
    d_wm |>
      select(
        scientific_name,
        worms_id,
        worms_id_url,
        taxon_rank,
        is_marine
      ),
    by = join_by(scientific_name)
  ) |>
  filter(!is.na(worms_id) & !is.na(botw_id))

d_b_notwm <- d_bm |>
  anti_join(
    d_b_wm,
    by = join_by(scientific_name)
  ) |>
  filter(!is.na(botw_id))

d_wm_notb <- d_wm |>
  anti_join(
    d_b_wm,
    by = join_by(scientific_name)
  ) |>
  filter(!is.na(worms_id))

d_bwm <- bind_rows(
  d_b_wm,
  d_b_notwm,
  d_wm_notb
) |>
  arrange(scientific_name, scientific_name_accepted, botw_id, worms_id) |>
  filter(is.na(is_marine) | is_marine == TRUE)

d_m_bw <- d_mbta |>
  left_join(d_bwm, by = "scientific_name") |>
  filter(!is.na(worms_id) | !is.na(botw_id)) |>
  # collapse synonyms: when multiple MBTA names resolve to same worms_id,
  # keep the row with botw_id (accepted species in BOTW)
  arrange(is.na(botw_id)) |>
  filter(is.na(worms_id) | !duplicated(worms_id))


# check duplicated
stopifnot(sum(duplicated(na.omit(d_m_bw$worms_id))) == 0)
stopifnot(sum(duplicated(na.omit(d_m_bw$botw_id))) == 0)

write_csv(d_m_bw, mbta_botw_worms_csv)

d_m_bw |>
  datatable_with_links(
    caption = HTML(mark(
      "MBTA-protected birds matched to WoRMS and Birds of the World."
    ))
  )
```

## Combine and compute `extrisk_code`

Priority: NMFS EN > FWS EN > NMFS TN > FWS TN > NMFS LC > FWS LC

```{r}
#| label: combine_listings

# combine NMFS and FWS listings
d_listing <- d_nw |>
  select(worms_id, nmfs_esa, is_mmpa) |>
  full_join(
    d_f_bw |>
      select(worms_id, botw_id, fws_esa, is_bcc),
    by = "worms_id",
    na_matches = "never"
  ) |>
  left_join(
    d_m_bw |>
      select(botw_id, is_mbta, redlist_code),
    by = "botw_id",
    na_matches = "never"
  ) |>
  mutate(
    is_mmpa = replace_na(is_mmpa, FALSE),
    is_mbta = replace_na(is_mbta, FALSE),
    is_bcc  = replace_na(is_bcc, FALSE)
  )

# assign extrisk_code with priority ordering
d_listing <- d_listing |>
  mutate(
    extrisk_code = case_when(
      nmfs_esa == "EN" ~ "NMFS:EN",
      nmfs_esa == "TN" ~ "NMFS:TN",
      nmfs_esa == "LC" ~ "NMFS:LC",
      fws_esa == "EN" ~ "FWS:EN",
      fws_esa == "TN" ~ "FWS:TN",
      fws_esa == "LC" ~ "FWS:LC",
      TRUE ~ NA_character_
    ),
    er_score = msens::compute_er_score(extrisk_code, is_mmpa, is_mbta)
  )

d_listing |>
  count(extrisk_code, name = "n_species") |>
  arrange(desc(n_species))
```

## Apply Species Overrides

Manual overrides from `species_overrides_v3.csv`.

```{r}
#| label: apply_overrides

d_overrides <- read_csv(overrides_csv)

# apply extrisk_code overrides
ovr_extrisk <- d_overrides |>
  filter(override == "extrisk_code") |>
  select(worms_id, extrisk_code_ovr = value)

ovr_is_mmpa <- d_overrides |>
  filter(override == "is_mmpa") |>
  mutate(value = as.logical(value)) |>
  select(worms_id, is_mmpa_ovr = value)

ovr_er_score <- d_overrides |>
  filter(override == "er_score") |>
  mutate(value = as.integer(value)) |>
  select(worms_id, er_score_ovr = value)

# merge overrides
d_listing <- d_listing |>
  left_join(ovr_extrisk, by = "worms_id") |>
  left_join(ovr_is_mmpa, by = "worms_id") |>
  left_join(ovr_er_score, by = "worms_id") |>
  mutate(
    extrisk_code = coalesce(extrisk_code_ovr, extrisk_code),
    is_mmpa = coalesce(is_mmpa_ovr, is_mmpa),
    # recompute er_score with updated extrisk_code and is_mmpa
    er_score = coalesce(
      er_score_ovr,
      msens::compute_er_score(extrisk_code, is_mmpa, is_mbta)
    )
  ) |>
  select(-ends_with("_ovr"))

# ensure species with overrides but not already in listing are added
ovr_missing <- d_overrides |>
  filter(!worms_id %in% d_listing$worms_id) |>
  distinct(worms_id)

if (nrow(ovr_missing) > 0) {
  message(glue(
    "Adding {nrow(ovr_missing)} override-only species to listing"
  ))
  d_ovr_add <- ovr_missing |>
    left_join(ovr_extrisk, by = "worms_id") |>
    left_join(ovr_is_mmpa, by = "worms_id") |>
    left_join(ovr_er_score, by = "worms_id") |>
    mutate(
      extrisk_code = extrisk_code_ovr,
      is_mmpa = replace_na(is_mmpa_ovr, FALSE),
      is_mbta = FALSE,
      is_bcc = FALSE,
      er_score = coalesce(
        er_score_ovr,
        msens::compute_er_score(extrisk_code, is_mmpa, is_mbta)
      )
    ) |>
    select(-ends_with("_ovr"))
  d_listing <- bind_rows(d_listing, d_ovr_add)
}

# spot-check overrides
d_listing |>
  filter(worms_id %in% d_overrides$worms_id) |>
  select(worms_id, extrisk_code, er_score, is_mmpa, is_mbta) |>
  arrange(worms_id)
```

## Write staging `listing` table

```{r}
#| label: write_listing

d_listing_out <- d_listing |>
  filter(!is.na(extrisk_code)) |>
  mutate(spp_id = coalesce(worms_id, botw_id)) |>
  select(
    spp_id,
    worms_id,
    botw_id,
    extrisk_code,
    er_score,
    is_mmpa,
    is_mbta,
    is_bcc
  ) |>
  distinct(spp_id, .keep_all = TRUE)

if (dbExistsTable(con_sdm, "listing")) {
  dbExecute(con_sdm, "DROP TABLE listing")
}
dbWriteTable(con_sdm, "listing", d_listing_out, overwrite = TRUE)

message(glue(
  "Wrote {nrow(d_listing_out)} rows to listing table in sdm_2026.duckdb"
))

tbl(con_sdm, "listing") |>
  count(extrisk_code, name = "n") |>
  arrange(desc(n))
```

## Score Scale Impact

Comparison of old `rl_score` vs new `er_score` scoring:

| Species type | Old rl_score | New er_score | Change |
|---|---|---|---|
| ESA Endangered mammal | 0.8 | 100 | +25% |
| ESA Threatened | 0.6 | 50 | -17% |
| MMPA-only mammal | 0.2 (LC default) | 20 (max of 1, 20) | +5% |
| MBTA-only bird | 0.2 (LC default) | 10 (max of 1, 10) | -45% |
| IUCN CR (no US listing) | 1.0 | 50 | -50% |
| IUCN EN (no US listing) | 0.8 | 25 | -69% |
| No listing at all | 0.2 | 1 | -95% |

## Audit Outputs

```{r}
#| label: audit

# IUCN CR/EN species NOT covered by NMFS/FWS listings
d_iucn_rl <- tbl(con_spp, "iucn_redlist") |>
  filter(red_list_category_code %in% c("CR", "EN")) |>
  select(
    taxon_scientific_name,
    iucn_code = red_list_category_code
  ) |>
  left_join(
    tbl(con_spp, "worms") |>
      filter(taxonomicStatus == "accepted") |>
      select(
        taxon_scientific_name = scientificName,
        worms_id = taxonID
      ),
    by = "taxon_scientific_name"
  ) |>
  filter(!is.na(worms_id)) |>
  collect()

d_iucn_gap <- d_iucn_rl |>
  anti_join(
    d_listing_out |> filter(!is.na(extrisk_code)),
    by = "worms_id"
  )

if (nrow(d_iucn_gap) > 0) {
  message(glue(
    "Gap analysis: {nrow(d_iucn_gap)} IUCN CR/EN species without US listing"
  ))
  write_csv(
    d_iucn_gap,
    overrides_audit_csv
  )
}

# all species where US listing overrides IUCN
d_us_override <- d_listing_out |>
  filter(str_detect(extrisk_code, "^(NMFS|FWS):")) |>
  inner_join(d_iucn_rl, by = "worms_id")

if (nrow(d_us_override) > 0) {
  write_csv(
    d_us_override,
    here("data/audit_us_overrides_iucn_v3.csv")
  )
}

# summary of er_score distribution
d_listing_out |>
  mutate(
    er_bucket = cut(
      er_score,
      breaks = c(0, 1, 5, 25, 50, 100),
      include.lowest = TRUE
    )
  ) |>
  count(er_bucket, name = "n_species")
```

## Verification vs Reference

```{r}
#| label: verify_reference
#| eval: false

con_ref <- DBI::dbConnect(duckdb::duckdb(
  dbdir = msens::sdm_db_path("2026-01-29"),
  read_only = TRUE
))

# compare old redlist_code distribution
d_old <- tbl(con_ref, "taxon") |>
  count(redlist_code, name = "n_old") |>
  collect()

d_new <- tbl(con_sdm, "listing") |>
  count(extrisk_code, name = "n_new") |>
  collect()

message("Old redlist_code distribution:")
print(d_old)
message("\nNew extrisk_code distribution:")
print(d_new)

DBI::dbDisconnect(con_ref, shutdown = TRUE)
```

```{r}
#| label: cleanup

DBI::dbDisconnect(con_spp, shutdown = TRUE)
DBI::dbDisconnect(con_sdm, shutdown = TRUE)
```
