---
title: "Ingest NMFS ESA Critical Habitat → global 0.05° cells"
msens:
  target_name: ingest_ch_nmfs
  workflow_type: ingest
  dependency: [build_cell_grid]
  output: data/manifests/ingest_ch_nmfs.json
  dataset: {ds_key: ch_nmfs, response_type: range, source_authority: NOAA NMFS, temporal_interval: static, native_format: vector}
editor_options:
  chunk_output_type: console
---

Rasterize each species' **NMFS ESA critical-habitat** polygon(s) onto the **global
0.05° cell grid** and write one Parquet per species to the marine-atlas release,
keyed by `mdl_key = "ch_nmfs|{sp_id}"` (`sp_id` = a slug of the scientific name —
NMFS critical habitat has no numeric species id).

**Method.** `msens::cells_from_ranges()` (exactextractr) marks every cell the
(unioned) polygon overlaps, capturing the whole footprint with no land mask (here
ocean anyway, critical habitat being marine). The **val is the species'
extinction-risk score** via `msens::compute_er_score()` — never hard-coded — from
its most-protective ESA status (`NMFS:EN` = 100, `NMFS:TN` = 50). This is the shared
**vector → cells** pattern reused by every range dataset (BOTW, FWS, NMFS core
areas, turtles).

Source `ply.gpkg` is derived from the NMFS ESA Critical Habitat geodatabase (see
`old/ingest_fisheries.noaa.gov_critical-habitat_v7.qmd` for that one-time
extraction + QA maps). Re-run control: `REDO_INGEST=1` rebuilds all models (else
the loop resumes, skipping species whose Parquet exists).

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "NMFS ESA critical habitat → extinction-risk value on the global 0.05° grid"
flowchart LR
  src["ply.gpkg<br/>per-species CH polygons"] --> mdl["one model per SCIENAME<br/>most-protective ESA status"]
  mdl --> er["val = compute_er_score('NMFS:code')<br/>EN100 TN50"]
  er --> ee["cells_from_ranges()<br/>exact_extract vs 0.05° cellid COG"]
  ee --> pq[("dist/dataset=ch_nmfs/{sp_id}.parquet<br/>(mdl_key, cell_id, val) · V2/zstd")]
  pq --> mf["hash_parquet → content-addressed manifest"]
```

## Setup

```{r}
#| label: setup

librarian::shelf(
  arrow, DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, readr, sf, terra,
  tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))   # dir_raw, dir_big_v, cellid_tif, ver, s3_atlas
source(here("libs/vars.R"))    # redo_ingest
options(readr.show_col_types = F)
sf_use_s2(FALSE)               # critical-habitat polygons are planar-valid, s2-invalid (self-
                               # crossing loops); planar GEOS unions/valids them (matches ch_fws/rng_fws)

ds_key   <- "ch_nmfs"
ply_geo  <- glue("{dir_raw}/fisheries.noaa.gov/ply.gpkg")   # per-species critical-habitat polygons
dir_dist <- glue("{dir_big_v}/marine-atlas/dist/dataset={ds_key}")
manifest <- here("data/manifests/ingest_ch_nmfs.json")
dir_create(c(dir_dist, path_dir(manifest)))

stopifnot(
  "run build_cell_grid.qmd first (need cellid_tif)" = file_exists(cellid_tif),
  "ply.gpkg missing (derive it, see old/..._v7.qmd)" = file_exists(ply_geo))
```

## Read polygons + per-species model crosswalk

One model per species (`SCIENAME`); its multiple critical-habitat polygons are
unioned at rasterize time.

```{r}
#| label: model_csv

ply <- st_read(ply_geo, quiet = TRUE)
sp  <- ply |>
  st_drop_geometry() |>
  group_by(SCIENAME) |>                       # one model per species (polygons unioned below)
  summarise(
    TAXON = first(TAXON), COMNAME = first(COMNAME),
    # most-protective ESA status among the species' critical-habitat features
    status = if_else(any(grepl("Endangered", LISTSTATUS)), "EN", "TN"), .groups = "drop") |>
  arrange(SCIENAME) |>
  mutate(
    sp_id    = gsub("^_|_$", "", gsub("[^A-Za-z0-9]+", "_", SCIENAME)),
    mdl_key  = mdl_key_raw(ds_key, sp_id),
    er_code  = paste0("NMFS:", status),
    er_score = compute_er_score(er_code))     # val via compute_er_score(), never hard-coded
stopifnot(!any(duplicated(sp$sp_id)))
write_csv(sp, glue("{dir_dist}/../model_{ds_key}.csv"))
nrow(sp)
```

## Rasterize each range → one Parquet each

```{r}
#| label: interpolate

if (redo_ingest && dir_exists(dir_dist)) { dir_delete(dir_dist); dir_create(dir_dist) }
done <- path_ext_remove(path_file(dir_ls(dir_dist, glob = "*.parquet")))
todo <- sp |> filter(!sp_id %in% done)
log_info("{ds_key}: {length(done)} done; rasterizing {nrow(todo)} of {nrow(sp)} ranges ...")

for (i in seq_len(nrow(todo))) {
  row    <- todo[i, ]
  ply_sp <- ply |> filter(SCIENAME == row$SCIENAME)   # no pre-union: cells_from_ranges handles multi-polygon
  d <- cells_from_ranges(ply_sp, cellid_tif, value = row$er_score)   # ER score, not hard-coded
  if (nrow(d) == 0) { log_warn("  {row$sp_id}: no ocean cells covered"); next }
  msens::write_atlas_parquet(
    tibble(mdl_key = row$mdl_key, cell_id = d$cell_id, val = d$val),
    path(dir_dist, row$sp_id, ext = "parquet"))
}
log_info("{ds_key}: done")
```

## Verify

```{r}
#| label: verify

pq   <- dir_ls(dir_dist, glob = "*.parquet")
con  <- dbConnect(duckdb())
smry <- dbGetQuery(con, glue(
  "SELECT count(DISTINCT mdl_key) n_models, count(*) n_cells,
          min(val) v_min, max(val) v_max
     FROM read_parquet('{dir_dist}/*.parquet')"))
# order-independent content fingerprint of the on-disk surface (not the ingest run)
h <- msens::hash_parquet(glue("{dir_dist}/*.parquet"), con)
dbDisconnect(con, shutdown = TRUE)
msens::report_table(smry, caption = "ch_nmfs model_cell surface")
```

## Manifest (target output)

```{r}
#| label: manifest

# content-addressed manifest: deterministic, no wall-clock, no machine paths ->
# downstream targets re-run only when this surface's content actually changes
msens::write_manifest(
  manifest, target = "ingest_ch_nmfs", content_hash = h,
  stats = list(ds_key = ds_key, ver = ver, n_models = nrow(sp),
               n_cells = smry$n_cells, v_min = smry$v_min, v_max = smry$v_max,
               n_parquet = length(pq), source = path_file(ply_geo)),
  force = msens::force_target("ingest_ch_nmfs"))
```
