---
title: "Ingest SWOT + NMFS-DPS sea turtle ranges → global 0.05° cells"
msens:
  target_name: ingest_turtles_swot_dps
  workflow_type: ingest
  dependency: [build_cell_grid]
  output: data/manifests/ingest_turtles_swot_dps.json
  dataset: {ds_key: rng_turtle_swot_dps, response_type: range, source_authority: SWOT/NOAA NMFS, temporal_interval: static, native_format: vector}
editor_options:
  chunk_output_type: console
---

Refined sea-turtle ranges onto the **global 0.05° cell grid**, replacing the overly
broad AquaMaps/IUCN turtle ranges. Range polygons come from **SWOT** global
distributions; **NMFS DPS** boundaries mark the Endangered sub-populations.

**Value = extinction-risk score** via `msens::compute_er_score()` (never hard-coded).
For the three species with DPS data (Loggerhead CC, Green CM, Olive Ridley LO):
cells inside an **Endangered DPS → `compute_er_score("NMFS:EN")` = 100**, the rest
of the range (Threatened) **→ `compute_er_score("NMFS:TN")` = 50**. The three
pure-Endangered species (Leatherback DC, Hawksbill EI, Kemp's Ridley LK): 100
throughout. The EN/TN split is done at the **cell level** (rasterize both, combine)
— no fragile geometry intersection. Whole range, no land mask (turtles are marine,
so these are ocean cells). `mdl_key = "rng_turtle_swot_dps|{code}"`.

## Setup

```{r}
#| label: setup

librarian::shelf(
  arrow, DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, readr, sf, terra,
  tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))   # dir_raw, dir_big_v, cellid_tif, ver
source(here("libs/vars.R"))    # redo_ingest
options(readr.show_col_types = F)

ds_key   <- "rng_turtle_swot_dps"
dir_swot <- glue("{dir_raw}/swot_seamap.env.duke.edu/swot_distribution")
dir_dps  <- glue("{dir_raw}/fisheries.noaa.gov/turtle-DPS_endangered")
dir_dist <- glue("{dir_big_v}/marine-atlas/dist/dataset={ds_key}")
manifest <- here("data/manifests/ingest_turtles_swot_dps.json")
dir_create(c(dir_dist, path_dir(manifest)))
stopifnot("run build_cell_grid.qmd first" = file_exists(cellid_tif), dir_exists(dir_swot))

# range value = extinction-risk score via msens::compute_er_score() (NEVER hard-code).
# Sea turtles are ESA-listed marine species (NMFS authority): EN=100, TN=50.
er_en <- compute_er_score("NMFS:EN")   # 100
er_tn <- compute_er_score("NMFS:TN")   # 50

swot_spp <- tribble(
  ~code, ~sci_name,                ~common_name,    ~has_dps, ~esa_pure,
  "CC",  "Caretta caretta",        "Loggerhead",    TRUE,     NA,
  "CM",  "Chelonia mydas",         "Green",         TRUE,     NA,
  "DC",  "Dermochelys coriacea",   "Leatherback",   FALSE,    "EN",
  "EI",  "Eretmochelys imbricata", "Hawksbill",     FALSE,    "EN",
  "LK",  "Lepidochelys kempii",    "Kemp's Ridley", FALSE,    "EN",
  "LO",  "Lepidochelys olivacea",  "Olive Ridley",  TRUE,     NA) |>
  mutate(mdl_key = mdl_key_raw(ds_key, code))
write_csv(swot_spp, glue("{dir_dist}/../model_{ds_key}.csv"))
```

## Rasterize each range (EN=100 / TN=50) → one Parquet each

```{r}
#| label: interpolate

if (redo_ingest && dir_exists(dir_dist)) { dir_delete(dir_dist); dir_create(dir_dist) }

for (i in seq_len(nrow(swot_spp))) {
  s      <- swot_spp[i, ]
  out_pq <- fs::path(dir_dist, s$code, ext = "parquet")
  if (file_exists(out_pq)) { log_info("{s$code}: skip (exists)"); next }

  swot <- st_read(glue("{dir_swot}/Global_Distribution_{s$code}.shp"), quiet = TRUE)
  if (s$has_dps) {
    dps     <- st_read(glue("{dir_dps}/{tolower(s$code)}_endangered.gpkg"), quiet = TRUE)
    d_range <- cells_from_ranges(swot, cellid_tif, value = er_tn)  # whole range = Threatened
    d_en    <- cells_from_ranges(dps,  cellid_tif, value = er_en)  # Endangered DPS areas
    d <- d_range |> mutate(value = if_else(cell_id %in% d_en$cell_id, er_en, er_tn))
  } else {
    d <- cells_from_ranges(swot, cellid_tif, value = er_en)        # pure Endangered
  }
  if (nrow(d) == 0) { log_warn("{s$code}: no cells"); next }
  write_parquet(tibble(mdl_key = s$mdl_key, cell_id = d$cell_id, value = d$value), out_pq)
  log_info("{s$code} ({s$common_name}): {nrow(d)} cells")
}
```

## Verify

```{r}
#| label: verify

pq   <- dir_ls(dir_dist, glob = "*.parquet")
con  <- dbConnect(duckdb())
smry <- dbGetQuery(con, glue(
  "SELECT replace(mdl_key,'rng_turtle_swot_dps|','') code, count(*) n_cells,
          count(*) FILTER (WHERE value = 100) n_en, count(*) FILTER (WHERE value = 50) n_tn
     FROM read_parquet('{dir_dist}/*.parquet') GROUP BY mdl_key ORDER BY code"))
dbDisconnect(con, shutdown = TRUE)
knitr::kable(smry)
```

## Manifest (target output)

```{r}
#| label: manifest

write_json(
  list(target = "ingest_turtles_swot_dps", ds_key = ds_key, ver = ver,
       n_species = nrow(swot_spp), n_parquet = length(pq),
       dir_dist = dir_dist, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
manifest
```
