---
title: "Ingest Turtle Distributions from SWOT Global Distributions and NMFS Distinct Population Segments (DPS)"
format: html
---

## Overview

The problem is that sea turtle distributions were too broad in the original Aquamaps, and some species have differential protection status across their range. The solution is to use the more refined distributions from the SWOT Global Distributions and NMFS Distinct Population Segments (DPS) datasets.

```{r}
#| label: setup

librarian::shelf(
  arcgislayers,
  arcgisutils,
  dplyr,
  DT,
  ggplot2,
  glue,
  here,
  janitor,
  litedown,
  mapview,
  purrr,
  readr,
  rlang,
  sf,
  stringr,
  terra,
  tibble,
  tidyr,
  quiet = T
)
options(readr.show_col_types = F)

# helper function to convert ArcGIS FeatureService to sf
fs_to_sf <- function(url, where = "LISTSTATUS IN ('Endangered')") {
  fs <- arcgislayers::arc_open(url)
  lyr <- arcgislayers::get_layer(fs, 0)

  sf <- arcgislayers::arc_select(
    lyr,
    where = where
  ) |>
    tibble::tibble() |>
    sf::st_as_sf() |>
    janitor::clean_names()

  return(sf)
}

# paths
spp_csv <- here("data/nmfs_species-directory.csv")
dir_dps <- "~/My Drive/projects/msens/data/raw/fisheries.noaa.gov/turtle-DPS"

# variables
dps = list(
  lo = "https://services2.arcgis.com/C8EMgrsFcRFL6LrL/arcgis/rest/services/SeaTurtleOliveRidley_PacificMexicoPopulation/FeatureServer",
  cc = "https://services2.arcgis.com/C8EMgrsFcRFL6LrL/arcgis/rest/services/SeaTurtleLoggerhead_AllDPS_20250609/FeatureServer",
  cm = "https://services2.arcgis.com/C8EMgrsFcRFL6LrL/arcgis/rest/services/SeaTurtleGreen_AllDPS_20250609/FeatureServer"
)
```

## NMFS Species Directory

- [Species Directory | NOAA Fisheries](https://www.fisheries.noaa.gov/species-directory)\
  CSV: [`nmfs_species-directory.csv`](https://github.com/MarineSensitivity/workflows/blob/main/data/nmfs_species-directory.csv)\
  generated by MarineSensitivity/workflows:ingest_fisheries.noaa.gov_critical-habitat.qmd
  

```{r}
#| label: spp_nmfs

d_spp <- read_csv(spp_csv) |>
  mutate(
    category = strsplit(category, ";\\s*"),
    is_en = str_detect(protected_status, "ESA Endangered"),
    is_tn = str_detect(protected_status, "ESA Threatened"),
    is_en_or_tn = (is_en | is_tn),
    is_en_and_tn = (is_en & is_tn)
  ) |>
  unnest(category) |>
  filter(category == "Sea Turtles")

d_spp |>
  rowwise() |>
  mutate(
    name = litedown::mark(
      text = glue("[{common_name}]({species_url}) (_{scientific_name}_)")
    ),
    across(where(is.logical), ~ ifelse(.x, "✅", "❌"))
  ) |>
  relocate(name) |>
  select(
    -c(common_name, scientific_name, species_url, also_known_as, category)
  ) |>
  datatable(escape = F)
```

In summary:

- **All** turtles are either ESA Endangered _**or**_ Threatened: `is_en_or_tn` = ✅
- **Some** turtles are both ESA Endangered _**and**_ Threatened: `is_en_and_tn` = ✅\
  By extracting the Distinct Population Segments that are Endangered, we can then assign the rest of the global population to Threatened.

## Fetch NOAA Fisheries Distinct Population Segments (DPS) from ArcGIS Feature Services

```{r}
#| label: fetch_dps

for (sp in names(dps)) {
  url <- dps[[sp]]
  gpkg <- glue("{dir_dps}/{sp}_endangered.gpkg")

  if (file.exists(gpkg)) {
    message(glue(
      "{sp}: SKIP fetching DPS since already exists:  {basename(gpkg)}"
    ))
    next
  }
  fs_to_sf(url) |>
    write_sf(gpkg, delete_dsn = TRUE)
  message(glue("{sp}: FETCHED DPS and saved to: {basename(gpkg)}"))
}

# read all gpkg files and bind into one sf object
sf <- list.files(dir_dps, pattern = "\\.gpkg$", full.names = TRUE) |>
  purrr::map_dfr(read_sf)

mapview::mapView(sf, zcol = "comname")
```


## ToDo

- [ ] Investigate other species for differential DPS status, ie:
    - ie: `is_en` = ✅ & `is_tn` = ✅
    - Find NOAA NMFS DPS, eg [NOAA.maps.arcgis.com Search 'AllDPS'](https://noaa.maps.arcgis.com/home/search.html?restrict=true&sortField=relevance&sortOrder=desc&searchTerm=AllDPS#content). 
- [ ] 
