---
title: "Ingest AquaMaps → global 0.05° cells (bilinear-weight interpolation)"
msens:
  target_name: ingest_aquamaps
  workflow_type: ingest
  dependency: [build_cell_grid]
  output: data/manifests/ingest_aquamaps.json
  dataset: {ds_key: am, response_type: suitability, source_authority: AquaMaps, temporal_interval: static, native_format: raster}
editor_options:
  chunk_output_type: console
---

Reingest **all** AquaMaps species (native HCAF 0.5°) onto the **global 0.05° cell
grid** ([-180,180], from `build_cell_grid.qmd`), writing one Parquet per species to
the marine-atlas release keyed by `mdl_key = "am|{sp_key}"`.

**Method — bilinear interpolation as a precomputed weight join (fast + edge-safe).**
The 0.5°→0.05° bilinear resample has *fixed* geometric weights (independent of
species), so we build them **once** as `w05(cell_id, loiczid, w)`: each 0.05° ocean
cell → its 4 surrounding HCAF 0.5° grid cells (`loiczid`) + bilinear weights (which
sum to 1). Then each species is interpolated by an indexed DuckDB join — no terra,
no per-species raster I/O — at ~0.1 s/species (~50× faster than the per-species
terra resample on shared storage).

**Edges.** Absent HCAF cells simply aren't in `spp_cells`, so they contribute 0 to
the weighted sum — the surface fades to 0 within ~one cell of the distribution edge
(the v7 "surround by zeros" behaviour). Validated cell-for-cell against the terra
bilinear result: values `cor > 0.999` (identical where both defined), edge fade
visually indistinguishable, no bleed.

Re-run control (`libs/vars.R`, override via env): `REDO_AM_W05=1` rebuilds the
weight table; `REDO_AM_INGEST=1` re-interpolates every species (otherwise the loop
resumes, skipping species whose Parquet already exists).

## Setup

```{r}
#| label: setup

librarian::shelf(
  DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, readr, terra, tibble,
  quiet = T)
source(here("libs/paths.R"))   # dir_derived, dir_big_v, cellid_tif, am_db, w05_db, ver, s3_atlas
source(here("libs/vars.R"))    # redo_am_w05, redo_am_ingest
options(readr.show_col_types = F)

ds_key    <- "am"
threshold <- 1                                                # drop interpolated suitability < 1
dir_dist  <- glue("{dir_big_v}/marine-atlas/dist/dataset={ds_key}")
manifest  <- here("data/manifests/ingest_aquamaps.json")
dir_create(c(dir_dist, path_dir(manifest)))

stopifnot(
  "run build_cell_grid.qmd first (need cellid_tif)" = file_exists(cellid_tif),
  "AquaMaps source am.duckdb not found"             = file_exists(am_db))
```

## Bilinear weight table `w05` (built once)

`w05` has 4 rows per 0.05° ocean cell — its bilinear stencil onto the HCAF grid.
`loiczid = row*720 + col + 1` is the HCAF grid position (verified `== cells.loiczid`).
~68M rows, ~30 s, indexed on `loiczid` so the per-species join is a fast lookup.

```{r}
#| label: build_w05

if (redo_am_w05 || !file_exists(w05_db)) {
  log_info("building AquaMaps bilinear weight table w05 (once) ...")
  if (file_exists(w05_db)) file_delete(w05_db)
  cw <- dbConnect(duckdb(w05_db))
  # AquaMaps is marine: restrict weights to OCEAN cells (cellid_tif is the GLOBAL
  # grid — land + ocean — so mask by the Bio-Oracle depth layer to keep ocean only)
  oc <- as.data.frame(
    mask(rast(cellid_tif), rast(bio_oracle_tif, lyrs = "depth_mean")),
    xy = TRUE, na.rm = TRUE)
  names(oc) <- c("lon", "lat", "cell_id")
  duckdb_register(cw, "cell_xy", oc)
  dbExecute(cw, "CREATE TABLE w05 AS
    WITH f AS (
      SELECT CAST(cell_id AS INTEGER)           AS cell_id,
             floor((lon+179.75)/0.5)            AS cl,
             floor((89.75-lat)/0.5)             AS rt,
             (lon-(-179.75+0.5*floor((lon+179.75)/0.5)))/0.5 AS fx,
             ((89.75-0.5*floor((89.75-lat)/0.5))-lat)/0.5    AS fy
      FROM cell_xy)
    SELECT cell_id, CAST(rt*720+cl+1     AS BIGINT) AS loiczid, (1-fx)*(1-fy) AS w FROM f WHERE (1-fx)*(1-fy) > 0
    UNION ALL SELECT cell_id, CAST(rt*720+cl+2     AS BIGINT), fx*(1-fy) FROM f WHERE fx*(1-fy) > 0
    UNION ALL SELECT cell_id, CAST((rt+1)*720+cl+1 AS BIGINT), (1-fx)*fy FROM f WHERE (1-fx)*fy > 0
    UNION ALL SELECT cell_id, CAST((rt+1)*720+cl+2 AS BIGINT), fx*fy     FROM f WHERE fx*fy     > 0")
  dbExecute(cw, "CREATE INDEX w05_loiczid ON w05(loiczid)")
  n_w <- dbGetQuery(cw, "SELECT count(*) n FROM w05")$n
  dbDisconnect(cw, shutdown = TRUE)
  log_info("w05 built: {format(n_w, big.mark=',')} weight rows")
}
```

## Species list + model crosswalk

```{r}
#| label: model_csv

con_am <- dbConnect(duckdb(am_db, read_only = TRUE))
sp_all <- tbl(con_am, "spp") |> select(sp_key, genus, species, iucn_id) |> collect()
dbDisconnect(con_am, shutdown = TRUE)
sp_all <- sp_all |>
  mutate(mdl_key = glue("{ds_key}|{sp_key}"), taxa = glue("{genus} {species}"))
write_csv(sp_all |> select(mdl_key, sp_key, taxa, iucn_id),
          glue("{dir_dist}/../model_{ds_key}.csv"))
nrow(sp_all)
```

## Interpolate every species → one Parquet each

Per species: join its present HCAF cells (`spp_cells → cells → loiczid`) to `w05`,
sum `w × probability×100` per 0.05° cell, keep `≥ threshold`. DuckDB `COPY` writes
the Parquet directly. Resumable; `REDO_AM_INGEST=1` starts fresh.

```{r}
#| label: interpolate

if (redo_am_ingest && dir_exists(dir_dist)) { dir_delete(dir_dist); dir_create(dir_dist) }
done <- path_ext_remove(path_file(dir_ls(dir_dist, glob = "*.parquet")))
todo <- sp_all$sp_key[!sp_all$sp_key %in% done]
log_info("AquaMaps: {length(done)} done; interpolating {length(todo)} of {nrow(sp_all)} species ...")

con <- dbConnect(duckdb(w05_db, read_only = TRUE))
dbExecute(con, glue("ATTACH '{am_db}' AS am (READ_ONLY)"))
dbExecute(con, glue("SET threads = {max(1L, parallel::detectCores())}"))
t0 <- Sys.time()
for (i in seq_along(todo)) {
  k <- todo[i]
  dbExecute(con, glue("
    COPY (
      SELECT 'am|{k}' AS mdl_key, w.cell_id, round(SUM(w.w * sc.prob), 2) AS \"val\"
      FROM w05 w
      JOIN (SELECT c.loiczid, s.probability*100 AS prob
              FROM am.spp_cells s JOIN am.cells c USING (cell_id)
             WHERE s.sp_key = '{k}') sc  ON w.loiczid = sc.loiczid
      GROUP BY w.cell_id
      HAVING SUM(w.w * sc.prob) >= {threshold}
    ) TO '{fs::path(dir_dist, k, ext = \"parquet\")}' (FORMAT parquet, COMPRESSION zstd)"))
  if (i %% 1000 == 0)
    log_info("  {i}/{length(todo)} ({round(as.numeric(Sys.time()-t0, units='mins'),1)} min)")
}
dbDisconnect(con, shutdown = TRUE)
log_info("interpolation done in {round(as.numeric(Sys.time()-t0, units='mins'),1)} min")
```

## Verify

```{r}
#| label: verify

n_pq <- length(dir_ls(dir_dist, glob = "*.parquet"))
con  <- dbConnect(duckdb())
smp  <- dbGetQuery(con, glue(
  "SELECT count(*) n_cells, min(val) v_min, max(val) v_max
     FROM read_parquet('{fs::path(dir_dist, sp_all$sp_key[1], ext=\"parquet\")}')"))
dbDisconnect(con, shutdown = TRUE)
glue("{n_pq} species Parquet written; sample {sp_all$sp_key[1]}: ",
     "{smp$n_cells} cells, val {smp$v_min}-{smp$v_max}")
```

## Manifest (target output)

```{r}
#| label: manifest

write_json(
  list(target = "ingest_aquamaps", ds_key = ds_key, ver = ver,
       n_species = nrow(sp_all), n_parquet = n_pq,
       dir_dist = dir_dist, w05_db = w05_db, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
manifest
```

## Publish to S3

```{r}
#| label: publish
#| eval: false

# aws s3 sync {dir_big_v}/marine-atlas s3://oceanmetrics.io-public/marine-atlas/v8
system(glue("aws s3 sync {dir_big_v}/marine-atlas {s3_atlas}/{ver}"))
```
