---
title: "Ingest NMFS Core Distribution Areas"
editor_options:
  chunk_output_type: console
---

## Ingest NMFS Core Distribution Areas

Rice's whale (*Balaenoptera ricei*) has a designated NMFS Core Distribution Area
that matches its IUCN range map footprint. Rather than rasterizing the shapefile
from scratch, we copy cell_ids from the existing IUCN range map (mdl_seq: 19014)
and assign maximum weighting (value = 100).

- [Rice's Whale | NOAA Fisheries](https://www.fisheries.noaa.gov/species/rices-whale)

```{r setup}
librarian::shelf(
  DBI, dplyr, duckdb, glue, here, sf, terra,
  quiet = T)

ds_key    <- "ca_nmfs"
is_server <- Sys.info()[["sysname"]] == "Linux"
dir_data  <- ifelse(is_server, "/share/data", "~/My Drive/projects/msens/data")
sdm_db    <- glue("{dir_data}/derived/sdm_2026.duckdb")
cell_tif  <- glue("{dir_data}/derived/r_bio-oracle_planarea.tif")

con_sdm <- dbConnect(duckdb(dbdir = sdm_db, read_only = F))
```

### Add `dataset`

```{r insert_dataset_row}
row_dataset <- tibble(
  ds_key          = !!ds_key,
  name_short      = "NMFS Core Areas, 2019",
  name_original   = "NMFS Core Distribution Areas",
  description     = "Core distribution areas designated by NOAA Fisheries for species with restricted ranges, used as high-confidence habitat indicators.",
  citation        = "",
  source_broad    = "NMFS",
  source_detail   = "https://www.fisheries.noaa.gov/resource/map/rices-whale-core-distribution-area-map-gis-data",
  regions         = "USA",
  response_type   = "binary",
  taxa_groups     = "cetaceans",
  year_pub        = 2019,
  date_obs_beg    = NA,
  date_obs_end    = NA,
  date_env_beg    = NA,
  date_env_end    = NA,
  link_info       = "https://www.fisheries.noaa.gov/species/rices-whale",
  link_download   = "https://www.fisheries.noaa.gov/s3/2021-09/shapefile_Rices_whale_core_distribution_area_Jun19_SERO.zip",
  link_metadata   = NA_character_,
  links_other     = NA_character_,
  spatial_res_deg = 0.05,
  temporal_res    = "static")

if (dbExistsTable(con_sdm, "dataset"))
  dbExecute(con_sdm, glue("DELETE FROM dataset WHERE ds_key = '{ds_key}'"))

dbWriteTable(con_sdm, "dataset", row_dataset, append = TRUE)
tbl(con_sdm, "dataset") |> filter(ds_key == !!ds_key) |> collect()
```

### Delete existing ca_nmfs data (idempotent re-runs)

```{r delete_existing}
mdl_seqs <- tbl(con_sdm, "model") |>
  filter(ds_key == !!ds_key) |>
  pull(mdl_seq)

if (length(mdl_seqs) > 0) {
  dbExecute(con_sdm, glue(
    "DELETE FROM model_cell WHERE mdl_seq IN ({paste(mdl_seqs, collapse = ',')})"))
  dbExecute(con_sdm, glue("DELETE FROM model   WHERE ds_key = '{ds_key}'"))
  dbExecute(con_sdm, glue("DELETE FROM species WHERE ds_key = '{ds_key}'"))
}
```

### Add Rice's whale: `model`, `species`, `model_cell`

Copy cell_ids from the existing IUCN range map (mdl_seq: 19014) for Rice's whale.

```{r add_rices_whale}
# rice's whale parameters
sp_sci       <- "Balaenoptera ricei"
sp_cmn       <- "Rice's whale"
worms_id     <- 1576133L
iucn_mdl_seq <- 19014L
sp_key       <- glue("{ds_key}:{sp_sci}")

# append: model ----
d_model <- tibble(
  ds_key      = ds_key,
  taxa        = sp_key,
  time_period = "2019",
  region      = "Gulf of Mexico",
  mdl_type    = "binary",
  description = glue(
    "NMFS core distribution area for {sp_sci}, copied from IUCN range map cells"))
dbWriteTable(con_sdm, "model", d_model, append = TRUE)

# get the mdl_seq that was just created
mdl_seq <- dbGetQuery(con_sdm, glue("
  SELECT mdl_seq FROM model
  WHERE
    ds_key = '{ds_key}' AND
    taxa   = '{sp_key}'
  ORDER BY mdl_seq DESC LIMIT 1 "))$mdl_seq
message(glue("New mdl_seq for {sp_sci}: {mdl_seq}"))

# append: species ----
d_species <- tibble(
  ds_key                  = ds_key,
  taxa                    = sp_key,
  sp_key                  = sp_key,
  worms_id                = worms_id,
  botw_id                 = NA_integer_,
  gbif_id                 = NA_integer_,
  itis_id                 = NA_integer_,
  iucn_id                 = NA_integer_,
  scientific_name_dataset  = sp_sci,
  common_name_dataset      = sp_cmn,
  scientific_name_accepted = sp_sci,
  common_name_accepted     = sp_cmn,
  redlist_code             = "EN",
  redlist_year             = NA_integer_,
  sp_cat                   = "mammal",
  worms_is_marine          = TRUE,
  worms_is_extinct         = FALSE)

stopifnot(length(setdiff(names(d_species), dbListFields(con_sdm, "species"))) == 0)
stopifnot(setdiff(dbListFields(con_sdm, "species"), names(d_species)) == "sp_seq")
dbWriteTable(con_sdm, "species", d_species, append = T)

# append: model_cell (copy from IUCN range map with value = 100) ----
d_iucn_cells <- tbl(con_sdm, "model_cell") |>
  filter(mdl_seq == !!iucn_mdl_seq) |>
  select(cell_id) |>
  collect()
message(glue("Copying {nrow(d_iucn_cells)} cells from IUCN mdl_seq {iucn_mdl_seq}"))

d_mdl_cell <- d_iucn_cells |>
  mutate(
    mdl_seq = mdl_seq,
    value   = 100L) |>
  arrange(cell_id)
dbWriteTable(con_sdm, "model_cell", d_mdl_cell, append = T)
```

### Verify

```{r verify}
# verify dataset row
tbl(con_sdm, "dataset") |> filter(ds_key == !!ds_key) |> collect()

# verify model row
tbl(con_sdm, "model") |> filter(ds_key == !!ds_key) |> collect()

# verify species row
tbl(con_sdm, "species") |> filter(ds_key == !!ds_key) |> collect()

# verify model_cell count matches IUCN range
n_ca   <- tbl(con_sdm, "model_cell") |> filter(mdl_seq == !!mdl_seq) |> count() |> pull(n)
n_iucn <- tbl(con_sdm, "model_cell") |> filter(mdl_seq == !!iucn_mdl_seq) |> count() |> pull(n)
message(glue("ca_nmfs cells: {n_ca}, IUCN cells: {n_iucn}"))
stopifnot(n_ca == n_iucn)

# verify all values are 100
d_vals <- tbl(con_sdm, "model_cell") |>
  filter(mdl_seq == !!mdl_seq) |>
  summarize(v_min = min(value), v_max = max(value)) |>
  collect()
message(glue("value range: {d_vals$v_min} - {d_vals$v_max}"))
stopifnot(d_vals$v_min == 100, d_vals$v_max == 100)
```

```{r disconnect}
dbDisconnect(con_sdm, shutdown = TRUE)
```
