---
title: "Ingest SDMs: NCCOS Atlantic & Pacific Seabird Densities"
subtitle: "Seasonal seabird density (#/km²) → suitability [0,100] on the global 0.05° grid"
msens:
  target_name: ingest_sdm_nc
  workflow_type: ingest
  dependency: [build_cell_grid]
  output: data/manifests/ingest_sdm_nc.json
  dataset: {ds_key: nc, response_type: density, source_authority: NOAA NCCOS, temporal_interval: seasonal, native_format: raster}
editor_options:
  chunk_output_type: console
---

NCCOS at-sea seabird **density** models, Atlantic [@winship2018] + Pacific [@leirness2021], as
seasonal COGs (`n_per_km2`, 2-km, oblique-mercator). v8 rasterizes each (species × season) onto
the global 0.05° cell grid and rescales the heavy-tailed density to a graded **suitability [0,100]**
per species (so seasons/regions share a scale), plus an **annual** mean surface for scoring.

- **Source COGs** (in-repo): `data/sdm/raw/nc_atl_birds_dens/*.tif` (140) + `nc_pac_birds_dens/*.tif`
  (135); band 1 = `n_per_km2` (bands 2/3 = CV/CI). Also public at
  `file.marinesensitivity.org/cog/sdm/raw/{ds}/{sp_code}_{season}.tif` (served natively via titiler).
- **`mdl_key`**: seasonal `nc|{sp_id}|{season}` + annual `nc|{sp_id}` (`sp_id` = `GBIF:…`, or
  `GUILD:{sp_code}` for the multi-species guilds that lack a single taxon id). Regions (atl/pac) are
  spatially disjoint and **unioned** per `sp_id`.

> **Density → [0,100] is a modeling choice**, not a v7-faithful recipe (v7 never scored nc). Here:
> per-`sp_id` linear rescale to the pooled **p99.5** of non-zero density (caps outliers), stored as
> `val_cap` in `model_nc.csv` for reversibility. **NOT yet folded into merge/scoring** — see the note
> at the end (merge treats non-`am` datasets as binary ranges; folding density in is a scoring change
> to validate with `pra_score_delta`).

## Setup

```{r}
#| label: setup

librarian::shelf(arrow, dplyr, fs, glue, here, jsonlite, logger, purrr, readr, sf, terra, tibble,
                 tidyr, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))
source(here("libs/vars.R"))
options(readr.show_col_types = FALSE)
`%||%` <- function(a, b) if (is.null(a) || length(a) == 0 || all(is.na(a))) b else a

ds_key   <- "nc"
dir_dist <- glue("{dir_big_v}/marine-atlas/dist/dataset={ds_key}"); dir_create(dir_dist)
mdl_csv  <- glue("{dir_big_v}/marine-atlas/dist/model_{ds_key}.csv")
manifest <- here("data/manifests/ingest_sdm_nc.json"); dir_create(path_dir(manifest))
cog_base <- "https://file.marinesensitivity.org/cog/sdm/raw"
p_cap    <- 0.995                                        # density percentile used as the 100 point
stopifnot(file_exists(cellid_tif))

if (redo_ingest && dir_exists(dir_dist)) { dir_delete(dir_dist); dir_create(dir_dist) }

# write one model surface as (mdl_key, cell_id, val) parquet. ssn = NA -> annual key/file.
pq_path <- function(safe, ssn) if (is.na(ssn)) glue("{dir_dist}/{ds_key}_{safe}.parquet") else
                                              glue("{dir_dist}/{ds_key}_{safe}_{ssn}.parquet")
write_parquet_row <- function(d, sid, ssn, safe) {
  mk <- if (is.na(ssn)) glue("{ds_key}|{sid}") else glue("{ds_key}|{sid}|{ssn}")
  msens::write_atlas_parquet(                              # V2/zstd/80MB row groups (Phase A)
    tibble(mdl_key = as.character(mk), cell_id = as.integer(d$cell_id), val = d$val),
    pq_path(safe, ssn))
}
```

## Models + species crosswalk

```{r}
#| label: models

# one row per (region, sp_code, season) at band 1 (density)
m <- read_csv(here("data/nc_models.csv")) |>
  filter(var == "n_per_km2") |>
  transmute(region_ds = ds_key, region, sp_code, season, cog_url, bidx)

# sp_code -> sp_id / names, unioned across the two regions' crosswalks; guilds get GUILD:{sp_code}
read_xw <- function(f) {
  if (!file_exists(f)) return(tibble())
  read_csv(f) |> transmute(sp_code, sp_id = as.character(sp_id), sp_common, sp_scientific)
}
xw <- bind_rows(
  read_xw(here("data/sdm/derived/nc_atl_birds_dens/mw_spp.csv")),
  read_xw(here("data/sdm/derived/nc_pac_birds_dens/m_spp.csv")),
  read_xw(here("data/sdm/derived/nc_pac_birds_dens/m_spp-multiple.csv")),
  read_xw(here("data/sdm/derived/nc_atl_birds_dens/mw_spp-multiple.csv"))) |>
  distinct(sp_code, .keep_all = TRUE)

m <- m |> left_join(xw, by = "sp_code") |>
  mutate(
    is_guild = is.na(sp_id) | grepl("/", sp_scientific %||% ""),
    sp_id    = ifelse(is.na(sp_id), glue("GUILD:{sp_code}"), sp_id),
    scientific_name = sp_scientific %||% sp_code)
log_info("nc models: {nrow(m)} (species×season×region); {n_distinct(m$sp_id)} sp_id; guilds: {sum(m$is_guild & !duplicated(m$sp_id))}")
knitr::kable(m |> count(region, season, name = "n") |> pivot_wider(names_from = season, values_from = n))
```

## Rasterize + rescale per species

```{r}
#| label: rasterize

# local source cog for a model row (reproject omerc -> 4326, band 1)
read_cog <- function(region_ds, sp_code, season) {
  f <- here(glue("data/sdm/raw/{region_ds}/{sp_code}_{season}.tif"))
  if (!file_exists(f)) return(NULL)
  r <- rast(f, lyrs = 1)
  terra::project(r, "EPSG:4326", method = "bilinear")
}

# process per sp_id: pool all its (season×region) densities for a shared p99.5 cap, then write one
# seasonal surface per season (regions unioned) + an annual mean. resumable (skip existing parquet).
sp_ids   <- sort(unique(m$sp_id))
reg_rows <- list()
for (sid in sp_ids) {
  mi   <- m |> filter(sp_id == sid)
  safe <- gsub("[^A-Za-z0-9._-]", "_", sid)
  done <- all(file_exists(glue("{dir_dist}/{ds_key}_{safe}_{unique(mi$season)}.parquet"))) &&
          file_exists(glue("{dir_dist}/{ds_key}_{safe}.parquet"))
  if (done && !redo_ingest) next

  # read every source raster for this sp_id, pool non-zero densities -> cap
  rs   <- pmap(list(mi$region_ds, mi$sp_code, mi$season), read_cog)
  keep <- !vapply(rs, is.null, logical(1)); mi <- mi[keep, ]; rs <- rs[keep]
  if (!length(rs)) { log_warn("no source rasters for {sid}"); next }
  vals <- unlist(lapply(rs, function(r) { v <- values(r, mat = FALSE); v[!is.na(v) & v > 0] }))
  cap  <- as.numeric(quantile(vals, p_cap, names = FALSE))
  if (!is.finite(cap) || cap <= 0) cap <- max(vals, na.rm = TRUE)

  # per season: union regions' cells, rescale density->[0,100] by the shared cap
  cells_season <- list()
  for (ssn in unique(mi$season)) {
    idx <- which(mi$season == ssn)
    d <- map_dfr(idx, function(j) {
      r <- rs[[j]]; r <- terra::clamp(r / cap * 100, upper = 100, values = TRUE)
      msens::cells_from_raster(r, cellid_tif, min_value = 1)
    })
    if (!nrow(d)) next
    d <- d |> group_by(cell_id) |> summarise(val = max(val), .groups = "drop")   # union regions (disjoint)
    write_parquet_row(d, sid, ssn, safe)
    cells_season[[ssn]] <- d
  }
  # annual = mean across the available seasons (per cell)
  if (length(cells_season)) {
    da <- bind_rows(cells_season, .id = "season") |>
      group_by(cell_id) |> summarise(val = round(mean(val), 2), .groups = "drop")
    write_parquet_row(da, sid, NA, safe)
  }
  reg_rows[[sid]] <- mi |> distinct(sp_id, sp_code, scientific_name, is_guild) |>
    slice(1) |> mutate(val_cap = round(cap, 4),
                       seasons = paste(sort(unique(mi$season)), collapse = "|"),
                       regions = paste(sort(unique(mi$region)), collapse = "|"))
}
```

## Model registry (`model_nc.csv`)

```{r}
#| label: model-csv

# resumable: a re-render with every surface already on disk leaves reg_rows empty (the rasterize
# loop `next`s), so keep the existing registry rather than rebuilding it from nothing.
if (length(reg_rows) == 0 && file_exists(mdl_csv)) {
  model_nc <- read_csv(mdl_csv)
} else {
  reg <- bind_rows(reg_rows)
  # expand to one mdl_key per (seasonal + annual) surface actually written
  model_nc <- reg |> rowwise() |> do({
    r <- .; ssns <- strsplit(r$seasons, "\\|")[[1]]
    tibble(
      mdl_key = c(glue("{ds_key}|{r$sp_id}|{ssns}"), glue("{ds_key}|{r$sp_id}")),
      interval = c(ssns, "annual"),
      sp_id = r$sp_id, sp_code = r$sp_code, scientific_name = r$scientific_name,
      is_guild = r$is_guild, val_cap = r$val_cap, regions = r$regions,
      cog_url = c(glue("{cog_base}/nc_{ifelse(r$regions=='Atlantic','atl','pac')}_birds_dens/{r$sp_code}_{ssns}.tif"), NA))
  }) |> ungroup() |>
    rowwise() |>
    filter(file_exists(glue("{dir_dist}/{ds_key}_{gsub('[^A-Za-z0-9._-]','_',sp_id)}{ifelse(interval=='annual','', paste0('_',interval))}.parquet"))) |>
    ungroup()
  write_csv(model_nc, mdl_csv)
}
log_info("model_nc.csv: {nrow(model_nc)} mdl_keys ({sum(model_nc$interval=='annual')} annual)")
knitr::kable(model_nc |> count(interval, name = "n"))
```

## Manifest

```{r}
#| label: manifest

# content-addressed, deterministic manifest (Phase A) — no wall-clock, so an identical re-ingest
# produces byte-identical bytes and does NOT invalidate downstream targets.
h    <- msens::hash_parquet(glue("{dir_dist}/*.parquet"))
smry <- arrow::open_dataset(dir_dist) |>
  summarise(n_rows = n(), n_mdl = n_distinct(mdl_key),
            val_min = round(min(val), 2), val_max = round(max(val), 2)) |> collect()
msens::report_table(smry, caption = "nc density model_cell surface (density → suitability [0,100])")

msens::write_manifest(
  manifest, target = "ingest_sdm_nc", content_hash = h,
  stats = list(
    ds_key = ds_key, n_sp_id = n_distinct(model_nc$sp_id), n_mdl = nrow(model_nc), p_cap = p_cap,
    note = "seasonal + annual density->[0,100]; NOT yet folded into merge/scoring"),
  force = msens::force_target("ingest_sdm_nc"))
```
