---
title: "Ingest USFWS species ranges → global 0.05° cells (v7-matched + marine-relevance)"
msens:
  target_name: ingest_rng_fws
  workflow_type: ingest
  dependency: [build_cell_grid]
  output: data/manifests/ingest_rng_fws.json
  dataset: {ds_key: rng_fws, response_type: range, source_authority: USFWS, temporal_interval: static, native_format: vector}
editor_options:
  chunk_output_type: console
---

USFWS complete-species current ranges onto the **global 0.05° cell grid**. FWS is a
**terrestrial-heavy** agency (2,205 species: bats, bears, plants…), so only the
coastal/marine species belong in a marine-sensitivity model. Handled in two phases:

1. **Exact v7-species match** — ingest exactly the **106** FWS species v7 kept (its
   ocean mask's marine-relevant subset), for v7↔v8 comparability
   (`data/v7_rng_fws_spcodes.csv`).
2. **Marine-relevance analysis** — compute `pct_marine` (area-weighted ocean
   fraction) for **all** FWS species; compare the v7 species' distribution (esp.
   `min`) against the non-v7 species to surface any marine-relevant species v7
   **missed**. Reported for review (not auto-ingested), so expanding the set is a
   deliberate, documented choice.

**Value = extinction-risk score** via `msens::compute_er_score("FWS:{code}")` (never
hard-coded): `EN` = 100, `TN` = 50, `LC` = 1, from the species' most-protective
(precautionary) ESA status. Whole range, no land mask (`pct_marine` is the metric).

## Setup

```{r}
#| label: setup

librarian::shelf(
  arrow, DBI, dplyr, duckdb, exactextractr, fs, glue, here, janitor, jsonlite,
  logger, readr, sf, terra, tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))   # dir_raw, dir_big_v, cellid_tif, bio_oracle_tif, ver
source(here("libs/vars.R"))    # redo_ingest
options(readr.show_col_types = F)
sf_use_s2(FALSE)               # FWS range polygons are planar-valid, s2-invalid

ds_key   <- "rng_fws"
dir_rng  <- glue("{dir_raw}/fws.gov/usfws_complete_species_current_range")
rng_shp  <- glue("{dir_rng}/usfws_complete_species_current_range_{1:2}.shp")
dir_dist <- glue("{dir_big_v}/marine-atlas/dist/dataset={ds_key}")
manifest <- here("data/manifests/ingest_rng_fws.json")
pctm_csv <- here("data/ingest_rng_fws_pct_marine.csv")
dir_create(c(dir_dist, path_dir(manifest)))
stopifnot("run build_cell_grid.qmd first" = file_exists(cellid_tif), all(file_exists(rng_shp)))

v7_spcodes <- read_csv(here("data/v7_rng_fws_spcodes.csv"))$spcode   # 106 v7 FWS species

# FWS STATUS_ABB -> ESA code (the v7 mapping); val via compute_er_score (never hard-coded)
esa_from_status <- function(status_abb) dplyr::case_when(
  status_abb %in% c("Endangered", "Experimental Population, Non-Essential",
                    "Proposed Endangered") ~ "EN",
  status_abb %in% c("Threatened", "Proposed Threatened",
                    "Similarity of Appearance (Threatened)",
                    "Proposed Similarity of Appearance (Threatened)") ~ "TN",
  TRUE ~ "LC")   # Not Listed / Recovery / Species of Concern / Under Review / etc.
```

## Read all FWS ranges + per-species ESA status + er_score

```{r}
#| label: read

rng <- bind_rows(lapply(rng_shp, st_read, quiet = TRUE)) |>
  clean_names() |>
  filter(status_abb != "Extinction") |>            # drop extinct
  mutate(esa_code = ordered(esa_from_status(status_abb), c("LC", "TN", "EN")))

# one model per species (spcode); most-protective (precautionary) status
sp <- rng |>
  st_drop_geometry() |>
  group_by(spcode, sciname, comname, family) |>
  summarise(n_features = n(), esa_code = max(esa_code), .groups = "drop") |>
  mutate(
    er_code  = paste0("FWS:", as.character(esa_code)),
    er_score = compute_er_score(er_code),
    is_v7    = spcode %in% v7_spcodes,
    mdl_key  = mdl_key_raw(ds_key, spcode)) |>
  arrange(sciname)
write_csv(sp, glue("{dir_dist}/../model_{ds_key}.csv"))
glue("{nrow(sp)} FWS species | {sum(sp$is_v7)} in v7 | ER: {sum(sp$esa_code=='EN')} EN / {sum(sp$esa_code=='TN')} TN / {sum(sp$esa_code=='LC')} LC")
```

## Phase 2 — marine-relevance analysis (pct_marine)

`pct_marine` = area-weighted ocean fraction of each range (exactextractr over the
Bio-Oracle ocean mask). Heavy (all species); cached to `pctm_csv`.

```{r}
#| label: pct_marine

if (redo_ingest || !file_exists(pctm_csv)) {
  r_ocean <- !is.na(rast(bio_oracle_tif, lyrs = "depth_mean"))   # 1 ocean, 0 land
  log_info("computing pct_marine for {nrow(sp)} FWS species ...")
  pm <- exact_extract(r_ocean, rng, "mean", append_cols = "spcode", progress = FALSE) |>
    group_by(spcode) |> summarise(pct_marine = round(100 * max(mean, na.rm = TRUE), 1), .groups = "drop")
  sp <- sp |> left_join(pm, by = "spcode")
  write_csv(sp, pctm_csv)
} else {
  sp <- read_csv(pctm_csv)
}

# v7 threshold vs non-v7 candidates v7 may have missed
thr    <- min(sp$pct_marine[sp$is_v7], na.rm = TRUE)
missed <- sp |> filter(!is_v7, pct_marine >= thr) |> arrange(desc(pct_marine))
log_info("v7 pct_marine: min={thr}% median={median(sp$pct_marine[sp$is_v7],na.rm=T)}%; ",
         "{nrow(missed)} non-v7 species >= v7-min threshold (candidates v7 missed)")
knitr::kable(t(summary(sp$pct_marine[sp$is_v7])))
knitr::kable(head(missed |> select(spcode, sciname, comname, esa_code, pct_marine), 20))
```

## Phase 1 — ingest the v7-matched species → one Parquet each

Ingest exactly the v7 species (comparability). To expand to the Phase-2 candidates,
add them to `to_ingest` after review.

```{r}
#| label: interpolate

if (redo_ingest && dir_exists(dir_dist)) { dir_delete(dir_dist); dir_create(dir_dist) }
to_ingest <- sp |> filter(is_v7)                       # v7 exact match (deliberate)
done <- path_ext_remove(path_file(dir_ls(dir_dist, glob = "*.parquet")))
todo <- to_ingest |> filter(!spcode %in% done)
log_info("{ds_key}: {length(done)} done; rasterizing {nrow(todo)} of {nrow(to_ingest)} v7-matched ranges ...")

for (i in seq_len(nrow(todo))) {
  r      <- todo[i, ]
  ply_sp <- rng |> filter(spcode == r$spcode) |> summarise(.groups = "drop")
  d <- cells_from_ranges(ply_sp, cellid_tif, value = r$er_score)   # whole range, ER-scored
  if (nrow(d) == 0) { log_warn("  {r$spcode} ({r$sciname}): no cells"); next }
  write_parquet(tibble(mdl_key = r$mdl_key, cell_id = d$cell_id, val = d$val),
                fs::path(dir_dist, r$spcode, ext = "parquet"))
}
log_info("{ds_key}: done")
```

## Verify + manifest

```{r}
#| label: verify

pq  <- dir_ls(dir_dist, glob = "*.parquet")
con <- dbConnect(duckdb())
smry <- dbGetQuery(con, glue(
  "SELECT count(DISTINCT mdl_key) n_models, count(*) n_cells, min(val) v_min, max(val) v_max
     FROM read_parquet('{dir_dist}/*.parquet')"))
dbDisconnect(con, shutdown = TRUE)
knitr::kable(smry)

write_json(
  list(target = "ingest_rng_fws", ds_key = ds_key, ver = ver,
       n_species_all = nrow(sp), n_v7 = sum(sp$is_v7), n_parquet = length(pq),
       dir_dist = dir_dist, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
```
