---
title: "USGS Gap Analysis Project Species Range Maps CONUS_2001"
subtitle: "Ingest and map sea turtle range shapefiles from USGS GAP"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Overview

Dataset:

- [U.S. Geological Survey - Gap Analysis Project Species Range Maps CONUS\_2001 | U.S. Geological Survey](https://www.usgs.gov/data/us-geological-survey-gap-analysis-project-species-range-maps-conus2001)
- [U.S. Geological Survey - Gap Analysis Project Species Range Maps CONUS\_2001 - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/item/5951527de4b062508e3b1e79)

Species shapefiles:

- [sea turtle - ScienceBase - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/items/get?q=sea%20turtle&max=20&parentId=5951527de4b062508e3b1e79)
- [Leatherback Sea Turtle (Dermochelys coriacea) rLEATx\_CONUS\_2001v1 Range Map - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/item/59f5ec32e4b063d5d307e4f5)
- [Loggerhead Sea Turtle (Caretta caretta) rLOGTx\_CONUS\_2001v1 Range Map - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/item/59f5e200e4b063d5d307dc0b)
- [Green Sea Turtle (Chelonia mydas) rGSTUx\_CONUS\_2001v1 Range Map - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/item/59f5eba8e4b063d5d307e069)
- [Olive Ridley Sea Turtle (Lepidochelys olivacea) rORSTx\_CONUS\_2001v1 Range Map - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/item/59f5ebabe4b063d5d307e081)
- [Kemp's Ridley Sea Turtle (Lepidochelys kempii) rKRSTx\_CONUS\_2001v1 Range Map - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/item/59f5ec0fe4b063d5d307e3c1)
- [Hawksbill Sea Turtle (Eretmochelys imbricata) rHSTUx\_CONUS\_2001v1 Range Map - ScienceBase-Catalog](https://www.sciencebase.gov/catalog/item/59f5ec2ae4b063d5d307e4bd)

## Setup

```{r}
#| label: setup
librarian::shelf(
  dplyr,
  glue,
  here,
  mapgl,
  purrr,
  sf,
  tibble,
  quiet = T
)

source(here("libs/paths.R"))
```

## Load Shapefiles

```{r}
#| label: load_shapefiles

# sea turtle species codes and shapefile subdirectories
gap_spp <- tribble(
  ~code    , ~short_name     , ~common_name               , ~sci_name                ,
  "rGSTUx" , "Green"         , "Green Sea Turtle"         , "Chelonia mydas"         ,
  "rHSTUx" , "Hawksbill"     , "Hawksbill Sea Turtle"     , "Eretmochelys imbricata" ,
  "rKRSTx" , "Kemp's Ridley" , "Kemp's Ridley Sea Turtle" , "Lepidochelys kempii"    ,
  "rLEATx" , "Leatherback"   , "Leatherback Sea Turtle"   , "Dermochelys coriacea"   ,
  "rLOGTx" , "Loggerhead"    , "Loggerhead Sea Turtle"    , "Caretta caretta"        ,
  "rORSTx" , "Olive Ridley"  , "Olive Ridley Sea Turtle"  , "Lepidochelys olivacea"
)

dir_gap <- glue("{dir_raw}/usgs.gov")

# read all shapefiles and bind into one sf object
sf_gap <- gap_spp |>
  purrr::pmap_dfr(\(code, common_name, sci_name, ...) {
    shp_dir <- glue("{dir_gap}/{code}_CONUS_Range_2001v1")
    shp_file <- glue("{shp_dir}/{code}_CONUS_Range_2001v1.shp")
    st_read(shp_file, quiet = TRUE) |>
      st_transform(4326) |>
      mutate(
        common_name = common_name,
        sci_name = sci_name
      )
  })

message(glue(
  "Loaded {nrow(sf_gap)} features across ",
  "{length(unique(sf_gap$sci_name))} species"
))

# dissolve by species for a cleaner overview map
sf_gap_dissolved <- sf_gap |>
  group_by(sci_name, common_name) |>
  summarise(.groups = "drop")

# assign a color per species
gap_colors <- c(
  "Chelonia mydas" = "#1b9e77",
  "Eretmochelys imbricata" = "#d95f02",
  "Lepidochelys kempii" = "#7570b3",
  "Dermochelys coriacea" = "#e7298a",
  "Caretta caretta" = "#66a61e",
  "Lepidochelys olivacea" = "#e6ab02"
)
```

## Maps

::: {.panel-tabset}

### All Species

```{r}
#| label: map_all

# named colors keyed by common name for legend
gap_colors_common <- setNames(
  unname(gap_colors),
  gap_spp$common_name[match(names(gap_colors), gap_spp$sci_name)]
)

bbox <- st_bbox(sf_gap_dissolved) |> as.numeric()

# one layer per species with common name as layer id for toggle control
m_gap <- maplibre(style = maptiler_style("bright", variant = "dark")) |>
  fit_bounds(bbox, animate = FALSE)

for (j in seq_len(nrow(gap_spp))) {
  sp_j <- gap_spp$sci_name[j]
  sp_lbl <- gap_spp$common_name[j]
  sf_j <- sf_gap_dissolved |> filter(sci_name == sp_j)
  m_gap <- m_gap |>
    add_fill_layer(
      id = sp_lbl,
      source = sf_j,
      fill_color = unname(gap_colors[sp_j]),
      fill_opacity = 0.5,
      popup = glue("<b>{sp_lbl}</b><br><i>{sp_j}</i>")
    )
}

m_gap |>
  add_layers_control(
    position = "top-left",
    collapsible = TRUE
  ) |>
  mapgl::add_legend(
    legend_title = "USGS GAP Range",
    values = names(gap_colors_common),
    colors = unname(gap_colors_common),
    type = "categorical",
    position = "bottom-right"
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```

### Green

```{r}
#| label: map_green
sf_sp <- sf_gap_dissolved |> filter(sci_name == "Chelonia mydas")
bbox_sp <- st_bbox(sf_sp) |> as.numeric()

maplibre(style = maptiler_style("bright", variant = "dark")) |>
  fit_bounds(bbox_sp, animate = FALSE) |>
  add_fill_layer(
    id = "green",
    source = sf_sp,
    fill_color = unname(gap_colors["Chelonia mydas"]),
    fill_opacity = 0.5,
    popup = "<b>Green Sea Turtle</b><br><i>Chelonia mydas</i>"
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```

### Hawksbill

```{r}
#| label: map_hawksbill
sf_sp <- sf_gap_dissolved |> filter(sci_name == "Eretmochelys imbricata")
bbox_sp <- st_bbox(sf_sp) |> as.numeric()

maplibre(style = maptiler_style("bright", variant = "dark")) |>
  fit_bounds(bbox_sp, animate = FALSE) |>
  add_fill_layer(
    id = "hawksbill",
    source = sf_sp,
    fill_color = unname(gap_colors["Eretmochelys imbricata"]),
    fill_opacity = 0.5,
    popup = "<b>Hawksbill Sea Turtle</b><br><i>Eretmochelys imbricata</i>"
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```

### Kemp's Ridley

```{r}
#| label: map_kemps
sf_sp <- sf_gap_dissolved |> filter(sci_name == "Lepidochelys kempii")
bbox_sp <- st_bbox(sf_sp) |> as.numeric()

maplibre(style = maptiler_style("bright", variant = "dark")) |>
  fit_bounds(bbox_sp, animate = FALSE) |>
  add_fill_layer(
    id = "kemps",
    source = sf_sp,
    fill_color = unname(gap_colors["Lepidochelys kempii"]),
    fill_opacity = 0.5,
    popup = "<b>Kemp's Ridley Sea Turtle</b><br><i>Lepidochelys kempii</i>"
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```

### Leatherback

```{r}
#| label: map_leatherback
sf_sp <- sf_gap_dissolved |> filter(sci_name == "Dermochelys coriacea")
bbox_sp <- st_bbox(sf_sp) |> as.numeric()

maplibre(style = maptiler_style("bright", variant = "dark")) |>
  fit_bounds(bbox_sp, animate = FALSE) |>
  add_fill_layer(
    id = "leatherback",
    source = sf_sp,
    fill_color = unname(gap_colors["Dermochelys coriacea"]),
    fill_opacity = 0.5,
    popup = "<b>Leatherback Sea Turtle</b><br><i>Dermochelys coriacea</i>"
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```

### Loggerhead

```{r}
#| label: map_loggerhead
sf_sp <- sf_gap_dissolved |> filter(sci_name == "Caretta caretta")
bbox_sp <- st_bbox(sf_sp) |> as.numeric()

maplibre(style = maptiler_style("bright", variant = "dark")) |>
  fit_bounds(bbox_sp, animate = FALSE) |>
  add_fill_layer(
    id = "loggerhead",
    source = sf_sp,
    fill_color = unname(gap_colors["Caretta caretta"]),
    fill_opacity = 0.5,
    popup = "<b>Loggerhead Sea Turtle</b><br><i>Caretta caretta</i>"
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```

### Olive Ridley

```{r}
#| label: map_olive
sf_sp <- sf_gap_dissolved |> filter(sci_name == "Lepidochelys olivacea")
bbox_sp <- st_bbox(sf_sp) |> as.numeric()

maplibre(style = maptiler_style("bright", variant = "dark")) |>
  fit_bounds(bbox_sp, animate = FALSE) |>
  add_fill_layer(
    id = "olive_ridley",
    source = sf_sp,
    fill_color = unname(gap_colors["Lepidochelys olivacea"]),
    fill_opacity = 0.5,
    popup = "<b>Olive Ridley Sea Turtle</b><br><i>Lepidochelys olivacea</i>"
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```

:::
