---
title: "Investigate Sea Turtle Model Outputs"
subtitle: "Diagnose spatial issues in turtle SDM/range models flagged by BOEM reviewers"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Overview

BOEM team (Tim White, Gwendolyn Gallagher) raised concerns about sea turtle
model outputs in v3:

1. **Leatherback** (*Dermochelys coriacea*, mdl_seq 27735): IUCN range map
   doesn't match the masked model — possible **dateline/antimeridian issue**
   since the species ranges across the Pacific.
2. **Hawksbill** (*Eretmochelys imbricata*, mdl_seq 27734): NOAA Fisheries
   range is more restricted than IUCN — Hawksbill likely only occurs in
   Southern California, not further north. **IUCN (broad) vs NMFS (narrow)**.
3. **Olive Ridley** (*Lepidochelys olivacea*, mdl_seq 27749): Model shows
   presence **north of the Aleutian Islands / Bering Sea**, unsupported by
   literature.
4. **General**: Any sea turtle presence north of the Aleutians needs
   literature justification or model correction.

```{r}
#| label: setup
librarian::shelf(
  DBI,
  dplyr,
  duckdb,
  DT,
  glue,
  here,
  purrr,
  RColorBrewer,
  sf,
  terra,
  tibble,
  quiet = T
)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE
)
terraOptions(progress = 0)

source(here("libs/paths.R"))

cell_tif <- glue("{dir_derived}/r_bio-oracle_planarea.tif")
r_cell <- rast(cell_tif, lyrs = "cell_id")

# iucn paths
dir_iucn_raw <- glue("{dir_raw}/iucnredlist.org")
dir_iucn_derived <- glue("{dir_big}/iucnredlist.org")

librarian::shelf(mapgl, quiet = T)

# database connection
con_sdm <- dbConnect(duckdb(dbdir = sdm_db, read_only = TRUE))

# sea turtle species
sci_turtles <- c(
  "Chelonia mydas", # green
  "Caretta caretta", # loggerhead
  "Eretmochelys imbricata", # hawksbill
  "Lepidochelys kempii", # Kemp's ridley
  "Lepidochelys olivacea", # olive ridley
  "Natator depressus", # flatback
  "Dermochelys coriacea"
) # leatherback

# shared color palette and fixed range (matches apps/mapsp)
n_cols <- 11
cols_spectral <- rev(brewer.pal(n_cols, "Spectral"))
FIXED_RANGE <- c(1, 100)

# helper: convert model_cell rows to a trimmed raster (0-360 coords, no rotate)
get_rast <- function(mdl_seq) {
  d <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq == !!mdl_seq) |>
    select(cell_id, value) |>
    collect()
  r <- init(r_cell[[1]], NA)
  r[d$cell_id] <- d$value
  names(r) <- "value"
  terra::trim(r)
}

# helper: add raster with fixed color range (from apps/mapsp)
add_fixed_range_raster <- function(
  map,
  data,
  id,
  fixed_range = FIXED_RANGE,
  colors = cols_spectral,
  ...
) {
  dr <- terra::minmax(data) |> as.numeric()
  # clamp to fixed range
  dr[1] <- max(dr[1], fixed_range[1])
  dr[2] <- min(dr[2], fixed_range[2])
  n_clrs <- diff(fixed_range) + 1
  clrs <- colorRampPalette(colors)(n_clrs)
  clrs_dr <- clrs[seq.int(dr[1], dr[2])]

  map |>
    add_image_source(
      id = paste0(id, "_src"),
      data = terra::clamp(
        data,
        lower = fixed_range[1],
        upper = fixed_range[2],
        values = TRUE
      ),
      colors = clrs_dr
    ) |>
    add_raster_layer(
      id = id,
      source = paste0(id, "_src"),
      ...
    )
}

# helper: convert 0-360 extent bbox to -180/180 for fit_bounds
bbox_0360_to_180 <- function(r) {
  e <- as.vector(ext(r)) # xmin, xmax, ymin, ymax
  c(
    ifelse(e[1] > 180, e[1] - 360, e[1]), # xmin
    e[3], # ymin
    ifelse(e[2] > 180, e[2] - 360, e[2]), # xmax
    e[4] # ymax
  )
}

# helper: build a mapgl raster map (rasters in 0-360 coords)
map_rast <- function(r, title = "value") {
  bbox <- bbox_0360_to_180(r)

  maplibre(
    style = maptiler_style("bright", variant = "dark")
  ) |>
    fit_bounds(bbox, animate = FALSE) |>
    add_fixed_range_raster(
      data = r,
      id = "r_lyr",
      raster_opacity = 0.7,
      raster_resampling = "nearest"
    ) |>
    mapgl::add_legend(
      legend_title = title,
      values = FIXED_RANGE,
      colors = cols_spectral,
      position = "bottom-right"
    ) |>
    add_fullscreen_control() |>
    add_navigation_control() |>
    add_scale_control()
}

# helper: build a mapgl polygon map
map_poly <- function(sf_data, title = "IUCN range") {
  bbox <- st_bbox(sf_data) |> as.numeric()

  maplibre(
    style = maptiler_style("bright", variant = "dark")
  ) |>
    fit_bounds(bbox, animate = FALSE) |>
    add_fill_layer(
      id = "range_lyr",
      source = sf_data,
      fill_color = "red",
      fill_opacity = 0.5,
      popup = "sci_name"
    ) |>
    add_fullscreen_control() |>
    add_navigation_control() |>
    add_scale_control()
}
```

## Query All Turtle Taxa & Models in Database

```{r}
#| label: turtle_models
# get turtle taxon records
d_turtles <- tbl(con_sdm, "taxon") |>
  filter(sp_cat == "turtle") |>
  collect()

# get all models per turtle species (taxon_model already has ds_key)
d_models <- tbl(con_sdm, "taxon_model") |>
  filter(taxon_id %in% !!d_turtles$taxon_id) |>
  collect() |>
  left_join(
    d_turtles |> select(taxon_id, scientific_name, common_name),
    by = "taxon_id"
  )

# cell counts per model
d_counts <- tbl(con_sdm, "model_cell") |>
  filter(mdl_seq %in% !!d_models$mdl_seq) |>
  group_by(mdl_seq) |>
  summarize(
    n_cells = n(),
    v_min = min(value, na.rm = TRUE),
    v_max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  collect()

d_summary <- d_models |>
  left_join(d_counts, by = "mdl_seq") |>
  select(
    scientific_name,
    common_name,
    ds_key,
    mdl_seq,
    n_cells,
    v_min,
    v_max
  ) |>
  arrange(scientific_name, ds_key)

datatable(
  d_summary,
  caption = "Sea turtle models in sdm.duckdb",
  filter = "top",
  options = list(dom = "ft")
)
```

## Load IUCN Range Polygons

```{r}
#| label: load_iucn_polys_tifs
# turtle shapefiles
turtle_shp <- glue("{dir_iucn_raw}/TURTLES/TURTLES.shp")
if (!file.exists(turtle_shp)) {
  turtle_shp <- Sys.glob(glue("{dir_iucn_raw}/TURTLES*/*.shp"))[1]
}

sf_turtles_iucn <- NULL
if (!is.na(turtle_shp) && file.exists(turtle_shp)) {
  sf_rep <- st_read(turtle_shp, quiet = TRUE)
  sf_turtles_iucn <- sf_rep |>
    filter(sci_name %in% sci_turtles)
  message(glue(
    "Loaded {nrow(sf_turtles_iucn)} IUCN features for ",
    "{length(unique(sf_turtles_iucn$sci_name))} turtle species"
  ))
} else {
  message(glue("IUCN TURTLES shapefile not found at {turtle_shp}"))
}

# check which species have derived tifs
tif_paths <- tibble(
  scientific_name = sci_turtles
) |>
  mutate(
    tif = glue("{dir_iucn_derived}/TURTLES/{scientific_name}.tif"),
    exists = file.exists(tif)
  )

datatable(
  tif_paths,
  caption = "Derived IUCN raster TIFs per turtle species",
  filter = "top",
  options = list(dom = "ft")
)
```

## Leatherback (*Dermochelys coriacea*) {#sec-leatherback}

::: {.panel-tabset}

### IUCN Polygon

```{r}
#| label: leatherback_polygon
sp_sci <- "Dermochelys coriacea"
sp_lbl <- "Leatherback"

if (!is.null(sf_turtles_iucn)) {
  sf_sp <- sf_turtles_iucn |> filter(sci_name == sp_sci)
  if (nrow(sf_sp) > 0) {
    map_poly(sf_sp, glue("{sp_lbl} IUCN range"))
  } else {
    message(glue("No IUCN polygons for {sp_sci}"))
  }
}
```

### IUCN Rasterized vs Merged

```{r}
#| label: leatherback_iucn_vs_merged
d_sp <- d_summary |> filter(scientific_name == sp_sci)

mdl_iucn <- d_sp |> filter(ds_key == "rng_iucn") |> pull(mdl_seq)
mdl_merge <- d_sp |> filter(ds_key == "ms_merge") |> pull(mdl_seq)

if (length(mdl_iucn) > 0 && length(mdl_merge) > 0) {
  r_iucn <- get_rast(mdl_iucn)
  r_merge <- get_rast(mdl_merge)

  m1 <- map_rast(r_iucn, glue("{sp_lbl} IUCN rasterized"))
  m2 <- map_rast(r_merge, glue("{sp_lbl} merged model"))
  mapgl::compare(m1, m2) |> suppressMessages()
}
```

### AquaMaps vs Merged

```{r}
#| label: leatherback_am_vs_merged
mdl_am <- d_sp |> filter(ds_key == "am_0.05") |> pull(mdl_seq)

if (length(mdl_am) > 0 && length(mdl_merge) > 0) {
  r_am <- get_rast(mdl_am)
  r_merge <- get_rast(mdl_merge)

  m1 <- map_rast(r_am, glue("{sp_lbl} AquaMaps"))
  m2 <- map_rast(r_merge, glue("{sp_lbl} merged model"))
  mapgl::compare(m1, m2)
} else {
  message(glue("Missing am or ms_merge for {sp_sci}"))
}
```

### Dateline Analysis
  
```{r}
#| label: leatherback_dateline
# check cell counts near the antimeridian (±180° longitude)
# r_cell uses 0-360 coords; antimeridian is at x=180
d_iucn <- tbl(con_sdm, "model_cell") |>
  filter(mdl_seq == !!mdl_iucn) |>
  select(cell_id, value) |>
  collect()

# get coordinates of cells
xy <- xyFromCell(r_cell, d_iucn$cell_id)
d_iucn$lon_0360 <- xy[, 1]
d_iucn$lat <- xy[, 2]
d_iucn <- d_iucn |>
  mutate(lon = ifelse(lon_0360 > 180, lon_0360 - 360, lon_0360))

# count cells in longitude bins near dateline
d_dateline <- d_iucn |>
  mutate(
    lon_bin = cut(
      lon,
      breaks = c(-180, -170, -160, 160, 170, 180),
      labels = c("-180:-170", "-170:-160", "160:170", "170:180", "near 180")
    )
  ) |>
  filter(!is.na(lon_bin)) |>
  group_by(lon_bin) |>
  summarize(
    n_cells = n(),
    lat_min = min(lat),
    lat_max = max(lat),
    .groups = "drop"
  )

datatable(
  d_dateline,
  caption = glue("{sp_lbl}: cell counts near antimeridian (±180°)")
)

# show raw IUCN polygon extent if available
if (!is.null(sf_turtles_iucn)) {
  sf_sp <- sf_turtles_iucn |> filter(sci_name == sp_sci)
  bbox <- st_bbox(sf_sp)
  message(glue(
    "IUCN polygon bbox: xmin={round(bbox['xmin'], 1)}, ",
    "xmax={round(bbox['xmax'], 1)}, ",
    "ymin={round(bbox['ymin'], 1)}, ",
    "ymax={round(bbox['ymax'], 1)}"
  ))
} else {
  message("No rng_iucn model found for Leatherback")
}
```

:::

## Hawksbill (*Eretmochelys imbricata*) {#sec-hawksbill}

::: {.panel-tabset}

### IUCN Range

```{r}
#| label: hawksbill_polygon
sp_sci <- "Eretmochelys imbricata"
sp_lbl <- "Hawksbill"

if (!is.null(sf_turtles_iucn)) {
  sf_sp <- sf_turtles_iucn |> filter(sci_name == sp_sci)
  if (nrow(sf_sp) > 0) {
    map_poly(sf_sp, glue("{sp_lbl} IUCN range"))
  } else {
    message(glue("No IUCN polygons for {sp_sci}"))
  }
}
```

### IUCN vs FWS Range

```{r}
#| label: hawksbill_iucn_vs_fws
d_sp <- d_summary |> filter(scientific_name == sp_sci)

mdl_iucn <- d_sp |> filter(ds_key == "rng_iucn") |> pull(mdl_seq)
mdl_fws <- d_sp |> filter(ds_key == "rng_fws") |> pull(mdl_seq)

if (length(mdl_iucn) > 0 && length(mdl_fws) > 0) {
  r_iucn <- get_rast(mdl_iucn)
  r_fws <- get_rast(mdl_fws)

  m1 <- map_rast(r_iucn, glue("{sp_lbl} IUCN"))
  m2 <- map_rast(r_fws, glue("{sp_lbl} FWS range"))
  mapgl::compare(m1, m2)
} else {
  message(glue(
    "Missing rng_iucn ({length(mdl_iucn)}) or rng_fws ({length(mdl_fws)}) ",
    "for {sp_sci}. Available ds_keys: {paste(d_sp$ds_key, collapse=', ')}"
  ))
}
```

### AquaMaps vs Merged

```{r}
#| label: hawksbill_am_vs_merged
mdl_am <- d_sp |> filter(ds_key == "am_0.05") |> pull(mdl_seq)
mdl_merge <- d_sp |> filter(ds_key == "ms_merge") |> pull(mdl_seq)

if (length(mdl_am) > 0 && length(mdl_merge) > 0) {
  r_am <- get_rast(mdl_am)
  r_merge <- get_rast(mdl_merge)

  m1 <- map_rast(r_am, glue("{sp_lbl} AquaMaps"))
  m2 <- map_rast(r_merge, glue("{sp_lbl} merged model"))
  mapgl::compare(m1, m2)
} else {
  message(glue("Missing am or ms_merge for {sp_sci}"))
}
```

### Pacific North Extent

```{r}
#| label: hawksbill_north_extent
# compare IUCN vs FWS range in the Pacific, north of Southern California (~34°N)
if (length(mdl_iucn) > 0) {
  d_iucn <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq == !!mdl_iucn) |>
    select(cell_id, value) |>
    collect()

  xy <- xyFromCell(r_cell, d_iucn$cell_id)
  d_iucn$lon_0360 <- xy[, 1]
  d_iucn$lat <- xy[, 2]
  d_iucn <- d_iucn |>
    mutate(lon = ifelse(lon_0360 > 180, lon_0360 - 360, lon_0360))

  # cells north of 34°N in the Pacific (lon < -100)
  d_north_pac <- d_iucn |>
    filter(lat > 34, lon < -100) |>
    arrange(desc(lat))

  message(glue(
    "{sp_lbl} IUCN cells north of 34°N in Pacific: {nrow(d_north_pac)}"
  ))

  if (nrow(d_north_pac) > 0) {
    r_north <- init(r_cell[[1]], NA)
    r_north[d_north_pac$cell_id] <- d_north_pac$value
    r_north <- terra::trim(r_north)
    map_rast(r_north, glue("{sp_lbl} IUCN > 34°N Pacific"))
  }

  # compare with FWS if available
  if (length(mdl_fws) > 0) {
    d_fws <- tbl(con_sdm, "model_cell") |>
      filter(mdl_seq == !!mdl_fws) |>
      select(cell_id, value) |>
      collect()
    xy_fws <- xyFromCell(r_cell, d_fws$cell_id)
    d_fws$lat <- xy_fws[, 2]
    d_fws$lon <- ifelse(xy_fws[, 1] > 180, xy_fws[, 1] - 360, xy_fws[, 1])

    d_fws_north <- d_fws |> filter(lat > 34, lon < -100)
    message(glue(
      "{sp_lbl} FWS cells north of 34°N in Pacific: {nrow(d_fws_north)}"
    ))
  }
}
```

:::

## Olive Ridley (*Lepidochelys olivacea*) {#sec-olive-ridley}

::: {.panel-tabset}

### IUCN Range

```{r}
#| label: olive_polygon
sp_sci <- "Lepidochelys olivacea"
sp_lbl <- "Olive Ridley"

if (!is.null(sf_turtles_iucn)) {
  sf_sp <- sf_turtles_iucn |> filter(sci_name == sp_sci)
  if (nrow(sf_sp) > 0) {
    map_poly(sf_sp, glue("{sp_lbl} IUCN range"))
  } else {
    message(glue("No IUCN polygons for {sp_sci}"))
  }
}
```

### AquaMaps vs IUCN

```{r}
#| label: olive_am_vs_iucn
d_sp <- d_summary |> filter(scientific_name == sp_sci)

mdl_am <- d_sp |> filter(ds_key == "am_0.05") |> pull(mdl_seq)
mdl_iucn <- d_sp |> filter(ds_key == "rng_iucn") |> pull(mdl_seq)

if (length(mdl_am) > 0 && length(mdl_iucn) > 0) {
  r_am <- get_rast(mdl_am)
  r_iucn <- get_rast(mdl_iucn)

  m1 <- map_rast(r_am, glue("{sp_lbl} AquaMaps"))
  m2 <- map_rast(r_iucn, glue("{sp_lbl} IUCN rasterized"))
  mapgl::compare(m1, m2)
} else {
  message(glue("Missing am or rng_iucn for {sp_sci}"))
}
```

### Merged Model

```{r}
#| label: olive_merged
mdl_merge <- d_sp |> filter(ds_key == "ms_merge") |> pull(mdl_seq)

if (length(mdl_merge) > 0) {
  r_merge <- get_rast(mdl_merge)
  map_rast(r_merge, glue("{sp_lbl} merged model"))
} else {
  message(glue("No ms_merge model for {sp_sci}"))
}
```

### Bering Sea Cells

```{r}
#| label: olive_bering
# identify cells north of ~52°N (Aleutian Islands)
lat_threshold <- 52

if (length(mdl_merge) > 0) {
  sp_mdl_seqs <- d_sp$mdl_seq

  d_all <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq %in% !!sp_mdl_seqs) |>
    select(mdl_seq, cell_id, value) |>
    collect()

  xy <- xyFromCell(r_cell, d_all$cell_id)
  d_all$lat <- xy[, 2]

  d_bering <- d_all |>
    filter(lat > lat_threshold) |>
    left_join(
      d_sp |> select(mdl_seq, ds_key),
      by = "mdl_seq"
    ) |>
    group_by(ds_key) |>
    summarize(
      n_cells = n_distinct(cell_id),
      lat_min = min(lat),
      lat_max = max(lat),
      .groups = "drop"
    ) |>
    arrange(desc(n_cells))

  datatable(
    d_bering,
    caption = glue("{sp_lbl}: dataset contributions north of {lat_threshold}°N")
  )

  # map cells north of threshold
  d_north <- d_all |>
    filter(lat > lat_threshold, mdl_seq == mdl_merge)
  if (nrow(d_north) > 0) {
    r_north <- init(r_cell[[1]], NA)
    r_north[d_north$cell_id] <- d_north$value
    r_north <- terra::trim(r_north)
    map_rast(r_north, glue("{sp_lbl} > {lat_threshold}°N"))
  }
}
```

:::

## All Turtles — North of Aleutians Check

```{r}
#| label: all_turtles_north
lat_threshold <- 52

# for each turtle species, count cells above threshold per dataset
d_north_all <- map_dfr(d_turtles$taxon_id, \(tid) {
  sp_name <- d_turtles |>
    filter(taxon_id == tid) |>
    pull(scientific_name)

  sp_models <- d_models |>
    filter(taxon_id == tid)

  if (nrow(sp_models) == 0) {
    return(tibble())
  }

  d_mc <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq %in% !!sp_models$mdl_seq) |>
    select(mdl_seq, cell_id, value) |>
    collect()

  if (nrow(d_mc) == 0) {
    return(tibble())
  }

  xy <- xyFromCell(r_cell, d_mc$cell_id)
  d_mc$lat <- xy[, 2]

  d_mc |>
    filter(lat > lat_threshold) |>
    left_join(
      sp_models |> select(mdl_seq, ds_key),
      by = "mdl_seq"
    ) |>
    group_by(ds_key) |>
    summarize(
      n_cells = n_distinct(cell_id),
      lat_max = max(lat),
      .groups = "drop"
    ) |>
    mutate(scientific_name = sp_name)
})

if (nrow(d_north_all) > 0) {
  d_north_summary <- d_north_all |>
    select(scientific_name, ds_key, n_cells, lat_max) |>
    arrange(scientific_name, desc(n_cells))

  datatable(
    d_north_summary,
    caption = glue(
      "Turtle cells north of {lat_threshold}°N by species and dataset"
    )
  )
} else {
  message(glue("No turtle cells found north of {lat_threshold}°N"))
}
```

## Summary & Recommendations

### Findings

- **Leatherback**: Check whether the IUCN polygon crosses the antimeridian
  and if the `rasterize() → rotate() → crop() → mask()` pipeline in
  `merge_models_prep.qmd` (lines 2930-2942) drops cells at ±180° longitude.
  The dateline analysis table above shows cell counts near the antimeridian.

- **Hawksbill**: Compare the IUCN range extent in the Pacific vs FWS/NMFS
  range. If IUCN extends significantly north of Southern California (~34°N)
  in the Pacific while FWS does not, consider using FWS range as an
  additional spatial constraint or manually clipping the IUCN range.

- **Olive Ridley**: The Bering Sea cells table above shows which dataset(s)
  contribute cells north of the Aleutian Islands (~52°N). If AquaMaps
  extends beyond the IUCN mask, verify the mask is correctly applied in
  `merge_models.qmd` (lines 275-324).

### Proposed Solutions

1. **Leatherback dateline fix**: If cells are missing near ±180°, the IUCN
   polygon likely needs to be split at the antimeridian before rasterization
   (use `sf::st_wrap_dateline()` before converting to `terra::vect()`).

2. **Hawksbill range constraint**: Consider using the intersection of IUCN
   and FWS ranges in the Pacific, or manually restrict the mask to reflect
   the known US distribution (Southern California and south).

3. **Olive Ridley Bering Sea**: If the merged model includes cells above
   ~52°N, investigate which source dataset places them there. If AquaMaps
   habitat suitability leaks through, the IUCN mask should exclude those
   cells — verify masking logic is working correctly.

4. **General turtle check**: The "All Turtles — North of Aleutians" table
   above identifies all species and datasets with cells above 52°N. Any
   species showing presence in the Bering Sea should be flagged for review.

## Cleanup

```{r}
#| label: disconnect
dbDisconnect(con_sdm, shutdown = TRUE)
```
