---
title: "Investigate IUCN Range Map Masking"
subtitle: "Impact of IUCN range masking on species removed from US EEZ analysis"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Overview

Two issues were discovered with IUCN range map ingestion in
`merge_models_prep.qmd` that caused spurious AquaMaps SDM predictions to
persist in the US EEZ for species that don't actually occur there:

1. **Missing mammal shapefiles** — 3 mammals (*Phoca vitulina* (harbor seal),
   *Trichechus manatus* (West Indian manatee), *Pusa hispida* (ringed seal))
   were present in the newer
   [IUCN Spatial Download](https://www.iucnredlist.org/resources/spatial-data-download)
   `MAMMALS.zip` but not in the originally downloaded
   `MAMMALS_MARINE_ONLY.zip` & `MAMMALS_MARINE_AND_TERRESTRIAL.zip`. These
   species now have rng_iucn entries with cells in the US EEZ, and their
   merged models are properly clipped.

2. **Species with IUCN ranges outside the US EEZ** — 2,247 species across all
   taxa groups (fish, corals, invertebrates, mammals, etc.) had IUCN range
   polygons that produced 0 cells when rasterized to the US EEZ grid. These
   were previously skipped entirely during ingestion, leaving their AquaMaps
   predictions unmasked (e.g., *Sotalia guianensis* appearing in the Gulf of
   America). Now these species have 0-cell rng_iucn entries in the database,
   so the merge clips AquaMaps to an empty mask → 0 merged cells →
   `is_ok=FALSE`.

This notebook quantifies the impact: how many species were affected, broken
down by component (`sp_cat`), and how many had their merged models truncated.

```{r}
#| label: setup
#| warning: false

librarian::shelf(
  DBI,
  dplyr,
  duckdb,
  DT,
  glue,
  here,
  readr,
  stringr,
  tibble,
  tidyr,
  quiet = T)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

source(here("libs/paths.R"))

con_sdm <- dbConnect(duckdb(dbdir = sdm_db, read_only = TRUE))
```

## 1. Species with 0-cell rng_iucn (range outside US EEZ)

These are species whose IUCN expert range polygon does not intersect the US EEZ
at all. They have `rng_iucn` model entries in the database with 0 rows in
`model_cell`.

```{r}
#| label: zero_cell_iucn

# all rng_iucn models with their cell counts
d_iucn_models <- tbl(con_sdm, "taxon_model") |>
  filter(ds_key == "rng_iucn") |>
  left_join(
    tbl(con_sdm, "model_cell") |>
      group_by(mdl_seq) |>
      summarize(n_cells = n(), .groups = "drop"),
    by = "mdl_seq") |>
  collect() |>
  mutate(n_cells = coalesce(n_cells, 0L))

d_taxon <- tbl(con_sdm, "taxon") |> collect()

d_iucn_spp <- d_iucn_models |>
  left_join(
    d_taxon |>
      select(taxon_id, scientific_name, common_name, sp_cat, is_ok),
    by = "taxon_id")

d_zero_cell <- d_iucn_spp |>
  filter(n_cells == 0)

d_zero_cell |>
  count(sp_cat, name = "n_species") |>
  arrange(desc(n_species)) |>
  datatable(
    caption = glue(
      "{nrow(d_zero_cell)} species with 0-cell rng_iucn ",
      "(IUCN range outside US EEZ), by component"),
    options = list(dom = "t"))
```

## 2. Merged models truncated

Species that previously had a merged model (`ms_merge` with cells) but now
have no merged model because the IUCN mask clipped them to 0 cells. We
identify these as species with 0-cell rng_iucn that are now `is_ok=FALSE`.

```{r}
#| label: truncated

# species with 0-cell rng_iucn that had AquaMaps predictions (am_0.05)
d_am <- tbl(con_sdm, "taxon_model") |>
  filter(ds_key == "am_0.05") |>
  select(taxon_id, am_mdl_seq = mdl_seq) |>
  collect()

d_had_aquamaps <- d_zero_cell |>
  inner_join(d_am, by = "taxon_id")

# of those, how many now have is_ok=FALSE?
d_truncated <- d_had_aquamaps |>
  filter(!is_ok)

d_truncated |>
  count(sp_cat, name = "n_truncated") |>
  arrange(desc(n_truncated)) |>
  datatable(
    caption = glue(
      "{nrow(d_truncated)} species with AquaMaps in US EEZ truncated by ",
      "0-cell IUCN mask (now is_ok=FALSE), by component"),
    options = list(dom = "t"))
```

### Truncated species listing

```{r}
#| label: truncated_listing

app_base <- "https://app.marinesensitivity.org/mapsp"

d_truncated_tbl <- d_truncated |>
  mutate(
    species = glue(
      "<a href='{app_base}/?mdl_seq={am_mdl_seq}' target='_blank'>",
      "<em>{scientific_name}</em></a>")) |>
  select(
    species, common_name, component = sp_cat,
    am_mdl_seq) |>
  arrange(component, common_name)

datatable(
  d_truncated_tbl,
  caption = glue(
    "{nrow(d_truncated_tbl)} species truncated by 0-cell IUCN mask ",
    "(click species for AquaMaps model in app)"),
  escape  = FALSE,
  filter  = "top",
  options = list(
    dom        = "ftp",
    pageLength = 25))
```

### Stale ms_merge check

Verify that truncated species had their old merged models properly deleted.

```{r}
#| label: stale_merge_check

d_stale <- d_truncated |>
  inner_join(
    tbl(con_sdm, "taxon_model") |>
      filter(ds_key == "ms_merge") |>
      select(taxon_id, mdl_seq) |>
      collect(),
    by = "taxon_id",
    suffix = c("_iucn", "_merge")) |>
  inner_join(
    tbl(con_sdm, "model_cell") |>
      group_by(mdl_seq) |>
      summarize(n_merge_cells = n(), .groups = "drop") |>
      collect(),
    by = c("mdl_seq_merge" = "mdl_seq"))

if (nrow(d_stale) > 0) {
  datatable(
    d_stale |>
      select(scientific_name, common_name, sp_cat, n_merge_cells),
    caption = glue(
      "WARNING: {nrow(d_stale)} species still have stale ms_merge cells"),
    options = list(dom = "ft", pageLength = 20))
} else {
  cat(glue(
    "all truncated species have no stale ms_merge entries ",
    "— merge cleanup is complete.\n"))
}
```

## 3. Impact on Program Area scores (v5 → v6)

Despite 371 species changing from `is_ok=TRUE` to `is_ok=FALSE`, the impact on
Program Area component scores is **negligible**. Only Southern California (SOC)
shows a visible change in rounded integer scores. This section explains why.

### Species count changes (v5 → v6)

```{r}
#| label: is_ok_changes

# connect to v5 for comparison
sdm_db_v5 <- gsub(ver, ver_prev, sdm_db)
con_v5 <- dbConnect(duckdb(dbdir = sdm_db_v5, read_only = TRUE))

d_v5 <- tbl(con_v5, "taxon") |>
  select(taxon_id, scientific_name, sp_cat,
         is_ok_v5 = is_ok) |>
  collect()
d_v6 <- d_taxon |>
  select(taxon_id, is_ok_v6 = is_ok)

d_cmp <- d_v5 |> inner_join(d_v6, by = "taxon_id")

d_lost <- d_cmp |> filter(is_ok_v5 & !is_ok_v6)

d_lost |>
  count(sp_cat, name = "n_lost") |>
  arrange(desc(n_lost)) |>
  datatable(
    caption = glue(
      "{nrow(d_lost)} species changed is_ok TRUE → FALSE (v5 → v6), by component"),
    options = list(dom = "t"))
```

### Score differences per Program Area and component

```{r}
#| label: score_diff

tbl_pra_v5 <- gsub(ver, ver_prev, tbl_pra)

get_zt <- function(con, tbl_z) {
  dbGetQuery(con, paste0(
    "SELECT zone_value AS pra, sp_cat, ",
    "COUNT(*) AS n_spp, ",
    "ROUND(AVG(suit_rl), 6) AS avg_score ",
    "FROM zone_taxon ",
    "WHERE zone_tbl = '", tbl_z, "' ",
    "GROUP BY zone_value, sp_cat ",
    "ORDER BY zone_value, sp_cat"))
}

d_zt5 <- get_zt(con_v5, tbl_pra_v5) |>
  rename(n5 = n_spp, s5 = avg_score)
d_zt6 <- get_zt(con_sdm, tbl_pra) |>
  rename(n6 = n_spp, s6 = avg_score)

d_zt <- d_zt5 |>
  full_join(d_zt6, by = c("pra", "sp_cat")) |>
  mutate(
    n_diff = coalesce(n6, 0L) - coalesce(n5, 0L),
    s_diff = round(coalesce(s6, 0) - coalesce(s5, 0), 6)) |>
  filter(n_diff != 0 | s_diff != 0) |>
  arrange(pra, sp_cat)

datatable(
  d_zt,
  caption = glue(
    "{nrow(d_zt)} PRA × component rows with changes ",
    "(n_diff = species count change, s_diff = avg score change)"),
  filter  = "top",
  options = list(dom = "ftp", pageLength = 20)) |>
  formatRound(columns = c("s5", "s6", "s_diff"), digits = 4)
```

### Why the impact is negligible

The removed species all had very low `suit_rl` scores (suitability × extinction
risk). They were range-edge artifacts from AquaMaps with minimal suitability
values in US waters, typically LC (Least Concern) with suitability of 1–5 out
of 100, producing `suit_rl` contributions near 0.0003.

```{r}
#| label: why_negligible

# show that removed species had low scores in v5
d_lost_scores <- dbGetQuery(con_v5, paste0(
  "SELECT sp_scientific, sp_cat, zone_value AS pra, ",
  "suit_rl, avg_suit, er_score ",
  "FROM zone_taxon ",
  "WHERE zone_tbl = '", tbl_pra_v5, "' ",
  "AND sp_scientific IN (",
  paste0("'", d_lost$scientific_name, "'", collapse = ","), ")"))

d_score_summary <- d_lost_scores |>
  as_tibble() |>
  group_by(sp_cat) |>
  summarize(
    n_zone_entries  = n(),
    mean_suit_rl    = round(mean(suit_rl, na.rm = TRUE), 6),
    mean_avg_suit   = round(mean(avg_suit, na.rm = TRUE), 4),
    mean_er_score   = round(mean(er_score, na.rm = TRUE), 2),
    .groups = "drop") |>
  arrange(desc(n_zone_entries))

datatable(
  d_score_summary,
  caption = glue(
    "average scores of removed species in v5 zone_taxon — ",
    "very low suit_rl confirms range-edge artifacts"),
  options = list(dom = "t")) |>
  formatRound(columns = c("mean_suit_rl"), digits = 6)
```

Because each Program Area has hundreds to thousands of species per component,
removing a few dozen low-scoring species barely moves the averaged component
scores. For example, SOC has 1,168 fish; removing 156 with near-zero scores
shifts the fish average by only +0.001 on a 0–1 scale. Only Southern California
crosses an integer rounding boundary in the final 0–100 display (coral: 15→18,
fish: 24→23).

**Bottom line**: the fix is correct and important for data integrity — it removes
species that expert IUCN range maps say do not occur in US waters. The
quantitative impact on final scores is negligible because the removed species
were AquaMaps edge-of-range artifacts with minimal suitability values.

## 4. Newly ingested mammals

Three mammals missing from the original IUCN download but present in the newer
`MAMMALS.zip`:

```{r}
#| label: new_mammals

new_mammals <- c("Phoca vitulina", "Trichechus manatus", "Pusa hispida")

d_new <- d_iucn_spp |>
  filter(scientific_name %in% new_mammals) |>
  select(scientific_name, common_name, sp_cat, is_ok, n_cells)

datatable(
  d_new,
  caption = "3 mammals newly ingested from MAMMALS.zip (with US EEZ cells)",
  options = list(dom = "t"))
```

## 5. Full summary

```{r}
#| label: full_summary

n_total_iucn <- nrow(d_iucn_spp)
n_zero       <- nrow(d_zero_cell)
n_with_cells <- n_total_iucn - n_zero
n_had_am     <- nrow(d_had_aquamaps)
n_truncated  <- nrow(d_truncated)
n_stale      <- nrow(d_stale)
n_lost       <- nrow(d_lost)

v5_ok <- sum(d_cmp$is_ok_v5)
v6_ok <- sum(d_cmp$is_ok_v6)

cat(glue(
  "## summary\n\n",
  "- total species with rng_iucn in database: {n_total_iucn}\n",
  "  - with US EEZ cells: {n_with_cells}\n",
  "  - with 0 cells (range outside US EEZ): {n_zero}\n",
  "- of 0-cell species, had AquaMaps in US EEZ: {n_had_am}\n",
  "- of those, now is_ok=FALSE (truncated): {n_truncated}\n",
  "- stale ms_merge entries remaining: {n_stale}\n",
  "- is_ok=TRUE species: v5={v5_ok}, v6={v6_ok} (net change: {v6_ok - v5_ok})\n",
  "- species changed is_ok TRUE→FALSE: {n_lost}\n",
  "- score impact: negligible (removed species were AquaMaps edge-of-range ",
  "artifacts with near-zero suitability × extinction risk)\n\n"))

# breakdown by component
d_summary <- d_iucn_spp |>
  group_by(sp_cat) |>
  summarize(
    n_total      = n(),
    n_with_cells = sum(n_cells > 0),
    n_zero_cell  = sum(n_cells == 0),
    n_is_ok_T    = sum(is_ok, na.rm = TRUE),
    n_is_ok_F    = sum(!is_ok, na.rm = TRUE),
    .groups      = "drop") |>
  arrange(desc(n_total))

datatable(
  d_summary,
  caption = glue(
    "rng_iucn species by component: total, with/without US EEZ cells, ",
    "is_ok status"),
  options = list(dom = "t"))
```

```{r}
#| label: cleanup
#| include: false

dbDisconnect(con_v5, shutdown = TRUE)
dbDisconnect(con_sdm)
```
