---
title: "Merge models — combine each taxon's models per cell (max-merge + range mask; turtles ×)"
msens:
  target_name: merge_models
  workflow_type: merge
  dependency: [merge_models_prep]
  output: data/manifests/merge_models.json
editor_options:
  chunk_output_type: console
---

Combine every taxon's per-dataset models into one **merged `model_cell`**
(`mdl_key = ms_merge|WORMS:…` / `BOTW:…`), reproducing v7's logic:

- **Default = max-merge, masked to the range footprint.** A taxon's merged cells are
  its **range** cells (any of `rng_iucn` / `bl` / `rng_fws` / `ch_nmfs` / `ch_fws` /
  `ca_nmfs` / `rng_turtle_swot_dps`) if it has any; a taxon with only AquaMaps keeps
  its full `am` footprint. At each kept cell, `value = max` across all the taxon's
  models (the expert range bounds the extent; suitability refines the value within).
- **Turtles = multiplicative.** `value = pmax(1, round(er_value × suit_value / 100))`
  (SWOT+DPS ER × AquaMaps suitability), with critical-habitat cells overriding via max.

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, quiet = T)
source(here("libs/paths.R"))
source(here("libs/vars.R"))

dir_atlas <- glue("{dir_big_v}/marine-atlas")
dir_dist  <- glue("{dir_atlas}/dist")
merge_db  <- glue("{dir_atlas}/merge.duckdb")
manifest  <- here("data/manifests/merge_models.json")
dir_create(path_dir(manifest))
stopifnot("run merge_models_prep first" = file_exists(merge_db))

suit_ds  <- "am"                    # suitability dataset(s); everything else is a range/mask
turtle_ds <- "rng_turtle_swot_dps"
ch_ds     <- c("ch_nmfs", "ch_fws", "ca_nmfs")

con <- dbConnect(duckdb(merge_db))
dbExecute(con, glue("CREATE OR REPLACE VIEW model_cell_all AS
  SELECT tm.ms_merge_key, tm.taxon_authority, split_part(mc.mdl_key,'|',1) AS ds_key,
         mc.cell_id, mc.value
  FROM read_parquet('{dir_dist}/*/*.parquet') mc
  JOIN taxon_model tm USING (mdl_key)"))
```

## Turtle taxa (multiplicative path)

```{r}
#| label: turtles

turtle_keys <- dbGetQuery(con, glue(
  "SELECT DISTINCT ms_merge_key FROM model_cell_all WHERE ds_key = '{turtle_ds}'"))$ms_merge_key
log_info("{length(turtle_keys)} turtle taxa (multiplicative merge)")

dbExecute(con, glue("CREATE OR REPLACE TABLE mc_turtle AS
  WITH er AS (   -- SWOT+DPS extinction-risk surface
    SELECT ms_merge_key, cell_id, max(value) er_value FROM model_cell_all
     WHERE ds_key = '{turtle_ds}' GROUP BY ms_merge_key, cell_id),
  suit AS (      -- AquaMaps suitability
    SELECT ms_merge_key, cell_id, max(value) suit_value FROM model_cell_all
     WHERE ds_key = '{suit_ds}' GROUP BY ms_merge_key, cell_id),
  ch AS (        -- critical habitat (max override)
    SELECT ms_merge_key, cell_id, max(value) ch_value FROM model_cell_all
     WHERE ds_key IN ('{paste(ch_ds, collapse=\"','\")}') GROUP BY ms_merge_key, cell_id),
  mult AS (      -- er * suit / 100 over the ER footprint
    SELECT er.ms_merge_key, er.cell_id,
           greatest(1, CAST(round(er.er_value * coalesce(suit.suit_value,1) / 100.0) AS INTEGER)) AS value
    FROM er LEFT JOIN suit USING (ms_merge_key, cell_id))
  SELECT ms_merge_key, cell_id, CAST(greatest(m.value, coalesce(ch.ch_value, 0)) AS DOUBLE) AS value
  FROM mult m LEFT JOIN ch USING (ms_merge_key, cell_id)
  WHERE ms_merge_key IN (SELECT DISTINCT ms_merge_key FROM model_cell_all WHERE ds_key='{turtle_ds}')"))
```

## Non-turtle taxa (max-merge, masked to range footprint)

```{r}
#| label: max_merge

dbExecute(con, glue("CREATE OR REPLACE TABLE mc_maxmerge AS
  WITH mc AS (
    SELECT *, (ds_key <> '{suit_ds}') AS is_range FROM model_cell_all
     WHERE ms_merge_key NOT IN (SELECT DISTINCT ms_merge_key FROM mc_turtle)),
  taxa_range AS (SELECT ms_merge_key, bool_or(is_range) taxon_has_range FROM mc GROUP BY ms_merge_key),
  cell_range AS (SELECT ms_merge_key, cell_id, bool_or(is_range) cell_has_range
                   FROM mc GROUP BY ms_merge_key, cell_id)
  SELECT mc.ms_merge_key, mc.cell_id, max(mc.value) AS value
  FROM mc JOIN taxa_range tr USING (ms_merge_key) JOIN cell_range cr USING (ms_merge_key, cell_id)
  WHERE NOT tr.taxon_has_range OR cr.cell_has_range    -- am-only taxon keeps all cells; else range footprint
  GROUP BY mc.ms_merge_key, mc.cell_id"))
```

## Merged model_cell + partitioned Parquet

```{r}
#| label: write

dbExecute(con, "CREATE OR REPLACE TABLE model_cell AS
  SELECT ms_merge_key AS mdl_key, cell_id, value FROM mc_maxmerge
  UNION ALL SELECT ms_merge_key AS mdl_key, cell_id, value FROM mc_turtle")

dir_merged <- glue("{dir_atlas}/dist_merged/dataset=ms_merge")
if (dir_exists(dir_merged)) dir_delete(dir_merged)
dir_create(dir_merged)
dbExecute(con, glue("COPY (SELECT * FROM model_cell) TO '{dir_merged}'
  (FORMAT parquet, PARTITION_BY (mdl_key), COMPRESSION zstd, OVERWRITE_OR_IGNORE)"))

smry <- dbGetQuery(con, "SELECT count(DISTINCT mdl_key) n_taxa, count(*) n_cells,
  min(value) v_min, max(value) v_max FROM model_cell")
knitr::kable(smry)
dbDisconnect(con, shutdown = TRUE)
```

## Manifest

```{r}
#| label: manifest

write_json(
  list(target = "merge_models", ver = ver, n_taxa = smry$n_taxa, n_cells = smry$n_cells,
       dir_merged = dir_merged, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
```
