---
title: "Merge models — combine each taxon's models per cell (max-merge + range mask; turtles ×)"
msens:
  target_name: merge_models
  workflow_type: merge
  dependency: [merge_models_prep]
  output: data/manifests/merge_models.json
editor_options:
  chunk_output_type: console
---

Combine every taxon's per-dataset models into one **merged `model_cell`**
(`mdl_key = ms_merge|WORMS:…` / `BOTW:…`), reproducing v7's logic at the v8 global
scale (~13B am cell-rows), so it must stream + spill rather than materialize:

- **Default = max-merge, masked to the range footprint.** A taxon's merged cells are
  its **range** cells if it has any (`rng_*`/`ch_*`/`ca_*`/`bl`), else its full `am`
  footprint; at each kept cell `val = max` across the taxon's models. Streamed
  straight to Parquet via a single `COPY` (one grouped scan), so no 13B-row table.
- **Turtles = multiplicative.** `val = pmax(1, round(er × suit / 100))` with
  critical-habitat max-override — computed from **only the 6 turtle taxa's files**
  (never a full am scan, which is what OOM'd the naive version).

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "US-scoped per-taxon merge: max-merge over range footprint (governing er) + multiplicative turtles → streamed Parquet"
flowchart LR
  raw["dist/*/*.parquet<br/>(raw model_cell)"] --> mca["model_cell_all<br/>US-scoped, by ms_merge_key"]
  mca --> mx["max-merge over range<br/>(er_score at range cells)"]
  mca --> tt["turtles ×<br/>pmax(1, er·suit/100)"]
  mx --> pq[("dist_merged/dataset=ms_merge/**<br/>V2/zstd via copy_atlas_parquet")]
  tt --> pq
  pq --> mf["hash_parquet → content-addressed manifest"]
```

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, quiet = T)
source(here("libs/paths.R"))
source(here("libs/vars.R"))

dir_atlas   <- glue("{dir_big_v}/marine-atlas")
dir_dist    <- glue("{dir_atlas}/dist")
merge_db    <- glue("{dir_atlas}/merge.duckdb")
# ONE global merge → two surfaces from the same data:
dir_global <- glue("{dir_atlas}/dist_merged_global/dataset=ms_merge")  # whole-range (land+ocean) → viz COGs
dir_merged <- glue("{dir_atlas}/dist_merged/dataset=ms_merge")          # US-study-area trim → model_cell scoring
manifest   <- here("data/manifests/merge_models.json")
dir_create(path_dir(manifest))
stopifnot("run merge_models_prep first" = file_exists(merge_db),
          "run build_cell_grid first"   = file_exists(sdm_db))

suit_ds   <- "am"
turtle_ds <- "rng_turtle_swot_dps"
ch_ds     <- c("ch_nmfs", "ch_fws", "ca_nmfs")
ch_sql    <- paste(sprintf("'%s'", ch_ds), collapse = ",")

con <- dbConnect(duckdb(merge_db))
# spill config so the multi-billion-row aggregation goes out-of-core instead of OOM. Keep the
# memory limit WELL under physical RAM (other processes — browser, renders — compete), cap threads
# (each parallel aggregate/join holds hash tables), and allow a large temp spill.
mem_gb <- tryCatch(
  max(4, min(8, floor(as.numeric(system("sysctl -n hw.memsize", intern = TRUE)) / 1e9 * 0.3))),
  error = function(e) 6)
dbExecute(con, glue("PRAGMA memory_limit='{mem_gb}GB'"))
dbExecute(con, glue("PRAGMA temp_directory='{dir_atlas}/duckdb_tmp'"))
dbExecute(con, glue("PRAGMA max_temp_directory_size='400GB'"))
dbExecute(con, glue("PRAGMA threads={min(4L, max(1L, parallel::detectCores() - 1L))}"))

# ONE GLOBAL merge, two outputs from the same data: the whole-range merged surface
# (dist_merged_global → viz COGs) and, trimmed to the US study area, the model_cell scoring surface
# (dist_merged). am is DENSE (~13B rows, 50G) so it is NEVER grouped: range cells (non-am) are
# materialized once + indexed (mc_range, ~3B); am is a VIEW (mc_am) probed only at the sparse range
# cells (both/turtle) or streamed for am-only taxa. us_cells is applied AFTER the global merge.
dbExecute(con, glue("ATTACH '{sdm_db}' AS grid (READ_ONLY)"))
dbExecute(con, "CREATE OR REPLACE TABLE us_cells AS SELECT cell_id FROM grid.cell WHERE in_usa")
n_us <- dbGetQuery(con, "SELECT count(*) n FROM us_cells")$n
# PARTITION the full merge input (ranges + am, ~16B rows) by ms_merge_key — a STREAMING write that
# never holds it in memory. The merge below then reads one BATCH of taxa at a time via
# `ms_merge_key IN (...)`, so DuckDB partition-prunes to just those taxa's slices: bounded memory,
# no OOM (a bulk join over the ~3B `both` range cells otherwise blows the memory limit).
try(dbExecute(con, "DROP TABLE IF EXISTS model_cell_all"), silent = TRUE)
try(dbExecute(con, "DROP TABLE IF EXISTS mc_range"),       silent = TRUE)
# ms_merge_key holds '|' and ':' which break hive partition PATHS (…/mc_parts/ms_merge_key=ms_merge|…
# can't be parsed back). Partition on an INTEGER surrogate mkey_id — exactly what the serve surface
# does with mdl_id. ms_merge_key stays a stored column.
dbExecute(con, "CREATE OR REPLACE TABLE mkey_map AS
  SELECT ms_merge_key, CAST(dense_rank() OVER (ORDER BY ms_merge_key) AS INTEGER) AS mkey_id
  FROM (SELECT DISTINCT ms_merge_key FROM taxon_model)")
dir_parts <- glue("{dir_atlas}/mc_parts")
# the partition write is the ~40-min step; skip it if the input already exists (REDO_MC_PARTS=1 forces)
if (!dir_exists(dir_parts) || nzchar(Sys.getenv("REDO_MC_PARTS"))) {
  if (dir_exists(dir_parts)) dir_delete(dir_parts)
  dbExecute(con, "SET partitioned_write_max_open_files=512")
  msens::copy_atlas_parquet(con, glue(
    "SELECT mm.mkey_id, tm.ms_merge_key, split_part(mc.mdl_key,'|',1) AS ds_key, mc.cell_id, mc.val
     FROM read_parquet('{dir_dist}/*/*.parquet') mc JOIN taxon_model tm USING (mdl_key)
     JOIN mkey_map mm USING (ms_merge_key)"),
    dir_parts, partition_by = "mkey_id")
} else log_info("mc_parts exists — skipping repartition (REDO_MC_PARTS=1 to force)")
# VIEW over the partitioned input — EVERY query below filters mkey_id (integer) so reads are pruned
dbExecute(con, glue("CREATE OR REPLACE VIEW mc_parts AS
  SELECT mkey_id, ms_merge_key, ds_key, cell_id, val
  FROM read_parquet('{dir_parts}/**/*.parquet', hive_partitioning = true)"))
all_ids <- dbGetQuery(con, "SELECT mkey_id FROM mkey_map ORDER BY mkey_id")$mkey_id
log_info("partitioned merge input: {length(all_ids)} taxa by mkey_id; US trim {n_us} in_usa cells; memory_limit={mem_gb}GB")
```

## Turtle taxa (multiplicative, from only their files)

```{r}
#| label: turtles

turtle_keys <- dbGetQuery(con, glue(
  "SELECT DISTINCT ms_merge_key FROM taxon_model WHERE mdl_key LIKE '{turtle_ds}|%'"))$ms_merge_key
tkeys_sql   <- paste(sprintf("'%s'", turtle_keys), collapse = ",")
turtle_ids  <- dbGetQuery(con, glue("SELECT mkey_id FROM mkey_map WHERE ms_merge_key IN ({tkeys_sql})"))$mkey_id
tids_sql    <- paste(turtle_ids, collapse = ",")
log_info("{length(turtle_keys)} turtle taxa (multiplicative merge)")

# whole-range (global) multiplicative turtle merge, reading ONLY the turtle taxa's partitions
# (mkey_id IN (...) → partition-pruned, tiny). er/ch/suit all come from the one pruned slice.
dbExecute(con, glue("CREATE OR REPLACE TABLE mc_turtle AS
  WITH t AS (SELECT ms_merge_key, ds_key, cell_id, val FROM mc_parts WHERE mkey_id IN ({tids_sql})),
       er   AS (SELECT ms_merge_key, cell_id, max(val) er_value   FROM t WHERE ds_key='{turtle_ds}' GROUP BY 1,2),
       suit AS (SELECT ms_merge_key, cell_id, max(val) suit_value FROM t WHERE ds_key='{suit_ds}'   GROUP BY 1,2),
       ch   AS (SELECT ms_merge_key, cell_id, max(val) ch_value   FROM t WHERE ds_key IN ({ch_sql})  GROUP BY 1,2),
       mult AS (SELECT er.ms_merge_key, er.cell_id,
                  greatest(1, CAST(round(er.er_value * coalesce(suit.suit_value,1) / 100.0) AS INTEGER)) AS val
                FROM er LEFT JOIN suit USING (ms_merge_key, cell_id))
  SELECT m.ms_merge_key AS mdl_key, m.cell_id,
         CAST(greatest(m.val, coalesce(ch.ch_value, 0)) AS DOUBLE) AS val
  FROM mult m LEFT JOIN ch USING (ms_merge_key, cell_id)"))
log_info("mc_turtle (whole-range): {dbGetQuery(con, 'SELECT count(*) n FROM mc_turtle')$n} cells")
```

## Max-merge (masked) + turtles → streamed to Parquet

```{r}
#| label: merge_write

# taxon flags from the crosswalk (no cell scan)
dbExecute(con, "CREATE OR REPLACE TABLE taxon_flags AS
  SELECT ms_merge_key,
         bool_or(split_part(mdl_key,'|',1)='am')  AS has_am,
         bool_or(split_part(mdl_key,'|',1)<>'am') AS has_range
  FROM taxon_model GROUP BY ms_merge_key")

if (dir_exists(dir_global)) dir_delete(dir_global); dir_create(dir_global)
if (dir_exists(dir_merged)) dir_delete(dir_merged); dir_create(dir_merged)

# turtles (done) → whole-range global + US-trim
msens::copy_atlas_parquet(con, "SELECT mdl_key, cell_id, val FROM mc_turtle",
  glue("{dir_global}/turtle"), per_thread = TRUE)
msens::copy_atlas_parquet(con,
  "SELECT mt.mdl_key, mt.cell_id, mt.val FROM mc_turtle mt JOIN us_cells u ON mt.cell_id = u.cell_id",
  glue("{dir_merged}/turtle"), per_thread = TRUE)

# BATCH the rest so each merge reads ONLY its taxa's pruned partitions (bounded memory). The
# whole-range merge per cell = max(the taxon's GOVERNING er over its range footprint, AquaMaps
# suitability AT those range cells) — am OUTSIDE the range is masked out (range-only → am NULL).
# am-only taxa have no range footprint → omitted from the GLOBAL surface (reuse am COGs), but their
# US am cells go into the scoring surface.
# Batch by CELL COUNT, not taxa count: a few taxa have global ranges (10-25M cells each), so a
# fixed 150-taxa batch OOMs. Pack taxa until ~40M cells/batch (a single taxon maxes at the 25.9M
# grid, so it always fits one batch). Counts come from parquet metadata (fast, no data scan).
nonturtle <- setdiff(all_ids, turtle_ids)
cnt <- dbGetQuery(con, "SELECT mkey_id, count(*) AS n FROM mc_parts GROUP BY mkey_id")
cnt <- cnt[cnt$mkey_id %in% nonturtle, ]; cnt <- cnt[order(cnt$mkey_id), ]
batches <- unname(split(cnt$mkey_id, cumsum(as.numeric(cnt$n)) %/% 40e6))
log_info("merging {length(nonturtle)} non-turtle taxa in {length(batches)} cell-bounded batches (~40M cells each)")
glob_sql <- "SELECT br.ms_merge_key AS mdl_key, br.cell_id,
               greatest(br.er, coalesce(ba.am_val, 0))::DOUBLE AS val
             FROM b_range br LEFT JOIN b_am ba USING (ms_merge_key, cell_id)"
for (i in seq_along(batches)) {
  bk <- paste(batches[[i]], collapse = ",")   # integer mkey_ids → partition-pruned
  dbExecute(con, glue("CREATE OR REPLACE TABLE b AS
    SELECT ms_merge_key, ds_key, cell_id, val FROM mc_parts WHERE mkey_id IN ({bk})"))
  # range footprint (non-am) valued by the taxon's governing er_score
  dbExecute(con, "CREATE OR REPLACE TABLE b_range AS
    SELECT DISTINCT b.ms_merge_key, b.cell_id, t.er_score::DOUBLE AS er
    FROM b JOIN taxon t ON b.ms_merge_key = t.ms_merge_key WHERE b.ds_key <> 'am'")
  # am AT those range cells (both taxa)
  dbExecute(con, "CREATE OR REPLACE TABLE b_am AS
    SELECT b.ms_merge_key, b.cell_id, max(b.val) am_val
    FROM b JOIN b_range USING (ms_merge_key, cell_id) WHERE b.ds_key = 'am' GROUP BY 1, 2")
  msens::copy_atlas_parquet(con, glob_sql, glue("{dir_global}/b{i}"), per_thread = TRUE)
  msens::copy_atlas_parquet(con, glue(
    "SELECT g.mdl_key, g.cell_id, g.val FROM ({glob_sql}) g JOIN us_cells u ON g.cell_id = u.cell_id
     UNION ALL
     SELECT b.ms_merge_key AS mdl_key, b.cell_id, b.val FROM b JOIN us_cells u ON b.cell_id = u.cell_id
      WHERE b.ds_key='am' AND b.ms_merge_key IN (SELECT ms_merge_key FROM taxon_flags WHERE has_am AND NOT has_range)"),
    glue("{dir_merged}/b{i}"), per_thread = TRUE)
  if (i %% 10 == 0) log_info("  batch {i}/{length(batches)}")
}
dbExecute(con, "DROP TABLE IF EXISTS b; DROP TABLE IF EXISTS b_range; DROP TABLE IF EXISTS b_am")

# summaries: global (viz) + US (scoring); fingerprint the US scoring surface (the tracked output)
smry_g <- dbGetQuery(con, glue("SELECT count(DISTINCT mdl_key) n_taxa, count(*) n_cells,
  min(val) v_min, max(val) v_max FROM read_parquet('{dir_global}/**/*.parquet')"))
smry   <- dbGetQuery(con, glue("SELECT count(DISTINCT mdl_key) n_taxa, count(*) n_cells,
  min(val) v_min, max(val) v_max FROM read_parquet('{dir_merged}/**/*.parquet')"))
h <- msens::hash_parquet(glue("{dir_merged}/**/*.parquet"), con)
msens::report_table(smry_g, caption = "GLOBAL whole-range merged surface (ms_merge) → viz COGs")
msens::report_table(smry,   caption = "US-trimmed merged model_cell (ms_merge) → scoring")
dbDisconnect(con, shutdown = TRUE)
```

## Manifest

```{r}
#| label: manifest

# content-addressed: fingerprint of the US scoring surface (the tracked target output); the global
# viz surface (dist_merged_global) is recorded in stats but doesn't drive the hash.
msens::write_manifest(
  manifest, target = "merge_models", content_hash = h,
  stats = list(ver = ver, n_taxa = smry$n_taxa, n_cells = smry$n_cells,
               v_min = smry$v_min, v_max = smry$v_max,
               n_taxa_global = smry_g$n_taxa, n_cells_global = smry_g$n_cells),
  force = msens::force_target("merge_models"))
```
