---
title: "Merge prep — taxon crosswalk (each model → WORMS / BOTW taxon)"
msens:
  target_name: merge_models_prep
  workflow_type: merge
  dependency: [ingest_aquamaps, ingest_rng_iucn, ingest_birdlife_botw, ingest_rng_fws, ingest_ch_nmfs, ingest_ch_fws, ingest_turtles_swot_dps, ingest_ca_nmfs]
  output: data/manifests/merge_models_prep.json
editor_options:
  chunk_output_type: console
---

Resolve every per-dataset model to a **common taxon**, so models of the same species
across datasets merge together. Two taxon namespaces (as in v7):

- **Birds → `BOTW:{sisid}`** — native BirdLife id. WoRMS has poor bird coverage, so
  birds are NOT resolved through WoRMS.
- **Non-birds → `WORMS:{worms_id}`** — resolved from `scientific_name` by **reusing
  v7's taxon table** (`scientific_name → worms_id`, cheap + already curated) and
  `msens::match_taxa()` (ITIS→WoRMS crosswalk, exact WoRMS match, then WoRMS REST) for
  species v7 didn't have.

Emits `taxon` (one row per taxon: id, authority, name, `worms_is_marine`…) and
`taxon_model` (`mdl_key ↔ taxon_id`), keyed for merge by
`mdl_key_merged(authority, taxon_id)` → `ms_merge|WORMS:…` / `ms_merge|BOTW:…`.

## Setup + read all model crosswalks

```{r}
#| label: setup

librarian::shelf(
  DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, purrr, readr, stringr,
  tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))   # dir_big_v, spp_db, ver, ver_prev, dir_big
source(here("libs/vars.R"))
options(readr.show_col_types = F)

dir_atlas <- glue("{dir_big_v}/marine-atlas")
dir_dist  <- glue("{dir_atlas}/dist")
v7_sdm_db <- glue("{dir_big}/{ver_prev}/sdm.duckdb")   # reuse v7 taxon resolution
manifest  <- here("data/manifests/merge_models_prep.json")
dir_create(path_dir(manifest))

# model_{ds}.csv is the dataset's species crosswalk (may list MORE than were
# ingested, e.g. FWS lists all 2,196 but ingests 106) — so key off the mdl_keys that
# actually have model_cell Parquet, joined to the CSVs for scientific_name.
sci_cols <- c("scientific_name", "sci_name", "sciname", "SCIENAME", "taxa")
lookup <- dir_ls(dir_dist, glob = "*model_*.csv") |>
  set_names(\(f) str_remove(path_ext_remove(path_file(f)), "^model_")) |>
  imap(\(f, ds) {
    m    <- read_csv(f)
    sci  <- intersect(sci_cols, names(m))[1]
    iucn <- intersect(c("category", "code", "iucn"), names(m))[1]   # IUCN Red List code (rng_iucn/bl)
    tibble(
      mdl_key         = m$mdl_key,
      ds_key          = ds,
      scientific_name = m[[sci]],
      iucn_code       = if (is.na(iucn)) NA_character_ else as.character(m[[iucn]]),
      sisid           = if ("sisid" %in% names(m)) as.character(m$sisid) else NA_character_) }) |>
  bind_rows() |> distinct(mdl_key, .keep_all = TRUE)

# read ingested mdl_keys per dataset so a mid-transfer / corrupt dataset (e.g. am
# actively rsyncing, with a truncated "no magic bytes" parquet) is skipped with a
# warning rather than crashing the whole glob — its taxa resolve on the next run.
con0     <- dbConnect(duckdb())
ds_dirs  <- dir_ls(dir_dist, type = "directory", regexp = "dataset=")
ingested <- unlist(lapply(ds_dirs, \(d) tryCatch(
  dbGetQuery(con0, glue("SELECT DISTINCT mdl_key FROM read_parquet('{d}/*.parquet')"))$mdl_key,
  error = \(e) { log_warn("skip {path_file(d)}: {conditionMessage(e)}"); character(0) })))
dbDisconnect(con0, shutdown = TRUE)
models <- lookup |> filter(mdl_key %in% ingested)
count(models, ds_key)
```

## Non-bird taxa → worms_id (reuse v7, then match_taxa)

```{r}
#| label: worms

nonbird <- models |> filter(ds_key != "bl") |> distinct(scientific_name) |>
  mutate(sci_clean = clean_sci_name(scientific_name),                 # drop (=synonym), ssp. markers
         sci_binom = clean_sci_name(scientific_name, binomial = TRUE)) # Genus species fallback

# reuse v7's taxon resolution, matched on the cleaned name
con_v7  <- dbConnect(duckdb(v7_sdm_db, read_only = TRUE))
v7_tax  <- tbl(con_v7, "taxon") |> filter(taxon_authority == "worms") |>
  select(scientific_name, worms_id, worms_is_marine, worms_is_extinct) |> collect()
dbDisconnect(con_v7, shutdown = TRUE)
v7_tax  <- v7_tax |> mutate(sci_clean = clean_sci_name(scientific_name)) |>
  distinct(sci_clean, .keep_all = TRUE) |> select(-scientific_name)
resolved <- nonbird |> left_join(v7_tax, by = "sci_clean")
log_info("non-bird taxa: {nrow(resolved)} | reused from v7: {sum(!is.na(resolved$worms_id))}")

# match_taxa (ITIS->WoRMS, exact, REST) for the rest: cleaned name, then binomial fallback
match_names <- function(v) {
  con_spp <- dbConnect(duckdb(spp_db, read_only = TRUE)); on.exit(dbDisconnect(con_spp, shutdown = TRUE))
  match_taxa(tibble(scientific_name = unique(v)), con_spp) |>
    filter(!is.na(worms_id)) |> distinct(scientific_name, .keep_all = TRUE)
}
for (col in c("sci_clean", "sci_binom")) {
  todo <- resolved |> filter(is.na(worms_id)) |> pull(.data[[col]]) |> unique()
  if (length(todo) == 0) next
  m <- match_names(todo) |> transmute(!!col := scientific_name, wid = worms_id)
  resolved <- resolved |> left_join(m, by = col) |>
    mutate(worms_id = coalesce(worms_id, wid)) |> select(-wid)
}
log_info("non-bird taxa resolved: {sum(!is.na(resolved$worms_id))} of {nrow(resolved)}")
```

## Build taxon + taxon_model

```{r}
#| label: taxon

# non-bird models -> worms taxon
tm_worms <- models |> filter(ds_key != "bl") |>
  left_join(resolved |> select(scientific_name, worms_id), by = "scientific_name") |>
  filter(!is.na(worms_id)) |>
  transmute(mdl_key, taxon_authority = "worms", taxon_id = as.character(worms_id))

# bird models -> botw taxon (native sisid)
tm_botw <- models |> filter(ds_key == "bl", !is.na(sisid)) |>
  transmute(mdl_key, taxon_authority = "botw", taxon_id = as.character(sisid))

taxon_model <- bind_rows(tm_worms, tm_botw) |>
  mutate(ms_merge_key = mdl_key_merged(taxon_authority, taxon_id))

taxon <- taxon_model |>
  left_join(models |> select(mdl_key, scientific_name, iucn_code), by = "mdl_key") |>
  group_by(taxon_authority, taxon_id, ms_merge_key) |>
  summarise(
    scientific_name = first(scientific_name),
    iucn_code       = first(c(iucn_code[!is.na(iucn_code)], NA_character_)),  # IUCN category if any model has it
    n_models        = n_distinct(mdl_key),
    n_datasets      = n_distinct(str_extract(mdl_key, "^[^|]+")), .groups = "drop") |>
  left_join(resolved |> select(scientific_name, worms_is_marine, worms_is_extinct),
            by = "scientific_name")

con <- dbConnect(duckdb(glue("{dir_atlas}/merge.duckdb")))
dbWriteTable(con, "taxon_model", taxon_model, overwrite = TRUE)
dbWriteTable(con, "taxon", taxon, overwrite = TRUE)
dbDisconnect(con, shutdown = TRUE)

glue("{nrow(taxon)} taxa ({sum(taxon$taxon_authority=='worms')} WORMS + ",
     "{sum(taxon$taxon_authority=='botw')} BOTW) from {nrow(taxon_model)} models")
```

## Manifest

```{r}
#| label: manifest

write_json(
  list(target = "merge_models_prep", ver = ver,
       n_models = nrow(taxon_model), n_taxa = nrow(taxon),
       n_worms = sum(taxon$taxon_authority == "worms"),
       n_botw  = sum(taxon$taxon_authority == "botw"),
       built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
```
