---
title: "Merge taxon metrics — validity flags + range / rarity / endemism / pct_marine"
msens:
  target_name: merge_taxon
  workflow_type: merge
  dependency: [merge_models, build_cell_grid, ingest_listings]
  output: data/manifests/merge_taxon.json
editor_options:
  chunk_output_type: console
---

Per-taxon attributes on the merged `model_cell`, joined to the `cell` table (ocean
cells with `area_km2` / `in_usa` / `in_pra`):

- **3 nested validity flags** (`is_valid_global ⊇ is_valid_usa ⊇ is_valid_pra`):
  a taxon is *valid_global* if it has a mapped distribution reaching the ocean, isn't
  extinct, and isn't flagged non-marine (`worms_is_marine`); *valid_usa* / *valid_pra*
  add "has cells in the US study area / Program Areas".
- **range / rarity / endemism / pct_marine**: `range_km2` = Σ `area_km2` over the
  taxon's ocean cells (v7-comparable marine range); `pct_marine` = ocean-cell share of
  the whole (land+ocean) range; `us_endemism` = US share of the marine range; rarity
  class from the marine `range_km2` distribution.

## Setup + compute

```{r}
#| label: compute

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, stringr,
                 tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))
source(here("libs/vars.R"))

dir_atlas <- glue("{dir_big_v}/marine-atlas")
merged    <- glue("{dir_atlas}/dist_merged/dataset=ms_merge")
merge_db  <- glue("{dir_atlas}/merge.duckdb")
manifest  <- here("data/manifests/merge_taxon.json")
dir_create(path_dir(manifest))
stopifnot("run merge_models first"    = dir_exists(merged),
          "run build_cell_grid first" = file_exists(sdm_db))

con <- dbConnect(duckdb(merge_db))
dbExecute(con, glue("ATTACH '{sdm_db}' AS grid (READ_ONLY)"))

# per-taxon cell tallies against the ocean cell table
dbExecute(con, glue("CREATE OR REPLACE TABLE taxon_cell AS
  WITH mc AS (SELECT mdl_key, cell_id FROM read_parquet('{merged}/**/*.parquet')),
  j AS (SELECT mc.mdl_key, c.area_km2, c.in_usa, c.in_pra,
               (c.cell_id IS NOT NULL) AS is_ocean
        FROM mc LEFT JOIN grid.cell c USING (cell_id))
  SELECT mdl_key,
    count(*)                              AS n_cells,
    count(*) FILTER (WHERE is_ocean)      AS n_ocean,
    count(*) FILTER (WHERE in_usa)        AS n_usa,
    count(*) FILTER (WHERE in_pra)        AS n_pra,
    round(sum(area_km2) FILTER (WHERE is_ocean), 1) AS range_km2,
    round(sum(area_km2) FILTER (WHERE in_usa),   1) AS range_usa_km2
  FROM j GROUP BY mdl_key"))
```

## Flags + range/rarity/endemism → taxon

```{r}
#| label: flags

# rarity thresholds on the marine range (km2); tune vs v7
dbExecute(con, "CREATE OR REPLACE TABLE taxon AS
  WITH t AS (
    SELECT x.*, tc.n_cells, tc.n_ocean, tc.n_usa, tc.n_pra, tc.range_km2, tc.range_usa_km2
    FROM taxon x LEFT JOIN taxon_cell tc ON x.ms_merge_key = tc.mdl_key)
  SELECT *,
    -- 3 nested validity flags (global superset of usa superset of pra)
    (n_ocean > 0 AND coalesce(worms_is_extinct, FALSE) = FALSE
       AND coalesce(worms_is_marine, TRUE) = TRUE)        AS is_valid_global,
    (n_usa > 0)                                            AS is_valid_usa,
    (n_pra > 0)                                            AS is_valid_pra,
    round(100.0 * n_ocean / nullif(n_cells, 0), 1)         AS pct_marine,
    round(100.0 * n_usa   / nullif(n_ocean, 0), 1)         AS us_endemism,
    CASE
      WHEN range_km2 IS NULL       THEN NULL
      WHEN range_km2 <     10000   THEN 'very_rare'
      WHEN range_km2 <    100000   THEN 'rare'
      WHEN range_km2 <   1000000   THEN 'uncommon'
      WHEN range_km2 <  10000000   THEN 'common'
      ELSE                              'widespread' END   AS rarity
  FROM t")

smry <- dbGetQuery(con, "SELECT
  count(*) n_taxa,
  count(*) FILTER (WHERE is_valid_global) valid_global,
  count(*) FILTER (WHERE is_valid_usa)    valid_usa,
  count(*) FILTER (WHERE is_valid_pra)    valid_pra,
  round(median(pct_marine),1) med_pct_marine FROM taxon")
knitr::kable(smry)
dbDisconnect(con, shutdown = TRUE)
```

## Taxon extinction-risk (`extrisk_code` + `er_score`)

Governing code by precedence: **US listing** (ESA status / MMPA / MBTA from
`ingest_listings`) if present, else the **IUCN** category; `er_score` via
`msens::compute_er_score()` with the MMPA/MBTA floors.

```{r}
#| label: extrisk

con <- dbConnect(duckdb(merge_db))
tx  <- dbGetQuery(con, "SELECT ms_merge_key, scientific_name, iucn_code FROM taxon") |>
  mutate(sci = clean_sci_name(scientific_name))
if ("listing" %in% dbListTables(con)) {
  tx <- tx |> left_join(
    dbGetQuery(con, "SELECT sci, nmfs_esa, fws_esa, is_mmpa, is_mbta, is_bcc FROM listing"),
    by = "sci")
} else {
  log_warn("no listing table — run ingest_listings; er_score falls back to IUCN only")
  tx <- tx |> mutate(nmfs_esa = NA, fws_esa = NA, is_mmpa = FALSE, is_mbta = FALSE, is_bcc = FALSE)
}
esa_rank <- function(x) match(coalesce(x, "LC"), c("LC", "TN", "EN"))
tx <- tx |> mutate(
  across(c(is_mmpa, is_mbta, is_bcc), \(v) coalesce(v, FALSE)),
  us_code = c("LC", "TN", "EN")[pmax(esa_rank(nmfs_esa), esa_rank(fws_esa))],
  is_us   = us_code != "LC" | is_mmpa | is_mbta,
  extrisk_code = case_when(
    is_us &  is_mmpa ~ paste0("NMFS:", us_code),
    is_us            ~ paste0("FWS:",  us_code),
    iucn_code %in% c("CR","EN","VU","NT","LC","DD") ~ paste0("IUCN:", iucn_code),
    TRUE ~ NA_character_))
tx$er_score <- 1L
ok <- !is.na(tx$extrisk_code)
tx$er_score[ok] <- compute_er_score(tx$extrisk_code[ok], is_mmpa = tx$is_mmpa[ok], is_mbta = tx$is_mbta[ok])

dbWriteTable(con, "taxon_er",
  tx |> select(ms_merge_key, extrisk_code, er_score, is_mmpa, is_mbta, is_bcc), overwrite = TRUE)
dbExecute(con, "CREATE OR REPLACE TABLE taxon AS
  SELECT t.*, e.extrisk_code, e.er_score, e.is_mmpa, e.is_mbta, e.is_bcc
  FROM taxon t LEFT JOIN taxon_er e USING (ms_merge_key)")
log_info("extrisk: {sum(tx$is_us)} US-listed ({sum(tx$is_mmpa)} MMPA, {sum(tx$is_mbta)} MBTA) of {nrow(tx)} taxa")
dbDisconnect(con, shutdown = TRUE)
```

## Manifest

```{r}
#| label: manifest

write_json(
  list(target = "merge_taxon", ver = ver, n_taxa = smry$n_taxa,
       valid_global = smry$valid_global, valid_usa = smry$valid_usa,
       valid_pra = smry$valid_pra, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
```
