---
title: "Merge Models"
subtitle: "Merge species distribution models and produce tabular summaries for score calculation"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Overview

Merge species distribution models from multiple datasets, flag valid taxa
(`is_ok`), and produce tabular summaries as input to `calc_scores.qmd`.

Data preparation steps (taxonomic ID resolution, ad-hoc species fixes,
dataset ingestion) are in `merge_models_prep.qmd`.


```{r}
#| label: setup
librarian::shelf(
  DBI,
  dplyr,
  DT,
  duckdb,
  glue,
  here,
  knitr,
  readr,
  stringr,
  tibble,
  tidyr,
  quiet = T
)
options(readr.show_col_types = F)

# dirs, high-level
is_server <- Sys.info()[["sysname"]] == "Linux"
dir_data <- ifelse(is_server, "/share/data", "~/My Drive/projects/msens/data")
dir_derived <- glue("{dir_data}/derived")
dir_raw <- glue("{dir_data}/raw")
dir_big <- ifelse(is_server, "/share/data/big", "~/_big/msens/derived")
dir_private <- ifelse(is_server, "/share/private", "~/My Drive/private")

# version
v_int <- 3
v_sfx <- paste0("_v", v_int)
v_dir <- paste0("v", v_int)

# dirs, versioned
dir_v <- glue("{dir_derived}/{v_dir}")
dir_big_v <- glue("{dir_big}/{v_dir}")

# files
spp_db <- glue("{dir_big}/spp.duckdb")
sdm_db <- glue("{dir_big_v}/sdm.duckdb")

# database connections
con_sdm <- dbConnect(duckdb(dbdir = sdm_db, read_only = F))

tbl(con_sdm, "taxon") |>
  filter(scientific_name == "Balaenoptera ricei") |>
  select(worms_id = taxon_id) |>
  left_join(
    tbl(con_sdm, "species"),
    by = join_by(worms_id)
  ) |>
  select(worms_id, ds_key, taxa, scientific_name_accepted) |>
  collect()


tbl(con_sdm, "model") |> collect() |> head()

# section eval flags (set to TRUE to re-run) ----
do_merge <- T # create ms_merge merged models
do_is_ok <- F # flag valid taxa

# TODO: move to merge_models_prep.qmd after merged models created.
```

## Add Derived Datasets

### Separating Suitability and Extinction Risk

The model pipeline separates two concepts that were previously conflated in
`model_cell.value`:

- **Suitability** (`is_suit`): likelihood of species presence (0-100).
  AquaMaps SDM provides continuous suitability; endemism (`ms_endemism`)
  converts binary range maps to a suitability proxy via log-scaled range
  restriction (10=widespread, 90=endemic).
  
- **Extinction Risk** (`is_er`): conservation status encoded as cell values
  (e.g., EN:100, TN:50, LC:1). Applies to range/CH/CA maps where values
  represent ESA/IUCN listing status.

The final merged model is the product:
`ms_merge = max(1, round(ms_suit * ms_er / 100))`

This ensures species never "disappear" (floor of 1) and that suitability
and extinction risk are independently interpretable layers.

### Insert derived dataset rows

```{r}
#| label: insert_dataset_derived
#| eval: !expr do_merge

# dataset templates for all derived layers
derived_datasets <- tribble(
  ~ds_key       , ~name_short                                        , ~description                                                                                                                                                                                                                           ,
  "ms_endemism" , glue("Endemism (suitability proxy), {Sys.Date()}") , "Log-scaled endemism score (10-90) derived from species range size relative to total Program Area cells. Small ranges score high (endemic), large ranges score low (dispersed). Used as a suitability proxy for species lacking SDMs." ,
  "ms_suit"     , glue("Merged Suitability, {Sys.Date()}")           , "Per-cell maximum of all suitability layers (AquaMaps SDM and endemism). Masked to IUCN range when available. Values 1-100 representing likelihood of species presence."                                                               ,
  "ms_er"       , glue("Merged Extinction Risk, {Sys.Date()}")       , "Per-cell extinction risk from the highest-priority ER layer (US > Intl, NMFS > FWS > IUCN). Cells in species range but without ER layer coverage receive the species-level max ER value. Values 1-100."                               ,
  "ms_merge"    , glue("Final Merged Model, {Sys.Date()}")           , "Product of suitability and extinction risk: max(1, round(ms_suit * ms_er / 100)). This is the final per-species model layer used for scoring."
)

for (i in 1:nrow(derived_datasets)) {
  dd <- derived_datasets[i, ]

  row_dataset <- tibble(
    ds_key = dd$ds_key,
    name_short = dd$name_short,
    name_original = dd$name_short,
    description = dd$description,
    citation = "",
    source_broad = "BOEM",
    source_detail = "https://marinesensitivity.org/docs",
    regions = "USA",
    response_type = "mixed",
    taxa_groups = "all taxa",
    year_pub = 2025,
    date_obs_beg = NA,
    date_obs_end = NA,
    date_env_beg = NA,
    date_env_end = NA,
    link_info = "https://github.com/MarineSensitivity",
    link_download = NA,
    link_metadata = NA,
    links_other = NA,
    spatial_res_deg = 0.05,
    temporal_res = "static"
  )

  if (dbExistsTable(con_sdm, "dataset")) {
    dbExecute(con_sdm, glue("DELETE FROM dataset WHERE ds_key = '{dd$ds_key}'"))
  }

  dbWriteTable(con_sdm, "dataset", row_dataset, append = TRUE)
}

# tbl(con_sdm, "dataset")
```

### Dataset metadata

```{r}
#| label: dataset_metadata
#| eval: !expr do_merge

# add metadata columns to dataset table ----
for (col_def in list(
  c("name_display", "VARCHAR"),
  c("value_info", "VARCHAR"),
  c("is_mask", "BOOLEAN"),
  c("is_suit", "BOOLEAN"),
  c("is_er", "BOOLEAN"),
  c("sort_order", "INTEGER")
)) {
  tryCatch(
    dbExecute(
      con_sdm,
      glue(
        "ALTER TABLE dataset ADD COLUMN {col_def[1]} {col_def[2]}"
      )
    ),
    error = function(e) {
      if (!grepl("already exists", e$message, ignore.case = TRUE)) {
        stop(e)
      }
    }
  )
}

# populate metadata for each dataset
# is_suit: dataset contributes to suitability (likelihood of presence)
# is_er:   dataset contributes to extinction risk (conservation status)
metadata <- tribble(
  ~ds_key       , ~name_display            , ~value_info                            , ~is_mask , ~is_suit , ~is_er , ~sort_order ,
  "am_0.05"     , "AquaMaps SDM"           , NA                                     , FALSE    , TRUE     , FALSE  ,  1L         ,
  "ca_nmfs"     , "NMFS Core Area"         , "Core: 100"                            , TRUE     , FALSE    , TRUE   ,  2L         ,
  "ch_nmfs"     , "NMFS Critical Habitat"  , "EN:100, TN:50"                        , TRUE     , FALSE    , TRUE   ,  3L         ,
  "ch_fws"      , "FWS Critical Habitat"   , "EN:100, TN:50"                        , TRUE     , FALSE    , TRUE   ,  4L         ,
  "rng_fws"     , "FWS Range"              , "EN:100, TN:50, LC:1"                  , TRUE     , FALSE    , TRUE   ,  5L         ,
  "bl"          , "BirdLife Range"         , "CR:50, EN:25, VU:5, NT:2, LC:1, DD:1" , TRUE     , FALSE    , TRUE   ,  6L         ,
  "rng_iucn"    , "IUCN Range"             , "CR:50, EN:25, VU:5, NT:2, LC:1, DD:1" , TRUE     , FALSE    , TRUE   ,  7L         ,
  "ms_endemism" , "Endemism (suitability)" , "10=dispersed, 90=endemic"             , FALSE    , TRUE     , FALSE  , 10L         ,
  "ms_suit"     , "Merged Suitability"     , NA                                     , FALSE    , FALSE    , FALSE  , 11L         ,
  "ms_er"       , "Merged Extinction Risk" , NA                                     , FALSE    , FALSE    , FALSE  , 12L         ,
  "ms_merge"    , "Final Merged Model"     , NA                                     , FALSE    , FALSE    , FALSE  ,  0L
)

for (i in 1:nrow(metadata)) {
  m <- metadata[i, ]
  vi <- ifelse(is.na(m$value_info), "NULL", glue("'{m$value_info}'"))
  dbExecute(
    con_sdm,
    glue(
      "UPDATE dataset
     SET name_display = '{m$name_display}',
         value_info   = {vi},
         is_mask      = {tolower(m$is_mask)},
         is_suit      = {tolower(m$is_suit)},
         is_er        = {tolower(m$is_er)},
         sort_order   = {m$sort_order}
     WHERE ds_key = '{m$ds_key}'"
    )
  )
}

tbl(con_sdm, "dataset") |>
  select(ds_key, name_display, is_mask, is_suit, is_er, sort_order) |>
  collect()
```

### Fix BirdLife range map values

BirdLife models currently have a flat `model_cell.value = 50` regardless of
the species' IUCN Red List status. This should match the `rng_iucn` encoding
scheme so that the ER layer values correctly reflect conservation status:
CR=50, EN=25, VU=5, NT=2, LC=1, DD=1.

```{r}
#| label: fix_birdlife_values
#| eval: !expr do_merge

bl_value_map <- c(CR = 50L, EN = 25L, VU = 5L, NT = 2L, LC = 1L, DD = 1L)

# get BirdLife models with species redlist codes
d_bl <- tbl(con_sdm, "taxon_model") |>
  filter(ds_key == "bl") |>
  inner_join(
    tbl(con_sdm, "taxon") |>
      select(taxon_id, redlist_code),
    by = "taxon_id"
  ) |>
  collect()

# map redlist codes to new values
d_bl <- d_bl |>
  mutate(
    new_value = as.integer(bl_value_map[redlist_code])
  ) |>
  # species without a recognized redlist code get DD default
  mutate(
    new_value = ifelse(is.na(new_value), 1L, new_value)
  )

# batch update model_cell values per mdl_seq
n_updated <- 0L
for (i in seq_len(nrow(d_bl))) {
  ms <- d_bl$mdl_seq[i]
  nv <- d_bl$new_value[i]

  n <- dbExecute(
    con_sdm,
    glue("UPDATE model_cell SET value = {nv} WHERE mdl_seq = {ms}")
  )
  n_updated <- n_updated + n
}

message(glue(
  "Updated {n_updated} BirdLife cells across {nrow(d_bl)} species"
))

# verify: show species counts by redlist code and new value
d_bl |>
  count(redlist_code, new_value) |>
  print()
```

### Create Endemism Layer (`ms_endemism`)

Endemism is a species-level suitability proxy derived from range restriction.
Species with smaller ranges (fewer cells in Program Areas) get higher scores,
reflecting that being within a restricted range is stronger evidence of
suitability. The log scaling provides better differentiation among endemic
species compared to linear scaling.

**Range**: 10 (most widespread) to 90 (most endemic)

- **Floor of 10**: Even the most widespread species have meaningful
  suitability — being in a known range is stronger evidence than absence
- **Ceiling of 90**: Highly endemic species likely have high suitability
  where found, but range maps carry more spatial uncertainty than SDMs
  (which can reach 100%)
- **Log scaling**: Approximate values: 1 cell → 90, 1K cells → ~60,
  10K → ~43, 50% of area → ~22, all cells → 10

```{r}
#| label: create_ms_endemism
#| eval: !expr do_merge

# get total program area cells as a lazy query (for efficient semi_joins)
q_pra_cells <- tbl(con_sdm, "zone") |>
  filter(
    tbl == !!glue("ply_programareas_2026{v_sfx}"),
    fld == "programarea_key"
  ) |>
  inner_join(
    tbl(con_sdm, "zone_cell") |> select(zone_seq, cell_id),
    by = "zone_seq"
  ) |>
  distinct(cell_id)
n_total_cells <- q_pra_cells |> collect() |> nrow()
message(glue("Total Program Area cells: {n_total_cells}"))

# dataset key definitions
ds_keys_mask <- tbl(con_sdm, "dataset") |>
  filter(is_mask) |>
  pull(ds_key)

ds_keys_input <- tbl(con_sdm, "dataset") |>
  filter(!ds_key %in% c("ms_endemism", "ms_suit", "ms_er", "ms_merge")) |>
  pull(ds_key)

# get all valid taxa
d_taxa <- tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  collect()

# clean up existing ms_endemism models
existing_ms <- tbl(con_sdm, "model") |>
  filter(ds_key == "ms_endemism") |>
  pull(mdl_seq)
if (length(existing_ms) > 0) {
  ms_str <- paste(existing_ms, collapse = ",")
  dbExecute(
    con_sdm,
    glue("DELETE FROM model_cell  WHERE mdl_seq IN ({ms_str})")
  )
  dbExecute(
    con_sdm,
    glue("DELETE FROM taxon_model WHERE ds_key = 'ms_endemism'")
  )
  dbExecute(
    con_sdm,
    glue("DELETE FROM model       WHERE ds_key = 'ms_endemism'")
  )
  dbExecute(
    con_sdm,
    glue("DELETE FROM species     WHERE ds_key = 'ms_endemism'")
  )
}

for (i in seq_len(nrow(d_taxa))) {
  # i = 1
  d_sp <- d_taxa[i, ]

  # get this species' input models (excluding derived layers)
  d_sp_l <- tbl(con_sdm, "taxon_model") |>
    filter(
      taxon_id == !!d_sp$taxon_id,
      ds_key %in% ds_keys_input
    ) |>
    collect()

  if (nrow(d_sp_l) == 0) {
    next
  }

  # determine the mask: if has rng_iucn, use union of is_mask layers;
  # otherwise use union of ALL distributions
  has_iucn <- "rng_iucn" %in% d_sp_l$ds_key
  if (has_iucn) {
    mask_mdl_seqs <- d_sp_l |>
      filter(ds_key %in% ds_keys_mask) |>
      pull(mdl_seq)
  } else {
    mask_mdl_seqs <- d_sp_l$mdl_seq
  }

  # count cells in program areas
  q_sp_pra <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq %in% mask_mdl_seqs) |>
    distinct(cell_id) |>
    semi_join(q_pra_cells, by = "cell_id")

  n_sp_cells <- q_sp_pra |> collect() |> nrow()

  if (n_sp_cells == 0) {
    next
  }

  # log-scaled endemism: 10 (widespread) to 90 (endemic)
  endemism <- round(pmax(
    10,
    pmin(90, 10 + 80 * log(n_total_cells / n_sp_cells) / log(n_total_cells))
  ))

  # get all cells in the species mask (within program areas)
  mask_cell_ids <- q_sp_pra |> pull(cell_id)

  sp_key <- glue("ms_endemism:{d_sp$taxon_id}")

  if (i %% 100 == 0 || i == 1) {
    message(glue(
      "ms_endemism {i}/{nrow(d_taxa)}: {d_sp$scientific_name} ",
      "({n_sp_cells} cells, endemism={endemism})"
    ))
  }

  # insert model row
  dbWriteTable(
    con_sdm,
    "model",
    tibble(
      ds_key = "ms_endemism",
      taxa = sp_key,
      time_period = "2025",
      region = "USA",
      mdl_type = "derived",
      description = glue(
        "Endemism score {endemism} for {d_sp$scientific_name} ",
        "({n_sp_cells}/{n_total_cells} PA cells)"
      )
    ),
    append = TRUE
  )

  mdl_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT mdl_seq FROM model
       WHERE ds_key = 'ms_endemism' AND taxa = '{sp_key}'
       ORDER BY mdl_seq DESC LIMIT 1"
    )
  )$mdl_seq

  # insert model_cell rows (all mask cells get same endemism value)
  dbWriteTable(
    con_sdm,
    "model_cell",
    tibble(
      mdl_seq = as.integer(mdl_seq),
      cell_id = as.integer(mask_cell_ids),
      value = as.integer(endemism)
    ),
    append = TRUE
  )

  # insert taxon_model junction
  dbWriteTable(
    con_sdm,
    "taxon_model",
    tibble(
      taxon_id = d_sp$taxon_id,
      ds_key = "ms_endemism",
      mdl_seq = as.integer(mdl_seq)
    ),
    append = TRUE
  )
}

message("ms_endemism complete")
```

### Create Merged Suitability (`ms_suit`)

Per species, per cell: `ms_suit = max(all is_suit layers for that cell)`.
Species with AquaMaps + ms_endemism get `max(AquaMaps, endemism)` per cell.
Species with only ms_endemism get their endemism score.
Masked to IUCN range when available.

```{r}
#| label: create_ms_suit
#| eval: !expr do_merge

ds_keys_suit <- tbl(con_sdm, "dataset") |>
  filter(is_suit) |>
  pull(ds_key)
# c("am_0.05", "ms_endemism")

# clean up existing
existing_ms <- tbl(con_sdm, "model") |>
  filter(ds_key == "ms_suit") |>
  pull(mdl_seq)
if (length(existing_ms) > 0) {
  ms_str <- paste(existing_ms, collapse = ",")
  dbExecute(
    con_sdm,
    glue("DELETE FROM model_cell  WHERE mdl_seq IN ({ms_str})")
  )
  dbExecute(con_sdm, glue("DELETE FROM taxon_model WHERE ds_key = 'ms_suit'"))
  dbExecute(con_sdm, glue("DELETE FROM model       WHERE ds_key = 'ms_suit'"))
  dbExecute(con_sdm, glue("DELETE FROM species     WHERE ds_key = 'ms_suit'"))
}

for (i in seq_len(nrow(d_taxa))) {
  d_sp <- d_taxa[i, ]

  # get suitability layers for this species
  d_sp_suit <- tbl(con_sdm, "taxon_model") |>
    filter(
      taxon_id == !!d_sp$taxon_id,
      ds_key %in% ds_keys_suit
    ) |>
    collect()

  if (nrow(d_sp_suit) == 0) {
    next
  }

  suit_mdl_seqs <- d_sp_suit$mdl_seq

  # per-cell max across suitability layers
  q_suit <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq %in% suit_mdl_seqs) |>
    group_by(cell_id) |>
    summarize(value = max(value, na.rm = TRUE), .groups = "drop")

  # apply IUCN mask if available
  d_sp_all <- tbl(con_sdm, "taxon_model") |>
    filter(
      taxon_id == !!d_sp$taxon_id,
      ds_key %in% ds_keys_input
    ) |>
    collect()

  has_iucn <- "rng_iucn" %in% d_sp_all$ds_key
  if (has_iucn) {
    mask_mdl_seqs <- d_sp_all |>
      filter(ds_key %in% ds_keys_mask) |>
      pull(mdl_seq)
    q_mask <- tbl(con_sdm, "model_cell") |>
      filter(mdl_seq %in% mask_mdl_seqs) |>
      distinct(cell_id)
    d_r_sp <- q_suit |>
      semi_join(q_mask, by = "cell_id") |>
      collect()
  } else {
    d_r_sp <- q_suit |> collect()
  }

  if (nrow(d_r_sp) == 0) {
    next
  }

  # clamp to 1-100
  d_r_sp <- d_r_sp |>
    mutate(value = pmax(1L, pmin(100L, as.integer(value))))

  sp_key <- glue("ms_suit:{d_sp$taxon_id}")

  if (i %% 100 == 0 || i == 1) {
    message(glue(
      "ms_suit {i}/{nrow(d_taxa)}: {d_sp$scientific_name} ",
      "({nrow(d_r_sp)} cells)"
    ))
  }

  suit_ds_str <- paste(d_sp_suit$ds_key, collapse = ", ")
  mask_str <- if (has_iucn) "; IUCN masked" else ""

  dbWriteTable(
    con_sdm,
    "model",
    tibble(
      ds_key = "ms_suit",
      taxa = sp_key,
      time_period = "2025",
      region = "USA",
      mdl_type = "derived",
      description = glue(
        "Merged suitability (max of: {suit_ds_str}{mask_str})"
      )
    ),
    append = TRUE
  )

  mdl_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT mdl_seq FROM model
       WHERE ds_key = 'ms_suit' AND taxa = '{sp_key}'
       ORDER BY mdl_seq DESC LIMIT 1"
    )
  )$mdl_seq

  dbWriteTable(
    con_sdm,
    "model_cell",
    tibble(
      mdl_seq = as.integer(mdl_seq),
      cell_id = as.integer(d_r_sp$cell_id),
      value = as.integer(d_r_sp$value)
    ),
    append = TRUE
  )

  dbWriteTable(
    con_sdm,
    "taxon_model",
    tibble(
      taxon_id = d_sp$taxon_id,
      ds_key = "ms_suit",
      mdl_seq = as.integer(mdl_seq)
    ),
    append = TRUE
  )
}

message("ms_suit complete")
```

### Create Merged Extinction Risk (`ms_er`)

Per species, per cell: take value from highest-priority (lowest `sort_order`)
ER layer. Cells in distribution but without ER layer coverage get the
species-level max ER value across all its ER layers (or the scalar
`er_score` from `taxon` if no ER layers exist).

Priority ordering (US > International, NMFS > FWS > IUCN):
ca_nmfs (2) > ch_nmfs (3) > ch_fws (4) > rng_fws (5) > bl (6) > rng_iucn (7)

```{r}
#| label: create_ms_er
#| eval: !expr do_merge

ds_er_priority <- tbl(con_sdm, "dataset") |>
  filter(is_er) |>
  arrange(sort_order) |>
  collect()
# ca_nmfs, ch_nmfs, ch_fws, rng_fws, bl, rng_iucn

# clean up existing
existing_ms <- tbl(con_sdm, "model") |>
  filter(ds_key == "ms_er") |>
  pull(mdl_seq)
if (length(existing_ms) > 0) {
  ms_str <- paste(existing_ms, collapse = ",")
  dbExecute(
    con_sdm,
    glue("DELETE FROM model_cell  WHERE mdl_seq IN ({ms_str})")
  )
  dbExecute(con_sdm, glue("DELETE FROM taxon_model WHERE ds_key = 'ms_er'"))
  dbExecute(con_sdm, glue("DELETE FROM model       WHERE ds_key = 'ms_er'"))
  dbExecute(con_sdm, glue("DELETE FROM species     WHERE ds_key = 'ms_er'"))
}

for (i in seq_len(nrow(d_taxa))) {
  d_sp <- d_taxa[i, ]

  # get ER layers for this species, ordered by priority
  d_sp_er <- tbl(con_sdm, "taxon_model") |>
    filter(
      taxon_id == !!d_sp$taxon_id,
      ds_key %in% !!ds_er_priority$ds_key
    ) |>
    collect() |>
    left_join(
      ds_er_priority |> select(ds_key, sort_order),
      by = "ds_key"
    ) |>
    arrange(sort_order)

  # get all input models for mask determination
  d_sp_all <- tbl(con_sdm, "taxon_model") |>
    filter(
      taxon_id == !!d_sp$taxon_id,
      ds_key %in% ds_keys_input
    ) |>
    collect()

  has_iucn <- "rng_iucn" %in% d_sp_all$ds_key

  # determine the species distribution mask
  if (has_iucn) {
    mask_mdl_seqs <- d_sp_all |>
      filter(ds_key %in% ds_keys_mask) |>
      pull(mdl_seq)
  } else {
    mask_mdl_seqs <- d_sp_all$mdl_seq
  }

  mask_cells <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq %in% mask_mdl_seqs) |>
    distinct(cell_id) |>
    pull(cell_id)

  if (length(mask_cells) == 0) {
    next
  }

  if (nrow(d_sp_er) > 0) {
    # build per-cell ER values using priority ordering:
    # for each cell, take value from highest-priority ER layer present
    d_er_cells <- NULL
    for (j in seq_len(nrow(d_sp_er))) {
      d_layer <- tbl(con_sdm, "model_cell") |>
        filter(mdl_seq == !!d_sp_er$mdl_seq[j]) |>
        select(cell_id, value) |>
        collect()

      if (is.null(d_er_cells)) {
        d_er_cells <- d_layer
      } else {
        # only add cells not already covered by higher-priority layer
        new_cells <- d_layer |>
          filter(!cell_id %in% d_er_cells$cell_id)
        d_er_cells <- bind_rows(d_er_cells, new_cells)
      }
    }

    # restrict to mask
    d_er_cells <- d_er_cells |>
      filter(cell_id %in% mask_cells)

    # cells in mask but not covered by any ER layer:
    # assign max ER value across all ER layers for this species
    uncovered_cells <- setdiff(mask_cells, d_er_cells$cell_id)
    if (length(uncovered_cells) > 0) {
      max_er_value <- max(d_er_cells$value, na.rm = TRUE)
      d_er_cells <- bind_rows(
        d_er_cells,
        tibble(
          cell_id = as.integer(uncovered_cells),
          value = as.integer(max_er_value)
        )
      )
    }
  } else {
    # no ER layers at all: use scalar er_score from taxon table
    fallback_er <- ifelse(
      is.na(d_sp$er_score),
      1L,
      as.integer(d_sp$er_score)
    )
    d_er_cells <- tibble(
      cell_id = as.integer(mask_cells),
      value = fallback_er
    )
  }

  # clamp to 1-100
  d_er_cells <- d_er_cells |>
    mutate(value = pmax(1L, pmin(100L, as.integer(value))))

  sp_key <- glue("ms_er:{d_sp$taxon_id}")

  if (i %% 100 == 0 || i == 1) {
    message(glue(
      "ms_er {i}/{nrow(d_taxa)}: {d_sp$scientific_name} ",
      "({nrow(d_er_cells)} cells)"
    ))
  }

  er_ds_str <- paste(d_sp_er$ds_key, collapse = ", ")
  if (nrow(d_sp_er) == 0) {
    er_ds_str <- glue("scalar er_score={d_sp$er_score}")
  }

  dbWriteTable(
    con_sdm,
    "model",
    tibble(
      ds_key = "ms_er",
      taxa = sp_key,
      time_period = "2025",
      region = "USA",
      mdl_type = "derived",
      description = glue(
        "Merged extinction risk (priority: {er_ds_str})"
      )
    ),
    append = TRUE
  )

  mdl_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT mdl_seq FROM model
       WHERE ds_key = 'ms_er' AND taxa = '{sp_key}'
       ORDER BY mdl_seq DESC LIMIT 1"
    )
  )$mdl_seq

  dbWriteTable(
    con_sdm,
    "model_cell",
    tibble(
      mdl_seq = as.integer(mdl_seq),
      cell_id = as.integer(d_er_cells$cell_id),
      value = as.integer(d_er_cells$value)
    ),
    append = TRUE
  )

  dbWriteTable(
    con_sdm,
    "taxon_model",
    tibble(
      taxon_id = d_sp$taxon_id,
      ds_key = "ms_er",
      mdl_seq = as.integer(mdl_seq)
    ),
    append = TRUE
  )
}

message("ms_er complete")
```

### Create Final Merged Model (`ms_merge`)

The final merged model combines suitability and extinction risk:
`ms_merge = max(1, round(ms_suit * ms_er / 100))`

The floor of 1 ensures species with low suitability x low ER (e.g.,
1% x 1% = 0.01%) still appear on the map rather than rounding to 0
and disappearing. Since ms_merge already incorporates ER per cell,
the downstream scoring in `calc_scores.qmd` no longer needs to
multiply by a scalar `er_score`.

```{r}
#| label: create_ms_merge
#| eval: !expr do_merge

# clean up existing ms_merge models
existing_ms <- tbl(con_sdm, "model") |>
  filter(ds_key == "ms_merge") |>
  pull(mdl_seq)
if (length(existing_ms) > 0) {
  ms_str <- paste(existing_ms, collapse = ",")
  dbExecute(
    con_sdm,
    glue("DELETE FROM model_cell  WHERE mdl_seq IN ({ms_str})")
  )
  dbExecute(con_sdm, glue("DELETE FROM taxon_model WHERE ds_key = 'ms_merge'"))
  dbExecute(con_sdm, glue("DELETE FROM model       WHERE ds_key = 'ms_merge'"))
  dbExecute(con_sdm, glue("DELETE FROM species     WHERE ds_key = 'ms_merge'"))
  dbExecute(
    con_sdm,
    glue(
      "UPDATE taxon SET mdl_seq = NULL WHERE mdl_seq IN ({ms_str})"
    )
  )
}

for (i in seq_len(nrow(d_taxa))) {
  d_sp <- d_taxa[i, ]

  # get ms_suit and ms_er models for this species
  d_sp_derived <- tbl(con_sdm, "taxon_model") |>
    filter(
      taxon_id == !!d_sp$taxon_id,
      ds_key %in% c("ms_suit", "ms_er")
    ) |>
    collect()

  suit_mdl <- d_sp_derived |> filter(ds_key == "ms_suit")
  er_mdl <- d_sp_derived |> filter(ds_key == "ms_er")

  if (nrow(suit_mdl) == 0 || nrow(er_mdl) == 0) {
    next
  }

  # get per-cell values from both layers
  d_suit <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq == !!suit_mdl$mdl_seq) |>
    select(cell_id, suit = value) |>
    collect()

  d_er <- tbl(con_sdm, "model_cell") |>
    filter(mdl_seq == !!er_mdl$mdl_seq) |>
    select(cell_id, er = value) |>
    collect()

  # join and compute product
  d_merge <- d_suit |>
    inner_join(d_er, by = "cell_id") |>
    mutate(
      value = pmax(1L, pmin(100L, as.integer(round(suit * er / 100))))
    ) |>
    select(cell_id, value)

  if (nrow(d_merge) == 0) {
    next
  }

  sp_key <- glue("ms_merge:{d_sp$taxon_id}")

  if (i %% 100 == 0 || i == 1) {
    message(glue(
      "ms_merge {i}/{nrow(d_taxa)}: {d_sp$scientific_name} ",
      "({nrow(d_merge)} cells)"
    ))
  }

  # insert model
  dbWriteTable(
    con_sdm,
    "model",
    tibble(
      ds_key = "ms_merge",
      taxa = sp_key,
      time_period = "2025",
      region = "USA",
      mdl_type = "derived",
      description = glue(
        "Final merged model: max(1, round(ms_suit * ms_er / 100))"
      )
    ),
    append = TRUE
  )

  mdl_seq <- dbGetQuery(
    con_sdm,
    glue(
      "SELECT mdl_seq FROM model
       WHERE ds_key = 'ms_merge' AND taxa = '{sp_key}'
       ORDER BY mdl_seq DESC LIMIT 1"
    )
  )$mdl_seq

  # insert model_cell
  dbWriteTable(
    con_sdm,
    "model_cell",
    tibble(
      mdl_seq = as.integer(mdl_seq),
      cell_id = as.integer(d_merge$cell_id),
      value = as.integer(d_merge$value)
    ),
    append = TRUE
  )

  # insert species row
  sp_sci <- d_sp$scientific_name
  sp_cmn <- d_sp$common_name
  d_species <- tibble(
    ds_key = "ms_merge",
    taxa = sp_key,
    sp_key = sp_key,
    worms_id = ifelse(
      d_sp$taxon_authority == "worms",
      d_sp$taxon_id,
      NA_integer_
    ),
    botw_id = ifelse(
      d_sp$taxon_authority == "botw",
      d_sp$taxon_id,
      NA_integer_
    ),
    gbif_id = NA_integer_,
    itis_id = NA_integer_,
    iucn_id = NA_integer_,
    scientific_name_dataset = sp_sci,
    common_name_dataset = sp_cmn,
    scientific_name_accepted = sp_sci,
    common_name_accepted = sp_cmn,
    redlist_code = d_sp$redlist_code,
    redlist_year = NA_integer_,
    sp_cat = d_sp$sp_cat,
    worms_is_marine = d_sp$worms_is_marine,
    worms_is_extinct = d_sp$worms_is_extinct
  )
  dbWriteTable(con_sdm, "species", d_species, append = TRUE)

  # update taxon with mdl_seq
  dbExecute(
    con_sdm,
    glue(
      "UPDATE taxon SET mdl_seq = {mdl_seq}
       WHERE taxon_id = {d_sp$taxon_id}"
    )
  )

  # insert taxon_model junction
  dbWriteTable(
    con_sdm,
    "taxon_model",
    tibble(
      taxon_id = d_sp$taxon_id,
      ds_key = "ms_merge",
      mdl_seq = as.integer(mdl_seq)
    ),
    append = TRUE
  )
}

message("ms_merge complete")
```

## Set taxon.is_ok

Add `is_ok`: a simple logical field for flagging valid taxa, ie (so far):

- birds:
  - [x] add worms_id (if available)
  - `redlist_code` != "EX"
  - has a `botw_id`
  - if has worms_id:
    - `worms_is_marine`  != F
    - `worms_is_extinct` != T
- no model cells overlap with Program Areas
- category is "reptile" but not "turtle" (handled by reclassify_reptiles)
- not birds:
  - has a `worms_id`
  - `worms_is_marine`  != F
  - `worms_is_extinct` != T

### Flag valid taxa

```{r}
#| label: taxon_is_ok
#| eval: !expr do_is_ok

d <- tbl(con_sdm, "taxon") |>
  collect() # 17,561 × 16

# birds ----
# d |>
#   filter(
#     taxon_authority == "botw") |>
#   pull(redlist_code) |>
#   table(useNA = "ifany")
#  CR  DD  EN  LC  NT  TN  VU
#   3   1  58 452  45  14  41
#
# d |>
#   filter(
#     taxon_authority == "botw") |>
#   select(worms_is_marine, worms_is_extinct) |>
#   table(useNA = "ifany")
#                worms_is_extinct
# worms_is_marine <NA>
#           FALSE  118
#           TRUE   221
#           <NA>   275

d_b <- d |>
  filter(
    taxon_authority == "botw"
  ) |>
  mutate(
    is_ok = case_when(
      is.na(taxon_id) ~ F,
      is.na(mdl_seq) ~ F,
      !is.na(redlist_code) & redlist_code == "EX" ~ F, # 0
      !is.na(worms_id) & worms_is_marine == F ~ F, # 118
      !is.na(worms_id) & worms_is_extinct == T ~ F, # 0
      .default = T
    )
  )
# d_b$is_ok |> table(useNA = "ifany")
# FALSE  TRUE
#   118   496

# worms ----

# d |>
#   filter(
#     taxon_authority == "worms") |>
#   pull(redlist_code) |>
#   table(useNA = "ifany")
# CR    DD    EN    EX    LC    NT    TN    VU  <NA>
# 47   340   235     1  5574   107    11   154 10478
#
# d |>
#   filter(
#     taxon_authority == "worms") |>
#   select(worms_is_marine, worms_is_extinct) |>
#   table(useNA = "ifany")
#                worms_is_extinct
# worms_is_marine FALSE  TRUE  <NA>
#           FALSE    19     0    24
#           TRUE   4171    20 12708
#           <NA>      0     0     5

d_w <- d |>
  filter(
    taxon_authority == "worms"
  ) |>
  mutate(
    is_ok = case_when(
      is.na(taxon_id) ~ F,
      is.na(mdl_seq) ~ F,
      !is.na(redlist_code) & redlist_code == "EX" ~ F, # 1
      !is.na(worms_id) & worms_is_marine == F ~ F, # 19 + 24
      !is.na(worms_id) & worms_is_extinct == T ~ F, # 20
      sp_cat == "reptile" ~ F,
      sp_cat == "turtle" ~ T,
      .default = T
    )
  )
# d_w$is_ok |> table(useNA = "ifany")
# FALSE   TRUE
#    64 16,883

d2 <- bind_rows(
  d_b,
  d_w
) |>
  select(taxon_id, is_ok)

# flag taxa with no distribution inside program areas ----
# (mirrors r_mask approach in apps_2026/mapsp/app.R but with DB queries)
taxa_in_pra <- tbl(con_sdm, "taxon_model") |>
  inner_join(
    tbl(con_sdm, "model_cell") |> select(mdl_seq, cell_id),
    by = "mdl_seq"
  ) |>
  inner_join(
    tbl(con_sdm, "zone_cell") |> select(zone_seq, cell_id),
    by = "cell_id"
  ) |>
  inner_join(
    tbl(con_sdm, "zone") |>
      filter(
        tbl == !!glue("ply_programareas_2026{v_sfx}"),
        fld == "programarea_key"
      ) |>
      select(zone_seq),
    by = "zone_seq"
  ) |>
  distinct(taxon_id) |>
  pull(taxon_id)

n_outside <- sum(d2$is_ok & !(d2$taxon_id %in% taxa_in_pra))
message(glue(
  "{n_outside} taxa flagged is_ok=F (no distribution in program areas)"
))

d2 <- d2 |>
  mutate(
    is_ok = is_ok & taxon_id %in% taxa_in_pra
  )

stopifnot(sum(duplicated(d2$taxon_id)) == 0)

dbExecute(
  con_sdm,
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS is_ok BOOLEAN"
)

duckdb_register(con_sdm, "d2", d2)
dbExecute(
  con_sdm,
  "UPDATE taxon
    SET is_ok  = d2.is_ok
    FROM d2
    WHERE taxon.taxon_id = d2.taxon_id"
) # 17,561
duckdb_unregister(con_sdm, "d2")
```

## Taxon Summary

### Export taxon summary

```{r}
#| label: taxon_summary
taxon_csv <- here("data/taxon.csv")

# dbListFields(con_sdm, "taxon") |> paste(collapse = ", ") |> cat()
d_taxon <- tbl(con_sdm, "taxon") |>
  select(
    is_ok,
    component = sp_cat,
    common_name,
    scientific_name,
    redlist_code_max = redlist_code,
    extrisk_code,
    er_score,
    is_mmpa,
    is_mbta,
    is_bcc,
    worms_is_marine,
    worms_is_extinct,
    n_datasets = n_ds,
    taxon_authority,
    taxon_id,
    model_id = mdl_seq
  ) |>
  arrange(desc(is_ok), component, common_name) |>
  collect()

write_csv(d_taxon, taxon_csv)
```

## Dataset Summary

```{r}
#| label: dataset_overview
tbl(con_sdm, "dataset") |>
  select(sort_order, ds_key, name_display, value_info, is_mask) |>
  arrange(sort_order) |>
  collect() |>
  DT::datatable()
```

```{r}
#| label: model_counts_by_ds
tbl(con_sdm, "model") |>
  group_by(ds_key) |>
  summarize(n_models = n()) |>
  arrange(ds_key) |>
  collect() |>
  DT::datatable()
```

## Species Summary

### Species counts by sp_cat and ds_key

```{r}
#| label: spp_counts_by_cat_ds
tbl(con_sdm, "species") |>
  group_by(sp_cat, ds_key) |>
  summarize(n = n(), .groups = "drop") |>
  arrange(sp_cat, ds_key) |>
  collect() |>
  DT::datatable()
```

### Taxonomic authority coverage

```{r}
#| label: auth_coverage
tbl(con_sdm, "species") |>
  group_by(sp_cat, ds_key) |>
  summarize(
    n = n(),
    n_worms = sum(!is.na(worms_id), na.rm = T),
    n_itis = sum(!is.na(itis_id), na.rm = T),
    n_gbif = sum(!is.na(gbif_id), na.rm = T),
    n_allna = sum(
      is.na(worms_id) & is.na(itis_id) & is.na(gbif_id),
      na.rm = T
    ),
    .groups = "drop"
  ) |>
  mutate(
    pct_worms = round(n_worms / n * 100, 1),
    pct_itis = round(n_itis / n * 100, 1),
    pct_gbif = round(n_gbif / n * 100, 1),
    pct_allna = round(n_allna / n * 100, 1)
  ) |>
  arrange(sp_cat, ds_key) |>
  collect() |>
  DT::datatable()
```

### WoRMS marine/extinct percentages

```{r}
#| label: worms_marine_extinct
tbl(con_sdm, "species") |>
  group_by(sp_cat) |>
  summarize(
    n = n(),
    n_worms_marine = sum(worms_is_marine, na.rm = T),
    n_worms_extinct = sum(worms_is_extinct, na.rm = T)
  ) |>
  mutate(
    pct_worms_marine = round(n_worms_marine / n * 100, 1),
    pct_worms_extinct = round(n_worms_extinct / n * 100, 1)
  ) |>
  collect() |>
  DT::datatable()
```

## Taxon Table Summary

### Taxon counts by sp_cat and is_ok

```{r}
#| label: taxon_counts_by_cat_ok
tbl(con_sdm, "taxon") |>
  group_by(sp_cat, is_ok) |>
  summarize(n = n(), .groups = "drop") |>
  collect() |>
  pivot_wider(
    names_from = is_ok,
    values_from = n,
    names_prefix = "is_ok_"
  ) |>
  DT::datatable()
```

### Redlist code distribution

```{r}
#| label: redlist_distribution
tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  group_by(sp_cat, redlist_code) |>
  summarize(n = n(), .groups = "drop") |>
  collect() |>
  pivot_wider(
    names_from = redlist_code,
    values_from = n,
    values_fill = 0
  ) |>
  DT::datatable()
```

### Extinction risk authority summary

```{r}
#| label: extrisk_authority
tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  mutate(
    authority = case_when(
      str_starts(extrisk_code, "NMFS") ~ "NMFS",
      str_starts(extrisk_code, "FWS") ~ "FWS",
      str_starts(extrisk_code, "IUCN") ~ "IUCN",
      TRUE ~ "none"
    )
  ) |>
  count(authority) |>
  collect() |>
  DT::datatable()
```

### Number of datasets per taxon

```{r}
#| label: n_ds_distribution
tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  count(n_ds) |>
  collect() |>
  DT::datatable()
```

## Taxon x Dataset Matrix

```{r}
#| label: taxon_dataset_matrix
tbl(con_sdm, "taxon_model") |>
  inner_join(
    tbl(con_sdm, "taxon") |>
      filter(is_ok) |>
      select(taxon_id, sp_cat),
    by = "taxon_id"
  ) |>
  group_by(sp_cat, ds_key) |>
  summarize(n = n(), .groups = "drop") |>
  collect() |>
  pivot_wider(
    names_from = ds_key,
    values_from = n,
    values_fill = 0
  ) |>
  DT::datatable()
```

## Taxon Detail Table

```{r}
#| label: taxon_detail
tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  select(
    sp_cat,
    scientific_name,
    common_name,
    redlist_code,
    extrisk_code,
    er_score,
    n_ds,
    taxon_id,
    taxon_authority
  ) |>
  arrange(sp_cat, scientific_name) |>
  collect() |>
  DT::datatable(
    filter = "top",
    options = list(pageLength = 25)
  )
```
