---
title: "Merge taxon metrics — validity flags + range / rarity / endemism / pct_marine"
msens:
  target_name: merge_taxon
  workflow_type: merge
  dependency: [merge_models, build_cell_grid]
  output: data/manifests/merge_taxon.json
editor_options:
  chunk_output_type: console
---

Per-taxon attributes on the merged `model_cell`, joined to the `cell` table (ocean
cells with `area_km2` / `in_usa` / `in_pra`):

- **3 nested validity flags** (`is_valid_global ⊇ is_valid_usa ⊇ is_valid_pra`):
  a taxon is *valid_global* if it has a mapped distribution reaching the ocean, isn't
  extinct, and isn't flagged non-marine (`worms_is_marine`); *valid_usa* / *valid_pra*
  add "has cells in the US study area / Program Areas".
- **range / rarity / endemism / pct_marine**: `range_km2` = Σ `area_km2` over the
  taxon's ocean cells (v7-comparable marine range); `pct_marine` = ocean-cell share of
  the whole (land+ocean) range; `us_endemism` = US share of the marine range; rarity
  class from the marine `range_km2` distribution.
- **`is_marine`** (marine-relevance filter): keeps genuinely marine/coastal **birds** in the
  score and culls terrestrial ones (v8 captures each bird's whole, largely-terrestrial range).

## Setup + compute

```{r}
#| label: compute

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, readr, stringr,
                 tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))
source(here("libs/vars.R"))

dir_atlas <- glue("{dir_big_v}/marine-atlas")
merged    <- glue("{dir_atlas}/dist_merged/dataset=ms_merge")
dir_dist  <- glue("{dir_atlas}/dist")   # raw per-dataset (global) ranges, for bird pct_marine
merge_db  <- glue("{dir_atlas}/merge.duckdb")
manifest  <- here("data/manifests/merge_taxon.json")
dir_create(path_dir(manifest))
stopifnot("run merge_models first"    = dir_exists(merged),
          "run build_cell_grid first" = file_exists(sdm_db))

con <- dbConnect(duckdb(merge_db))
dbExecute(con, glue("ATTACH '{sdm_db}' AS grid (READ_ONLY)"))
dbExecute(con, glue("ATTACH '{spp_db}' AS spp (READ_ONLY)"))   # WoRMS class/phylum for sp_cat

# per-taxon cell tallies against the ocean cell table
dbExecute(con, glue("CREATE OR REPLACE TABLE taxon_cell AS
  WITH mc AS (SELECT mdl_key, cell_id FROM read_parquet('{merged}/**/*.parquet')),
  j AS (SELECT mc.mdl_key, c.area_km2, c.in_usa, c.in_pra,
               (c.cell_id IS NOT NULL) AS is_ocean
        FROM mc LEFT JOIN grid.cell c USING (cell_id))
  SELECT mdl_key,
    count(*)                              AS n_cells,
    count(*) FILTER (WHERE is_ocean)      AS n_ocean,
    count(*) FILTER (WHERE in_usa)        AS n_usa,
    count(*) FILTER (WHERE in_pra)        AS n_pra,
    round(sum(area_km2) FILTER (WHERE is_ocean), 1) AS range_km2,
    round(sum(area_km2) FILTER (WHERE in_usa),   1) AS range_usa_km2
  FROM j GROUP BY mdl_key"))
```

## Flags + range/rarity/endemism → taxon

```{r}
#| label: flags

# idempotent: drop any derived columns a prior run added so `x.*` stays base taxon (+ er)
for (col in c("n_cells","n_ocean","n_usa","n_pra","range_km2","range_usa_km2",
              "is_valid_global","is_valid_usa","is_valid_pra","pct_marine","us_endemism",
              "rarity","is_marine","pct_marine_bl","sp_cat"))
  try(dbExecute(con, glue("ALTER TABLE taxon DROP COLUMN IF EXISTS {col}")), silent = TRUE)

# rarity thresholds on the marine range (km2); tune vs v7
dbExecute(con, "CREATE OR REPLACE TABLE taxon AS
  WITH t AS (
    SELECT x.*, tc.n_cells, tc.n_ocean, tc.n_usa, tc.n_pra, tc.range_km2, tc.range_usa_km2
    FROM taxon x LEFT JOIN taxon_cell tc ON x.ms_merge_key = tc.mdl_key)
  SELECT *,
    -- 3 nested validity flags (global superset of usa superset of pra)
    (n_ocean > 0 AND coalesce(worms_is_extinct, FALSE) = FALSE
       AND coalesce(worms_is_marine, TRUE) = TRUE)        AS is_valid_global,
    (n_usa > 0)                                            AS is_valid_usa,
    (n_pra > 0)                                            AS is_valid_pra,
    round(100.0 * n_ocean / nullif(n_cells, 0), 1)         AS pct_marine,
    round(100.0 * n_usa   / nullif(n_ocean, 0), 1)         AS us_endemism,
    CASE
      WHEN range_km2 IS NULL       THEN NULL
      WHEN range_km2 <     10000   THEN 'very_rare'
      WHEN range_km2 <    100000   THEN 'rare'
      WHEN range_km2 <   1000000   THEN 'uncommon'
      WHEN range_km2 <  10000000   THEN 'common'
      ELSE                              'widespread' END   AS rarity
  FROM t")

smry <- dbGetQuery(con, "SELECT
  count(*) n_taxa,
  count(*) FILTER (WHERE is_valid_global) valid_global,
  count(*) FILTER (WHERE is_valid_usa)    valid_usa,
  count(*) FILTER (WHERE is_valid_pra)    valid_pra,
  round(median(pct_marine),1) med_pct_marine FROM taxon")
knitr::kable(smry)
```

## Species category (`sp_cat`) by taxonomy

`sp_cat` is assigned by **taxonomy** (WoRMS class/phylum/kingdom + BirdLife→bird), not a v7
crosswalk — so new taxa land in the right class instead of a catch-all `other` (eliminated).
v7's six scored classes (bird, mammal, turtle, fish, coral, invertebrate) **plus
`primary_producer`** (marine plants/algae/diatoms — distinct from the carbon `primprod`
metric). **Reptiles/amphibians are categorized but EXCLUDED from scoring** (sea turtles stay
`turtle`; `score_cell_metrics` scores only the seven classes). AphiaIDs missing from the
local WoRMS snapshot (mostly algae/seagrass) fall back to `data/worms_taxonomy_supplement.csv`.

```{r}
#| label: sp_cat

supp <- read_csv(here("data/worms_taxonomy_supplement.csv"), show_col_types = FALSE) |>
  transmute(taxon_id = as.character(taxon_id), s_kingdom = kingdom, s_phylum = phylum, s_class = class)
turtle_keys <- dbGetQuery(con,
  "SELECT DISTINCT ms_merge_key FROM taxon_model WHERE mdl_key LIKE 'rng_turtle_swot_dps|%'")$ms_merge_key

tx <- dbGetQuery(con, "
  SELECT t.ms_merge_key, t.taxon_authority AS auth, t.taxon_id,
         w.kingdom, w.phylum, w.\"class\" AS cls
  FROM taxon t
  LEFT JOIN spp.worms w ON t.taxon_authority='worms' AND CAST(w.scientificNameID AS VARCHAR)=t.taxon_id") |>
  left_join(supp, by = "taxon_id") |>
  mutate(kingdom = coalesce(kingdom, s_kingdom),
         phylum  = coalesce(phylum,  s_phylum),
         cls     = coalesce(cls,     s_class))

fish_cls    <- c("Teleostei","Elasmobranchii","Holocephali","Myxini","Chondrostei","Petromyzonti",
                 "Actinopteri","Actinopterygii","Chondrichthyes","Coelacanthi","Dipneusti","Cladistia","Sarcopterygii")
reptile_cls <- c("Reptilia","Crocodylia","Squamata","Testudines","Lepidosauria","Sauropsida","Archosauria")
tx <- tx |> mutate(sp_cat = case_when(
  ms_merge_key %in% turtle_keys                    ~ "turtle",
  auth == "botw" | cls == "Aves"                   ~ "bird",
  cls == "Mammalia"                                ~ "mammal",
  cls %in% reptile_cls                             ~ "reptile",        # excluded from scoring
  cls == "Amphibia"                                ~ "amphibian",      # excluded from scoring
  cls %in% fish_cls                                ~ "fish",
  cls %in% c("Hexacorallia","Octocorallia")        ~ "coral",
  kingdom %in% c("Plantae","Chromista","Bacteria") ~ "primary_producer",
  TRUE                                             ~ "invertebrate"))

duckdb_register(con, "tmp_spcat", tx |> select(ms_merge_key, sp_cat))
dbExecute(con, "CREATE OR REPLACE TABLE taxon AS
  SELECT t.*, s.sp_cat FROM taxon t LEFT JOIN tmp_spcat s USING (ms_merge_key)")
duckdb_unregister(con, "tmp_spcat")
knitr::kable(dbGetQuery(con, "SELECT sp_cat, count(*) n, count(*) FILTER(WHERE is_valid_usa) valid_usa
  FROM taxon GROUP BY 1 ORDER BY n DESC"))
```

## Marine relevance (birds) — `is_marine`

A bird's whole global range is largely terrestrial (v8 keeps the entire range, no land
mask), so only genuinely **marine/coastal** birds should drive the marine-sensitivity
score. `is_marine` (bird) = member of a marine/coastal **family**
(`data/marine_bird_families.csv`) **and** whole-range `pct_marine ≥ 5%`, **or** on the
curated **include** list — minus the curated **exclude** list
(`data/marine_birds_curation.csv`); both are GitHub-tracked + reviewable. The share is
computed from the **raw global** `bl` ranges (the merged `model_cell` is US-scoped, so its
`pct_marine` is ~100% and unusable here). Non-bird taxa are marine by dataset → `TRUE`.

```{r}
#| label: marine_relevance

marine_fam <- read_csv(here("data/marine_bird_families.csv"), show_col_types = FALSE)$family
cur     <- read_csv(here("data/marine_birds_curation.csv"), show_col_types = FALSE)
inc_sci <- clean_sci_name(cur$scientific_name[cur$action == "include"])
exc_sci <- clean_sci_name(cur$scientific_name[cur$action == "exclude"])

# whole-range ocean share per bird from the RAW global bl ranges ∩ ocean cell table
bl_pm <- dbGetQuery(con, glue("
  WITH b AS (SELECT mc.mdl_key, (c.cell_id IS NOT NULL) AS is_ocean
    FROM read_parquet('{dir_dist}/dataset=bl/*.parquet') mc
    LEFT JOIN grid.cell c USING (cell_id))
  SELECT mdl_key, round(100.0 * count(*) FILTER (WHERE is_ocean) / count(*), 1) AS pct_marine_bl
  FROM b GROUP BY mdl_key"))

# family + cleaned name per bl model; map to the merged taxon via taxon_model
bl_mod <- read_csv(glue("{dir_dist}/model_bl.csv"), show_col_types = FALSE) |>
  transmute(mdl_key, family, sci = clean_sci_name(sci_name))
tmod <- dbGetQuery(con, "SELECT ms_merge_key, mdl_key FROM taxon_model WHERE mdl_key LIKE 'bl|%'")

bird <- bl_pm |>
  left_join(bl_mod, by = "mdl_key") |>
  left_join(tmod,   by = "mdl_key") |>
  mutate(is_marine_bird =
    ((family %in% marine_fam & pct_marine_bl >= 5) | sci %in% inc_sci) & !sci %in% exc_sci) |>
  group_by(ms_merge_key) |>
  summarise(pct_marine_bl  = max(pct_marine_bl),
            is_marine_bird = any(is_marine_bird), .groups = "drop")

duckdb_register(con, "tmp_bird_marine", bird)
dbExecute(con, "CREATE OR REPLACE TABLE taxon AS
  SELECT t.*,
    CASE WHEN t.taxon_authority = 'botw'
         THEN coalesce(b.is_marine_bird, FALSE) ELSE TRUE END AS is_marine,
    b.pct_marine_bl
  FROM taxon t LEFT JOIN tmp_bird_marine b USING (ms_merge_key)")
duckdb_unregister(con, "tmp_bird_marine")

mr <- dbGetQuery(con, "SELECT
  count(*) FILTER (WHERE taxon_authority='botw')                            n_birds,
  count(*) FILTER (WHERE taxon_authority='botw' AND is_marine)              n_marine,
  count(*) FILTER (WHERE taxon_authority='botw' AND is_valid_usa AND is_marine) n_marine_usa
  FROM taxon")
log_info("marine birds: {mr$n_marine} of {mr$n_birds} ({mr$n_marine_usa} valid_usa)")
knitr::kable(mr)
dbDisconnect(con, shutdown = TRUE)
```

## Taxon extinction-risk

The governing `extrisk_code` / `er_score` (US-national/floored, via
`msens::compute_er_score()`) are computed upstream in `merge_models_prep` — so
`merge_models` can apply them to the range-cell values — and carried through here on the
`taxon` table (via `x.*` in the flags chunk above). No recompute needed.

## Manifest

```{r}
#| label: manifest

write_json(
  list(target = "merge_taxon", ver = ver, n_taxa = smry$n_taxa,
       valid_global = smry$valid_global, valid_usa = smry$valid_usa,
       valid_pra = smry$valid_pra, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
```
