---
title: "Merge models — combine each taxon's models per cell (max-merge + range mask; turtles ×)"
msens:
  target_name: merge_models
  workflow_type: merge
  dependency: [merge_models_prep]
  output: data/manifests/merge_models.json
editor_options:
  chunk_output_type: console
---

Combine every taxon's per-dataset models into one **merged `model_cell`**
(`mdl_key = ms_merge|WORMS:…` / `BOTW:…`), reproducing v7's logic at the v8 global
scale (~13B am cell-rows), so it must stream + spill rather than materialize:

- **Default = max-merge, masked to the range footprint.** A taxon's merged cells are
  its **range** cells if it has any (`rng_*`/`ch_*`/`ca_*`/`bl`), else its full `am`
  footprint; at each kept cell `val = max` across the taxon's models. Streamed
  straight to Parquet via a single `COPY` (one grouped scan), so no 13B-row table.
- **Turtles = multiplicative.** `val = pmax(1, round(er × suit / 100))` with
  critical-habitat max-override — computed from **only the 6 turtle taxa's files**
  (never a full am scan, which is what OOM'd the naive version).

## Design

```{mermaid}
%%| label: fig-design
%%| fig-cap: "US-scoped per-taxon merge: max-merge over range footprint (governing er) + multiplicative turtles → streamed Parquet"
flowchart LR
  raw["dist/*/*.parquet<br/>(raw model_cell)"] --> mca["model_cell_all<br/>US-scoped, by ms_merge_key"]
  mca --> mx["max-merge over range<br/>(er_score at range cells)"]
  mca --> tt["turtles ×<br/>pmax(1, er·suit/100)"]
  mx --> pq[("dist_merged/dataset=ms_merge/**<br/>V2/zstd via copy_atlas_parquet")]
  tt --> pq
  pq --> mf["hash_parquet → content-addressed manifest"]
```

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, quiet = T)
source(here("libs/paths.R"))
source(here("libs/vars.R"))

dir_atlas  <- glue("{dir_big_v}/marine-atlas")
dir_dist   <- glue("{dir_atlas}/dist")
merge_db   <- glue("{dir_atlas}/merge.duckdb")
dir_merged <- glue("{dir_atlas}/dist_merged/dataset=ms_merge")
manifest   <- here("data/manifests/merge_models.json")
dir_create(path_dir(manifest))
stopifnot("run merge_models_prep first" = file_exists(merge_db),
          "run build_cell_grid first"   = file_exists(sdm_db))

suit_ds   <- "am"
turtle_ds <- "rng_turtle_swot_dps"
ch_ds     <- c("ch_nmfs", "ch_fws", "ca_nmfs")
ch_sql    <- paste(sprintf("'%s'", ch_ds), collapse = ",")

con <- dbConnect(duckdb(merge_db))
# spill config so the 13B-row aggregation goes out-of-core instead of OOM
mem_gb <- tryCatch(
  max(4, floor(as.numeric(system("sysctl -n hw.memsize", intern = TRUE)) / 1e9 * 0.5)),
  error = function(e) 8)
dbExecute(con, glue("PRAGMA memory_limit='{mem_gb}GB'"))
dbExecute(con, glue("PRAGMA temp_directory='{dir_atlas}/duckdb_tmp'"))
dbExecute(con, glue("PRAGMA threads={max(1L, parallel::detectCores() - 1L)}"))

# US-SCOPE: restrict the merge to in_usa cells (the global whole-range merge is a
# separate bigger-RAM run; a global GROUP BY over 13B near-unique (taxon,cell) rows
# won't fit here). Taxon range/pct_marine metrics are computed globally in merge_taxon.
dbExecute(con, glue("ATTACH '{sdm_db}' AS grid (READ_ONLY)"))
dbExecute(con, "CREATE OR REPLACE TABLE us_cells AS SELECT cell_id FROM grid.cell WHERE in_usa")
n_us <- dbGetQuery(con, "SELECT count(*) n FROM us_cells")$n
# MATERIALIZE the US-filtered model cells once (a streamed scan+join, no aggregate) so
# the parts below don't each re-scan the 66G am parquet.
# model_cell_all may persist as a view OR a table from a prior run; drop whichever it is
try(dbExecute(con, "DROP TABLE IF EXISTS model_cell_all"), silent = TRUE)
try(dbExecute(con, "DROP VIEW IF EXISTS model_cell_all"),  silent = TRUE)
dbExecute(con, glue("CREATE OR REPLACE TABLE model_cell_all AS
  SELECT tm.ms_merge_key, split_part(mc.mdl_key,'|',1) AS ds_key, mc.cell_id, mc.val
  FROM read_parquet('{dir_dist}/*/*.parquet') mc
  JOIN taxon_model tm USING (mdl_key)
  JOIN us_cells u ON mc.cell_id = u.cell_id"))
dbExecute(con, "CREATE INDEX mca_key ON model_cell_all(ms_merge_key)")
log_info("US-scoped merge: {n_us} in_usa cells, {dbGetQuery(con,'SELECT count(*) n FROM model_cell_all')$n} model-cells; memory_limit={mem_gb}GB")
```

## Turtle taxa (multiplicative, from only their files)

```{r}
#| label: turtles

turtle_keys <- dbGetQuery(con, glue(
  "SELECT DISTINCT ms_merge_key FROM taxon_model WHERE mdl_key LIKE '{turtle_ds}|%'"))$ms_merge_key
tkeys_sql   <- paste(sprintf("'%s'", turtle_keys), collapse = ",")
log_info("{length(turtle_keys)} turtle taxa (multiplicative merge)")

# turtle taxa's US cells from the materialized model_cell_all (indexed on ms_merge_key,
# already US-scoped) — the dist parquets are consolidated so per-model files aren't addressable
dbExecute(con, glue("CREATE OR REPLACE TABLE mc_turtle AS
  WITH t AS (
    SELECT ms_merge_key, ds_key, cell_id, val FROM model_cell_all WHERE ms_merge_key IN ({tkeys_sql})),
  er   AS (SELECT ms_merge_key, cell_id, max(val) er_value   FROM t WHERE ds_key='{turtle_ds}' GROUP BY 1,2),
  suit AS (SELECT ms_merge_key, cell_id, max(val) suit_value FROM t WHERE ds_key='{suit_ds}'   GROUP BY 1,2),
  ch   AS (SELECT ms_merge_key, cell_id, max(val) ch_value   FROM t WHERE ds_key IN ({ch_sql})  GROUP BY 1,2),
  mult AS (SELECT er.ms_merge_key, er.cell_id,
             greatest(1, CAST(round(er.er_value * coalesce(suit.suit_value,1) / 100.0) AS INTEGER)) AS val
           FROM er LEFT JOIN suit USING (ms_merge_key, cell_id))
  SELECT m.ms_merge_key AS mdl_key, m.cell_id,
         CAST(greatest(m.val, coalesce(ch.ch_value, 0)) AS DOUBLE) AS val
  FROM mult m LEFT JOIN ch USING (ms_merge_key, cell_id)"))
log_info("mc_turtle: {dbGetQuery(con, 'SELECT count(*) n FROM mc_turtle')$n} cells")
```

## Max-merge (masked) + turtles → streamed to Parquet

```{r}
#| label: merge_write

# A single GROUP BY over the dense am produces ~600M near-unique (taxon,cell) groups
# that won't fit even spilling. Decompose so we NEVER group dense am:
#   Part 1 am-only taxa  -> stream cells (dense, but no aggregate)
#   Part 2 range-only    -> group-by (sparse ranges, small)
#   Part 3 both (am+range) -> range footprint + am AT those cells via JOIN (no am group)
#   Part 4 turtles        -> mc_turtle (done)
dbExecute(con, "CREATE OR REPLACE TABLE taxon_flags AS
  SELECT ms_merge_key, bool_or(ds_key='am') AS has_am, bool_or(ds_key<>'am') AS has_range
  FROM model_cell_all GROUP BY ms_merge_key")

if (dir_exists(dir_merged)) dir_delete(dir_merged)
dir_create(dir_merged)

# Part 1: am-only taxa -> stream US am cells directly (no aggregate)
msens::copy_atlas_parquet(con,
  "SELECT mc.ms_merge_key AS mdl_key, mc.cell_id, mc.val FROM model_cell_all mc
    WHERE mc.ms_merge_key IN (SELECT ms_merge_key FROM taxon_flags WHERE has_am AND NOT has_range)",
  glue("{dir_merged}/p_amonly"), per_thread = TRUE)

# ER RECONCILIATION (the "fitting point"): a range presence cell carries the taxon's
# GOVERNING er_score (US-national/floored, from merge_models_prep) — NOT the raw
# per-dataset source ER the ingest layer held. So Part 2/3 val the range footprint
# by taxon.er_score, matching v7. (Turtles keep their spatially-varying DPS ER surface.)

# Part 2: range-only taxa -> governing er_score over their range footprint
dbExecute(con, glue("CREATE OR REPLACE TABLE mc_rangeonly AS
    SELECT DISTINCT mc.ms_merge_key AS mdl_key, mc.cell_id, t.er_score::DOUBLE AS \"val\"
    FROM model_cell_all mc JOIN taxon t ON mc.ms_merge_key = t.ms_merge_key
    WHERE mc.ms_merge_key IN (SELECT ms_merge_key FROM taxon_flags WHERE has_range AND NOT has_am)
      AND mc.ms_merge_key NOT IN ({tkeys_sql})"))

# Part 3: both taxa -> range footprint (er), then am AT those cells via a join that
# builds the small range set and probes am (never a dense-am group-by)
dbExecute(con, glue("CREATE OR REPLACE TABLE both_range AS
    SELECT DISTINCT mc.ms_merge_key, mc.cell_id, t.er_score AS er
    FROM model_cell_all mc JOIN taxon t ON mc.ms_merge_key = t.ms_merge_key
    WHERE mc.ms_merge_key IN (SELECT ms_merge_key FROM taxon_flags WHERE has_am AND has_range)
      AND mc.ds_key <> 'am' AND mc.ms_merge_key NOT IN ({tkeys_sql})"))
dbExecute(con, "CREATE OR REPLACE TABLE both_am AS
    SELECT am.ms_merge_key, am.cell_id, max(am.val) am_value
    FROM model_cell_all am JOIN both_range br USING (ms_merge_key, cell_id)
    WHERE am.ds_key='am' GROUP BY am.ms_merge_key, am.cell_id")
dbExecute(con, "CREATE OR REPLACE TABLE mc_both AS
    SELECT br.ms_merge_key AS mdl_key, br.cell_id,
           greatest(br.er, coalesce(ba.am_value, 0))::DOUBLE AS \"val\"
    FROM both_range br LEFT JOIN both_am ba USING (ms_merge_key, cell_id)")
log_info("range-only {dbGetQuery(con,'SELECT count(*) n FROM mc_rangeonly')$n}, both {dbGetQuery(con,'SELECT count(*) n FROM mc_both')$n} cells")

# write parts 2+3+turtles alongside the streamed part 1
msens::copy_atlas_parquet(con,
  "SELECT mdl_key, cell_id, val FROM mc_rangeonly
    UNION ALL SELECT mdl_key, cell_id, val FROM mc_both
    UNION ALL SELECT mdl_key, cell_id, val FROM mc_turtle",
  glue("{dir_merged}/p_agg"), per_thread = TRUE)

smry <- dbGetQuery(con, glue("SELECT count(DISTINCT mdl_key) n_taxa, count(*) n_cells,
  min(val) v_min, max(val) v_max FROM read_parquet('{dir_merged}/**/*.parquet')"))
# order-independent fingerprint of the merged surface (before disconnect)
h <- msens::hash_parquet(glue("{dir_merged}/**/*.parquet"), con)
msens::report_table(smry, caption = "merged model_cell surface (ms_merge)")
dbDisconnect(con, shutdown = TRUE)
```

## Manifest

```{r}
#| label: manifest

# content-addressed: fingerprint of the merged Parquet surface (no wall-clock, no path)
msens::write_manifest(
  manifest, target = "merge_models", content_hash = h,
  stats = list(ver = ver, n_taxa = smry$n_taxa, n_cells = smry$n_cells,
               v_min = smry$v_min, v_max = smry$v_max),
  force = msens::force_target("merge_models"))
```
