---
title: "Environmental Sensitivity Summary by Program Areas"
subtitle: "for the _National Outer Continental Shelf Oil and Gas Leasing Proposed Final Program_"
author: "Marine Sensitivity Toolkit Team"
editor_options:
  chunk_output_type: console
---

## Summary

Program Area–level summaries of environmental sensitivity scores, primary productivity, and species group metrics. Adapted from `msens-summary_env-national-ocs.qmd` (Planning Area version) to use the 2026 Program Area polygons with versioned outputs.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
librarian::shelf(
  bslib, DBI, dplyr, duckdb, DT, flextable, fs, ftExtra, ggiraph, ggplot2,
  glue, gt, here, htmlwidgets, kableExtra, knitr, mapview, openxlsx, purrr,
  RColorBrewer, readr, scales, sf, shiny, stringr,
  terra, tibble, tidyr, tidyverse, units, webshot2,
  quiet = T)
options(readr.show_col_types = F)

# version ----
v_int <- 3
v_sfx <- paste0("_v", v_int)

# variables ----
verbose        <- F
is_server      <- Sys.info()[["sysname"]] == "Linux"
dir_private    <- ifelse(
  is_server,
  "/share/private",
  "~/My Drive/private")
dir_data       <- ifelse(
  is_server,
  "/share/data",
  "~/My Drive/projects/msens/data")
dir_derived    <- glue("{dir_data}/derived")
dir_figs       <- here("figs/msens-summary_programareas")
dir_dl         <- glue("{dir_derived}/downloads")
mapbox_tkn_txt <- glue("{dir_private}/mapbox_token_bdbest.txt")

# input files (versioned) ----
sdm_dd         <- glue("{dir_derived}/sdm_2026.duckdb")
pra_gpkg       <- glue("{dir_derived}/ply_programareas_2026{v_sfx}.gpkg")
er_gpkg        <- glue("{dir_derived}/ply_ecoregions_2025{v_sfx}.gpkg")
metrics_tif    <- glue("{dir_derived}/r_metrics_2026{v_sfx}.tif")
sr_pra_csv     <- glue("{dir_derived}/subregion_programareas.csv")

# output files (versioned) ----
er_pra_gpkg    <- glue("{dir_derived}/ply_ecoregion_programarea{v_sfx}.gpkg")
er_pra_csv     <- glue("{dir_derived}/ply_ecoregion_programarea{v_sfx}.csv")
er_pra_n_csv   <- glue("{dir_derived}/tbl_er_pra_scores{v_sfx}.csv")
pra_n_csv      <- glue("{dir_derived}/tbl_pra_scores{v_sfx}.csv")

Sys.setenv(MAPBOX_PUBLIC_TOKEN = readLines(mapbox_tkn_txt))
librarian::shelf(mapgl)

dir_create(dir_figs)

# db connections ----
source(here("libs/db.R")) # con (PostGIS)
con_sdm <- dbConnect(duckdb(), dbdir = sdm_dd, read_only = T)

# flower plot function ----
plot_flower <- function(
    data,
    fld_category,
    fld_height,
    fld_width,
    colors,
    tooltip_expr = NULL,
    score        = NULL,
    title        = NULL,
    interactive  = T) {

  stopifnot(is.numeric(data |> pull({{ fld_height }})))
  stopifnot(is.numeric(data |> pull({{ fld_width }})))

  if (is.null(score)) {
    score <- data |>
      mutate(
        "{{fld_height}}" := as.double({{ fld_height }}),
        "{{fld_width}}"  := as.double({{ fld_width  }}) ) |>
      summarize(
        score = weighted.mean({{ fld_height }}, {{ fld_width }}, na.rm = T)) |>
      pull(score)
  }

  d <- data |>
    arrange({{ fld_category }}) |>
    mutate(across(!where(is.character), as.double)) |>
    mutate(
      ymax = cumsum({{ fld_width }}),
      ymin = lag(ymax, default = 0),
      xmax = {{ fld_height }},
      xmin = 0)

  sym_category <- ensym(fld_category)
  sym_height   <- ensym(fld_height)
  sym_width    <- ensym(fld_width)

  if (!is.null(tooltip_expr)) {
    d <- d |>
      mutate(tooltip = glue(tooltip_expr))
  } else {
    d <- d |>
      mutate(tooltip = glue("{!!fld_category}"))
  }

  g <- ggplot(d) +
    geom_rect_interactive(aes(
      xmin    = xmin,
      xmax    = xmax,
      ymin    = ymin,
      ymax    = ymax,
      fill    = {{ fld_category }},
      color   = "white",
      data_id = {{ fld_category }},
      tooltip = tooltip),
      color = "white",
      alpha = 0.5) +
    coord_polar(theta = "y") +
    xlim(c(-10, max(data |> pull({{ fld_height }})))) +
    annotate(
      "text", x = -10, y = 0,
      label = round(score),
      size  = 8,
      fontface = "bold") +
    scale_fill_manual(values = colors) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(20, 20, 20, 20), "pt"))

  if (!is.null(title))
    g <- g + ggtitle(title)

  if (interactive)
    g <- girafe(
      ggobj = g,
      options = list(
        opts_sizing(rescale = TRUE, width = 1),
        opts_tooltip(
          css = "background-color:white;color:black;padding:5px;border-radius:3px;")))

  g
}
```

## Ecoregion × Program Area Polygons

```{r}
#| label: read_plys
er  <- read_sf(er_gpkg)
pra <- read_sf(pra_gpkg)

if (!file.exists(er_pra_gpkg)) {
  er_pra <- st_intersection(
    er |>
      select(ecoregion_key, ecoregion_name),
    pra |>
      select(region_key, region_name, programarea_key, programarea_name)) |>
    st_make_valid() |>
    mutate(
      area_km2 = st_area(geom) |> set_units(km^2) |> drop_units(),
      ctr_lon  = st_coordinates(st_centroid(geom))[, 1],
      ctr_lat  = st_coordinates(st_centroid(geom))[, 2]) |>
    filter(area_km2 > 1) |>
    arrange(programarea_key, ecoregion_key) |>
    group_by(programarea_key) |>
    mutate(
      pct_programarea_in_ecoregion = round(
        area_km2 / sum(area_km2) * 100, 2),
      pct_programarea_in_ecoregion = na_if(
        pct_programarea_in_ecoregion, 100.0)) |>
    ungroup()

  er_pra <- er_pra |>
    select(
      region_key, region_name,
      ecoregion_key, ecoregion_name,
      programarea_key, programarea_name,
      pct_programarea_in_ecoregion,
      area_km2, ctr_lon, ctr_lat) |>
    arrange(region_key, ecoregion_key, programarea_key) |>
    mutate(
      area_km2 = round(area_km2, 2),
      ctr_lon  = round(ctr_lon, 4),
      ctr_lat  = round(ctr_lat, 4))

  er_pra |>
    st_drop_geometry() |>
    write_csv(er_pra_csv, na = "")

  write_sf(er_pra, er_pra_gpkg, delete_dsn = T)
}
er_pra <- read_sf(er_pra_gpkg)
```

## Label Points

```{r}
#| label: ply_lbl_pts
lbls_csv <- glue("{dir_derived}/ply_label_placement_pra.csv")

er_pts <- er |>
  st_drop_geometry() |>
  select(
    ecoregion_key, ecoregion_name, region_name, region_key,
    ctr_lon, ctr_lat) |>
  st_as_sf(coords = c("ctr_lon", "ctr_lat"), crs = 4326)
pra_pts <- pra |>
  st_drop_geometry() |>
  select(
    programarea_key, programarea_name, region_name, region_key,
    ctr_lon, ctr_lat) |>
  st_as_sf(coords = c("ctr_lon", "ctr_lat"), crs = 4326)

if (!file.exists(lbls_csv)) {
  bind_rows(
    er_pts  |> st_drop_geometry() |> select(ecoregion_key),
    pra_pts |> st_drop_geometry() |> select(programarea_key)) |>
    arrange(ecoregion_key, programarea_key) |>
    mutate(
      text_anchor = ifelse(
        !is.na(programarea_key),
        "center",
        "top"),
      text_justify      = "center",
      text_offset_right = 0,
      text_offset_down  = 0) |>
    write_csv(lbls_csv, na = "")
}
```

## Program Area Maps

```{r}
#| label: map_imgs_pra_ply_l48_ak
d_lbls <- read_csv(lbls_csv)

var_pra <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
n_cols  <- 11
rng_pra <- pra |>
  pull({{ var_pra }}) |>
  range()
cols_pra <- rev(RColorBrewer::brewer.pal(n_cols, "Spectral"))
brks_pra <- seq(rng_pra[1], rng_pra[2], length.out = n_cols)

for (is_ak in c(T, F)) {

  if (is_ak) {
    # alaska
    er_area  <- er |>
      filter(region_key == "AK") |>
      st_shift_longitude()
    pra_area <- pra |>
      filter(region_key == "AK") |>
      st_shift_longitude()

    er_pts_area <- er_pts |>
      filter(region_key == "AK") |>
      st_shift_longitude() |>
      left_join(
        d_lbls |> select(-programarea_key),
        by = "ecoregion_key")
    pra_pts_area <- pra_pts |>
      filter(region_key == "AK") |>
      st_shift_longitude() |>
      left_join(
        d_lbls |> select(-ecoregion_key),
        by = "programarea_key")

    lgnd_pos <- "top-left"
  } else {
    # lower 48
    box_lower48 <- pra |>
      filter(region_key != "AK") |>
      st_bbox() |>
      st_as_sfc()

    er_area <- er |>
      st_filter(box_lower48, .predicate = st_intersects)
    pra_area <- pra |>
      st_filter(box_lower48, .predicate = st_intersects)

    er_pts_area <- er_pts |>
      filter(ecoregion_key %in% er_area$ecoregion_key) |>
      left_join(
        d_lbls |> select(-programarea_key),
        by = "ecoregion_key")
    pra_pts_area <- pra_pts |>
      filter(programarea_key %in% pra_area$programarea_key) |>
      left_join(
        d_lbls |> select(-ecoregion_key),
        by = "programarea_key")

    lgnd_pos <- "bottom-left"
  }
  scale_pos <- "bottom-right"

  er_url  <- glue("https://api.marinesensitivity.org/tilejson?table=public.ply_ecoregions_2025{v_sfx}&use_cache=FALSE")
  pra_url <- glue("https://api.marinesensitivity.org/tilejson?table=public.ply_programareas_2026{v_sfx}&use_cache=FALSE")
  er_filter  <- c("in", "ecoregion_key",   er_area$ecoregion_key)
  pra_filter <- c("in", "programarea_key", pra_area$programarea_key)

  m <- mapboxgl(
    style      = mapbox_style("dark"),
    projection = "globe") |>
    fit_bounds(bbox = er_area) |>
    add_vector_source(
      id  = "pra_src",
      url = pra_url) |>
    add_vector_source(
      id  = "er_src",
      url = er_url) |>
    add_fill_layer(
      id           = "pra_fill",
      source       = "pra_src",
      source_layer = glue("public.ply_programareas_2026{v_sfx}"),
      filter       = pra_filter,
      fill_color   = mapgl::interpolate(
        column = var_pra,
        values = brks_pra,
        stops  = cols_pra)) |>
    add_line_layer(
      id           = "er_ln",
      source       = "er_src",
      source_layer = glue("public.ply_ecoregions_2025{v_sfx}"),
      filter       = er_filter,
      line_color   = "lightgray",
      line_opacity = 1,
      line_width   = 5) |>
    add_line_layer(
      id           = "pra_ln",
      source       = "pra_src",
      source_layer = glue("public.ply_programareas_2026{v_sfx}"),
      filter       = pra_filter,
      line_color   = "black",
      line_opacity = 1,
      line_width   = 1) |>
    add_symbol_layer(
      id                 = "pra_symbol",
      source             = pra_pts_area,
      filter             = pra_filter,
      text_field         = get_column("programarea_key"),
      text_allow_overlap = T,
      text_anchor        = get_column("text_anchor"),
      text_justify       = get_column("text_justify"),
      text_offset        = list(
        get_column("text_offset_right"),
        get_column("text_offset_down"))) |>
    add_symbol_layer(
      id                 = "er_symbol",
      source             = er_pts_area,
      filter             = er_filter,
      text_field         = get_column("ecoregion_name"),
      text_color         = "black",
      text_allow_overlap = T,
      text_anchor        = get_column("text_anchor"),
      text_justify       = get_column("text_justify"),
      text_offset        = list(
        get_column("text_offset_right"),
        get_column("text_offset_down")),
      text_halo_color    = "lightgray",
      text_halo_width    = 2,
      text_halo_blur     = 2,
      text_line_height   = 1.5,
      text_size          = 20) |>
    mapgl::add_legend(
      "Score",
      values   = rng_pra,
      colors   = cols_pra,
      position = lgnd_pos) |>
    add_scale_control(position = scale_pos)

  b       <- glue("{dir_figs}/map_pra_{ifelse(is_ak, 'ak', 'l48')}")
  b_html  <- glue("{b}.html")
  b_files <- glue("{b}_files")
  b_png   <- glue("{b}.png")
  saveWidget(m, b_html, selfcontained = F, background = "transparent")
  webshot(
    b_html, b_png,
    vwidth   = 1200, vheight = 800, zoom = 2,
    selector = "#htmlwidget_container",
    delay    = 2, quiet = T)
  file_delete(b_html)
  dir_delete(b_files)
}
```

## Cell Score & NPP Maps

```{r}
#| label: map_imgs_cell_l48_ak
vgpm_tif <- glue("{dir_data}/raw/oregonstate.edu/vgpm.r2022.v.chl.v.sst.2160x4320_2014-2023.avg.sd_planareas.tif")

r_metrics <- rast(metrics_tif)
r_vgpm    <- rast(vgpm_tif)
d_lbls    <- read_csv(lbls_csv)

lyr_score <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
lyr_npp   <- "npp_avg"

er_ak  <- er |>
  filter(region_key == "AK") |>
  pull(ecoregion_key)
er_l48 <- c("CAC", "WAOR", "SECS", "WCGOA", "EGOA", "NECS")

r_score <- r_metrics[[lyr_score]] |>
  mask(r_metrics$ecoregion_key %in% c(er_l48, er_ak), maskvalues = F)

r_npp <- r_vgpm |>
  log() |>
  mask(r_metrics$ecoregion_key %in% c(er_l48, er_ak), maskvalues = F)

n_cols <- 11
rng_r_score  <- range(values(r_score, na.rm = T))
cols_r_score <- rev(RColorBrewer::brewer.pal(n_cols, "Spectral"))
brks_r_score <- seq(rng_r_score[1], rng_r_score[2], length.out = n_cols)

rng_r_npp  <- range(values(r_npp, na.rm = T))
cols_r_npp <- pal_viridis()(n_cols)
brks_r_npp <- seq(rng_r_npp[1], rng_r_npp[2], length.out = n_cols)

for (is_vgpm in c(T, F)) {

  if (is_vgpm) {
    r      <- r_npp
    rng_r  <- round(rng_r_npp, 2)
    cols_r <- cols_r_npp
    brks_r <- brks_r_npp
    lyr    <- lyr_npp
    title  <- "Primary Productivity (log(NPP))"
  } else {
    r      <- r_score
    rng_r  <- rng_r_score
    cols_r <- cols_r_score
    brks_r <- brks_r_score
    lyr    <- lyr_score
    title  <- "Score"
  }

  for (is_ak in c(T, F)) {

    if (is_ak) {
      r_area <- r |>
        mask(r_metrics$ecoregion_key %in% er_ak, maskvalues = F) |>
        trim()

      er_area <- er |>
        filter(region_key == "AK") |>
        st_shift_longitude()
      pra_area <- pra |>
        filter(region_key == "AK") |>
        st_shift_longitude()

      er_pts_area <- er_pts |>
        filter(region_key == "AK") |>
        st_shift_longitude() |>
        left_join(
          d_lbls |> select(-programarea_key),
          by = "ecoregion_key")
      pra_pts_area <- pra_pts |>
        filter(region_key == "AK") |>
        st_shift_longitude() |>
        left_join(
          d_lbls |> select(-ecoregion_key),
          by = "programarea_key")

      lgnd_pos <- "top-left"
    } else {
      r_area <- r |>
        mask(r_metrics$ecoregion_key %in% er_l48, maskvalues = F) |>
        trim()

      box_lower48 <- pra |>
        filter(region_key != "AK") |>
        st_bbox() |>
        st_as_sfc()

      er_area <- er |>
        st_filter(box_lower48, .predicate = st_intersects)
      pra_area <- pra |>
        st_filter(box_lower48, .predicate = st_intersects)

      er_pts_area <- er_pts |>
        filter(ecoregion_key %in% er_area$ecoregion_key) |>
        left_join(
          d_lbls |> select(-programarea_key),
          by = "ecoregion_key")
      pra_pts_area <- pra_pts |>
        filter(programarea_key %in% pra_area$programarea_key) |>
        left_join(
          d_lbls |> select(-ecoregion_key),
          by = "programarea_key")

      lgnd_pos <- "bottom-left"
    }
    scale_pos <- "bottom-right"

    er_url  <- glue("https://api.marinesensitivity.org/tilejson?table=public.ply_ecoregions_2025{v_sfx}&use_cache=FALSE")
    pra_url <- glue("https://api.marinesensitivity.org/tilejson?table=public.ply_programareas_2026{v_sfx}&use_cache=FALSE")
    er_filter  <- c("in", "ecoregion_key",   er_area$ecoregion_key)
    pra_filter <- c("in", "programarea_key", pra_area$programarea_key)

    m <- mapboxgl(
      style      = mapbox_style("dark"),
      projection = "globe") |>
      fit_bounds(bbox = er_area) |>
      add_image_source(
        id     = "r_src",
        data   = r_area,
        colors = cols_r) |>
      add_vector_source(
        id  = "pra_src",
        url = pra_url) |>
      add_vector_source(
        id  = "er_src",
        url = er_url) |>
      add_raster_layer(
        id                = "r_lyr",
        source            = "r_src",
        raster_opacity    = 0.9,
        raster_resampling = "nearest") |>
      add_line_layer(
        id           = "er_ln",
        source       = "er_src",
        source_layer = glue("public.ply_ecoregions_2025{v_sfx}"),
        filter       = er_filter,
        line_color   = "lightgray",
        line_opacity = 1,
        line_width   = 5) |>
      add_line_layer(
        id           = "pra_ln",
        source       = "pra_src",
        source_layer = glue("public.ply_programareas_2026{v_sfx}"),
        filter       = pra_filter,
        line_color   = "black",
        line_opacity = 1,
        line_width   = 1) |>
      add_symbol_layer(
        id                 = "pra_symbol",
        source             = pra_pts_area,
        filter             = pra_filter,
        text_field         = get_column("programarea_key"),
        text_allow_overlap = T,
        text_anchor        = get_column("text_anchor"),
        text_justify       = get_column("text_justify"),
        text_offset        = list(
          get_column("text_offset_right"),
          get_column("text_offset_down"))) |>
      add_symbol_layer(
        id                 = "er_symbol",
        source             = er_pts_area,
        filter             = er_filter,
        text_field         = get_column("ecoregion_name"),
        text_color         = "black",
        text_allow_overlap = T,
        text_anchor        = get_column("text_anchor"),
        text_justify       = get_column("text_justify"),
        text_offset        = list(
          get_column("text_offset_right"),
          get_column("text_offset_down")),
        text_halo_color    = "lightgray",
        text_halo_width    = 2,
        text_halo_blur     = 2,
        text_line_height   = 1.5,
        text_size          = 20) |>
      mapgl::add_legend(
        title,
        values   = rng_r,
        colors   = cols_r,
        position = lgnd_pos) |>
      add_scale_control(position = scale_pos)

    b       <- glue("{dir_figs}/map_cell_{ifelse(is_vgpm, 'npp', 'score')}_{ifelse(is_ak, 'ak', 'l48')}")
    b_html  <- glue("{b}.html")
    b_files <- glue("{b}_files")
    b_png   <- glue("{b}.png")
    saveWidget(m, b_html, selfcontained = F, background = "transparent")
    webshot(
      b_html, b_png,
      vwidth   = 1200, vheight = 800, zoom = 2,
      selector = "#htmlwidget_container",
      delay    = 2, quiet = T)
    file_delete(b_html)
    dir_delete(b_files)
  }
}
```

## Program Area Score Table

```{r}
#| label: pra_score_table
d_pra <- read_sf(pra_gpkg) |>
  st_drop_geometry()
d_er <- read_sf(er_gpkg) |>
  st_drop_geometry()
flds <- intersect(names(d_pra), names(d_er)) |>
  str_subset("rescaled") |> sort() |> rev()
flds_rn <- flds |>
  str_replace(
    "score_extriskspcat_primprod_ecoregionrescaled_equalweights", "score") |>
  str_replace("_ecoregion_rescaled", "") |>
  str_replace("extrisk_", "") |>
  setNames(flds)

d_er_n <- read_csv(er_pra_csv) |>
  distinct(
    region_key, region_name,
    ecoregion_key, ecoregion_name) |>
  left_join(
    d_er |>
      select(all_of(c("ecoregion_key", "area_km2", flds))) |>
      rename_with(~ flds_rn[.x], all_of(flds)),
    by = "ecoregion_key")

d_pra_n <- read_csv(er_pra_csv) |>
  select(
    region_name, ecoregion_name,
    programarea_key, programarea_name,
    pct_programarea_in_ecoregion) |>
  left_join(
    d_pra |>
      select(all_of(c("programarea_key", "area_km2", flds))) |>
      rename_with(~ flds_rn[.x], all_of(flds)),
    by = "programarea_key")

d_er_pra_n <- bind_rows(
  d_er_n |>
    mutate(row_type = "ecoregion"),
  d_pra_n |>
    mutate(row_type = "programarea")) |>
  relocate(
    row_type, programarea_key, programarea_name,
    pct_programarea_in_ecoregion, .after = ecoregion_name) |>
  arrange(region_name, ecoregion_name, row_type, programarea_name) |>
  mutate(
    pct = ifelse(
      !is.na(pct_programarea_in_ecoregion),
      glue(" [{round(pct_programarea_in_ecoregion, 1)}%]", .trim = F),
      ""),
    row_lbl = case_match(
      row_type,
      "ecoregion"   ~ glue("{region_name}: {ecoregion_name} ({ecoregion_key})"),
      "programarea" ~ glue("· {programarea_name} ({programarea_key}){pct}"))) |>
  relocate(row_lbl)

d_er_pra_n |>
  select(
    row_type,
    region_name,
    ecoregion_name, ecoregion_key,
    programarea_name, programarea_key, pct_programarea_in_ecoregion,
    row_lbl,
    score,
    primprod,
    bird, coral, fish, invertebrate, mammal, reptile, other,
    area_km2) |>
  write_excel_csv(er_pra_n_csv)
```

## Ecoregion × Program Area Table (docx)

```{r}
#| label: flextable_er_pra
tbl_docx <- glue("{dir_figs}/table_ecoregion_programarea_scores{v_sfx}.docx")

set_flextable_defaults(
  font.family = "Arial", font.size = 10,
  border.color = "gray", big.mark = ",")

d_er_pra_n |>
  mutate(
    area_kkm2 = area_km2 / 1000) |>
  mutate(across(where(is.numeric), round)) |>
  mutate(
    row_lbl = ifelse(
      row_type == "ecoregion",
      glue("**{row_lbl}**"),
      row_lbl)) |>
  select(
    Area                 = row_lbl,
    Score                = score,
    `Prim. Prod.`        = primprod,
    Bird                 = bird,
    Coral                = coral,
    Fish                 = fish,
    Invertebrate         = invertebrate,
    Mammal               = mammal,
    Reptile              = reptile,
    Other                = other,
    `area (1,000 km^2^)` = area_kkm2) |>
  flextable() |>
  ftExtra::colformat_md(part = "all") |>
  theme_vanilla() |>
  save_as_docx(path = tbl_docx)

browseURL(tbl_docx)
```

## Program Area Table (docx + xlsx)

```{r}
#| label: flextable_pra
tbl_docx <- glue("{dir_figs}/table_programarea_scores{v_sfx}.docx")
tbl_xlsx <- glue("{dir_figs}/table_programarea_scores{v_sfx}.xlsx")

set_flextable_defaults(
  font.family = "Arial", font.size = 10,
  border.color = "gray", big.mark = ",")

d_pra_only <- d_er_pra_n |>
  filter(row_type == "programarea") |>
  arrange(desc(score)) |>
  select(
    -row_type,
    -ecoregion_name, -ecoregion_key,
    -pct_programarea_in_ecoregion) |>
  mutate(
    area_kkm2 = area_km2 / 1000) |>
  mutate(across(where(is.numeric), round)) |>
  mutate(
    row_lbl = glue("{programarea_name} ({programarea_key})")) |>
  distinct(
    row_lbl, score, primprod,
    bird, coral, fish, invertebrate, mammal, reptile, other,
    area_kkm2) |>
  select(
    `Program Area`       = row_lbl,
    Score                = score,
    `Primary Production` = primprod,
    Bird                 = bird,
    Coral                = coral,
    Fish                 = fish,
    Invertebrate         = invertebrate,
    Mammal               = mammal,
    Reptile              = reptile,
    Other                = other,
    `Area (1,000 km^2^)` = area_kkm2)

write_excel_csv(d_pra_only, pra_n_csv, na = "")

d_pra_only |>
  flextable() |>
  ftExtra::colformat_md(part = "all") |>
  theme_vanilla() |>
  save_as_docx(path = tbl_docx)

# xlsx version (plain column names without superscript)
d_pra_only |>
  rename(`Area (1,000 km2)` = `Area (1,000 km^2^)`) |>
  openxlsx::write.xlsx(tbl_xlsx)

browseURL(tbl_docx)
```

## Ecoregion Flower Plots

```{r}
#| label: plot_ecoregion_flowers
source(here("libs/db.R")) # con

# contributing metrics (same as PA version — ecoregion-level)
sp_cats_u <- tbl(con_sdm, "species") |>
  distinct(sp_cat) |>
  pull(sp_cat) |>
  str_replace(" ", "_")
metric_input_keys <- c(
  glue("extrisk_{sp_cats_u}_ecoregion_rescaled"),
  "primprod_ecoregion_rescaled")
m_input_seqs <- tbl(con_sdm, "metric") |>
  filter(metric_key %in% metric_input_keys) |>
  pull(metric_seq)

d_fl <- tbl(con_sdm, "zone") |>
  filter(tbl == "ply_ecoregions_2025") |>
  select(ecoregion_key = value, zone_seq) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "zone_seq") |>
  left_join(
    tbl(con_sdm, "metric") |>
      filter(metric_seq %in% m_input_seqs) |>
      select(metric_seq, metric_key),
    by = "metric_seq") |>
  filter(!is.na(metric_key)) |>
  select(ecoregion_key, metric_key, value) |>
  arrange(ecoregion_key, metric_key) |>
  collect() |>
  left_join(
    tbl(con, "ply_ecoregions_2025") |>
      filter(
        region_key == "AK" |
          ecoregion_key %in% c("CAC", "WAOR", "SECS", "WCGOA", "EGOA", "NECS")) |>
      select(ecoregion_key, ecoregion_name) |>
      collect(),
    by = "ecoregion_key") |>
  mutate(
    component = metric_key |>
      str_replace("extrisk_", "") |>
      str_replace("_ecoregion_rescaled", "") |>
      str_replace("_", " "),
    even = 1) |>
  filter(component != "all") |>
  filter(
    !is.na(ecoregion_name),
    !is.na(metric_key),
    !is.na(value))

components <- c(
  "invertebrate", "mammal", "other", "primprod", "reptile", "bird", "coral", "fish")
cols <- setNames(hue_pal()(length(components)), components)

# all flowers ----
d <- d_fl |>
  group_by(ecoregion_name) |>
  arrange(component) |>
  mutate(across(!where(is.character), as.double)) |>
  mutate(
    ymax = cumsum(even),
    ymin = lag(ymax, default = 0),
    xmax = value,
    xmin = 0)

d_scores <- d_fl |>
  group_by(ecoregion_key, ecoregion_name) |>
  summarize(
    score = weighted.mean(value, even, na.rm = T) |> round(),
    .groups = "drop")

g <- d |>
  ggplot() +
  geom_rect_interactive(aes(
    xmin    = xmin,
    xmax    = xmax,
    ymin    = ymin,
    ymax    = ymax,
    fill    = component,
    color   = "white",
    data_id = component),
    color = "white",
    alpha = 0.5) +
  scale_fill_manual(values = cols) +
  facet_wrap(~ecoregion_key) +
  coord_polar(theta = "y") +
  xlim(c(-20, max(d_scores$score))) +
  geom_text(
    aes(x, y, label = score),
    data = d_scores |>
      mutate(x = -20, y = 0),
    size     = 4,
    fontface = "bold") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.margin     = unit(c(20, 20, 20, 20), "pt"),
    strip.text.x    = element_text(size = 8),
    axis.title.x    = element_blank(),
    axis.text.x     = element_blank(),
    axis.ticks.x    = element_blank(),
    axis.title.y    = element_blank(),
    axis.text.y     = element_blank(),
    axis.ticks.y    = element_blank())

er_fl_png <- glue("{dir_figs}/ecoregion_flower-plot_scores{v_sfx}.png")
er_fl_pdf <- glue("{dir_figs}/ecoregion_flower-plot_scores{v_sfx}.pdf")
png(er_fl_png, width = 1200, height = 800, res = 150)
print(g)
dev.off()
pdf(er_fl_pdf, width = 7, height = 5)
print(g)
dev.off()
```

## Program Area Flower Plots

```{r}
#| label: plot_programarea_flowers
source(here("libs/db.R")) # con

# pra list from PostGIS ----
pra_all <- tbl(con, glue("ply_programareas_2026{v_sfx}")) |>
  select(programarea_key, programarea_name) |>
  collect()

# components ----
sp_cats <- tbl(con_sdm, "taxon") |>
  filter(is_ok) |>
  distinct(sp_cat) |>
  pull(sp_cat)
metric_input_keys <- c(
  glue("extrisk_{sp_cats}_ecoregion_rescaled"),
  "primprod_ecoregion_rescaled")
m_input_seqs <- tbl(con_sdm, "metric") |>
  filter(metric_key %in% metric_input_keys) |>
  pull(metric_seq)

components <- c(
  "invertebrate", "mammal", "other", "primary production", "reptile", "bird",
  "coral", "fish") |> str_to_title()
component_cols <- setNames(
  hue_pal()(length(components)),
  components)

# d_scores ----
m_score <- "score_extriskspcat_primprod_ecoregionrescaled_equalweights"
m_key_score <- tbl(con_sdm, "metric") |>
  filter(metric_key == !!m_score) |>
  pull(metric_seq)

d_scores <- tbl(con_sdm, "zone") |>
  filter(
    tbl == glue("ply_programareas_2026{v_sfx}"),
    fld == "programarea_key",
    value %in% pra_all$programarea_key) |>
  select(programarea_key = value, zone_seq) |>
  inner_join(
    tbl(con_sdm, "zone_metric") |>
      filter(metric_seq == !!m_key_score),
    by = join_by(zone_seq)) |>
  select(programarea_key, score = value) |>
  mutate(score = round(score)) |>
  collect() |>
  arrange(desc(score), programarea_key)

d_scores$programarea_lbl <- factor(
  d_scores$programarea_key,
  levels = d_scores$programarea_key, ordered = T)

# d_fl ----
d_fl <- tbl(con_sdm, "zone") |>
  filter(
    tbl == glue("ply_programareas_2026{v_sfx}"),
    fld == "programarea_key",
    value %in% pra_all$programarea_key) |>
  select(programarea_key = value, zone_seq) |>
  inner_join(
    tbl(con_sdm, "zone_metric"),
    by = "zone_seq") |>
  inner_join(
    tbl(con_sdm, "metric") |>
      filter(metric_seq %in% m_input_seqs) |>
      select(metric_seq, metric_key),
    by = "metric_seq") |>
  select(programarea_key, metric_key, value) |>
  arrange(programarea_key, metric_key) |>
  collect() |>
  mutate(
    component = metric_key |>
      str_replace("extrisk_", "") |>
      str_replace("_ecoregion_rescaled", "") |>
      str_replace("_", " ")) |>
  select(programarea_key, component, value) |>
  arrange(programarea_key, component, value)

# add missing components with value 0
d_0 <- expand_grid(
  programarea_key = d_scores$programarea_key,
  component       = sort(unique(d_fl$component)),
  value           = 0) |>
  anti_join(
    d_fl,
    by = join_by(programarea_key, component))

d_fl <- d_fl |>
  bind_rows(d_0) |>
  arrange(programarea_key, component) |>
  mutate(even = 1)

d_fl <- d_fl |>
  mutate(
    programarea_lbl = factor(
      programarea_key,
      levels = d_scores$programarea_key, ordered = T),
    component = case_match(
      component,
      "primprod" ~ "Primary Production",
      .default   = component),
    component = str_to_title(component)) |>
  arrange(programarea_lbl, component)

# all flowers ----
d <- d_fl |>
  arrange(programarea_lbl, component) |>
  group_by(programarea_lbl) |>
  mutate(across(!where(is.character), as.double)) |>
  mutate(
    ymax = cumsum(even),
    ymin = lag(ymax, default = 0),
    xmax = value,
    xmin = 0)

for (theme_name in c("minimal", "linedraw", "min_line", "dark", "void")) {

  theme_fxn <- switch(
    theme_name,
    "minimal"  = theme_minimal,
    "linedraw" = theme_linedraw,
    "min_line" = function(...) {
      theme_minimal(...) %+replace%
        theme(
          panel.grid       = element_line(colour = "black"),
          panel.grid.major = element_line(linewidth = rel(0.1)),
          panel.grid.minor = element_line(linewidth = rel(0.05)))
    },
    "dark" = theme_dark,
    "void" = theme_void)

  g <- d |>
    ggplot() +
    geom_rect_interactive(aes(
      xmin    = xmin,
      xmax    = xmax,
      ymin    = ymin,
      ymax    = ymax,
      fill    = component,
      color   = "white",
      data_id = component),
      color = "white",
      alpha = 1) +
    scale_fill_manual(
      values = component_cols,
      name   = "Component") +
    facet_wrap(~programarea_lbl) +
    coord_polar(theta = "y") +
    xlim(c(-20, max(d_scores$score, d$xmax))) +
    annotate(
      "rect", xmin = -20, xmax = 0, ymin = 0, ymax = max(d$ymax),
      fill = "white", color = NA) +
    geom_text(
      aes(x, y, label = score),
      data = d_scores |>
        mutate(x = -20, y = 0),
      size     = 1.5,
      fontface = "bold",
      color    = ifelse(theme_name == "dark", "white", "black")) +
    theme_fxn() +
    theme(
      legend.position = "bottom",
      legend.title    = element_text(size = 8),
      legend.text     = element_text(size = 6),
      plot.margin     = unit(c(1, 1, 1, 1), "pt"),
      strip.text.x    = element_text(size = 6),
      axis.title.x    = element_blank(),
      axis.text.x     = element_blank(),
      axis.ticks.x    = element_blank(),
      axis.title.y    = element_blank(),
      axis.text.y     = element_blank(),
      axis.ticks.y    = element_blank())

  pra_fl_png <- glue("{dir_figs}/programarea_flower-plot_scores_{theme_name}{v_sfx}.png")
  pra_fl_pdf <- glue("{dir_figs}/programarea_flower-plot_scores_{theme_name}{v_sfx}.pdf")
  png(pra_fl_png, width = 1200, height = 800, res = 150)
  print(g)
  dev.off()
  pdf(pra_fl_pdf, width = 7, height = 5)
  print(g)
  dev.off()
}
```

## Primary Productivity by Program Area

```{r}
#| label: plot_primprod
metric_pp_keys <- glue("primprod_{c('avg', 'stddev')}")

d_pp <- tbl(con_sdm, "zone") |>
  filter(
    tbl == glue("ply_programareas_2026{v_sfx}"),
    fld == "programarea_key") |>
  select(programarea_key = value, zone_seq) |>
  left_join(
    tbl(con_sdm, "zone_metric"),
    by = "zone_seq") |>
  left_join(
    tbl(con_sdm, "metric") |>
      filter(metric_key %in% metric_pp_keys),
    by = "metric_seq") |>
  filter(
    !is.na(programarea_key),
    !is.na(metric_key),
    !is.na(value)) |>
  select(programarea_key, metric_key, value) |>
  arrange(programarea_key, metric_key) |>
  collect() |>
  left_join(
    tbl(con, glue("ply_programareas_2026{v_sfx}")) |>
      select(programarea_key, programarea_name) |>
      collect(),
    by = "programarea_key") |>
  filter(!is.na(programarea_name)) |>
  tidyr::pivot_wider(
    names_from  = metric_key,
    values_from = value) |>
  rename(
    avg = primprod_avg,
    sd  = primprod_stddev) |>
  mutate(
    across(where(is.numeric),
      ~ set_units(.x, mg/m^2/day) |>
        set_units(t/km^2/yr)))

pp_csv <- glue("{dir_figs}/programarea_primprod{v_sfx}.csv")
d_pp |>
  mutate(across(where(is.numeric), drop_units)) |>
  write_csv(pp_csv)

g <- d_pp |>
  arrange(desc(avg)) |>
  mutate(across(where(is.numeric), drop_units)) |>
  mutate(programarea_name = factor(programarea_name, levels = programarea_name)) |>
  ggplot(aes(x = programarea_name, y = avg)) +
  geom_bar(
    position = position_dodge(), stat = "identity", fill = "darkgreen") +
  geom_errorbar(aes(ymin = avg - sd, ymax = avg + sd), width = .4) +
  labs(
    title = NULL,
    x     = NULL,
    y     = expression("Areal NPP (metric tons C km"^-2~"yr"^-1*")")) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(
    axis.text.x = element_text(
      angle = 45, vjust = 1, hjust = 1))

pp_png <- glue("{dir_figs}/programarea_primprod{v_sfx}.png")
pp_pdf <- glue("{dir_figs}/programarea_primprod{v_sfx}.pdf")
png(pp_png, width = 1200, height = 800, res = 150)
print(g)
dev.off()
pdf(pp_pdf, width = 7, height = 5)
print(g)
dev.off()
```

## Copy to Downloads

```{r}
#| label: copy_to_downloads
dir_dl_figs <- glue("{dir_dl}/figures")
dir_create(dir_dl)
dir_create(dir_dl_figs)

# derived data files → downloads/
dl_data <- c(
  pra_gpkg, er_gpkg, er_pra_gpkg,
  er_pra_n_csv, pra_n_csv)

for (f in dl_data) {
  if (file.exists(f))
    file_copy(f, glue("{dir_dl}/{path_file(f)}"), overwrite = T)
}

# figures → downloads/figures/
fig_files <- dir_ls(dir_figs, regexp = "\\.(png|pdf|docx|xlsx|csv)$")
for (f in fig_files) {
  file_copy(f, glue("{dir_dl_figs}/{path_file(f)}"), overwrite = T)
}

message(glue("Copied {length(dl_data)} data files and {length(fig_files)} figure files to downloads/"))
```
