---
title: "OBIS-method runs compared: published `mpaeu`, our `og` / `ogm` and the controls"
author: "Ben Best"
date: "2026-10-08"
editor_options:
  chunk_output_type: console
---

## Why

`obis_range_constraint.qmd` scores the range rules A–E on one model run at a time (OBIS's published
models and, through `obis_range_constraint_{og,ogm}.qmd`, each of ours). This page puts the runs side by
side: how good each model is, how many cells each rule keeps per US EEZ subregion under each run, where the
regional effort test agrees, and what each run cost. It reads only the committed csvs under `data/og/`
(`scripts/og/cv_metrics.R`, `scripts/og/timings.R` and the scorecard notebooks write them), so it renders on
any machine.

| run | fit points | background | layers | species |
|---|---|---|---|---|
| `mpaeu` | OBIS's published models (their thinning, their tuning) | random cells of the masked stack | OBIS's depth choice | all seven |
| `ogp` | pinned to OBIS's fit points and hyperparameters (control) | random | surface | leatherback, right whale |
| `ogc` | our thinning, our tuning | random | surface | leatherback, right whale |
| `og` | our thinning, our tuning | target group = the species' ORDER (turtle records for turtles) | surface | six (loggerhead apart) |
| `ogm` | our thinning, our tuning | target group = air-breathing megafauna | surface | six (loggerhead running) |
| `ogmp` | pinned to OBIS's fit points (1,160), our tuning | target group = air-breathing megafauna | surface | loggerhead |

```{r}
#| label: setup
#| message: false
librarian::shelf(dplyr, ggplot2, glue, here, knitr, purrr, readr, stringr, tidyr, quiet = TRUE)
acros  <- c(mpaeu = "published", ogp = "ogp", ogc = "ogc", og = "og", ogm = "ogm", ogmp = "ogmp")
read_set <- function(what) imap_dfr(acros, \(f, a) {
  p <- here(glue("data/og/{f}_{what}.csv"))
  if (!file.exists(p)) return(NULL)
  read_csv(p, show_col_types = FALSE) |> mutate(acro = a, taxon = as.character(taxon)) })
card <- read_set("scorecard") |> mutate(acro = factor(acro, names(acros)))
rtest <- read_set("region_test")
cvm  <- read_csv(here("data/og/cv_metrics.csv"), show_col_types = FALSE) |> mutate(taxon = as.character(taxon))
tim  <- read_csv(here("data/og/timings.csv"),    show_col_types = FALSE) |> mutate(taxon = as.character(taxon))
common <- card |> distinct(taxon, common)
sp_order <- c("Loggerhead", "Green", "Hawksbill", "Kemp's ridley", "Leatherback", "Olive ridley", "North Atlantic right whale")
```

## Model quality

The pipeline reports two sets of metrics per algorithm: the spatial-block cross-validation (`cv`, what its
acceptance gate reads, CBI ≥ 0.3) and the independent evaluation records (`eval`, records held back before
the fit; for these species only `r min(cvm$n_eval)`–`r max(cvm$n_eval)` presences, so a single eval AUC is
noisy — read the pattern across runs, not one cell). The ensemble row is the mean over the accepted
algorithms.

```{r}
#| label: quality
#| tbl-cap: "Ensemble metrics per run: AUC / CBI / TSS at P10, on the independent evaluation records (eval) and on the spatial-block CV folds (cv)."
q <- cvm |> filter(method == "ensemble") |>
  mutate(v = sprintf("%.2f / %.2f / %.2f", auc, cbi, tss_p10), common = factor(common, levels = c("loggerhead", "green", "hawksbill", "kemps_ridley", "leatherback", "olive_ridley", "right_whale"))) |>
  select(common, origin, acro, v) |> pivot_wider(names_from = acro, values_from = v) |> arrange(common, desc(origin))
kable(q |> select(common, origin, any_of(names(acros))))
```

```{r}
#| label: accepted
#| tbl-cap: "Fit points, evaluation presences, background and the algorithms the gate accepted, per run."
kable(cvm |> filter(method != "ensemble", origin == "cv") |>
  summarise(n_fit = first(n_fit), n_eval = first(n_eval), background = first(background),
            algos_accepted = paste(method[accepted], collapse = "+"), algos_rejected = paste(method[!accepted], collapse = "+"),
            .by = c(common, acro)) |> arrange(common, acro))
```

The order-level target group fails as a background: for a turtle it is the turtle records, which for the
loggerhead are mostly its own 80,012 occurrences, so presences and background are the same places. Every
`og` run scores below its `ogm` twin on the evaluation records (leatherback 0.63 vs 0.90, hawksbill 0.75
vs 0.93), and the loggerhead `og` ensemble is below chance (0.42) while its CV CBI is 0.93–0.96 — the gate
accepted all three algorithms. The megafauna background (`ogm`) and the random background (`ogc`, `ogp`)
are in the same band as OBIS's published models.

## Cells kept per EEZ subregion

```{r}
#| label: kept
#| tbl-cap: "Cells kept per US EEZ subregion, rule P10 (the published threshold inside the fitted-record hull) and rule E (held-out threshold, uncertainty, regional effort test), by run."
kept <- card |> filter(rule %in% c("P10", "E"), str_starts(region, "EEZ")) |>
  mutate(region = str_remove(region, "EEZ "), common = factor(common, sp_order)) |>
  select(common, region, rule, acro, n_kept) |>
  pivot_wider(names_from = c(acro, rule), values_from = n_kept, names_sep = " ") |> arrange(common, region)
kable(kept, format.args = list(big.mark = ","))
```

```{r}
#| label: fig-kept
#| fig-cap: "Rule E cells kept per EEZ subregion and run (log scale; a missing bar is zero)."
#| fig-width: 10
#| fig-height: 6
card |> filter(rule == "E", str_starts(region, "EEZ"), n_kept > 0) |>
  mutate(region = str_remove(region, "EEZ "), common = factor(common, sp_order)) |>
  ggplot(aes(region, n_kept, fill = acro)) + geom_col(position = position_dodge(preserve = "single")) +
  scale_y_log10(labels = scales::comma) + facet_wrap(~ common, ncol = 4) +
  labs(x = NULL, y = "cells kept (rule E)", fill = "run") + theme_minimal() + theme(legend.position = "bottom")
```

```{r}
#| label: variants
#| tbl-cap: "Rule E and its two variants for the runs that have a held-out threshold: `E mega` runs the effort test on megafauna records, `E no guard` drops the ratio guard."
kable(card |> filter(rule %in% c("E", "E mega", "E no guard"), str_starts(region, "EEZ"), acro %in% c("og", "ogm")) |>
  mutate(region = str_remove(region, "EEZ "), common = factor(common, sp_order)) |>
  select(common, region, acro, rule, n_kept) |>
  pivot_wider(names_from = c(acro, rule), values_from = n_kept, names_sep = " ") |> arrange(common, region),
  format.args = list(big.mark = ","))
```

## Regional effort test

```{r}
#| label: verdicts
#| tbl-cap: "Verdict of the regional effort test (order-level effort, cell unit, ratio 0.25) per species x EEZ subregion and run."
v <- rtest |> filter(effort == "order", unit == "cell", str_starts(region, "EEZ")) |>
  left_join(common, by = "taxon") |>
  mutate(region = str_remove(region, "EEZ "), common = factor(common, sp_order),
         verdict = if_else(!is.na(label) & verdict == "dropped", paste0(verdict, " (", label, ")"), verdict)) |>
  select(common, region, acro, verdict) |> pivot_wider(names_from = acro, values_from = verdict) |> arrange(common, region)
kable(v |> select(common, region, any_of(names(acros))))
agree <- v |> filter(!is.na(og), !is.na(ogm)) |> summarise(n = n(), same = sum(og == ogm))
```

`og` and `ogm` reach the same verdict in `r agree$same` of `r agree$n` species × subregion cells; the test
reads records and effort, not the model, so a background change moves it only through the P10 footprint it
is applied to.

## Cost

```{r}
#| label: timings
#| tbl-cap: "Minutes per run (fit only, bootstrap excluded) and the share of the slowest step."
t <- tim |> summarise(total = first(total_mins), n_fit = first(n_fit),
                      slowest = step[which.max(mins)], pct = max(pct), .by = c(common, acro)) |>
  mutate(acro = factor(acro, names(acros)))
kable(t |> mutate(v = sprintf("%s (%s %s %%)", round(total), slowest, pct)) |> select(common, acro, v) |>
  pivot_wider(names_from = acro, values_from = v) |> arrange(common))
```

Maxent is ~91–95 % of a fit and roughly linear in fit points (5.5 k → 3 h 51, 8.5 k → 6 h 20, 24 k →
13 h 20); rf and xgboost take ≤ 10 min each. The loggerhead's 23,666 fit points (OBIS: 1,160) come from the
Mantel distance being drawn from a 1,000-record random sample, which sends the >15,000-point branch to a
grid prune at a different resolution.

## What this settles, and what it does not

- The ORDER-level target-group background is not usable for these species; carry the megafauna (`ogm`) or
  random (`ogc`) background forward. The `og` scorecard stays as the record of why.
- The acceptance gate (CV CBI ≥ 0.3) passed a model that is below chance on independent records; a floor on
  CV AUC or on the evaluation metrics would have caught it. That is a method change in the pipeline, not made.
- The loggerhead on the megafauna background with OBIS's 1,160 fit points (`ogmp`) is a model again
  (eval AUC 0.85, 90 min end to end against 875 for the 23,666-point `og` fit) and keeps no Alaska cell under
  P10 where the published model keeps 55,807; its P10 Pacific footprint (198,141) is three times the
  published one (61,327).
- Open (Ben): the loggerhead thinning (cap the fit points, pin OBIS's `fitocc` as `ogmp` does, or a fixed
  thinning distance); which background to carry; the `ratio = 0.25` guard and the 0.5° unit of the effort test;
  whether the leatherback's P10 count of 51 vs 57 is acceptable.
