---
title: "Publish native-format input surfaces — range PMTiles + AquaMaps COGs"
msens:
  target_name: publish_native
  workflow_type: release
  dependency: [merge_taxon]
  output: data/manifests/publish_native.json
editor_options:
  chunk_output_type: console
---

Phase 4b: publish each **raw input** model in a cloud-native, pyramided format so the
species app can overlay the whole (often global) range a source contributes — surfaces
far too large (up to ~19M cells) for the dense per-cell SQL tiler:

- **vector ranges** (`rng_iucn`, `rng_fws`, `ca_nmfs`, `ch_fws`, `ch_nmfs`, `bl`,
  `rng_turtle_swot_dps`) → **one PMTiles per dataset**, features tagged with `mdl_key`
  (client filters by it), served from the file host `/pmtiles`.
- **AquaMaps** (`am`, raster) → **one COG per model** (bbox-cropped, overviews), served
  from S3 by titiler-v8 `/cog`.

Both drive a `native_asset` registry (`mdl_key → asset_type + url`) that STAC + the app read.
We only publish inputs that feed a **served taxon** (`is_valid_usa & is_marine`, non
reptile/amphibian) — the set the species app actually lists.

**Flags:** `REDO_NATIVE=1` (rebuild PMTiles + re-sort am), `NATIVE_SKIP_AM` / `NATIVE_SKIP_PMTILES`,
`NATIVE_TEST_N=<n>` (only first n am COGs, smoke test), `NATIVE_NO_S3` (stage locally, skip upload),
`NATIVE_WORKERS=<n>`.

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, furrr, glue, here, jsonlite, logger, purrr,
                 readr, sf, stringr, terra, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))
source(here("../msens/R/publish.R"))   # override any stale install

dir_atlas  <- glue("{dir_big_v}/marine-atlas")
merge_db   <- glue("{dir_atlas}/merge.duckdb")
dir_native <- glue("{dir_atlas}/native"); dir_create(dir_native)
dir_cog    <- glue("{dir_native}/am");     dir_create(dir_cog)
dir_pmt    <- glue("{dir_native}/pmtiles"); dir_create(dir_pmt)
manifest   <- here("data/manifests/publish_native.json"); dir_create(path_dir(manifest))
stopifnot(all(file_exists(c(sdm_db, merge_db, cellid_tif))))

redo        <- nzchar(Sys.getenv("REDO_NATIVE"))
do_s3       <- !nzchar(Sys.getenv("NATIVE_NO_S3"))
skip_am     <- nzchar(Sys.getenv("NATIVE_SKIP_AM"))
skip_pmt    <- nzchar(Sys.getenv("NATIVE_SKIP_PMTILES"))
test_n      <- as.integer(Sys.getenv("NATIVE_TEST_N", "0"))
n_workers   <- as.integer(Sys.getenv("NATIVE_WORKERS", "6"))
s3_ver      <- glue("{s3_atlas}/{ver}")
s3_http     <- glue("https://s3.us-east-1.amazonaws.com/{sub('^s3://', '', s3_ver)}")
pmt_base    <- glue("https://file.marinesensitivity.org/pmtiles/{ver}")

safe_key <- function(k) gsub("[^A-Za-z0-9._-]", "_", k)   # mdl_key -> filesystem/URL-safe
```

## Served raw mdl_keys (the set the app lists)

```{r}
#| label: served

con <- dbConnect(duckdb(), config = list(threads = "6", memory_limit = "10GB"))
dbExecute(con, glue("SET temp_directory='{dir_atlas}/duckdb_tmp'"))
dbExecute(con, glue("ATTACH '{merge_db}' AS m (READ_ONLY)"))
served <- dbGetQuery(con, "
  SELECT DISTINCT tm.mdl_key, split_part(tm.mdl_key,'|',1) AS ds_key
  FROM m.taxon_model tm
  JOIN m.taxon t USING(ms_merge_key)
  WHERE t.is_valid_usa AND t.is_marine AND t.sp_cat NOT IN ('reptile','amphibian')")
dbDisconnect(con, shutdown = TRUE)
served_by_ds <- split(served$mdl_key, served$ds_key)
knitr::kable(served |> count(ds_key, name = "n_mdl") |> arrange(desc(n_mdl)))
```

## Vector ranges → per-model PMTiles

Each dataset's source polygons are read, `mdl_key` derived from the documented source id,
then **inner-joined to the served set** (reproducing each ingest's presence/marine/category
filter — only ingested-and-served models survive the join) and split into **one PMTiles per
model** at `{ds}/{sp_id}.pmtiles`. A per-model file — like the per-model AquaMaps COGs — needs
no client-side filter and its low-zoom tiles stay a few KB. Packing thousands of overlapping
global ranges into one per-dataset archive overflowed the z0–z2 world tiles past maplibre's
per-tile budget, so a selected range silently failed to draw at global zoom.

IUCN's thousands of ranges are read from a **v8 indexed GeoPackage** (`native/src/rng_iucn.gpkg`,
built once from the source shapefiles with the ingest's presence/marine/category filter,
dissolved per `id_no`, indexed) — the derived per-group gpkgs are stale v7 artifacts (410 served
species short).

```{r}
#| label: pmtiles-specs

# reader per dataset: returns an sf with a `mdl_key` column (unfiltered — the served
# inner-join happens after). paths mirror the ingest notebooks (see Explore recon).
Sys.setenv(OGR_STROKE_CURVE = "TRUE")   # linearize BOTW MULTISURFACE curves
slug <- function(x) gsub("^_|_$", "", gsub("[^A-Za-z0-9]+", "_", x))

read_iucn <- function() {
  # read the v8 indexed ranges gpkg (one dissolved feature per served id_no), built by the
  # `iucn-src` chunk below — fast + off-Drive vs. crawling thousands of source shapefiles.
  gpkg <- glue("{dir_native}/src/rng_iucn.gpkg")
  stopifnot("run the iucn-src chunk first" = file_exists(gpkg))
  st_read(gpkg, quiet = TRUE) |> transmute(mdl_key = as.character(mdl_key)) |> st_zm(drop = TRUE)
}
read_fws_rng <- function() {
  shps <- glue("{dir_raw}/fws.gov/usfws_complete_species_current_range/usfws_complete_species_current_range_{1:2}.shp")
  map_dfr(shps[file_exists(shps)], function(s) {
    x <- st_read(s, quiet = TRUE); names(x) <- tolower(names(x))
    x |> transmute(mdl_key = glue("rng_fws|{spcode}"))          # source is NAD83 -> publish_pmtiles reprojects
  })
}
read_ca_nmfs <- function() {
  s <- glue("{dir_raw}/fisheries.noaa.gov/core-areas/shapefile_Rices_whale_core_distribution_area_Jun19_SERO/shapefile_Rices_whale_core_distribution_area_Jun19_SERO.shp")
  st_read(s, quiet = TRUE) |> transmute(mdl_key = "ca_nmfs|Balaenoptera_ricei")
}
read_ch_fws <- function() {
  x <- st_read(glue("{dir_raw}/fws.gov/crithab_all_layers/crithab_poly.shp"), quiet = TRUE)
  names(x) <- tolower(names(x)); x |> transmute(mdl_key = glue("ch_fws|{spcode}"))
}
read_ch_nmfs <- function() {
  x <- st_read(glue("{dir_raw}/fisheries.noaa.gov/ply.gpkg"), layer = "ply", quiet = TRUE)
  x |> transmute(mdl_key = glue("ch_nmfs|{slug(SCIENAME)}"))
}
read_turtle <- function() {
  codes <- c("CC","CM","DC","EI","LK","LO")
  map_dfr(codes, function(cd) {
    s <- glue("{dir_raw}/swot_seamap.env.duke.edu/swot_distribution/Global_Distribution_{cd}.shp")
    if (!file_exists(s)) return(NULL)
    st_read(s, quiet = TRUE) |> transmute(mdl_key = glue("rng_turtle_swot_dps|{cd}")) |> st_zm(drop = TRUE)
  })
}
read_bl <- function(sisids) {
  gpkg <- c(glue("{dir_big}/../raw/BOTW_2024_2.gpkg"),
            glue("{dir_raw}/birdlife.org/BOTW_GPKG_2024_2/BOTW_2024_2.gpkg"))
  gpkg <- gpkg[file_exists(gpkg)][1]; stopifnot(!is.na(gpkg))
  ids  <- paste(unique(sub("^bl\\|", "", sisids)), collapse = ",")
  q    <- glue("SELECT sisid, geom FROM all_species WHERE sisid IN ({ids}) AND presence IN (1,2,3)")
  st_read(gpkg, query = q, quiet = TRUE) |> transmute(mdl_key = glue("bl|{sisid}"))
}

pmt_specs <- list(
  # fast/small first (seconds each), then bl (one local gpkg query), then rng_iucn
  # LAST (thousands of shapefiles off Google Drive — minutes even warm).
  ca_nmfs             = read_ca_nmfs,
  ch_fws              = read_ch_fws,
  ch_nmfs             = read_ch_nmfs,
  rng_turtle_swot_dps = read_turtle,
  rng_fws             = read_fws_rng,
  bl                  = function() read_bl(served_by_ds[["bl"]]),
  rng_iucn            = read_iucn)
```

### IUCN v8 indexed ranges GeoPackage

Built once from the source shapefiles with the ingest's presence/marine/category filter,
dissolved to one feature per served `id_no`, indexed on `id_no`. `read_iucn()` then reads it
in one fast, off-Drive query (the derived per-group gpkgs are stale v7 and miss 410 species).

```{r}
#| label: iucn-src

iucn_gpkg <- glue("{dir_native}/src/rng_iucn.gpkg"); dir_create(path_dir(iucn_gpkg))
if (!file_exists(iucn_gpkg) || redo) {
  iucn_ok <- c("CR","EN","VU","NT","LC","DD")
  ids_srv <- as.integer(sub("^rng_iucn\\|", "", served_by_ds[["rng_iucn"]]))
  log_info("building {path_file(iucn_gpkg)} for {length(ids_srv)} served ids (source shapefiles) …")
  shps <- dir_ls(glue("{dir_raw}/iucnredlist.org"), glob = "*.shp", recurse = TRUE)
  gi <- map_dfr(shps, function(s) {
    y <- tryCatch(st_read(s, quiet = TRUE), error = function(e) NULL)
    if (is.null(y)) return(NULL); names(y) <- tolower(names(y))
    if (!all(c("id_no","presence","marine","category") %in% names(y))) return(NULL)
    y |> filter(presence %in% 1:3, tolower(marine) == "true", category %in% iucn_ok,
                id_no %in% ids_srv) |>
      transmute(id_no = as.integer(id_no)) |> st_zm(drop = TRUE)
  })
  # keep each presence polygon as its own feature (same mdl_key) — publish_pmtiles_models
  # groups a model's features into one file; dissolving would trip s2 on invalid rings.
  gi <- gi |> mutate(mdl_key = as.character(glue("rng_iucn|{id_no}")), ds_key = "rng_iucn")
  gi <- gi[!st_is_empty(gi), ]
  st_write(gi, iucn_gpkg, layer = "rng_iucn", quiet = TRUE, delete_dsn = TRUE)
  system2("ogrinfo", c(shQuote(iucn_gpkg), "-sql",
                       shQuote("CREATE INDEX idx_idno ON rng_iucn(id_no)")))
  log_info("wrote {path_file(iucn_gpkg)}: {nrow(gi)} species")
} else log_info("{path_file(iucn_gpkg)} exists — skip (REDO_NATIVE=1 to rebuild)")
```

```{r}
#| label: pmtiles-build

pmt_only <- Sys.getenv("NATIVE_PMT_DS")                    # optional comma-list to restrict (partial reruns/tests)
pmt_ds   <- if (nzchar(pmt_only)) intersect(names(pmt_specs), strsplit(pmt_only, ",")[[1]]) else names(pmt_specs)
pmt_rows <- list()
if (!skip_pmt) for (ds in pmt_ds) {
  keys <- served_by_ds[[ds]]
  if (is.null(keys)) { log_warn("no served keys for {ds} — skip PMTiles"); next }
  dir_ds <- glue("{dir_pmt}/{ds}"); dir_create(dir_ds)
  # one PMTiles per model (resumable — skips existing unless REDO_NATIVE). z6 is plenty for
  # coarse range overlays; a single range tiles to a few KB so no per-dataset density blowup.
  if (ds == "rng_iucn") {
    # IUCN's full-res source is ~7 GB — read each model on demand from the indexed gpkg
    # (never load it all into memory), one index hit per species.
    log_info("{ds}: {length(keys)} models -> per-model PMTiles from indexed gpkg ({n_workers} workers)")
    built <- publish_pmtiles_from_gpkg(iucn_gpkg, "rng_iucn", keys, dir_ds,
                                       workers = n_workers, redo = redo, minzoom = 0, maxzoom = 6)
  } else {
    log_info("reading source polygons for {ds} ({length(keys)} served models) …")
    x <- pmt_specs[[ds]]()
    x <- x |> mutate(mdl_key = as.character(mdl_key)) |>
      filter(mdl_key %in% keys) |> mutate(ds_key = ds)
    x <- x[!st_is_empty(x), ]
    if (!nrow(x)) { log_warn("{ds}: 0 features after served-join — skip"); next }
    log_info("{ds}: {nrow(x)} features, {n_distinct(x$mdl_key)} models -> per-model PMTiles ({n_workers} workers)")
    built <- publish_pmtiles_models(x, dir_ds, layer = ds, workers = n_workers,
                                    redo = redo, minzoom = 0, maxzoom = 6)
  }
  if (is.null(built) || !nrow(built)) { log_warn("{ds}: no PMTiles built"); next }
  pmt_rows[[ds]] <- built |> transmute(
    mdl_key, ds_key = ds, asset_type = "pmtiles",
    # per-model URL; ?v={mtime} busts the browser/PMTiles-protocol cache on rebuild
    # (Caddy ignores the query and serves the file).
    asset_url = glue("{pmt_base}/{ds}/{path_file(file)}?v={as.integer(file_info(file)$modification_time)}"),
    source_layer = ds)
}
pmt_reg <- bind_rows(pmt_rows)   # now ONE ROW PER MODEL (mdl_key, asset_url)
if (nrow(pmt_reg)) knitr::kable(pmt_reg |> count(ds_key, name = "n_models"))
```

## AquaMaps → per-model COGs

Consolidate the served `am` cells by **integer** key (`mid`) via `PARTITION_BY` — one small
Parquet dir per model (`_am_parts/mid={mid}/`, columns `cell_id,val` only). This drops the
40-byte `mdl_key` string from the ~12B-row intermediate and needs no global sort (a plain
ORDER BY would spill ~500GB), so each per-model read is a direct partition scan. Then paint
each model onto the global grid, cropped to its bbox, as a COG with overviews. Parallel +
resumable (skip existing COGs).

```{r}
#| label: am-partition

# test mode uses a separate parts dir + only the test subset, so it never poisons a full run
am_keys <- if (length(served_by_ds[["am"]])) sort(served_by_ds[["am"]]) else character(0)
am_take <- if (test_n > 0) head(am_keys, test_n) else am_keys
am_map  <- tibble(mdl_key = am_take, mid = seq_along(am_take))       # stable int key per model
am_parts<- glue("{dir_native}/_am_parts{if (test_n > 0) '_test' else ''}")
if (!skip_am && length(am_take)) {
  if (!dir_exists(am_parts) || redo) {
    if (dir_exists(am_parts)) dir_delete(am_parts)
    log_info("partitioning {length(am_take)} served am models by mid -> _am_parts …")
    con <- dbConnect(duckdb(), config = list(threads = "6", memory_limit = "10GB"))
    dbExecute(con, glue("SET temp_directory='{dir_atlas}/duckdb_tmp'"))
    dbExecute(con, "SET partitioned_write_max_open_files=400")
    dbWriteTable(con, "am_map", am_map, overwrite = TRUE)
    dbExecute(con, glue("COPY (
      SELECT map.mid, p.cell_id, p.val
      FROM read_parquet('{dir_atlas}/dist/dataset=am/*.parquet') p
      JOIN am_map map ON p.mdl_key = map.mdl_key
    ) TO '{am_parts}' (FORMAT parquet, COMPRESSION zstd, PARTITION_BY (mid), OVERWRITE_OR_IGNORE)"))
    dbDisconnect(con, shutdown = TRUE)
    log_info("_am_parts: {length(dir_ls(am_parts))} partitions, {round(sum(file_size(dir_ls(am_parts, recurse=TRUE, type='file')))/1e9,1)} GB")
  } else log_info("_am_parts exists — skip repartition (REDO_NATIVE=1 to force)")
}
```

```{r}
#| label: am-cogs

am_reg <- tibble()
if (!skip_am && length(am_take) && dir_exists(am_parts)) {
  todo <- am_map[!file_exists(glue("{dir_cog}/{safe_key(am_map$mdl_key)}.tif")), ]
  log_info("am COGs: {nrow(am_map)} target, {nrow(todo)} to build ({nrow(am_map)-nrow(todo)} exist)")

  grid <- grid_spec(rast(cellid_tif))
  build_batch <- function(rows) {
    suppressMessages({library(DBI); library(duckdb); library(terra)})
    con <- dbConnect(duckdb())
    on.exit(dbDisconnect(con, shutdown = TRUE))
    for (i in seq_len(nrow(rows))) {
      k <- rows$mdl_key[i]; f <- sprintf("%s/%s.tif", dir_cog, safe_key(k))
      if (file.exists(f)) next
      pf <- sprintf("%s/mid=%d", am_parts, rows$mid[i])
      if (!dir.exists(pf)) next
      d <- dbGetQuery(con, sprintf("SELECT cell_id, val FROM read_parquet('%s/*.parquet')", pf))
      if (nrow(d)) msens::publish_cog(d$cell_id, d$val, f, grid)
    }
    nrow(rows)
  }

  if (nrow(todo)) {
    plan(multisession, workers = n_workers)
    batches <- split(todo, (seq_len(nrow(todo)) - 1) %% n_workers)
    future_map(batches, build_batch,
               .options = furrr_options(globals = c("am_parts","dir_cog","grid","safe_key"),
                                        packages = c("msens"), seed = TRUE))
    plan(sequential)
  }
  built <- glue("{dir_cog}/{safe_key(am_map$mdl_key)}.tif")
  am_reg <- tibble(
    mdl_key = am_map$mdl_key, ds_key = "am", asset_type = "cog",
    asset_url = glue("{s3_http}/native/am/{safe_key(am_map$mdl_key)}.tif"),
    rescale_min = 1L, rescale_max = 100L, colormap = "spectral_r")[file_exists(built), ]
  log_info("am COGs on disk: {nrow(am_reg)} / {nrow(am_map)}")
}
```

## native_asset registry → sdm.duckdb

```{r}
#| label: registry

# per-input-model asset rows: am COGs AND vector PMTiles are now both per-model — pmt_reg
# already carries one (mdl_key, asset_url) row per model (its own PMTiles file).
pmt_model_rows <- if (nrow(pmt_reg)) pmt_reg else tibble()

asset_info <- bind_rows(am_reg, pmt_model_rows) |>
  mutate(across(everything(), ~ if (is.factor(.)) as.character(.) else .))

# per-model geographic bbox (cell_id row/col arithmetic on the global grid) so the app fits
# to each input's actual extent — NOT the whole world (lands on the globe antipode). am from
# the compact _am_parts (mid->mdl_key); vector ranges from their small dist partitions.
bbox_cols <- c("xmin","xmax","ymin","ymax")
bb_am <- tibble()
if (nrow(am_reg) && exists("am_parts") && dir_exists(am_parts)) {
  con <- dbConnect(duckdb(), config = list(threads = "6"))
  bb_am <- dbGetQuery(con, glue("
    WITH rc AS (SELECT mid, (cell_id-1) % 7200 AS col, (cell_id-1) // 7200 AS row
                FROM read_parquet('{am_parts}/**/*.parquet', hive_partitioning=true))
    SELECT mid,
      -180 + min(col)*0.05       AS xmin, -180 + (max(col)+1)*0.05 AS xmax,
       90 - (max(row)+1)*0.05    AS ymin,   90 - min(row)*0.05     AS ymax
    FROM rc GROUP BY mid")) |>
    dplyr::left_join(am_map, by = "mid") |> dplyr::select(-mid)
  dbDisconnect(con, shutdown = TRUE)
}
bb_vec <- tibble()
vec_ds <- unique(pmt_model_rows$ds_key)                          # the pmtiles datasets
if (length(vec_ds)) {
  con   <- dbConnect(duckdb(), config = list(threads = "6"))
  globs <- paste0("'", dir_atlas, "/dist/dataset=", vec_ds, "/*.parquet'", collapse = ", ")
  dbWriteTable(con, "vk", tibble(mdl_key = unique(pmt_model_rows$mdl_key)), overwrite = TRUE)
  bb_vec <- dbGetQuery(con, glue("
    WITH rc AS (SELECT p.mdl_key, (p.cell_id-1) % 7200 AS col, (p.cell_id-1) // 7200 AS row
                FROM read_parquet([{globs}]) p SEMI JOIN vk USING(mdl_key))
    SELECT mdl_key,
      -180 + min(col)*0.05       AS xmin, -180 + (max(col)+1)*0.05 AS xmax,
       90 - (max(row)+1)*0.05    AS ymin,   90 - min(row)*0.05     AS ymax
    FROM rc GROUP BY mdl_key"))
  dbDisconnect(con, shutdown = TRUE)
}
bb_all <- bind_rows(bb_am, bb_vec)
if (nrow(bb_all)) asset_info <- asset_info |> left_join(bb_all, by = "mdl_key")

# attach ms_merge_key (the merged taxon key) so the species app can join a taxon
# directly to its input surfaces. A raw model may feed >1 merged taxon -> one row each.
con <- dbConnect(duckdb(sdm_db))
dbExecute(con, glue("ATTACH '{merge_db}' AS m (READ_ONLY)"))
tm <- dbGetQuery(con, "SELECT DISTINCT tm.mdl_key, tm.ms_merge_key
  FROM m.taxon_model tm JOIN m.taxon t USING(ms_merge_key)
  WHERE t.is_valid_usa AND t.is_marine AND t.sp_cat NOT IN ('reptile','amphibian')")
dbExecute(con, "DETACH m")

native_asset <- asset_info |>
  inner_join(tm, by = "mdl_key", relationship = "many-to-many") |>
  relocate(ms_merge_key)

dbWriteTable(con, "native_asset", native_asset, overwrite = TRUE)
smry <- dbGetQuery(con, "SELECT asset_type, count(*) n_rows, count(DISTINCT mdl_key) n_models,
  count(DISTINCT ms_merge_key) n_taxa, count(DISTINCT ds_key) n_ds FROM native_asset GROUP BY 1")
dbDisconnect(con, shutdown = TRUE)
knitr::kable(smry)
```

## Sync → S3 (COGs) + file host (PMTiles)

```{r}
#| label: sync

if (do_s3) {
  if (nrow(am_reg)) {
    st <- system2("aws", c("s3","sync", dir_cog, glue("{s3_ver}/native/am"),
                           "--only-show-errors","--no-progress"), stdout=TRUE, stderr=TRUE)
    if (!is.null(attr(st,"status")) && attr(st,"status")!=0) stop(paste(st, collapse="\n"))
    log_info("synced am COGs -> {s3_ver}/native/am")
  }
  # PMTiles -> server file host (/share/data/derived/pmtiles/{ver}); also keep on S3
  if (nrow(pmt_reg)) {
    system2("ssh", c("msens", glue("mkdir -p /share/data/derived/pmtiles/{ver}")))
    system2("rsync", c("-aq", glue("{dir_pmt}/"), glue("msens:/share/data/derived/pmtiles/{ver}/")))
    st <- system2("aws", c("s3","sync", dir_pmt, glue("{s3_ver}/native/pmtiles"),
                           "--only-show-errors","--no-progress"), stdout=TRUE, stderr=TRUE)
    log_info("synced PMTiles -> file host /pmtiles/{ver} + {s3_ver}/native/pmtiles")
  }
} else log_info("NATIVE_NO_S3 set — staged locally only ({dir_native})")
```

## Manifest

```{r}
#| label: manifest

write_json(list(
  target = "publish_native", ver = ver,
  pmtiles_datasets = if (nrow(pmt_reg)) pmt_reg$ds_key else character(0),
  am_cogs = nrow(am_reg),
  native_assets = nrow(native_asset), pushed_s3 = do_s3,
  built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
manifest
```
