---
title: "Release marine-atlas — versioned Parquet → S3, view-DB serving, STAC"
msens:
  target_name: release_marine-atlas
  workflow_type: release
  dependency: [auto]
  output: data/manifests/release_marine-atlas.json
editor_options:
  chunk_output_type: console
---

Publish the **v8 marine-atlas** and stand up its serving, reproducibly:

1. **Data → S3** `s3://oceanmetrics.io-public/marine-atlas/{ver}/`:
   - `tables/` — derived scoring tables (`cell`, `taxon`, `dataset`, `model`,
     `cell_metric`, `zone*`, `metric`).
   - `dist_merged/` — merged per-taxon surfaces (`ms_merge|…`).
   - `serve/model_cell.parquet` — the **serving-optimized** surface: one file, globally
     **sorted by `mdl_key`** (row-group zone-map pruning) so a titiler tile does an
     HTTP-range point read; single file ⇒ anonymous GET (no S3 LIST/creds).
   - `registry/` — `dataset`/`model` Parquet.
   - `dist/` — raw per-dataset surfaces (~74G) — opt-in (`RELEASE_RAW=1`).
2. **View DB** — a **tiny** `serve.duckdb` whose tables are *views* over the S3 Parquet
   (path-style HTTPS URLs). Titiler reads this; the big data never rsyncs to the server.
3. **STAC** — `stac_build(version)` → deploy the catalog to the file host `/stac`.
4. **Deploy** (`RELEASE_DEPLOY=1`) — rsync the KB view DB + STAC to `msens1`, (re)build the
   parallel `titiler-v8` service, restart caddy. v7 serving is untouched (A/B).

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, quiet = T)
source(here("libs/paths.R"))
dir_atlas <- glue("{dir_big_v}/marine-atlas")
tbl_dir   <- glue("{dir_atlas}/tables");  dir_create(tbl_dir)
serve_dir <- glue("{dir_atlas}/serve");   dir_create(serve_dir)
manifest  <- here("data/manifests/release_marine-atlas.json"); dir_create(path_dir(manifest))
s3_ver    <- glue("{s3_atlas}/{ver}")                      # s3://…/marine-atlas/v8
s3_http   <- glue("https://s3.us-east-1.amazonaws.com/{sub('^s3://', '', s3_ver)}")  # path-style
do_s3     <- !nzchar(Sys.getenv("RELEASE_NO_S3"))
do_raw    <-  nzchar(Sys.getenv("RELEASE_RAW"))
do_deploy <-  nzchar(Sys.getenv("RELEASE_DEPLOY"))
# server targets (parallel titiler-v8)
srv_repo  <- "/share/github/MarineSensitivity/server"
srv_viewdb<- glue("/share/data/big/{ver}/serve.duckdb")
srv_stac  <- glue("/share/data/derived/stac/{ver}")
titiler_v8<- glue("https://titiler-{ver}.marinesensitivity.org/msens")
stopifnot("run build_registry first" = {
  con0 <- dbConnect(duckdb(sdm_db, read_only = TRUE)); on.exit(dbDisconnect(con0, shutdown = TRUE))
  all(c("dataset", "model", "model_cell") %in% dbListTables(con0)) })
```

## Stage derived tables + serving surface → Parquet

```{r}
#| label: stage

con <- dbConnect(duckdb(sdm_db, read_only = TRUE))
dbExecute(con, "PRAGMA memory_limit='12GB'"); dbExecute(con, "PRAGMA threads=6")
dbExecute(con, glue("PRAGMA temp_directory='{dir_atlas}/duckdb_tmp'"))

# small/derived scoring tables (model_cell is released as dist_merged + serve/, not here).
# native_asset (mdl_key -> COG/PMTiles input surfaces) is included when publish_native has run.
rel_tables <- c("cell", "taxon", "dataset", "model", "metric",
                "cell_metric", "zone", "zone_cell", "zone_metric", "native_asset")
rel_tables <- intersect(rel_tables, dbListTables(con))
for (t in rel_tables)
  dbExecute(con, glue("COPY (SELECT * FROM {t}) TO '{tbl_dir}/{t}.parquet' (FORMAT parquet, COMPRESSION zstd)"))

# serving surface: ONE file, globally sorted by mdl_key -> row-group min/max stats let a
# titiler point query (WHERE mdl_key='…') fetch only the matching row groups over HTTP range.
# skip the ~3GB re-sort if it already exists (model_cell rarely changes); RELEASE_REDO_SERVE=1 forces.
serve_pq <- glue("{serve_dir}/model_cell.parquet")
if (!file_exists(serve_pq) || nzchar(Sys.getenv("RELEASE_REDO_SERVE"))) {
  dbExecute(con, glue("COPY (SELECT mdl_key, cell_id, val FROM model_cell ORDER BY mdl_key)
    TO '{serve_pq}' (FORMAT parquet, COMPRESSION zstd, ROW_GROUP_SIZE 100000)"))
} else {
  log_info("serve/model_cell.parquet exists — skipping re-sort (RELEASE_REDO_SERVE=1 to force)")
}
dbDisconnect(con, shutdown = TRUE)
knitr::kable(tibble(
  file = c(glue("tables/{rel_tables}.parquet"), "serve/model_cell.parquet"),
  mb   = round(file_size(c(glue("{tbl_dir}/{rel_tables}.parquet"),
                           glue("{serve_dir}/model_cell.parquet"))) / 1e6, 1)))
```

## Sync data → S3

```{r}
#| label: s3

sync <- function(local, prefix) {
  if (!dir_exists(local)) { log_warn("skip (missing): {local}"); return(invisible()) }
  st <- system2("aws", c("s3", "sync", local, glue("{s3_ver}/{prefix}"),
                         "--only-show-errors", "--no-progress"), stdout = TRUE, stderr = TRUE)
  if (!is.null(attr(st, "status")) && attr(st, "status") != 0) stop(paste(st, collapse = "\n"))
  log_info("synced {path_file(local)} -> {s3_ver}/{prefix}")
}
if (do_s3) {
  sync(tbl_dir,                         "tables")
  sync(serve_dir,                       "serve")
  sync(glue("{dir_atlas}/dist_merged"), "dist_merged")
  sync(glue("{dir_atlas}/registry"),    "registry")
  if (do_raw) sync(glue("{dir_atlas}/dist"), "dist")
} else log_info("RELEASE_NO_S3 set — staged locally only")
```

## Build the tiny view DB (views over the S3 Parquet)

```{r}
#| label: viewdb

viewdb <- glue("{serve_dir}/serve.duckdb")
if (file_exists(viewdb)) file_delete(viewdb)
cv <- dbConnect(duckdb(viewdb))
dbExecute(cv, "INSTALL httpfs; LOAD httpfs;")
# explicit path-style HTTPS single-file URLs -> anonymous GET + HTTP range (the dotted bucket
# breaks virtual-hosted TLS, so never use s3://… here). model_cell = the sorted serve file.
dbExecute(cv, glue("CREATE VIEW model_cell AS SELECT * FROM read_parquet('{s3_http}/serve/model_cell.parquet')"))
for (t in rel_tables)
  dbExecute(cv, glue("CREATE VIEW {t} AS SELECT * FROM read_parquet('{s3_http}/tables/{t}.parquet')"))
dbDisconnect(cv, shutdown = TRUE)
log_info("view DB {viewdb} ({round(file_size(viewdb)/1e3,1)} KB) — views over {s3_http}")
```

## Emit STAC catalog (tile links → titiler-{ver})

```{r}
#| label: stac

suppressWarnings(try(library(msens), silent = TRUE))
source(here("../msens/R/stac.R"))   # override any stale install with the current (v8) generators
stac_dir <- glue("{dir_atlas}/stac")
cfg <- stac_cfg(ver); cfg$titiler_base <- titiler_v8
con <- dbConnect(duckdb(sdm_db, read_only = TRUE))
stac_build(version = ver, dir_out = stac_dir, cfg = cfg, con = con)
dbDisconnect(con, shutdown = TRUE)
log_info("STAC catalog: {stac_dir}/{ver}/collection.json")
```

## Deploy to msens1 (parallel titiler-{ver})

```{r}
#| label: deploy

if (!do_deploy) {
  log_info("RELEASE_DEPLOY unset — skipping server deploy (S3 data + local view DB/STAC ready)")
} else {
# 1. the KB view DB + the STAC version subtree (data stays on S3 — no big rsync)
system2("ssh", c("msens", glue("mkdir -p {path_dir(srv_viewdb)} {srv_stac}")))
system2("rsync", c("-aq", viewdb, glue("msens:{srv_viewdb}")))
system2("rsync", c("-aq", glue("{stac_dir}/{ver}/"), glue("msens:{srv_stac}/")))
# add this version as a child of the STAC root (keep other versions)
system2("ssh", c("msens", shQuote(glue(
  "python3 - <<'PY'\nimport json\np='/share/data/derived/stac/catalog.json'\nc=json.load(open(p))\nh=[l['href'] for l in c['links'] if l.get('rel')=='child']\nif './{ver}/collection.json' not in h:\n c['links'].append({{'rel':'child','href':'./{ver}/collection.json'}}); json.dump(c,open(p,'w'),indent=2)\nPY"))))
# 2. (re)build + start the parallel titiler-{ver}; restart caddy for its route
system2("ssh", c("msens", glue(
  "cd {srv_repo} && git pull --quiet && docker compose up -d --build titiler-{ver} && docker compose restart caddy")))
# 3. smoke test
sql <- glue("SELECT cell_id, val AS value FROM model_cell WHERE mdl_key='{dbGetQuery(dbConnect(duckdb(sdm_db,read_only=TRUE)),\"SELECT mdl_key FROM model WHERE ds_key='ms_merge' LIMIT 1\")$mdl_key}'")
b64 <- sdm_sql_b64(sql)
code <- system2("curl", c("-s","-o","/dev/null","-w","%{http_code}",
  glue("{titiler_v8}/statistics?sql={b64}")), stdout = TRUE)
log_info("titiler-{ver} /statistics smoke test: HTTP {code}")
}
```

## Manifest

```{r}
#| label: manifest

write_json(
  list(target = "release_marine-atlas", ver = ver, s3 = s3_ver,
       tables = rel_tables, serve = "serve/model_cell.parquet (sorted by mdl_key)",
       view_db = "serve.duckdb (views over S3)", stac = glue("{s3_http} + file host /stac/{ver}"),
       titiler = titiler_v8, raw_included = do_raw, pushed_s3 = do_s3, deployed = do_deploy,
       built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
manifest
```
