---
title: "Release marine-atlas — versioned Parquet → S3"
msens:
  target_name: release_marine-atlas
  workflow_type: release
  dependency: [auto]
  output: data/manifests/release_marine-atlas.json
editor_options:
  chunk_output_type: console
---

Publish the **v8 marine-atlas** as partitioned Parquet to
`s3://oceanmetrics.io-public/marine-atlas/{ver}/` — a self-describing, versioned release
consumable directly over `httpfs` (DuckDB) or any Parquet reader:

- **`tables/`** — the derived scoring tables: `cell` (0.05° ocean grid + env + `in_usa`/
  `in_pra`), `taxon` (sp_cat, er_score, validity flags, `is_marine`, rarity), `dataset` +
  `model` (registry), `cell_metric` (per-cell scores), `zone`/`zone_cell`/`zone_metric`,
  `metric`.
- **`dist_merged/`** — the merged per-taxon model surfaces (`ms_merge|…`).
- **`registry/`** — `dataset.parquet` + `model.parquet` (also in `tables/`).
- **`dist/`** — the raw per-dataset per-species surfaces (~74G, AquaMaps-heavy) — **opt-in**
  (`RELEASE_RAW=1`); AquaMaps originals are already public.

The STAC catalog is emitted separately (needs the v8 `stac.R` pass).

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, quiet = T)
source(here("libs/paths.R"))
dir_atlas <- glue("{dir_big_v}/marine-atlas")
tbl_dir   <- glue("{dir_atlas}/tables"); dir_create(tbl_dir)
manifest  <- here("data/manifests/release_marine-atlas.json"); dir_create(path_dir(manifest))
s3_ver    <- glue("{s3_atlas}/{ver}")           # s3://oceanmetrics.io-public/marine-atlas/v8
do_s3     <- !nzchar(Sys.getenv("RELEASE_NO_S3"))
do_raw    <- nzchar(Sys.getenv("RELEASE_RAW"))
stopifnot("run build_registry first" = {
  con0 <- dbConnect(duckdb(sdm_db, read_only = TRUE)); on.exit(dbDisconnect(con0, shutdown = TRUE))
  all(c("dataset", "model") %in% dbListTables(con0)) })
```

## Stage derived tables → Parquet

```{r}
#| label: stage

con <- dbConnect(duckdb(sdm_db, read_only = TRUE))
# small/derived scoring tables (model_cell is released as dist_merged, not re-exported)
rel_tables <- c("cell", "taxon", "dataset", "model", "metric",
                "cell_metric", "zone", "zone_cell", "zone_metric")
for (t in rel_tables) {
  out <- glue("{tbl_dir}/{t}.parquet")
  dbExecute(con, glue("COPY (SELECT * FROM {t}) TO '{out}' (FORMAT parquet, COMPRESSION zstd)"))
}
sizes <- tibble(table = rel_tables,
                mb = round(file_size(glue("{tbl_dir}/{rel_tables}.parquet")) / 1e6, 1))
dbDisconnect(con, shutdown = TRUE)
knitr::kable(sizes)
```

## Sync to S3

```{r}
#| label: s3

# each release component -> its prefix under s3://…/marine-atlas/{ver}/
sync <- function(local, prefix) {
  if (!dir_exists(local)) { log_warn("skip (missing): {local}"); return(invisible()) }
  args <- c("s3", "sync", local, glue("{s3_ver}/{prefix}"), "--only-show-errors", "--no-progress")
  log_info("aws s3 sync {path_file(local)} -> {s3_ver}/{prefix}")
  st <- system2("aws", args, stdout = TRUE, stderr = TRUE)
  if (!is.null(attr(st, "status")) && attr(st, "status") != 0) stop(paste(st, collapse = "\n"))
}
if (do_s3) {
  sync(tbl_dir,                     "tables")
  sync(glue("{dir_atlas}/dist_merged"), "dist_merged")
  sync(glue("{dir_atlas}/registry"),    "registry")
  if (do_raw) sync(glue("{dir_atlas}/dist"), "dist")
  # top-level listing
  ls_s3 <- system2("aws", c("s3", "ls", glue("{s3_ver}/"), "--recursive", "--summarize"),
                   stdout = TRUE, stderr = TRUE)
  log_info("released to {s3_ver}/ ({tail(ls_s3, 2)[1]})")
} else log_info("RELEASE_NO_S3 set — staged locally only ({tbl_dir})")
```

## Manifest

```{r}
#| label: manifest

write_json(
  list(target = "release_marine-atlas", ver = ver, s3 = s3_ver,
       tables = rel_tables, table_mb = sizes$mb, raw_included = do_raw,
       pushed_s3 = do_s3, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
manifest
```
