---
title: "Rendering Guide: Derived Products v3"
subtitle: "Step-by-step instructions for regenerating all versioned outputs"
author: "Marine Sensitivity Toolkit Team"
date: "2026-02-10"
editor_options:
  chunk_output_type: console
---

## Overview

This guide documents how to regenerate all versioned derived products (`_v3` suffix) from
`calc_scores.qmd` and `msens-summary_programareas.qmd`. All output files are cataloged in
`workflows/data/derived_products.csv`.

**Prerequisites:**

- Updated `sdm_v3.duckdb` (with Rice's whale ca_nmfs model and NMFS/FWS listings)
- PostGIS database accessible via `libs/db.R`
- Mapbox token configured in private directory

## Step 1: Archive Old Downloads

Move existing unversioned files from `derived/downloads/` to an archive directory
so they are preserved but won't conflict with new versioned outputs.

```{r}
#| label: archive_old
#| eval: false
librarian::shelf(fs, glue, quiet = T)

is_server <- Sys.info()[["sysname"]] == "Linux"
dir_data <- ifelse(
  is_server,
  "/share/data",
  "~/My Drive/projects/msens/data"
)
dir_dl <- glue("{dir_data}/derived/downloads")
dir_archive <- glue("{dir_dl}/archive_2026-02-10")
dir_create(dir_archive)

# files to archive (old unversioned products)
old_files <- c(
  "ply_planareas_2025.gpkg",
  "ply_ecoregions_2025.gpkg",
  "table_planarea_scores.docx",
  "table_planarea_scores.xlsx",
  "table_ecoregion_planarea_scores.docx",
  "tbl_er_pa_scores.csv",
  "map_pa_l48.png",
  "map_pa_ak.png",
  "map_cell_score_l48.png",
  "map_cell_score_ak.png",
  "map_cell_npp_l48.png",
  "map_cell_npp_ak.png",
  "r_metrics_akl48.tif",
  "zone_taxon.csv",
  "zone_taxon_akl48.csv",
  "subregion_planareas.csv"
)

for (f in old_files) {
  src <- glue("{dir_dl}/{f}")
  if (file_exists(src)) {
    file_move(src, glue("{dir_archive}/{f}"))
  }
}

# archive old figures
dir_dl_figs <- glue("{dir_dl}/figures")
dir_archive_figs <- glue("{dir_archive}/figures")
if (dir_exists(dir_dl_figs)) {
  dir_create(dir_archive_figs)
  old_figs <- dir_ls(dir_dl_figs, regexp = "planarea|primprod|flower")
  for (f in old_figs) {
    file_move(f, glue("{dir_archive_figs}/{path_file(f)}"))
  }
}

message(glue("Archived old files to {dir_archive}"))
```

## Step 2: Run `calc_scores.qmd` Chunks

Run these chunks **interactively** in order (do not render the full document):

| Order | Chunk Label | Purpose |
|-------|-------------|---------|
| 1 | `setup` (line ~53) | load packages, set version vars, db connections |
| 2 | `zone_taxon` (line ~2600) | rebuild zone × taxon matrix |
| 3 | `setup_update_pgtable_with_clear` (line ~2730) | define PostGIS update helper |
| 4 | `pg_update_planarea_ecoregion_metrics_2025` (line ~2785) | update PA & ER PostGIS tables |
| 5 | `pg_update_programareas_metrics_2026` (line ~2870) | update PRA PostGIS table |
| 6 | `pg_rm_old_metrics` (line ~2918) | remove obsolete metric columns |
| 7 | `add_index_for_vector_tiles` (line ~2985) | add spatial indexes |
| 8 | `re-add_con_sdm.cell` (line ~3010) | refresh DuckDB cell table |
| 9 | `update_cached_downloads` (line ~3029) | generate gpkg, csv, tif files |
| 10 | `r_metrics_akl48_primprod_log` (line ~3208) | add primprod_log layer to AKL48 raster |
| 11 | `pa_metrics_geo` (line ~3223) | generate GeoJSON for planning areas |
| 12 | `export_parquet` (line ~3377) | export DuckDB to parquet (optional) |

**Key outputs from this step:**

- `ply_planareas_2025_v3.gpkg`
- `ply_programareas_2026_v3.gpkg`
- `ply_ecoregions_2025_v3.gpkg`
- `layers_2026_v3.csv`
- `r_metrics_2026_v3.tif`
- `r_metrics_akl48_v3.tif`
- `zone_taxon_2026_v3.csv`

## Step 3: Copy calc_scores Outputs to Downloads

```{r}
#| label: copy_calc_scores
#| eval: false
librarian::shelf(fs, glue, quiet = T)

is_server <- Sys.info()[["sysname"]] == "Linux"
dir_data <- ifelse(
  is_server,
  "/share/data",
  "~/My Drive/projects/msens/data"
)
dir_derived <- glue("{dir_data}/derived")
dir_dl <- glue("{dir_derived}/downloads")
dir_create(dir_dl)

v_sfx <- "_v3"

# copy versioned gpkg files
gpkg_files <- c(
  glue("ply_planareas_2025{v_sfx}.gpkg"),
  glue("ply_programareas_2026{v_sfx}.gpkg"),
  glue("ply_ecoregions_2025{v_sfx}.gpkg")
)

for (f in gpkg_files) {
  src <- glue("{dir_derived}/{f}")
  if (file_exists(src)) {
    file_copy(src, glue("{dir_dl}/{f}"), overwrite = T)
  }
}

# copy other versioned files
other_files <- c(
  glue("zone_taxon_2026{v_sfx}.csv"),
  glue("layers_2026{v_sfx}.csv"),
  glue("r_metrics_akl48{v_sfx}.tif"),
  "subregion_programareas.csv"
)

for (f in other_files) {
  src <- glue("{dir_derived}/{f}")
  if (file_exists(src)) {
    file_copy(src, glue("{dir_dl}/{f}"), overwrite = T)
  }
}

message("Copied calc_scores outputs to downloads/")
```

## Step 4: Render `msens-summary_programareas.qmd`

Run all chunks interactively (or render the full document):

```{r}
#| label: render_pra_summary
#| eval: false
quarto::quarto_render(here::here("workflows/msens-summary_programareas.qmd"))
```

This generates all Program Area products:

- Ecoregion × Program Area intersection polygons (`.gpkg`, `.csv`)
- Score tables (`.csv`, `.docx`, `.xlsx`)
- Maps (`.png`): PRA boundaries, cell scores, cell NPP
- Flower plots: ecoregion and program area (5 themes × 2 formats)
- Primary productivity bar chart (`.png`, `.pdf`, `.csv`)
- Copies everything to `downloads/` and `downloads/figures/`

## Step 5: Verify Outputs

```{r}
#| label: verify
#| eval: false
librarian::shelf(fs, glue, readr, terra, quiet = T)

is_server <- Sys.info()[["sysname"]] == "Linux"
dir_data <- ifelse(
  is_server,
  "/share/data",
  "~/My Drive/projects/msens/data"
)
dir_derived <- glue("{dir_data}/derived")
v_sfx <- "_v3"

# check dates — all v3 files should have today's date
v3_files <- dir_ls(dir_derived, regexp = "_v3\\.", recurse = F)
message(glue("Found {length(v3_files)} versioned files in derived/"))
for (f in v3_files) {
  mod <- file_info(f)$modification_time
  message(glue("  {path_file(f)}: {mod}"))
}

# check Rice's whale in zone_taxon
zt <- read_csv(
  glue("{dir_derived}/zone_taxon_2026{v_sfx}.csv"),
  show_col_types = F
)
rice <- zt |> dplyr::filter(grepl("Rice", sp_scientific, ignore.case = T))
message(glue("Rice's whale rows in zone_taxon: {nrow(rice)}"))

# check raster layers include programarea_key
r <- rast(glue("{dir_derived}/r_metrics_2026{v_sfx}.tif"))
message(glue("Raster layers: {paste(names(r), collapse = ', ')}"))
stopifnot("programarea_key" %in% names(r))

# check PRA flower plot has correct facets
message("Verify PRA flower plots manually — should show ~20 facets (not 27)")
```

## Appendix: Full Product Catalog

See `workflows/data/derived_products.csv` for a complete listing of every output file,
its source QMD, chunk label, and dependencies.
