---
title: "Rendering Guide: Derived Products v3"
subtitle: "Step-by-step instructions for regenerating all versioned outputs"
author: "Marine Sensitivity Toolkit Team"
date: "2026-02-10"
editor_options:
  chunk_output_type: console
---

## Overview

This guide documents how to regenerate all versioned derived products (`_v3` suffix) from
`calc_scores.qmd` and `msens-summary_programareas.qmd`. All output files are cataloged in
`workflows/data/derived_products.csv`.

**Prerequisites:**

- Updated `sdm.duckdb` in big-files `v3/` (with Rice's whale ca_nmfs model and NMFS/FWS listings)
- PostGIS database accessible via `libs/db.R`
- Mapbox token configured in private directory

## Step 1: Run `calc_scores.qmd` Chunks

Run these chunks **interactively** in order (do not render the full document):

| Order | Chunk Label | Purpose |
|-------|-------------|---------|
| 1 | `setup` (line ~53) | load packages, set version vars, db connections |
| 2 | `zone_taxon` (line ~2600) | rebuild zone x taxon matrix |
| 3 | `setup_update_pgtable_with_clear` (line ~2730) | define PostGIS update helper |
| 4 | `pg_update_planarea_ecoregion_metrics_2025` (line ~2785) | update PA & ER PostGIS tables |
| 5 | `pg_update_programareas_metrics_2026` (line ~2870) | update PRA PostGIS table |
| 6 | `pg_rm_old_metrics` (line ~2918) | remove obsolete metric columns |
| 7 | `add_index_for_vector_tiles` (line ~2985) | add spatial indexes |
| 8 | `re-add_con_sdm.cell` (line ~3010) | refresh DuckDB cell table |
| 9 | `update_cached_downloads` (line ~3029) | generate gpkg, csv, tif files |
| 10 | `r_metrics_akl48_primprod_log` (line ~3208) | add primprod_log layer to AKL48 raster |
| 11 | `pa_metrics_geo` (line ~3223) | generate GeoJSON for planning areas |
| 12 | `export_parquet` (line ~3377) | export DuckDB to parquet (optional) |

**Key outputs** (all written to `derived/v3/`):

- `ply_planareas_2025_v3.gpkg`
- `ply_programareas_2026_v3.gpkg`
- `ply_ecoregions_2025_v3.gpkg`
- `layers_v3.csv`
- `r_metrics_v3.tif`
- `r_metrics_programareas_v3.tif`
- `zone_taxon_v3.csv`

## Step 2: Render `msens-summary_programareas.qmd`

Run all chunks interactively (or render the full document):

```{r}
#| label: render_pra_summary
#| eval: false
quarto::quarto_render(here::here("workflows/msens-summary_programareas.qmd"))
```

This generates all Program Area products directly into `derived/v3/`:

- Ecoregion x Program Area intersection polygons (`.gpkg`, `.csv`)
- Score tables (`.csv`, `.docx`, `.xlsx`)
- Maps (`.png`): PRA boundaries, cell scores, cell NPP
- Flower plots: ecoregion and program area (5 themes x 2 formats)
- Primary productivity bar chart (`.png`, `.pdf`, `.csv`)

## Step 3: Verify Outputs

```{r}
#| label: verify
#| eval: false
librarian::shelf(fs, glue, readr, terra, quiet = T)

source(here::here("libs/paths.R"))

# check dates — all v3 files should have today's date
v3_files <- dir_ls(dir_v, regexp = "_v3\\.", recurse = F)
message(glue("Found {length(v3_files)} versioned files in derived/{v_dir}/"))
for (f in v3_files) {
  mod <- file_info(f)$modification_time
  message(glue("  {path_file(f)}: {mod}"))
}

# check Rice's whale in zone_taxon
zt <- read_csv(
  glue("{dir_v}/zone_taxon{v_sfx}.csv"),
  show_col_types = F
)
rice <- zt |> dplyr::filter(grepl("Rice", sp_scientific, ignore.case = T))
message(glue("Rice's whale rows in zone_taxon: {nrow(rice)}"))

# check raster layers include programarea_key
r <- rast(glue("{dir_v}/r_metrics{v_sfx}.tif"))
message(glue("Raster layers: {paste(names(r), collapse = ', ')}"))
stopifnot("programarea_key" %in% names(r))

# check PRA flower plot has correct facets
message("Verify PRA flower plots manually — should show ~20 facets (not 27)")
```

## Appendix: Full Product Catalog

See `workflows/data/derived_products.csv` for a complete listing of every output file,
its source QMD, chunk label, and dependencies.
