---
title: "Score zone metrics — program-area composite + v7 equivalence gate"
msens:
  target_name: score_zone_metrics
  workflow_type: score
  dependency: [score_cell_metrics]
  output: data/manifests/score_zone_metrics.json
editor_options:
  chunk_output_type: console
---

Aggregate the composite components to zones and compute the **program-area score**
(v7's distinct path, not a roll-up of the cell composite):

1. component → zone = `SUM(cm.val × pct_covered) / SUM(pct_covered)`
2. PA pct_area down-weight: `val × (data_coverage / total_coverage)` (backs up `_prepctareaweighting`)
3. PA score = `SUM(val)/COUNT(val)` (plain mean) of the pct_area-weighted components

Then the **`pra_score_delta` gate**: compare the v8 PA composite to v7's per program area.

## Setup

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, fs, glue, here, jsonlite, logger, sf, quiet = T)
source(here("libs/paths.R")); source(here("libs/vars.R"))
manifest <- here("data/manifests/score_zone_metrics.json"); dir_create(path_dir(manifest))
pa_gpkg  <- glue("{dir_data}/derived/{ver_prev}/ply_programareas_2026_{ver_prev}.gpkg")
stopifnot(file_exists(sdm_db), file_exists(pa_gpkg))

con <- dbConnect(duckdb(sdm_db))
mseq_of <- function(k) dbGetQuery(con, glue("SELECT metric_seq FROM metric WHERE metric_key='{k}'"))$metric_seq[1]
mseq_new <- function(k, desc="") { ex<-dbGetQuery(con,glue("SELECT metric_seq FROM metric WHERE metric_key='{k}'"))
  s <- if(nrow(ex)) ex$metric_seq[1] else { s0<-dbGetQuery(con,"SELECT coalesce(max(metric_seq),0)+1 s FROM metric")$s
    dbExecute(con,glue("INSERT INTO metric VALUES ({s0},'{k}','{desc}')")); s0 }
  dbExecute(con, glue("DELETE FROM zone_metric WHERE metric_seq={s} AND zone_seq IN (SELECT zone_seq FROM zone WHERE fld='programarea_key')")); s }

sp_cats   <- c("bird","coral","fish","invertebrate","mammal","primary_producer","turtle")
comp_keys <- c(glue("extrisk_{sp_cats}_ecoregion_rescaled"), "primprod_ecoregion_rescaled")
comp_seqs <- sapply(comp_keys, mseq_of)
```

## Component → program-area (coverage-weighted mean) + pct_area down-weight

```{r}
#| label: zone_agg

for (k in comp_keys) {
  bs <- mseq_of(k)
  bk <- mseq_new(glue("{k}_prepctareaweighting"), glue("pre-pct_area backup of {k}"))
  # 2a coverage-weighted mean over PA cells, into zone_metric (this metric_seq, PA zones)
  dbExecute(con, glue("DELETE FROM zone_metric WHERE metric_seq={bs} AND zone_seq IN (SELECT zone_seq FROM zone WHERE fld='programarea_key')"))
  dbExecute(con, glue("INSERT INTO zone_metric (zone_seq, metric_seq, val)
    SELECT zc.zone_seq, {bs}, SUM(cm.val*zc.pct_covered)/SUM(zc.pct_covered)
    FROM zone_cell zc JOIN zone z USING(zone_seq) JOIN cell_metric cm ON zc.cell_id=cm.cell_id
    WHERE z.fld='programarea_key' AND cm.metric_seq={bs} AND cm.val IS NOT NULL
    GROUP BY zc.zone_seq"))
  # backup pre-weight val
  dbExecute(con, glue("INSERT INTO zone_metric (zone_seq, metric_seq, val)
    SELECT zone_seq, {bk}, val FROM zone_metric WHERE metric_seq={bs}
      AND zone_seq IN (SELECT zone_seq FROM zone WHERE fld='programarea_key')"))
  # 2b pct_area = (coverage of cells with data for this metric) / (coverage of all cells in zone); val *= pct_area
  dbExecute(con, glue("UPDATE zone_metric zm SET val = val * (
      SELECT SUM(zc.pct_covered) FILTER (WHERE cm.val IS NOT NULL) * 1.0 / SUM(zc.pct_covered)
      FROM zone_cell zc LEFT JOIN cell_metric cm ON zc.cell_id=cm.cell_id AND cm.metric_seq={bs}
      WHERE zc.zone_seq = zm.zone_seq)
    WHERE zm.metric_seq={bs} AND zm.zone_seq IN (SELECT zone_seq FROM zone WHERE fld='programarea_key')"))
}
log_info("zone components + pct_area done for {length(comp_keys)} metrics")
```

## Program-area composite score

```{r}
#| label: pa_score

sc <- mseq_new("score_extriskspcat_primprod_ecoregionrescaled_equalweights", "Equal-weight composite (PA)")
seqs_sql <- paste(comp_seqs, collapse=",")
dbExecute(con, glue("INSERT INTO zone_metric (zone_seq, metric_seq, val)
  SELECT zone_seq, {sc}, SUM(val)/COUNT(val)
  FROM zone_metric
  WHERE metric_seq IN ({seqs_sql}) AND val IS NOT NULL
    AND zone_seq IN (SELECT zone_seq FROM zone WHERE fld='programarea_key')
  GROUP BY zone_seq"))
v8_pa <- dbGetQuery(con, glue("SELECT z.val AS programarea_key, round(zm.val,2) AS score_v8
  FROM zone z JOIN zone_metric zm USING(zone_seq) WHERE zm.metric_seq={sc} ORDER BY 1"))
dbDisconnect(con, shutdown=TRUE)
```

## v7 equivalence gate (pra_score_delta)

```{r}
#| label: validate

v7_pa <- st_read(pa_gpkg, quiet=TRUE) |> st_drop_geometry() |>
  transmute(programarea_key, score_v7 = round(score_extriskspcat_primprod_ecoregionrescaled_equalweights, 2))
cmp <- v8_pa |> inner_join(v7_pa, by="programarea_key") |>
  mutate(delta = round(score_v8 - score_v7, 2)) |> arrange(desc(abs(delta)))
knitr::kable(cmp)
gate <- with(cmp, list(n=nrow(cmp), mean_abs=round(mean(abs(delta)),2), max_abs=round(max(abs(delta)),2),
  rmse=round(sqrt(mean(delta^2)),2), cor=round(cor(score_v8, score_v7),3)))
str(gate)
write_json(c(list(target="score_zone_metrics", ver=ver), gate, list(built=as.character(Sys.time()))),
  manifest, auto_unbox=TRUE, pretty=TRUE)
```
