---
title: "Score setup — zones (program areas + ecoregions) + taxon prep"
msens:
  target_name: score_zones
  workflow_type: score
  dependency: [merge_taxon, build_cell_grid]
  output: data/manifests/score_zones.json
editor_options:
  chunk_output_type: console
---

Foundation for `calc_scores` (v8): assemble the scoring DuckDB from the merged
`model_cell` + `taxon` + `cell`, prep the taxon for scoring (`sp_cat`,
`is_er_spatial`), and build `zone` + `zone_cell` — cells ↔ program areas / ecoregions
with `pct_covered` (0–100) via exactextractr coverage of each zone polygon over the
cell-id grid. Scoring formulas need `zone_cell.pct_covered` at every step.

## Setup — assemble the scoring DB

```{r}
#| label: setup

librarian::shelf(DBI, dplyr, duckdb, exactextractr, fs, glue, here, jsonlite, logger,
                 sf, terra, tibble, MarineSensitivity/msens, quiet = T)
source(here("libs/paths.R"))
source(here("libs/vars.R"))
sf_use_s2(FALSE)

dir_atlas <- glue("{dir_big_v}/marine-atlas")
merged    <- glue("{dir_atlas}/dist_merged/dataset=ms_merge")
merge_db  <- glue("{dir_atlas}/merge.duckdb")
pa_gpkg   <- glue("{dir_data}/derived/{ver_prev}/ply_programareas_2026_{ver_prev}.gpkg")
er_gpkg   <- glue("{dir_data}/derived/{ver_prev}/ply_ecoregions_2025.gpkg")
manifest  <- here("data/manifests/score_zones.json")
dir_create(path_dir(manifest))
stopifnot(all(file_exists(c(sdm_db, merge_db, cellid_tif, pa_gpkg, er_gpkg))),
          dir_exists(merged))

con <- dbConnect(duckdb(sdm_db))                    # sdm_db already has `cell` (build_cell_grid)
dbExecute(con, glue("ATTACH '{merge_db}' AS m (READ_ONLY)"))

# model_cell = merged US model (mdl_key, cell_id, val); materialize for scoring joins
dbExecute(con, glue("CREATE OR REPLACE TABLE model_cell AS
  SELECT mdl_key, cell_id, val FROM read_parquet('{merged}/**/*.parquet')"))
dbExecute(con, "CREATE INDEX IF NOT EXISTS mc_cell ON model_cell(cell_id)")
```

## Taxon prep — sp_cat + is_er_spatial

```{r}
#| label: taxon

# reuse v7 sp_cat (bird/coral/fish/invertebrate/mammal/other/turtle) by worms_id/sisid.
# v7 taxon_id is DOUBLE, v8 is VARCHAR -> cast via BIGINT in SQL (R as.character gives
# scientific notation for large ids and won't match). New v8 taxa default to 'other'.
# is_er_spatial = turtles (SWOT+DPS ER baked into the merged cell val -> er_score=100
# at scoring so it isn't double-counted).
dbExecute(con, glue("ATTACH '{dir_big}/{ver_prev}/sdm.duckdb' AS v7 (READ_ONLY)"))
dbExecute(con, "CREATE OR REPLACE TABLE v7_cat AS
  SELECT taxon_authority, CAST(CAST(taxon_id AS BIGINT) AS VARCHAR) AS taxon_id,
         any_value(sp_cat) AS sp_cat, bool_or(is_ok) AS is_ok
  FROM v7.taxon WHERE taxon_id IS NOT NULL GROUP BY 1, 2")
dbExecute(con, "CREATE OR REPLACE TABLE taxon AS
  SELECT t.*,
    coalesce(c.sp_cat, 'other')                              AS sp_cat,
    coalesce(c.is_ok, FALSE)                                 AS in_v7,   -- in v7's scored set (same-species compare)
    (t.ms_merge_key IN (
       SELECT DISTINCT ms_merge_key FROM m.taxon_model WHERE mdl_key LIKE 'rng_turtle_swot_dps|%'
     ))                                                      AS is_er_spatial
  FROM m.taxon t
  LEFT JOIN v7_cat c ON t.taxon_authority = c.taxon_authority AND t.taxon_id = c.taxon_id")
dbExecute(con, "DETACH v7")
knitr::kable(dbGetQuery(con, "SELECT sp_cat, count(*) n FROM taxon WHERE is_valid_usa GROUP BY 1 ORDER BY 2 DESC"))
```

## Zones + zone_cell (program areas + ecoregions)

```{r}
#| label: zones

zone_specs <- list(
  list(tbl = glue("ply_programareas_2026_{ver}"), fld = "programarea_key", gpkg = pa_gpkg),
  list(tbl = "ply_ecoregions_2025",               fld = "ecoregion_key",   gpkg = er_gpkg))

r_cell <- rast(cellid_tif)
zone_rows <- list(); zc_rows <- list(); zseq <- 0L
for (zs in zone_specs) {
  ply <- st_read(zs$gpkg, quiet = TRUE) |> st_make_valid() |>
    st_transform(4326) |> select(all_of(zs$fld))
  for (i in seq_len(nrow(ply))) {
    zseq <- zseq + 1L
    zval <- ply[[zs$fld]][i]
    zone_rows[[length(zone_rows) + 1]] <- tibble(
      zone_seq = zseq, tbl = zs$tbl, fld = zs$fld, val = as.character(zval))
    # coverage fraction of this zone polygon over each cell -> pct_covered (0-100)
    df <- exact_extract(r_cell, ply[i, ], progress = FALSE)[[1]]   # exactextractr's own `value` col
    df <- df[!is.na(df$value) & df$coverage_fraction > 0, , drop = FALSE]
    if (nrow(df) > 0) zc_rows[[length(zc_rows) + 1]] <- tibble(
      zone_seq = zseq, cell_id = as.integer(df$value),
      pct_covered = as.integer(round(df$coverage_fraction * 100)))
  }
  log_info("{zs$fld}: {nrow(ply)} zones")
}
dbWriteTable(con, "zone",      bind_rows(zone_rows), overwrite = TRUE)
dbWriteTable(con, "zone_cell", bind_rows(zc_rows) |> filter(pct_covered > 0), overwrite = TRUE)

smry <- dbGetQuery(con, "SELECT z.fld, count(DISTINCT z.zone_seq) zones, count(*) zone_cells
  FROM zone z JOIN zone_cell zc USING(zone_seq) GROUP BY 1")
knitr::kable(smry)
dbDisconnect(con, shutdown = TRUE)
```

## Manifest

```{r}
#| label: manifest

write_json(list(target = "score_zones", ver = ver, built = as.character(Sys.time())),
  manifest, auto_unbox = TRUE, pretty = TRUE)
```
