---
title: "Study Area"
format:
  html:
    embed-resources: true
---

## Proposed Planning Areas

The [Proposed Planning Areas](https://marinesensitivity.org/workflows/ingest_productivity.html)
span the US Exclusive Economic Zone (EEZ), originally created in
[`ingest_productivity.qmd`](https://marinesensitivity.org/workflows/ingest_productivity.html).

The "to shore" extension fills the gap between federal planning areas and the
coastline, which varies by jurisdiction:

- **Most US coasts**: state waters extend **3 nautical miles** from shore
- **Texas, Gulf coast of Florida, Puerto Rico**: state waters extend
  **9 nautical miles** from shore

Nearshore territory is partitioned among Planning Areas using Voronoi
tessellation (`st_voronoi()`) seeded from densely sampled PA boundary points,
so nearshore boundaries follow the natural extension of PA boundaries to shore.

```{r}
#| label: setup
#| warning: false
#| message: false
librarian::shelf(
  dplyr,
  mapgl,
  mregions2,
  rnaturalearth,
  rnaturalearthhires,
  sf,
  quiet = T
)

dir_data <- "~/My Drive/projects/msens/data/derived/v1"
pa_gpkg <- file.path(dir_data, "ply_planareas_2025.gpkg")
pa_toshore_gpkg <- file.path(dir_data, "ply_planareas_2025_toshore.gpkg")
```

```{r}
#| label: read-planning-areas
#| warning: false
#| message: false
pa <- read_sf(pa_gpkg)
```

```{r}
#| label: extend-to-shore
#| warning: false
#| message: false
if (!file.exists(pa_toshore_gpkg)) {
  sf_use_s2(FALSE)

  # high-res land: Natural Earth countries + minor islands (≤2 km²)
  land_countries <- ne_countries(
    scale       = 10,
    sovereignty = "United States of America",
    returnclass = "sf") |>
    st_geometry()
  land_minor <- ne_download(
    scale    = 10,
    type     = "minor_islands",
    category = "physical",
    returnclass = "sf") |>
    st_geometry()
  land <- c(land_countries, land_minor) |>
    st_union() |>
    st_make_valid()

  # US EEZ from mregions2
  eez <- mrp_get("eez", cql_filter = "sovereign1 = 'United States'") |>
    st_geometry() |>
    st_union() |>
    st_make_valid()

  # zero-width buffer to clean up topology
  land     <- st_buffer(land, 0) |> st_make_valid()
  eez      <- st_buffer(eez, 0)  |> st_make_valid()
  land_buf <- st_buffer(land, 0.3) |> st_make_valid()

  # nearshore zone: buffer land ~12nm (0.3°), clip to US EEZ, subtract land
  nearshore <- st_intersection(land_buf, eez) |>
    st_make_valid() |>
    st_difference(land) |>
    st_make_valid()

  # supplement nearshore with NPS + NOAA marine boundaries for islands
  # too small for Natural Earth (Dry Tortugas, Farallons, Florida Keys)
  nps_url <- paste0(
    "https://services1.arcgis.com/fBc8EJBxQRMcHlei/arcgis/rest/services/",
    "NPS_Land_Resources_Division_Boundary_and_Tract_Data_Service/",
    "FeatureServer/2/query",
    "?where=UNIT_CODE+IN+(%27DRTO%27,%27CHIS%27,%27GUIS%27)",
    "&outFields=UNIT_NAME&f=geojson")
  noaa_url <- paste0(
    "https://raw.githubusercontent.com/noaa-onms/onmsR/",
    "master/data-raw/sanctuaries.geojson")

  nps_marine <- tryCatch(
    read_sf(nps_url) |> st_geometry() |> st_union() |> st_make_valid(),
    error = function(e) st_sfc(crs = 4326))
  noaa_marine <- tryCatch({
    sanc <- read_sf(noaa_url)
    # only Greater Farallones + Florida Keys (the relevant sanctuaries)
    # only Greater Farallones (covers Farallon Islands)
    sanc <- sanc[sanc$sanctuary == "Greater Farallones", ]
    sanc |> st_geometry() |> st_union() |> st_make_valid()
  }, error = function(e) st_sfc(crs = 4326))

  # add marine park/sanctuary areas to nearshore (minus land);
  # these are already within US EEZ so no need to clip to eez
  marine_supp <- tryCatch(
    c(nps_marine, noaa_marine) |>
      st_union() |>
      st_make_valid() |>
      st_difference(land) |>
      st_make_valid(),
    error = function(e) st_sfc(crs = 4326))
  nearshore <- st_union(nearshore, marine_supp) |> st_make_valid()

  # voronoi from subsampled PA boundary vertices so that nearshore
  # boundaries naturally extend the existing PA boundary lines to shore;
  # subsample to ~200 vertices per PA for tractable Voronoi computation
  pts_list <- lapply(seq_len(nrow(pa)), function(i) {
    pts_i <- tryCatch(
      st_cast(st_geometry(pa)[i], "MULTIPOINT") |>
        st_cast("POINT"),
      error = function(e) st_sfc(crs = st_crs(pa)))
    n <- length(pts_i)
    if (n < 2) return(NULL)
    step <- max(1, floor(n / 200))
    idx  <- seq(1, n, by = step)
    result <- st_sf(
      planarea_key = rep(pa$planarea_key[i], length(idx)),
      geometry     = pts_i[idx])

    # wrap dateline-spanning PA points into [-180, 180] for Voronoi
    bb <- st_bbox(pa[i, ])
    if (bb["xmin"] < -170 && bb["xmax"] > 170) {
      coords <- st_coordinates(result)
      coords[, "X"] <- ifelse(
        coords[, "X"] > 180, coords[, "X"] - 360, coords[, "X"])
      result <- st_sf(
        planarea_key = result$planarea_key,
        geometry     = st_sfc(
          lapply(seq_len(nrow(coords)), function(j) st_point(coords[j, ])),
          crs = st_crs(pa)))
    }
    result
  })
  pa_pts_all <- do.call(rbind, Filter(Negate(is.null), pts_list))
  message("  voronoi seed points: ", nrow(pa_pts_all))

  # build voronoi tessellation (sf#824: must union points first)
  envelope  <- st_as_sfc(st_bbox(nearshore))
  pts_union <- st_union(pa_pts_all)
  vor_cells <- st_voronoi(pts_union, envelope = envelope) |>
    st_collection_extract()

  # assign each voronoi cell to its seed point's PA (first match for
  # shared boundary points)
  cell_matches <- st_intersects(vor_cells, pa_pts_all)
  cell_idx     <- sapply(cell_matches, `[`, 1)
  keep         <- !is.na(cell_idx)
  vor_sf <- st_sf(
    planarea_key = pa_pts_all$planarea_key[cell_idx[keep]],
    geometry     = vor_cells[keep])

  # dissolve voronoi cells by PA → one territory polygon per PA
  vor_dissolved <- vor_sf |>
    group_by(planarea_key) |>
    summarize(.groups = "drop") |>
    st_make_valid()

  # nearshore gap: only the nearshore NOT already covered by any PA
  # (prevents voronoi slivers from overwriting original PA territory)
  pa_union      <- st_union(pa) |> st_make_valid()
  nearshore_gap <- st_difference(nearshore, pa_union) |> st_make_valid()

  # clip voronoi territories to the nearshore gap only
  nearshore_by_pa <- st_intersection(vor_dissolved, nearshore_gap) |>
    st_make_valid()

  # union each PA with its nearshore gap territory
  pa_toshore <- pa
  for (i in seq_len(nrow(pa))) {
    key_i   <- pa$planarea_key[i]
    coast_i <- nearshore_by_pa |> filter(planarea_key == key_i)
    if (nrow(coast_i) == 0) next

    bb          <- st_bbox(pa[i, ])
    is_dateline <- bb["xmin"] < -170 && bb["xmax"] > 170

    if (is_dateline) {
      # shift nearshore territory to 0-360 space to match PA geometry
      coast_geom <- tryCatch(
        st_shift_longitude(st_geometry(coast_i)) |> st_make_valid(),
        error = function(e) NULL)
    } else {
      coast_geom <- st_geometry(coast_i)
    }

    if (!is.null(coast_geom)) {
      new_geom <- tryCatch(
        st_union(c(st_geometry(pa)[i], coast_geom)) |> st_make_valid(),
        error = function(e) NULL)
      if (!is.null(new_geom) && length(new_geom) > 0 &&
          inherits(new_geom[[1]], "MULTIPOLYGON")) {
        st_geometry(pa_toshore)[i] <- new_geom
      }
    }
  }

  # subtract land from final output so PAs are ocean-only (fixes PR, etc.)
  for (i in seq_len(nrow(pa_toshore))) {
    bb <- st_bbox(pa_toshore[i, ])
    if (bb["xmin"] < -170 && bb["xmax"] > 170) next  # skip dateline PAs
    new_geom <- tryCatch(
      st_difference(st_geometry(pa_toshore)[i], land) |> st_make_valid(),
      error = function(e) NULL)
    if (!is.null(new_geom) && length(new_geom) > 0 &&
        inherits(new_geom[[1]], c("POLYGON", "MULTIPOLYGON"))) {
      st_geometry(pa_toshore)[i] <- new_geom
    }
  }

  sf_use_s2(TRUE)

  write_sf(pa_toshore, pa_toshore_gpkg)
} else {
  pa_toshore <- read_sf(pa_toshore_gpkg)
}
```

```{r}
#| label: map
#| warning: false
#| message: false

# color palette and label points for planning areas
pa_keys <- sort(unique(pa_toshore$planarea_key))
pa_colors <- hcl.colors(length(pa_keys), palette = "Dynamic")

sf_use_s2(FALSE)
pa_labels <- pa |>
  select(planarea_key) |>
  st_point_on_surface()
sf_use_s2(TRUE)

maplibre(
  style = maptiler_style("bright"),
  center = c(-95, 38),
  zoom = 2
) |>
  add_fill_layer(
    id = "pa_toshore_fill",
    source = pa_toshore,
    fill_color = match_expr(
      column = "planarea_key",
      values = pa_keys,
      stops = pa_colors
    ),
    fill_opacity = 0.5,
    tooltip = "planarea_name",
    popup = "planarea_name"
  ) |>
  add_line_layer(
    id = "pa_outline",
    source = pa,
    line_color = "darkblue",
    line_width = 1
  ) |>
  add_symbol_layer(
    id = "pa_labels",
    source = pa_labels,
    text_field = get_column("planarea_key"),
    text_size = 12,
    text_color = "black",
    text_halo_color = "white",
    text_halo_width = 1,
    text_allow_overlap = TRUE
  ) |>
  add_layers_control(
    position = "top-left",
    layers = c(
      "PA to Shore" = "pa_toshore_fill",
      "PA Outlines" = "pa_outline",
      "PA Labels" = "pa_labels"
    )
  ) |>
  add_fullscreen_control() |>
  add_navigation_control() |>
  add_scale_control()
```
