---
title: "Update Scores"
subtitle: "Remap model_cell values, listing join, re-export taxon CSV"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Overview

Focused update notebook to:

1. Remap old model_cell values (70/90 for ch, 50/70/90 for rng) to the new `er_score` scale
2. Recompute merged model cells for affected taxa
3. Run the listing join to populate `extrisk_code`/`er_score` on the `taxon` table
4. Re-export `data/taxon.csv`

After running this notebook, re-run `calc_scores.qmd` interactively starting from
chunk `delete_extrisk_metrics` (line 472) to recalculate downstream scoring metrics.

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(
  DBI, dplyr, duckdb, glue, here, readr, stringr, tibble, quiet = T)
options(readr.show_col_types = F)

# set up paths
is_server <- Sys.info()[["sysname"]] == "Linux"
dir_data  <- ifelse(
  is_server,
  "/share/data",
  "~/My Drive/projects/msens/data")
sdm_db <- glue("{dir_data}/derived/sdm_2026.duckdb")

con_sdm <- dbConnect(duckdb(), sdm_db)
```

## Step 1: Remap ch/rng model_cell values to er_score scale

Remap old ad-hoc values to match the `compute_er_score()` scale: EN=100, TN=50, LC=1.

```{r}
#| label: remap_model_cell_values

# check current values before remapping
d_before <- dbGetQuery(con_sdm, "
  SELECT m.ds_key, mc.value, count(*) AS n
  FROM model_cell mc
  JOIN model m ON mc.mdl_seq = m.mdl_seq
  WHERE m.ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
  GROUP BY m.ds_key, mc.value
  ORDER BY m.ds_key, mc.value")
message("before remapping:")
print(d_before)

# remap: 90->100 (EN), 70->50 (TN), 50->1 (LC, rng_fws only)
n_updated <- dbExecute(con_sdm, "
  UPDATE model_cell SET value = CASE value
    WHEN 90 THEN 100
    WHEN 70 THEN 50
    WHEN 50 THEN 1
  END
  WHERE mdl_seq IN (
    SELECT mdl_seq FROM model
    WHERE ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
  )")
message(glue("remapped {n_updated} model_cell rows"))

# verify
d_after <- dbGetQuery(con_sdm, "
  SELECT m.ds_key, mc.value, count(*) AS n
  FROM model_cell mc
  JOIN model m ON mc.mdl_seq = m.mdl_seq
  WHERE m.ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
  GROUP BY m.ds_key, mc.value
  ORDER BY m.ds_key, mc.value")
message("after remapping:")
print(d_after)
stopifnot(all(d_after$value %in% c(1, 50, 100)))
```

## Step 2: Recompute merged model cells for affected taxa

For taxa with merged models (n_ds > 1) that include ch/rng data, recompute cell
values as `max(all source values)` since the source ch/rng values changed.

```{r}
#| label: recompute_merged_models

# count affected taxa
d_affected <- dbGetQuery(con_sdm, "
  SELECT count(DISTINCT t.taxon_id) AS n_taxa
  FROM taxon t
  WHERE t.mdl_seq IS NOT NULL
    AND t.n_ds > 1
    AND t.taxon_id IN (
      SELECT DISTINCT taxon_id FROM taxon_model
      WHERE ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
    )")
message(glue("recomputing merged models for {d_affected$n_taxa} taxa"))

n_updated <- dbExecute(con_sdm, "
  UPDATE model_cell mc SET value = src.max_val
  FROM (
    SELECT t.mdl_seq AS merged_seq, all_src.cell_id,
           MAX(all_src.value) AS max_val
    FROM taxon t
    JOIN taxon_model tm ON t.taxon_id = tm.taxon_id
    JOIN model_cell all_src ON tm.mdl_seq = all_src.mdl_seq
    WHERE t.mdl_seq IS NOT NULL
      AND t.n_ds > 1
      AND t.taxon_id IN (
        SELECT DISTINCT taxon_id FROM taxon_model
        WHERE ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
      )
    GROUP BY t.mdl_seq, all_src.cell_id
  ) AS src
  WHERE mc.mdl_seq = src.merged_seq
    AND mc.cell_id = src.cell_id")
message(glue("updated {n_updated} merged model_cell rows"))
```

## Step 3: Listing join

Populate `extrisk_code`, `er_score`, `is_mmpa`, `is_mbta`, `is_bcc` on the `taxon` table
from the `listing` table. Includes a secondary join by `botw_id` for BOTW-only birds.

```{r}
#| label: listing_join

# add columns if not exist
for (col_sql in c(
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS extrisk_code VARCHAR",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS er_score INTEGER",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS is_mmpa BOOLEAN DEFAULT FALSE",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS is_mbta BOOLEAN DEFAULT FALSE",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS is_bcc  BOOLEAN DEFAULT FALSE")) {
  dbExecute(con_sdm, col_sql)
}

# clear any existing values so we start fresh
dbExecute(con_sdm, "
  UPDATE taxon SET
    extrisk_code = NULL, er_score = NULL,
    is_mmpa = FALSE, is_mbta = FALSE, is_bcc = FALSE")

# primary: join by worms_id (covers non-birds + birds with worms mapping)
n_worms <- dbExecute(con_sdm, "
  UPDATE taxon SET
    extrisk_code = l.extrisk_code,
    er_score     = l.er_score,
    is_mmpa      = l.is_mmpa,
    is_mbta      = l.is_mbta,
    is_bcc       = l.is_bcc
  FROM listing l
  WHERE taxon.worms_id = l.worms_id")
message(glue("listing join by worms_id: {n_worms} taxa matched"))

# secondary: join by botw_id for BOTW-only birds not matched above
n_botw <- dbExecute(con_sdm, "
  UPDATE taxon SET
    extrisk_code = l.extrisk_code,
    er_score     = l.er_score,
    is_mmpa      = l.is_mmpa,
    is_mbta      = l.is_mbta,
    is_bcc       = l.is_bcc
  FROM listing l
  WHERE taxon.taxon_authority = 'botw'
    AND taxon.taxon_id = l.botw_id
    AND taxon.extrisk_code IS NULL")
message(glue("listing join by botw_id: {n_botw} taxa matched"))

# IUCN fallback for unmatched taxa
n_iucn <- dbExecute(con_sdm, "
  UPDATE taxon SET
    extrisk_code = 'IUCN:' || redlist_code,
    er_score     = CASE redlist_code
      WHEN 'CR' THEN 50
      WHEN 'EN' THEN 25
      WHEN 'VU' THEN 5
      WHEN 'NT' THEN 2
      ELSE 1 END
  WHERE extrisk_code IS NULL AND redlist_code IS NOT NULL")
message(glue("IUCN fallback: {n_iucn} taxa"))

# default for taxa with no listing and no redlist_code
n_default <- dbExecute(con_sdm, "
  UPDATE taxon SET er_score = 1
  WHERE er_score IS NULL")
message(glue("default er_score=1: {n_default} taxa"))

# verify no nulls remain
n_null <- dbGetQuery(con_sdm,
  "SELECT count(*) AS n FROM taxon WHERE er_score IS NULL")$n
stopifnot(n_null == 0)

# summary by authority
tbl(con_sdm, "taxon") |>
  mutate(
    authority = CASE_WHEN(
      str_starts(extrisk_code, "NMFS") ~ "NMFS",
      str_starts(extrisk_code, "FWS")  ~ "FWS",
      str_starts(extrisk_code, "IUCN") ~ "IUCN",
      TRUE                             ~ "none")) |>
  count(authority) |>
  collect() |>
  print()

# summary by taxon_authority
tbl(con_sdm, "taxon") |>
  filter(!is.na(extrisk_code)) |>
  count(taxon_authority) |>
  collect() |>
  print()
```

## Step 4: Re-export taxon summary CSV

```{r}
#| label: taxon_summary

taxon_csv <- here("data/taxon.csv")

d_taxon <- tbl(con_sdm, "taxon") |>
  select(
    is_ok,
    component     = sp_cat,
    common_name,
    scientific_name,
    redlist_code_max = redlist_code,
    extrisk_code,
    er_score,
    is_mmpa,
    is_mbta,
    is_bcc,
    worms_is_marine,
    worms_is_extinct,
    n_datasets    = n_ds,
    taxon_authority,
    taxon_id,
    model_id      = mdl_seq) |>
  arrange(desc(is_ok), component, common_name) |>
  collect()

write_csv(d_taxon, taxon_csv)
message(glue("wrote {nrow(d_taxon)} rows to {taxon_csv}"))
```

## Cleanup

```{r}
#| label: cleanup

dbDisconnect(con_sdm, shutdown = TRUE)
```

## Next steps

Re-run `calc_scores.qmd` interactively in Positron starting from chunk
`delete_extrisk_metrics` (line 472):

1. `delete_extrisk_metrics` -- clear old metrics
2. `calc_cell_metric_redlist` -- recompute cell-level extinction risk
3. `calc_ecoregion_minmax` through `cell_metric_score` -- rescale and aggregate
4. `zone_taxon` -- per-zone species sensitivity with updated er_score
