---
title: "Update Scores"
subtitle: "Remap model_cell values, listing join, re-export taxon CSV"
format:
  html:
    code-fold: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Overview

Focused update notebook to:

1. Remap old model_cell values (70/90 for ch, 50/70/90 for rng) to the new `er_score` scale
2. Recompute merged model cells for affected taxa
3. Run the listing join to populate `extrisk_code`/`er_score` on the `taxon` table
4. Re-export `data/taxon.csv`

After running this notebook, re-run `calc_scores.qmd` interactively starting from
chunk `delete_extrisk_metrics` (line 472) to recalculate downstream scoring metrics.

```{r}
#| label: setup
#| message: false
#| warning: false

librarian::shelf(
  DBI,
  dplyr,
  duckdb,
  glue,
  here,
  readr,
  stringr,
  tibble,
  tidyr,
  quiet = T
)
options(readr.show_col_types = F)

# set up paths
is_server <- Sys.info()[["sysname"]] == "Linux"
dir_data <- ifelse(
  is_server,
  "/share/data",
  "~/My Drive/projects/msens/data"
)
v_int <- 3
v_sfx <- paste0("_v", v_int)
sdm_db <- glue("{dir_data}/derived/sdm{v_sfx}.duckdb")

con_sdm <- dbConnect(duckdb(), sdm_db)
```

## Step 0: Create taxon_model from tmp_taxon.csv

The `taxon_model` junction table (taxon_id, ds_key, mdl_seq) was defined in
`merge_models.qmd` but never written to DB. Reconstruct it from `data/tmp_taxon.csv`
which has per-dataset `mdl_seq` columns.

```{r}
#| label: create_taxon_model

# read tmp_taxon.csv (has per-dataset mdl_seq columns)
d_taxon_full <- read_csv(here("data/tmp_taxon.csv"))

# pivot to junction table: (taxon_id, ds_key, mdl_seq)
ds_cols <- c("am_0.05", "ch_nmfs", "ch_fws", "rng_fws", "bl")
d_taxon_model <- d_taxon_full |>
  select(taxon_id, any_of(ds_cols)) |>
  pivot_longer(
    -taxon_id,
    names_to = "ds_key",
    values_to = "mdl_seq"
  ) |>
  filter(!is.na(mdl_seq))

# write to DB
if (dbExistsTable(con_sdm, "taxon_model")) {
  dbExecute(con_sdm, "DROP TABLE taxon_model")
}
dbWriteTable(con_sdm, "taxon_model", d_taxon_model, overwrite = TRUE)
message(glue("wrote {nrow(d_taxon_model)} rows to taxon_model"))
# wrote 17668 rows to taxon_model

# verify
dbGetQuery(
  con_sdm,
  "SELECT ds_key, count(*) AS n FROM taxon_model GROUP BY ds_key ORDER BY ds_key"
) |>
  print()
#    ds_key     n
# 1 am_0.05 16935
# 2      bl   571
# 3  ch_fws    28
# 4 ch_nmfs    33
# 5 rng_fws   101
```

## Step 1: Remap ch/rng model_cell values to er_score scale

Remap old ad-hoc values to match the `compute_er_score()` scale: EN=100, TN=50, LC=1.

```{r}
#| label: remap_model_cell_values

# check current values before remapping
d_before <- dbGetQuery(
  con_sdm,
  "
  SELECT m.ds_key, mc.value, count(*) AS n
  FROM model_cell mc
  JOIN model m ON mc.mdl_seq = m.mdl_seq
  WHERE m.ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
  GROUP BY m.ds_key, mc.value
  ORDER BY m.ds_key, mc.value"
)
message("before remapping:")
print(d_before)
# before remapping:
#    ds_key value      n
# 1  ch_fws    70  51655
# 2  ch_fws    90    150
# 3 ch_nmfs    70 188503
# 4 ch_nmfs    90  56886
# 5 rng_fws    50   9459
# 6 rng_fws    70 377995
# 7 rng_fws    90 888062

# only run if values are as expected
if (all(d_before$value %in% c(50, 70, 90))) {
  message("model_cell values are OLD — proceeding with remapping")

  # remap: 90->100 (EN), 70->50 (TN), 50->1 (LC, rng_fws only)
  n_updated <- dbExecute(
    con_sdm,
    "
  UPDATE model_cell SET value = CASE value
    WHEN 90 THEN 100
    WHEN 70 THEN 50
    WHEN 50 THEN 1
  END
  WHERE mdl_seq IN (
    SELECT mdl_seq FROM model
    WHERE ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
  )"
  )
  message(glue("remapped {n_updated} model_cell rows"))

  # verify
  d_after <- dbGetQuery(
    con_sdm,
    "
  SELECT m.ds_key, mc.value, count(*) AS n
  FROM model_cell mc
  JOIN model m ON mc.mdl_seq = m.mdl_seq
  WHERE m.ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
  GROUP BY m.ds_key, mc.value
  ORDER BY m.ds_key, mc.value"
  )
  message("after remapping:")
  print(d_after)
  # after remapping:
  #    ds_key value      n
  # 1  ch_fws    50  51655
  # 2  ch_fws   100    150
  # 3 ch_nmfs    50 188503
  # 4 ch_nmfs   100  56886
  # 5 rng_fws     1   9459
  # 6 rng_fws    50 377995
  # 7 rng_fws   100 888062
} else if (all(d_before$value %in% c(1, 50, 100))) {
  stop("model_cell values are NEW — aborting remapping")
} else {
  stop("unexpected model_cell values — aborting")
}

stopifnot(all(d_after$value %in% c(1, 50, 100)))
```

## Step 2: Recompute merged model cells for affected taxa

For taxa with merged models (n_ds > 1) that include ch/rng data, recompute cell
values as `max(all source values)` since the source ch/rng values changed.

```{r}
#| label: recompute_merged_models

# check if ms_merge models exist (they are created by merge_models.qmd merge loop)
n_merge <- dbGetQuery(
  con_sdm,
  "SELECT count(*) AS n FROM model WHERE ds_key = 'ms_merge'"
)$n

if (n_merge == 0) {
  message("no ms_merge models found in model table — skipping recompute")
} else {
  # count affected taxa
  d_affected <- dbGetQuery(
    con_sdm,
    "
    SELECT count(DISTINCT t.taxon_id) AS n_taxa
    FROM taxon t
    WHERE t.mdl_seq IS NOT NULL
      AND t.n_ds > 1
      AND t.taxon_id IN (
        SELECT DISTINCT taxon_id FROM taxon_model
        WHERE ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
      )"
  )
  message(glue("recomputing merged models for {d_affected$n_taxa} taxa"))
  # recomputing merged models for 87 taxa

  n_updated <- dbExecute(
    con_sdm,
    "
    UPDATE model_cell mc SET value = src.max_val
    FROM (
      SELECT t.mdl_seq AS merged_seq, all_src.cell_id,
             MAX(all_src.value) AS max_val
      FROM taxon t
      JOIN taxon_model tm ON t.taxon_id = tm.taxon_id
      JOIN model_cell all_src ON tm.mdl_seq = all_src.mdl_seq
      WHERE t.mdl_seq IS NOT NULL
        AND t.n_ds > 1
        AND t.taxon_id IN (
          SELECT DISTINCT taxon_id FROM taxon_model
          WHERE ds_key IN ('ch_nmfs', 'ch_fws', 'rng_fws')
        )
      GROUP BY t.mdl_seq, all_src.cell_id
    ) AS src
    WHERE mc.mdl_seq = src.merged_seq
      AND mc.cell_id = src.cell_id"
  )
  message(glue("updated {n_updated} merged model_cell rows"))
  # updated 5548640 merged model_cell rows
}
```

## Step 3: Listing join

Populate `extrisk_code`, `er_score`, `is_mmpa`, `is_mbta`, `is_bcc` on the `taxon` table
from the `listing` table. Includes a secondary join by `botw_id` for BOTW-only birds.

```{r}
#| label: listing_join

# add columns if not exist
for (col_sql in c(
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS extrisk_code VARCHAR",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS er_score INTEGER",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS is_mmpa BOOLEAN DEFAULT FALSE",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS is_mbta BOOLEAN DEFAULT FALSE",
  "ALTER TABLE taxon ADD COLUMN IF NOT EXISTS is_bcc  BOOLEAN DEFAULT FALSE"
)) {
  dbExecute(con_sdm, col_sql)
}

# clear any existing values so we start fresh
dbExecute(
  con_sdm,
  "
  UPDATE taxon SET
    extrisk_code = NULL, er_score = NULL,
    is_mmpa = FALSE, is_mbta = FALSE, is_bcc = FALSE"
) # 17561

# primary: join by worms_id (covers non-birds + birds with worms mapping)
n_worms <- dbExecute(
  con_sdm,
  "
  UPDATE taxon SET
    extrisk_code = l.extrisk_code,
    er_score     = l.er_score,
    is_mmpa      = l.is_mmpa,
    is_mbta      = l.is_mbta,
    is_bcc       = l.is_bcc
  FROM listing l
  WHERE taxon.worms_id = l.worms_id"
)
message(glue("listing join by worms_id: {n_worms} taxa matched"))
# listing join by worms_id: 491 taxa matched

# secondary: join by botw_id for BOTW-only birds not matched above
n_botw <- dbExecute(
  con_sdm,
  "
  UPDATE taxon SET
    extrisk_code = l.extrisk_code,
    er_score     = l.er_score,
    is_mmpa      = l.is_mmpa,
    is_mbta      = l.is_mbta,
    is_bcc       = l.is_bcc
  FROM listing l
  WHERE taxon.taxon_authority = 'botw'
    AND taxon.taxon_id = l.botw_id
    AND taxon.extrisk_code IS NULL"
)
message(glue("listing join by botw_id: {n_botw} taxa matched"))
# listing join by botw_id: 2 taxa matched

# IUCN fallback for unmatched taxa
n_iucn <- dbExecute(
  con_sdm,
  "
  UPDATE taxon SET
    extrisk_code = 'IUCN:' || redlist_code,
    er_score     = CASE redlist_code
      WHEN 'CR' THEN 50
      WHEN 'EN' THEN 25
      WHEN 'VU' THEN 5
      WHEN 'NT' THEN 2
      ELSE 1 END
  WHERE extrisk_code IS NULL AND redlist_code IS NOT NULL"
)
message(glue("IUCN fallback: {n_iucn} taxa"))
# IUCN fallback: 6640 taxa

# default for taxa with no listing and no redlist_code
n_default <- dbExecute(
  con_sdm,
  "
  UPDATE taxon SET er_score = 1
  WHERE er_score IS NULL"
)
message(glue("default er_score=1: {n_default} taxa"))
# default er_score=1: 10428 taxa

# verify no nulls remain
n_null <- dbGetQuery(
  con_sdm,
  "SELECT count(*) AS n FROM taxon WHERE er_score IS NULL"
)$n
stopifnot(n_null == 0)

# summary by authority
tbl(con_sdm, "taxon") |>
  mutate(
    authority = case_when(
      str_starts(extrisk_code, "NMFS") ~ "NMFS",
      str_starts(extrisk_code, "FWS") ~ "FWS",
      str_starts(extrisk_code, "IUCN") ~ "IUCN",
      TRUE ~ "none"
    )
  ) |>
  count(authority) |>
  collect() |>
  arrange(desc(n)) |>
  print()
#   authority      n
#   <chr>      <dbl>
# 1 none      10,428
# 2 IUCN       6,640
# 3 FWS          259
# 4 NMFS         234

# summary by taxon_authority
tbl(con_sdm, "taxon") |>
  filter(!is.na(extrisk_code)) |>
  count(taxon_authority) |>
  collect() |>
  print()
#   taxon_authority     n
#   <chr>           <dbl>
# 1 botw              614
# 2 worms            6519
```

## Step 4: Re-export taxon summary CSV

```{r}
#| label: taxon_summary

taxon_csv <- here("data/taxon.csv")

d_taxon <- tbl(con_sdm, "taxon") |>
  select(
    is_ok,
    component = sp_cat,
    common_name,
    scientific_name,
    redlist_code_max = redlist_code,
    extrisk_code,
    er_score,
    is_mmpa,
    is_mbta,
    is_bcc,
    worms_is_marine,
    worms_is_extinct,
    n_datasets = n_ds,
    taxon_authority,
    taxon_id,
    model_id = mdl_seq
  ) |>
  arrange(desc(is_ok), component, common_name) |>
  collect()

write_csv(d_taxon, taxon_csv)
message(glue("wrote {nrow(d_taxon)} rows to {taxon_csv}"))
# wrote 17561 rows to /Users/bbest/Github/MarineSensitivity/workflows/data/taxon.csv
```

## Cleanup

```{r}
#| label: cleanup

dbDisconnect(con_sdm, shutdown = TRUE)
```

## Next steps

Re-run `merge_models.qmd` interactively in Positron starting from chunk
`notbirds_wormsid_ds_duplicates` (line 795). This rebuilds `taxon` (without
per-dataset columns) and `taxon_model` (junction table with all dataset
mdl_seq values including `ca_nmfs` and `rng_iucn`), re-runs the merge loop
to create `ms_merge` models, populates listing/ESA/is_ok fields, and exports
`data/taxon.csv`.

Then re-run `calc_scores.qmd` starting from chunk `delete_extrisk_metrics`
(line 472):

1. `delete_extrisk_metrics` -- clear old metrics
2. `calc_cell_metric_redlist` -- recompute cell-level extinction risk
3. `calc_ecoregion_minmax` through `cell_metric_score` -- rescale and aggregate
4. `zone_taxon` -- per-zone species sensitivity with updated er_score
